### 2021-12-06 paper_overleaf/main.tex
paper latex files added

\documentclass[12pt, letterpaper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage[super]{nth}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage[
singlelinecheck=false
]{caption}
\usepackage{amsmath}
\usepackage{bbm}
\title{Method to infer relatedness using low coverage ancient DNA}
\author{Divya Ratan Popli, Benjamin M. Peter}
\date{5 August 2021}
\linenumbers

\setlength{\parskip}{1em}
\setlength{\parindent}{0em}

\newcommand{\BZ}{\mathbf{Z}}
\newcommand{\BD}{\mathbf{D}}
\newcommand{\BH}{\mathbf{H}}

\begin{document}

\maketitle

\begin{abstract}
\noindent Knowledge of familial relationships is an important part of several research fields. However, estimating relatedness from ancient DNA can be be difficult since DNA sequences extracted from ancient bones may have low coverage, ascertainment bias (if the sequences have been captured with DNA probes), and contamination from modern humans. In addition, the population of interest may have long runs of homozygosity (ROH) due to recent inbreeding or/and small population size. This can affect the relatedness estimates. Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and pairwise relatedness upto 3rd degree, while additionally differentiating between siblings and parent-child. We have developed another HMM to estimate ROH proportions, that can be utilized by the relatedness estimation model to output better estimates. 
\end{abstract}

\section{Introduction}

\subsection{Why study relatedness?}
Identifying related individuals is a common task in applications of genetics. Relatedness is of direct interest in e.g. DNA forensics, where familial search can aid in solving criminal cases, and to identify missing persons following a disaster \cite{murphy_law_2018,ram_genealogy_2018}. Genetic paternity tests have an important application in resolving family relation, e.g. in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}. It is also an essential preprocessing step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.

In ancient DNA, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an  understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_kinship-based_2019,sikora_ancient_2017}. For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals in the mating animals is important to ensure diversity. ~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}

\subsection{Simple way to get relatedness, methods that do it with a lot of data}
Commonly, pairs of relateded individuals are  identified by looking for parts of the genome that are identical by descent (IBD), ie. inherited from a recent common ancestor. Due to the laws of Mendelian segregation, a parent, for example, will share exactly one set of chromosome IBD with its offspring, with the other set of chromosomes in the offspring coming from the other parent. A grandparent will, on average share a quarter of its genome with a grand-child, because recombination broke up the chromosomes.

However, it is not possible to directly observe IBD segments, and so most methods infer them by first identifying segments of the genome that are Identical by State (IBS) and using population allele frequencies to calculate the probability of IBD given IBS \cite{vai_kinship_2020}.  One exception is SNPduo, a software that uses IBS direclty to visualize and analyze pairwise relatedness. \cite{roberson_visualization_2009}. Most methods, however, use IBD as an intermediary in inferring relatedness \cite{boehnke_accurate_1997,lynch_estimation_1999,mcpeek_statistical_2000}. For a non-inbred population, there are three IBD states possible: either no chromosomes are shared, only one chromosome is shared, or both chromosomes are shared. The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals. For example, a pair of siblings are expected to have all three possible IBD states with probabilities (0.25,0.5,0.25) as shown in fig 1. These IBD probabilities can directly be used to categorize a pair of individuals as related by a certain degree and nature of relatedness as shown in table1. One can also use these probabilities to estimate the coefficient of relatedness as $r= k_1/2 + k_2$. The coefficient of relatedness (r) is defined as the proportion of genome IBD for a pair of individuals.
There are numerous methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to estimate the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}. PLINK, a popular tool to infer relatedness coefficient in  diploid genotype data, estimates IBD probabilities from average of observed IBS, and using the allele frequencies at each SNP in samples \cite{purcell_plink_2007}. KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}. .

\begin{table}
\caption{\label{tab:Table 1}IBD sharing probabilities for different relations}
\begin{tabular}{|c|c|c|c|}
    \hline
    Relatedness & $k_0$ & $k_1$ & $k_2$\\
    \hline
    Unrelated & 1 & 0 & 0\\
    \hline
    3rd Degree & 0.75 & 0.25 & 0\\
    \hline
    2nd Degree & 0.5 & 0.5 & 0\\
    \hline
    Siblings & 0.25 & 0.5 & 0.25\\
    \hline
    Parent-Child & 0 & 1 & 0\\
    \hline
    Identical/twins & 0 & 0 & 1\\
    \hline
\end{tabular}

\end{table}

\subsection{Problems with ancient data, and methods that address these issues}
One major issue with applying above mentioned methods to ancient DNA is that, frequently, the amount of endogenous DNA is very low, making it very difficult to get genotype calls.
Several methods solve this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}. These methods account for the uncertainity in genotype calls by summing over all possible genotypes, weighted by the genotype likelihoods. However, these approaches typically still require at least 2x coverage, since genotype likelihoods may not be very informative at lower coverages. Apart from the low amount of endogenous DNA sequences present, ancient DNA analyses often have additional problems like contamination from modern populations \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in case of captured sequences. In cases where allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations. However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}. Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4x$ coverage to get a good estimate of genotype likelihoods\cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. In this study, we show that contamination from modern humans can cause some of the individuals in ancient populations to look more unrelated than they are. This happens because of comparison of endogenous reads to contaminating reads which would give higher differences than comparison of endogenous reads from individuals from ancient population. Hence, contamination can reduce the power to detect related pairs, especially when the target population is quite diverged from contaminating population(s). In addition, the target populations may have long runs of homozygosity (ROH) due to a small population size, or recent inbreeding. Long ROH causes related individuals to seem genetically closer to each other, while does not affect the genetic distance between unrelated individuals. This effect may cause an increase in false positives. In many cases, ancient DNA is captured with a SNP array, and the SNPs are ascertained to variable sites determined from certain individuals. Relatedness methods based on the fraction of sites in different IBS states are highly sensitive to ascertainment bias. KING robust, for example, differentiates between different relatedness by plotting KING‐robust kinship against fraction of sites in IBS0 \cite{manichaikul_robust_2010}. Another method from Rosenberg,2006 \cite{rosenberg_standardized_2006} uses the fraction of sites in all three IBS states to classify related individuals in the HGDP-CEPH Human Genome Diversity Cell Line Pane. These methods may give different results when applied to data with different ascertainment schemes \cite{waples_allele_2019}.

READ is a software that does not require population allele frequencies or a reference panel, and can work with around 0.5 x coverage in presence of ascertainment bias. The software solves the problem of unavailability of genotype calls by randomly sampling alleles from each individual. A string of these alleles at each position (called pseudohaploids) are then compared to other individuals to calculate pairwise genetic distances, which in turn are used to infer relatedness. The limitations are that the software can identify upto only second degree relatives, without differentiating between siblings and parent-child, which may be crucial when making pedigrees.

\subsection{How our method works}
Here, we present KIn (Kinship Inference), a Hidden Markov Model (HMM) to estimate relatedness from low-coverage ancient DNA data. KIn can detect up to \nth{3} degree relatives, and differentiates between siblings and parent-child. KIn is also able to incorporate ROH, contamination and ascertainment bias. We validate the performance of KIn using simulations and show that we are able to infer relatedness in an archaic human data set.

\section{Methods}

The core of  KIn, is a HMM that aims to infer relatedness and IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts in each sample, and contamination estimates into account. If the locations of ROH tracts are unknown, we provide another HMM to estimate ROH in the first step for a sample of sufficient coverage ($\geq 0.5x$). In a second step, the relatedness-HMM uses the contamination-corrected differences between a pair of individuals, and the positions of ROH tracts, to infer relatedness.  Our method is available on \url{https://github.com/DivyaratanPopli/KIn_snakemake} along with a \textt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.

\subsection{Relatedness Model overview} 
To estimate the relatedness of two individuals, we subdivide the genome into $L$ large windows $w$ (typically 10Mb). The inputs of our algorithm are i) the number of sites $N_w$ for which both samples have at least one read available, ii) the number of pairwise differences $D_w$ at these sites, and iii) position of ROH, by default obtained from ROH-HMM described in a following section.

Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$.

The model then uses this data for a given pair of individuals to classify each window into three hidden states $Z_w$ reflecting zero, one or two chromosomes shared IBD. Likewise, we classify each window into one of three possible ROH state $H_w$, reflecting, neither, one or both indivuals being in a run-of-homozygosity at this genomic location.

Ideally, we would like to compute the likelihood $P(\BD | \BZ, \BH, \theta$), where  $\theta$ is a vector of parameters (initial transition probability $\pi$, transition matrix $A$ and $N_w$).

However, calculating this likelihood is hard because the number of possible states for $\mathbf{Z}$ is very large, and we use a standard trick for fitting HMMs, to base or inference on the complete data likelihood:

\begin{align}
P(\mathbf{D},\BZ, \BH|\theta) &= P(\mathbf{D}|\mathbf{Z},\BH, \theta) P(\BZ |\theta)P(\BH |\theta) \nonumber\\
&= \prod_w P(D_w|Z_w,H_w, \theta) \prod_w P(H_w |\theta) \prod_w P(Z_w |Z_{w-1},\theta) P(Z_0|\theta)
\end{align}

Here, $P(D_w|Z_w,  H_w, \theta)$ is the emission probability for window $w$,  $P(Z_w|Z_{w-1},\theta)$ is the transition probability, and $P(Z_0| \theta)$ is the initial probability $\pi$. $P(H_w |\theta)$ is the output of ROH HMM described in a following section.

Using the complete data likelihood allows us to factor the emission and transition probabilities, and hence we can maximize them independently.
Transition matrix for different cases of relatedness are obtained from simulations (see section). Initial transition probability $\pi_i$ is same for all the $i$.

\subsection{Emission probability}
In equation \ref{eq:cll}  the term $P(D_w | Z_w, H_w, \theta) = P(D_w | Z_w, H_w, N_w)$ is the emission probability, i.e. the probability of our data , given a particular hidden state. The number of shared sites $N_w$ is the only parameter in $\theta$ that affects the emissions.

Assuming sites are equally distributed and independent, we could use a  binomial likelihood
$$P(D_w|Z_w, N_w, H_w) \sim \text{Binom}[D_w ;N_w, p(Z_w, H_w)] \text{,}$$

where, $p$  is the expected proportion of differences we would expect for a particular IBD and ROH state.

If the two individuals are unrelated in a particular window and neither is in a ROH region (i.e. $Z_w = H_w = 0$), then the expected proportion of pairwise differences depends solely on the effective population size, and we denote this proportion with $p_0$. If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes.

$p_0$, $p_1$, and $p_2$ are the expected proportion of differences in $i_0$, $i_1$, and $i_2$. $p_0$ is calculated directly from the data by calculating median of differences for all possible pairs of specimens. We calculate $p_1$ as $(3/4)p_1$ and $p_2$ as $p_1/2$. However, we need one more parameter (represented as $p_4$ in the table) in cases where IBD state is $i_1$ and ROH state is $\omega_2$, or IBD state is $i_2$ and ROH state is $\omega_1$ or $\omega_2$. In such cases the expected proportion of differences between the individuals should be 0. However, as mentioned before, we do all our calculations in windows of 10 Mb, and start/end positions of these windows may not coincide with that of ROH tracts. And so, the pairwise differences in this state (called $p_4$ in the table) can lie anywhere between 0 to $p_2$. We found $p_4$ = $p_2$/2 to be a reasonable assumption for expected pairwise proportion of differences in this IBD/ROH state.

Taken together, we can summarize $p$ as follows 
\begin{equation}\label{eq:p}
    p(Z_w, H_w) = \left[\begin{array}
{rrr}
p_0 & p_1 & p_2 \\
p_0 & p_2 & p_4 \\
p_0 & p_4 & p_4
\end{array}\right]
\end{equation}

The effect of these considerations is that even though we have nine possible combinations of $Z$ and $H$ for each window, there are actually only four different $p$-parameters.

However, we find that the data often has considerably higher variance than would be expected from a binomial distribution, likely because we average over a large number of SNPs with different allele frequencies.

We therefore add an overdispersion parameter $\delta$ that allows for variation in the $p$ between SNPs. This results in the betabinomial likelihood, (we show that such a distribution fits the data well (supplementary fig)).

\begin{equation}\label{eq:bbsimple}
P(D_{w}|p_i,\delta_i,N_w) = \binom{N_w}{D_w}\frac{B(D_w+p_i \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})}
\end{equation}

\subsubsection{Estimation}

Emission probabilities are estimated with Expectation-Maximization (EM) algorithm. In each iteration of the algorithm, the Expectation step involves forward-backward algorithm to calculate joint probability of the observed data and the IBD states $Z$, given an initial guess of $\delta$. In maximization step we use these joint probabilities to optimize $\delta$, and update emission probabilities. Iterations are run, until the complete data likelihood stops changing.

\paragraph{Initialization}
The value of $\delta$ is unknown to start with, and is randomly assigned, such that the mean of beta distribution and variance is within constraints described in maximization section.

\paragraph{Expectation step}

In this step, using the standard forward-backward algorithm, we calculate the posterior probability of each IBD state in each window $\gamma_{wi}$. Expectation step, among other parameters, requires emission probabilities. We calculate emission probabilities for each case of $Z_w$ and $H_w$ from $\delta$ and $p$ using equation 3, and we get the total emission probabilities for each $Z_w$ as $\sum_{H_w} P(D_w|Z_w,N_w,H_w) P(H_w|\theta)$

\paragraph{Maximization step}

The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_j$. We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_wi$ and the ROH state-probabilities and $h_{w\omega}$.

\begin{equation}
\mathcal{C} = \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta, p)h_{w\omega}\gamma_{wi}
\end{equation}

Using equation \ref{eq:p}, we simplify this by grouping all the terms that would result in the same $p$, i.e.

\begin{align*}
k_{w0} &= \gamma_{w0}\\
k_{w1} &= \gamma_{w1} h_{w0}\\
k_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\\
k_{w4} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2}
\end{align*}

So we can rewrite Eqn 4 as:
\begin{align}
\mathcal{C} &= \sum_{w=0}^L\sum_{j=0}^4  \log P(D_{w}|Z_w, H_w, N_w, \delta_j, p_j)k_{wj}
\end{align}

Since each term in this equation depends on only one $\delta_j$, we can optimize for them independently.

We observed that estimating the $\delta_j$ without constraints can result in wide Beta distributions with very high variance, leading to over fitting of data. For example, a parent  shares exactly one chromosome IBD with its child throughout the genome, so we expect only one IBD state. However,  running an unconstrained model could use this data to fit a pair of siblings with a very wide beta-distribution (with a very high $\delta_1$, which causes problems in differentiating between parent-child and siblings.

We avoid this problem by constraining the parameter space of $\delta$. 
In particular, 
Variance of beta distribution is shown below:
\begin{align}
    X \sim B(p,\delta)\\
    var(X) = \frac{p(1-p)}{\delta + 1}
\end{align}
We find the value of $\delta$ for which the variance is less than a threshold t.

We solve the following equation for $\delta$:
\begin{equation}
    \frac{p(1-p)}{\delta + 1}  < t
\end{equation}

We choose t for each beta distribution to be proportional to the mean ($p$), since the variances of beta distribution corresponding to different $x$ states follow the same relation as the distribution means (supplement figure). 
\begin{align}
    Var(X_{x_1}) = 3/4 Var(X_{x_0})\\
    Var(X_{x_2}) = 1/2 Var(X_{x_0})
\end{align}

Ideally, t for $x_4$ should be zero, but since our genomic windows may not coincide with the IBD fragments or ROH, the variance can be much higher. We find that fixing t for $x_4$ at the same value as $x_0$ works well.

\subsection{Relatedness classes}
This model is run for all the different cases of relatedness using the corresponding transition matrix, and the likelihood for all the relatedness models is compared. The relatedness for the maximum likelihood model is reported, and the confidence is given by the log likelihood ratio between the two highest likelihood models.  
Hidden state of each window (IBD state) is estimated using standard Viterbi algorithm, and the IBD states corresponding to the most likely model are returned.

\subsection{ROH estimation model}

Our HMM to detect ROH tracts works similar to the relatedness model described above. We sample all possible pairs of reads from each individual at all the positions where there are at least two reads, and calculate the proportion of read pairs that are different. These proportion of differences at each position are summed up in windows along the genome. This forms the input to the HMM, along with the number of overlapping sites, and the output is the posterior probability of a window being homozygous. This probability can also be inferred as the proportion of a window that has ROH. Our model has two hidden states: homozygous ($y_0$), and non homozygous ($y_1$). $Y_w$ denotes the state $y$ in a window $w$, and the vector of $Y_w$'s is denoted with $\mathbf{Y}$. The complete data likelihood for the model in this case is shown below:

\begin{align}
    P(\mathbf{\Delta},\mathbf{Y}|\Theta) &= P(\mathbf{\Delta}|\mathbf{Y},\Theta) P(\mathbf{Y}|\Theta)\nonumber\\
 &= [\prod_{w} P(\Delta_w|Y_w, \Theta) \prod_{w} P(Y_w|Y_w-1, \Theta)] P(Y_0| \Theta)
\end{align}

In this case, transitions are not known, and both transitions and emissions are estimated with standard EM algorithm. The emissions are calculated using betabinomial likelihood, and the mean of the distributions are constrained at expected proportion of differences in a ROH tract ($\rho_{0}$) and expected proportion of differences in a non-ROH tract $\rho_{1}$. Parameters $\rho_{1}$ and $\rho_{0}$ are the same as $p_{2}$ and $p_4$ in previous model.

\subsection{Emissions}
Similar to our relatedness model, we estimate emissions using EM algorithm with expectation and maximization steps. The expectation step outputs the posterior probability $\Gamma$ of being in $y_0$ and $y_1$ in each window. Maximization step uses the $\Gamma$ to optimize the $\Delta$ and $\rho$ parameters with a betabinomial cost function:

\begin{align}
\mathcal{C} = \sum_{w} \sum_{y} P(\Delta_w|\eta_{wy},\rho_{wy}) \Gamma_{wy} 
\end{align}

In case of very noisy data, we see high pairwise differences in some windows that are not explained with the beta binomial distributions estimated for $y_0$ and $y_1$. To avoid our distributions from flattening to explain these high differences, we force these high difference windows to $y_1$ (see supplement figure).

\subsection{Transitions}
Transitions are estimated using standard Baum-Welch algorithm. Expectation step is the same as that for emissions, and gives the posterior probability. In maximization step, we use the posterior probability to count how many times transitions are made from state i to j, normalized by the total number of times we are in state i. We observe that the final results are not affected by initial guess of the transition probabilities and assign a probability of 0.8 to stay in the same state for both $y_0$ and $y_1$.

\subsection{Contamination}

Contamination by present-day people is a common feature in ancient DNA \cite{peyregne_present-day_2020} . Even low levels of contamination may make samples look less similar, and thus distort relatedness analyses. This issue is particularly pronounced  when analyzing populations with  low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations (supplementary figure with READ?, lcMLkin at 4x). To address this issue, we  adjust the  pairwise differences $D_w$, given the contamination estimates of both samples and the average divergence between the target population and a putative contaminating population.

For each pair of individuals $i$ and $j$ we approximate the contamination $C_{i,j}$ as the sum of contamination for each individual $C_i + C_j$. Then, $C_{i,j}$ represents the probability of comparing a contaminant read from an individual to an endogenous read from the other individual. Throughout, we assume that contamination levels are low, so that we can ignore comparison between contaminant reads, i.e. $C_iC_j \approx 0$.  We denote the average divergence between target and contaminating populations as $p_c$, and the observed proportion of differences between individuals $i$ and $j$ as $p_{aij}$. We calculate average endogenous pairwise difference $p_eij$ for individuals i and j as follows \cite{noauthor_ancient_nodate}:
\begin{align}
    p_{aij} = C_{ij}  p_c + (1-C_{ij})  p_{eij}\\
    or, p_{eij} = \frac{p_{aij} - C_{ij}  p_c}{1-C_{ij}} 
\end{align}

Given $p_{eij}$, we would ideally like to calculate the complete data log likelihood for the relatedness model while summing over all possible differences we could have in a window weighted by the probability of seeing that difference.

\paragraph{Short version}
With contamination, we do not observe the number of differences $D$ and total number of comparisons directly $N$, but instead need to approximate them. The window-subscript is omitted in the following for clarity:
\begin{equation}
    P(D_{\text{obs}} | Z, N_\text{obs}, \theta) \approx 
    P( \hat{D_w} | \hat{N_w}, Z, \theta)
\end{equation}
We do that by calculating the probability that a single read is endogenous given it has a difference.
\begin{align}
    \hat{D} = \mathbb{E}[D | D_{\text{obs}}, c] &= 
    D_{\text{obs}} P(E_i | D_i=1) \\
    &= D_\text{obs} \frac{P(D_i=1 | E_i)P(E_i)}{P(D_i=1)}\\
    &= D_\text{obs} \frac{(1-c)p_e}{(1-c)p_e + c p_c}
\end{align}

and likewise for reads that do not have a difference
\begin{align}
    \hat{S} = \mathbb{E}[S | S_{\text{obs}}, c] &= 
     S_\text{obs} \frac{(1-c)(1-p_e)}{(1-c)(1-p_e) + c (1-p_c)}
\end{align}

and $\hat{N} = \hat{S} + \hat{D}$

\begin{equation}
    P(D_{\text{obs}} | Z, N_\text{obs}, \theta) = \sum_n\sum_d P(D_{\text{obs}}, N_\text{obs} | D_w=d, N_w, C_i, C_j)\underbrace{P(D_w=d | N_w, Z, \theta)}_\text{old likelihood}
\end{equation}

i.e. we sum over all possible $D_w$, $N_w$. Instead, we use

\begin{align}
    P(D_{cor},Z|D,H,\theta,c,p_c) &= P(D_{cor}|Z,D,H,\theta,c,p_c) P(Z|D,H,\theta,c,p_c)\nonumber\\
    &= [\prod_{w} [\sum_\kappa P(D_{cor,w}|Z_w, H_w, \theta,c,p_c) P(D_{cor,w}=\kappa)] \prod_{w} P(Z_w|Z_{w-1}, \theta)] P(Z_0| \theta)
\end{align}

We realize that this calculation would be time consuming, and would further complicate the HMM. To avoid this, instead of using weighted sum of all possible $\kappa$, we use $E(D_{cor,w}|Z_w, H_w, \theta,c,p_c)$. This simplifies the equation 21 to :

\begin{align}
    P(D_{cor},Z|D,H,\theta,c,p_c) &= P(D_{cor}|Z,D,H,\theta,c,p_c) P(Z|D,H,\theta,c,p_c)\nonumber\\
    &= [\prod_{w} P(E(D_{cor,w})|Z_w, H_w, \theta,c,p_c) \prod_{w} P(Z_w|Z_{w-1}, \theta)] P(Z_0| \theta)
\end{align}

We calculate $E(D_{cor,w})$ as shown:

\begin{align}
    E[N_e] + E[N_c] = N\\
    E[D_e] + E[D_c] = D
\end{align}

Probability of comparing endogenous reads at a site where we see a difference:
$$P(\xi | N = 1,D = 1)=\frac{(1-c) p_e}{c p_c + (1-c) p_e} $$
Probability of comparing endogenous reads at a site where we see no difference:
$$P(\xi | N = 1,D = 0)=\frac{(1-c)(1-p_e)}{c (1-p_c) + (1-c) (1-p_e)} $$
Expectation of number of endogenous comparisons in a window:
\begin{align}
    E[\xi | N, D] = D E[\xi | N=1, D=1] + (N-D) E[\xi | N=1, D=0]\\
    = D P(\xi | N=1, D=1) + (N-D) P(\xi | N=1, D=0)
\end{align}

Total number of endogenous sites showing a difference in a window: 
$$D_{cor,w} = D_w* \frac{p_e(1-c)}{p_e(1-c) + c p_c} $$
Total number of endogenous sites showing 0 difference in a window: 
$$S_{cor,w} = S_w* \frac{(1-p_e)(1-c)}{(1-p_e)(1-c) + c (1-p_c)} $$
Total number of endogenous sites in a window:
$$N_{cor,w} = D_{cor,w} + S_{cor,w}$$

This model outputs the contamination corrected number of differences and overlapping sites in each window.

\subsection{Simulations}

\paragraph{ Simulations for estimation of transition matrix}
We use simulations both for estimating the transition matrices for all our models, and for testing and validating our algorithm. All simulations are performed using \textt{msprime} \cite{kelleher_efficient_2016}, followed by simulations of related indivudals using a predetermined pedigree (Figure S1).

For our simulations of background diversity, we simulated a haploid population (pop1) with constant effective size of 10,000 and sampled 16 haploids from 2500 generations ago to randomly combine into 8 unrelated diploid ancient individuals. For each individual, we simulated 22 chromosomes with length $L=100$ Mb. The mutation rate was set as $\mu= 10^{-8}$ per base pair per generation and recombination rate used was $r=10^{-8}$

For artificially mating two individuals, we simulated a recombined chromosome from each individual to be passed on to the progeny. For simulating a recombined chromosome from a diploid individual, we drew the number of recombination points from a Poisson distribution with parameter $rL$ as the recombination rate, and used a uniform distribution to sample the positions of recombination points. We estimated transition matrices corresponding to different cases of relatedness by counting transition between IBD states for a pair of  individuals at a particular relatedness level. We realize that we have low power to detect $4^{th}$ and $5{th}$ degree and cna differentiate these from unrelated. In our final output, we show classifications of $4_{th}$ and $5_{th}$ degrees as unrelated.

\paragraph{Model Comparison}
model with maximum likelihood is reported, and log likelihood ratio is calculated with the two highest likelihood models.

#### Simulations for model evaluation
Apart from the related and unrelated individuals in pop1, we simulated more haploids in three other populations to create scenarios with ascertainment and contamination. We simulated two haploids to form an individual each from two other populations (pop2, pop3) with split time of 3500 and 4500 generations with pop1, and sampling time of 4000 generations and 2000 generations ago respectively. We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from pop1. We test out the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a markov chain using transition matrix shown below:

$$\mathbf{A} = \left[\begin{array}
{rr}
1-10/l & 10/l \\
2/l & 1-2/l  \\
\end{array}\right]
$$

From the steps described above, we got genotypes of 17 individuals in pop1 in presence/absence of ROH and ascertainment. We further simulated 10 haploids (5 diploid individuals) from another population (pop4) with split time of 20,000 generations with pop1, sampled from the present time. We calculated the allele frequency for pop4 at each site, and used it to introduce contamination (with fixed contamination percentage for each individual ($<5\%$) in samples from pop1. We generated reads (derived/ancestral) for different genomic coverages ranging from 4x to 0.03x at each position for each individual. We specifically generated reads for four different cases: Ancestral-Endogenous, Ancestral-Contaminant, Derived-Endogenous, Derived-Contaminant. The number of reads for each case was sampled from a Poisson distribution with $\lambda$ calculated from the coverage, allele frequency or genotype (G $\in$ 0,0.5,1), and contamination estimate:

\begin{align}
    \lambda_{DE} = \zeta (1-c) G\nonumber\\
    \lambda_{AE} = \zeta (1-c) (1-G)\nonumber\\
    \lambda_{DC} = \zeta c \upsilon\nonumber\\
    \lambda_{AC} = \zeta c (1-\upsilon)
\end{align}

\section{Notation summary}
Notation used in the paper:

\begin{itemize}
\item W: Total number of windows
\item w: Index of window
\item $I$:IBD states base
\item $Z_w$: IBD state in window w
\item $\mathbf{Z}$: Vector of $Z_w$'s
\item $\omega$:ROH states base
\item $H_{w\omega}$: ROH state in $w$
\item $s$: SNP
\item $d_s$: pairwise difference at $s$
\item $\mathbf{R_{si}}$: Vector of reads mapped to snp s for individual $i$
\item $n_{si}$: Number of reads at SNP s for individual $i$
\item $a_{si}$: Number of reads with alternate allele at SNP s for individual $i$
\item $N_{wij}$: Number of SNPs overlapping for individuals $i$,$j$ in window $w$
\item $D_{w}$: Number of SNPs in window w where a pair of individuals differ
\item $\mathbf{D}$: Vector of $D_w$
\item $\pi$: Initial transition probability
\item $A$: Transition matrix
\item $theta$: vector of additional parameters ($\pi$, $A$ and $N_w$)
\item $\psi$: Matrix of parameters representing expected proportion of differences in different $I$ and $\omega$
\item $psi_{I\omega}$: Element on row $I$ and column $\omega$ of $\psi$ matrix
\item $x$: All combinations of $I$ and $\omega$ states with unique $\psi_{I\omega}$
\item $p$: Proportion of differences in $x$
\item $\alpha,\beta$: Parameters of Beta distribution
\item $\delta$: $\alpha + \beta$ for relatedness model
\item $\gamma_{wI}$: posterior probability of $I$ in $w$
\item $k_{wI}$: Posterior probability of $x$ in $w$ 
\item $\nu_{w}$: Allele frequency in window w
\item X: Random variable
\item r: Relatedness between individuals
\item $\theta$: Vector of initial probability of being in an IBD state $I_l$
\item $/mathbf{B}$: Beta function
\item $B$: Reparameterized Beta function
\item $Delta$: Heterozygosity in an individual
\item Y: Hidden Homozygosity state
\item $\Theta$: vector of additional parameters ($\pi$ $N_w$)
\item $Gamma_{wY}$: Posterior probability of Y in w
\item $/eta$: $/alpha + /beta$ for ROH model
\item $c$: Contamination estimate for a library
\item $p_c$: Average proportion of differences between endogenous and contaminating reads
\item $p_e$:Average proportion of differences between endogenous reads
\item $p_a$: Average observed proportion of differences
\item $D_{cor,w}$ Contamination corrected number of differences in $w$
\item $N_{cor,w}$ Contamination corrected number of overlapping sites in $w$
\item $\Xi$: Endogenous read
\item $S_{cor,w}$ $N_{cor,w}$ - $D_{cor,w}$
\item $/mathbf{D_cor}$ Vector of $D_{cor,w}$ 
\item E: Expectation
\item P: Probability
\end{itemize}

\bibliographystyle{plain}
\bibliography{KIn_ref.bib}

\section{Results}
We predict the following categories of relatedness with KIn: Identical individuals (twins), First degree (parent-child), First degree (siblings), Second degree, Third degree, Unrelated ( including higher than third degree relatives). We apply KIn ROH model on simulated data with different coverages:

\section{Discussion}

We show that READ outputs higher false positives in case of recent inbreeding in the population, and higher false negatives in case of contamination.

\section{extra}
Many methods exist to estimate relatedness. Most of these work with high coverage, phased data, reference panel, population allele frequency. Other methods that work with low coverage ancient data use likelihoods, but still require 2x coverage.
methods that require a lot of things: PLINK, SNPduo,ERSA,KING,REAP,GRAB,

methods that require low coverage: SEEKIN

mtDNA and Y in relatedness READ paper[15,16,25,16]
snp data better than str when there is damage, low data
excluding related individuals, haak indo europian

READ is a very popular software to estimate relatedness in low coverage ancient DNA. It calculates distance between a pair of individuals using pseudo haploid sequences. These pairwise differences are then normalized by the median of pairwise differences in the population, and compared to the expected pairwise difference given a degree of relatedness. Expected pairwise difference for each case of relatedness is calculated using identity by descent (IBD) probabilities.

lcMLkin is another widely used method that uses genotype likelihoods to estimate the number of positions at which a pair of individuals have 0/1/2 chromosomes in IBD.

REAP is a software to estimate relatedness coefficient and IBD prbabilities in admixed populations, accounting for structure. It requires the number of ancestral populations, and a representation of sub population allele frequencies.

Likelihood estimators:
Milligan BG (2003) Maximum-likelihood estimation of relatedness. Genetics
Anderson AD, Weir BS (2007) A maximum-likelihood method for the estimation of pairwise relatedness in structured populations. Genetics
Choi Y, Wijsman EM, Weir BS (2009) Case-control association testing in the presence of unknown relationships

There are various ways to measure the coefficient of relatedness r. Most methods use the probabilties of IBD to calculate r. There can be, in total 9 modes of identity described by Jacquard (1974), including the cases for 3 IBD probabilities and 3 cases of IBD. Methods assuming outbred population use only the three modes accounting for IBD. Most methods used maximum likelihood approach to estimate r (). Other methods are moment estimators that assume Hardy-Weinberg equilibrium in the population (KING, GCTA).

#Methods
$$\sum_{w} \sum_{I}\sum_\omega \log P(D_{w}|Z_w, H_w)k_{w\omega i}$$

$$P(D_w | N_w, Z_w) = \sum_{k=0}^4 
\underbrace{P(D_w | N_w, k)}_{Betabinomial}
\underbrace{P(k | Z_w, H_w)}_{p matrix}
\underbrace{P(H_w =h)}_{\mathbf{h}_w}
$$

Emission probability in a window depends on both IBD state and ROH state. We obtain posterior probability of one or both individuals being ROH in each window from another HMM described in section 3.2. We calculate the cost function with Betabinomial likelihood in log space for the entire genome as:

The model can be summarised as:

\begin{equation}\label{eq:6}
P(D_w|Z_w,H_w) \sim {\sf Binom}(D_w, N_w, p_{Z,H,w})
\end{equation}

\begin{equation}\label{eq:7}
p_{Z,H,w} \sim {\sf Beta}(\alpha_{Z,H}, \beta_{Z,H})
\end{equation}

Emission probability in a window depends on both IBD state and ROH state. We calculate a cost function as the betabinomial loglikelihood weighted by the posterior probability represented by \gamma.
likelihood:
\begin{align}
P(\mathbf{D},\BZ, H|\theta) &= P(\mathbf{D}|\mathbf{Z},H,\theta) P(H |\theta) P(\BZ|\theta)\nonumber\\
&= \prod_{w}  P(D_w|Z_w,  H_w, \theta) \prod_w P(H_w | \theta) \prod_{w} P(Z_w|Z_{w-1}, \theta)] P(Z_0| \theta) 
\end{align}

cost function:
\begin{equation}
    \mathcal{C} = 
       \sum_{w} \sum_{i} \log P(D_{w}|Z_w, H_w)\gamma_{wi}=
    \sum_{w} \sum_{i} 
    \log \left[\sum_{\omega}P(D_{w}|\psi_{i\omega},\delta_{i\omega}) \right] \gamma_{wi}
\end{equation}

\end{document}

### 2021-12-07 paper_overleaf/main.tex
added text to result section

@@
 \usepackage{bbm}
+\usepackage{graphicx}
 \title{Method to infer relatedness using low coverage ancient DNA}
@@
 
-\bibliographystyle{plain}
-\bibliography{KIn_ref.bib}
+\section{Results}
+\graphicspath{{plots/plotimg/}}
+We predict the following categories of relatedness with KIn: Identical individuals (twins), First degree (parent-child), First degree (siblings), Second degree, Third degree, Unrelated (including higher than third degree relatives).
+We show the working of Relatedness HMM with fig.1 showing performance of each model, when applied to a pair of siblings.
+We see that the siblings model predictions match true IBD states the most.
+This is reflected in the log likelihoods values of the models as well, where siblings have the highest value shown in table 3.
+We take a loglikelihood ratio of the two most likely models to get a confidence in the relatedness estimate.
+All the models in fig.1 utilize information about ROH given by the ROH-HMM.
+Performance of ROH-HMM under different cases of ascertainment, contamination and long ROH, and for different coverages is shown in fig 2.
+We applied the ROH-HMM and relatedness HMM on 60 runs of simulations, to get relatedness estimates, and compared the final relatedness estimates to the true relatedness for different pairs at coverages ranging from 4x to 0.1x.
+We plotted the true positive and false positive rates with a cutoff on the loglikelihood ratio in fig.3.
+Here, and in fig.4, we have two different relatedness labels: 'Unrelated' and 'Unrelated w/o 3rd Degree'.
+Unrelated label here refers to KIn performance results when all unrelated, degree5, degree4, degree3 pairs are labelled as unrelated.
+This is useful for comparison to READ, since READ classifies relatedness upto degree2.
+'Unrelated w/o degree3'  refers to the performance results when degree 3 is considered as separate from the unrelated individuals.
+Fig.3 shows that both for identical, 1st degree, unrelated pairs, the false positive rate is below $5\%$ even when no cut off is used.
+For second degree, at 0.1x we see a high false positive rate of under 0.25, and does not go down much with an increase in cutoff.
+For the same relationship, a cutoff of 1 seems appropriate to reduce false positive rate to much below $5\%$.
+Similarly, for siblings at 0.1x, a cutoff of 1 works well to keep false positives low.
+For parent-child, we find that a cutoff of about 2 is a better choice.
+Degree 3 shows higher false positive rates in general, and we see that only for 4x coverage and a cutoff of 2, do we see a false positive rate of below $5\%$.
+In conclusion, it is recommended to use a cutoff of 1 for most cases, with the exception of differentiation between paret-child and siblings, where a cutoff of 2 may be more appropriate, and degree 3 where cutoff of 2 at 4x coverage seems reasonable.
+We use a cutoff of 1 to compare our method performance to READ, while using a cutoff of 1 standard deviation for READ.
+Fig.4 shows this comparison using false positive and true positive rates for both the mothods at different coverages under different simulations.
+We show that both methods have similar performance in the control case (except for 0.1x coverage, where KIn has higher power to detect degree1 and degree2, and higher false positive rate for degree 2).
+However, in case of ascertainment, ROH, or comtamination KIn shows higher power to detect relatedness cases while maintaining a low false positive rate.
+To test KIn on real ancient dataset, we apply it to look for neandertal specimens from Chagyrskaya cave in Siberia, Russia.
+This dataset comprised of 14 specimens that came from neighbouring layers, and hence could possibly belong to contemporary neandertals who occupied the cave 63 ± 4 to 48 ± 3 ka [\cite{}]
 
-\section{Results}
-We predict the following categories of relatedness with KIn: Identical individuals (twins), First degree (parent-child), First degree (siblings), Second degree, Third degree, Unrelated ( including higher than third degree relatives).
-We apply KIn ROH model on simulated data with different coverages:
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{IBDplot.png}
+    \caption{IBD fragments detected for a pair of simulated siblings (proportion of differences in each window shown on top) by the relatedness HMM with different models.}
+    \label{fig:ibd}
+\end{figure}
+
+\begin{table}
+\caption{\label{tab:Table 3}Log likelihoods for different relatedness models}
+\begin{tabular}{|c|c|c|c|c|c|c|}
+    \hline
+    Relatedness & $Identical$ & $Parent-Child$ & $Siblings$ & $Degree2$ & $Degree3$ & $Unrelated$\\
+    \hline
+    Loglikelihood & -1478.24 & -1259.44 & -1213.61 & -1236.09 & -1261.19 & -1281.77\\
+    \hline
+
+\end{tabular}
+
+\end{table}
+
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{ROHplot.png}
+    \caption{ROH stuff}
+    \label{fig:galaxy}
+\end{figure}
+
+And then cutoff:
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{contam0_inbred0_model_performance_allroc_asc0_plot.png}
+    \caption{cutoff stuff}
+    \label{fig:galaxy}
+\end{figure}
+
+And then comparison with read:
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{comparison_plot.png}
+    \caption{comparison stuff}
+    \label{fig:galaxy}
+\end{figure}
+
+Comparison with Chagyrskaya:
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{fil0_relatable_plot.png}
+    \caption{Chagyrskaya stuff}
+    \label{fig:galaxy}
+\end{figure}
 
@@
 \end{document}
+
+\bibliographystyle{plain}
+\bibliography{KIn_ref.bib}

### 2021-12-08 paper_overleaf/main.tex
references with zotero working again

@@
 To test KIn on real ancient dataset, we apply it to look for neandertal specimens from Chagyrskaya cave in Siberia, Russia.
-This dataset comprised of 14 specimens that came from neighbouring layers, and hence could possibly belong to contemporary neandertals who occupied the cave 63 ± 4 to 48 ± 3 ka [\cite{}]
+This dataset comprised of 14 specimens that came from neighbouring layers, and hence could possibly belong to contemporary neandertals who occupied the cave 63 ± 4 to 48 ± 3 ka \cite{kolobova_archaeological_2020-1}
 
@@
 We show that READ outputs higher false positives in case of recent inbreeding in the population, and higher false negatives in case of contamination.
+
+\bibliographystyle{plain}
+\bibliography{references.bib}
 
@@
 
-Emission probability in a window depends on both IBD state and ROH state. We calculate a cost function as the betabinomial loglikelihood weighted by the posterior probability represented by \gamma.
+Emission probability in a window depends on both IBD state and ROH state. We calculate a cost function as the betabinomial loglikelihood weighted by the posterior probability represented by $\gamma$.
 likelihood:
@@
 \end{document}
-
-\bibliographystyle{plain}
-\bibliography{KIn_ref.bib}

### 2021-12-08 paper_overleaf/main.tex
added more to results

@@
 \section{Results}
-\graphicspath{{plots/plotimg/}}
+
 We predict the following categories of relatedness with KIn: Identical individuals (twins), First degree (parent-child), First degree (siblings), Second degree, Third degree, Unrelated (including higher than third degree relatives).
 We show the working of Relatedness HMM with fig.1 showing performance of each model, when applied to a pair of siblings.
-We see that the siblings model predictions match true IBD states the most.
-This is reflected in the log likelihoods values of the models as well, where siblings have the highest value shown in table 3.
-We take a loglikelihood ratio of the two most likely models to get a confidence in the relatedness estimate.
-All the models in fig.1 utilize information about ROH given by the ROH-HMM.
-Performance of ROH-HMM under different cases of ascertainment, contamination and long ROH, and for different coverages is shown in fig 2.
-We applied the ROH-HMM and relatedness HMM on 60 runs of simulations, to get relatedness estimates, and compared the final relatedness estimates to the true relatedness for different pairs at coverages ranging from 4x to 0.1x.
-We plotted the true positive and false positive rates with a cutoff on the loglikelihood ratio in fig.3.
-Here, and in fig.4, we have two different relatedness labels: 'Unrelated' and 'Unrelated w/o 3rd Degree'.
-Unrelated label here refers to KIn performance results when all unrelated, degree5, degree4, degree3 pairs are labelled as unrelated.
-This is useful for comparison to READ, since READ classifies relatedness upto degree2.
-'Unrelated w/o degree3'  refers to the performance results when degree 3 is considered as separate from the unrelated individuals.
-Fig.3 shows that both for identical, 1st degree, unrelated pairs, the false positive rate is below $5\%$ even when no cut off is used.
-For second degree, at 0.1x we see a high false positive rate of under 0.25, and does not go down much with an increase in cutoff.
-For the same relationship, a cutoff of 1 seems appropriate to reduce false positive rate to much below $5\%$.
-Similarly, for siblings at 0.1x, a cutoff of 1 works well to keep false positives low.
-For parent-child, we find that a cutoff of about 2 is a better choice.
-Degree 3 shows higher false positive rates in general, and we see that only for 4x coverage and a cutoff of 2, do we see a false positive rate of below $5\%$.
-In conclusion, it is recommended to use a cutoff of 1 for most cases, with the exception of differentiation between paret-child and siblings, where a cutoff of 2 may be more appropriate, and degree 3 where cutoff of 2 at 4x coverage seems reasonable.
-We use a cutoff of 1 to compare our method performance to READ, while using a cutoff of 1 standard deviation for READ.
-Fig.4 shows this comparison using false positive and true positive rates for both the mothods at different coverages under different simulations.
-We show that both methods have similar performance in the control case (except for 0.1x coverage, where KIn has higher power to detect degree1 and degree2, and higher false positive rate for degree 2).
-However, in case of ascertainment, ROH, or comtamination KIn shows higher power to detect relatedness cases while maintaining a low false positive rate.
-To test KIn on real ancient dataset, we apply it to look for neandertal specimens from Chagyrskaya cave in Siberia, Russia.
-This dataset comprised of 14 specimens that came from neighbouring layers, and hence could possibly belong to contemporary neandertals who occupied the cave 63 ± 4 to 48 ± 3 ka \cite{kolobova_archaeological_2020-1}
 
@@
     \centering
-    \includegraphics[width=18cm]{IBDplot.png}
+    \includegraphics[width=18cm]{plots/plotimg/IBDplot.png}
     \caption{IBD fragments detected for a pair of simulated siblings (proportion of differences in each window shown on top) by the relatedness HMM with different models.}
@@
 \end{figure}
+
+We see that in this case, the siblings model predictions match true IBD states the most.
+This is reflected in the log likelihoods values of the models as well, where siblings have the highest value shown in table 3.
+We take a loglikelihood ratio ($\Lambda$) of the two most likely models to get a confidence in the relatedness estimate (in this case $\Lambda$ = 22.48).
+All the models in fig.1 utilize information about ROH given by the ROH-HMM.
 
@@
 
+Performance of ROH-HMM under different cases of ascertainment, contamination and long ROH, and for different coverages is shown in fig 2.
+We applied the ROH-HMM and relatedness HMM on 60 runs of simulations, to get relatedness estimates, and compared the final relatedness estimates to the true relatedness for different pairs at coverages ranging from 4x to 0.1x.
+
 \begin{figure}[htp]
     \centering
-    \includegraphics[width=18cm]{ROHplot.png}
+    \includegraphics[width=18cm]{plots/plotimg/ROHplot.png}
     \caption{ROH stuff}
@@
 
-And then cutoff:
+We plotted the true positive and false positive rates with a cutoff on the loglikelihood ratio in fig.3.
+Here, and in fig.4, we have two relatedness labels: 'Unrelated' and 'Unrelated w/o 3rd Degree'.
+Unrelated label here refers to KIn performance results when all unrelated, degree5, degree4, degree3 pairs are labelled as unrelated.
+This is useful for comparison to READ, since READ classifies relatedness upto degree2.
+'Unrelated w/o degree3'  refers to the performance results when degree 3 is considered as separate from the unrelated individuals.
+Fig.3 shows that both for identical, 1st degree, unrelated pairs, the false positive rate is below $5\%$ even when no cut off is used.
+For second degree, at 0.1x we see a high false positive rate, that does not go down much with an increase in cutoff.
+For the same relationship, at a coverage of 0.2x, a cutoff of 1 seems appropriate to reduce false positive rate to below $5\%$.
+Similarly, for siblings at 0.1x, a cutoff of 1 works well to keep false positives low.
+For parent-child, we find that a cutoff of about 2 is a better choice.
+Degree 3 shows higher false positive rates in general, and we see that only for 4x coverage and a cutoff of 2, do we see a false positive rate of below $5\%$.
+In conclusion, it is recommended to use a cutoff of 1 for most cases, with the exception of differentiation between paret-child and siblings, where a cutoff of 2 may be more appropriate, and degree 3 where cutoff of 2 at 4x coverage seems reasonable.
+
 \begin{figure}[htp]
     \centering
-    \includegraphics[width=18cm]{contam0_inbred0_model_performance_allroc_asc0_plot.png}
+    \includegraphics[width=18cm]{plots/plotimg/contam0_inbred0_model_performance_allroc_asc0_plot.png}
     \caption{cutoff stuff}
@@
 
-And then comparison with read:
+We use a cutoff of 1 to compare our method performance to READ, while using a cutoff of 1 standard deviation for READ.
+Fig.4 shows this comparison using false positive and true positive rates for both the methods at different coverages under different simulations.
+We show that both methods have similar performance in the control case (except for 0.1x coverage, where KIn has higher power to detect degree1 and degree2, and higher false positive rate for degree 2).
+However, in case of ascertainment, ROH, or comtamination KIn shows higher power to detect relatedness cases while maintaining a low false positive rate.
+
 \begin{figure}[htp]
     \centering
-    \includegraphics[width=18cm]{comparison_plot.png}
+    \includegraphics[width=18cm]{plots/plotimg/comparison_plot.png}
     \caption{comparison stuff}
@@
 
-Comparison with Chagyrskaya:
+To test KIn on real ancient dataset, we applied it to look for neandertal specimens from Chagyrskaya cave in Siberia, Russia.
+This dataset comprised of 14 specimens that came from neighbouring layers, and hence could possibly belong to contemporary neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya \cite{kolobova_archaeological_2020-1}.
+These specimens showed low coverage of nuclear genome, with 8 samples out of 14 showing <1x coverage.
+Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena.
+All these samples were captured with an array targeting for Neanderthal variation, and we further ascertained this data to variable sites in high coverage Neanderthal genomes: Altai Neanderthal (Denisova 5) (Prüfer et al., 2014) and Vindija 33.19 (Prüfer et al., 2017).
+This dataset had all the issues that KIn deals with, and hence we decided to apply it here to see if we could find family relationships that would not be possible with other available methods.
+Our results for pairwise relatedness for these individuals is shown in fig.5.
+We found that identical individuals (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), first degree relatives (Chagyrskaya07-Chagyrskaya17, Chagyrskaya06-Chagyrskaya14), and a second degree relative (Chagyrskaya01-Chagyrskaya60).
+These results agree with READ (supplementary fig).
+However, we find that Chagyrskaya07-Chagyrskaya17 are parent-child.
+This matches with the conclusions of the authors based on additional information from mtDNA.
+Further, we find a possible third degree relative (Chagyrskaya07-Chagyrskaya60).
+We compare our results to lcMLkin, and show that the software does not perform well with these low coverage samples since it is not possible to obtain reliable genotype likelihoods for many of these samples (supplementary fig).
+
 \begin{figure}[htp]
     \centering
-    \includegraphics[width=18cm]{fil0_relatable_plot.png}
+    \includegraphics[width=18cm]{plots/plotimg/fil0_relatable_plot.png}
     \caption{Chagyrskaya stuff}

### 2021-12-13 paper_overleaf/main.tex
started supplementary info section
added figure legends in results

@@
     \includegraphics[width=18cm]{plots/plotimg/ROHplot.png}
-    \caption{ROH stuff}
+    \caption{Estimation of ROH probabilities along the genome.
+Top panel shows proportion of differences along the genome, and the bottom panel shows probability of no ROH.
+A.Simulation with no ascertainment, contamination, or ROH.
+B.
+Simulation with ROH C.
+Simulation with ROH and ascertainment D.
+Simulation with ascertainment}
     \label{fig:galaxy}
@@
     \includegraphics[width=18cm]{plots/plotimg/contam0_inbred0_model_performance_allroc_asc0_plot.png}
-    \caption{cutoff stuff}
-    \label{fig:galaxy}
+    \caption{False positive and negative rates as a function of cutoff on log likelihood ratio. Dotted line represents $5\%$$}
+    \label{fig3}
 \end{figure}
@@
     \includegraphics[width=18cm]{plots/plotimg/comparison_plot.png}
-    \caption{comparison stuff}
-    \label{fig:galaxy}
+    \caption{Comparison of KIn with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH}
+    \label{fig4}
 \end{figure}
@@
     \includegraphics[width=18cm]{plots/plotimg/fil0_relatable_plot.png}
-    \caption{Chagyrskaya stuff}
-    \label{fig:galaxy}
+    \caption{Application of KIn on Neandertal specimens from Chagyrskaya}
+    \label{fig5}
 \end{figure}
@@
 We show that READ outputs higher false positives in case of recent inbreeding in the population, and higher false negatives in case of contamination.
+
+\section{Supplementary Information}
+-binomial,betabinomial
 

### 2021-12-13 supplementary_info/Rscripts/binom.R
added supplementary_plot folder woth binom plot

@@ -1,0 +1,96 @@
+library(oompaBase)
+library(TailRank)
+library(ggplot2)
+library(cowplot)
+library(dplyr)
+library(tidyr)
+library(readr)
+
+#dfile <-'/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz_highDiversity/contam0/inbred0/run1/coverage4/asc0/inputMode_hapProbs_fil0/merged_wind.csv.gz'
+#tfile <-'/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz_highDiversity/contam0/inbred0/run1/coverage4/asc0/inputMode_hapProbs_fil0/merged_wint.csv.gz'
+#p1file<-'/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz_highDiversity/contam0/inbred0/run1/coverage4/asc0/inputMode_hapProbs_fil0/hmm_parameters/p1_file.gz'
+#deltafile0<-'/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/supplementary_info/datafiles/contam0_inbred0_run1_coverage4_asc0_inputMode_hapProbs_fil0_delta_0_1_un.csv.gz'
+#deltafile1<-'/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/supplementary_info/datafiles/contam0_inbred0_run1_coverage4_asc0_inputMode_hapProbs_fil0_delta_0_8_pc.csv.gz'
+#deltafile2<-'/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/supplementary_info/datafiles/contam0_inbred0_run1_coverage4_asc0_inputMode_hapProbs_fil0_delta_0_15_id.csv.gz'
+#mergePanels(dfile, tfile, p1file, deltafile0, deltafile1, deltafile2, '/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/supplementary_info/test.png')
+
+makepanelVar <- function(dfile, tfile, p1file, ind, pratio, deltaind,deltafile){
+
+  diff=read.csv(file = dfile, header = FALSE)[,ind]
+  total=read.csv(file = tfile, header = FALSE)[,ind]
+
+  deltalist=read.csv(file = deltafile, header = FALSE)
+
+  prop=diff/total
+  propdf=as.data.frame(prop)
+
+  p <- read_file(p1file)
[…]

### 2021-12-13 supplementary_info/delta_functions.py

@@ -1,0 +1,588 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Jan 21 21:36:58 2021
+
+@author: divyaratan_popli
+"""
+
+
+import scipy
+import numpy as np
+import random
+import pandas as pd
+from scipy.stats import binom
+from math import isclose
+from scipy.special import gammaln
+from scipy import optimize
+import math
+import matplotlib.pyplot as plt
+from scipy.special import loggamma
+from scipy.special import beta as Betaf
+from scipy.optimize import minimize_scalar
+import pylab
+from numpy.linalg import matrix_power
+from scipy.special import logsumexp
+import numba
+
+#print(scipy.__version__)
+
+
+
+"""
+#to stop the program if there is a warning
+import warnings
+
+with warnings.catch_warnings():
+    warnings.simplefilter('error')
+    for chrm in range(chrm_no):
+
+            for t in range(pos[chrm],pos[chrm+1]-1):
+
+
+                for i in range(M):
+                    print.info(chrm,t,i)
+                    numerator_l = np.log(alpha[t, i]) + np.log(A[i, :]) + np.log(B[:, t + 1].T) + np.log(beta[t + 1, :].T) - np.log(scale[t+1])  #it is the probability that hmm path has particular arc at time t
+                    xi_l[i, :, t] = numerator_l
+
+
+"""
+@numba.vectorize
+def Betal(x, y):
+    # numba doesn't understand scipy.special.gammaln, but it understands math.lgamma,
+    # which is the same, except that it doesn't understand arrays.
+    return math.lgamma(x) + math.lgamma(y) - math.lgamma(x+y)
+
+@numba.vectorize
+def fact(x):
+    return math.lgamma(x + 1)
+
+@numba.njit
+def forward(A,B,pi,data):
+    N = A.shape[0]
+    T = data.shape[0]
+    alpha = np.zeros((T,N))
+    alpha_p = np.zeros((T,N))
[…]

### 2021-12-13 supplementary_info/Rscripts/binom.R
expanded C in binom plot

@@ -85,8 +85,8 @@ mergePanels <- function(dfile, tfile, p1file, deltafile0, deltafile1, deltafile2
     theme_bw() + theme(axis.title.x = element_blank(), axis.title.y = element_blank())
 
 
-  plot_grid(panel0binom,panel1binom,panel2binom, labels= c("A","B","C"),
-            ncol=3, rel_widths = c(1,1,1), rel_heights =  c(1,1,1), align = "h", axis="b")
+    plot_grid(panel0binom,panel1binom,panel2binom, labels= c("A","B","C"),
+              ncol=3, rel_widths = c(1,1,1.45), align = "h", axis="b")
   ggsave(outf,
          width = 8, height = 5, dpi = 150, units = "in", device='png')
 }

### 2021-12-13 paper_overleaf/main.tex

@@
     \includegraphics[width=18cm]{plots/plotimg/fil0_relatable_plot.png}
-    \caption{Application of KIn on Neandertal specimens from Chagyrskaya}
+    \caption{Application of KIn on Neandertal specimens from Chagyrskaya. Color of a square represents the relatedness, while the number denotes loglikelihood ratio.}
     \label{fig5}
@@
 -binomial,betabinomial
+
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/binom.png}
+    \caption{Beta binomial distribution is a better fit compared to Binomial for proportion of difference in windows}
+    \label{figS1}
+\end{figure}
 

### 2021-12-16 supplementary_info/Rscripts/binom.R
added lcMLkin plot

@@ -86,7 +86,7 @@ mergePanels <- function(dfile, tfile, p1file, deltafile0, deltafile1, deltafile2
 
 
     plot_grid(panel0binom,panel1binom,panel2binom, labels= c("A","B","C"),
-              ncol=3, rel_widths = c(1,1,1.45), align = "h", axis="b")
+              ncol=3, rel_widths = c(1,1,1.5), align = "h", axis="b")
   ggsave(outf,
          width = 8, height = 5, dpi = 150, units = "in", device='png')
 }

### 2021-12-16 supplementary_info/Rscripts/lcPlot.R

@@ -1,0 +1,109 @@
+library(oompaBase)
+library(TailRank)
+library(ggplot2)
+library(cowplot)
+library(tidyr)
+library(tidyverse)
+library(plyr)
+
+#lcres <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/laurits_writeup/Ch9_rem/neavar/noCh12_9mar_newmodel/lcMLkin/output.relate"
+#likf <-"/mnt/diversity/divyaratan_popli/100arc/inbreeding/laurits_writeup/Ch9_rem/neavar/newSim_ibd/hmm_numba/filtered0/merged_relatable_allLikelihoods.csv"
+#kinf <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/laurits_writeup/Ch9_rem/neavar/newSim_ibd/hmm_numba/filtered0/IBDstates_allLikelihoods.csv"
+
+
+lcplot <- function(lcres,likf,kinf,outplot){
+
+  lc<- read.csv(file = lcres, header = TRUE,sep = "\t")
+  lik <-read.csv(file = likf, header = TRUE, sep = "\t")
+
+  lc %>%
+    separate(Ind1, c("name1", "bam1"), "\\.") %>% separate(Ind2, c("name2", "bam2"), "\\.") ->lc
+  lc <-unite(lc, pair, c(name1, name2), remove=FALSE)
+
+  drops <- c("name1","name2","bam1","bam2","nbSNP","pi_HAT")
+  lc <-lc[ , !(names(lc) %in% drops)]
+
+  lc <-merge(lik[c('pair','loglik_ratio')], lc, by = c("pair"))
+
+  lc$r_hat <- lc$k1_hat/2 + lc$k2_hat
+  lc$Relatedness <- 'Unrelated'
+  lc[lc$pair=='Chagyrskaya07_Chagyrskaya17','Relatedness'] <- 'Parent-Child'
+  lc[lc$pair=='Chagyrskaya13_Chagyrskaya19','Relatedness'] <- 'Identical'
+  lc[lc$pair=='Chagyrskaya1141_Chagyrskaya13','Relatedness'] <- 'Identical'
+  lc[lc$pair=='Chagyrskaya1141_Chagyrskaya19','Relatedness'] <- 'Identical'
+  lc[lc$pair=='Chagyrskaya01_Chagyrskaya60','Relatedness'] <- 'Degree2'
+  lc[lc$pair=='Chagyrskaya06_Chagyrskaya14','Relatedness'] <- 'Identical'
[…]

### 2021-12-20 supplementary_info/Rscripts/degree2_cluster.R
added plot function for degree 2

@@ -1,0 +1,53 @@
+library(ggplot2)
+library(tidyr)
+library(tidyverse)
+library(plyr)
+
+hsib=c('8_9','8_10')
+avu='9_12'
+gr=c('1_12','2_12','3_12','4_12','10_13','11_13','12_14','12_16','5_14')
+
+
+#fnames=c("/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz/hmm_numba_corr_IBD/contam0/inbred0/run1/coverage4/filtered0/asc0/IBDstates_allLikelihoods_inphapProbs.csv.gz", "/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz/hmm_numba_corr_IBD/contam0/inbred0/run2/coverage4/filtered0/asc0/IBDstates_allLikelihoods_inphapProbs.csv.gz")
+#outplot="/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz/test.png"
+deg2_plot <- function(fnames, outplot){
+
+  list_all <<- NULL
+  for (r in 1:60){
+
+
+    deg2=read.csv(file = fnames[r], header = TRUE,sep = ",")
+
+    list_hsib=deg2[deg2$pair %in% hsib,]
+    list_avu=deg2[deg2$pair %in% avu,]
+    list_gr=deg2[deg2$pair %in% gr,]
+
+    list_run <- rbind(list_avu,list_gr,list_hsib)
+    list_run$rel='gr'
+    list_run[list_run$pair %in% hsib,'rel']= 'hsib'
+    list_run[list_run$pair %in% avu,'rel']= 'avu'
+
+    drops <- c("pair","rel","IBD_len","IBD_num")
+    list_run <-list_run[, (names(list_run) %in% drops)]
+
+    list_all <- rbind(list_all, list_run)
+
+
+    }
+
+
+  colors <- c("avu" = "red", "hsib" = "blue", "gr" = "green")
+  #alphas <-c("0"=0.01,"1"=1)
+
+  ggplot(data=list_all, aes(x=IBD_len, y=IBD_num, color=rel)) +
+    geom_point(alpha=0.5) + theme_bw() + labs(x="Total length of IBD fragments", y= "Total number of IBD fragments") +
+    scale_color_manual("Relatedness",values = colors, labels = c("avu" = "Avuncular", "hsib" = "Half-siblings", "gr" = "Grandparent-Grandchild"))
+
[…]

### 2021-12-22 supplementary_info/Rscripts/emission_bnds.R
added supplementary plots for change in emission,beta distribution with bnds

@@ -1,0 +1,64 @@
+library(cowplot)
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+library(readr)
+
+#B1file <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/supplementary_info/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_B1file_0_15_relid"
+#B2file <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/supplementary_info/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_B2file_0_15_relid"
+#xnew1 <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/supplementary_info/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_xnew1file_0_15_relid"
+#xnew2 <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/supplementary_info/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_xnew2file_0_15_relid"
+#dfile <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz_highDiversity/contam0/inbred1/run57/coverage0.2/asc0/inputMode_hapProbs_fil0/remHighdiv_diff.csv.gz"
+#tfile <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz_highDiversity/contam0/inbred1/run57/coverage0.2/asc0/inputMode_hapProbs_fil0/remHighdiv_total.csv.gz"
+#p1f <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz_highDiversity/contam0/inbred1/run57/coverage0.2/asc0/inputMode_hapProbs_fil0/hmm_parameters/p1_file.gz"
+
+plot_b01 <- function(B1file,B2file,xnew1,xnew2,dfile,tfile,p1f,outf){
+
+  ind=15
+  df=read.csv(file = dfile, header = FALSE,sep = ",")[,ind]
+  tf=read.csv(file = tfile, header = FALSE,sep = ",")[,ind]
+  B1=read.csv(file = B1file, header = FALSE,sep = ",")
+  B2=read.csv(file = B2file, header = FALSE,sep = ",")
+  x1=read.csv(file = xnew1, header = FALSE,sep = ",")
[…]

### 2021-12-22 supplementary_info/beta_bnds.py

@@ -1,0 +1,670 @@
+
+import scipy
+import numpy as np
+import random
+import pandas as pd
+from scipy.stats import binom
+from math import isclose
+from scipy.special import gammaln
+from scipy import optimize
+import math
+from scipy.special import loggamma
+from scipy.special import beta as Betaf
+from scipy.optimize import minimize_scalar
+import pylab
+from numpy.linalg import matrix_power
+from scipy.special import logsumexp
+import numba
+from scipy.stats import beta
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+#print(scipy.__version__)
+
+
+
+"""
+#to stop the program if there is a warning
+import warnings
+
+with warnings.catch_warnings():
+    warnings.simplefilter('error')
+    for chrm in range(chrm_no):
+
+            for t in range(pos[chrm],pos[chrm+1]-1):
+
+
+                for i in range(M):
+                    print.info(chrm,t,i)
+                    numerator_l = np.log(alpha[t, i]) + np.log(A[i, :]) + np.log(B[:, t + 1].T) + np.log(beta[t + 1, :].T) - np.log(scale[t+1])  #it is the probability that hmm path has particular arc at time t
+                    xi_l[i, :, t] = numerator_l
+
+
+"""
+@numba.vectorize
+def Betal(x, y):
+    # numba doesn't understand scipy.special.gammaln, but it understands math.lgamma,
+    # which is the same, except that it doesn't understand arrays.
+    return math.lgamma(x) + math.lgamma(y) - math.lgamma(x+y)
+
+@numba.vectorize
+def fact(x):
+    return math.lgamma(x + 1)
+
+@numba.njit
+def forward(A,B,pi,data):
+    N = A.shape[0]
+    T = data.shape[0]
+    alpha = np.zeros((T,N))
+    alpha_p = np.zeros((T,N))
+    scale = np.zeros(T)
+
+    alpha[0,:] = pi * B[:,0]
+    alpha_p[0,:] = alpha[0,:] / alpha[0,:].sum()
[…]

### 2021-12-22 paper_overleaf/main.tex
Updates from Overleaf

@@
 
+- Variances are proportional to mean
+
+- READ with Chagyrskaya
+
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/fil0_read_plot.png}
+    \caption{Application of READ on Neandertal specimens from Chagyrskaya. Color of a square represents the relatedness, while the number denotes loglikelihood ratio.}
+    \label{figS2}
+\end{figure}
+
+- lcMLKIn and kin ibd
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/lcPlot.png}
+    \caption{Application of lcMLkin on Neandertal specimens from Chagyrskaya.}
+    \label{figS3}
+\end{figure}
+
+-2nd degree with 10M windows
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/degree2_10Mwin.png}
+    \caption{2deg}
+    \label{figS4}
+\end{figure}
+
+-ROH high diff
+
 \bibliographystyle{plain}

### 2021-12-23 supplementary_info/forcedROH.py
added forced ROH plot

@@ -1,0 +1,479 @@
+
+import scipy
+import numpy as np
+import random
+import pandas as pd
+from scipy.stats import binom
+from math import isclose
+from scipy.special import gammaln
+from scipy import optimize
+import math
+import matplotlib.pyplot as plt
+from scipy.special import loggamma
+from scipy.special import beta as Betaf
+from scipy.optimize import minimize_scalar
+import pylab
+from numpy.linalg import matrix_power
+from scipy.stats import beta
+
+
+
+def Betal(x,y):
+    return gammaln(x)+gammaln(y) - gammaln(x+y)
+
+def fact(x):
+    return(gammaln(x+1))
+
+
+def scaleB(B):
+    scale=B.max(axis=0)
+    Bscaled=B/scale
+    return Bscaled, np.sum(np.log(scale))
+    #return B, np.zeros(np.shape(B)[1])
+
+def forward(A,B,pi,data):
+    N= np.shape(A)[0]
+    T= np.shape(data)[0]
+    alpha=np.zeros((T,N))
+    alpha_p=np.zeros((T,N))
+    scale=np.zeros(T)
+
+    alpha[0,:]= (pi) * B[:,0]
+    alpha_p[0,:]= alpha[0,:]/sum(alpha[0,:])
+    scale[0]=sum(alpha[0,:])
+
+    for t in range(1,T):
+        alpha[t,:]= (alpha_p[t-1,:] @ A) * B[:,t]
+        alpha_p[t,:]= alpha[t,:]/sum(alpha[t,:])
+        scale[t]=sum(alpha[t,:])
+
+    return alpha_p, scale
+
+
+
+def backward(A, B, data, scale):
+
+    N= np.shape(A)[0]
+    T= np.shape(data)[0]
+    beta=np.zeros((T,N))
+    beta_p=np.zeros((T,N))
+
+    beta[-1,:]=np.ones(N)
+    beta_p[-1,:]=np.ones(N)
+
+    for t in reversed(range(T-1)):
+        for n in range(N):
+            beta[t,n]= sum( A[n,:] * beta_p[t+1,:] * B[:,t+1])
+            beta_p[t,n]= beta[t,n]/scale[t+1]
+
+    return beta_p
+
+
+
+def objective_old(x,gamma,data):
+    if len(x)==6:
+        a0=x[0]
+        b0=x[1]
[…]

### 2021-12-23 paper_overleaf/main.tex

@@
 
+-effect of bounding variance on emission. Here is an identical pair, with identical model transitions with and without variance constraints
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_emissions_bnds.png}
+    \caption{Emission with bnds}
+    \label{figS3}
+\end{figure}
+
+-effect of bounding variance on beta. Here is an identical pair, with identical model transitions with and without emissions
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_betaplot.png}
+    \caption{beta with bnds}
+    \label{figS3}
+\end{figure}
+
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_reli
+    \caption{beta with bnds}
+    \label{figS3}
+\end{figure}
+
 - lcMLKIn and kin ibd
@@
     \caption{Application of lcMLkin on Neandertal specimens from Chagyrskaya.}
-    \label{figS3}
+    \label{figS4}
 \end{figure}
@@
     \caption{2deg}
-    \label{figS4}
+    \label{figS5}
 \end{figure}

### 2021-12-23 paper_overleaf/main.tex
added supplementary info legends

--- 2021-12-13 paper_overleaf/main.tex
@@
     \includegraphics[width=18cm]{plots/plotimg/contam0_inbred0_model_performance_allroc_asc0_plot.png}
-    \caption{False positive and negative rates as a function of cutoff on log likelihood ratio. Dotted line represents $5\%$$}
+    \caption{False positive and negative rates as a function of cutoff on log likelihood ratio. Dotted line represents $5\%$}
     \label{fig3}
@@
 \section{Supplementary Information}
--binomial,betabinomial
-
+Beta binomial distribution is a better fit compared to Binomial distribution for proportion of differences in windows.
 \begin{figure}[htp]
@@
     \includegraphics[width=18cm]{supplementary_info/plots/binom.png}
-    \caption{Beta binomial distribution is a better fit compared to Binomial for proportion of difference in windows}
+    \caption{Comparison of fit with Betabinomial and Binomial distributions on (A) unrelated, (B) Parent-Child, (C) Identical pair of individuals.}
     \label{figS1}
 \end{figure}
+
+Results from KIn agree with results from READ, and provide some more information about siblings/parent-child, and $3^{rd}$ Degree relatedness. 
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/fil0_read_plot.png}
+    \caption{Application of READ on Neandertal specimens from Chagyrskaya cave.
+Color of a square represents the relatedness, while the number denotes standard deviations away from the upper threshold (we show lower threshold for unrelated pairs since upper threshold is not available).}
+    \label{figS2}
+\end{figure}
+
+Constraining variance of beta distributions in case of an identical pair, when identical relatedness hmm is applied, gives more accurate emission probabilities.
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_emissions_bnds.png}
+    \caption{Application of identical relatedness HMM on a pair of identical individuals with low coverage (0.2x), and ROH tracts.
+Top panel shows pairwise proportion of differences is close to expectation, but dips in some windows due to presence of ROH tracts.
+Bottom two panels show $P(Data|Z=2)$ without constraints and with constraints respectively.}
+    \label{figS3}
+\end{figure}
+
+In figS4, the emissions improve in accuracy with constraints on variance of Beta distributions.
+figS5 shows why we see this improvement.
+With variance constraints, beta distributions corresponding to Z=1 and Z=2 do not overfit the data, and higher emission probabilities seen in (A) at $D_w/N_w$=0.05 disappear.
+All probability distributions are scaled by the same factor so they are visible with the data histogram.
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_betaplot.png}
+    \caption{Comparison of beta distributions estimated with the HMM (A) without and (B) with variance constraints.}
+    \label{figS3}
+\end{figure}
+
+Restricting the emission probabilities at 1 for Y=1, in windows that have higher than $p_0$ difference improves the estimated Beta parameters. 
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_ind0_foreced_roh.png}
+    \caption{Comparison of Beta distributions (A) without and (B) with modified emissions}
+    \label{figS3}
+\end{figure}
+
+*Remember to use updated lcMLkin in the plot.
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/lcPlot.png}
+    \caption{Comparison of IBD states estimated for Chagyrskaya specimens using (A) lcMLkin and (B) KIn.}
+    \label{figS4}
+\end{figure}
+
+This figure shows that it is possible to differentiate between Avuncular and Grandparent-Grandchild pairs.
+However, differentiating between Half-siblings may not be possible with our approach. 
+*Add 1M windows result
+\begin{figure}[htp]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/degree2_10Mwin.png}
+    \caption{Clustering of Avuncular, Half-siblings and Grandparent-Grandchild pairs using IBD estimation in simulations.}
+    \label{figS5}
+\end{figure}
+
+-ROH high diff
 
@@
 
-#Methods
+Methods
 $$\sum_{w} \sum_{I}\sum_\omega \log P(D_{w}|Z_w, H_w)k_{w\omega i}$$

### 2022-01-20 plots/Rscripts/IBDplot.R
added chromosome differentiator

@@ -24,32 +24,95 @@ ibd_plot <- function(inf,p1f,outf){
   d3 <-d2 %>% pivot_longer(Data:Identical)
   d3$name_f = factor(d3$name, levels=c('Data','True','Identical','Parent-Child','Siblings','Degree2','Degree3','Unrelated'))
 
+  chrmlist=list()
+  for (val in 1:22)
+  {
+    if (val%%2==1){
+      lchr=rep(0,10*8)
+    }
+    else if (val%%2==0){
+      lchr=rep(1,10*8)
+    }
+    chrmlist <- append(chrmlist, lchr)
+
+  }
+  nalist=rep(NA,16)
+  chrmlist <-append(chrmlist,nalist)
+
+  d3$chrm =as.integer(chrmlist)
+
+
   d_ibd<-subset(d3, name!="Data")
   d_prop<-subset(d3, name=="Data")
 
   dat_text <- data.frame(
     label = as.factor(c("", "LL=-1478.24", "LL=-1259.44", "LL=-1213.61", "LL=-1236.09", "LL=-1261.19", "LL=-1281.77")),
     name_f = as.factor(c("True", "Identical", "Parent-Child", "Siblings", "Degree2", "Degree3", "Unrelated")),
-    x     = c(10,10,10,10,10,10,10),
+    x     = c(5,5,5,5,5,5,5),
     y     = c(0.85,0.85,0.85,0.85,0.85,0.85,0.85)
   )
 
-  plot1i<-ggplot(data=d_ibd,aes(x=Wins, y=value)) + geom_line() + facet_grid(name_f~., scales='free') +
-  scale_y_continuous(breaks=c(0.5,0.75,1.0), labels = c("2","1","0")) +theme_bw() + labs(y="IBD state", x = "Windows along the genome") +
-  theme(strip.text = element_text(size = 6))
+  datcol <-data.frame(
+    from=c(1,20,40),
+    to=c(10,30,50)
+  )
+
+  plot1i<-ggplot(data=d_ibd,aes(x=Wins, y=value)) +
+    geom_rect(aes(
+      xmin = Wins-1,
+      xmax = Wins,
+      ymin = -Inf,
+      ymax = +Inf,
+      fill = factor(chrm),
+      #alpha=0.8
+    )) +
+    scale_fill_manual(values = c("grey90", "white")) +
+    theme_void() +
+    geom_line() + facet_grid(name_f~., scales='free') +
[…]

### 2022-01-20 paper_overleaf/main.tex

@@
     \includegraphics[width=18cm]{plots/plotimg/IBDplot.png}
-    \caption{IBD fragments detected for a pair of simulated siblings (proportion of differences in each window shown on top) by the relatedness HMM with different models.}
+    \caption{Comparison of IBD fragments predicted by different relatedness models.
+Top panel shows the proportion of differences in each window along the genome for a pair of simulated siblings.
+The panel below shows the true IBD state for each window.
+Each panel below shows the IBD states predicted by a particular relatedness model.
+The log likelihood values for each model is shown on upper left corner of the panel.
+}
     \label{fig:ibd}
@@
 We see that in this case, the siblings model predictions match true IBD states the most.
-This is reflected in the log likelihoods values of the models as well, where siblings have the highest value shown in table 3.
+This is reflected in the log likelihoods values of the models as well, where siblings model has the highest value.
 We take a loglikelihood ratio ($\Lambda$) of the two most likely models to get a confidence in the relatedness estimate (in this case $\Lambda$ = 22.48).
 All the models in fig.1 utilize information about ROH given by the ROH-HMM.
-
-\begin{table}
-\caption{\label{tab:Table 3}Log likelihoods for different relatedness models}
-\begin{tabular}{|c|c|c|c|c|c|c|}
-    \hline
-    Relatedness & $Identical$ & $Parent-Child$ & $Siblings$ & $Degree2$ & $Degree3$ & $Unrelated$\\
-    \hline
-    Loglikelihood & -1478.24 & -1259.44 & -1213.61 & -1236.09 & -1261.19 & -1281.77\\
-    \hline
-
-\end{tabular}
-
-\end{table}
 

### 2022-01-20 plots/Rscripts/ROHplot.R
chrm differentiation for roh plot

@@ -1,15 +1,41 @@
 library(ggplot2)
 library(cowplot)
 
-make_panel1 <- function(data,acrval,leg){
-  plot1<-ggplot(data=data[data$acr==acrval,], aes(x=Wins, y=Prop))
-  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.4, size=0.8)
-  plot1<-plot1 + theme_bw()+ theme(legend.position = "none", axis.title.x = element_blank())+labs(y= "Heterozygosity", x = "Windows along genome")
 
+
+make_panel1 <- function(data,acrval,leg){
+  plot1<-ggplot(data=data[data$acr==acrval,], aes(x=Wins, y=Prop)) +
+  geom_rect(aes(
+    xmin = Wins-1,
+    xmax = Wins,
+    ymin = -Inf,
+    ymax = +Inf,
+    fill = factor(chrm),
+    #alpha=0.8
+  )) +
+    scale_fill_manual(values = c("grey90", "white")) +
+    theme_void()
+  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.4, size=0.8)
+  plot1<-plot1 + theme_bw()+ theme(legend.position = "none", axis.title.x = element_blank())+labs(y= "Heterozygosity", x = "Windows along genome") +
+    theme(panel.grid.major = element_blank(),
+          panel.grid.minor = element_blank(),
+          strip.background = element_blank(),
+          panel.border = element_rect(colour = "black", fill = NA))
   return(plot1)
 }
+
 make_panel2 <- function(data,acrval,leg){
-  plot2<-ggplot(data=data[data$acr==acrval,])
+  plot2<-ggplot(data=data[data$acr==acrval,]) +
+    geom_rect(aes(
+      xmin = Wins-1,
+      xmax = Wins,
+      ymin = -Inf,
+      ymax = +Inf,
+      fill = factor(chrm),
+      #alpha=0.8
+    )) +
+      scale_fill_manual(values = c("grey90", "white")) +
+      theme_void()
   plot2<-plot2+geom_line(aes(x=Wins, y=True_ROH, linetype="Truth"))
   #plot2
   plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.4, size=0.8)
@@ -22,6 +48,11 @@ make_panel2 <- function(data,acrval,leg){
[…]

### 2022-01-26 supplementary_info/supplementary_scripts.py
added supplementary table 2

@@ -1,0 +1,51 @@
+import pandas as pd
+
+def suppTable2(resultsf,kinf,outf):
+    res = pd.read_excel(resultsf,usecols=(0,1,5,6,7,8,16,17),header=None)
+
+    res.columns=list(res.loc[3,:])
+    res=res.drop([0,1,2,3],axis=0)
+    res.reset_index(drop=True, inplace=True)
+
+    res['IND1']=res['IND1'].replace({'_':''}, regex=True)
+    res['IND2']=res['IND2'].replace({'_':''}, regex=True)
+
+    res['pair']=res['IND1'] + '_' + res['IND2']
+    res.loc[0,'IND1']
+
+
+    kin=pd.read_csv(kinf, sep="\t", header=0, index_col=0)
+    kin.loc[0,:]
+
+    kin.loc[kin['relatedness']=='sib','relatedness_r'] ='First Degree'
+    kin.loc[kin['relatedness']=='pc','relatedness_r'] ='First Degree'
+    kin.loc[kin['relatedness']=='sec','relatedness_r'] ='Second Degree'
+    kin.loc[kin['relatedness']=='id','relatedness_r'] ='IdenticalTwins/SameIndividual'
+    kin.loc[kin['relatedness']=='un','relatedness_r'] ='Unrelated'
+    kin.loc[kin['relatedness']=='deg3','relatedness_r'] ='Third Degree'
+
+    kin.loc[kin['relatedness']=='sib','relatedness_l'] ='Sibling'
+    kin.loc[kin['relatedness']=='pc','relatedness_l'] ='Parent-offspring'
+    kin.loc[kin['relatedness']=='sec','relatedness_l'] ='Second Degree'
+    kin.loc[kin['relatedness']=='id','relatedness_l'] ='Identical'
+    kin.loc[kin['relatedness']=='un','relatedness_l'] ='Unrelated'
+    kin.loc[kin['relatedness']=='deg3','relatedness_l'] ='Third Degree'
+
+
+    ares=pd.merge(res, kin, how='inner', on=['pair'])
+    ares.loc[ares['Relationship (r/k0)* ']=='unrelated', 'Relationship (r/k0)* ']='Unrelated'
+    ares.loc[ares['Relationship (r/k0)* ']=='2nd degree', 'Relationship (r/k0)* ']='Second Degree'
+
+    #where do read and kin differ?
+
[…]

### 2022-01-26 paper_overleaf/main.tex

@@
 We use simulations both for estimating the transition matrices for all our models, and for testing and validating our algorithm.
-All simulations are performed using \textt{msprime} \cite{kelleher_efficient_2016}, followed by simulations of related indivudals using a predetermined pedigree (Figure S1).
+All simulations are performed using \textt{msprime} \cite{kelleher_efficient_2016}, followed by simulations of related individuals using a predetermined pedigree (Figure S1).
 
@@
 
-\begin{figure}[htp]
+\begin{figure}[h]
     \centering
@@
 Top panel shows the proportion of differences in each window along the genome for a pair of simulated siblings.
-The panel below shows the true IBD state for each window.
+Dashed lines in this panel represent $p_0$, $p_1$ and $p_2$ estimates.
+The panel just below shows the true IBD state for each window.
 Each panel below shows the IBD states predicted by a particular relatedness model.
-The log likelihood values for each model is shown on upper left corner of the panel.
+The log likelihood value for each model is shown on upper left corner of the panel.
 }
@@
     \caption{Estimation of ROH probabilities along the genome.
-Top panel shows proportion of differences along the genome, and the bottom panel shows probability of no ROH.
-A.Simulation with no ascertainment, contamination, or ROH.
-B.
-Simulation with ROH C.
-Simulation with ROH and ascertainment D.
-Simulation with ascertainment}
+Top panel shows proportion of differences along the genome, and the bottom panel shows probability of seeing no ROH.
+(A) Simulation with no ascertainment, contamination, or ROH.
+(B) Simulation with ROH.
+(C) Simulation with ROH and ascertainment.
+(D) Simulation with ROH and contamination.
+Legend to be added with inkscape: blue:4x, green:0.5x, orange:0.2x}
     \label{fig:galaxy}
@@
 'Unrelated w/o degree3'  refers to the performance results when degree 3 is considered as separate from the unrelated individuals.
-Fig.3 shows that both for identical, 1st degree, unrelated pairs, the false positive rate is below $5\%$ even when no cut off is used.
+Fig.3 shows that for identical, 1st degree, unrelated pairs, the false positive rate is below $5\%$ even when no cut off is used.
 For second degree, at 0.1x we see a high false positive rate, that does not go down much with an increase in cutoff.
@@
 For parent-child, we find that a cutoff of about 2 is a better choice.
-Degree 3 shows higher false positive rates in general, and we see that only for 4x coverage and a cutoff of 2, do we see a false positive rate of below $5\%$.
+Degree 3 shows higher false positive rates in general, and we see that only for 4x coverage and a cutoff of 2, do we see a false positive rate of around $5\%$.
 In conclusion, it is recommended to use a cutoff of 1 for most cases, with the exception of differentiation between paret-child and siblings, where a cutoff of 2 may be more appropriate, and degree 3 where cutoff of 2 at 4x coverage seems reasonable.
 
-\begin{figure}[htp]
+\begin{figure}[h]
     \centering
     \includegraphics[width=18cm]{plots/plotimg/contam0_inbred0_model_performance_allroc_asc0_plot.png}
-    \caption{False positive and negative rates as a function of cutoff on log likelihood ratio. Dotted line represents $5\%$}
+    \caption{False positive and True Positive rates as a function of cutoff on log likelihood ratio. In top panel, dotted line represents $5\%$}.
     \label{fig3}
@@
 This dataset comprised of 14 specimens that came from neighbouring layers, and hence could possibly belong to contemporary neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya \cite{kolobova_archaeological_2020-1}.
-These specimens showed low coverage of nuclear genome, with 8 samples out of 14 showing <1x coverage.
+These specimens showed low coverage of nuclear genome, with 8 samples out of 14 showing $<1x$ coverage.
 Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena.
@@
 \end{figure}
+
+We further applied KIn to a genome-wide dataset of 118 ancient individuals sequenced and analyzed by Mittnik et al.,2019.
+We compared our relatedness estimates to those obtained from the authors using READ and lcMLkin, and found that in most cases the three methods agree.
+Out of 6903 pairs, 344 pairs show different relatedness for KIn and READ (suppl table1).
+We find only three pairs in this subset where lcMLkin has sufficient data, and for two of these pairs, KIn and lcMLKin predict relatedness as second degree, and READ predicts unrelated individuals.
+This is concurrent with the results from simulations where we find that KIn has higher power to detect second degree relatedness, compared to READ.
+We find only one pair in the entire dataset for which all three methods differ.
+For pair (POST131_POST28), READ predicts unrelated, KIn predicts second degree, while lcMLkin shows 3rd-5th degree relatedness.
+KIn is able to differentiate between parent-child and siblings in many cases where READ can identify the pairs as first degree but there is not enough data for lcMLkin to give a result (suppl table2).
+In many of these cases, KIn can differentiate between parent-child and siblings.
+Interestingly, we also find cases where READ predicts first degree, but KIn and lcMLKin predict differ in prediction of siblings and parent-child (supple table 3).
+Four such pairs have good amount of data, and a visual inspection shows that in all these cases, KIn seems to have the correct estimate (supp fig.
+).
 
@@
 Beta binomial distribution is a better fit compared to Binomial distribution for proportion of differences in windows.
-\begin{figure}[htp]
+\begin{figure}[h]
     \centering
     \includegraphics[width=18cm]{supplementary_info/plots/binom.png}
-    \caption{Comparison of fit with Betabinomial and Binomial distributions on (A) unrelated, (B) Parent-Child, (C) Identical pair of individuals.}
+    \caption{Comparison of fit with Betabinomial and Binomial distributions.
+Data is represented by the histogram of proportion of differences from all windows of a (A) unrelated, (B) Parent-Child, (C) Identical pair of individuals.
+Binomial parameter p is calculated from the data, while Betabinomial parameters are estimated with the relatedness HMM.}
     \label{figS1}
@@
 Results from KIn agree with results from READ, and provide some more information about siblings/parent-child, and $3^{rd}$ Degree relatedness. 
-\begin{figure}[htp]
+\begin{figure}[h]
     \centering
@@
 Constraining variance of beta distributions in case of an identical pair, when identical relatedness hmm is applied, gives more accurate emission probabilities.
-\begin{figure}[htp]
+\begin{figure}[h]
     \centering
@@
 All probability distributions are scaled by the same factor so they are visible with the data histogram.
-\begin{figure}[htp]
+\begin{figure}[h]
     \centering
@@
 Restricting the emission probabilities at 1 for Y=1, in windows that have higher than $p_0$ difference improves the estimated Beta parameters. 
-\begin{figure}[htp]
+\begin{figure}[h]
     \centering
@@
 *Remember to use updated lcMLkin in the plot.
-\begin{figure}[htp]
+\begin{figure}[h]
     \centering
     \includegraphics[width=18cm]{supplementary_info/plots/lcPlot.png}
-    \caption{Comparison of IBD states estimated for Chagyrskaya specimens using (A) lcMLkin and (B) KIn.}
+    \caption{Comparison of IBD states estimated for Chagyrskaya specimens using (A) lcMLkin and (B) KIn.
+    The relatedness shown with different colors is estimated with and matched with READ and KIn.}
     \label{figS4}
@@
 *Add 1M windows result
-\begin{figure}[htp]
+\begin{figure}[h]
     \centering
     \includegraphics[width=18cm]{supplementary_info/plots/degree2_10Mwin.png}
-    \caption{Clustering of Avuncular, Half-siblings and Grandparent-Grandchild pairs using IBD estimation in simulations.}
+    \caption{Avuncular and Grandparent-grandchild form different clusters, when total number of IBD fragments are plotted against total length of IBD fragments.}
     \label{figS5}

### 2022-01-26 paper_overleaf/main.tex
Updates from Overleaf

@@
 We compared our relatedness estimates to those obtained from the authors using READ and lcMLkin, and found that in most cases the three methods agree.
-Out of 6903 pairs, 344 pairs show different relatedness for KIn and READ (suppl table1).
-We find only three pairs in this subset where lcMLkin has sufficient data, and for two of these pairs, KIn and lcMLKin predict relatedness as second degree, and READ predicts unrelated individuals.
-This is concurrent with the results from simulations where we find that KIn has higher power to detect second degree relatedness, compared to READ.
-We find only one pair in the entire dataset for which all three methods differ.
-For pair (POST131_POST28), READ predicts unrelated, KIn predicts second degree, while lcMLkin shows 3rd-5th degree relatedness.
+Out of 6903 pairs, 346 pairs show different relatedness for KIn and READ (suppl table2).
+However, there are only 31 pairs where KIn has $log likelihood >1$.
+We looked in detail at some of these pairs, and we found that it is possible to deduct the correct relatedness for several of these pairs by plotting the proportion of differences along the genome.
+In all the pairs that we analyzed in this way, KIn seems to predict the true relatedness.
+We show few examples of such cases in supple fig.
+We found 20 pairs in this subset of 346 pairs where lcMLkin has sufficient data.
+Among these 20 pairs, we found that lcMLkin matches KIn estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLkin predicts 3rd-5th degree, KIn predicts second degree) and pair AITI72-AITI77A (READ and lcMLKin predict unrelated, while KIn estimates 3rd degree with log likelihood=1.7).
 KIn is able to differentiate between parent-child and siblings in many cases where READ can identify the pairs as first degree but there is not enough data for lcMLkin to give a result (suppl table2).
-In many of these cases, KIn can differentiate between parent-child and siblings.
 Interestingly, we also find cases where READ predicts first degree, but KIn and lcMLKin predict differ in prediction of siblings and parent-child (supple table 3).

### 2022-01-28 supplementary_info/Rscripts/egplot1.R
added egplot1

@@ -1,0 +1,158 @@
+library(cowplot)
+library(ggplot2)
+library(readr)
+
+#datadf <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/bronze_age_newSim/mergedwin_remHighdiv_fil0/pw_diff.csv"
+#datatf <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/bronze_age_newSim/mergedwin_remHighdiv_fil0/pw_total.csv"
+#p1f <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/bronze_age_newSim/hmm_parameters_fil0/p1_file"
+#chrmf <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/bronze_age_newSim/mergedwin_fil0/merged_chrm.csv"
+egplot1 <- function(p1f, datadf, datatf, chrmf, outimg){
+
+  ind1=1907+1 #AITI62B_OTTM156
+  ind2=6783+1 #UNTA5867_UNTA5868Sk1
+  ind3=1029+1 #AITI40_AITI72
+  ind4=244+1 #AITI2_AITI55
+  p1 <- read_file(p1f)
+  p1 <- as.double(gsub("[\r\n]", "", p1))
+  p12=p1/2
+  p34=p1*3/4
+
+  datad=read.csv(file = datadf, header = FALSE)
+  datat=read.csv(file = datatf, header = FALSE)
+  chrm=read.csv(file = chrmf, header = FALSE)
+
+  d1=datad[,ind1]/datat[,ind1]
+  d2=datad[,ind2]/datat[,ind2]
+  d3=datad[,ind3]/datat[,ind3]
+  d4=datad[,ind4]/datat[,ind4]
+
+  wins=seq(1,300,1)
+
+  chrm1=chrm%%2
+  df1 <- data.frame(d1,wins,chrm1)
+  df2 <- data.frame(d2,wins,chrm1)
+  df3 <- data.frame(d3,wins,chrm1)
+  df4 <- data.frame(d4,wins,chrm1)
+
+  plot1 <- ggplot(data=df1, aes(x=wins,y=d1)) +
+    geom_rect(aes(
+      xmin = wins-1,
+      xmax = wins,
+      ymin = -Inf,
+      ymax = +Inf,
+      fill = factor(V1),
+      #alpha=0.8
+    )) +
+    scale_fill_manual(values = c("grey90", "white")) +
+    theme_void() +
[…]

### 2022-01-28 paper_overleaf/main.tex
edited suplementary info

@@
 \usepackage{graphicx}
+\usepackage{csvsimple}
+
 \title{Method to infer relatedness using low coverage ancient DNA}
@@
 Among these 20 pairs, we found that lcMLkin matches KIn estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLkin predicts 3rd-5th degree, KIn predicts second degree) and pair AITI72-AITI77A (READ and lcMLKin predict unrelated, while KIn estimates 3rd degree with log likelihood=1.7).
-KIn is able to differentiate between parent-child and siblings in many cases where READ can identify the pairs as first degree but there is not enough data for lcMLkin to give a result (suppl table2).
-Interestingly, we also find cases where READ predicts first degree, but KIn and lcMLKin predict differ in prediction of siblings and parent-child (supple table 3).
-Four such pairs have good amount of data, and a visual inspection shows that in all these cases, KIn seems to have the correct estimate (supp fig.
-).
+We found 43 pairs for which READ estimates first degree, and compared the estimates of lcMLkin and KIn for these cases (suppl table3).
+There were a total of 20 pairs where lcMLkin had sufficient data, and 26 pairs where $loglikelihood ratio>2$ for KIn.
+Among the 19 overlapping pairs, we found four pairs where KIn and lcMLkin differ, and a visual inspection shows that in these cases, KIn seems to have the correct estimate (supp fig).
 
@@
 
--ROH high diff
+-READ and KIn differ
+
+-supplementary table2
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/}
+    \caption{Avuncular and Grandparent-grandchild form different clusters, when total number of IBD fragments are plotted against total length of IBD fragments.}
+    \label{figS5}
+\end{figure}
 

### 2022-01-28 supplementary_info/supplementary_scripts.py
edited the 2 supp tables

@@ -1,6 +1,6 @@
 import pandas as pd
 
-def suppTable2(resultsf,kinf,outf):
+def suppTable2(resultsf,kinf,outf2,outf3):
     res = pd.read_excel(resultsf,usecols=(0,1,5,6,7,8,16,17),header=None)
 
     res.columns=list(res.loc[3,:])
@@ -43,9 +43,17 @@ def suppTable2(resultsf,kinf,outf):
 
 
     dtable=diff.copy()
-    dtable.columns=['Relatedness KIn', 'relatedness_l', 'Log liklihood ratio (KIn)', 'Relatedness READ','Z_upper (READ)', 'Z_lower (READ)', 'Relatedness lcMLkin', 'pair', 'nSNPs']
-    dtable=dtable.drop('relatedness_l',1)
+    dtable.columns=['relatedness_r', 'Relatedness KIn', 'Log liklihood ratio (KIn)', 'Relatedness READ','Z_upper (READ)', 'Z_lower (READ)', 'Relatedness lcMLkin', 'pair', 'nSNPs']
+    dtable=dtable.drop('relatedness_r',1)
 
+    #looking at cases where read says first degree and others disagree
+    fread=ares.loc[ares['Relationship (READ)']=='First Degree',['relatedness_r','relatedness_l','withinDeg_ll','Relationship (READ)','Z_upper','Z_lower','Relationship (r/k0)* ','pair','nSNPs']]
+    rtable=fread.copy()
+    rtable.columns=['relatedness_r', 'Relatedness KIn', 'Log liklihood ratio (KIn)', 'Relatedness READ','Z_upper (READ)', 'Z_lower (READ)', 'Relatedness lcMLkin', 'pair', 'nSNPs']
+    rtable=rtable.drop('relatedness_r',1)
 
     with pd.option_context('display.max_rows', len(dtable.index), 'display.max_columns', len(dtable.columns)):
-                dtable.to_csv(outf, sep=',')
+                dtable.to_csv(outf2, sep=',',index=False)
+
+    with pd.option_context('display.max_rows', len(rtable.index), 'display.max_columns', len(rtable.columns)):
+                rtable.to_csv(outf3, sep=',',index=False)

### 2022-01-28 supplementary_info/Rscripts/egplot2.R
added egplot2 and edited egplot1 to have ylim=0.4

@@ -6,12 +6,11 @@ library(readr)
 #datatf <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/bronze_age_newSim/mergedwin_remHighdiv_fil0/pw_total.csv"
 #p1f <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/bronze_age_newSim/hmm_parameters_fil0/p1_file"
 #chrmf <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/bronze_age_newSim/mergedwin_fil0/merged_chrm.csv"
-egplot1 <- function(p1f, datadf, datatf, chrmf, outimg){
+egplot2 <- function(p1f, datadf, datatf, chrmf, outimg){
 
-  ind1=1907+1 #AITI62B_OTTM156
-  ind2=6783+1 #UNTA5867_UNTA5868Sk1
-  ind3=1029+1 #AITI40_AITI72
-  ind4=244+1 #AITI2_AITI55
+  ind1=1128+1 #AITI43_AITI55
+  ind2=2247+1 #AITI70_AITI72
+
   p1 <- read_file(p1f)
   p1 <- as.double(gsub("[\r\n]", "", p1))
   p12=p1/2
@@ -23,16 +22,14 @@ egplot1 <- function(p1f, datadf, datatf, chrmf, outimg){
 
   d1=datad[,ind1]/datat[,ind1]
   d2=datad[,ind2]/datat[,ind2]
-  d3=datad[,ind3]/datat[,ind3]
-  d4=datad[,ind4]/datat[,ind4]
+
 
   wins=seq(1,300,1)
 
   chrm1=chrm%%2
   df1 <- data.frame(d1,wins,chrm1)
   df2 <- data.frame(d2,wins,chrm1)
-  df3 <- data.frame(d3,wins,chrm1)
-  df4 <- data.frame(d4,wins,chrm1)
+
 
   plot1 <- ggplot(data=df1, aes(x=wins,y=d1)) +
     geom_rect(aes(
@@ -62,65 +59,8 @@ egplot1 <- function(p1f, datadf, datatf, chrmf, outimg){
     annotate(geom="text", x=-2, y=0.14, label="p2",
              color="black", size=3)
 
+
   plot2 <- ggplot(data=df2, aes(x=wins,y=d2)) +
-    geom_rect(aes(
-      xmin = wins-1,
-      xmax = wins,
-      ymin = -Inf,
-      ymax = +Inf,
[…]

### 2022-02-01 plots/Rscripts/ROHplot.R
fied error in roh plot

@@ -41,7 +41,7 @@ make_panel2 <- function(data,acrval,leg){
   plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.4, size=0.8)
 
   plot2<-plot2+scale_linetype_manual(values=c("Truth"="dashed", "Inferred"="solid"))
-  plot2 <- plot2 + scale_color_manual(values=c("0.2"="Green", "0.5"="Blue","4"="Red"))
+  plot2 <- plot2 + scale_color_manual(values=c("0.2"="Red", "0.5"="Green","4"="Blue"))
   if (leg==1){
     plot2<-plot2 + theme_bw() + theme(legend.position = "bottom")+labs(y= "1 - P(ROH)", x = "Windows along genome")
     plot2<- plot2+ guides(linetype = guide_legend(nrow = 1, byrow = TRUE))+ guides(colour = guide_legend(nrow = 1, byrow = TRUE))

### 2022-02-01 paper_overleaf/main.tex
Updates from Overleaf

@@
 We see that in this case, the siblings model predictions match true IBD states the most.
-This is reflected in the log likelihoods values of the models as well, where siblings model has the highest value.
+This is reflected in the log likelihoods values (LL) of the models as well, where siblings model has the highest value.
 We take a loglikelihood ratio ($\Lambda$) of the two most likely models to get a confidence in the relatedness estimate (in this case $\Lambda$ = 22.48).
@@
 Degree 3 shows higher false positive rates in general, and we see that only for 4x coverage and a cutoff of 2, do we see a false positive rate of around $5\%$.
-In conclusion, it is recommended to use a cutoff of 1 for most cases, with the exception of differentiation between paret-child and siblings, where a cutoff of 2 may be more appropriate, and degree 3 where cutoff of 2 at 4x coverage seems reasonable.
+In conclusion, it is recommended to use a cutoff of 1 for most cases, with the exception of differentiation between parent-child and siblings, where a cutoff of 2 may be more appropriate, and degree 3 where cutoff of 2 at 4x coverage seems reasonable.
 
@@
 
-We show that READ outputs higher false positives in case of recent inbreeding in the population, and higher false negatives in case of contamination.
+Several methods are available to estimate relatedness from modern as well as ancient DNA data.
+We develop a new method to estimate genetic kinship from low coverage data.
+We show that it is possible to accurately infer relatedness, along with IBD tracts, from such data in presence of long ROH, contamination, as well as ascertainment bias.
+Our method utilizes Hidden Markov Models to estimate IBD tracts, conditional on a particular relatedness, and we use log likelihoods of these HMMs to find the best fit.
+This approach of using Hidden Markov Model makes use of the fact that DNA is inherited in blocks, and hence the IBD state in a given window is dependent on the IBD state in the previous window.
+We use this approach to identify the following relatives: Identical individuals, Parent-Child, Siblings, Second degree, Third degree, Unrelated individuals.
+We show that it is possible to differentiate between avuncular and grand-parent-grandchild relations.
+We have developed a similar Hidden Markov Model to infer long ROH, and use the posterior probabilities to make better relatedness inference.
+We further use a simple model to correct the pairwise differences in genomic windows for contamination.
+
+We demonstrate using simulations, that our method performs well at low coverage.
+We compared KIn's model performance to a widely used software READ to show that our approach works at least as good as READ on ideal data without long ROH, ascertainment and contamination, while additionally distinguishing between parent-child and siblings, and identifying third degree relatives.
+We demonstrate that at low coverage (0.1x) we have higher power to detect first degree relatives compared to READ with similar false positives, while a higher power to identify second degree relatives at the cost of slightly higher false positive rate.
+Contrary to the intuition, we see that adding long ROH ($~17\%$) to simulations improves the performance of both KIn and READ.
+READ's performance improves because ROH does not cause any change in pairwise differences between unrelated individuals, while decreasing pairwise differences between related individuals.
+More is the relatedness, the more is the decrease in pairwise differences in presence of ROH.
+And so long ROH causes higher power for READ which otherwise is a conservative method.
+KIn performs well for the same reasons, but can miss-classify second degree individuals with ROH as siblings, if ROH is not accounted for.
+Hence, we implement ROH states in the relatedness model to avoid such miss classifications.
+Contamination, on the other hand can make the performance of READ much worse depending on the amount of contamination, and the number of individuals contaminated.
+We show that even at contamination levels  $<=3\%$, and contamination in only 8 out of 17 individuals, READ performance falls dramatically.
+Contamination in more individuals will cause a shift in median difference in the population and may case non contaminated individuals to look more related, thus increasing the false positive rates.
+On the other hand, increase in amount of contamination in few individuals can make them look more unrelated.
+KIn, with the implementation of a model to account for contamination performs well in this case.
+Ascertainment causes a decrease in power for both the methods, since it reduces the amount of data.
+However, the decrease in power is more for READ compared to KIn.
+
+DNA data from Bronze age, Lech valley has low contamination, and no ROH.
+We demonstrate that KIn matches the results from READ and lcMLkin for pairs with good amount of overlaps, while adds more information for low coverage samples.
+As expected from the simulations, KIn is helpful in differentiating between parent-child from siblings, and identifying second and third degree individuals in cases where there is an overlap of only few thousand sites or less, and the other methods are not informative.
+We show that when applied to neandertal specimens from Chagyrskaya cave, READ and KIn agree in relatedness estimates for all pairs where both methods have sufficient data.
+KIn additionally identifys a pair of first degree relatives as parent-child, which is also confirmed with mtDNA.
+In addition, KIn identifies a third degree relative.
+We compare IBD estimates from KIn to lcMLkin and show that KIn's IBD estimates are much closer to the expected estimates, given the relatedness between individuals that has been estimated with both READ and KIn.
+Output of KIn is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio.
+Due to the tabular output, KIn is easy to automatise and apply on large datasets.
+One limitation with our approach is that this method assumes a single population.
+In case of population structure, with the subpopulations with very different diversity, KIn may show inaccurate inferences.
+Our method makes another assumption, that the dataset comprises of at least some unrelated individuals.
+We estimate $p_0$ as the median of all pairwise differences, and if more than half of the pairs are related to each other, $p_0$ estimate would be biased, and may affect the relatedness inference.
+In this study, we identify six different cases of relatedness, but in theory our approach can be extended to identify many more such cases, such as double first cousins, using a corresponding IBD state transition matrix.
 
@@
     \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/}
-    \caption{Avuncular and Grandparent-grandchild form different clusters, when total number of IBD fragments are plotted against total length of IBD fragments.}
+    \includegraphics[width=18cm]{supplementary_info/plots/egplot1.png}
+    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIn differs from READ.
+(A) AITI62B-OTTM156 (READ estimates Identical individuals, lcMLkin does not have sufficient data, and KIn predicts parent-child.
+(B) UNTA5867-UNTA5868Sk1 (READ estimates Second Degree, lcMLkin does not have sufficient data, and KIn predicts parent-child).
+(C) AITI40-AITI72 (READ estimates Unrelated individuals, lcMLkin shows 3rd-5th degree, and KIn predicts 3rd degree.)
+    (D) AITI2-AITI55 (READ estimates Unrelated individuals, lcMLkin and KIn predicts predict second degree.)}
+    \label{figS5}
+\end{figure}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/egplot2.png}
+    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIn differs from lcMLkin.
+(A) AITI43-AITI55 (READ estimates First degree, lcMLkin predicts siblings, while KIn predicts parent-child.
+(B) AITI70-AITI72 (READ estimates First degree, lcMLkin predicts parent-child, while KIn predicts siblings.}
     \label{figS5}
@@
 
+Discussion
+
+Our contamination correction model requires an estimate of divergence between the contaminating and the target populations.
+We provide the users an option to provide a file with average pairwise difference between the two populations, or to provide a vcf file with an individual from each population.
+Pipeline has p(diff) instead of pseudohap
 Methods

### 2022-02-02 plots/Rscripts/ROHplot.R
edited colors

@@ -17,6 +17,7 @@ make_panel1 <- function(data,acrval,leg){
     theme_void()
   plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.4, size=0.8)
   plot1<-plot1 + theme_bw()+ theme(legend.position = "none", axis.title.x = element_blank())+labs(y= "Heterozygosity", x = "Windows along genome") +
+    scale_color_manual(values=c("0.2"="Green", "0.5"="Blue","4"="Red")) +
     theme(panel.grid.major = element_blank(),
           panel.grid.minor = element_blank(),
           strip.background = element_blank(),

### 2022-02-02 plots/Rscripts/ROHlegend.R
ROH legend added

@@ -1,0 +1,107 @@
+library(ggplot2)
+library(cowplot)
+
+
+
+make_panel1 <- function(data,acrval,leg){
+  plot1<-ggplot(data=data[data$acr==acrval,], aes(x=Wins, y=Prop)) +
+    geom_rect(aes(
+      xmin = Wins-1,
+      xmax = Wins,
+      ymin = -Inf,
+      ymax = +Inf,
+      fill = factor(chrm),
+      #alpha=0.8
+    )) +
+    scale_fill_manual(values = c("grey90", "white")) +
+    theme_void()
+  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.4, size=0.8)
+  plot1<-plot1 + theme_bw()+ theme(legend.position = "none", axis.title.x = element_blank())+labs(y= "Heterozygosity", x = "Windows along genome") +
+    scale_color_manual(values=c("0.2"="Green", "0.5"="Blue","4"="Red")) +
+    theme(panel.grid.major = element_blank(),
+          panel.grid.minor = element_blank(),
+          strip.background = element_blank(),
+          panel.border = element_rect(colour = "black", fill = NA))
+  return(plot1)
+}
+
+make_panel2 <- function(data,acrval,leg){
+  plot2<-ggplot(data=data[data$acr==acrval,]) +
+    geom_rect(aes(
+      xmin = Wins-1,
+      xmax = Wins,
+      ymin = -Inf,
+      ymax = +Inf,
+      fill = factor(chrm),
+      #alpha=0.8
+    )) +
+    scale_fill_manual(values = c("grey90", "white")) +
+    theme_void()
+  plot2<-plot2+geom_line(aes(x=Wins, y=True_ROH, linetype="Truth"))
+  #plot2
+  plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.4, size=0.8)
+
+  plot2<-plot2+scale_linetype_manual(values=c("Truth"="dashed", "Inferred"="solid"))
+  plot2 <- plot2 + scale_color_manual(values=c("0.2"="Green", "0.5"="Blue","4"="Red"))
+  if (leg==1){
+    plot2<-plot2 + theme_bw() + theme(legend.position = "bottom")+labs(y= "1 - P(ROH)", x = "Windows along genome")
[…]

### 2022-02-03 supplementary_info/Rscripts/emission_bnds.R
added chrm lines

@@ -41,16 +41,49 @@ plot_b01 <- function(B1file,B2file,xnew1,xnew2,dfile,tfile,p1f,outf){
   Blabel <- c('B1'='Emission w/o constraint', 'B2'='Emission with constraint')
 
 
-  plot1<-ggplot(data=df_prop,aes(x=Wins, y=value)) + geom_line() + facet_grid(Bname~., scales='free') +
+  chrm1=rep(c(rep(0,10),rep(1,10)),11)
+  df_prop$V1=chrm1
+  plot1<-ggplot(data=df_prop,aes(x=Wins, y=value)) +
+    geom_rect(aes(
+      xmin = Wins-1,
+      xmax = Wins,
+      ymin = -Inf,
+      ymax = +Inf,
+      fill = factor(V1),
+      #alpha=0.8
+    )) +
+    scale_fill_manual(values = c("grey90", "white")) +
+    theme_void() +
+    geom_line() + facet_grid(Bname~., scales='free') +
     geom_segment(aes(x=0,xend=220,y=p1,yend=p1),linetype='dotted') + geom_segment(aes(x=0,xend=220,y=p12,yend=p12),linetype='dotted') +
     geom_segment(aes(x=0,xend=220,y=p34,yend=p34),linetype='dotted') +theme_bw()+
     theme(legend.position = "none", axis.title.x = element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(), strip.text = element_text(size = 6))+
-    labs(y=expression(D[w]/N[w])) + theme(strip.text = element_text(size = 7))
+    labs(y=expression(D[w]/N[w])) + theme(strip.text = element_text(size = 9)) +
+    theme(panel.grid.major = element_blank(),
+          panel.grid.minor = element_blank(),
+          strip.background = element_blank(),
+          panel.border = element_rect(colour = "black", fill = NA))
 
-
-  plot2<-ggplot(data=B_all,aes(x=Wins, y=value)) + geom_line() + facet_grid(Bname~., scales='free', labeller = as_labeller(Blabel)) +
+  B_all$V1=as.factor(rep(chrm1,2))
+  plot2<-ggplot(data=B_all,aes(x=Wins, y=value)) +
+    geom_rect(aes(
+      xmin = Wins-1,
+      xmax = Wins,
+      ymin = -Inf,
+      ymax = +Inf,
+      fill = factor(V1),
+      #alpha=0.8
+    )) +
+    scale_fill_manual(values = c("grey90", "white")) +
+    theme_void() +
[…]

### 2022-02-03 paper_overleaf/main.tex
Updates from Overleaf

@@
     \centering
-    \includegraphics[width=18cm]{plots/plotimg/ROHplot.png}
-    \caption{Estimation of ROH probabilities along the genome.
-Top panel shows proportion of differences along the genome, and the bottom panel shows probability of seeing no ROH.
+    \includegraphics[width=18cm]{plots/inkscape_finalImg/ROHplot.png}
+    \caption{Estimation of ROH probabilities along the genome in simulations with different cases of ROH, contamination, and ascertainment.
+Top panel shows proportion of differences in a simulated individual along the genome, and the bottom panel shows probability of seeing no ROH.
 (A) Simulation with no ascertainment, contamination, or ROH.
@@
 (C) Simulation with ROH and ascertainment.
-(D) Simulation with ROH and contamination.
-Legend to be added with inkscape: blue:4x, green:0.5x, orange:0.2x}
+(D) Simulation with ROH and contamination.}
     \label{fig:galaxy}
@@
     \includegraphics[width=18cm]{plots/plotimg/fil0_relatable_plot.png}
-    \caption{Application of KIn on Neandertal specimens from Chagyrskaya. Color of a square represents the relatedness, while the number denotes loglikelihood ratio.}
+    \caption{Application of KIn on Neandertal specimens from Chagyrskaya.
+Color of a square represents the relatedness, while the number denotes loglikelihood ratio between the two maximum likelihood models.}
     \label{fig5}

### 2022-02-03 supplementary_info/beta_bnds.py
edited supplementary plots for lrgrnds

@@ -636,28 +636,29 @@ def plot_betas(dfile, tfile, pair, thresh, rel, libraries, listf, Afile, pfile, 
         #no. of points in each bin=1000*0.08/0.15
         #I normalize with this value to make it comparable to the prop. of diff. histogram which is also normalized by the
         #no. of total windows
-        a=betas[0]
-        b=betas[1]
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'b-', lw=3, alpha=0.6, label='beta pdf')
 
         a=betas[2]
         b=betas[3]
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'r-', lw=3, alpha=0.6, label='beta pdf')
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'r-', lw=3, alpha=0.6, label="$Z_w$=0")
         #plt.hist(diff/total,alpha=0.3)
 
+        a=betas[0]
+        b=betas[1]
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'b-', lw=3, alpha=0.6, label="$Z_w$=1")
 
         a=betas[4]
         b=betas[5]
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'g-', lw=3, alpha=0.6, label='beta pdf')
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'g-', lw=3, alpha=0.6, label="$Z_w$=2")
         #plt.hist(diff/total,alpha=0.3)
 
         a=betas[6]
         b=betas[7]
 
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'y-', lw=3, alpha=0.6, label='beta pdf')
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'y-', lw=3, alpha=0.6, label="$Z_w$=4")
         ax[i].hist(diff/total,alpha=0.3,weights=np.ones(len(total)) / len(total))
         ax[i].set_ylim(0,0.4)
 
+    ax[1].legend(loc="upper right")
     fig.text(0.5, 0.04, 'Frequency', ha='center')
     fig.text(0.04, 0.5, 'Proportion of differences', va='center', rotation='vertical')
     plt.savefig(betaplot, bbox_inches='tight')

### 2022-02-03 supplementary_info/forcedROH.py

@@ -464,16 +464,17 @@ def plotbetas(dfile, tfile, ind, chrmf, p1file, upA, targets, outplot):
 
         a=betas[0]
         b=betas[1]
-        ax[i].plot(x, beta.pdf(x, a, b)/500,'b-', lw=3, alpha=0.6, label='beta pdf')
+        ax[i].plot(x, beta.pdf(x, a, b)/500,'b-', lw=3, alpha=0.6, label="$Y_w$=0")
 
         a=betas[2]
         b=betas[3]
-        ax[i].plot(x, beta.pdf(x, a, b)/500,'r-', lw=3, alpha=0.6, label='beta pdf')
+        ax[i].plot(x, beta.pdf(x, a, b)/500,'r-', lw=3, alpha=0.6, label="$Y_w$=1")
         ax[i].hist(diff/total,alpha=0.3, weights=np.ones(len(diff)) / len(diff))
 
         ax[i].set_ylim(0,1)
         ax[i].set_xlim(0,0.04)
-
+        
+    ax[1].legend(loc="upper right")
     fig.text(0.5, 0.04, 'Proportion of differences', ha='center')
     fig.text(0.04, 0.5, 'Frequency', va='center', rotation='vertical')
     plt.savefig(outplot, bbox_inches='tight')

### 2022-02-15 paper_overleaf/main.tex
Updates from Overleaf

@@
 \usepackage{amsmath}
+\usepackage{bm}
 \usepackage{bbm}
@@
 \newcommand{\BH}{\mathbf{H}}
+\newcommand{\Btheta}{\pmb{\theta}}
 
@@
 This can affect the relatedness estimates.
-Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and pairwise relatedness upto 3rd degree, while additionally differentiating between siblings and parent-child.
+Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness upto 3rd degree, while additionally differentiating between siblings and parent-child.
 We have developed another HMM to estimate ROH proportions, that can be utilized by the relatedness estimation model to output better estimates.
@@
 READ is a software that does not require population allele frequencies or a reference panel, and can work with around 0.5 x coverage in presence of ascertainment bias.
-The software solves the problem of unavailability of genotype calls by randomly sampling alleles from each individual.
+The software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual.
 A string of these alleles at each position (called pseudohaploids) are then compared to other individuals to calculate pairwise genetic distances, which in turn are used to infer relatedness.
@@
 KIn is also able to incorporate ROH, contamination and ascertainment bias.
-We validate the performance of KIn using simulations and show that we are able to infer relatedness in an archaic human data set.
+We validate the performance of KIn using simulations and show that we are able to infer relatedness in a neandertal data set, and a Bronze age dataset.
 
@@
 \subsection{Relatedness Model overview} 
-To estimate the relatedness of two individuals, we subdivide the genome into $L$ large windows $w$ (typically 10Mb).
-The inputs of our algorithm are i) the number of sites $N_w$ for which both samples have at least one read available, ii) the number of pairwise differences $D_w$ at these sites, and iii) position of ROH, by default obtained from ROH-HMM described in a following section.
+To estimate the relatedness of two individuals, we subdivide the genome into $L$ large windows $w$ (typically of size 10Mb).
+The inputs of our algorithm are i) the number of sites $N_w$ for which both samples have at least one read available, ii) the number of pairwise differences $D_w$ at these sites, and iii) probability of ROH in in windows, by default obtained from ROH-HMM described in a following section.
+Calculation of $N_w$ and $D_w$ requires comparison of reads at all positions for a pair of individuals.
+For low coverage genomes, this can be done by comparing randomly sampled read at each position for a pair of individuals (or by comparing pseudo haploid sequences).
+However, we realized that such random sampling can throw away a lot of useful data.
+To avoid this wastage, our input generation pipeline for Relatedness HMM calculates proportion of reads with alternate allele, at all positions for each genome.
+We use these proportions to find pairwise heterozygosity as the probability of comparing different alleles.
+These probabilities are summed up in each window to give $D_w$, along with the number of overlapping sites $N_w$.
 
@@
 
-The model then uses this data for a given pair of individuals to classify each window into three hidden states $Z_w$ reflecting zero, one or two chromosomes shared IBD.
-Likewise, we classify each window into one of three possible ROH state $H_w$, reflecting, neither, one or both indivuals being in a run-of-homozygosity at this genomic location.
+The model then uses this data for a given pair of individuals to classify each window into three hidden states $Z_w \in$ ($i_0$, $i_1$, $i_2$), reflecting zero, one or two chromosomes shared IBD.
+Likewise, we classify each window into one of three possible ROH state $H_w \in$ ($\omega_0$, $\omega_1$, $\omega_2$), reflecting, neither, one or both individuals being in a run-of-homozygosity at this genomic location.
 
-Ideally, we would like to compute the likelihood $P(\BD | \BZ, \BH, \theta$), where  $\theta$ is a vector of parameters (initial transition probability $\pi$, transition matrix $A$ and $N_w$).
+Ideally, we would like to compute the likelihood $P(\BD | \BZ, \BH, \theta)$, where $\theta$ is a vector of parameters (initial transition probability $\pi$, transition matrix $A$ and $N_w$).
 
@@
 &= \prod_w P(D_w|Z_w,H_w, \theta) \prod_w P(H_w |\theta) \prod_w P(Z_w |Z_{w-1},\theta) P(Z_0|\theta)
+\label{eq:1}
 \end{align}
@@
 Using the complete data likelihood allows us to factor the emission and transition probabilities, and hence we can maximize them independently.
-Transition matrix for different cases of relatedness are obtained from simulations (see section). Initial transition probability $\pi_i$ is same for all the $i$.
+Transition matrix for different cases of relatedness are either derived from theory, or obtained from simulations (see section). Initial transition probability $\pi_i$ is same for all the $i$.
 
 \subsection{Emission probability}
-In equation \ref{eq:cll}  the term $P(D_w | Z_w, H_w, \theta) = P(D_w | Z_w, H_w, N_w)$ is the emission probability, i.e.
+In equation \ref{eq:1} the term $P(D_w | Z_w, H_w, \theta) = P(D_w | Z_w, H_w, N_w)$ is the emission probability, i.e.
 the probability of our data , given a particular hidden state.
@@
 If the two individuals are unrelated in a particular window and neither is in a ROH region (i.e.
-$Z_w = H_w = 0$), then the expected proportion of pairwise differences depends solely on the effective population size, and we denote this proportion with $p_0$.
+$Z_w = i_0$, and $H_w = \omega_0$), then the expected proportion of pairwise differences depends solely on the effective population size, and we denote this proportion with $p_0$.
 If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes.
@@
 We calculate $p_1$ as $(3/4)p_1$ and $p_2$ as $p_1/2$.
-However, we need one more parameter (represented as $p_4$ in the table) in cases where IBD state is $i_1$ and ROH state is $\omega_2$, or IBD state is $i_2$ and ROH state is $\omega_1$ or $\omega_2$.
+However, we need one more parameter (represented as $p_3$ in the matrix below) in cases where IBD state is $i_1$ and ROH state is $\omega_2$, or IBD state is $i_2$ and ROH state is $\omega_1$ or $\omega_2$.
 In such cases the expected proportion of differences between the individuals should be 0.
-However, as mentioned before, we do all our calculations in windows of 10 Mb, and start/end positions of these windows may not coincide with that of ROH tracts.
-And so, the pairwise differences in this state (called $p_4$ in the table) can lie anywhere between 0 to $p_2$.
-We found $p_4$ = $p_2$/2 to be a reasonable assumption for expected pairwise proportion of differences in this IBD/ROH state.
+However, as mentioned before, we do all our calculations in large windows, and start/end positions of these windows may not coincide with that of ROH tracts.
+And so, the pairwise differences in this state (called $p_3$ in the table) can lie anywhere between 0 to $p_2$.
+We found $p_3$ = $p_2$/2 to be a reasonable assumption for expected pairwise proportion of differences in this IBD/ROH state.
 
 Taken together, we can summarize $p$ as follows 
-\begin{equation}\label{eq:p}
+\begin{equation}\label{eq:2}
     p(Z_w, H_w) = \left[\begin{array}
@@
 p_0 & p_1 & p_2 \\
-p_0 & p_2 & p_4 \\
-p_0 & p_4 & p_4
+p_0 & p_2 & p_3 \\
+p_0 & p_4 & p_3
 \end{array}\right]
@@
 We therefore add an overdispersion parameter $\delta$ that allows for variation in the $p$ between SNPs.
-This results in the betabinomial likelihood, (we show that such a distribution fits the data well (supplementary fig)).
+This results in the betabinomial likelihood, (we show that such a distribution fits the data well (fig.
+S\ref{FigS1:binom})).
 
-\begin{equation}\label{eq:bbsimple}
-P(D_{w}|p_i,\delta_i,N_w) = \binom{N_w}{D_w}\frac{B(D_w+p_i \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})}
+\begin{equation}\label{eq:3}
+P(D_{w}|p_j,\delta_j,N_w) = \binom{N_w}{D_w}\frac{B(D_w+p_j \delta_{j}, N_w-D_w+ \delta_{j}(1-p_{j}))}{ B(p_{j}\delta_{j}, (1-p_{j})\delta_{j})}
 \end{equation}
@@
 
-In this step, using the standard forward-backward algorithm, we calculate the posterior probability of each IBD state in each window $\gamma_{wi}$.
+In this step, using the standard forward-backward algorithm, we calculate the posterior probability of each IBD state in each window ($\gamma_{wi}$).
 Expectation step, among other parameters, requires emission probabilities.
-We calculate emission probabilities for each case of $Z_w$ and $H_w$ from $\delta$ and $p$ using equation 3, and we get the total emission probabilities for each $Z_w$ as $\sum_{H_w} P(D_w|Z_w,N_w,H_w) P(H_w|\theta)$
+We calculate emission probabilities for each case of $Z_w$ and $H_w$ from $\delta$ and $p$ using equation \ref{eq:3}, and we get the total emission probabilities for each $Z_w$ as $\sum_{H_w} P(D_w|Z_w,N_w,H_w) P(H_w|\theta)$
 
@@
 The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_j$.
-We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_wi$ and the ROH state-probabilities and $h_{w\omega}$.
+We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wi}$ and the ROH state-probabilities and $h_{w\omega}$.
 
-\begin{equation}
+\begin{equation}\label{eq:4}
 \mathcal{C} = \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta, p)h_{w\omega}\gamma_{wi}
@@
 
-Using equation \ref{eq:p}, we simplify this by grouping all the terms that would result in the same $p$, i.e.
+Using equation \ref{eq:2}, we simplify this by grouping all the terms that would result in the same $p$, i.e.
 
@@
 k_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\\
-k_{w4} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2}
+k_{w3} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2}
 \end{align*}
 
-So we can rewrite Eqn 4 as:
+So we can rewrite equation \ref{eq:4} as:
 \begin{align}
-\mathcal{C} &= \sum_{w=0}^L\sum_{j=0}^4  \log P(D_{w}|Z_w, H_w, N_w, \delta_j, p_j)k_{wj}
+\mathcal{C} &= \sum_{w=1}^L\sum_{j=0}^3 \log P(D_{w}|Z_w, H_w, N_w, \delta_j, p_j)k_{wj}
 \end{align}
 
-Since each term in this equation depends on only one $\delta_j$, we can optimize for them independently.
+Since each term in this equation depends on only one $\delta_j$, we can optimize for them independently. For convenience, we refer to these four states corresponding to different $p$ as $x$ states.
 
-We observed that estimating the $\delta_j$ without constraints can result in wide Beta distributions with very high variance, leading to over fitting of data.
+We observed that estimating the $\delta_j$ without constraints can result in wide Beta distributions with very high variance, leading to over fitting of data (fig.
+\ref{figS4:bndsbeta}), and resulting in incorrect posterior probabilities for $Z_w$ (fig.
+\ref{figS3:bnds}).
 For example, a parent shares exactly one chromosome IBD with its child throughout the genome, so we expect only one IBD state.
-However,  running an unconstrained model could use this data to fit a pair of siblings with a very wide beta-distribution (with a very high $\delta_1$, which causes problems in differentiating between parent-child and siblings.
+However, running an unconstrained model could use this data to fit a pair of siblings with a very wide beta-distribution (with a very high $\delta_1$, which may result in problems in differentiating between parent-child and siblings.
 
@@
 
-We choose t for each beta distribution to be proportional to the mean ($p$), since the variances of beta distribution corresponding to different $x$ states follow the same relation as the distribution means (supplement figure).
+We choose t for each beta distribution to be proportional to the mean ($p$), since the variances of beta distribution corresponding to different $x$ states follow the same relation as the distribution means.
 \begin{align}
@@
 \end{align}
+
+Even though this may not hold true when we bin sites in large windows, we still find that $Var(X_{x_0}) > Var(X_{x_1}) > Var(X_{x_2})$ (fig. \ref{FigS1:binom}).
 
@@
 The expectation step outputs the posterior probability $\Gamma$ of being in $y_0$ and $y_1$ in each window.
-Maximization step uses the $\Gamma$ to optimize the $\Delta$ and $\rho$ parameters with a betabinomial cost function:
+Maximization step uses the $\Gamma$ to optimize overdispersion parameter $\eta$ with a betabinomial cost function:
 
@@
 
-In case of very noisy data, we see high pairwise differences in some windows that are not explained with the beta binomial distributions estimated for $y_0$ and $y_1$.
-To avoid our distributions from flattening to explain these high differences, we force these high difference windows to $y_1$ (see supplement figure).
+In case of very noisy data, we see high pairwise differences in some windows that are not explained with the betabinomial distributions estimated for $y_0$ and $y_1$.
+To avoid our distributions from flattening to explain these high differences, we force these high difference windows to $y_1$ (fig.
+S\ref{figS5:ROHforced}).
 
@@
 Even low levels of contamination may make samples look less similar, and thus distort relatedness analyses.
-This issue is particularly pronounced  when analyzing populations with  low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations (supplementary figure with READ?, lcMLkin at 4x).
-To address this issue, we  adjust the  pairwise differences $D_w$, given the contamination estimates of both samples and the average divergence between the target population and a putative contaminating population.
+This issue is particularly pronounced  when analyzing populations with  low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations.
+To address this issue, we  adjust $D_w$ and $N_w$, given the contamination estimates of both samples and the average divergence between the target population and a putative contaminating population.
 
@@
 $C_iC_j \approx 0$.
-We denote the average divergence between target and contaminating populations as $p_c$, and the observed proportion of differences between individuals $i$ and $j$ as $p_{aij}$.
-We calculate average endogenous pairwise difference $p_eij$ for individuals i and j as follows \cite{noauthor_ancient_nodate}:
+We denote the average divergence between target and contaminating populations as $p_c$, and the observed proportion of differences between individuals $i$ and $j$ as $p_{oij}$.
+We calculate average endogenous pairwise difference $p_{eij}$ for individuals i and j as follows \cite{noauthor_ancient_nodate}:
 \begin{align}
-    p_{aij} = C_{ij}  p_c + (1-C_{ij})  p_{eij}\\
-    or, p_{eij} = \frac{p_{aij} - C_{ij}  p_c}{1-C_{ij}} 
+    p_{oij} = C_{ij}  p_c + (1-C_{ij})  p_{eij}\\
+    or, p_{eij} = \frac{p_{oij} - C_{ij}  p_c}{1-C_{ij}} 
 \end{align}
@@
 \paragraph{Short version}
-With contamination, we do not observe the number of differences $D$ and total number of comparisons directly $N$, but instead need to approximate them.
+With contamination, we do not observe the number of differences $D_{true}$ and total number of comparisons $N_{true}$ directly, but instead need to approximate them.
 The window-subscript is omitted in the following for clarity:
 \begin{equation}
-    P(D_{\text{obs}} | Z, N_\text{obs}, \theta) \approx 
-    P( \hat{D_w} | \hat{N_w}, Z, \theta)
+    P(D_{\text{obs}} | N_\text{obs}, Z, \theta) \approx 
+    P( \hat{D} | \hat{N}, Z, \theta)
 \end{equation}
@@
 \begin{itemize}
-\item W: Total number of windows
+\item L: Total number of windows
 \item w: Index of window
-\item $I$:IBD states base
-\item $Z_w$: IBD state in window w
+\item $N_w$: Number of sites in w
+\item $D_{w}$: Number of pairwise differences in w
+\item $Z_w$: Hidden IBD state in window w
+\item $H_w$: ROH state in window w
+\item A: Transition matrix
+\item $\pi$: Vector of initial transition probabilities
+\item $\theta$: Vector of $\pi$, A, $N_w$
 \item $\mathbf{Z}$: Vector of $Z_w$'s
@@
 We predict the following categories of relatedness with KIn: Identical individuals (twins), First degree (parent-child), First degree (siblings), Second degree, Third degree, Unrelated (including higher than third degree relatives).
-We show the working of Relatedness HMM with fig.1 showing performance of each model, when applied to a pair of siblings.
+We show the working of Relatedness HMM with Fig.\ref{fig1:ibd} showing performance of each model, when applied to a pair of siblings.
 
@@
 }
-    \label{fig:ibd}
+    \label{fig1:ibd}
 \end{figure}
@@
 We take a loglikelihood ratio ($\Lambda$) of the two most likely models to get a confidence in the relatedness estimate (in this case $\Lambda$ = 22.48).
-All the models in fig.1 utilize information about ROH given by the ROH-HMM.
+All the models in Fig.\ref{fig1:ibd} utilize information about ROH given by the ROH-HMM.
 
-Performance of ROH-HMM under different cases of ascertainment, contamination and long ROH, and for different coverages is shown in fig 2.
+Performance of ROH-HMM under different cases of ascertainment, contamination and long ROH, and for different coverages is shown in Fig.\ref{fig2:ROH}.
 We applied the ROH-HMM and relatedness HMM on 60 runs of simulations, to get relatedness estimates, and compared the final relatedness estimates to the true relatedness for different pairs at coverages ranging from 4x to 0.1x.
@@
 (D) Simulation with ROH and contamination.}
-    \label{fig:galaxy}
+    \label{fig2:ROH}
 \end{figure}
 
-We plotted the true positive and false positive rates with a cutoff on the loglikelihood ratio in fig.3.
-Here, and in fig.4, we have two relatedness labels: 'Unrelated' and 'Unrelated w/o 3rd Degree'.
+We plotted the true positive and false positive rates with a cutoff on the loglikelihood ratio in Fig.\ref{fig3:cutoff}.
+Here, and in Fig.\ref{fig4:Comparison_READ_KIn}, we have two relatedness labels: 'Unrelated' and 'Unrelated w/o 3rd Degree'.
 Unrelated label here refers to KIn performance results when all unrelated, degree5, degree4, degree3 pairs are labelled as unrelated.
@@
 'Unrelated w/o degree3'  refers to the performance results when degree 3 is considered as separate from the unrelated individuals.
-Fig.3 shows that for identical, 1st degree, unrelated pairs, the false positive rate is below $5\%$ even when no cut off is used.
+Fig.\ref{fig3:cutoff} shows that for identical, 1st degree, unrelated pairs, the false positive rate is below $5\%$ even when no cut off is used.
 For second degree, at 0.1x we see a high false positive rate, that does not go down much with an increase in cutoff.
@@
     \caption{False positive and True Positive rates as a function of cutoff on log likelihood ratio. In top panel, dotted line represents $5\%$}.
-    \label{fig3}
+    \label{fig3:cutoff}
 \end{figure}
@@
 We use a cutoff of 1 to compare our method performance to READ, while using a cutoff of 1 standard deviation for READ.
-Fig.4 shows this comparison using false positive and true positive rates for both the methods at different coverages under different simulations.
-We show that both methods have similar performance in the control case (except for 0.1x coverage, where KIn has higher power to detect degree1 and degree2, and higher false positive rate for degree 2).
+Fig.\ref{fig4:Comparison_READ_KIn} shows this comparison using false positive and true positive rates for both the methods at different coverages under different simulations.
+We show that both methods have similar performance in the control case (except for 0.1x coverage, where KIn has higher power to detect degree1 and degree2, and slightly higher false positive rate for degree 2).
 However, in case of ascertainment, ROH, or comtamination KIn shows higher power to detect relatedness cases while maintaining a low false positive rate.
@@
     \caption{Comparison of KIn with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH}
-    \label{fig4}
+    \label{fig4:Comparison_READ_KIn}
 \end{figure}
 
-To test KIn on real ancient dataset, we applied it to look for neandertal specimens from Chagyrskaya cave in Siberia, Russia.
+To test KIn on real ancient dataset, we applied it to neandertal specimens from Chagyrskaya cave in Siberia, Russia.
 This dataset comprised of 14 specimens that came from neighbouring layers, and hence could possibly belong to contemporary neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya \cite{kolobova_archaeological_2020-1}.
@@
 This dataset had all the issues that KIn deals with, and hence we decided to apply it here to see if we could find family relationships that would not be possible with other available methods.
-Our results for pairwise relatedness for these individuals is shown in fig.5.
-We found that identical individuals (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), first degree relatives (Chagyrskaya07-Chagyrskaya17, Chagyrskaya06-Chagyrskaya14), and a second degree relative (Chagyrskaya01-Chagyrskaya60).
-These results agree with READ (supplementary fig).
-However, we find that Chagyrskaya07-Chagyrskaya17 are parent-child.
-This matches with the conclusions of the authors based on additional information from mtDNA.
+Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_kin}.
+We found identical individuals (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), first degree relatives (Chagyrskaya07-Chagyrskaya17, Chagyrskaya06-Chagyrskaya14), and a second degree relative (Chagyrskaya01-Chagyrskaya60).
+These results agree with READ (fig.
+S\ref{figS2:Chagyrskaya_READ}).
+Interestingly, for the pair of Chagyrskaya07-Chagyrskaya17, READ shows that they are first degree relatives, while KIn further identifies them as parent-child.
+Our result matches with the conclusions of the authors based on additional information from mtDNA.
 Further, we find a possible third degree relative (Chagyrskaya07-Chagyrskaya60).
-We compare our results to lcMLkin, and show that the software does not perform well with these low coverage samples since it is not possible to obtain reliable genotype likelihoods for many of these samples (supplementary fig).
+We compare our results to lcMLkin, and show that the software does not perform well with these low coverage samples since it is not possible to obtain reliable genotype likelihoods for many of these samples (fig.
+S\ref{figS6:Chagyrskaya_ibd}).
 
@@
 Color of a square represents the relatedness, while the number denotes loglikelihood ratio between the two maximum likelihood models.}
-    \label{fig5}
+    \label{fig5:Chagyrskaya_kin}
 \end{figure}
@@
 We compared our relatedness estimates to those obtained from the authors using READ and lcMLkin, and found that in most cases the three methods agree.
-Out of 6903 pairs, 346 pairs show different relatedness for KIn and READ (suppl table2).
+Out of 6903 pairs, 346 pairs show different relatedness for KIn and READ (table S2).
 However, there are only 31 pairs where KIn has $log likelihood >1$.
@@
 In all the pairs that we analyzed in this way, KIn seems to predict the true relatedness.
-We show few examples of such cases in supple fig.
+We show few examples of such cases in fig.
+S\ref{figS8:eg1}.
 We found 20 pairs in this subset of 346 pairs where lcMLkin has sufficient data.
@@
 There were a total of 20 pairs where lcMLkin had sufficient data, and 26 pairs where $loglikelihood ratio>2$ for KIn.
-Among the 19 overlapping pairs, we found four pairs where KIn and lcMLkin differ, and a visual inspection shows that in these cases, KIn seems to have the correct estimate (supp fig).
+Among the 19 overlapping pairs, we found four pairs where KIn and lcMLkin differ, and a visual inspection shows that in these cases, KIn seems to have the correct estimate (fig.
+S\ref{figS9:eg2}).
 
@@
 We show that even at contamination levels  $<=3\%$, and contamination in only 8 out of 17 individuals, READ performance falls dramatically.
-Contamination in more individuals will cause a shift in median difference in the population and may case non contaminated individuals to look more related, thus increasing the false positive rates.
+Contamination in more individuals will cause a shift in median difference in the population and may cause non contaminated individuals to look more related, thus increasing the false positive rates.
 On the other hand, increase in amount of contamination in few individuals can make them look more unrelated.
@@
 Beta binomial distribution is a better fit compared to Binomial distribution for proportion of differences in windows.
+
+\renewcommand{\figurename}{Fig. S}
 \begin{figure}[h]
@@
 Binomial parameter p is calculated from the data, while Betabinomial parameters are estimated with the relatedness HMM.}
-    \label{figS1}
+    \label{FigS1:binom}
+
 \end{figure}
@@
 Color of a square represents the relatedness, while the number denotes standard deviations away from the upper threshold (we show lower threshold for unrelated pairs since upper threshold is not available).}
-    \label{figS2}
+    \label{figS2:Chagyrskaya_READ}
 \end{figure}
@@
 Bottom two panels show $P(Data|Z=2)$ without constraints and with constraints respectively.}
-    \label{figS3}
+    \label{figS3:bnds}
 \end{figure}
@@
     \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_betaplot.png}
-    \caption{Comparison of beta distributions estimated with the HMM (A) without and (B) with variance constraints.}
-    \label{figS3}
+    \caption{Comparison of beta distributions estimated with the HMM (left) without and (right) with variance constraints.}
+    \label{figS4:bndsbeta}
 \end{figure}
@@
     \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_ind0_foreced_roh.png}
-    \caption{Comparison of Beta distributions (A) without and (B) with modified emissions}
-    \label{figS3}
+    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_ind0_forced_roh.png}
+    \caption{Comparison of Beta distributions (left) without and (right) with modified emissions}
+    \label{figS5:ROHforced}
 \end{figure}
@@
     The relatedness shown with different colors is estimated with and matched with READ and KIn.}
-    \label{figS4}
+    \label{figS6:Chagyrskaya_ibd}
 \end{figure}
@@
     \caption{Avuncular and Grandparent-grandchild form different clusters, when total number of IBD fragments are plotted against total length of IBD fragments.}
-    \label{figS5}
+    \label{figS7:second_degree}
 \end{figure}
@@
     (D) AITI2-AITI55 (READ estimates Unrelated individuals, lcMLkin and KIn predicts predict second degree.)}
-    \label{figS5}
+    \label{figS8:eg1}
 \end{figure}
@@
 (B) AITI70-AITI72 (READ estimates First degree, lcMLkin predicts parent-child, while KIn predicts siblings.}
-    \label{figS5}
+    \label{figS9:eg2}
 \end{figure}

### 2022-02-17 paper_overleaf/main.tex

@@
 \usepackage{csvsimple}
+\usepackage[section]{placeins}
 
@@
 \begin{document}
+\nolinenumbers
 
@@
 \begin{abstract}
+
 \noindent Knowledge of familial relationships is an important part of several research fields.
@@
 \subsection{Why study relatedness?}
+
 Identifying related individuals is a common task in applications of genetics.
@@
 \subsection{Simple way to get relatedness, methods that do it with a lot of data}
+
 Commonly, pairs of relateded individuals are  identified by looking for parts of the genome that are identical by descent (IBD), ie.
@@
 The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals.
-For example, a pair of siblings are expected to have all three possible IBD states with probabilities (0.25,0.5,0.25) as shown in fig 1.
-These IBD probabilities can directly be used to categorize a pair of individuals as related by a certain degree and nature of relatedness as shown in table1.
+For example, a pair of siblings are expected to have all three possible IBD states with probabilities (0.25,0.5,0.25) as shown in Fig \ref{fig0:schematic}.
+These IBD probabilities can directly be used to categorize a pair of individuals as related by a certain degree and nature of relatedness as shown in table \ref{tab:Table 1}.
 One can also use these probabilities to estimate the coefficient of relatedness as $r= k_1/2 + k_2$.
 The coefficient of relatedness (r) is defined as the proportion of genome IBD for a pair of individuals.
+
+\begin{figure}[h!]
+    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic1.png}
+    \centering
+    \caption{Overview of KIn.
+A.
+In top row, recombined chromosomes inherited from parents by a pair of siblings are shown.
+On left, the parents have unrelated chromosomes, and on right one parent has homozygous pair of chromosomes.
+Bottom row shows the expected proportion of differences along the chromosome in both the cases.
+B.
+Overall schematic of the method.
+Blue boxes show the data files, while green boxes represent scripts.}
+    \label{fig0:schematic}
+\end{figure}
+
 There are numerous methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to estimate the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}.
@@
 \end{tabular}
-
+\label{table1}
 \end{table}
@@
 
-The effect of these considerations is that even though we have nine possible combinations of $Z$ and $H$ for each window, there are actually only four different $p$-parameters.
+The effect of these considerations is that even though we have nine possible combinations of $Z$ and $H$ for each window, there are actually only four different $p$-parameters $p_x$ with $x$ $\in$ (0,1,2,3).
 
@@
 \begin{equation}\label{eq:3}
-P(D_{w}|p_j,\delta_j,N_w) = \binom{N_w}{D_w}\frac{B(D_w+p_j \delta_{j}, N_w-D_w+ \delta_{j}(1-p_{j}))}{ B(p_{j}\delta_{j}, (1-p_{j})\delta_{j})}
+P(D_{w}|p_x,\delta_x,N_w) = \binom{N_w}{D_w}\frac{B(D_w+p_x \delta_{x}, N_w-D_w+ \delta_{x}(1-p_{x}))}{ B(p_{x}\delta_{x}, (1-p_{x})\delta_{x})}
 \end{equation}
@@
 The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_j$.
-We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wi}$ and the ROH state-probabilities and $h_{w\omega}$.
+We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wi}$ and the ROH state-probabilities $h_{w\omega}$.
 
@@
 \begin{align}
-\mathcal{C} &= \sum_{w=1}^L\sum_{j=0}^3 \log P(D_{w}|Z_w, H_w, N_w, \delta_j, p_j)k_{wj}
+\mathcal{C} &= \sum_{w=1}^L\sum_{x=0}^3 \log P(D_{w}|Z_w, H_w, N_w, \delta_x, p_x)k_{wx}
 \end{align}
 
-Since each term in this equation depends on only one $\delta_j$, we can optimize for them independently. For convenience, we refer to these four states corresponding to different $p$ as $x$ states.
+Since each term in this equation depends on only one $\delta_x$, we can optimize for them independently. For convenience, we refer to these four states corresponding to different $p$ as $x$ states.
 
-We observed that estimating the $\delta_j$ without constraints can result in wide Beta distributions with very high variance, leading to over fitting of data (fig.
+We observed that estimating the $\delta_w$ without constraints can result in wide Beta distributions with very high variance, leading to over fitting of data (fig.
 \ref{figS4:bndsbeta}), and resulting in incorrect posterior probabilities for $Z_w$ (fig.
@@
 The emissions are calculated using betabinomial likelihood, and the mean of the distributions are constrained at expected proportion of differences in a ROH tract ($\rho_{0}$) and expected proportion of differences in a non-ROH tract $\rho_{1}$.
-Parameters $\rho_{1}$ and $\rho_{0}$ are the same as $p_{2}$ and $p_4$ in previous model.
+Parameters $\rho_{1}$ and $\rho_{0}$ are the same as $p_{2}$ and $p_3$ in previous model.
 
@@
 
-\section{Notation summary}
-Notation used in the paper:
-
-\begin{itemize}
-\item L: Total number of windows
-\item w: Index of window
-\item $N_w$: Number of sites in w
-\item $D_{w}$: Number of pairwise differences in w
-\item $Z_w$: Hidden IBD state in window w
-\item $H_w$: ROH state in window w
-\item A: Transition matrix
-\item $\pi$: Vector of initial transition probabilities
-\item $\theta$: Vector of $\pi$, A, $N_w$
-\item $\mathbf{Z}$: Vector of $Z_w$'s
-\item $\omega$:ROH states base
-\item $H_{w\omega}$: ROH state in $w$
-\item $s$: SNP
-\item $d_s$: pairwise difference at $s$
-\item $\mathbf{R_{si}}$: Vector of reads mapped to snp s for individual $i$
-\item $n_{si}$: Number of reads at SNP s for individual $i$
-\item $a_{si}$: Number of reads with alternate allele at SNP s for individual $i$
-\item $N_{wij}$: Number of SNPs overlapping for individuals $i$,$j$ in window $w$
-\item $D_{w}$: Number of SNPs in window w where a pair of individuals differ
-\item $\mathbf{D}$: Vector of $D_w$
-\item $\pi$: Initial transition probability
-\item $A$: Transition matrix
-\item $theta$: vector of additional parameters ($\pi$, $A$ and $N_w$)
-\item $\psi$: Matrix of parameters representing expected proportion of differences in different $I$ and $\omega$
-\item $psi_{I\omega}$: Element on row $I$ and column $\omega$ of $\psi$ matrix
-\item $x$: All combinations of $I$ and $\omega$ states with unique $\psi_{I\omega}$
-\item $p$: Proportion of differences in $x$
-\item $\alpha,\beta$: Parameters of Beta distribution
-\item $\delta$: $\alpha + \beta$ for relatedness model
-\item $\gamma_{wI}$: posterior probability of $I$ in $w$
-\item $k_{wI}$: Posterior probability of $x$ in $w$ 
-\item $\nu_{w}$: Allele frequency in window w
-\item X: Random variable
-\item r: Relatedness between individuals
-\item $\theta$: Vector of initial probability of being in an IBD state $I_l$
-\item $/mathbf{B}$: Beta function
-\item $B$: Reparameterized Beta function
-\item $Delta$: Heterozygosity in an individual
-\item Y: Hidden Homozygosity state
-\item $\Theta$: vector of additional parameters ($\pi$ $N_w$)
-\item $Gamma_{wY}$: Posterior probability of Y in w
-\item $/eta$: $/alpha + /beta$ for ROH model
-\item $c$: Contamination estimate for a library
-\item $p_c$: Average proportion of differences between endogenous and contaminating reads
-\item $p_e$:Average proportion of differences between endogenous reads
-\item $p_a$: Average observed proportion of differences
-\item $D_{cor,w}$ Contamination corrected number of differences in $w$
-\item $N_{cor,w}$ Contamination corrected number of overlapping sites in $w$
-\item $\Xi$: Endogenous read
-\item $S_{cor,w}$ $N_{cor,w}$ - $D_{cor,w}$
-\item $/mathbf{D_cor}$ Vector of $D_{cor,w}$ 
-\item E: Expectation
-\item P: Probability
-\end{itemize}
-
 \section{Results}
@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
+    \includegraphics[width=16cm]{plots/plotimg/IBDplot.png}
     \centering
-    \includegraphics[width=18cm]{plots/plotimg/IBDplot.png}
     \caption{Comparison of IBD fragments predicted by different relatedness models.
@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
+    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot.png}
     \centering
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/ROHplot.png}
     \caption{Estimation of ROH probabilities along the genome in simulations with different cases of ROH, contamination, and ascertainment.
@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
+    \includegraphics[width=16cm]{plots/plotimg/contam0_inbred0_model_performance_allroc_asc0_plot.png}
     \centering
-    \includegraphics[width=18cm]{plots/plotimg/contam0_inbred0_model_performance_allroc_asc0_plot.png}
     \caption{False positive and True Positive rates as a function of cutoff on log likelihood ratio. In top panel, dotted line represents $5\%$}.
@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
+    \includegraphics[width=16cm]{plots/plotimg/comparison_plot.png}
     \centering
-    \includegraphics[width=18cm]{plots/plotimg/comparison_plot.png}
     \caption{Comparison of KIn with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH}
@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
+    \includegraphics[width=18cm]{plots/plotimg/fil0_relatable_plot.png}
     \centering
-    \includegraphics[width=18cm]{plots/plotimg/fil0_relatable_plot.png}
     \caption{Application of KIn on Neandertal specimens from Chagyrskaya.
@@
 \renewcommand{\figurename}{Fig. S}
-\begin{figure}[h]
+\begin{figure}[h!]
+    \includegraphics[width=18cm]{supplementary_info/plots/binom.png}
     \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/binom.png}
     \caption{Comparison of fit with Betabinomial and Binomial distributions.
@@
 Results from KIn agree with results from READ, and provide some more information about siblings/parent-child, and $3^{rd}$ Degree relatedness. 
-\begin{figure}[h]
+\begin{figure}[h!]
+    \includegraphics[width=18cm]{supplementary_info/plots/fil0_read_plot.png}
     \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/fil0_read_plot.png}
     \caption{Application of READ on Neandertal specimens from Chagyrskaya cave.
@@
 Constraining variance of beta distributions in case of an identical pair, when identical relatedness hmm is applied, gives more accurate emission probabilities.
-\begin{figure}[h]
+\begin{figure}[h!]
+    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_emissions_bnds.png}
     \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_emissions_bnds.png}
     \caption{Application of identical relatedness HMM on a pair of identical individuals with low coverage (0.2x), and ROH tracts.
@@
 All probability distributions are scaled by the same factor so they are visible with the data histogram.
-\begin{figure}[h]
+\begin{figure}[h!]
+    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_betaplot.png}
     \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_betaplot.png}
     \caption{Comparison of beta distributions estimated with the HMM (left) without and (right) with variance constraints.}
@@
 Restricting the emission probabilities at 1 for Y=1, in windows that have higher than $p_0$ difference improves the estimated Beta parameters. 
-\begin{figure}[h]
+\begin{figure}[h!]
+    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_ind0_forced_roh.png}
      \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_ind0_forced_roh.png}
     \caption{Comparison of Beta distributions (left) without and (right) with modified emissions}
@@
 *Remember to use updated lcMLkin in the plot.
-\begin{figure}[h]
+\begin{figure}[h!]
     \centering
@@
 *Add 1M windows result
-\begin{figure}[h]
+\begin{figure}[h!]
     \centering
@@
 -supplementary table2
-\begin{figure}[h]
+\begin{figure}[!ht]
     \centering
@@
 \bibliography{references.bib}
+
+\section{Notation summary}
+Notation used in the paper:
+
+\begin{itemize}
+\item L: Total number of windows
+\item w: Index of window
+\item $N_w$: Number of sites in w
+\item $D_{w}$: Number of pairwise differences in w
+\item $Z_w$: Hidden IBD state in window w
+\item $i$: States that $Z_w$ can take 
+\item $H_w$: ROH state in window w
+\item $\omega$ States that $H_w$ can take
+\item A: Transition matrix
+\item $\pi$: Vector of initial transition probabilities
+\item $\theta$: Vector of $\pi$, A, $N_w$ for relatedness HMM
+\item $p(Z_w,H_w)$: Expected proportion of differences for given $Z_w$ and $H_w$
+\item $x$: Combinations of $i$ and $\omega$ states with unique $p$
+\item $\delta$: Over-dispersion parameter
+\item $\gamma_{wi}$: posterior probability of $H_w$
+\item $h_{\omega w}$: ROH state-probabilities
+\item $k_{wx}$: Posterior probability of $x$ in $w$
+\item $t$: Threshold on variance of Beta distribution
+\item $X$: Random variable from a Beta distribution
+\item $y$: Hidden homozygosity states for ROH HMM
+\item $Y_w$: Homozygosity state in $w$
+\item $\Theta$: Vector of $\pi$, A, $N_w$ for ROH HMM
+\item $\Delta_w$: Number of differences within a library 
+\item $\rho$: Expected proportion of differences in a library
+\item $Gamma$: Posterior probabilities of ROH
+\item $\eta$: Overdispersion parameter used in ROH HMM
+\item $\mathcal{C}$: Cost function
+\item $C_l$: Contamination in library l
+
+\item $c$: Contamination estimate for a library
+\item $p_c$: Average proportion of differences between endogenous and contaminating reads
+\item $p_e$:Average proportion of differences between endogenous reads
+\item $p_a$: Average observed proportion of differences
+\item $D_{cor,w}$ Contamination corrected number of differences in $w$
+\item $N_{cor,w}$ Contamination corrected number of overlapping sites in $w$
+\item $\Xi$: Endogenous read
+\item $S_{cor,w}$ $N_{cor,w}$ - $D_{cor,w}$
+\item $/mathbf{D_cor}$ Vector of $D_{cor,w}$ 
+\item E: Expectation
+\item P: Probability
+\end{itemize}
 

### 2022-02-18 supplementary_info/Rscripts/lcPlot.R
removed 6_14

@@ -33,6 +33,9 @@ lcplot <- function(lcres,likf,kinf,outplot){
   lc[lc$pair=='Chagyrskaya1141_Chagyrskaya19','Relatedness'] <- 'Identical'
   lc[lc$pair=='Chagyrskaya01_Chagyrskaya60','Relatedness'] <- 'Degree2'
   lc[lc$pair=='Chagyrskaya06_Chagyrskaya14','Relatedness'] <- 'Identical'
+
+  lc=lc[lc$pair!='Chagyrskaya06_Chagyrskaya14',]
+
   colors1 <- c("Unrelated" = "grey80", "Degree2" = "darkorchid1", "Parent-Child" = "orangered3", "Identical" = "darkgoldenrod1")
 
   panel1<-ggplot(data=lc[which(lc$loglik_ratio >=1),], mapping=aes(x=k0_hat, y=r_hat, color=Relatedness)) +
@@ -57,7 +60,7 @@ lcplot <- function(lcres,likf,kinf,outplot){
   newdf[newdf$pair=='Chagyrskaya01_Chagyrskaya60','Relatedness'] <- 'Degree2'
   newdf[newdf$pair=='Chagyrskaya06_Chagyrskaya14','Relatedness'] <- 'Identical'
 
-
+  newdf=newdf[newdf$pair!='Chagyrskaya06_Chagyrskaya14',]
 
 
   newdf$r_prop <- newdf$k1_prop/2 + newdf$k2_prop

### 2022-02-18 paper_overleaf/main.tex

@@
 \begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic1.png}
+    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib.png}
     \centering
-    \caption{Overview of KIn.
-A.
+    \caption{IBD sharing between siblings with and without ROH.
 In top row, recombined chromosomes inherited from parents by a pair of siblings are shown.
 On left, the parents have unrelated chromosomes, and on right one parent has homozygous pair of chromosomes.
-Bottom row shows the expected proportion of differences along the chromosome in both the cases.
-B.
-Overall schematic of the method.
-Blue boxes show the data files, while green boxes represent scripts.}
+Bottom row shows the expected proportion of differences along the chromosome in both the cases.}
     \label{fig0:schematic}
@@
 \subsection{Relatedness classes}
-This model is run for all the different cases of relatedness using the corresponding transition matrix, and the likelihood for all the relatedness models is compared.
+This model is run for different cases of relatedness including (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{rd}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, Identical) using the corresponding transition matrix, and the likelihood for all the relatedness models is compared.
+We realized that we have low power to detect $4^{th}$ and $5^{th}$ degree and we can not always differentiate these from unrelated.
+In our final output, we show classifications of $4^{th}$ and $5^{th}$ degrees as unrelated.
+Similarly, we observe that we have low power to differentiate between half-siblings, avuncular and grandparent-grandchild, and we merge these classifications as $2^{nd}$ Degree in the final results.
 The relatedness for the maximum likelihood model is reported, and the confidence is given by the log likelihood ratio between the two highest likelihood models.
-Hidden state of each window (IBD state) is estimated using standard Viterbi algorithm, and the IBD states corresponding to the most likely model are returned.
+Hidden IBD state of each window ($Z_w$) is estimated using standard Viterbi algorithm, and $\BZ$ corresponding to the most likely model are returned.
 
@@
 
-\paragraph{ Simulations for estimation of transition matrix}
 We use simulations both for estimating the transition matrices for all our models, and for testing and validating our algorithm.
 All simulations are performed using \textt{msprime} \cite{kelleher_efficient_2016}, followed by simulations of related individuals using a predetermined pedigree (Figure S1).
+
+\paragraph{ Simulations for estimation of transition matrix}
 
@@
 We estimated transition matrices corresponding to different cases of relatedness by counting transition between IBD states for a pair of  individuals at a particular relatedness level.
-We realize that we have low power to detect $4^{th}$ and $5{th}$ degree and cna differentiate these from unrelated.
-In our final output, we show classifications of $4_{th}$ and $5_{th}$ degrees as unrelated.
+We compared the transition matrices for grandparent-grandchild and siblings estimated in this way to a theoretical calculation, and verified that they are very close.
+However, it is difficult to theoretically calculate transition matrices for other cases like third degree, and so for these cases we use the transition matrices obtained from simulations.
 
 \paragraph{Model Comparison}
-model with maximum likelihood is reported, and log likelihood ratio is calculated with the two highest likelihood models.
+Model with maximum likelihood is reported, and log likelihood ratio is calculated with the two highest likelihood models.
 
-#### Simulations for model evaluation
 Apart from the related and unrelated individuals in pop1, we simulated more haploids in three other populations to create scenarios with ascertainment and contamination.
@@
 \section{Supplementary Information}
+
+Pipeline of KIn
+\begin{figure}[h!]
+    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sup.png}
+    \centering
+    \caption{Overall schematic of the method. Blue boxes show the data files, while green boxes represent scripts.}
+    \label{figS0:schematic}
+\end{figure}
+
 Beta binomial distribution is a better fit compared to Binomial distribution for proportion of differences in windows.

### 2022-02-18 paper_overleaf/main.tex

@@
 
-For our simulations of background diversity, we simulated a haploid population (pop1) with constant effective size of 10,000 and sampled 16 haploids from 2500 generations ago to randomly combine into 8 unrelated diploid ancient individuals.
+For our simulations of background diversity, we simulated a haploid population (Pop1) with constant effective size of 10,000 and sampled 16 haploids from 2500 generations ago to randomly combine into 8 unrelated diploid ancient individuals (fig.
+\ref{figS10:pedigree}).
 For each individual, we simulated 22 chromosomes with length $L=100$ Mb.
@@
 
-Apart from the related and unrelated individuals in pop1, we simulated more haploids in three other populations to create scenarios with ascertainment and contamination.
+Apart from the related and unrelated individuals in pop1, we simulated more haploids in three other populations to create scenarios with ascertainment and contamination (fig.
+\ref{figS10:pedigree}).
 We simulated two haploids to form an individual each from two other populations (pop2, pop3) with split time of 3500 and 4500 generations with pop1, and sampling time of 4000 generations and 2000 generations ago respectively.
@@
 
+Here, l is length of the chromosome.
 From the steps described above, we got genotypes of 17 individuals in pop1 in presence/absence of ROH and ascertainment.
-We further simulated 10 haploids (5 diploid individuals) from another population (pop4) with split time of 20,000 generations with pop1, sampled from the present time.
+We further simulated 10 haploids (5 diploid individuals) from pop4 with split time of 20,000 generations with pop1, sampled from the present time.
 We calculated the allele frequency for pop4 at each site, and used it to introduce contamination (with fixed contamination percentage for each individual ($<5\%$) in samples from pop1.
@@
 We specifically generated reads for four different cases: Ancestral-Endogenous, Ancestral-Contaminant, Derived-Endogenous, Derived-Contaminant.
-The number of reads for each case was sampled from a Poisson distribution with $\lambda$ calculated from the coverage, allele frequency or genotype (G $\in$ 0,0.5,1), and contamination estimate:
+The number of reads for each case was sampled from a Poisson distribution with $\lambda$ calculated from the coverage $\zeta$, allele frequency or genotype (G $\in$ 0,0.5,1), and contamination estimate:
 
@@
 Performance of ROH-HMM under different cases of ascertainment, contamination and long ROH, and for different coverages is shown in Fig.\ref{fig2:ROH}.
+We observe that as coverage decreases, data becomes more noisy.
+This causes ROH HMM to predict some probability of ROH in few windows when there is no ROH in data.
+However, at 4x and 0.5x, this is very rare.
+In presence of ROH, we see a similar trend.
+At 4x the model can estimate ROH probability well, but loses accuracy at 0.2x.
+In presence of ascertainment, overlapping sites are reduced, but they are informative.
+We see that the model performs well at 0.5x.
+New plot for contamination to be included.
 We applied the ROH-HMM and relatedness HMM on 60 runs of simulations, to get relatedness estimates, and compared the final relatedness estimates to the true relatedness for different pairs at coverages ranging from 4x to 0.1x.
@@
 Here, and in Fig.\ref{fig4:Comparison_READ_KIn}, we have two relatedness labels: 'Unrelated' and 'Unrelated w/o 3rd Degree'.
-Unrelated label here refers to KIn performance results when all unrelated, degree5, degree4, degree3 pairs are labelled as unrelated.
-This is useful for comparison to READ, since READ classifies relatedness upto degree2.
-'Unrelated w/o degree3'  refers to the performance results when degree 3 is considered as separate from the unrelated individuals.
-Fig.\ref{fig3:cutoff} shows that for identical, 1st degree, unrelated pairs, the false positive rate is below $5\%$ even when no cut off is used.
+Unrelated label here refers to KIn performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
+This is useful for comparison to READ, since READ classifies relatedness upto Second Degree.
+'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when Third Degree is considered as separate from the unrelated individuals.
+Fig.\ref{fig3:cutoff} shows that for Identical, First Degree, Unrelated pairs, the false positive rate is below $5\%$ even when no cut off is used.
 For second degree, at 0.1x we see a high false positive rate, that does not go down much with an increase in cutoff.
@@
 Fig.\ref{fig4:Comparison_READ_KIn} shows this comparison using false positive and true positive rates for both the methods at different coverages under different simulations.
-We show that both methods have similar performance in the control case (except for 0.1x coverage, where KIn has higher power to detect degree1 and degree2, and slightly higher false positive rate for degree 2).
-However, in case of ascertainment, ROH, or comtamination KIn shows higher power to detect relatedness cases while maintaining a low false positive rate.
+We show that both methods have similar performance in the control case (except for 0.1x coverage, where KIn has higher power to detect First Degree and Second Degree, and slightly higher false positive rate for degree 2).
+However, in case of ascertainment, ROH, or contamination KIn shows higher power to detect relatedness cases while maintaining a low false positive rate.
 
@@
 
-To test KIn on real ancient dataset, we applied it to neandertal specimens from Chagyrskaya cave in Siberia, Russia.
+To test KIn on real ancient dataset, we applied it to Neandertal specimens from Chagyrskaya cave in Siberia, Russia.
 This dataset comprised of 14 specimens that came from neighbouring layers, and hence could possibly belong to contemporary neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya \cite{kolobova_archaeological_2020-1}.
@@
 Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena.
-All these samples were captured with an array targeting for Neanderthal variation, and we further ascertained this data to variable sites in high coverage Neanderthal genomes: Altai Neanderthal (Denisova 5) (Prüfer et al., 2014) and Vindija 33.19 (Prüfer et al., 2017).
+All these samples were captured with an array targeting for Neanderthal variation, and we further ascertained this data to variable sites in high coverage Neanderthal genomes: Altai Neanderthal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}.
 This dataset had all the issues that KIn deals with, and hence we decided to apply it here to see if we could find family relationships that would not be possible with other available methods.
@@
 S\ref{figS2:Chagyrskaya_READ}).
-Interestingly, for the pair of Chagyrskaya07-Chagyrskaya17, READ shows that they are first degree relatives, while KIn further identifies them as parent-child.
+Interestingly, for the pair of Chagyrskaya07-Chagyrskaya17, READ shows that they are first degree relatives, and KIn further identifies them as parent-child.
 Our result matches with the conclusions of the authors based on additional information from mtDNA.
 Further, we find a possible third degree relative (Chagyrskaya07-Chagyrskaya60).
-We compare our results to lcMLkin, and show that the software does not perform well with these low coverage samples since it is not possible to obtain reliable genotype likelihoods for many of these samples (fig.
+We compared our results to lcMLkin, and show that the software does not perform well with these low coverage samples since it is not possible to obtain reliable genotype likelihoods for many of these samples (fig.
 S\ref{figS6:Chagyrskaya_ibd}).
@@
 Pipeline of KIn
+
+\renewcommand{\figurename}{Fig. S}
 \begin{figure}[h!]
@@
 
-\renewcommand{\figurename}{Fig. S}
 \begin{figure}[h!]
@@
 Binomial parameter p is calculated from the data, while Betabinomial parameters are estimated with the relatedness HMM.}
-    \label{FigS1:binom}
+    \label{figS1:binom}
+
+\end{figure}
+
+\begin{figure}[h!]
+    \includegraphics[width=18cm]{plots/inkscape_finalImg/pedigree.png}
+    \centering
+    \caption{Overview of simulated dataset.
+We simulate pop1, and artificially mate unrelated individuals to make a family with upto 5th degree relatives, along with other unrelated and identical individuals (not shown here).
+Pop2 and Pop3 are simulated so as to introduce ascertainment, while Pop4 is a present day population used to simulate contamination.
+Split times and age of the populations are shown in generations.}
+    \label{figS10:pedigree}
 
@@
 \item $p_a$: Average observed proportion of differences
-\item $D_{cor,w}$ Contamination corrected number of differences in $w$
-\item $N_{cor,w}$ Contamination corrected number of overlapping sites in $w$
+\item $D_{cor,w}$: Contamination corrected number of differences in $w$
+\item $N_{cor,w}$: Contamination corrected number of overlapping sites in $w$
 \item $\Xi$: Endogenous read
@@
 \item $/mathbf{D_cor}$ Vector of $D_{cor,w}$
+\item $G$: Genotype (0,0.5,1)
+\item $\lambda$: Expected rate following Poisson distribution
+\item $\nu$: Population allele frequency
+\item $Lambda$: Log likelihood ratio
 \item E: Expectation

### 2022-02-18 plots/Rscripts/kinPlot.R
edited the kin and read plots to remove labels on x and y

@@ -59,12 +59,12 @@ bubbleplot_kin <- function(overlap_table, relatable, out_rel, out_over) {
     geom_tile(aes(fill=relatedness), col="white")  +
     scale_fill_manual(values = c("Identical"="khaki1","Parent-Child"="orangered3", "Siblings"="orange2","Unrelated"="gray80","Second Degree"="darkorchid1","Third Degree"="aquamarine2")) +
 
-    labs(title = "relatedness", x = "lib_ID", y = "lib_ID")+
     theme(
       panel.border = element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
       panel.background = element_rect(fill = "white",colour = "white", size = 0.5, linetype = "solid"))+
     theme(axis.text.x = element_text(angle = 90,hjust = 1, colour = colors, size = 14)) +
     theme(axis.text.y = element_text(angle = 0,hjust = 1, colour = colors, size = 14)) +
+    theme(axis.title.x=element_blank(), axis.title.y=element_blank()) +
     geom_text(aes(lib1, lib2, label = matr1$loglik_ratio), color = "black", size = 5)#Plot overlap +
   theme(legend.text = element_text(size=14),
         legend.title=element_text(size=15))
@@ -74,12 +74,12 @@ bubbleplot_kin <- function(overlap_table, relatable, out_rel, out_over) {
   gfig_over=ggplot(matr1, aes(lib1, lib2), col= "white") +
     geom_tile(aes(fill=overlap), col="white")  +
     scale_fill_manual(values = c("gray90","darkslateblue","dodgerblue3","mediumseagreen","palegreen","yellowgreen","skyblue4","khaki1","firebrick")) +
-    labs(title = "Overlap filter", x = "lib_ID", y = "lib_ID")+
     theme(
       panel.border = element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
       panel.background = element_rect(fill = "white",colour = "white", size = 0.5, linetype = "solid"))+
     theme(axis.text.x = element_text(angle = 90,hjust = 1, colour = colors, size = 14)) +
     theme(axis.text.y = element_text(angle = 0,hjust = 1, colour = colors, size = 14)) +
+    theme(axis.title.x=element_blank(), axis.title.y=element_blank()) +
     theme(legend.text = element_text(size=14),
           legend.title=element_text(size=15))
 

### 2022-02-18 supplementary_info/Rscripts/readPlot.R

@@ -66,12 +66,12 @@ bubbleplot_read <- function(overlap_table, relatable, out_rel, out_over) {
   gfig_rel=ggplot(matr1, aes(lib1, lib2), col= "white") +
     geom_tile(aes(fill=relatedness), col="white")  +
     scale_fill_manual(values = c("khaki1","orangered3","mediumorchid4","gray80")) +
-    labs(title = "relatedness", x = "lib_ID", y = "lib_ID")+
     theme(
       panel.border = element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
       panel.background = element_rect(fill = "white",colour = "white", size = 0.5, linetype = "solid"))+
     theme(axis.text.x = element_text(angle = 90,hjust = 1, colour = colors, size = 14)) +
     theme(axis.text.y = element_text(angle = 0,hjust = 1, colour = colors, size = 14)) +
+    theme(axis.title.x=element_blank(), axis.title.y=element_blank()) +
     geom_text(aes(lib1, lib2, label = matr1$dist_thresh), color = "black", size = 5) + #Plot overlap
     theme(legend.text = element_text(size=14),
           legend.title=element_text(size=15))
@@ -79,14 +79,15 @@ bubbleplot_read <- function(overlap_table, relatable, out_rel, out_over) {
   gfig_over=ggplot(matr1, aes(lib1, lib2), col= "white") +
     geom_tile(aes(fill=overlap), col="white")  +
     scale_fill_manual(values = c("gray90","darkslateblue","dodgerblue3","mediumseagreen","palegreen","yellowgreen","skyblue4","khaki1","firebrick")) +
-    labs(title = "Overlap filter", x = "lib_ID", y = "lib_ID")+
     theme(
       panel.border = element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
       panel.background = element_rect(fill = "white",colour = "white", size = 0.5, linetype = "solid"))+
     theme(axis.text.x = element_text(angle = 90,hjust = 1, colour = colors, size = 14)) +
     theme(axis.text.y = element_text(angle = 0,hjust = 1, colour = colors, size = 14)) +
+    theme(axis.title.x=element_blank(), axis.title.y=element_blank()) +
     theme(legend.text = element_text(size=14),
           legend.title=element_text(size=15))
+
 
 
   ggsave(out_rel,height=12,width=14, plot = gfig_rel)

### 2022-02-18 paper_overleaf/main.tex

@@
 
-Even though this may not hold true when we bin sites in large windows, we still find that $Var(X_{x_0}) > Var(X_{x_1}) > Var(X_{x_2})$ (fig. \ref{FigS1:binom}).
+Even though this may not hold true when we bin sites in large windows, we still find that $Var(X_{x_0}) > Var(X_{x_1}) > Var(X_{x_2})$ (fig. \ref{figS1:binom}).
 
@@
 Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_kin}.
-We found identical individuals (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), first degree relatives (Chagyrskaya07-Chagyrskaya17, Chagyrskaya06-Chagyrskaya14), and a second degree relative (Chagyrskaya01-Chagyrskaya60).
+We found identical individuals (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), first degree relative (Chagyrskaya07-Chagyrskaya17), and a second degree relative (Chagyrskaya01-Chagyrskaya60).
 These results agree with READ (fig.
 S\ref{figS2:Chagyrskaya_READ}).
+Chagyrskaya06-Chagyrskaya14 were estimated to be First Degree by both the methods, but have been classified as Identical pair in the original study, since Chagyrskaya06 is a tooth sample that fits in mandible which is Chgyrskaya14.
+We believe that the contradiction is due to non human contamination present in these libraries ($1\%$ and $2.9\%$ respectively).
 Interestingly, for the pair of Chagyrskaya07-Chagyrskaya17, READ shows that they are first degree relatives, and KIn further identifies them as parent-child.
@@
 Further, we find a possible third degree relative (Chagyrskaya07-Chagyrskaya60).
-We compared our results to lcMLkin, and show that the software does not perform well with these low coverage samples since it is not possible to obtain reliable genotype likelihoods for many of these samples (fig.
+We compared our IBD estimates to lcMLkin for all pairs for whom READ results (with s.d.
+>1) and KIn results ($\Lambda$>1) match (fig.
 S\ref{figS6:Chagyrskaya_ibd}).
+Chagyrskaya06-Chagyrskaya14 is removed from the analysis.
+We find that lcMLkin does create four different clusters for different relatedness.
+However, the clusters are quite deviated from the expected values, which makes it difficult to identify the relatedness.
+Identical individuals, for example, are expected to be at $k_0 = 0$ and $r = 1$, and Parent-Child at $k_0 = 0$ and $r = 0.5$.
+These deviations can be due to ascertainment, contamination, ROH as well as low coverage of samples.
 
@@
 Out of 6903 pairs, 346 pairs show different relatedness for KIn and READ (table S2).
-However, there are only 31 pairs where KIn has $log likelihood >1$.
-We looked in detail at some of these pairs, and we found that it is possible to deduct the correct relatedness for several of these pairs by plotting the proportion of differences along the genome.
+However, there are only 31 pairs where KIn has $\Lambda >1$.
+We looked in detail at some of these pairs, and we found that it is possible to deduce the correct relatedness for several of these pairs by plotting the proportion of differences along the genome.
 In all the pairs that we analyzed in this way, KIn seems to predict the true relatedness.
@@
 S\ref{figS8:eg1}.
-We found 20 pairs in this subset of 346 pairs where lcMLkin has sufficient data.
+We found 20 pairs in this subset of 346 pairs where lcMLkin has sufficient data (Overlapping SNPs $>$ 10000).
 Among these 20 pairs, we found that lcMLkin matches KIn estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLkin predicts 3rd-5th degree, KIn predicts second degree) and pair AITI72-AITI77A (READ and lcMLKin predict unrelated, while KIn estimates 3rd degree with log likelihood=1.7).
@@
 Several methods are available to estimate relatedness from modern as well as ancient DNA data.
-We develop a new method to estimate genetic kinship from low coverage data.
+We developed a new method to estimate genetic kinship from low coverage data.
 We show that it is possible to accurately infer relatedness, along with IBD tracts, from such data in presence of long ROH, contamination, as well as ascertainment bias.
@@
 We use this approach to identify the following relatives: Identical individuals, Parent-Child, Siblings, Second degree, Third degree, Unrelated individuals.
-We show that it is possible to differentiate between avuncular and grand-parent-grandchild relations.
-We have developed a similar Hidden Markov Model to infer long ROH, and use the posterior probabilities to make better relatedness inference.
+We have developed a similar Hidden Markov Model to infer long ROH, and we use the posterior ROH probabilities to make better relatedness inference.
 We further use a simple model to correct the pairwise differences in genomic windows for contamination.
@@
 We demonstrate using simulations, that our method performs well at low coverage.
-We compared KIn's model performance to a widely used software READ to show that our approach works at least as good as READ on ideal data without long ROH, ascertainment and contamination, while additionally distinguishing between parent-child and siblings, and identifying third degree relatives.
+We compared KIn's model performance to a widely used software READ to show that our approach gives similar predictions as READ on ideal data with 4x coverage, and without long ROH, ascertainment and contamination, while additionally distinguishing between parent-child and siblings, and identifying third degree relatives.
 We demonstrate that at low coverage (0.1x) we have higher power to detect first degree relatives compared to READ with similar false positives, while a higher power to identify second degree relatives at the cost of slightly higher false positive rate.
@@
 And so long ROH causes higher power for READ which otherwise is a conservative method.
-KIn performs well for the same reasons, but can miss-classify second degree individuals with ROH as siblings, if ROH is not accounted for.
-Hence, we implement ROH states in the relatedness model to avoid such miss classifications.
+KIn performs well for the same reasons, but can additionally differentiate between different combinations of IBD and ROH in genomic windows.
 Contamination, on the other hand can make the performance of READ much worse depending on the amount of contamination, and the number of individuals contaminated.
@@
 On the other hand, increase in amount of contamination in few individuals can make them look more unrelated.
-KIn, with the implementation of a model to account for contamination performs well in this case.
+KIn, with the implementation of a model to correct for contamination performs well in this case.
 Ascertainment causes a decrease in power for both the methods, since it reduces the amount of data.
@@
 As expected from the simulations, KIn is helpful in differentiating between parent-child from siblings, and identifying second and third degree individuals in cases where there is an overlap of only few thousand sites or less, and the other methods are not informative.
-We show that when applied to neandertal specimens from Chagyrskaya cave, READ and KIn agree in relatedness estimates for all pairs where both methods have sufficient data.
-KIn additionally identifys a pair of first degree relatives as parent-child, which is also confirmed with mtDNA.
+We show that when applied to Neandertal specimens from Chagyrskaya cave, READ and KIn agree in relatedness estimates for all pairs where both methods have sufficient data.
+KIn additionally identifys a pair of first degree relatives as parent-child, which is in agreement with mtDNA analysis from the authors.
 In addition, KIn identifies a third degree relative.
@@
 Output of KIn is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio.
-Due to the tabular output, KIn is easy to automatise and apply on large datasets.
-One limitation with our approach is that this method assumes a single population.
-In case of population structure, with the subpopulations with very different diversity, KIn may show inaccurate inferences.
+Due to the tabular output, KIn is easy to automatise and apply to large datasets.
+One limitation with our approach is that it assumes a single population.
+In case of population structure, where the subpopulations have very different diversity, KIn may show inaccurate inferences.
 Our method makes another assumption, that the dataset comprises of at least some unrelated individuals.

### 2022-03-14 plots/plot_script.py
comparison plot from hmm_numba_id_corr

@@ -11,12 +11,12 @@ def ROH_plotf(fold,outfold):
     ind=0
     cov=4
     f=0
-    rohf=fold+"hbd_hmm_corr/contam%s/inbred%s/run%s/coverage%s/filtered%s/asc%s_res/gamma_hapProbs/pw_%s.csv.gz" %(c,i,r,cov,f,a,ind)
+    rohf=fold+"hbd_hmm/contam%s/inbred%s/run%s/coverage%s/filtered%s/asc%s_res/gamma_hapProbs/pw_%s.csv.gz" %(c,i,r,cov,f,a,ind)
     roh=pd.read_csv(rohf, sep=",", header=0,index_col=0)
     roh.loc[roh['g_noin']>1,'g_noin']=np.nan
 
-    difff=fold+"contam%s/inbred%s/run%s/coverage%s/asc%s/inputMode_idhapProbs_fil%s/id_remHighdiv_diff.csv.gz" %(c,i,r,cov,a,f)
-    totalf=fold+"contam%s/inbred%s/run%s/coverage%s/asc%s/inputMode_idhapProbs_fil%s/id_remHighdiv_total.csv.gz" %(c,i,r,cov,a,f)
+    difff=fold+"contam%s/inbred%s/run%s/coverage%s/asc%s/inputMode_idhapProbs_fil%s/id_wind.csv.gz" %(c,i,r,cov,a,f)
+    totalf=fold+"contam%s/inbred%s/run%s/coverage%s/asc%s/inputMode_idhapProbs_fil%s/id_wint.csv.gz" %(c,i,r,cov,a,f)
 
     diff=np.loadtxt(difff, dtype='float', delimiter = ",")[:,ind]
     total=np.loadtxt(totalf, dtype='float', delimiter = ",")[:,ind]
@@ -28,9 +28,9 @@ def ROH_plotf(fold,outfold):
     hbdd=np.loadtxt(truedf, dtype='float', delimiter = ",")[:,ind]
     hbdt=np.loadtxt(truetf, dtype='float', delimiter = ",")[:,ind]
 
-    plt.plot(diff/total)
-    plt.plot(roh['g_noin']/20)
-    plt.plot(hbdd/hbdt/20)
+    #plt.plot(diff/total)
+    #plt.plot(roh['g_noin']/20)
+    #plt.plot(hbdd/hbdt/20)
 
     df1=pd.DataFrame(
         columns=['Chrm', 'Prop', 'Coverage',
[…]

### 2022-03-14 paper_overleaf/main.tex

@@
 \usepackage{amsmath}
+\usepackage{amsfonts}
 \usepackage{bm}
@@
 \newcommand{\BD}{\mathbf{D}}
+\newcommand{\BN}{\mathbf{N}}
 \newcommand{\BH}{\mathbf{H}}
@@
 This can affect the relatedness estimates.
-Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness upto 3rd degree, while additionally differentiating between siblings and parent-child.
-We have developed another HMM to estimate ROH proportions, that can be utilized by the relatedness estimation model to output better estimates.
+Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness up to 3rd degree, while additionally differentiating between siblings and parent-child.
+The output can be further refined with another HMM that we developed to estimate ROH proportions.
 \end{abstract}
@@
 in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}.
-It is also an essential preprocessing step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.
+It is also an essential pre-processing step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.
 
-In ancient DNA, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an  understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_kinship-based_2019,sikora_ancient_2017}.
+In ancient DNA, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an  understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_KINship-based_2019,sikora_ancient_2017}.
 For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals in the mating animals is important to ensure diversity.
@@
 
+These analyses use cranial and dental metric and morphological and postcranial non-metric data to infer inter-individual relatedness based on the assumption that the phenotype reflects underlying genetic variation and patterns of relatedness.
+Multifactorial inheritance complicates straightforward interpretations, however, and given the moderate to low (and population specific) heritabilities of most morphometric traits (see Scott & Turner, 1997 for an overview; also Townsend et al., 2009), the effects of development on final phenotypic form are difficult to specify.
+
 \subsection{Simple way to get relatedness, methods that do it with a lot of data}
 
-Commonly, pairs of relateded individuals are  identified by looking for parts of the genome that are identical by descent (IBD), ie.
+Commonly, pairs of related individuals are  identified by looKINg for parts of the genome that are identical by descent (IBD), ie.
 inherited from a recent common ancestor.
@@
 
-However, it is not possible to directly observe IBD segments, and so most methods infer them by first identifying segments of the genome that are Identical by State (IBS) and using population allele frequencies to calculate the probability of IBD given IBS \cite{vai_kinship_2020}.
+However, it is not possible to directly observe IBD segments, and so most methods infer them by first identifying segments of the genome that are Identical by State (IBS) and using population allele frequencies to calculate the probability of IBD given IBS \cite{vai_KINship_2020}.
 One exception is SNPduo, a software that uses IBS direclty to visualize and analyze pairwise relatedness.
@@
 For example, a pair of siblings are expected to have all three possible IBD states with probabilities (0.25,0.5,0.25) as shown in Fig \ref{fig0:schematic}.
-These IBD probabilities can directly be used to categorize a pair of individuals as related by a certain degree and nature of relatedness as shown in table \ref{tab:Table 1}.
+These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}.
 One can also use these probabilities to estimate the coefficient of relatedness as $r= k_1/2 + k_2$.
@@
     \centering
-    \caption{IBD sharing between siblings with and without ROH.
+    \caption{IBD sharing between siblings without and with ROH.
 In top row, recombined chromosomes inherited from parents by a pair of siblings are shown.
-On left, the parents have unrelated chromosomes, and on right one parent has homozygous pair of chromosomes.
-Bottom row shows the expected proportion of differences along the chromosome in both the cases.}
+On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes.
+The bottom row shows the expected proportion of differences (p) along the chromosome in both the cases.
+p can take values of $p_0$, $p_1$ or $p_2$ which are expected proportion of differences in IBD states 0,1 and 2 respectively.
+}
     \label{fig0:schematic}
@@
 There are numerous methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to estimate the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}.
-PLINK, a popular tool to infer relatedness coefficient in  diploid genotype data, estimates IBD probabilities from average of observed IBS, and using the allele frequencies at each SNP in samples \cite{purcell_plink_2007}.
+PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from average of observed IBS, using the allele frequencies at each SNP in samples \cite{purcell_plink_2007}.
 KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}.
@@
 \subsection{Problems with ancient data, and methods that address these issues}
-One major issue with applying above mentioned methods to ancient DNA is that, frequently, the amount of endogenous DNA is very low, making it very difficult to get genotype calls.
-Several methods solve this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}.
-These methods account for the uncertainity in genotype calls by summing over all possible genotypes, weighted by the genotype likelihoods.
+One major issue with applying the above mentioned methods to ancient DNA is that, frequently, the amount of endogenous DNA is very low, maKINg it very difficult to get genotype calls.
+Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}.In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods.
 However, these approaches typically still require at least 2x coverage, since genotype likelihoods may not be very informative at lower coverages.
-Apart from the low amount of endogenous DNA sequences present, ancient DNA analyses often have additional problems like contamination from modern populations \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in case of captured sequences.
-In cases where allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations.
+Apart from the low amount of endogenous DNA sequences present, ancient DNA analyses often have additional challenges like contamination from modern populations \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in case of captured sequences.
+In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations.
 However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}.
-Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4x$ coverage to get a good estimate of genotype likelihoods\cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}.
+Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4x$ coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}.
 In this study, we show that contamination from modern humans can cause some of the individuals in ancient populations to look more unrelated than they are.
 This happens because of comparison of endogenous reads to contaminating reads which would give higher differences than comparison of endogenous reads from individuals from ancient population.
+It is worth mentioning that different DNA libraries may have different proportion of contamination levels, even when they have been sampled from the same specimen.
 Hence, contamination can reduce the power to detect related pairs, especially when the target population is quite diverged from contaminating population(s).
@@
 This effect may cause an increase in false positives.
-In many cases, ancient DNA is captured with a SNP array, and the SNPs are ascertained to variable sites determined from certain individuals.
-Relatedness methods based on the fraction of sites in different IBS states are highly sensitive to ascertainment bias.
-KING robust, for example, differentiates between different relatedness by plotting KING‐robust kinship against fraction of sites in IBS0 \cite{manichaikul_robust_2010}.
+Moreover, ancient DNA is commonly captured with a SNP array to enrich for endogenous sequences.
+The way that these SNPs are defined (e.g.
+selecting SNPs variable in certain individuals) becomes relevant, because relatedness methods based on the fraction of sites in different IBS states are highly sensitive to ascertainment bias.
+KING robust, for example, differentiates between different relatedness by plotting KING‐robust KINship against fraction of sites in IBS0 \cite{manichaikul_robust_2010}.
 Another method from Rosenberg,2006 \cite{rosenberg_standardized_2006} uses the fraction of sites in all three IBS states to classify related individuals in the HGDP-CEPH Human Genome Diversity Cell Line Pane.
@@
 
-READ is a software that does not require population allele frequencies or a reference panel, and can work with around 0.5 x coverage in presence of ascertainment bias.
-The software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual.
-A string of these alleles at each position (called pseudohaploids) are then compared to other individuals to calculate pairwise genetic distances, which in turn are used to infer relatedness.
-The limitations are that the software can identify upto only second degree relatives, without differentiating between siblings and parent-child, which may be crucial when making pedigrees.
+Several limitations with ancient DNA mentioned above, such as low coverage, the absence of population allele frequencies or a reference panel, and ascertainment bias are resolved by a software called READ.
+This software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual.
+A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to calculate average pairwise genetic distances, which in turn are used to infer relatedness.
+The limitations are that the software can identify only up to second degree relatives, without differentiating between siblings and parent-child, which may be crucial when maKINg pedigrees.
 
 \subsection{How our method works}
-Here, we present KIn (Kinship Inference), a Hidden Markov Model (HMM) to estimate relatedness from low-coverage ancient DNA data.
-KIn can detect up to \nth{3} degree relatives, and differentiates between siblings and parent-child.
-KIn is also able to incorporate ROH, contamination and ascertainment bias.
-We validate the performance of KIn using simulations and show that we are able to infer relatedness in a neandertal data set, and a Bronze age dataset.
+Here, we present KIN (KINship Inference), a Hidden Markov Model (HMM) to estimate relatedness and IBD from low-coverage ancient DNA data.
+KIN can detect up to \nth{3} degree relatives, and differentiates between siblings and parent-child.
+KIN is also able to take into account ROH, contamination and ascertainment bias.
+We validate the performance of KIN using simulations and show that we are able to infer relatedness in real data from two datasets: a group of Neandertals and a group of Bronze age individuals.
 
@@
 
-The core of  KIn, is a HMM that aims to infer relatedness and IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts in each sample, and contamination estimates into account.
-If the locations of ROH tracts are unknown, we provide another HMM to estimate ROH in the first step for a sample of sufficient coverage ($\geq 0.5x$).
-In a second step, the relatedness-HMM uses the contamination-corrected differences between a pair of individuals, and the positions of ROH tracts, to infer relatedness.
-Our method is available on \url{https://github.com/DivyaratanPopli/KIn_snakemake} along with a \textt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.
+The core of  KIN is a HMM that aims to infer relatedness and IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts and contamination estimates in each individual into account.
+If the locations of ROH tracts are unknown, we provide another HMM to coarsely estimate the location of ROH for samples  of sufficient coverage ($\geq 0.5x$).
+Our method is available on \url{https://github.com/DivyaratanPopli/KIN_snakemake} along with a \textt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.
 
 \subsection{Relatedness Model overview} 
-To estimate the relatedness of two individuals, we subdivide the genome into $L$ large windows $w$ (typically of size 10Mb).
-The inputs of our algorithm are i) the number of sites $N_w$ for which both samples have at least one read available, ii) the number of pairwise differences $D_w$ at these sites, and iii) probability of ROH in in windows, by default obtained from ROH-HMM described in a following section.
-Calculation of $N_w$ and $D_w$ requires comparison of reads at all positions for a pair of individuals.
-For low coverage genomes, this can be done by comparing randomly sampled read at each position for a pair of individuals (or by comparing pseudo haploid sequences).
-However, we realized that such random sampling can throw away a lot of useful data.
-To avoid this wastage, our input generation pipeline for Relatedness HMM calculates proportion of reads with alternate allele, at all positions for each genome.
-We use these proportions to find pairwise heterozygosity as the probability of comparing different alleles.
-These probabilities are summed up in each window to give $D_w$, along with the number of overlapping sites $N_w$.
+Our relatedness model assumes we have low-coverage genome sequences for a number of individuals from the same population, and our goal is to infer which pairs of individuals are related.
+We subdivide the genome into $L$ large windows (typically of size 10Mb).
+The inputs of our algorithm are i) the number of overlapping sites for the $w$-th window $N_w$ for which both samples have at least one read available, ii) the expected number of pairwise differences $D_w$ at these sites, and iii) the probability of ROH in windows, by default obtained from ROH-HMM described in a following section.
+$D_w$ is the number of sites that are different between a pair of individuals, and can be estimated accurately for high coverage data by comparing genotypes.
+For low coverage samples, $D_w$ is typically calculated by comparing randomly-sampled reads (or pseudo-haploid sequences) \cite{haak_massive_2015}.
+However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
+\begin{align}\label{eq:x}
+D_w &= \sum_s \nu_i(s) (1-\nu_j(s)) + (1-\nu_i(s)) \nu_j(s)
+\end{align}
+Here, $\nu_i(s)$ and $\nu_j(s)$ are the derived allele frequencies for individuals i and j respectively, at SNP index $s$ in a window.
+Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g.
+$\BD = (D_1, D_2, \dots D_L)$.
+The KIN-HMM uses $\BD$ and $\BN$  to classify each window into three hidden states $Z_w \in$ ($0$, $1$, $2$), reflecting zero, one or two chromosomes shared IBD.
+Likewise, we classify each window into one of three possible ROH state $H_w \in$ ($0$, $1$, $2$), reflecting, zero, one or both individuals being in a run-of-homozygosity at this genomic location.
 
-Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$.
+There are two additional parameters, $\pi$ and $\mathbf{A}$.
+The initial probabilities $\pi$ give the probabilities of being in each state $Z$ at the beginning of each chromosome, which we set to uniform for simplicity.
+The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed for each relatedness category (see section YYY).
+For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.
 
-The model then uses this data for a given pair of individuals to classify each window into three hidden states $Z_w \in$ ($i_0$, $i_1$, $i_2$), reflecting zero, one or two chromosomes shared IBD.
-Likewise, we classify each window into one of three possible ROH state $H_w \in$ ($\omega_0$, $\omega_1$, $\omega_2$), reflecting, neither, one or both individuals being in a run-of-homozygosity at this genomic location.
+We compute the complete data likelihood for the HMM:
 
-Ideally, we would like to compute the likelihood $P(\BD | \BZ, \BH, \theta)$, where $\theta$ is a vector of parameters (initial transition probability $\pi$, transition matrix $A$ and $N_w$).
-
-However, calculating this likelihood is hard because the number of possible states for $\mathbf{Z}$ is very large, and we use a standard trick for fitting HMMs, to base or inference on the complete data likelihood:
-
-\begin{align}
-P(\mathbf{D},\BZ, \BH|\theta) &= P(\mathbf{D}|\mathbf{Z},\BH, \theta) P(\BZ |\theta)P(\BH |\theta) \nonumber\\
-&= \prod_w P(D_w|Z_w,H_w, \theta) \prod_w P(H_w |\theta) \prod_w P(Z_w |Z_{w-1},\theta) P(Z_0|\theta)
-\label{eq:1}
+\begin{align}\label{eq:1}
+\log P(\BD,\BZ|\BH, \theta) &= \log P(\BD|\BZ,\BH, \theta) + \log P(\BZ |\theta) \nonumber\\
+&= \sum_w \log P(D_w|Z_w,H_w, \theta) + \sum_w \log P(Z_w |Z_{w-1},\theta) + \log P(Z_0|\theta)
 \end{align}
 
-Here, $P(D_w|Z_w,  H_w, \theta)$ is the emission probability for window $w$,  $P(Z_w|Z_{w-1},\theta)$ is the transition probability, and $P(Z_0| \theta)$ is the initial probability $\pi$.
-$P(H_w |\theta)$ is the output of ROH HMM described in a following section.
-
-Using the complete data likelihood allows us to factor the emission and transition probabilities, and hence we can maximize them independently.
-Transition matrix for different cases of relatedness are either derived from theory, or obtained from simulations (see section). Initial transition probability $\pi_i$ is same for all the $i$.
-
 \subsection{Emission probability}
-In equation \ref{eq:1} the term $P(D_w | Z_w, H_w, \theta) = P(D_w | Z_w, H_w, N_w)$ is the emission probability, i.e.
-the probability of our data , given a particular hidden state.
+Using this set up, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta)$ from equation \ref{eq:1} and optimize them alone.
 The number of shared sites $N_w$ is the only parameter in $\theta$ that affects the emissions.
 
-Assuming sites are equally distributed and independent, we could use a  binomial likelihood
-$$P(D_w|Z_w, N_w, H_w) \sim \text{Binom}[D_w ;N_w, p(Z_w, H_w)] \text{,}$$
+Assuming sites are equally distributed and independent would result in the binomial likelihood:
+$$P(D_w|Z_w, H_w, N_w) \sim \text{Binom}[D_w ; p(Z_w, H_w), N_w] \text{,}$$
 
 where, $p$ is the expected proportion of differences we would expect for a particular IBD and ROH state.
+If the two individuals are unrelated in a particular window (i.e.
+$Z_w = 0$), then the expected proportion of pairwise differences depends solely on the effective population size, and we denote this proportion with $p_0$.
+Note that $p_0$ does not depend on the presence of ROH, since this does not change expected proportion of differences between unrelated individuals.
+If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes \cite{kuhn_estimating_2018}.
+Thus, $p(Z_w=i, H_w=0) = p_i$.
 
-If the two individuals are unrelated in a particular window and neither is in a ROH region (i.e.
-$Z_w = i_0$, and $H_w = \omega_0$), then the expected proportion of pairwise differences depends solely on the effective population size, and we denote this proportion with $p_0$.
-If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes.
+The proportion of differences between unrelated individuals $p_0$ is an important parameter.
+We follow READ \cite{kuhn_estimating_2018} and estimate it as the median of differences for all possible pairs of individuals, which should work well if the majority of individuals in the samples are unrelated.
+We set $p_1$ to $(3/4)p_0$ and $p_2$ to $p_0/2$.
 
-$p_0$, $p_1$, and $p_2$ are the expected proportion of differences in $i_0$, $i_1$, and $i_2$.
-$p_0$ is calculated directly from the data by calculating median of differences for all possible pairs of specimens.
-We calculate $p_1$ as $(3/4)p_1$ and $p_2$ as $p_1/2$.
-However, we need one more parameter (represented as $p_3$ in the matrix below) in cases where IBD state is $i_1$ and ROH state is $\omega_2$, or IBD state is $i_2$ and ROH state is $\omega_1$ or $\omega_2$.
-In such cases the expected proportion of differences between the individuals should be 0.
-However, as mentioned before, we do all our calculations in large windows, and start/end positions of these windows may not coincide with that of ROH tracts.
-And so, the pairwise differences in this state (called $p_3$ in the table) can lie anywhere between 0 to $p_2$.
-We found $p_3$ = $p_2$/2 to be a reasonable assumption for expected pairwise proportion of differences in this IBD/ROH state.
+The presence of inbreeding adds an additional complication, as it is possible that all chromosomes are shared between individuals, so that the expected differences should be zero.
+An example would be inbred regions in two bones from the same individual.
+However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that setting expected proportion of differences in this case ($p_3$) at $p_2/2$ works well in practice.
 
-Taken together, we can summarize $p$ as follows 
+Taken together, we can summarize $p$ in the following matrix, where rows give the state of $Z_w$, and columns of $H_w$:
 \begin{equation}\label{eq:2}
@@
 {rrr}
-p_0 & p_1 & p_2 \\
-p_0 & p_2 & p_3 \\
-p_0 & p_4 & p_3
+p_0 & p_0 & p_0 \\
+p_1 & p_2 & p_3 \\
+p_2 & p_3 & p_3
 \end{array}\right]
@@
 
-The effect of these considerations is that even though we have nine possible combinations of $Z$ and $H$ for each window, there are actually only four different $p$-parameters $p_x$ with $x$ $\in$ (0,1,2,3).
+The effect of these considerations is that even though we have nine possible combinations of $Z_w$ and $H_w$ for each window, there are actually only four unique $p$-parameters $p_i$ with $i$ $\in$ (0,1,2,3).
 
-However, we find that the data often has considerably higher variance than would be expected from a binomial distribution, likely because we average over a large number of SNPs with different allele frequencies.
+\paragraph{Beta Binomial Model}
+The Binomial Model laid out above assumes that sites are independent.
+However, due to genetic linkage, nearby sites will be correlated and we find that the data often has considerably higher variance than would be expected from a binomial model.
+We take this into account by adding an overdispersion parameter $\delta$ that allows for variation in the $p$ between windows.
+Note that, just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ has effectively four unique parameters $\delta_i$ with $i$ $\in$ (0,1,2,3).
+This is due to the fact that several combinations of $Z_w$ and $H_w$ can lead to comparison between same number of unique chromosomes, and such states will have same mean as well as the same variance.
+This results in the beta-binomial likelihood, that can be fitted to the data (fig.
+S\ref{figS1:binom})).
 
-We therefore add an overdispersion parameter $\delta$ that allows for variation in the $p$ between SNPs.
-This results in the betabinomial likelihood, (we show that such a distribution fits the data well (fig.
-S\ref{FigS1:binom})).
+\begin{align}\label{eq:3}
+P(D_{w}|p_i,\delta_i,N_w) &\sim BB[D_w; N_w, p_i, \delta_i] \nonumber\\
+&= \binom{N_w}{D_w}\frac{B(D_w+p_x \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})}.
+\end{align}
 
-\begin{equation}\label{eq:3}
-P(D_{w}|p_x,\delta_x,N_w) = \binom{N_w}{D_w}\frac{B(D_w+p_x \delta_{x}, N_w-D_w+ \delta_{x}(1-p_{x}))}{ B(p_{x}\delta_{x}, (1-p_{x})\delta_{x})}
-\end{equation}
+This parameterization of the beta distribution in terms of expected value $p$ and overdispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parametrization in terms of $\alpha$ and $\beta$.
+It is worth mentioning here that $D_w$ is estimated as a sum of probabilities (see section XX), while both $D_w$ and $N_w$ may undergo contamination correction, which causes these estimates to be non-integral.
+We calculate factorials required in Beta Binomial likelihood with Gamma function, which is continuous and hence does not require integral values.
 
-\subsubsection{Estimation}
+\subsection{Estimation}
 
 Emission probabilities are estimated with Expectation-Maximization (EM) algorithm.
-In each iteration of the algorithm, the Expectation step involves forward-backward algorithm to calculate joint probability of the observed data and the IBD states $Z$, given an initial guess of $\delta$.
-In maximization step we use these joint probabilities to optimize $\delta$, and update emission probabilities.
-Iterations are run, until the complete data likelihood stops changing.
+In each iteration of the algorithm, the Expectation step involves forward-backward algorithm to calculate $P(D, Z| H, \theta, \delta)$.
+
+In maximization step we use these joint probabilities to optimize $\delta_i$ for each state i $\in$ (0,1,2,3), and update emission probabilities.
+Iterations are run, until the complete data log likelihood stops changing.
 
 \paragraph{Initialization}
-The value of $\delta$ is unknown to start with, and is randomly assigned, such that the mean of beta distribution and variance is within constraints described in maximization section.
+The value of $\delta_i$ is unknown to start with, and is set to a random value.
 
 \paragraph{Expectation step}
-
-In this step, using the standard forward-backward algorithm, we calculate the posterior probability of each IBD state in each window ($\gamma_{wi}$).
-Expectation step, among other parameters, requires emission probabilities.
-We calculate emission probabilities for each case of $Z_w$ and $H_w$ from $\delta$ and $p$ using equation \ref{eq:3}, and we get the total emission probabilities for each $Z_w$ as $\sum_{H_w} P(D_w|Z_w,N_w,H_w) P(H_w|\theta)$
+In the $t$-th iteration, we calculate the posterior probability of each IBD state in each window $\gamma^{(t)}_{wi} = P(Z_w=i | \BD, \BH, \theta, \delta^{(t)})$ using the forward-backward algorithm, where $\delta^{(t)}$ is the current guess for delta.
 
 \paragraph{Maximization step}
-
-The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_j$.
+The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_i$.
 We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wi}$ and the ROH state-probabilities $h_{w\omega}$.
 
-\begin{equation}\label{eq:4}
-\mathcal{C} = \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta, p)h_{w\omega}\gamma_{wi}
-\end{equation}
+\begin{align}\label{eq:4}
+\mathcal{C} &= \mathbb{E}[\log P(D_w|Z_w, H_w, \theta, \delta^{(t-1)}]\nonumber\\
+&= \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta^{(t-1)}, p)h_{w\omega}\gamma_{wi}
+\end{align}
 
@@
 
-\begin{align*}
-k_{w0} &= \gamma_{w0}\\
-k_{w1} &= \gamma_{w1} h_{w0}\\
-k_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\\
-k_{w3} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2}
+\begin{align*}\label{eq:5}
+g_{w0} &= \gamma_{w0}\\
+g_{w1} &= \gamma_{w1} h_{w0}\\
+g_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\\
+g_{w3} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2}
 \end{align*}
@@
 So we can rewrite equation \ref{eq:4} as:
-\begin{align}
-\mathcal{C} &= \sum_{w=1}^L\sum_{x=0}^3 \log P(D_{w}|Z_w, H_w, N_w, \delta_x, p_x)k_{wx}
+\begin{align}\label{eq:6}
+\mathcal{C} &= \sum_{w=1}^L\sum_{i=0}^3 \log P(D_{w}|Z_w, H_w, N_w, \delta_i, p_i)g_{wi}
 \end{align}
 
-Since each term in this equation depends on only one $\delta_x$, we can optimize for them independently. For convenience, we refer to these four states corresponding to different $p$ as $x$ states.
+Since each term in equation \ref{eq:6}  depends on only one $\delta_i$ and not on different IBD and ROH states, we can optimize for them independently.
 
-We observed that estimating the $\delta_w$ without constraints can result in wide Beta distributions with very high variance, leading to over fitting of data (fig.
-\ref{figS4:bndsbeta}), and resulting in incorrect posterior probabilities for $Z_w$ (fig.
-\ref{figS3:bnds}).
-For example, a parent shares exactly one chromosome IBD with its child throughout the genome, so we expect only one IBD state.
-However, running an unconstrained model could use this data to fit a pair of siblings with a very wide beta-distribution (with a very high $\delta_1$, which may result in problems in differentiating between parent-child and siblings.
-
+We realized that different cases of relatedness have different number of IBD states possible.
+For example, siblings may have all three IBD states present while a Parent-Child has only $Z_w=1$.
+This may cause some overfitting problem, when we apply a Parent-Child HMM to a dataset corresponding to siblings.
+We show a simple example of this problem in fig.
+\ref{figS4:bndsbeta}.
 We avoid this problem by constraining the parameter space of $\delta$.
-In particular, 
-Variance of beta distribution is shown below:
-\begin{align}
-    X \sim B(p,\delta)\\
+In particular, the variance of a beta distributed random variable $X$ is:
+\begin{align}\label{eq:7}
     var(X) = \frac{p(1-p)}{\delta + 1}
 \end{align}
-We find the value of $\delta$ for which the variance is less than a threshold t.
+We constrain $\delta$ such that  the variance is less than a threshold T:
 
-We solve the following equation for $\delta$:
-\begin{equation}
-    \frac{p(1-p)}{\delta + 1}  < t
+\begin{equation}\label{eq:8}
+    \frac{p(1-p)}{\delta + 1}  < T
 \end{equation}
 
-We choose t for each beta distribution to be proportional to the mean ($p$), since the variances of beta distribution corresponding to different $x$ states follow the same relation as the distribution means.
-\begin{align}
-    Var(X_{x_1}) = 3/4 Var(X_{x_0})\\
-    Var(X_{x_2}) = 1/2 Var(X_{x_0})
-\end{align}
+We choose T for each beta distribution such that the distributions do not overlap much.
 
-Even though this may not hold true when we bin sites in large windows, we still find that $Var(X_{x_0}) > Var(X_{x_1}) > Var(X_{x_2})$ (fig. \ref{figS1:binom}).
+\subsection{Relatedness cases}
+We run our model for the following relatedness cases: Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, Identical.
+We estimated the transition matrix corresponding to each relatedness case using simulations (see section XX).
+While testing KIN on simulations, we found that classification performance for $4^{th}$ and $5^{th}$ degree relatives is poor, and we have low power to distinguish them from unrelated pairs of individuals.
+However, we would like to classify 4-th and 5-th degree relatives as unrelated, and so we include these cases in the algorithm.
+Similarly, we observe that we have low power to differentiate between half-siblings, avuncular and grandparent-grandchild, and we merge these classifications as $2^{nd}$ degree in the final results.
+We report the final pairwise classification in the following categories: Unrelated, Third Degree, Second Degree, Parent-Child, Siblings, Identical individuals.
+We output the relatedness corresponding to the maximum likelihood model, and the confidence as given by the log likelihood ratio between the two highest likelihood models.
 
-Ideally, t for $x_4$ should be zero, but since our genomic windows may not coincide with the IBD fragments or ROH, the variance can be much higher.
-We find that fixing t for $x_4$ at the same value as $x_0$ works well.
-
-\subsection{Relatedness classes}
-This model is run for different cases of relatedness including (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{rd}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, Identical) using the corresponding transition matrix, and the likelihood for all the relatedness models is compared.
-We realized that we have low power to detect $4^{th}$ and $5^{th}$ degree and we can not always differentiate these from unrelated.
-In our final output, we show classifications of $4^{th}$ and $5^{th}$ degrees as unrelated.
-Similarly, we observe that we have low power to differentiate between half-siblings, avuncular and grandparent-grandchild, and we merge these classifications as $2^{nd}$ Degree in the final results.
-The relatedness for the maximum likelihood model is reported, and the confidence is given by the log likelihood ratio between the two highest likelihood models.
 Hidden IBD state of each window ($Z_w$) is estimated using standard Viterbi algorithm, and $\BZ$ corresponding to the most likely model are returned.
@@
 \subsection{ROH estimation model}
+Our HMM to detect ROH tracts works similar to the relatedness model described above, but we only consider one individual at a time, and only positions covered by at least two reads.
+For each site, we calculate the proportion of reads that carry different alleles, and sum them up in windows along the genome.
+We call the vector with the number of differences $\mathbf{\Delta}$, and the vector with the number of sites with at least two reads $\mathbf{N}$.
+Our model has two possible hidden states: homozygous (state $3$), and non homozygous (state $2$).
+As above, we collect the hidden states in a vector $\mathbf{Y} = (Y_1, \dots Y_w, \dots, Y_L)$.
+The complete data likelihood for the model in this case is then:
 
-Our HMM to detect ROH tracts works similar to the relatedness model described above.
-We sample all possible pairs of reads from each individual at all the positions where there are at least two reads, and calculate the proportion of read pairs that are different.
-These proportion of differences at each position are summed up in windows along the genome.
-This forms the input to the HMM, along with the number of overlapping sites, and the output is the posterior probability of a window being homozygous.
-This probability can also be inferred as the proportion of a window that has ROH.
-Our model has two hidden states: homozygous ($y_0$), and non homozygous ($y_1$).
-$Y_w$ denotes the state $y$ in a window $w$, and the vector of $Y_w$'s is denoted with $\mathbf{Y}$.
-The complete data likelihood for the model in this case is shown below:
-
-\begin{align}
-    P(\mathbf{\Delta},\mathbf{Y}|\Theta) &= P(\mathbf{\Delta}|\mathbf{Y},\Theta) P(\mathbf{Y}|\Theta)\nonumber\\
- &= [\prod_{w} P(\Delta_w|Y_w, \Theta) \prod_{w} P(Y_w|Y_w-1, \Theta)] P(Y_0| \Theta)
+\begin{align}\label{eq:10}
+    log P(\mathbf{\Delta},\mathbf{Y}|\Theta) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta) + log P(\mathbf{Y}|\Theta)\nonumber\\
+ &= [\sum_{w} log P(\Delta_w|Y_w, \Theta) + \sum_{w} log P(Y_w|Y_w-1, \Theta)] + P(Y_0| \Theta)
 \end{align}
 
-In this case, transitions are not known, and both transitions and emissions are estimated with standard EM algorithm.
-The emissions are calculated using betabinomial likelihood, and the mean of the distributions are constrained at expected proportion of differences in a ROH tract ($\rho_{0}$) and expected proportion of differences in a non-ROH tract $\rho_{1}$.
-Parameters $\rho_{1}$ and $\rho_{0}$ are the same as $p_{2}$ and $p_3$ in previous model.
+We give details on the transition, emission and their estimation below.
 
-\subsection{Emissions}
-Similar to our relatedness model, we estimate emissions using EM algorithm with expectation and maximization steps.
-The expectation step outputs the posterior probability $\Gamma$ of being in $y_0$ and $y_1$ in each window.
-Maximization step uses the $\Gamma$ to optimize overdispersion parameter $\eta$ with a betabinomial cost function:
+In this case, transitions are not known, and we estimate both transitions and emissions using an EM algorithm.
+We calculate the emissions using beta-binomial likelihood, and fix the mean of the distributions at expected proportion of differences in a ROH tract ($p_3$) and expected proportion of differences in a non-ROH tract ($p_2$).
+The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w=3$ or the state $Y_w=2$ in each window.
+The M-step for emissions in analogous to that in the KIN model (eq.
+\ref{eq:4}):
 
-\begin{align}
-\mathcal{C} = \sum_{w} \sum_{y} P(\Delta_w|\eta_{wy},\rho_{wy}) \Gamma_{wy} 
+\begin{align}\label{eq:11}
+\mathcal{C} = \sum_{w=1}^L \sum_{i=2}^3 P(\Delta_w|\eta_{i},p_{i}) \Gamma_{wi} ,
+\end{align}
+where $P(\Delta_w|\eta_{i},p_{i})$ is a beta-binomial probability given by eq. \ref{eq:3}.
+
+While testing this method on low coverage data, we observed that some pair of individuals may have few sites in each window, resulting in the proportion of differences to become close to 0 or 1.
+This may result in a high variance, and can potentially influence our estimation of $\eta$.
+To avoid the estimated Beta distributions from flattening to fit this high variance data, we set $Y_w=0$ for windows with high differences (fig.
+S\ref{figS5:ROHforced}).
+Finally, to estimate transitions, we initialize the transition matrix with 0.2 for the off-diagonal entries, and update it using the standard Baum-Welch update step \cite{baum_maximization_1970}.
+
+\subsection{Contamination}
+Contamination by present-day people is a common feature in ancient DNA \cite{peyregne_present-day_2020}.
+Even low levels of contamination may make samples look less similar, and thus distort relatedness analyses.
+This issue is particularly pronounced  when analyzing populations with low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations (Skov et al.,2022 (in press)).
+To address this issue, we develop a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.
+
+We conceptualize contamination in the following way.
+If data from individual $i$ has a known contamination rate of $C_i$, this means that any particular read drawn from this individual will actually come from that individual with probability $1-C_i$.
+With probability $C_i$ the read comes from a contaminant.
+Thus, when comparing two individuals $i$ and $j$, there are three possible cases: First, with probability $(1-C_i)(1-C_j)$, we compare endogenous reads, which we can use to infer relatedness.
+With probability $C_i(1-C_j) + C_j(1-C_i)$ we compare an endogenous read from an individual with a contaminant read from the other individual.
+Finally, with probability $C_iC_j$ we compare contaminant reads from both the samples.
+If we make the simplifying assumption that both $C_i$ and $C_j$ are small, the terms of the order of $C^2$ can be ignored.
+Then, we compare contaminant read to endogenous read with the probability $C_{ij} = C_i + C_j$.
+
+Our goal then, is to estimate an expectation of differences from comparison of endogenous reads $\mathbb{E}[D_w^{'}]$, and the total number of sites with overlapping endogenous reads $\mathbb{E}[N_w^{'}]$.
+To estimate $\mathbb{E}[D_w^{'}]$, we first need to calculate $P(c=0|d=1)$.
+Here, $c$ is a binary variable with value $0$ or $1$ denoting that the comparison at a site is between endogenous reads, or between endogenous and contaminant reads respectively.
+Similarly $d$ is a binary variable taking values 0 and 1 for seeing no difference, or a difference, when comparing the reads at a given site.
+We can further expand this probability:
+
+\begin{align}\label{eq:12a}
+    P(c=0|d=1) &= \frac{P(d=1|c=0)P(c=0)}{P(d=1)} \nonumber\\
 \end{align}
 
-In case of very noisy data, we see high pairwise differences in some windows that are not explained with the betabinomial distributions estimated for $y_0$ and $y_1$.
-To avoid our distributions from flattening to explain these high differences, we force these high difference windows to $y_1$ (fig.
-S\ref{figS5:ROHforced}).
+Here, $P(d=1|c=0)$, referred afterwards to as $\rho_{ij}$ for simplicity, is equivalent to the proportion of differences when we compare endogenous reads from individuals $i$ and $j$.
+This can be estimated by considering that proportion of pairwise differences ($\frac{\sum_w D_w}{\sum_w N_w}$), referred to as $o_{ij}$ afterwards, can be expressed as a sum of proportion of differences between endogenous reads and that between endogenous and contaminant reads \cite{peyregne_present-day_2020}.
+\begin{align}\label{eq:13}
+    o_{i,j} = (1-C_{ij}) \rho_{ij} + C_{ij} \phi \nonumber\\
+    or, \rho_{ij} = \frac{o_{ij} - C_{ij} \phi} {1 - C_{ij}}
+\end{align}
 
-\subsection{Transitions}
-Transitions are estimated using standard Baum-Welch algorithm.
-Expectation step is the same as that for emissions, and gives the posterior probability.
-In maximization step, we use the posterior probability to count how many times transitions are made from state i to j, normalized by the total number of times we are in state i.
-We observe that the final results are not affected by initial guess of the transition probabilities and assign a probability of 0.8 to stay in the same state for both $y_0$ and $y_1$.
+Here, $\phi$ is the probability of seeing a difference when endogenous read is compared to a contaminant read, and is equivalent to the average divergence between target and contaminating population.
+Given this information, we can rewrite eq. \ref{eq:12a}:
+\begin{align}\label{eq:12}
+    P(c=0|d=1) &= \frac{(1 - C_{ij}) \rho_{ij}}{(1 - C_{ij}) \rho_{ij} + C_{ij} \phi} \nonumber\\
+\end{align}
+We can further write $\mathbb{E}[D_w^{'}]$ as $D_w P(c=0|d=1)$:
+\begin{align}
+    \mathbb{E}[D_w^{'}] &= D_w\frac{(1-C_{ij})\rho_{ij}}{(1-C_{ij}) \rho_{ij} + C_{ij} \phi}\\
+\end{align}
 
-\subsection{Contamination}
+Similarly, we derive $\mathbb{E}[S_w^{'}]$ as the estimate for number of sites that do not show pairwise difference.
+\begin{align}
+    \mathbb{E}[S_w^{'}] &= (N_w-D_w) P(e=1|d=0) = (N_w - D_w)\frac{(1-\rho_{ij})(1-C_{ij})}{(1-\rho_{ij})(1-C_{ij}) + C_{ij}(1-\phi)}\\
+\end{align}
 
-Contamination by present-day people is a common feature in ancient DNA \cite{peyregne_present-day_2020} .
-Even low levels of contamination may make samples look less similar, and thus distort relatedness analyses.
-This issue is particularly pronounced  when analyzing populations with  low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations.
-To address this issue, we  adjust $D_w$ and $N_w$, given the contamination estimates of both samples and the average divergence between the target population and a putative contaminating population.
+We added $D_w^{'}$ to $S_w^{'}$ to get the corrected number of overlapping sites $N_w^{'}$
 
-For each pair of individuals $i$ and $j$ we approximate the contamination $C_{i,j}$ as the sum of contamination for each individual $C_i + C_j$.
-Then, $C_{i,j}$ represents the probability of comparing a contaminant read from an individual to an endogenous read from the other individual.
-Throughout, we assume that contamination levels are low, so that we can ignore comparison between contaminant reads, i.e.
-$C_iC_j \approx 0$.
-We denote the average divergence between target and contaminating populations as $p_c$, and the observed proportion of differences between individuals $i$ and $j$ as $p_{oij}$.
+------------------------------------------extra stuff-----------Do not read--------------------------------
+
+If we look at one site in the window, we could find either no difference in reads from individuals $i$ and $j$ ($d_s$=0), or a non-zero difference ($0<d_s<=1$).
+Here $d_s$ is the probability of seeing a difference at site $s$, and $\sum_{s \in w} d_s =D_w$.
+We ask, what is the
+
+If we make the simplifying assumption that both $C_i$ and $C_j$ are small, terms of the order of $C^2$ can be ignored and there are just two cases: with probability $C_{i,j} =C_i + C_j$, one of the reads stems from a contaminant, and we effectively compare a contaminant fragment with an endogenous one.
+The probability that they differ is the average divergence between target and contaminating population, which we denote with $\phi$.
+The other case (w.
+p.
+$(1-C_{i,j}$) is where we compare two endogenous reads, which will differ with probability
+
+\begin{equation}
+    p_{e} = \frac{p_{obs}-\phi C_{i,j}}{1-C_{i,j}}
+\end{equation}
+
+Here, $$p_{obs} = \frac{\sum_w D_w}{\sum_w N_w}$$
+
+We estimate $\hat{D_c} = C_{i,j} N_{tot} \phi$ and $\hat{D_e} = (1-C_{i,j}) N_{tot} p_{obs}\phi$.
+If there are $N_{obs}$ overlapping sites in a window, $N_{c} = C_{i,j}N_{obs}$ of them will include a contaminant allele, and $N_e = (1-C_{i,j})N_{obs}$ will be between endogenous alleles.
+
+For KIN, we can use $N_e$ as an input (for each window), but we also need the proportion of differences, $D_e$.
+For this, we use the assumption that the proportion of sites where the contaminant differs from our samples is
+
+$$\phi = \frac{D_c}{N_c} = \frac{D_c}{C_{i,j} N_{obs}}, $$
+and thus $D_c = C_{i,j} \phi N_{obs}$.
+
+We can think of $p_c$ and $\rho$ as empirical likelihoods that a particular site shows a difference between two reads, given that the comparisons contain only endogenous or one contaminant read, respectively.
+\begin{align}
+    P(E|D) &= \frac{P(D|E)P(E)}{P(D)} \nonumber\\
+    \mathbb{E}[D_e] &= D_{obs} P(E|D) = D_{obs}\frac{p_e (1-c)}{p_e(1-c) + c\rho}\\
+    \mathbb{E}[S_e] &= D_{obs} P(E|\text{not} D) = S_{obs}\frac{(1-p_e)(1-c)}{(1-p_e)(1-c) + c(1-\rho)}
+\end{align}
+
+And for calculating $N_{e}$, we added $D_{e}$ to $S_{e}$
+
+The goal is that given $C_{i,j}, p_e, \BN and \BD$ to calculate adjusted $\hat{\BN}$ and $\hat{\BD}$ that reflect the endogenous proportion only.
+
+, and the observed proportion of differences between individuals $i$ and $j$ as $p_{oij}$.
 We calculate average endogenous pairwise difference $p_{eij}$ for individuals i and j as follows \cite{noauthor_ancient_nodate}:
@@
 
-This model outputs the contamination corrected number of differences and overlapping sites in each window.
+This model outputs the contamination-corrected number of differences and overlapping sites in each window.
 
@@
 We use simulations both for estimating the transition matrices for all our models, and for testing and validating our algorithm.
-All simulations are performed using \textt{msprime} \cite{kelleher_efficient_2016}, followed by simulations of related individuals using a predetermined pedigree (Figure S1).
+All simulations are performed using \textt{msprime} \cite{kelleher_efficient_2016}, followed by formation of related individuals using a predetermined pedigree (Figure S\ref{figS10:pedigree}).
 
-\paragraph{ Simulations for estimation of transition matrix}
+\paragraph{Simulations for estimation of transition matrix}
 
-For our simulations of background diversity, we simulated a haploid population (Pop1) with constant effective size of 10,000 and sampled 16 haploids from 2500 generations ago to randomly combine into 8 unrelated diploid ancient individuals (fig.
+For our simulations of background diversity, we formed a haploid population (Pop1) with constant effective size of 10,000 and, sampled 16 haploids from 2500 generations ago to randomly combine into 8 unrelated diploid ancient individuals (fig.
 \ref{figS10:pedigree}).
@@
 For artificially mating two individuals, we simulated a recombined chromosome from each individual to be passed on to the progeny.
-For simulating a recombined chromosome from a diploid individual, we drew the number of recombination points from a Poisson distribution with parameter $rL$ as the recombination rate, and used a uniform distribution to sample the positions of recombination points.
+To simulate a recombined chromosome from a diploid individual, we drew the number of recombination points from a Poisson distribution with parameter $rL$ as the recombination rate, and used a uniform distribution to sample the positions of recombination points.
 We estimated transition matrices corresponding to different cases of relatedness by counting transition between IBD states for a pair of  individuals at a particular relatedness level.
@@
 \paragraph{Model Comparison}
-Model with maximum likelihood is reported, and log likelihood ratio is calculated with the two highest likelihood models.
+A model with maximum likelihood is reported, and log likelihood ratio is calculated with the two highest likelihood models.
 
-Apart from the related and unrelated individuals in pop1, we simulated more haploids in three other populations to create scenarios with ascertainment and contamination (fig.
+Apart from the related and unrelated individuals in Pop1, we simulated more haploids in three other populations to create scenarios with ascertainment and contamination (fig.
 \ref{figS10:pedigree}).
-We simulated two haploids to form an individual each from two other populations (pop2, pop3) with split time of 3500 and 4500 generations with pop1, and sampling time of 4000 generations and 2000 generations ago respectively.
-We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from pop1.
-We test out the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a markov chain using transition matrix shown below:
+We simulated two haploids to form an individual each from two other populations (Pop2, Pop3) with split time of 3500 and 4500 generations with Pop1, and sampling time of 4000 generations and 2000 generations ago respectively.
+We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from Pop1.
+We test out the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a Markov chain using transition matrix shown below:
 
@@
 {rr}
-1-10/l & 10/l \\
-2/l & 1-2/l  \\
+1-10/L & 10/L \\
+2/L & 1-2/L  \\
 \end{array}\right]
@@
 
-Here, l is length of the chromosome.
-From the steps described above, we got genotypes of 17 individuals in pop1 in presence/absence of ROH and ascertainment.
-We further simulated 10 haploids (5 diploid individuals) from pop4 with split time of 20,000 generations with pop1, sampled from the present time.
-We calculated the allele frequency for pop4 at each site, and used it to introduce contamination (with fixed contamination percentage for each individual ($<5\%$) in samples from pop1.
+From the steps described above, we got genotypes of 17 individuals in Pop1 in presence/absence of ROH and ascertainment.
+We further simulated 10 haploids (5 diploid individuals) from Pop4 with split time of 20,000 generations with pop1, sampled from the present time.
+We calculated the allele frequency for Pop4 at each site, and used it to introduce contamination (with fixed contamination percentage for each individual ($<5\%$) in samples from pop1.
 We generated reads (derived/ancestral) for different genomic coverages ranging from 4x to 0.03x at each position for each individual.
@@
 
-We predict the following categories of relatedness with KIn: Identical individuals (twins), First degree (parent-child), First degree (siblings), Second degree, Third degree, Unrelated (including higher than third degree relatives).
-We show the working of Relatedness HMM with Fig.\ref{fig1:ibd} showing performance of each model, when applied to a pair of siblings.
+We predict the following categories of relatedness with KIN: Identical individuals (twins), First degree (parent-child), First degree (siblings), Second degree, Third degree, Unrelated (including higher than third degree relatives).
+We show the output of different Relatedness HMMs in Fig.\ref{fig1:ibd}, when they are applied to a pair of siblings.
 
@@
 This is reflected in the log likelihoods values (LL) of the models as well, where siblings model has the highest value.
-We take a loglikelihood ratio ($\Lambda$) of the two most likely models to get a confidence in the relatedness estimate (in this case $\Lambda$ = 22.48).
+We take a log likelihood ratio ($\Lambda$) of the two most likely models to get a confidence in the relatedness estimate (in this case $\Lambda$ = 22.48).
 All the models in Fig.\ref{fig1:ibd} utilize information about ROH given by the ROH-HMM.
@@
 
-We plotted the true positive and false positive rates with a cutoff on the loglikelihood ratio in Fig.\ref{fig3:cutoff}.
-Here, and in Fig.\ref{fig4:Comparison_READ_KIn}, we have two relatedness labels: 'Unrelated' and 'Unrelated w/o 3rd Degree'.
-Unrelated label here refers to KIn performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
-This is useful for comparison to READ, since READ classifies relatedness upto Second Degree.
+We plotted the true positive and false positive rates with a cutoff on the log likelihood ratio in Fig.\ref{fig3:cutoff}.
+Here, and in Fig.\ref{fig4:Comparison_READ_KIN}, we have two relatedness labels: 'Unrelated' and 'Unrelated w/o 3rd Degree'.
+Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
+This is useful for comparison to READ, since READ classifies relatedness up to Second Degree.
 'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when Third Degree is considered as separate from the unrelated individuals.
@@
 Degree 3 shows higher false positive rates in general, and we see that only for 4x coverage and a cutoff of 2, do we see a false positive rate of around $5\%$.
-In conclusion, it is recommended to use a cutoff of 1 for most cases, with the exception of differentiation between parent-child and siblings, where a cutoff of 2 may be more appropriate, and degree 3 where cutoff of 2 at 4x coverage seems reasonable.
+In conclusion, it is recommended to use a cutoff of 1 for most cases, with the exception of differentiation between parent-child and siblings, where a cutoff of 2 may be more appropriate, and degree 3 where cutoff of 2 at around 4x coverage seems reasonable.
 
@@
 We use a cutoff of 1 to compare our method performance to READ, while using a cutoff of 1 standard deviation for READ.
-Fig.\ref{fig4:Comparison_READ_KIn} shows this comparison using false positive and true positive rates for both the methods at different coverages under different simulations.
-We show that both methods have similar performance in the control case (except for 0.1x coverage, where KIn has higher power to detect First Degree and Second Degree, and slightly higher false positive rate for degree 2).
-However, in case of ascertainment, ROH, or contamination KIn shows higher power to detect relatedness cases while maintaining a low false positive rate.
+Fig.\ref{fig4:Comparison_READ_KIN} shows this comparison using false positive and true positive rates for both the methods at different coverages under different simulations.
+We show that both methods have similar performance in the control case (except for 0.1x coverage, where KIN has higher power to detect First Degree and Second Degree, and slightly higher false positive rate for degree 2).
+However, in case of ascertainment, ROH, or contamination KIN shows higher power to detect relatedness cases while maintaining a low false positive rate.
 
@@
     \centering
-    \caption{Comparison of KIn with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH}
-    \label{fig4:Comparison_READ_KIn}
+    \caption{Comparison of KIN with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH}
+    \label{fig4:Comparison_READ_KIN}
 \end{figure}
 
-To test KIn on real ancient dataset, we applied it to Neandertal specimens from Chagyrskaya cave in Siberia, Russia.
-This dataset comprised of 14 specimens that came from neighbouring layers, and hence could possibly belong to contemporary neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya \cite{kolobova_archaeological_2020-1}.
-These specimens showed low coverage of nuclear genome, with 8 samples out of 14 showing $<1x$ coverage.
+To test KIN on real ancient dataset, we applied it to Neandertal specimens from Chagyrskaya cave in Siberia, Russia.
+This dataset comprised of 14 specimens that came from neighbouring layers, and hence could possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya \cite{kolobova_archaeological_2020-1}.
+These specimens showed low coverage of nuclear genome (ranging from 0.01x to 12.34x), with 8 samples out of 14 showing $<1x$ coverage.
 Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena.
-All these samples were captured with an array targeting for Neanderthal variation, and we further ascertained this data to variable sites in high coverage Neanderthal genomes: Altai Neanderthal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}.
-This dataset had all the issues that KIn deals with, and hence we decided to apply it here to see if we could find family relationships that would not be possible with other available methods.
-Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_kin}.
+All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes and common variations in Africans (Laurits et al.,2022), and we further ascertained this data to variable sites in high coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}.
+This dataset had all the issues that KIN deals with, and hence we decided to apply it here to see if we could find family relationships that would not be possible with other available methods.
+Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}.
 We found identical individuals (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), first degree relative (Chagyrskaya07-Chagyrskaya17), and a second degree relative (Chagyrskaya01-Chagyrskaya60).
@@
 S\ref{figS2:Chagyrskaya_READ}).
-Chagyrskaya06-Chagyrskaya14 were estimated to be First Degree by both the methods, but have been classified as Identical pair in the original study, since Chagyrskaya06 is a tooth sample that fits in mandible which is Chgyrskaya14.
+Chagyrskaya06-Chagyrskaya14 were estimated to be First Degree by both the methods (table S1), but have been classified as Identical pair in the original study, since Chagyrskaya06 is a tooth sample that fits in mandible which is Chagyrskaya14.
 We believe that the contradiction is due to non human contamination present in these libraries ($1\%$ and $2.9\%$ respectively).
-Interestingly, for the pair of Chagyrskaya07-Chagyrskaya17, READ shows that they are first degree relatives, and KIn further identifies them as parent-child.
+Interestingly, for the pair of Chagyrskaya07-Chagyrskaya17, READ shows that they are first degree relatives, and KIN further identifies them as parent-child.
 Our result matches with the conclusions of the authors based on additional information from mtDNA.
 Further, we find a possible third degree relative (Chagyrskaya07-Chagyrskaya60).
-We compared our IBD estimates to lcMLkin for all pairs for whom READ results (with s.d.
->1) and KIn results ($\Lambda$>1) match (fig.
+We compared our IBD estimates to lcMLKIN for all pairs for whom READ results (with s.d.
+>1) and KIN results ($\Lambda$>1) match (fig.
 S\ref{figS6:Chagyrskaya_ibd}).
 Chagyrskaya06-Chagyrskaya14 is removed from the analysis.
-We find that lcMLkin does create four different clusters for different relatedness.
+We find that lcMLKIN does create four different clusters for different relatedness.
 However, the clusters are quite deviated from the expected values, which makes it difficult to identify the relatedness.
@@
     \centering
-    \caption{Application of KIn on Neandertal specimens from Chagyrskaya.
-Color of a square represents the relatedness, while the number denotes loglikelihood ratio between the two maximum likelihood models.}
-    \label{fig5:Chagyrskaya_kin}
+    \caption{Application of KIN on Neandertal specimens from Chagyrskaya.
+The color of a square represents the relatedness, while the number within denotes log likelihood ratio between the two maximum likelihood models.}
+    \label{fig5:Chagyrskaya_KIN}
 \end{figure}
 
-We further applied KIn to a genome-wide dataset of 118 ancient individuals sequenced and analyzed by Mittnik et al.,2019.
-We compared our relatedness estimates to those obtained from the authors using READ and lcMLkin, and found that in most cases the three methods agree.
-Out of 6903 pairs, 346 pairs show different relatedness for KIn and READ (table S2).
-However, there are only 31 pairs where KIn has $\Lambda >1$.
+We further applied KIN to a genome-wide dataset of 118 ancient individuals sequenced and analyzed by Mittnik et al.,2019.
+We compared our relatedness estimates to those obtained from the authors using READ and lcMLKIN, and found that in most cases the three methods agree.
+Out of 6903 pairs, 346 pairs show different relatedness for KIN and READ (table S2).
+However, there are only 31 pairs where KIN has $\Lambda >1$.
 We looked in detail at some of these pairs, and we found that it is possible to deduce the correct relatedness for several of these pairs by plotting the proportion of differences along the genome.
-In all the pairs that we analyzed in this way, KIn seems to predict the true relatedness.
+In all the pairs that we analyzed in this way, KIN seems to predict the true relatedness.
 We show few examples of such cases in fig.
 S\ref{figS8:eg1}.
-We found 20 pairs in this subset of 346 pairs where lcMLkin has sufficient data (Overlapping SNPs $>$ 10000).
-Among these 20 pairs, we found that lcMLkin matches KIn estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLkin predicts 3rd-5th degree, KIn predicts second degree) and pair AITI72-AITI77A (READ and lcMLKin predict unrelated, while KIn estimates 3rd degree with log likelihood=1.7).
-We found 43 pairs for which READ estimates first degree, and compared the estimates of lcMLkin and KIn for these cases (suppl table3).
-There were a total of 20 pairs where lcMLkin had sufficient data, and 26 pairs where $loglikelihood ratio>2$ for KIn.
-Among the 19 overlapping pairs, we found four pairs where KIn and lcMLkin differ, and a visual inspection shows that in these cases, KIn seems to have the correct estimate (fig.
+We found 20 pairs in this subset of 346 pairs where lcMLKIN has sufficient data (Overlapping SNPs $>$ 10000).
+Among these 20 pairs, we found that lcMLKIN matches KIN estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLKIN predicts 3rd-5th degree, KIN predicts second degree) and pair AITI72-AITI77A (READ and lcMLKIN predict unrelated, while KIN estimates 3rd degree with log likelihood=1.7).
+We found 43 pairs for which READ estimates first degree, and compared the estimates of lcMLKIN and KIN for these cases (suppl table3).
+There were a total of 20 pairs where lcMLKIN had sufficient data, and 26 pairs where $loglikelihood ratio>2$ for KIN.
+Among the 19 overlapping pairs, we found four pairs where KIN and lcMLKIN differ, and a visual inspection shows that in these cases, KIN seems to have the correct estimate (fig.
 S\ref{figS9:eg2}).
@@
 Several methods are available to estimate relatedness from modern as well as ancient DNA data.
-We developed a new method to estimate genetic kinship from low coverage data.
+We developed a new method to estimate genetic KINship from low coverage data.
 We show that it is possible to accurately infer relatedness, along with IBD tracts, from such data in presence of long ROH, contamination, as well as ascertainment bias.
@@
 This approach of using Hidden Markov Model makes use of the fact that DNA is inherited in blocks, and hence the IBD state in a given window is dependent on the IBD state in the previous window.
-We use this approach to identify the following relatives: Identical individuals, Parent-Child, Siblings, Second degree, Third degree, Unrelated individuals.
-We have developed a similar Hidden Markov Model to infer long ROH, and we use the posterior ROH probabilities to make better relatedness inference.
-We further use a simple model to correct the pairwise differences in genomic windows for contamination.
+We use this approach to identify the following pairwise relations: Identical individuals, Parent-Child, Siblings, Second degree, Third degree, Unrelated individuals.
+We have developed a similar Hidden Markov Model to infer long ROH, and we use the posterior ROH probabilities to make better relatedness inferences.
+We further use a simple model to correct the pairwise differences in genomic windows in the presence of contamination.
 
-We demonstrate using simulations, that our method performs well at low coverage.
-We compared KIn's model performance to a widely used software READ to show that our approach gives similar predictions as READ on ideal data with 4x coverage, and without long ROH, ascertainment and contamination, while additionally distinguishing between parent-child and siblings, and identifying third degree relatives.
+We used simulated data to validate KIN's performance.
+We compared KIN's model performance to a widely used software READ to show that our approach gives similar predictions as READ on ideal data with 4x coverage, and without long ROH, ascertainment and contamination, while additionally distinguishing between parent-child and siblings, and identifying third degree relatives.
 We demonstrate that at low coverage (0.1x) we have higher power to detect first degree relatives compared to READ with similar false positives, while a higher power to identify second degree relatives at the cost of slightly higher false positive rate.
-Contrary to the intuition, we see that adding long ROH ($~17\%$) to simulations improves the performance of both KIn and READ.
+Counter-intuitively, we see that adding long ROH ($~17\%$) to simulations improves the performance of both KIN and READ.
 READ's performance improves because ROH does not cause any change in pairwise differences between unrelated individuals, while decreasing pairwise differences between related individuals.
-More is the relatedness, the more is the decrease in pairwise differences in presence of ROH.
-And so long ROH causes higher power for READ which otherwise is a conservative method.
-KIn performs well for the same reasons, but can additionally differentiate between different combinations of IBD and ROH in genomic windows.
-Contamination, on the other hand can make the performance of READ much worse depending on the amount of contamination, and the number of individuals contaminated.
-We show that even at contamination levels  $<=3\%$, and contamination in only 8 out of 17 individuals, READ performance falls dramatically.
-Contamination in more individuals will cause a shift in median difference in the population and may cause non contaminated individuals to look more related, thus increasing the false positive rates.
-On the other hand, increase in amount of contamination in few individuals can make them look more unrelated.
-KIn, with the implementation of a model to correct for contamination performs well in this case.
+The higher the relatedness, the higher is the decrease in pairwise differences in presence of ROH.
+And so, ROH causes higher power for READ which otherwise is a conservative method.
+KIN performs well for the same reasons, but can additionally differentiate between different combinations of IBD and ROH in genomic windows.
+Contamination, on the other hand, can affect the performance of READ adversely depending on the amount of exogenous DNA, and the number of individuals contaminated.
+We show that even at contamination levels  $<=3\%$, and contamination in only 8 out of 17 individuals, READ performance falls.
+Contamination in more individuals causes a shift in median difference in the population and may cause non contaminated individuals to look more related, thus increasing the false positive rates.
+On the other hand, an increase in the amount of contamination in a few individuals can make them look more unrelated.
+KIN, with the implementation of a model to correct for contamination, performs well in this case.
 Ascertainment causes a decrease in power for both the methods, since it reduces the amount of data.
-However, the decrease in power is more for READ compared to KIn.
+However, the decrease in power is more for READ compared to KIN.
 
 DNA data from Bronze age, Lech valley has low contamination, and no ROH.
-We demonstrate that KIn matches the results from READ and lcMLkin for pairs with good amount of overlaps, while adds more information for low coverage samples.
-As expected from the simulations, KIn is helpful in differentiating between parent-child from siblings, and identifying second and third degree individuals in cases where there is an overlap of only few thousand sites or less, and the other methods are not informative.
-We show that when applied to Neandertal specimens from Chagyrskaya cave, READ and KIn agree in relatedness estimates for all pairs where both methods have sufficient data.
-KIn additionally identifys a pair of first degree relatives as parent-child, which is in agreement with mtDNA analysis from the authors.
-In addition, KIn identifies a third degree relative.
-We compare IBD estimates from KIn to lcMLkin and show that KIn's IBD estimates are much closer to the expected estimates, given the relatedness between individuals that has been estimated with both READ and KIn.
-Output of KIn is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio.
-Due to the tabular output, KIn is easy to automatise and apply to large datasets.
-One limitation with our approach is that it assumes a single population.
-In case of population structure, where the subpopulations have very different diversity, KIn may show inaccurate inferences.
+We demonstrate that KIN matches the results from READ and lcMLKIN for pairs with good amount of overlaps, while adds more information for low coverage samples.
+As expected from the simulations, KIN is helpful in differentiating between parent-child from siblings, and identifying second and third degree individuals in cases where there is an overlap of only few thousand sites or less, and the other methods are not informative.
+We show that when applied to Neandertal specimens from Chagyrskaya cave, READ and KIN agree in relatedness estimates for all pairs where both methods have sufficient data.
+KIN additionally identifies a pair of first degree relatives as parent-child, which is in agreement with mtDNA analysis from the authors.
+In addition, KIN identifies a third degree relative.
+We compare IBD estimates from KIN to lcMLKIN and show that KIN's IBD estimates are much closer to the expected estimates, given the relatedness between individuals that has been estimated with both READ and KIN.
+The output of KIN is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio.
+Due to the tabular output, KIN is easy to automatise and apply to large datasets.
+One limitation of our approach is that it assumes a single population.
+In case of population structure, where the sub-populations have very different diversity, KIN may show inaccurate inferences.
 Our method makes another assumption, that the dataset comprises of at least some unrelated individuals.
@@
 \section{Supplementary Information}
-
-Pipeline of KIn
 
@@
 
-Beta binomial distribution is a better fit compared to Binomial distribution for proportion of differences in windows.
-
 \begin{figure}[h!]
@@
     \centering
-    \caption{Comparison of fit with Betabinomial and Binomial distributions.
+    \caption{Comparison of fit with Beta-binomial and Binomial distributions.
 Data is represented by the histogram of proportion of differences from all windows of a (A) unrelated, (B) Parent-Child, (C) Identical pair of individuals.
-Binomial parameter p is calculated from the data, while Betabinomial parameters are estimated with the relatedness HMM.}
+Binomial parameter p is calculated from the data, while Beta-binomial parameters are estimated with the relatedness HMM.}
     \label{figS1:binom}
@@
     \caption{Overview of simulated dataset.
-We simulate pop1, and artificially mate unrelated individuals to make a family with upto 5th degree relatives, along with other unrelated and identical individuals (not shown here).
+We simulated Pop1, and artificially mated unrelated individuals to make a family with upto 5th degree relatives, along with other unrelated and identical individuals (not shown here).
 Pop2 and Pop3 are simulated so as to introduce ascertainment, while Pop4 is a present day population used to simulate contamination.
-Split times and age of the populations are shown in generations.}
+Split times and sampling times of the populations are shown in generations.}
     \label{figS10:pedigree}
@@
 
-Results from KIn agree with results from READ, and provide some more information about siblings/parent-child, and $3^{rd}$ Degree relatedness. 
 \begin{figure}[h!]
@@
 
-Constraining variance of beta distributions in case of an identical pair, when identical relatedness hmm is applied, gives more accurate emission probabilities.
 \begin{figure}[h!]
@@
 
-In figS4, the emissions improve in accuracy with constraints on variance of Beta distributions.
-figS5 shows why we see this improvement.
-With variance constraints, beta distributions corresponding to Z=1 and Z=2 do not overfit the data, and higher emission probabilities seen in (A) at $D_w/N_w$=0.05 disappear.
-All probability distributions are scaled by the same factor so they are visible with the data histogram.
 \begin{figure}[h!]
@@
 
-Restricting the emission probabilities at 1 for Y=1, in windows that have higher than $p_0$ difference improves the estimated Beta parameters. 
 \begin{figure}[h!]
@@
 
-*Remember to use updated lcMLkin in the plot.
+*Remember to use updated lcMLKIN in the plot.
 \begin{figure}[h!]
@@
     \includegraphics[width=18cm]{supplementary_info/plots/lcPlot.png}
-    \caption{Comparison of IBD states estimated for Chagyrskaya specimens using (A) lcMLkin and (B) KIn.
-    The relatedness shown with different colors is estimated with and matched with READ and KIn.}
+    \caption{Comparison of IBD states estimated for Chagyrskaya specimens using (A) lcMLKIN and (B) KIN.
+    The relatedness shown with different colors is estimated with and matched with READ and KIN.}
     \label{figS6:Chagyrskaya_ibd}
@@
 
-This figure shows that it is possible to differentiate between Avuncular and Grandparent-Grandchild pairs.
-However, differentiating between Half-siblings may not be possible with our approach. 
-*Add 1M windows result
 \begin{figure}[h!]
@@
 
--READ and KIn differ
-
--supplementary table2
 \begin{figure}[!ht]
@@
     \includegraphics[width=18cm]{supplementary_info/plots/egplot1.png}
-    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIn differs from READ.
-(A) AITI62B-OTTM156 (READ estimates Identical individuals, lcMLkin does not have sufficient data, and KIn predicts parent-child.
-(B) UNTA5867-UNTA5868Sk1 (READ estimates Second Degree, lcMLkin does not have sufficient data, and KIn predicts parent-child).
-(C) AITI40-AITI72 (READ estimates Unrelated individuals, lcMLkin shows 3rd-5th degree, and KIn predicts 3rd degree.)
-    (D) AITI2-AITI55 (READ estimates Unrelated individuals, lcMLkin and KIn predicts predict second degree.)}
+    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIN differs from READ.
+(A) AITI62B-OTTM156 (READ estimates Identical individuals, lcMLKIN does not have sufficient data, and KIN predicts parent-child.
+(B) UNTA5867-UNTA5868Sk1 (READ estimates Second Degree, lcMLKIN does not have sufficient data, and KIN predicts parent-child).
+(C) AITI40-AITI72 (READ estimates Unrelated individuals, lcMLKIN shows 3rd-5th degree, and KIN predicts 3rd degree.)
+    (D) AITI2-AITI55 (READ estimates Unrelated individuals, lcMLKIN and KIN predicts predict second degree.)}
     \label{figS8:eg1}
@@
     \includegraphics[width=18cm]{supplementary_info/plots/egplot2.png}
-    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIn differs from lcMLkin.
-(A) AITI43-AITI55 (READ estimates First degree, lcMLkin predicts siblings, while KIn predicts parent-child.
-(B) AITI70-AITI72 (READ estimates First degree, lcMLkin predicts parent-child, while KIn predicts siblings.}
+    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIN differs from lcMLKIN.
+(A) AITI43-AITI55 (READ estimates First degree, lcMLKIN predicts siblings, while KIN predicts parent-child.
+(B) AITI70-AITI72 (READ estimates First degree, lcMLKIN predicts parent-child, while KIN predicts siblings.}
     \label{figS9:eg2}
@@
 \item $t$: Threshold on variance of Beta distribution
-\item $X$: Random variable from a Beta distribution
+\item $X$: Random variable from a beta distribution
 \item $y$: Hidden homozygosity states for ROH HMM
@@
 \item $Gamma$: Posterior probabilities of ROH
-\item $\eta$: Overdispersion parameter used in ROH HMM
+\item $\eta$: Over-dispersion parameter used in ROH HMM
 \item $\mathcal{C}$: Cost function
@@
 \section{extra}
+
+log lik:
+\begin{align}\label{eq:1}
+\log P(\BD|\theta) &= \log \sum_{\BZ} [\sum_{\BH} P(\BD|\BZ,\BH, \theta) P(\BH |\theta)] P(\BZ |\theta) \nonumber\\
+&= \log \sum_{\BZ} [\sum_{\BH} \prod_w P(\BD|\BZ,\BH, \theta) P(\BH |\theta)] P(\BZ |\theta)
+\end{align}
+
 Many methods exist to estimate relatedness.
@@
 
-lcMLkin is another widely used method that uses genotype likelihoods to estimate the number of positions at which a pair of individuals have 0/1/2 chromosomes in IBD.
+lcMLKIN is another widely used method that uses genotype likelihoods to estimate the number of positions at which a pair of individuals have 0/1/2 chromosomes in IBD.
 
-REAP is a software to estimate relatedness coefficient and IBD prbabilities in admixed populations, accounting for structure.
+REAP is a software to estimate relatedness coefficient and IBD probabilities in admixed populations, accounting for structure.
 It requires the number of ancestral populations, and a representation of sub population allele frequencies.
@@
 There are various ways to measure the coefficient of relatedness r.
-Most methods use the probabilties of IBD to calculate r.
+Most methods use the probabilities of IBD to calculate r.
 There can be, in total 9 modes of identity described by Jacquard (1974), including the cases for 3 IBD probabilities and 3 cases of IBD.
@@
 We obtain posterior probability of one or both individuals being ROH in each window from another HMM described in section 3.2.
-We calculate the cost function with Betabinomial likelihood in log space for the entire genome as:
+We calculate the cost function with beta-binomial likelihood in log space for the entire genome as:
 
@@
 
-Emission probability in a window depends on both IBD state and ROH state. We calculate a cost function as the betabinomial loglikelihood weighted by the posterior probability represented by $\gamma$.
+Emission probability in a window depends on both IBD state and ROH state.
+We calculate a cost function as the beta-binomial log-likelihood weighted by the posterior probability represented by $\gamma$.
 likelihood:
@@
 
+Supplementary info:
+Pipeline of KIN
+Beta binomial distribution is a better fit compared to Binomial distribution for proportion of differences in windows.
+Results from KIN agree with results from READ, and provide some more information about siblings/parent-child, and $3^{rd}$ Degree relatedness. 
+Constraining variance of beta distributions in case of an identical pair, when identical relatedness hmm is applied, gives more accurate emission probabilities.
+In figS4, the emissions improve in accuracy with constraints on variance of beta distributions.
+figS5 shows why we see this improvement.
+With variance constraints, beta distributions corresponding to Z=1 and Z=2 do not overfit the data, and higher emission probabilities seen in (A) at $D_w/N_w$=0.05 disappear.
+All probability distributions are scaled by the same factor so they are visible with the data histogram.
+Restricting the emission probabilities at 1 for Y=1, in windows that have higher than $p_0$ difference improves the estimated Beta parameters. 
+This figure shows that it is possible to differentiate between Avuncular and Grandparent-Grandchild pairs.
+However, differentiating between Half-siblings may not be possible with our approach. 
+*Add 1M windows result
+-READ and KIN differ
+
+-supplementary table2
+
 \end{document}

### 2022-03-15 plots/plot_script.py
location of read files changed

@@ -256,7 +256,7 @@ def cutoff_plotf(a,cnt,inb,hmmfold,outfold):
                 elif model1.split('_')[0]=="read":
                     un_flag=0
                     model=model1
-                    table=hmmfold+"rocNo3deg/contam%s/inbred%s/model_performance_%s/coverage%s/asc%s/filtered0_cut%s.csv.gz" %(cnt,inb,model,cov,a,ct)
+                    table="/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz_highDiversity/hmm_numba_corrA_relSim_fast_theoretical_sib_grA/rocNo3deg/contam%s/inbred%s/model_performance_%s/coverage%s/asc%s/filtered0_cut%s.csv.gz" %(cnt,inb,model,cov,a,ct)
 
                 t= pd.read_csv(table, sep=",", header=0,index_col=0)
                 t=all_manipulation(t)

### 2022-03-15 paper_overleaf/main.tex

@@
 
-These analyses use cranial and dental metric and morphological and postcranial non-metric data to infer inter-individual relatedness based on the assumption that the phenotype reflects underlying genetic variation and patterns of relatedness.
-Multifactorial inheritance complicates straightforward interpretations, however, and given the moderate to low (and population specific) heritabilities of most morphometric traits (see Scott & Turner, 1997 for an overview; also Townsend et al., 2009), the effects of development on final phenotypic form are difficult to specify.
-
 \subsection{Simple way to get relatedness, methods that do it with a lot of data}
 
-Commonly, pairs of related individuals are  identified by looKINg for parts of the genome that are identical by descent (IBD), ie.
+Commonly, pairs of related individuals are identified by looking for parts of the genome that are identical by descent (IBD), ie.
 inherited from a recent common ancestor.
@@
 \cite{roberson_visualization_2009}.
-Most methods, however, use IBD as an intermediary in inferring relatedness \cite{boehnke_accurate_1997,lynch_estimation_1999,mcpeek_statistical_2000}.
+Most methods, however, use IBD as an intermediary in inferring relatedness \cite{boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}.
 For a non-inbred population, there are three IBD states possible: either no chromosomes are shared, only one chromosome is shared, or both chromosomes are shared.
@@
 There are numerous methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to estimate the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}.
-PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from average of observed IBS, using the allele frequencies at each SNP in samples \cite{purcell_plink_2007}.
+PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each SNP in samples \cite{purcell_plink_2007}.
 KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}.
@@
 \subsection{Problems with ancient data, and methods that address these issues}
-One major issue with applying the above mentioned methods to ancient DNA is that, frequently, the amount of endogenous DNA is very low, maKINg it very difficult to get genotype calls.
+One major issue with applying the above mentioned methods to ancient DNA is that, frequently, the amount of endogenous DNA is very low, making it very difficult to get genotype calls.
 Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}.In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods.
@@
 A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to calculate average pairwise genetic distances, which in turn are used to infer relatedness.
-The limitations are that the software can identify only up to second degree relatives, without differentiating between siblings and parent-child, which may be crucial when maKINg pedigrees.
+The limitations are that the software can identify only up to second degree relatives, without differentiating between siblings and parent-child, which may be crucial when making pedigrees.
 
 \subsection{How our method works}
-Here, we present KIN (KINship Inference), a Hidden Markov Model (HMM) to estimate relatedness and IBD from low-coverage ancient DNA data.
+Here, we present KIN (Kinship INference), a Hidden Markov Model (HMM) to estimate relatedness and IBD from low-coverage ancient DNA data.
 KIN can detect up to \nth{3} degree relatives, and differentiates between siblings and parent-child.

### 2022-03-15 paper_overleaf/main.tex
Updates from Overleaf

@@
 \usepackage[section]{placeins}
+\usepackage{lineno}
+\linenumbers
 
@@
 
-\subsection{Emission probability}
+\subsubsection{Emission probability}
 Using this set up, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta)$ from equation \ref{eq:1} and optimize them alone.
@@
 
-\subsection{Estimation}
+\subsubsection{Estimation}
 
@@
 
-\subsection{Relatedness cases}
+\subsubsection{Relatedness cases}
 We run our model for the following relatedness cases: Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, Identical.
@@
 
-\subsection{ROH estimation model}
+\subsubsection{ROH estimation model}
 Our HMM to detect ROH tracts works similar to the relatedness model described above, but we only consider one individual at a time, and only positions covered by at least two reads.
@@
 
-\subsection{Contamination}
+\subsubsection{Contamination}
 Contamination by present-day people is a common feature in ancient DNA \cite{peyregne_present-day_2020}.
@@
 
-Here, $P(d=1|c=0)$, referred afterwards to as $\rho_{ij}$ for simplicity, is equivalent to the proportion of differences when we compare endogenous reads from individuals $i$ and $j$.
-This can be estimated by considering that proportion of pairwise differences ($\frac{\sum_w D_w}{\sum_w N_w}$), referred to as $o_{ij}$ afterwards, can be expressed as a sum of proportion of differences between endogenous reads and that between endogenous and contaminant reads \cite{peyregne_present-day_2020}.
+Here, $P(d=1|c=0)$, referred later to as $\rho_{ij}$ for simplicity, is equivalent to the proportion of differences when we compare endogenous reads from individuals $i$ and $j$.
+This can be estimated by considering that proportion of pairwise differences ($\frac{\sum_w D_w}{\sum_w N_w}$), referred to as $o_{ij}$ later, can be expressed as a sum of proportion of differences between endogenous reads and that between endogenous and contaminant reads \cite{peyregne_present-day_2020}.
 \begin{align}\label{eq:13}
@@
 
-We use simulations both for estimating the transition matrices for all our models, and for testing and validating our algorithm.
-All simulations are performed using \textt{msprime} \cite{kelleher_efficient_2016}, followed by formation of related individuals using a predetermined pedigree (Figure S\ref{figS10:pedigree}).
+We use simulations both for estimating the transition matrices for all relatedness cases, and for testing and validating our algorithm.
+All simulations are performed by simulating unrelated individuals using \textt{msprime} \cite{kelleher_efficient_2016}, followed by an additional step where we simulate related individuals using a predetermined pedigree (Figure S\ref{figS10:pedigree}).
 
@@
 
-For our simulations of background diversity, we formed a haploid population (Pop1) with constant effective size of 10,000 and, sampled 16 haploids from 2500 generations ago to randomly combine into 8 unrelated diploid ancient individuals (fig.
+For our simulations of background diversity, we form a population (Pop1) with constant effective size of 10,000 and sample eight diploid individuals (each made up of two haploid individuals) from 2500 generations ago  (fig.
 \ref{figS10:pedigree}).
-For each individual, we simulated 22 chromosomes with length $L=100$ Mb.
-The mutation rate was set as $\mu= 10^{-8}$ per base pair per generation and recombination rate used was $r=10^{-8}$
+For each individual, we simulate 22 chromosomes with length $L=100$ Mb and a recombination rate of $r=10^{-8}$ per base pair per generation.
+We introduce mutations using an infinite sites model with rate  $\mu= 10^{-8}$ per base pair per generation.
 
-For artificially mating two individuals, we simulated a recombined chromosome from each individual to be passed on to the progeny.
-To simulate a recombined chromosome from a diploid individual, we drew the number of recombination points from a Poisson distribution with parameter $rL$ as the recombination rate, and used a uniform distribution to sample the positions of recombination points.
-We estimated transition matrices corresponding to different cases of relatedness by counting transition between IBD states for a pair of  individuals at a particular relatedness level.
-We compared the transition matrices for grandparent-grandchild and siblings estimated in this way to a theoretical calculation, and verified that they are very close.
-However, it is difficult to theoretically calculate transition matrices for other cases like third degree, and so for these cases we use the transition matrices obtained from simulations.
+For the pedigree simulations, we first simulate a recombined set of chromosomes for either parent, and combine them to create the progeny.
+We simulate recombination by first drawing the number of breakpoints as a  Poisson random variable with parameter $rL$, and use a uniform distribution on $[1, L]$ to sample the positions of recombination points.
 
-\paragraph{Model Comparison}
-A model with maximum likelihood is reported, and log likelihood ratio is calculated with the two highest likelihood models.
+\paragraph{Estimating transition matrices}
+KIn requires a transition matrix for each relatedness case.
+For the simplest cases like grandparent-grandchild and siblings, it is possible to calculate the transition matrix from first principles, but for most cases this is difficult  because the relatedness cases include multiple cases, or they are non-Markovian, and so  and so for these cases we obtain transition matrices from our simulations.
+
+For each relatedness case, we estimate the transition matrix  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of XX simulations.
 
@@
 
-We predict the following categories of relatedness with KIN: Identical individuals (twins), First degree (parent-child), First degree (siblings), Second degree, Third degree, Unrelated (including higher than third degree relatives).
+\paragraph{Motivating Example}
+We predict different categories of relatedness with KIN, and test it's performance on simulated pedigrees.
+We use these simulations to find the limits of our method, and compare it to READ, a widely used method.
+We introduce cases of contamination, ascertainment, low coverages, and ROH to compare the two methods.
+Finally we apply KIN to published data and show that it can give reliable information about genetic kinship, that may not be available from other available methods.
 We show the output of different Relatedness HMMs in Fig.\ref{fig1:ibd}, when they are applied to a pair of siblings.
@@
     \centering
-    \caption{Comparison of IBD fragments predicted by different relatedness models.
-Top panel shows the proportion of differences in each window along the genome for a pair of simulated siblings.
-Dashed lines in this panel represent $p_0$, $p_1$ and $p_2$ estimates.
-The panel just below shows the true IBD state for each window.
-Each panel below shows the IBD states predicted by a particular relatedness model.
+    \caption{\textbf{Comparison of pairwise difference data  and inferred IBD fragments}:  The top panel shows the proportion of differences in each window along the genome for a pair of simulated siblings.
+Dashed lines represent $p_0$, $p_1$ and $p_2$ estimates.
+The second panel shows the true IBD state for each window.
+The remaining panels show the IBD states predicted by particular relatedness models.
 The log likelihood value for each model is shown on upper left corner of the panel.
-}
+Light and shaded background represent distinct chromosomes.}
     \label{fig1:ibd}
@@
 
-We see that in this case, the siblings model predictions match true IBD states the most.
-This is reflected in the log likelihoods values (LL) of the models as well, where siblings model has the highest value.
-We take a log likelihood ratio ($\Lambda$) of the two most likely models to get a confidence in the relatedness estimate (in this case $\Lambda$ = 22.48).
-All the models in Fig.\ref{fig1:ibd} utilize information about ROH given by the ROH-HMM.
+We see that in this case, the siblings model predictions match true IBD states the most, as reflected by the highest log-likelihood, and the close match between the estimated and true IBD states.
 
-Performance of ROH-HMM under different cases of ascertainment, contamination and long ROH, and for different coverages is shown in Fig.\ref{fig2:ROH}.
-We observe that as coverage decreases, data becomes more noisy.
-This causes ROH HMM to predict some probability of ROH in few windows when there is no ROH in data.
-However, at 4x and 0.5x, this is very rare.
-In presence of ROH, we see a similar trend.
-At 4x the model can estimate ROH probability well, but loses accuracy at 0.2x.
-In presence of ascertainment, overlapping sites are reduced, but they are informative.
-We see that the model performs well at 0.5x.
-New plot for contamination to be included.
-We applied the ROH-HMM and relatedness HMM on 60 runs of simulations, to get relatedness estimates, and compared the final relatedness estimates to the true relatedness for different pairs at coverages ranging from 4x to 0.1x.
+\paragraph{Evaluating the ROH detection}
+We applied the ROH-HMM and relatedness HMM on 60 runs of simulations, to get relatedness estimates, and compared the final relatedness estimates to the true relatedness for different pairs at coverages ranging from 4x to 0.05x.
+Performance of ROH-HMM under scenarios of ascertainment bias, contamination and long ROH (described in section Simulations), and for different coverages is shown in Fig.\ref{fig2:ROH}.
+We observe that as coverage decreases, the raw heterozygosity estimates become more noisy.
+In the case of no ROH in the data (Fig \ref{fig2:ROH}A, we correctly infer that there is no ROH for coverage of 0.2x and 4x, but at the lowest coverage of 0.1x there are few regions where we erroneously call some ROH.
+In all three scenarios with ROH ((Fig \ref{fig2:ROH}), the model prediction deviates from the true ROH probability by 8-12 \%) for the three coverages.
+Hence, we do not see any large adverse effects of ascertainment bias and contamination on the performance of ROH HMM.
 
@@
 
+\paragraph{Critical Values}
+The log likelihood-ratio ($\Lambda$) is commonly used in likelihood-ratio tests to compare nested models.
+However, in our cases, models are not nested, and so standard likelihood-ratio theory does not apply.
+For this reason, we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values.
 We plotted the true positive and false positive rates with a cutoff on the log likelihood ratio in Fig.\ref{fig3:cutoff}.
@@
 This is useful for comparison to READ, since READ classifies relatedness up to Second Degree.
-'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when Third Degree is considered as separate from the unrelated individuals.
-Fig.\ref{fig3:cutoff} shows that for Identical, First Degree, Unrelated pairs, the false positive rate is below $5\%$ even when no cut off is used.
-For second degree, at 0.1x we see a high false positive rate, that does not go down much with an increase in cutoff.
-For the same relationship, at a coverage of 0.2x, a cutoff of 1 seems appropriate to reduce false positive rate to below $5\%$.
-Similarly, for siblings at 0.1x, a cutoff of 1 works well to keep false positives low.
-For parent-child, we find that a cutoff of about 2 is a better choice.
-Degree 3 shows higher false positive rates in general, and we see that only for 4x coverage and a cutoff of 2, do we see a false positive rate of around $5\%$.
-In conclusion, it is recommended to use a cutoff of 1 for most cases, with the exception of differentiation between parent-child and siblings, where a cutoff of 2 may be more appropriate, and degree 3 where cutoff of 2 at around 4x coverage seems reasonable.
+'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is considered as separate from the unrelated individuals.
+Fig.\ref{fig3:cutoff} shows that for all relatedness cases except for $3^{rd}$ Degree, the false positive rate is below $5\%$ even when no cut off is used.
+We observe that for $3^{rd}$ Degree relation, using a cutoff of 1 brings down the false positive rate close to $5\%$ for all coverages except 0.05x.
+In conclusion, it is recommended to use a cutoff of 1 for most cases.
 
@@
 Fig.\ref{fig4:Comparison_READ_KIN} shows this comparison using false positive and true positive rates for both the methods at different coverages under different simulations.
-We show that both methods have similar performance in the control case (except for 0.1x coverage, where KIN has higher power to detect First Degree and Second Degree, and slightly higher false positive rate for degree 2).
-However, in case of ascertainment, ROH, or contamination KIN shows higher power to detect relatedness cases while maintaining a low false positive rate.
+We show that both methods have similar performance in the control case, and cases with ROH and ascertainment, except for $2^{nd}$ Degree relation, where KIN has higher power, especially at low coverages.
+We show that adding contamination adversely affects both the False Positve and True Positive rates corresponding to 1st and 2nd Degree relations for READ, while this has no effect on KIN performance.
 
@@
 Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}.
-We found identical individuals (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), first degree relative (Chagyrskaya07-Chagyrskaya17), and a second degree relative (Chagyrskaya01-Chagyrskaya60).
+We found Identical individuals (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), $1^{st}$ Degree relative (Chagyrskaya07-Chagyrskaya17), and a $2^{nd}$ Degree relative (Chagyrskaya01-Chagyrskaya60).
 These results agree with READ (fig.
 S\ref{figS2:Chagyrskaya_READ}).
-Chagyrskaya06-Chagyrskaya14 were estimated to be First Degree by both the methods (table S1), but have been classified as Identical pair in the original study, since Chagyrskaya06 is a tooth sample that fits in mandible which is Chagyrskaya14.
+Both the methods classified Chagyrskaya06-Chagyrskaya14 to be $1^{st}$ Degree relatives with low confidence (table S1), but have been classified as Identical pair in the original study, since Chagyrskaya06 is a tooth sample that fits in mandible which is Chagyrskaya14.
 We believe that the contradiction is due to non human contamination present in these libraries ($1\%$ and $2.9\%$ respectively).
@@
 Our result matches with the conclusions of the authors based on additional information from mtDNA.
-Further, we find a possible third degree relative (Chagyrskaya07-Chagyrskaya60).
+Further, we find a possible $3^{rd}$ degree relative (Chagyrskaya07-Chagyrskaya60).
 We compared our IBD estimates to lcMLKIN for all pairs for whom READ results (with s.d.

### 2022-03-21 paper_overleaf/main.tex

@@
 
-\title{Method to infer relatedness using low coverage ancient DNA}
+\title{KIN: Method to infer relatedness using low coverage ancient DNA}
 \author{Divya Ratan Popli, Benjamin M. Peter}
@@
 \noindent Knowledge of familial relationships is an important part of several research fields.
-However, estimating relatedness from ancient DNA can be be difficult since DNA sequences extracted from ancient bones may have low coverage, ascertainment bias (if the sequences have been captured with DNA probes), and contamination from modern humans.
+However, estimating relatedness from ancient DNA can be difficult since DNA sequences extracted from ancient remains may have low coverage, ascertainment bias (if the sequences have been captured with DNA probes), and contamination from modern human DNA.
 In addition, the population of interest may have long runs of homozygosity (ROH) due to recent inbreeding or/and small population size.
@@
 
-In ancient DNA, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an  understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_KINship-based_2019,sikora_ancient_2017}.
-For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals in the mating animals is important to ensure diversity.
+In ancient DNA studies, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_KINship-based_2019,sikora_ancient_2017}.
+For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to ensure diversity.
 ~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}
@@
 There are numerous methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to estimate the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}.
-PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each SNP in samples \cite{purcell_plink_2007}.
-KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}.
+PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each Single Nucleotide Polymorphism (SNP) in the analyzed samples \cite{purcell_plink_2007}.
+KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies in the samples (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}.
 .
@@
 However, these approaches typically still require at least 2x coverage, since genotype likelihoods may not be very informative at lower coverages.
-Apart from the low amount of endogenous DNA sequences present, ancient DNA analyses often have additional challenges like contamination from modern populations \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in case of captured sequences.
+Apart from the low amount of endogenous DNA sequences present, ancient DNA analyses often have additional challenges like contamination from present-day DNA \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in case of captured sequences.
 In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations.
@@
 Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4x$ coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}.
-In this study, we show that contamination from modern humans can cause some of the individuals in ancient populations to look more unrelated than they are.
-This happens because of comparison of endogenous reads to contaminating reads which would give higher differences than comparison of endogenous reads from individuals from ancient population.
-It is worth mentioning that different DNA libraries may have different proportion of contamination levels, even when they have been sampled from the same specimen.
+In this study, we show that contamination can cause some of the individuals in ancient populations to look more unrelated than they are.
+This happens because of comparison of endogenous reads to contaminating reads which yields more differences than comparison of endogenous reads from the ancient individuals.
+It is worth mentioning that different DNA libraries may have different levels of contamination, even when they have been sampled from the same specimen.
 Hence, contamination can reduce the power to detect related pairs, especially when the target population is quite diverged from contaminating population(s).
-In addition, the target populations may have long runs of homozygosity (ROH) due to a small population size, or recent inbreeding.
+In addition, the genomes of target individuals may have long runs of homozygosity (ROH) due to a small population size, or recent inbreeding.
 Long ROH causes related individuals to seem genetically closer to each other, while does not affect the genetic distance between unrelated individuals.
@@
 KING robust, for example, differentiates between different relatedness by plotting KING‐robust KINship against fraction of sites in IBS0 \cite{manichaikul_robust_2010}.
-Another method from Rosenberg,2006 \cite{rosenberg_standardized_2006} uses the fraction of sites in all three IBS states to classify related individuals in the HGDP-CEPH Human Genome Diversity Cell Line Pane.
+Another method from Rosenberg,2006 \cite{rosenberg_standardized_2006} uses the fraction of sites in all three IBS states to classify related individuals in the HGDP-CEPH Human Genome Diversity Cell Line Panel.
 These methods may give different results when applied to data with different ascertainment schemes \cite{waples_allele_2019}.
@@
 We output the relatedness corresponding to the maximum likelihood model, and the confidence as given by the log likelihood ratio between the two highest likelihood models.
+The log likelihood-ratio is commonly used in likelihood-ratio tests to compare nested models.
+However, in our cases, models are not nested, and so standard likelihood-ratio theory does not apply.
+For this reason, we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig \ref{fig3:cutoff}).
 
@@
 \subsubsection{Contamination}
-Contamination by present-day people is a common feature in ancient DNA \cite{peyregne_present-day_2020}.
+Contamination by the DNA from present-day people is a common feature in ancient DNA \cite{peyregne_present-day_2020}.
 Even low levels of contamination may make samples look less similar, and thus distort relatedness analyses.
@@
 With probability $C_i$ the read comes from a contaminant.
-Thus, when comparing two individuals $i$ and $j$, there are three possible cases: First, with probability $(1-C_i)(1-C_j)$, we compare endogenous reads, which we can use to infer relatedness.
-With probability $C_i(1-C_j) + C_j(1-C_i)$ we compare an endogenous read from an individual with a contaminant read from the other individual.
-Finally, with probability $C_iC_j$ we compare contaminant reads from both the samples.
+Thus, when comparing two reads, one drawn from individual $i$ and the other drawn from individual $j$, there are three possible cases: First, with probability $(1-C_i)(1-C_j)$, both reads are endogenous and we can use them to infer relatedness.
+With probability $C_i(1-C_j) + C_j(1-C_i)$, one read is endogenous and the other is contaminant.
+Finally, with probability $C_iC_j$, both reads are contaminant.
 If we make the simplifying assumption that both $C_i$ and $C_j$ are small, the terms of the order of $C^2$ can be ignored.
@@
 
-Our goal then, is to estimate an expectation of differences from comparison of endogenous reads $\mathbb{E}[D_w^{'}]$, and the total number of sites with overlapping endogenous reads $\mathbb{E}[N_w^{'}]$.
+Given that we know the level of contamination (estimated from other methods),our goal then, is to estimate the expected number of differences from comparison of endogenous reads $\mathbb{E}[D_w^{'}]$, and the total number of sites with overlapping endogenous reads $\mathbb{E}[N_w^{'}]$.
 To estimate $\mathbb{E}[D_w^{'}]$, we first need to calculate $P(c=0|d=1)$.
@@
 Here, $P(d=1|c=0)$, referred later to as $\rho_{ij}$ for simplicity, is equivalent to the proportion of differences when we compare endogenous reads from individuals $i$ and $j$.
-This can be estimated by considering that proportion of pairwise differences ($\frac{\sum_w D_w}{\sum_w N_w}$), referred to as $o_{ij}$ later, can be expressed as a sum of proportion of differences between endogenous reads and that between endogenous and contaminant reads \cite{peyregne_present-day_2020}.
+This can be estimated by considering that the proportion of pairwise differences $P(d=1) =\frac{\sum_w D_w}{\sum_w N_w}$, referred to as $o_{ij}$ later, can be expressed as a linear combination of the proportion of differences between endogenous reads and that between endogenous and contaminant reads \cite{peyregne_present-day_2020}.
 \begin{align}\label{eq:13}
@@
 
-Here, $\phi$ is the probability of seeing a difference when endogenous read is compared to a contaminant read, and is equivalent to the average divergence between target and contaminating population.
+Here, $\phi$ is the probability of seeing a difference when an endogenous read is compared to a contaminant read, and is equivalent to the average sequence divergence between individuals of the target and contaminating populations.
 Given this information, we can rewrite eq. \ref{eq:12a}:
@@
 
-Similarly, we derive $\mathbb{E}[S_w^{'}]$ as the estimate for number of sites that do not show pairwise difference.
+Similarly, we derive $\mathbb{E}[S_w^{'}]$ as the estimate for the number of sites that do not show pairwise differences.
 \begin{align}
@@
 \paragraph{Estimating transition matrices}
-KIn requires a transition matrix for each relatedness case.
-For the simplest cases like grandparent-grandchild and siblings, it is possible to calculate the transition matrix from first principles, but for most cases this is difficult  because the relatedness cases include multiple cases, or they are non-Markovian, and so  and so for these cases we obtain transition matrices from our simulations.
+KIN requires a transition matrix for each relatedness case.
+For the simplest cases like grandparent-grandchild and siblings, it is possible to calculate the transition matrix from first principles, but for most cases this is difficult  because the relatedness cases include multiple cases, or they are non-Markovian, and so for these cases we obtain transition matrices from our simulations.
 
@@
 
-Apart from the related and unrelated individuals in Pop1, we simulated more haploids in three other populations to create scenarios with ascertainment and contamination (fig.
+Apart from the related and unrelated individuals in Pop1, we simulated more haploids in three other populations to create scenarios with ascertainment bias and contamination (fig.
 \ref{figS10:pedigree}).
@@
 We further simulated 10 haploids (5 diploid individuals) from Pop4 with split time of 20,000 generations with pop1, sampled from the present time.
-We calculated the allele frequency for Pop4 at each site, and used it to introduce contamination (with fixed contamination percentage for each individual ($<5\%$) in samples from pop1.
+We calculated the allele frequency for Pop4 at each site, and used it to introduce contamination (with fixed contamination percentage for each individual ($<5\%$) in samples from Pop1.
 We generated reads (derived/ancestral) for different genomic coverages ranging from 4x to 0.03x at each position for each individual.
@@
 
+We run simulations with different cases referred to as control (with no ROH, contamination or ascertainment bias), R (with ROH), A (with ascertainment bias), and C (with fixed levels of contamination for each individual).
+
 \section{Results}
@@
 We predict different categories of relatedness with KIN, and test it's performance on simulated pedigrees.
-We use these simulations to find the limits of our method, and compare it to READ, a widely used method.
-We introduce cases of contamination, ascertainment, low coverages, and ROH to compare the two methods.
+We use these simulations to find the limits of KIN as well as ROH HMM.
+We introduce cases of contamination, ascertainment, ROH, and low coverages to compare our method to READ.
 Finally we apply KIN to published data and show that it can give reliable information about genetic kinship, that may not be available from other available methods.
-We show the output of different Relatedness HMMs in Fig.\ref{fig1:ibd}, when they are applied to a pair of siblings.
+Here, in Fig.\ref{fig1:ibd} we show the output of different Relatedness HMMs when they are applied to a pair of siblings.
 
@@
 
-\paragraph{Evaluating the ROH detection}
-We applied the ROH-HMM and relatedness HMM on 60 runs of simulations, to get relatedness estimates, and compared the final relatedness estimates to the true relatedness for different pairs at coverages ranging from 4x to 0.05x.
-Performance of ROH-HMM under scenarios of ascertainment bias, contamination and long ROH (described in section Simulations), and for different coverages is shown in Fig.\ref{fig2:ROH}.
-We observe that as coverage decreases, the raw heterozygosity estimates become more noisy.
-In the case of no ROH in the data (Fig \ref{fig2:ROH}A, we correctly infer that there is no ROH for coverage of 0.2x and 4x, but at the lowest coverage of 0.1x there are few regions where we erroneously call some ROH.
-In all three scenarios with ROH ((Fig \ref{fig2:ROH}), the model prediction deviates from the true ROH probability by 8-12 \%) for the three coverages.
-Hence, we do not see any large adverse effects of ascertainment bias and contamination on the performance of ROH HMM.
-
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot.png}
-    \centering
-    \caption{Estimation of ROH probabilities along the genome in simulations with different cases of ROH, contamination, and ascertainment.
-Top panel shows proportion of differences in a simulated individual along the genome, and the bottom panel shows probability of seeing no ROH.
-(A) Simulation with no ascertainment, contamination, or ROH.
-(B) Simulation with ROH.
-(C) Simulation with ROH and ascertainment.
-(D) Simulation with ROH and contamination.}
-    \label{fig2:ROH}
-\end{figure}
-
 \paragraph{Critical Values}
-The log likelihood-ratio ($\Lambda$) is commonly used in likelihood-ratio tests to compare nested models.
-However, in our cases, models are not nested, and so standard likelihood-ratio theory does not apply.
-For this reason, we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values.
  We plotted the true positive and false positive rates with a cutoff on the log likelihood ratio in Fig.\ref{fig3:cutoff}.
@@
 This is useful for comparison to READ, since READ classifies relatedness up to Second Degree.
-'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is considered as separate from the unrelated individuals.
+'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.
 Fig.\ref{fig3:cutoff} shows that for all relatedness cases except for $3^{rd}$ Degree, the false positive rate is below $5\%$ even when no cut off is used.
@@
 
+In our assessment of critical value for log likelihood ratio, we used the control simulations.
+However, we test performance of KIN on different simulation cases including case R, which is also based on performance of ROH HMM.
+
+\paragraph{Evaluating the ROH detection}
+ Fig.\ref{fig2:ROH} shows the performance of ROH-HMM under cases of control, R, A, C and simulations (described in section Simulations), and for different coverages.
+We observe that as coverage decreases, the raw heterozygosity estimates become more noisy.
+In the control case (Fig \ref{fig2:ROH}, we correctly infer that there is no ROH for most of the time, but occasionally we erroneously call some ROH as shown in the figure.
+We observe that the model predicts ROH probabilities with biases of upto -0.007, -0.081, -0.083, and -0.091 and variances upto 0.006, 0.036, 0.036, and 0.037 in control, R, C, and A cases respectively as shown in supplementary table \ref{tab:Table 2}.
+The negative bias in all cases shows that on an average, the model over-estimates probability of ROH, and this effect is seen more often at higer coverages.
+We do not see any large adverse effects of ascertainment bias and contamination on the performance of ROH HMM.
+
+\begin{figure}[h!]
+    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot_final.png}
+    \centering
+    \caption{Estimation of ROH probabilities along the genome in simulations with different cases of ROH, contamination, and ascertainment.
+Top panel shows proportion of differences in a simulated individual along the genome, and the bottom panel shows probability of seeing no ROH.
+(A) Simulation with no ascertainment, contamination, or ROH.
+(B) Simulation with ROH.
+(C) Simulation with ROH and ascertainment.
+(D) Simulation with ROH and contamination.}
+    \label{fig2:ROH}
+\end{figure}
 We use a cutoff of 1 to compare our method performance to READ, while using a cutoff of 1 standard deviation for READ.
@@
 \begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/plotimg/fil0_relatable_plot.png}
+    \includegraphics[width=18cm]{plots/inkscape_finalImg/kinplot.png}
     \centering
@@
 However, there are only 31 pairs where KIN has $\Lambda >1$.
-We looked in detail at some of these pairs, and we found that it is possible to deduce the correct relatedness for several of these pairs by plotting the proportion of differences along the genome.
+We looked in detail at these pairs, and we found that it is possible to deduce the correct relatedness for several of these pairs by plotting the proportion of differences along the genome.
 In all the pairs that we analyzed in this way, KIN seems to predict the true relatedness.
@@
 \section{Supplementary Information}
+
+\begin{table}
+\caption{\label{tab:Table 2}Model performance for ROH prediction}
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
+    \hline
+    Coverage & Control(bias) & Control(var) & R(bias) & R(var) & RC(bias) & RC(var) & RA(bias) & RA(var)\\
+    \hline
+    4x & -0.007 & 0.006 & -0.081 & 0.036 & -0.083 & 0.036 & -0.091 & 0.037\\
+    \hline
+    0.2x & -0.003 & 0.003 & -0.068 & 0.03 & -0.068 & 0.026 & -0.076 & 0.029\\
+    \hline
+    0.1x & -0.003 & 0.002 & -0.049 & 0.021 & -0.05 & 0.021 & -0.057 & 0.022\\
+    \hline
+\end{tabular}
+\label{table1}
+\end{table}
 

### 2022-03-31 supplementary_info/beta_bnds.py
added title to supp fig. 13 and 14 constraint figs

@@ -658,6 +658,9 @@ def plot_betas(dfile, tfile, pair, thresh, rel, libraries, listf, Afile, pfile, 
         ax[i].hist(diff/total,alpha=0.3,weights=np.ones(len(total)) / len(total))
         ax[i].set_ylim(0,0.4)
 
+    ax[0].set_title('Without Constraint')
+    ax[1].set_title('With Constraint')
+
     ax[1].legend(loc="upper right")
     fig.text(0.5, 0.04, 'Frequency', ha='center')
     fig.text(0.04, 0.5, 'Proportion of differences', va='center', rotation='vertical')

### 2022-03-31 supplementary_info/forcedROH.py

@@ -464,16 +464,19 @@ def plotbetas(dfile, tfile, ind, chrmf, p1file, upA, targets, outplot):
 
         a=betas[0]
         b=betas[1]
-        ax[i].plot(x, beta.pdf(x, a, b)/500,'b-', lw=3, alpha=0.6, label="$Y_w$=0")
+        ax[i].plot(x, beta.pdf(x, a, b)/500,'b-', lw=3, alpha=0.6, label="$Y_w$=3")
 
         a=betas[2]
         b=betas[3]
-        ax[i].plot(x, beta.pdf(x, a, b)/500,'r-', lw=3, alpha=0.6, label="$Y_w$=1")
+        ax[i].plot(x, beta.pdf(x, a, b)/500,'r-', lw=3, alpha=0.6, label="$Y_w$=2")
         ax[i].hist(diff/total,alpha=0.3, weights=np.ones(len(diff)) / len(diff))
 
         ax[i].set_ylim(0,1)
         ax[i].set_xlim(0,0.04)
-        
+
+    ax[0].set_title('Without Constraint')
+    ax[1].set_title('With Constraint')
+
     ax[1].legend(loc="upper right")
     fig.text(0.5, 0.04, 'Proportion of differences', ha='center')
     fig.text(0.04, 0.5, 'Frequency', va='center', rotation='vertical')

### 2022-03-31 paper_overleaf/main.tex

@@
 \title{KIN: Method to infer relatedness using low coverage ancient DNA}
-\author{Divya Ratan Popli, Benjamin M. Peter}
+\author{Divya Ratan Popli, Stéphane Peyrégne, Benjamin M. Peter}
 \date{5 August 2021}
@@
 \noindent Knowledge of familial relationships is an important part of several research fields.
-However, estimating relatedness from ancient DNA can be difficult since DNA sequences extracted from ancient remains may have low coverage, ascertainment bias (if the sequences have been captured with DNA probes), and contamination from modern human DNA.
+Knowing genetic kinship of ancient specimens can lead to interesting insights into the culture and social hierarchy of ancient populations, and is also relevant for downstream genetic analyses.
+However, estimating relatedness from ancient DNA can be difficult since DNA sequences extracted from ancient remains may have low coverage, ascertainment bias, and contamination from modern human DNA.
 In addition, the population of interest may have long runs of homozygosity (ROH) due to recent inbreeding or/and small population size.
 This can affect the relatedness estimates.
-Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness up to 3rd degree, while additionally differentiating between siblings and parent-child.
+Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness up to 3rd degree, while additionally differentiating between siblings and parent-child with as low as 0.05x coverage DNA data.
 The output can be further refined with another HMM that we developed to estimate ROH proportions.
+We focus on the application of KIN on ancient human DNA, but it is applicable to other diploid species as well.
+On a machine with 8 cores, the generation of input files from 17 bam files takes -- minutes, and running KIN software takes -- minutes.
 \end{abstract}
@@
 
-Identifying related individuals is a common task in applications of genetics.
+Identifying related individuals is a common task in genetic studies.
 Relatedness is of direct interest in e.g.
-DNA forensics, where familial search can aid in solving criminal cases, and to identify missing persons following a disaster \cite{murphy_law_2018,ram_genealogy_2018}.
+DNA forensics, where familial search can aid in solving criminal cases, and to identify unknown deceased persons \cite{murphy_law_2018,ram_genealogy_2018}.
 Genetic paternity tests have an important application in resolving family relation, e.g.
@@
 In ancient DNA studies, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_KINship-based_2019,sikora_ancient_2017}.
-For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to ensure diversity.
+For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to avoid inbreeding and ensure diversity.
 ~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}
 
-\subsection{Simple way to get relatedness, methods that do it with a lot of data}
+\subsection{Approaches to estimate relatedness from high coverage data}
 
@@
 inherited from a recent common ancestor.
-Due to the laws of Mendelian segregation, a parent, for example, will share exactly one set of chromosome IBD with its offspring, with the other set of chromosomes in the offspring coming from the other parent.
-A grandparent will, on average share a quarter of its genome with a grand-child, because recombination broke up the chromosomes.
+Due to the laws of Mendelian segregation, each parent, for example, will share exactly one set of chromosomes IBD with their offspring.
+A grandparent will, on average share a quarter of their genome with a grand-child due to recombination.
+Along the genomes of a pair of diploid individuals, there are three IBD states possible at any given position: either no chromosomes are IBD, only one chromosome is IBD, or both chromosomes are IBD.
+The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals.
+For example, a pair of siblings are expected to have all three possible IBD states with probabilities (0.25,0.5,0.25) in the absence of inbreeding, as shown in Fig \ref{fig0:schematic}.
+These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}.
+One can also use these probabilities to estimate the coefficient of relatedness $r$, which is defined as the proportion of the genome that is IBD.
+In the absence of inbreeding, this would be calculated as $r= k_1/2 + k_2$.
 
 However, it is not possible to directly observe IBD segments, and so most methods infer them by first identifying segments of the genome that are Identical by State (IBS) and using population allele frequencies to calculate the probability of IBD given IBS \cite{vai_KINship_2020}.
-One exception is SNPduo, a software that uses IBS direclty to visualize and analyze pairwise relatedness.
+One exception is SNPduo, a software that uses IBS directly from a pair of individuals to visualize and analyze their relatedness.
 \cite{roberson_visualization_2009}.
-Most methods, however, use IBD as an intermediary in inferring relatedness \cite{boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}.
-For a non-inbred population, there are three IBD states possible: either no chromosomes are shared, only one chromosome is shared, or both chromosomes are shared.
-The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals.
-For example, a pair of siblings are expected to have all three possible IBD states with probabilities (0.25,0.5,0.25) as shown in Fig \ref{fig0:schematic}.
-These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}.
-One can also use these probabilities to estimate the coefficient of relatedness as $r= k_1/2 + k_2$.
-The coefficient of relatedness (r) is defined as the proportion of genome IBD for a pair of individuals.
+Most methods, however, use IBD as a hidden state in inferring relatedness \cite{boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}.
 
@@
 
-There are numerous methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to estimate the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}.
+These methods utilize population allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}.
+There are many such methods available.
 PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each Single Nucleotide Polymorphism (SNP) in the analyzed samples \cite{purcell_plink_2007}.
@@
 \begin{table}
-\caption{\label{tab:Table 1}IBD sharing probabilities for different relations}
+\caption{\label{tab:Table 1}IBD sharing probabilities for different relations in absence of inbreeding}
 \begin{tabular}{|c|c|c|c|}
@@
 
-\subsection{Problems with ancient data, and methods that address these issues}
-One major issue with applying the above mentioned methods to ancient DNA is that, frequently, the amount of endogenous DNA is very low, making it very difficult to get genotype calls.
-Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}.In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods.
-However, these approaches typically still require at least 2x coverage, since genotype likelihoods may not be very informative at lower coverages.
-Apart from the low amount of endogenous DNA sequences present, ancient DNA analyses often have additional challenges like contamination from present-day DNA \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in case of captured sequences.
+\subsection{Approaches that address problems with ancient DNA data}
+One major issue with applying the above mentioned methods to ancient DNA data is that the sequence coverage is typically low, making it difficult to get genotype calls.
+Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}.
+In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods.
+However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be unreliable at lower coverages.
+Ancient DNA analyses often face additional challenges such as contamination from present-day DNA \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in the case of data generated from DNA capture.
 In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations.
 However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}.
-Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4x$ coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}.
+Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}.
 In this study, we show that contamination can cause some of the individuals in ancient populations to look more unrelated than they are.
-This happens because of comparison of endogenous reads to contaminating reads which yields more differences than comparison of endogenous reads from the ancient individuals.
+This happens because the DNA sequence of an ancient individual typically exhibits more differences to a contaminant sequence than that of another ancient individual from the same population.
 It is worth mentioning that different DNA libraries may have different levels of contamination, even when they have been sampled from the same specimen.
-Hence, contamination can reduce the power to detect related pairs, especially when the target population is quite diverged from contaminating population(s).
+Hence, contamination can reduce the power to detect related pairs, especially when the target population is quite diverged from the contaminating population(s).
 In addition, the genomes of target individuals may have long runs of homozygosity (ROH) due to a small population size, or recent inbreeding.
-Long ROH causes related individuals to seem genetically closer to each other, while does not affect the genetic distance between unrelated individuals.
-This effect may cause an increase in false positives.
-Moreover, ancient DNA is commonly captured with a SNP array to enrich for endogenous sequences.
+Long ROH cause related individuals to seem genetically closer to each other, but do not affect the genetic distance between unrelated individuals.
+This effect may cause an increase in falsely inferred relationships.
+Moreover, ancient DNA is commonly captured with a SNP array to enrich for sequences from the target species.
 The way that these SNPs are defined (e.g.
-selecting SNPs variable in certain individuals) becomes relevant, because relatedness methods based on the fraction of sites in different IBS states are highly sensitive to ascertainment bias.
-KING robust, for example, differentiates between different relatedness by plotting KING‐robust KINship against fraction of sites in IBS0 \cite{manichaikul_robust_2010}.
-Another method from Rosenberg,2006 \cite{rosenberg_standardized_2006} uses the fraction of sites in all three IBS states to classify related individuals in the HGDP-CEPH Human Genome Diversity Cell Line Panel.
-These methods may give different results when applied to data with different ascertainment schemes \cite{waples_allele_2019}.
+selecting SNPs variable in certain individuals) becomes relevant, because methods to infer relatedness based on the fraction of sites in different IBS states are highly sensitive to ascertainment bias, and may give different results when applied to data with different ascertainment schemes \cite{waples_allele_2019}.
 
@@
 
-\section{Methods}
+\section{New Approaches}\label{new_approaches}
 
-The core of  KIN is a HMM that aims to infer relatedness and IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts and contamination estimates in each individual into account.
-If the locations of ROH tracts are unknown, we provide another HMM to coarsely estimate the location of ROH for samples  of sufficient coverage ($\geq 0.5x$).
+To infer the case of relatedness, KIN fits one HMM for each pair of individual and for each relatedness case.
+The KIN-HMM infers IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts and contamination estimates in each individual into account.
+The best-fit model is then assigned as the inferred relatedness.
+If the locations of ROH tracts are unknown, we provide another HMM (ROH-HMM) to coarsely estimate the location of ROH for samples with sufficient coverage ($\geq 0.1x$).
 Our method is available on \url{https://github.com/DivyaratanPopli/KIN_snakemake} along with a \textt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.
 
-\subsection{Relatedness Model overview} 
-Our relatedness model assumes we have low-coverage genome sequences for a number of individuals from the same population, and our goal is to infer which pairs of individuals are related.
-We subdivide the genome into $L$ large windows (typically of size 10Mb).
-The inputs of our algorithm are i) the number of overlapping sites for the $w$-th window $N_w$ for which both samples have at least one read available, ii) the expected number of pairwise differences $D_w$ at these sites, and iii) the probability of ROH in windows, by default obtained from ROH-HMM described in a following section.
+\subsection{Model description}\label{method_overview} 
+To infer the IBD-state across the genome, we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb).
+The proportion of pairwise differences in those windows will differ based on the IBD-state, and the relatedness case will determine how often we expect to switch between different IBD-states.
+The cases we consider are:  Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical individuals.
+The details of the likelihood computation are given in section \ref{ll}.
+The likelihood computations require an emission (section \ref{B}) and transition probability.
+We estimate the transition matrix for each relatedness case from simulations (see section \ref{simulat}), and hence the transition probability for a HMM is fixed.
+To calculate the emission probabilities, we need to estimate a parameter $\delta$ that describes the variance in the data given the IBD state and the ROH state.
+The estimation of this parameter is essential to our inference of relatedness, since the variance in the data can change with the number of overlapping sites.
+So, our algorithm optimizes $\delta$, which is used in the calculation of emission probabilities, and we use these emission probabilities, along with the transition probabilities, to estimate hidden IBD states and to calculate the log likelihood for the HMM.
+We can then compare the log likelihood from each model to find the best fit along with a measure of uncertainty.
+We output the IBD states corresponding to the most likely model.
+Hidden IBD state of each window ($Z_w$) is estimated using standard Viterbi algorithm, and $\BZ$ corresponding to the most likely model are returned.
+
+\subsection{Input of KIN-HMM}\label{hmm_input}
+The inputs of our algorithm are i) the number of overlapping sites for the $w$-th window $N_w$ for which both samples have at least one read available, ii) the expected number of pairwise differences $D_w$ at these sites, and iii) the probability of ROH in windows, by default obtained from ROH-HMM described in section \ref{roh}.
+
 $D_w$ is the number of sites that are different between a pair of individuals, and can be estimated accurately for high coverage data by comparing genotypes.
-For low coverage samples, $D_w$ is typically calculated by comparing randomly-sampled reads (or pseudo-haploid sequences) \cite{haak_massive_2015}.
+For low-coverage samples, $D_w$ is often calculated by comparing randomly-sampled reads ("pseudo-haploid" sequences) \cite{haak_massive_2015}.
 However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
@@
 \end{align}
-Here, $\nu_i(s)$ and $\nu_j(s)$ are the derived allele frequencies for individuals i and j respectively, at SNP index $s$ in a window.
-Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g.
-$\BD = (D_1, D_2, \dots D_L)$.
-The KIN-HMM uses $\BD$ and $\BN$  to classify each window into three hidden states $Z_w \in$ ($0$, $1$, $2$), reflecting zero, one or two chromosomes shared IBD.
-Likewise, we classify each window into one of three possible ROH state $H_w \in$ ($0$, $1$, $2$), reflecting, zero, one or both individuals being in a run-of-homozygosity at this genomic location.
+Here, $\nu_i(s)$ and $\nu_j(s)$ are the proportions of reads carrying the derived allele at SNP index $s$ for individuals i and j respectively.
+Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$.
 
-There are two additional parameters, $\pi$ and $\mathbf{A}$.
+\subsection{Log likelihood of the KIN-HMM}\label{ll}
+The KIN-HMM uses $\BD$ and $\BN$ to classify each window into three hidden states $Z_w \in$ ($0$, $1$, $2$), reflecting zero, one or two chromosomes IBD.
+To take ROH into account, we define the variable $H_w \in$ ($0$, $1$, $2$) that designates that zero, one or both individuals are homozygous in window $w$.
+Since $H_w$ is unobserved, in practice we use the estimates from ROH-HMM $h_{wj} = P(H_w = j)$.
+
+There are three additional model parameters, $\pi, \mathbf{A}$ and $\delta$.
 The initial probabilities $\pi$ give the probabilities of being in each state $Z$ at the beginning of each chromosome, which we set to uniform for simplicity.
-The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed for each relatedness category (see section YYY).
+The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed for each relatedness case.
+The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).
 For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.
@@
 \begin{align}\label{eq:1}
-\log P(\BD,\BZ|\BH, \theta) &= \log P(\BD|\BZ,\BH, \theta) + \log P(\BZ |\theta) \nonumber\\
-&= \sum_w \log P(D_w|Z_w,H_w, \theta) + \sum_w \log P(Z_w |Z_{w-1},\theta) + \log P(Z_0|\theta)
+\log P(\BD,\BZ|\BH, \theta, \delta) &= \log P(\BD|\BZ,\BH, \theta, \delta) + \log P(\BZ |\theta) \nonumber\\
+&= \sum_w \log P(D_w|Z_w,H_w, \theta, \delta) + \sum_w \log P(Z_w |Z_{w-1},\theta) + \log P(Z_0|\theta)
 \end{align}
 
-\subsubsection{Emission probability}
-Using this set up, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta)$ from equation \ref{eq:1} and optimize them alone.
+\subsection{Emission probability}\label{B}
+
+Using this set up, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta, \delta)$ from equation \ref{eq:1} and optimize them alone.
 The number of shared sites $N_w$ is the only parameter in $\theta$ that affects the emissions.
 
-Assuming sites are equally distributed and independent would result in the binomial likelihood:
+Assuming the sites were equally distributed and independent would result in the binomial likelihood:
 $$P(D_w|Z_w, H_w, N_w) \sim \text{Binom}[D_w ; p(Z_w, H_w), N_w] \text{,}$$
 
-where, $p$ is the expected proportion of differences we would expect for a particular IBD and ROH state.
+where, $p$ is the proportion of differences expected for a particular IBD and ROH state.
 If the two individuals are unrelated in a particular window (i.e.
@@
 The proportion of differences between unrelated individuals $p_0$ is an important parameter.
-We follow READ \cite{kuhn_estimating_2018} and estimate it as the median of differences for all possible pairs of individuals, which should work well if the majority of individuals in the samples are unrelated.
+We follow READ \cite{kuhn_estimating_2018} and estimate $p_0$ as the median of differences for all possible pairs of individuals, which should work well if the majority of individuals in the samples are unrelated.
 We set $p_1$ to $(3/4)p_0$ and $p_2$ to $p_0/2$.
@@
 We take this into account by adding an overdispersion parameter $\delta$ that allows for variation in the $p$ between windows.
-Note that, just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ has effectively four unique parameters $\delta_i$ with $i$ $\in$ (0,1,2,3).
-This is due to the fact that several combinations of $Z_w$ and $H_w$ can lead to comparison between same number of unique chromosomes, and such states will have same mean as well as the same variance.
-This results in the beta-binomial likelihood, that can be fitted to the data (fig.
+Note that, just like $p(Z_w,H_w)$ the overdispersion $\delta(Z_w,H_w)$ depends on the number of chromosomes compared, there are only four parameters $\delta_i$ with $i$ $\in$ (0,1,2,3) analogous to the $p_i$.
+Empirically, we find that the beta-binomial likelihood provides a good fit to the data (fig.
 S\ref{figS1:binom})).
 
+Taken together, our emission probabilities are 
 \begin{align}\label{eq:3}
-P(D_{w}|p_i,\delta_i,N_w) &\sim BB[D_w; N_w, p_i, \delta_i] \nonumber\\
-&= \binom{N_w}{D_w}\frac{B(D_w+p_x \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})}.
+P(D_{w}|Z_w,H_w,N_w, \delta) = P(D_{w}|p_i,\delta_i,N_w) &\sim BB[D_w; p_i, \delta_i, N_w] \nonumber\\
+&= \binom{N_w}{D_w}\frac{B(D_w+p_i \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})},
 \end{align}
+where $i$ is fully determined by the combination of $Z_w$ and $H_w$ (see eq 3).
 
 This parameterization of the beta distribution in terms of expected value $p$ and overdispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parametrization in terms of $\alpha$ and $\beta$.
-It is worth mentioning here that $D_w$ is estimated as a sum of probabilities (see section XX), while both $D_w$ and $N_w$ may undergo contamination correction, which causes these estimates to be non-integral.
-We calculate factorials required in Beta Binomial likelihood with Gamma function, which is continuous and hence does not require integral values.
+We use this equation as well if some preprocessing steps result in non-integer $D_w$ and $N_w$, in which case we approximate the binomial coefficient using Gamma functions.
 
-\subsubsection{Estimation}
-
-Emission probabilities are estimated with Expectation-Maximization (EM) algorithm.
-In each iteration of the algorithm, the Expectation step involves forward-backward algorithm to calculate $P(D, Z| H, \theta, \delta)$.
-
-In maximization step we use these joint probabilities to optimize $\delta_i$ for each state i $\in$ (0,1,2,3), and update emission probabilities.
-Iterations are run, until the complete data log likelihood stops changing.
+\subsection{Estimation of $\delta$}\label{delta}
+We estimate the $\delta$-parameters using an Expectation-Maximization (EM) algorithm (REF). In an EM-algorithm, we iterate the Expecation and Maximiation Steps until the algorithm converges.
 
 \paragraph{Initialization}
-The value of $\delta_i$ is unknown to start with, and is set to a random value.
+The value of $\delta_i$ is unknown to start with, and we set it to a random value.
 
 \paragraph{Expectation step}
-In the $t$-th iteration, we calculate the posterior probability of each IBD state in each window $\gamma^{(t)}_{wi} = P(Z_w=i | \BD, \BH, \theta, \delta^{(t)})$ using the forward-backward algorithm, where $\delta^{(t)}$ is the current guess for delta.
+In the $t$-th iteration, we calculate the posterior probability of each IBD state in each window $\gamma^{(t)}_{wi} = P(Z_w=i | D_w, H_w, \theta_i, \delta_i^{(t)})$ using the forward-backward algorithm, where $\delta_i^{(t)}$ is the current guess for $\delta$ for a given IBD state.
 
@@
 The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_i$.
-We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wi}$ and the ROH state-probabilities $h_{w\omega}$.
+We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wi}$ and optionally the ROH state-probabilities $h_{w\omega}$ obtained from the ROH-HMM.
 
@@
 \mathcal{C} &= \mathbb{E}[\log P(D_w|Z_w, H_w, \theta, \delta^{(t-1)}]\nonumber\\
-&= \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta^{(t-1)}, p)h_{w\omega}\gamma_{wi}
+&= \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta_{i \omega}^{(t-1)}, p_{i \omega})h_{w\omega}\gamma_{wi}
 \end{align}
 
-Using equation \ref{eq:2}, we simplify this by grouping all the terms that would result in the same $p$, i.e.
+Using equation \ref{eq:2}, we simplify this by grouping all the terms that would result in the same $p_i$ and $\delta_i$, i.e.
 
@@
 \begin{align}\label{eq:6}
-\mathcal{C} &= \sum_{w=1}^L\sum_{i=0}^3 \log P(D_{w}|Z_w, H_w, N_w, \delta_i, p_i)g_{wi}
+\mathcal{C} &= \sum_{i=0}^3 \sum_{w=1}^L \log P(D_{w}|Z_w, H_w, N_w, \delta_i, p_i)g_{wi}
 \end{align}
 
-Since each term in equation \ref{eq:6}  depends on only one $\delta_i$ and not on different IBD and ROH states, we can optimize for them independently.
+The cost function \ref{eq:6} has one independent term for each $i$ and so we can separate them and estimate each $\delta_i$ independently using the XXX algorithm implemented in \texttt{scipy.optimize} (REF scipy).
 
-We realized that different cases of relatedness have different number of IBD states possible.
+%\paragraph{Constraints on $\delta$ optimization}\label{delta_constraint}
+
+Unconstrained optimization of the $\delta_i$ could result in some confounding of cases, We realized that different cases of relatedness have different number of IBD states possible.
 For example, siblings may have all three IBD states present while a Parent-Child has only $Z_w=1$.
-This may cause some overfitting problem, when we apply a Parent-Child HMM to a dataset corresponding to siblings.
-We show a simple example of this problem in fig.
-\ref{figS4:bndsbeta}.
-We avoid this problem by constraining the parameter space of $\delta$.
-In particular, the variance of a beta distributed random variable $X$ is:
-\begin{align}\label{eq:7}
-    var(X) = \frac{p(1-p)}{\delta + 1}
-\end{align}
-We constrain $\delta$ such that  the variance is less than a threshold T:
+However, the parent-child model could fit data generated under the sibling model by assigning it a very high $\delta_1$, which would reduce performance  \ref{figS4:bndsbeta}.
+We avoid this problem by constraining the $\delta$ such that the beta distributions for the different cases overlap by at most one standard deviation.
 
-\begin{equation}\label{eq:8}
-    \frac{p(1-p)}{\delta + 1}  < T
-\end{equation}
-
-We choose T for each beta distribution such that the distributions do not overlap much.
-
-\subsubsection{Relatedness cases}
-We run our model for the following relatedness cases: Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, Identical.
-We estimated the transition matrix corresponding to each relatedness case using simulations (see section XX).
-While testing KIN on simulations, we found that classification performance for $4^{th}$ and $5^{th}$ degree relatives is poor, and we have low power to distinguish them from unrelated pairs of individuals.
-However, we would like to classify 4-th and 5-th degree relatives as unrelated, and so we include these cases in the algorithm.
-Similarly, we observe that we have low power to differentiate between half-siblings, avuncular and grandparent-grandchild, and we merge these classifications as $2^{nd}$ degree in the final results.
-We report the final pairwise classification in the following categories: Unrelated, Third Degree, Second Degree, Parent-Child, Siblings, Identical individuals.
+\subsection{Model comparison}\label{model_comp}
+To infer the most probable relatedness case, we run our model on all cases ( section \ref{method_overview}) and compare the resulting likelihoods.
 We output the relatedness corresponding to the maximum likelihood model, and the confidence as given by the log likelihood ratio between the two highest likelihood models.
@@
 
-Hidden IBD state of each window ($Z_w$) is estimated using standard Viterbi algorithm, and $\BZ$ corresponding to the most likely model are returned.
+\paragraph{Grouping of cases}
+Particularly for low-quality data, we may not be able to distinguish all cases.
+Thus, we group $4^{th}$ and $5^{th}$ degree relatives together with unrelated.
+Similarly, we group  half-siblings, avuncular and grandparent-grandchild to  $2^{nd}$-degree relatives in the final results.
+We report the final pairwise classification in the following categories: Unrelated, Third Degree, Second Degree, Parent-Child, Siblings, Identical individuals.
 
-\subsubsection{ROH estimation model}
-Our HMM to detect ROH tracts works similar to the relatedness model described above, but we only consider one individual at a time, and only positions covered by at least two reads.
+\subsection{ROH estimation model}\label{roh}
+Our HMM to detect ROH tracts works similar to the KIN-HMM described above, but we only consider one individual at a time, and only positions covered by at least two reads.
 For each site, we calculate the proportion of reads that carry different alleles, and sum them up in windows along the genome.
 We call the vector with the number of differences $\mathbf{\Delta}$, and the vector with the number of sites with at least two reads $\mathbf{N}$.
-Our model has two possible hidden states: homozygous (state $3$), and non homozygous (state $2$).
+Our model has two possible hidden states: homozygous state ($Y_w=3$), and non homozygous state ($Y-w=2$).
 As above, we collect the hidden states in a vector $\mathbf{Y} = (Y_1, \dots Y_w, \dots, Y_L)$.
@@
 
-We give details on the transition, emission and their estimation below.
-
-In this case, transitions are not known, and we estimate both transitions and emissions using an EM algorithm.
-We calculate the emissions using beta-binomial likelihood, and fix the mean of the distributions at expected proportion of differences in a ROH tract ($p_3$) and expected proportion of differences in a non-ROH tract ($p_2$).
+Since the source of ROH may not be known,  we estimate both transitions and emissions.
+We calculate the emissions using a beta-binomial likelihood, and fix the mean of the distributions at expected proportion of differences in a homozygous tract ($p_3$) and expected proportion of differences in a non-homozygous tract ($p_2$).
 The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w=3$ or the state $Y_w=2$ in each window.
 The M-step for emissions in analogous to that in the KIN model (eq.
-\ref{eq:4}):
+\ref{eq:4}), and the optimization step here is done with the following cost function:
 
 \begin{align}\label{eq:11}
-\mathcal{C} = \sum_{w=1}^L \sum_{i=2}^3 P(\Delta_w|\eta_{i},p_{i}) \Gamma_{wi} ,
+\mathcal{C} = \sum_{w=1}^L \sum_{i=2}^3 P(\Delta_w|\delta_{i},p_{i}) \Gamma_{wi} ,
 \end{align}
-where $P(\Delta_w|\eta_{i},p_{i})$ is a beta-binomial probability given by eq. \ref{eq:3}.
+where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and overdispersion parameter $\eta$ similar to eq. \ref{eq:3}.
 
-While testing this method on low coverage data, we observed that some pair of individuals may have few sites in each window, resulting in the proportion of differences to become close to 0 or 1.
-This may result in a high variance, and can potentially influence our estimation of $\eta$.
-To avoid the estimated Beta distributions from flattening to fit this high variance data, we set $Y_w=0$ for windows with high differences (fig.
-S\ref{figS5:ROHforced}).
-Finally, to estimate transitions, we initialize the transition matrix with 0.2 for the off-diagonal entries, and update it using the standard Baum-Welch update step \cite{baum_maximization_1970}.
+To estimate transitions, we initialize the transition matrix with 0.2 for the off-diagonal entries, and update it using the standard Baum-Welch update step \cite{baum_maximization_1970}.
 
-\subsubsection{Contamination}
+Similar to the KIN-HMM, we avoid fitting issues by forcing all windows whose proportion of differences is larger than $p_2$ to be in the non-homozygous state (fig. S\ref{figS5:ROHforced}).
+
+\subsection{Contamination correction}\label{contam}
 Contamination by the DNA from present-day people is a common feature in ancient DNA \cite{peyregne_present-day_2020}.
@@
 
-------------------------------------------extra stuff-----------Do not read--------------------------------
-
-If we look at one site in the window, we could find either no difference in reads from individuals $i$ and $j$ ($d_s$=0), or a non-zero difference ($0<d_s<=1$).
-Here $d_s$ is the probability of seeing a difference at site $s$, and $\sum_{s \in w} d_s =D_w$.
-We ask, what is the
-
-If we make the simplifying assumption that both $C_i$ and $C_j$ are small, terms of the order of $C^2$ can be ignored and there are just two cases: with probability $C_{i,j} =C_i + C_j$, one of the reads stems from a contaminant, and we effectively compare a contaminant fragment with an endogenous one.
-The probability that they differ is the average divergence between target and contaminating population, which we denote with $\phi$.
-The other case (w.
-p.
-$(1-C_{i,j}$) is where we compare two endogenous reads, which will differ with probability
-
-\begin{equation}
-    p_{e} = \frac{p_{obs}-\phi C_{i,j}}{1-C_{i,j}}
-\end{equation}
-
-Here, $$p_{obs} = \frac{\sum_w D_w}{\sum_w N_w}$$
-
-We estimate $\hat{D_c} = C_{i,j} N_{tot} \phi$ and $\hat{D_e} = (1-C_{i,j}) N_{tot} p_{obs}\phi$.
-If there are $N_{obs}$ overlapping sites in a window, $N_{c} = C_{i,j}N_{obs}$ of them will include a contaminant allele, and $N_e = (1-C_{i,j})N_{obs}$ will be between endogenous alleles.
-
-For KIN, we can use $N_e$ as an input (for each window), but we also need the proportion of differences, $D_e$.
-For this, we use the assumption that the proportion of sites where the contaminant differs from our samples is
-
-$$\phi = \frac{D_c}{N_c} = \frac{D_c}{C_{i,j} N_{obs}}, $$
-and thus $D_c = C_{i,j} \phi N_{obs}$.
-
-We can think of $p_c$ and $\rho$ as empirical likelihoods that a particular site shows a difference between two reads, given that the comparisons contain only endogenous or one contaminant read, respectively.
-\begin{align}
-    P(E|D) &= \frac{P(D|E)P(E)}{P(D)} \nonumber\\
-    \mathbb{E}[D_e] &= D_{obs} P(E|D) = D_{obs}\frac{p_e (1-c)}{p_e(1-c) + c\rho}\\
-    \mathbb{E}[S_e] &= D_{obs} P(E|\text{not} D) = S_{obs}\frac{(1-p_e)(1-c)}{(1-p_e)(1-c) + c(1-\rho)}
-\end{align}
-
-And for calculating $N_{e}$, we added $D_{e}$ to $S_{e}$
-
-The goal is that given $C_{i,j}, p_e, \BN and \BD$ to calculate adjusted $\hat{\BN}$ and $\hat{\BD}$ that reflect the endogenous proportion only.
-
-, and the observed proportion of differences between individuals $i$ and $j$ as $p_{oij}$.
-We calculate average endogenous pairwise difference $p_{eij}$ for individuals i and j as follows \cite{noauthor_ancient_nodate}:
-\begin{align}
-    p_{oij} = C_{ij}  p_c + (1-C_{ij})  p_{eij}\\
-    or, p_{eij} = \frac{p_{oij} - C_{ij}  p_c}{1-C_{ij}} 
-\end{align}
-
-Given $p_{eij}$, we would ideally like to calculate the complete data log likelihood for the relatedness model while summing over all possible differences we could have in a window weighted by the probability of seeing that difference.
-
-\paragraph{Short version}
-With contamination, we do not observe the number of differences $D_{true}$ and total number of comparisons $N_{true}$ directly, but instead need to approximate them.
-The window-subscript is omitted in the following for clarity:
-\begin{equation}
-    P(D_{\text{obs}} | N_\text{obs}, Z, \theta) \approx 
-    P( \hat{D} | \hat{N}, Z, \theta)
-\end{equation}
-We do that by calculating the probability that a single read is endogenous given it has a difference.
-\begin{align}
-    \hat{D} = \mathbb{E}[D | D_{\text{obs}}, c] &= 
-    D_{\text{obs}} P(E_i | D_i=1) \\
-    &= D_\text{obs} \frac{P(D_i=1 | E_i)P(E_i)}{P(D_i=1)}\\
-    &= D_\text{obs} \frac{(1-c)p_e}{(1-c)p_e + c p_c}
-\end{align}
-
-and likewise for reads that do not have a difference
-\begin{align}
-    \hat{S} = \mathbb{E}[S | S_{\text{obs}}, c] &= 
-     S_\text{obs} \frac{(1-c)(1-p_e)}{(1-c)(1-p_e) + c (1-p_c)}
-\end{align}
-
-and $\hat{N} = \hat{S} + \hat{D}$
-
-\begin{equation}
-    P(D_{\text{obs}} | Z, N_\text{obs}, \theta) = \sum_n\sum_d P(D_{\text{obs}}, N_\text{obs} | D_w=d, N_w, C_i, C_j)\underbrace{P(D_w=d | N_w, Z, \theta)}_\text{old likelihood}
-\end{equation}
-
-i.e. we sum over all possible $D_w$, $N_w$. Instead, we use
-
-\begin{align}
-    P(D_{cor},Z|D,H,\theta,c,p_c) &= P(D_{cor}|Z,D,H,\theta,c,p_c) P(Z|D,H,\theta,c,p_c)\nonumber\\
-    &= [\prod_{w} [\sum_\kappa P(D_{cor,w}|Z_w, H_w, \theta,c,p_c) P(D_{cor,w}=\kappa)] \prod_{w} P(Z_w|Z_{w-1}, \theta)] P(Z_0| \theta)
-\end{align}
-
-We realize that this calculation would be time consuming, and would further complicate the HMM.
-To avoid this, instead of using weighted sum of all possible $\kappa$, we use $E(D_{cor,w}|Z_w, H_w, \theta,c,p_c)$.
-This simplifies the equation 21 to :
-
-\begin{align}
-    P(D_{cor},Z|D,H,\theta,c,p_c) &= P(D_{cor}|Z,D,H,\theta,c,p_c) P(Z|D,H,\theta,c,p_c)\nonumber\\
-    &= [\prod_{w} P(E(D_{cor,w})|Z_w, H_w, \theta,c,p_c) \prod_{w} P(Z_w|Z_{w-1}, \theta)] P(Z_0| \theta)
-\end{align}
-
-We calculate $E(D_{cor,w})$ as shown:
-
-\begin{align}
-    E[N_e] + E[N_c] = N\\
-    E[D_e] + E[D_c] = D
-\end{align}
-
-Probability of comparing endogenous reads at a site where we see a difference:
-$$P(\xi | N = 1,D = 1)=\frac{(1-c) p_e}{c p_c + (1-c) p_e} $$
-Probability of comparing endogenous reads at a site where we see no difference:
-$$P(\xi | N = 1,D = 0)=\frac{(1-c)(1-p_e)}{c (1-p_c) + (1-c) (1-p_e)} $$
-Expectation of number of endogenous comparisons in a window:
-\begin{align}
-    E[\xi | N, D] = D E[\xi | N=1, D=1] + (N-D) E[\xi | N=1, D=0]\\
-    = D P(\xi | N=1, D=1) + (N-D) P(\xi | N=1, D=0)
-\end{align}
-
-Total number of endogenous sites showing a difference in a window: 
-$$D_{cor,w} = D_w* \frac{p_e(1-c)}{p_e(1-c) + c p_c} $$
-Total number of endogenous sites showing 0 difference in a window: 
-$$S_{cor,w} = S_w* \frac{(1-p_e)(1-c)}{(1-p_e)(1-c) + c (1-p_c)} $$
-Total number of endogenous sites in a window:
-$$N_{cor,w} = D_{cor,w} + S_{cor,w}$$
-
-This model outputs the contamination-corrected number of differences and overlapping sites in each window.
-
-\subsection{Simulations}
-
-We use simulations both for estimating the transition matrices for all relatedness cases, and for testing and validating our algorithm.
-All simulations are performed by simulating unrelated individuals using \textt{msprime} \cite{kelleher_efficient_2016}, followed by an additional step where we simulate related individuals using a predetermined pedigree (Figure S\ref{figS10:pedigree}).
-
-\paragraph{Simulations for estimation of transition matrix}
-
-For our simulations of background diversity, we form a population (Pop1) with constant effective size of 10,000 and sample eight diploid individuals (each made up of two haploid individuals) from 2500 generations ago  (fig.
-\ref{figS10:pedigree}).
-For each individual, we simulate 22 chromosomes with length $L=100$ Mb and a recombination rate of $r=10^{-8}$ per base pair per generation.
-We introduce mutations using an infinite sites model with rate  $\mu= 10^{-8}$ per base pair per generation.
-
-For the pedigree simulations, we first simulate a recombined set of chromosomes for either parent, and combine them to create the progeny.
-We simulate recombination by first drawing the number of breakpoints as a  Poisson random variable with parameter $rL$, and use a uniform distribution on $[1, L]$ to sample the positions of recombination points.
-
-\paragraph{Estimating transition matrices}
-KIN requires a transition matrix for each relatedness case.
-For the simplest cases like grandparent-grandchild and siblings, it is possible to calculate the transition matrix from first principles, but for most cases this is difficult  because the relatedness cases include multiple cases, or they are non-Markovian, and so for these cases we obtain transition matrices from our simulations.
-
-For each relatedness case, we estimate the transition matrix  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of XX simulations.
-
-Apart from the related and unrelated individuals in Pop1, we simulated more haploids in three other populations to create scenarios with ascertainment bias and contamination (fig.
-\ref{figS10:pedigree}).
-We simulated two haploids to form an individual each from two other populations (Pop2, Pop3) with split time of 3500 and 4500 generations with Pop1, and sampling time of 4000 generations and 2000 generations ago respectively.
-We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from Pop1.
-We test out the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a Markov chain using transition matrix shown below:
-
-$$\mathbf{A} = \left[\begin{array}
-{rr}
-1-10/L & 10/L \\
-2/L & 1-2/L  \\
-\end{array}\right]
-$$
-
-From the steps described above, we got genotypes of 17 individuals in Pop1 in presence/absence of ROH and ascertainment.
-We further simulated 10 haploids (5 diploid individuals) from Pop4 with split time of 20,000 generations with pop1, sampled from the present time.
-We calculated the allele frequency for Pop4 at each site, and used it to introduce contamination (with fixed contamination percentage for each individual ($<5\%$) in samples from Pop1.
-We generated reads (derived/ancestral) for different genomic coverages ranging from 4x to 0.03x at each position for each individual.
-We specifically generated reads for four different cases: Ancestral-Endogenous, Ancestral-Contaminant, Derived-Endogenous, Derived-Contaminant.
-The number of reads for each case was sampled from a Poisson distribution with $\lambda$ calculated from the coverage $\zeta$, allele frequency or genotype (G $\in$ 0,0.5,1), and contamination estimate:
-
-\begin{align}
-    \lambda_{DE} = \zeta (1-c) G\nonumber\\
-    \lambda_{AE} = \zeta (1-c) (1-G)\nonumber\\
-    \lambda_{DC} = \zeta c \upsilon\nonumber\\
-    \lambda_{AC} = \zeta c (1-\upsilon)
-\end{align}
-
-We run simulations with different cases referred to as control (with no ROH, contamination or ascertainment bias), R (with ROH), A (with ascertainment bias), and C (with fixed levels of contamination for each individual).
-
-\section{Results}
+\section{Results}\label{results}
 
 \paragraph{Motivating Example}
-We predict different categories of relatedness with KIN, and test it's performance on simulated pedigrees.
+We predict different cases of relatedness with KIN, and test it's performance on simulated pedigrees.
 We use these simulations to find the limits of KIN as well as ROH HMM.
@@
 
+\paragraph{IBD prediction}
+We compared the IBD assignment for simulated pairs of individuals with different relations.
+
 \paragraph{Critical Values}
- We plotted the true positive and false positive rates with a cutoff on the log likelihood ratio in Fig.\ref{fig3:cutoff}.
+ To investigate the limits of our method, we plotted the true positive and false positive rates with a cutoff on the log likelihood ratio in Fig.\ref{fig3:cutoff}.
 Here, and in Fig.\ref{fig4:Comparison_READ_KIN}, we have two relatedness labels: 'Unrelated' and 'Unrelated w/o 3rd Degree'.
@@
 \paragraph{Evaluating the ROH detection}
- Fig.\ref{fig2:ROH} shows the performance of ROH-HMM under cases of control, R, A, C and simulations (described in section Simulations), and for different coverages.
+ Fig.\ref{fig2:ROH} shows the performance of ROH-HMM under cases of control, R, A, C and simulations (described in section \ref{simulat}), and for different coverages.
 We observe that as coverage decreases, the raw heterozygosity estimates become more noisy.
 In the control case (Fig \ref{fig2:ROH}, we correctly infer that there is no ROH for most of the time, but occasionally we erroneously call some ROH as shown in the figure.
-We observe that the model predicts ROH probabilities with biases of upto -0.007, -0.081, -0.083, and -0.091 and variances upto 0.006, 0.036, 0.036, and 0.037 in control, R, C, and A cases respectively as shown in supplementary table \ref{tab:Table 2}.
-The negative bias in all cases shows that on an average, the model over-estimates probability of ROH, and this effect is seen more often at higer coverages.
+We observe that the model predicts ROH probabilities with biases of upto -0.007, -0.081, -0.083, and -0.091 and variances upto 0.006, 0.036, 0.036, and 0.039 in control, R, A, and C cases respectively as shown in supplementary table \ref{tab:Table 2}.
+The negative bias in all cases shows that on an average, the model over-estimates probability of ROH, and this effect is seen slightly more often at higer coverages.
 We do not see any large adverse effects of ascertainment bias and contamination on the performance of ROH HMM.
@@
     \caption{Estimation of ROH probabilities along the genome in simulations with different cases of ROH, contamination, and ascertainment.
-Top panel shows proportion of differences in a simulated individual along the genome, and the bottom panel shows probability of seeing no ROH.
+Top panel shows proportion of differences in a simulated individual in windows along the genome, and the bottom panel shows probability of seeing no ROH.
 (A) Simulation with no ascertainment, contamination, or ROH.
@@
 
-\section{Discussion}
+\section{Discussion}\label{discussion}
 
@@
 We show that when applied to Neandertal specimens from Chagyrskaya cave, READ and KIN agree in relatedness estimates for all pairs where both methods have sufficient data.
-KIN additionally identifies a pair of first degree relatives as parent-child, which is in agreement with mtDNA analysis from the authors.
+KIN additionally identifies a pair of first degree relatives as parent-child, which is in agreement with mtDNA analysis from the authors (Laurits et al., 2022 (in press)).
 In addition, KIN identifies a third degree relative.
@@
 
+\section{Materials and methods}\label{method}
+
+\subsection{Simulations}\label{simulat}
+
+We use simulations both for estimating the transition matrices for all relatedness cases, and for testing and validating our algorithm.
+All simulations are performed by simulating unrelated individuals using $\texttt{msprime}$ \cite{kelleher_efficient_2016}, followed by an additional step where we simulate related individuals using a predetermined pedigree (Figure S\ref{figS10:pedigree}).
+
+\paragraph{Simulating pedigrees}
+
+For our simulations of background diversity, we form a population (Pop1) with constant effective size of 10,000 and sample eight diploid individuals (each made up of two haploid individuals) from 2500 generations ago (fig.
+\ref{figS10:pedigree}).
+For each individual, we simulate 22 chromosomes with length $L=100$ Mb and a recombination rate of $r=10^{-8}$ per base pair per generation.
+We introduce mutations using an infinite sites model with rate  $\mu= 10^{-8}$ per base pair per generation.
+
+For the pedigree simulations, we first simulate a recombined set of chromosomes for either parent, and combine them to create the progeny.
+We simulate recombination by first drawing the number of breakpoints as a  Poisson random variable with parameter $rL$, and use a uniform distribution on $[1, L]$ to sample the positions of recombination points.
+
+\paragraph{Estimating transition matrices}
+KIN requires a transition matrix for each relatedness case.
+For the simplest cases like grandparent-grandchild and siblings, it is possible to calculate the transition matrix from first principles (as shown in the section \ref{Aver}), but for most cases this is difficult  because the relatedness cases include multiple cases, or they are non-Markovian, and so for these cases we obtain transition matrices from our simulations.
+
+For each relatedness case, we estimate the transition matrix  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of XX simulations.
+
+\paragraph{Verifying accuracy of transition matrices}\label{A_ver}
+We verify that are estimation of transition matrices for grandparent-grandchild ($A_g$) and siblings ($A_s$) are very close to the respective transition matrices obtained from theoretical calculations.
+To calculate $A_g$, we count the number of IBD state change events that happen along chromosomes from grandparent-grandchild pair given a recombination event happens with rate $r$.
+In this case, every recombination event in the recombined chromosome inherited from grandparent leads to a state change.
+So, we define the rate matrix ($Q$) for our continuous time markov chain as $$Q = \left[\begin{array}
+{rrr}
+-r & r \\
+r & -r 
+\end{array}\right]$$
+
+Here the state space includes only $Z_w=0$ and $Z_w=1$. Similarly, for a pair of siblings, we calculate $$Q =  
+\left[\begin{array}
+{rrr}
+-4r & 4r & 0\\
+2r & -4r & 2r\\
+0 & 4r & -4r
+\end{array}\right]$$
+
+The state space includes all IBD states in this case.
+Then, we get the transition matrix in each case as $e^{Qb}$, where b is the window size.
+We show a comparison between these theoretically obtained transition matrices with the simulated ones in supp table YY.
+
+\paragraph{Simulations for method evaluation}
+Apart from the related and unrelated individuals in Pop1, we simulated more haploids in three other populations to create scenarios with ascertainment bias and contamination (fig.
+\ref{figS10:pedigree}).
+We simulated two haploids to form an individual each from two other populations (Pop2, Pop3) with split time of 3500 and 4500 generations with Pop1, and sampling time of 4000 generations and 2000 generations ago respectively.
+We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from Pop1.
+We teste out the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a Markov chain using transition matrix shown below:
+
+$$\mathbf{A} = \left[\begin{array}
+{rr}
+1-10/L & 10/L \\
+2/L & 1-2/L  \\
+\end{array}\right]
+$$
+
+From the steps described above, we got genotypes of 17 individuals in Pop1 in presence/absence of ROH and ascertainment.
+We further simulated 10 haploids (5 diploid individuals) from Pop4 with split time of 20,000 generations with pop1, sampled from the present time.
+We calculated the allele frequency for Pop4 at each site, and used it to introduce contamination (with fixed contamination percentage for each individual ($<5\%$) in samples from Pop1.
+We generated reads (derived/ancestral) for different genomic coverages ranging from 4x to 0.03x at each position for each individual.
+We specifically generated reads for four different cases: Ancestral-Endogenous, Ancestral-Contaminant, Derived-Endogenous, Derived-Contaminant.
+The number of reads for each case was sampled from a Poisson distribution with $\lambda$ calculated from the coverage $\zeta$, allele frequency or genotype (G $\in$ 0,0.5,1), and contamination estimate:
+
+\begin{align}
+    \lambda_{DE} = \zeta (1-c) G\nonumber\\
+    \lambda_{AE} = \zeta (1-c) (1-G)\nonumber\\
+    \lambda_{DC} = \zeta c \upsilon\nonumber\\
+    \lambda_{AC} = \zeta c (1-\upsilon)
+\end{align}
+
+We run simulations with different cases referred to as control (with no ROH, contamination or ascertainment bias), R (with ROH), A (with ascertainment bias), and C (with fixed levels of contamination for each individual).
+
 \section{Supplementary Information}
@@
     \hline
-    Coverage & Control(bias) & Control(var) & R(bias) & R(var) & RC(bias) & RC(var) & RA(bias) & RA(var)\\
+    Coverage & Control(bias) & Control(var) & R(bias) & R(var) & RA(bias) & RA(var) & RC(bias) & RC(var)\\
     \hline
-    4x & -0.007 & 0.006 & -0.081 & 0.036 & -0.083 & 0.036 & -0.091 & 0.037\\
+    4x & -0.007 & 0.006 & -0.081 & 0.036 & -0.083 & 0.036 & -0.091 & 0.039\\
     \hline
@@
     \centering
-    \caption{Comparison of beta distributions estimated with the HMM (left) without and (right) with variance constraints.}
+    \caption{Comparison of beta distributions estimated with the KIN-HMM (left) without and (right) with variance constrained optimization of $\delta$ parameters.The histogram in each plot shows pairwise differences in genomic windows.
+Colored lines shows the Beta probability distribution for each IBD state.}
     \label{figS4:bndsbeta}
@@
      \centering
-    \caption{Comparison of Beta distributions (left) without and (right) with modified emissions}
+    \caption{Comparison of Beta distributions estimated with ROH-HMM (left) without and (right) with constrained emissions.
+The histogram in each plot shows pairwise differences in genomic windows.
+Colored lines shows the Beta probability distribution for each homozygosity state.}
     \label{figS5:ROHforced}
@@
 
-\section{Notation summary}
-Notation used in the paper:
-
-\begin{itemize}
-\item L: Total number of windows
-\item w: Index of window
-\item $N_w$: Number of sites in w
-\item $D_{w}$: Number of pairwise differences in w
-\item $Z_w$: Hidden IBD state in window w
-\item $i$: States that $Z_w$ can take 
-\item $H_w$: ROH state in window w
-\item $\omega$ States that $H_w$ can take
-\item A: Transition matrix
-\item $\pi$: Vector of initial transition probabilities
-\item $\theta$: Vector of $\pi$, A, $N_w$ for relatedness HMM
-\item $p(Z_w,H_w)$: Expected proportion of differences for given $Z_w$ and $H_w$
-\item $x$: Combinations of $i$ and $\omega$ states with unique $p$
-\item $\delta$: Over-dispersion parameter
-\item $\gamma_{wi}$: posterior probability of $H_w$
-\item $h_{\omega w}$: ROH state-probabilities
-\item $k_{wx}$: Posterior probability of $x$ in $w$
-\item $t$: Threshold on variance of Beta distribution
-\item $X$: Random variable from a beta distribution
-\item $y$: Hidden homozygosity states for ROH HMM
-\item $Y_w$: Homozygosity state in $w$
-\item $\Theta$: Vector of $\pi$, A, $N_w$ for ROH HMM
-\item $\Delta_w$: Number of differences within a library 
-\item $\rho$: Expected proportion of differences in a library
-\item $Gamma$: Posterior probabilities of ROH
-\item $\eta$: Over-dispersion parameter used in ROH HMM
-\item $\mathcal{C}$: Cost function
-\item $C_l$: Contamination in library l
-
-\item $c$: Contamination estimate for a library
-\item $p_c$: Average proportion of differences between endogenous and contaminating reads
-\item $p_e$:Average proportion of differences between endogenous reads
-\item $p_a$: Average observed proportion of differences
-\item $D_{cor,w}$: Contamination corrected number of differences in $w$
-\item $N_{cor,w}$: Contamination corrected number of overlapping sites in $w$
-\item $\Xi$: Endogenous read
-\item $S_{cor,w}$ $N_{cor,w}$ - $D_{cor,w}$
-\item $/mathbf{D_cor}$ Vector of $D_{cor,w}$
-\item $G$: Genotype (0,0.5,1)
-\item $\lambda$: Expected rate following Poisson distribution
-\item $\nu$: Population allele frequency
-\item $Lambda$: Log likelihood ratio
-\item E: Expectation
-\item P: Probability
-\end{itemize}
-
-\section{extra}
-
-log lik:
-\begin{align}\label{eq:1}
-\log P(\BD|\theta) &= \log \sum_{\BZ} [\sum_{\BH} P(\BD|\BZ,\BH, \theta) P(\BH |\theta)] P(\BZ |\theta) \nonumber\\
-&= \log \sum_{\BZ} [\sum_{\BH} \prod_w P(\BD|\BZ,\BH, \theta) P(\BH |\theta)] P(\BZ |\theta)
-\end{align}
-
-Many methods exist to estimate relatedness.
-Most of these work with high coverage, phased data, reference panel, population allele frequency.
-Other methods that work with low coverage ancient data use likelihoods, but still require 2x coverage.
-methods that require a lot of things: PLINK, SNPduo,ERSA,KING,REAP,GRAB,
-
-methods that require low coverage: SEEKIN
-
-mtDNA and Y in relatedness READ paper[15,16,25,16]
-snp data better than str when there is damage, low data
-excluding related individuals, haak indo europian
-
-READ is a very popular software to estimate relatedness in low coverage ancient DNA.
-It calculates distance between a pair of individuals using pseudo haploid sequences.
-These pairwise differences are then normalized by the median of pairwise differences in the population, and compared to the expected pairwise difference given a degree of relatedness.
-Expected pairwise difference for each case of relatedness is calculated using identity by descent (IBD) probabilities.
-
-lcMLKIN is another widely used method that uses genotype likelihoods to estimate the number of positions at which a pair of individuals have 0/1/2 chromosomes in IBD.
-
-REAP is a software to estimate relatedness coefficient and IBD probabilities in admixed populations, accounting for structure.
-It requires the number of ancestral populations, and a representation of sub population allele frequencies.
-
-Likelihood estimators:
-Milligan BG (2003) Maximum-likelihood estimation of relatedness. Genetics
-Anderson AD, Weir BS (2007) A maximum-likelihood method for the estimation of pairwise relatedness in structured populations. Genetics
-Choi Y, Wijsman EM, Weir BS (2009) Case-control association testing in the presence of unknown relationships
-
-There are various ways to measure the coefficient of relatedness r.
-Most methods use the probabilities of IBD to calculate r.
-There can be, in total 9 modes of identity described by Jacquard (1974), including the cases for 3 IBD probabilities and 3 cases of IBD.
-Methods assuming outbred population use only the three modes accounting for IBD.
-Most methods used maximum likelihood approach to estimate r ().
-Other methods are moment estimators that assume Hardy-Weinberg equilibrium in the population (KING, GCTA).
-
-Discussion
-
-Our contamination correction model requires an estimate of divergence between the contaminating and the target populations.
-We provide the users an option to provide a file with average pairwise difference between the two populations, or to provide a vcf file with an individual from each population.
-Pipeline has p(diff) instead of pseudohap
-Methods
-$$\sum_{w} \sum_{I}\sum_\omega \log P(D_{w}|Z_w, H_w)k_{w\omega i}$$
-
-$$P(D_w | N_w, Z_w) = \sum_{k=0}^4 
-\underbrace{P(D_w | N_w, k)}_{Betabinomial}
-\underbrace{P(k | Z_w, H_w)}_{p matrix}
-\underbrace{P(H_w =h)}_{\mathbf{h}_w}
-$$
-
-Emission probability in a window depends on both IBD state and ROH state.
-We obtain posterior probability of one or both individuals being ROH in each window from another HMM described in section 3.2.
-We calculate the cost function with beta-binomial likelihood in log space for the entire genome as:
-
-The model can be summarised as:
-
-\begin{equation}\label{eq:6}
-P(D_w|Z_w,H_w) \sim {\sf Binom}(D_w, N_w, p_{Z,H,w})
-\end{equation}
-
-\begin{equation}\label{eq:7}
-p_{Z,H,w} \sim {\sf Beta}(\alpha_{Z,H}, \beta_{Z,H})
-\end{equation}
-
-Emission probability in a window depends on both IBD state and ROH state.
-We calculate a cost function as the beta-binomial log-likelihood weighted by the posterior probability represented by $\gamma$.
-likelihood:
-\begin{align}
-P(\mathbf{D},\BZ, H|\theta) &= P(\mathbf{D}|\mathbf{Z},H,\theta) P(H |\theta) P(\BZ|\theta)\nonumber\\
-&= \prod_{w}  P(D_w|Z_w,  H_w, \theta) \prod_w P(H_w | \theta) \prod_{w} P(Z_w|Z_{w-1}, \theta)] P(Z_0| \theta) 
-\end{align}
-
-cost function:
-\begin{equation}
-    \mathcal{C} = 
-       \sum_{w} \sum_{i} \log P(D_{w}|Z_w, H_w)\gamma_{wi}=
-    \sum_{w} \sum_{i} 
-    \log \left[\sum_{\omega}P(D_{w}|\psi_{i\omega},\delta_{i\omega}) \right] \gamma_{wi}
-\end{equation}
-
-Supplementary info:
-Pipeline of KIN
-Beta binomial distribution is a better fit compared to Binomial distribution for proportion of differences in windows.
-Results from KIN agree with results from READ, and provide some more information about siblings/parent-child, and $3^{rd}$ Degree relatedness. 
-Constraining variance of beta distributions in case of an identical pair, when identical relatedness hmm is applied, gives more accurate emission probabilities.
-In figS4, the emissions improve in accuracy with constraints on variance of beta distributions.
-figS5 shows why we see this improvement.
-With variance constraints, beta distributions corresponding to Z=1 and Z=2 do not overfit the data, and higher emission probabilities seen in (A) at $D_w/N_w$=0.05 disappear.
-All probability distributions are scaled by the same factor so they are visible with the data histogram.
-Restricting the emission probabilities at 1 for Y=1, in windows that have higher than $p_0$ difference improves the estimated Beta parameters. 
-This figure shows that it is possible to differentiate between Avuncular and Grandparent-Grandchild pairs.
-However, differentiating between Half-siblings may not be possible with our approach. 
-*Add 1M windows result
--READ and KIN differ
-
--supplementary table2
-
 \end{document}

### 2022-04-04 plots/Rscripts/IBD_accuracyPlot.R
added ibd plot

@@ -1,0 +1,61 @@
+library(ggplot2)
+
+#infile='/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/plots/plotdata/plot_IBDaccuracy.csv'
+IBD_accuracy_plot <- function(infile,outfile){
+
+  data=read.csv(file = infile, header = TRUE)
+
+
+  data[data$rel=='deg4','rel']='Unrelated'
+  data[data$rel=='deg5','rel']='Unrelated'
+  data[data$rel=='un','rel']='Unrelated'
+
+  data[data$rel=='avu','rel']='2nd Degree'
+  data[data$rel=='gr','rel']='2nd Degree'
+  data[data$rel=='hsib','rel']='2nd Degree'
+
+  data[data$rel=='id','rel']='Identical'
+  data[data$rel=='pc','rel']='Parent-Child'
+  data[data$rel=='sib','rel']='Siblings'
+
+  data[data$rel=='deg3','rel']='3rd Degree'
+
+  allrel=c("Unrelated", "3rd Degree", "2nd Degree", "Siblings", "Parent-Child", "Identical")
+  allcov=c(4,0.5,0.2,0.1,0.05,0.03)
+
+  accuracy=c()
+  stdev=c()
+  rel=c()
+  cov=c()
+  for(cv in allcov) {
+
+    datacv=data[data$cov==cv,]
+    for(i in allrel) {
+      temp=datacv[datacv$rel==i,]
+      accuracy=c(accuracy,mean(temp$accuracy))
+      stdev=c(stdev,sd(temp$accuracy))
+      rel=c(rel,i)
+      cov=c(cov,cv)
+    }
+  }
+  df=data.frame(accuracy,stdev,rel,cov)
+
+  df$cov=as.factor(df$cov)
+
+  colors1 <- c("Unrelated" = "grey30", "3rd Degree" = "blue", "2nd Degree" = "darkorchid1", "Siblings"= "Green", "Parent-Child" = "orangered3", "Identical" = "darkgoldenrod1")
+
+  f <- ggplot(
+    df,
+    aes(x = cov, y = accuracy, ymin = accuracy-stdev, ymax = accuracy+stdev, colour=rel)
+  )
+  f <- f + geom_errorbar(width = 0.2) +
+    geom_point(size = 1.5)
+  f <- f + labs(y="Accuracy", x="Coverage") +
[…]

### 2022-04-04 plots/plot_script.py

@@ -478,3 +478,55 @@ def comparison_plotf(hmmfold,outfold):
     outff=outfold+'comparison_plot_data.csv.gz'
     with pd.option_context('display.max_rows', len(dff.index), 'display.max_columns', len(dff.columns)):
                     dff.to_csv(outff, sep=',')
+
+
+
+def IBDstates(fold, outf, list_inds, runlist):
+
+    accuracy=[]
+    cov_all=[]
+    run_all=[]
+    pair_all=[]
+    rel_all=[]
+
+
+    for rel in list(list_inds.keys()):
+        pair=list_inds[rel]
+        for run in runlist:
+
+            if rel=='gr':
+                    rel1='grand'
+            else:
+                rel1=rel
+
+            if rel == 'un':
+                tr=np.ones(220)
+            elif rel == 'pc':
+                tr=np.ones(220)*0.75
+            elif rel == 'id':
+                tr=np.ones(220)*0.5
+            elif not rel in ['un','pc','id']:
+                trf=fold+'sanity_check/inbred0/run%s/coverage4/contam0/asc0/inputMode_hapProbs_fil0/%s_state_all.gz' %(run,rel1)
+                tr=np.loadtxt(trf, dtype='float', delimiter = ",")
+
+            for cov in [4,0.5,0.2,0.1,0.05,0.03]:
+                vname=fold+"hmm_numba_viterbi_all_win/contam0/inbred0/run%s/coverage%s/filtered0/asc0/relation_%s_file/res_inphapProbs/pw_%s.csv.gz" %(run,cov,rel,pair)
+                viterbi=np.loadtxt(vname, dtype='float', delimiter = ",")
+
+                ac=sum(viterbi == tr)/len(viterbi)
+                accuracy.append(ac)
+                rel_all.append(rel)
+                cov_all.append(cov)
+                pair_all.append(pair)
+                run_all.append(run)
+
+    df=pd.DataFrame({
+        'pair': pair_all,
+        'rel': rel_all,
+        'run': run_all,
+        'accuracy': accuracy,
+        'cov': cov_all
+        })
+
+    with pd.option_context('display.max_rows', len(df.index), 'display.max_columns', len(df.columns)):
+                df.to_csv(outf, sep=',',index=False)

### 2022-04-04 paper_overleaf/main.tex
Updates from Overleaf

@@
 in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}.
-It is also an essential pre-processing step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.
+It is also an essential  step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.
 
@@
 
-\subsection{Approaches to estimate relatedness from high coverage data}
+\subsection{Approaches to estimate relatedness from high-coverage data}
 
@@
 The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals.
-For example, a pair of siblings are expected to have all three possible IBD states with probabilities (0.25,0.5,0.25) in the absence of inbreeding, as shown in Fig \ref{fig0:schematic}.
+For example, a pair of siblings are expected to have all three possible IBD states with proportions of 0.25,0.5,0.25, respectively (Fig \ref{fig0:schematic}).
 These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}.
@@
 
-However, it is not possible to directly observe IBD segments, and so most methods infer them by first identifying segments of the genome that are Identical by State (IBS) and using population allele frequencies to calculate the probability of IBD given IBS \cite{vai_KINship_2020}.
-One exception is SNPduo, a software that uses IBS directly from a pair of individuals to visualize and analyze their relatedness.
-\cite{roberson_visualization_2009}.
-Most methods, however, use IBD as a hidden state in inferring relatedness \cite{boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}.
+However, it is not possible to directly observe IBD segments, and so a common approach is to first identify segments of the genome that are Identical by State (IBS) and using population allele frequencies obtained from a reference panel to calculate the probability of IBD given IBS \cite{vai_KINship_2020, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}.
+One exception is SNPduo, a software that uses IBS directly from a pair of individuals to visualize and analyze their relatedness \cite{roberson_visualization_2009}.
 
@@
 These methods utilize population allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}.
-There are many such methods available.
 PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each Single Nucleotide Polymorphism (SNP) in the analyzed samples \cite{purcell_plink_2007}.
 KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies in the samples (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}.
-.
 
@@
 In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods.
-However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be unreliable at lower coverages.
-Ancient DNA analyses often face additional challenges such as contamination from present-day DNA \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in the case of data generated from DNA capture.
+However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be imprecise at lower coverages .
+Ancient DNA analyses often face additional challenges such as contamination from present-day DNA \cite{peyregne_authentict_2020}, absence of suitable reference panels, and ascertainment bias in the case of data generated from DNA capture.
+
 In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations.
@@
 It is worth mentioning that different DNA libraries may have different levels of contamination, even when they have been sampled from the same specimen.
-Hence, contamination can reduce the power to detect related pairs, especially when the target population is quite diverged from the contaminating population(s).
-In addition, the genomes of target individuals may have long runs of homozygosity (ROH) due to a small population size, or recent inbreeding.
-Long ROH cause related individuals to seem genetically closer to each other, but do not affect the genetic distance between unrelated individuals.
-This effect may cause an increase in falsely inferred relationships.
+In addition, some analyzed  genomes  may have long runs of homozygosity (ROH), for example due to a small population size, or due to recent inbreeding.
+Long ROH cause related individuals to seem genetically more similar to each other, but do not affect the genetic distance between unrelated individuals.
+
 Moreover, ancient DNA is commonly captured with a SNP array to enrich for sequences from the target species.
-The way that these SNPs are defined (e.g.
-selecting SNPs variable in certain individuals) becomes relevant, because methods to infer relatedness based on the fraction of sites in different IBS states are highly sensitive to ascertainment bias, and may give different results when applied to data with different ascertainment schemes \cite{waples_allele_2019}.
+Particularly methods based  on the fraction of sites in different IBS states are sensiteive to the ascertainment bias caused by this non-random selection of targted sites \cite{waples_allele_2019}.
 
-Several limitations with ancient DNA mentioned above, such as low coverage, the absence of population allele frequencies or a reference panel, and ascertainment bias are resolved by a software called READ.
+Several limitations with ancient DNA mentioned above, such as low coverage, the absence of population allele frequencies or a reference panel, and ascertainment bias are resolved by a software called READ (REF).
 This software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual.
 A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to calculate average pairwise genetic distances, which in turn are used to infer relatedness.
-The limitations are that the software can identify only up to second degree relatives, without differentiating between siblings and parent-child, which may be crucial when making pedigrees.
+However, READ only infers the degree of relatedness, and only up to second degree.
 
 \subsection{How our method works}
-Here, we present KIN (Kinship INference), a Hidden Markov Model (HMM) to estimate relatedness and IBD from low-coverage ancient DNA data.
-KIN can detect up to \nth{3} degree relatives, and differentiates between siblings and parent-child.
-KIN is also able to take into account ROH, contamination and ascertainment bias.
+Here, we present KIN (Kinship INference), a Hidden Markov Model (HMM)-based approach to estimate relatedness and IBD from low-coverage ancient DNA data.
+KIN can detect up to \nth{3}-degree relatives, and differentiates between siblings and parent-child.
+KIN is also able to take into account ROH, contamination and ascertainment.
 We validate the performance of KIN using simulations and show that we are able to infer relatedness in real data from two datasets: a group of Neandertals and a group of Bronze age individuals.
@@
 \subsection{Model description}\label{method_overview}
-To infer the IBD-state across the genome, we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb).
-The proportion of pairwise differences in those windows will differ based on the IBD-state, and the relatedness case will determine how often we expect to switch between different IBD-states.
-The cases we consider are:  Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical individuals.
+
+The goal of KIN-HMM is to infer how two individuals are  relateded via the patterns of shared IBD-states along a pair of genomes.
+For this purpose,  we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb), and infer the pattern of IBD-sharing for each relatedness case we consider (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical).
+We then compare the likelihood between all models, and classify each pair to the model with the highest likelihood.
+We also return the most likely location of IBD tracts (using standard Viterbi algorithm), and the state posterior probablities $\BZ$.
+
 The details of the likelihood computation are given in section \ref{ll}.
-The likelihood computations require an emission (section \ref{B}) and transition probability.
-We estimate the transition matrix for each relatedness case from simulations (see section \ref{simulat}), and hence the transition probability for a HMM is fixed.
-To calculate the emission probabilities, we need to estimate a parameter $\delta$ that describes the variance in the data given the IBD state and the ROH state.
-The estimation of this parameter is essential to our inference of relatedness, since the variance in the data can change with the number of overlapping sites.
-So, our algorithm optimizes $\delta$, which is used in the calculation of emission probabilities, and we use these emission probabilities, along with the transition probabilities, to estimate hidden IBD states and to calculate the log likelihood for the HMM.
-We can then compare the log likelihood from each model to find the best fit along with a measure of uncertainty.
-We output the IBD states corresponding to the most likely model.
-Hidden IBD state of each window ($Z_w$) is estimated using standard Viterbi algorithm, and $\BZ$ corresponding to the most likely model are returned.
+As any HMM, KIN-HMM  require a set of n emission (section \ref{B}) and transition probability matrices.
+We assume the transition matrix for each relatedness case is fixed.
+Our emissions include a vector of parameters ($\delta$) that describes the variance in the data for each IBD state that we infer from the data (section ADD REF).
 
@@
 
-$D_w$ is the number of sites that are different between a pair of individuals, and can be estimated accurately for high coverage data by comparing genotypes.
-For low-coverage samples, $D_w$ is often calculated by comparing randomly-sampled reads ("pseudo-haploid" sequences) \cite{haak_massive_2015}.
+$D_w$ is the number of sites that are different between a pair of individuals.
+For high-coverage data, this can be estimated accurately by comparing genotypes, but for low-coverage samples, $D_w$ needs to be estimated.
+The simplest approach is is to  randomly-sampled a read from each position ("pseudo-haploid" sequences) \cite{haak_massive_2015}.
 However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
 \begin{align}\label{eq:x}
-D_w &= \sum_s \nu_i(s) (1-\nu_j(s)) + (1-\nu_i(s)) \nu_j(s)
+D_w &= \sum_{s=1}^{N_w} \nu_i(s) (1-\nu_j(s)) + (1-\nu_i(s)) \nu_j(s)
 \end{align}
@@
 \subsection{Log likelihood of the KIN-HMM}\label{ll}
-The KIN-HMM uses $\BD$ and $\BN$ to classify each window into three hidden states $Z_w \in$ ($0$, $1$, $2$), reflecting zero, one or two chromosomes IBD.
+The KIN-HMM uses $\BD$ and $\BN$ to classify each window into three hidden states $Z_w \in$ ($0$, $1$, $2$), reflecting zero, one or two shared chromosomes IBD, respectively.
 To take ROH into account, we define the variable $H_w \in$ ($0$, $1$, $2$) that designates that zero, one or both individuals are homozygous in window $w$.
@@
 The initial probabilities $\pi$ give the probabilities of being in each state $Z$ at the beginning of each chromosome, which we set to uniform for simplicity.
-The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed for each relatedness case.
+The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case.
 The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).
@@
 
-We compute the complete data likelihood for the HMM:
+Thus, the complete data likelihood for the HMM is
 
@@
 \log P(\BD,\BZ|\BH, \theta, \delta) &= \log P(\BD|\BZ,\BH, \theta, \delta) + \log P(\BZ |\theta) \nonumber\\
-&= \sum_w \log P(D_w|Z_w,H_w, \theta, \delta) + \sum_w \log P(Z_w |Z_{w-1},\theta) + \log P(Z_0|\theta)
+&= \sum_w \log P(D_w|Z_w,H_w, N_w, \delta) + \sum_w \log P(Z_w |Z_{w-1},A) + \log P(Z_0|\pi)\text{.}
 \end{align}
@@
 
-Using this set up, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta, \delta)$ from equation \ref{eq:1} and optimize them alone.
-The number of shared sites $N_w$ is the only parameter in $\theta$ that affects the emissions.
-
-Assuming the sites were equally distributed and independent would result in the binomial likelihood:
+Using this setup, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta, \delta)$ from equation \ref{eq:1} and optimize them for $\delta$.
+The simplest model is to assume that sites in each window are equally distributed and independent, in this case we could use the  binomial likelihood:
 $$P(D_w|Z_w, H_w, N_w) \sim \text{Binom}[D_w ; p(Z_w, H_w), N_w] \text{,}$$
@@
 If the two individuals are unrelated in a particular window (i.e.
-$Z_w = 0$), then the expected proportion of pairwise differences depends solely on the effective population size, and we denote this proportion with $p_0$.
-Note that $p_0$ does not depend on the presence of ROH, since this does not change expected proportion of differences between unrelated individuals.
+$Z_w = 0$), then the expected proportion of pairwise differences depends solely on the population history, and we denote this proportion with $p_0$.
 If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes \cite{kuhn_estimating_2018}.
@@
 We follow READ \cite{kuhn_estimating_2018} and estimate $p_0$ as the median of differences for all possible pairs of individuals, which should work well if the majority of individuals in the samples are unrelated.
-We set $p_1$ to $(3/4)p_0$ and $p_2$ to $p_0/2$.
+We set $p_1$ to $(3/4)p_0$ and $p_2$ to $p_0/2$, because one, respectively two of the four compared chromosomes are identical in these cases.
 
-The presence of inbreeding adds an additional complication, as it is possible that all chromosomes are shared between individuals, so that the expected differences should be zero.
-An example would be inbred regions in two bones from the same individual.
-However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that setting expected proportion of differences in this case ($p_3$) at $p_2/2$ works well in practice.
+The presence of inbreeding adds an additional complication, as the number of shared chromosome may  change.
+For example, when considering two bones from the same individual, we would expect the entire genome to have a pairwise difference of $p_2$, because two out of the four compared chromosomes are between identical copies.
+However, in inbred regions all four chromosomes will be identical, and so the expected pairwise differences are zero.
+Note however, that $p_0$ does not depend on the presence of ROH, since all comparisons are between unrelated chromosomes even if both individuals are homozygous at a particular locus.
 
@@
 
+As explained above, $p_3 = 0$.
+However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that we obtain better results by setting $p_3 = \frac{p_2}{2}$).
+
 The effect of these considerations is that even though we have nine possible combinations of $Z_w$ and $H_w$ for each window, there are actually only four unique $p$-parameters $p_i$ with $i$ $\in$ (0,1,2,3).
@@
 The Binomial Model laid out above assumes that sites are independent.
-However, due to genetic linkage, nearby sites will be correlated and we find that the data often has considerably higher variance than would be expected from a binomial model.
-We take this into account by adding an overdispersion parameter $\delta$ that allows for variation in the $p$ between windows.
-Note that, just like $p(Z_w,H_w)$ the overdispersion $\delta(Z_w,H_w)$ depends on the number of chromosomes compared, there are only four parameters $\delta_i$ with $i$ $\in$ (0,1,2,3) analogous to the $p_i$.
-Empirically, we find that the beta-binomial likelihood provides a good fit to the data (fig.
-S\ref{figS1:binom})).
+However, due to genetic linkage, nearby sites will be correlated.
+As a result, we empirically find that the data often has considerably higher variance than would be expected from a binomial model (fig.
+S\ref{figS1:binom}).
+We take this into account by adding an overdispersion parameter $\delta$.Just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ depends on the number of chromosomes compared and so each of the  four $p_i$ has a corresponding $\delta_i$ parameter
 
@@
 \begin{align}\label{eq:3}
-P(D_{w}|Z_w,H_w,N_w, \delta) = P(D_{w}|p_i,\delta_i,N_w) &\sim BB[D_w; p_i, \delta_i, N_w] \nonumber\\
+P(D_{w}|Z_w,H_w,N_w, \delta) &\sim BB[D_w; p_i, \delta_i, N_w] \nonumber\\
 &= \binom{N_w}{D_w}\frac{B(D_w+p_i \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})},
@@
 This parameterization of the beta distribution in terms of expected value $p$ and overdispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parametrization in terms of $\alpha$ and $\beta$.
-We use this equation as well if some preprocessing steps result in non-integer $D_w$ and $N_w$, in which case we approximate the binomial coefficient using Gamma functions.
+We use this equation even if preprocessing steps result in non-integer $D_w$ and $N_w$, in which case we approximate the binomial coefficient using Gamma functions.
 
@@
 \paragraph{Initialization}
-The value of $\delta_i$ is unknown to start with, and we set it to a random value.
+The value of $\delta_i$ is unknown to start with, and we set it to a random value between x and y.
 
@@
 \subsection{Contamination correction}\label{contam}
-Contamination by the DNA from present-day people is a common feature in ancient DNA \cite{peyregne_present-day_2020}.
-Even low levels of contamination may make samples look less similar, and thus distort relatedness analyses.
-This issue is particularly pronounced  when analyzing populations with low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations (Skov et al.,2022 (in press)).
+Contamination by the DNA from present-day people is a common feature in human ancient DNA \cite{peyregne_present-day_2020}.
+Even levels of contamination may make samples look less similar, and thus distort relatedness analyses.
+This issue is particularly pronounced  when analyzing populations with low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations .
 To address this issue, we develop a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.
 
-We conceptualize contamination in the following way.
-If data from individual $i$ has a known contamination rate of $C_i$, this means that any particular read drawn from this individual will actually come from that individual with probability $1-C_i$.
+We conceptualize contamination in the following way: If data from individual $i$ has a known contamination rate of $C_i$, this means that any particular read  will actually come from the desired individual with probability $1-C_i$.
 With probability $C_i$ the read comes from a contaminant.
@@
 Finally, with probability $C_iC_j$, both reads are contaminant.
-If we make the simplifying assumption that both $C_i$ and $C_j$ are small, the terms of the order of $C^2$ can be ignored.
+If we make the simplifying assumption that both $C_i$ and $C_j$ are small, the $C_iC_j$ terms  can be ignored.
 Then, we compare contaminant read to endogenous read with the probability $C_{ij} = C_i + C_j$.
 
-Given that we know the level of contamination (estimated from other methods),our goal then, is to estimate the expected number of differences from comparison of endogenous reads $\mathbb{E}[D_w^{'}]$, and the total number of sites with overlapping endogenous reads $\mathbb{E}[N_w^{'}]$.
+We assume that the level of contamination is known (cite Stephane and Kay's review).
+We  estimate the expected number of differences from comparison of endogenous reads $\mathbb{E}[D_w^{'}]$, and the total number of sites with overlapping endogenous reads $\mathbb{E}[N_w^{'}]$.
 To estimate $\mathbb{E}[D_w^{'}]$, we first need to calculate $P(c=0|d=1)$.
@@
 
+%adding back the previous versions:
+If we look at one site in the window, we could find either no difference in reads from individuals $i$ and $j$ ($d_s$=0), or a non-zero difference ($0<d_s<=1$).
+Here $d_s$ is the probability of seeing a difference at site $s$, and $\sum_{s \in w} d_s =D_w$.
+We ask, what is the
+
+If we make the simplifying assumption that both $C_i$ and $C_j$ are small, terms of the order of $C^2$ can be ignored and there are just two cases: with probability $C_{i,j} =C_i + C_j$, one of the reads stems from a contaminant, and we effectively compare a contaminant fragment with an endogenous one.
+The probability that they differ is the average divergence between target and contaminating population, which we denote with $\phi$.
+The other case (w.
+p.
+$(1-C_{i,j}$) is where we compare two endogenous reads, which will differ with probability
+
+\begin{equation}
+    p_{e} = \frac{p_{obs}-\phi C_{i,j}}{1-C_{i,j}}
+\end{equation}
+
+Here, $$p_{obs} = \frac{\sum_w D_w}{\sum_w N_w}$$
+
+We estimate $\hat{D_c} = C_{i,j} N_{tot} \phi$ and $\hat{D_e} = (1-C_{i,j}) N_{tot} p_{obs}\phi$.
+If there are $N_{obs}$ overlapping sites in a window, $N_{c} = C_{i,j}N_{obs}$ of them will include a contaminant allele, and $N_e = (1-C_{i,j})N_{obs}$ will be between endogenous alleles.
+
+For KIN, we can use $N_e$ as an input (for each window), but we also need the proportion of differences, $D_e$.
+For this, we use the assumption that the proportion of sites where the contaminant differs from our samples is
+
+$$\phi = \frac{D_c}{N_c} = \frac{D_c}{C_{i,j} N_{obs}}, $$
+and thus $D_c = C_{i,j} \phi N_{obs}$.
+
+We can think of $p_c$ and $\rho$ as empirical likelihoods that a particular site shows a difference between two reads, given that the comparisons contain only endogenous or one contaminant read, respectively.
+\begin{align}
+    P(E|D) &= \frac{P(D|E)P(E)}{P(D)} \nonumber\\
+    \mathbb{E}[D_e] &= D_{obs} P(E|D) = D_{obs}\frac{p_e (1-c)}{p_e(1-c) + c\rho}\\
+    \mathbb{E}[S_e] &= D_{obs} P(E|\text{not} D) = S_{obs}\frac{(1-p_e)(1-c)}{(1-p_e)(1-c) + c(1-\rho)}
+\end{align}
+
+And for calculating $N_{e}$, we added $D_{e}$ to $S_{e}$
+
+The goal is that given $C_{i,j}, p_e, \BN and \BD$ to calculate adjusted $\hat{\BN}$ and $\hat{\BD}$ that reflect the endogenous proportion only.
+
+, and the observed proportion of differences between individuals $i$ and $j$ as $p_{oij}$.
+We calculate average endogenous pairwise difference $p_{eij}$ for individuals i and j as follows \cite{noauthor_ancient_nodate}:
+\begin{align}
+    p_{oij} = C_{ij}  p_c + (1-C_{ij})  p_{eij}\\
+    or, p_{eij} = \frac{p_{oij} - C_{ij}  p_c}{1-C_{ij}} 
+\end{align}
+
+Given $p_{eij}$, we would ideally like to calculate the complete data log likelihood for the relatedness model while summing over all possible differences we could have in a window weighted by the probability of seeing that difference.
+
+\paragraph{Short version}
+With contamination, we do not observe the number of differences $D_{true}$ and total number of comparisons $N_{true}$ directly, but instead need to approximate them.
+The window-subscript is omitted in the following for clarity:
+\begin{equation}
+    P(D_{\text{obs}} | N_\text{obs}, Z, \theta) \approx 
+    P( \hat{D} | \hat{N}, Z, \theta)
+\end{equation}
+We do that by calculating the probability that a single read is endogenous given it has a difference.
+\begin{align}
+    \hat{D} = \mathbb{E}[D | D_{\text{obs}}, c] &= 
+    D_{\text{obs}} P(E_i | D_i=1) \\
+    &= D_\text{obs} \frac{P(D_i=1 | E_i)P(E_i)}{P(D_i=1)}\\
+    &= D_\text{obs} \frac{(1-c)p_e}{(1-c)p_e + c p_c}
+\end{align}
+
+and likewise for reads that do not have a difference
+\begin{align}
+    \hat{S} = \mathbb{E}[S | S_{\text{obs}}, c] &= 
+     S_\text{obs} \frac{(1-c)(1-p_e)}{(1-c)(1-p_e) + c (1-p_c)}
+\end{align}
+
+and $\hat{N} = \hat{S} + \hat{D}$
+
+\begin{equation}
+    P(D_{\text{obs}} | Z, N_\text{obs}, \theta) = \sum_n\sum_d P(D_{\text{obs}}, N_\text{obs} | D_w=d, N_w, C_i, C_j)\underbrace{P(D_w=d | N_w, Z, \theta)}_\text{old likelihood}
+\end{equation}
+
+i.e. we sum over all possible $D_w$, $N_w$. Instead, we use
+
+\begin{align}
+    P(D_{cor},Z|D,H,\theta,c,p_c) &= P(D_{cor}|Z,D,H,\theta,c,p_c) P(Z|D,H,\theta,c,p_c)\nonumber\\
+    &= [\prod_{w} [\sum_\kappa P(D_{cor,w}|Z_w, H_w, \theta,c,p_c) P(D_{cor,w}=\kappa)] \prod_{w} P(Z_w|Z_{w-1}, \theta)] P(Z_0| \theta)
+\end{align}
+
+We realize that this calculation would be time consuming, and would further complicate the HMM.
+To avoid this, instead of using weighted sum of all possible $\kappa$, we use $E(D_{cor,w}|Z_w, H_w, \theta,c,p_c)$.
+This simplifies the equation 21 to :
+
+\begin{align}
+    P(D_{cor},Z|D,H,\theta,c,p_c) &= P(D_{cor}|Z,D,H,\theta,c,p_c) P(Z|D,H,\theta,c,p_c)\nonumber\\
+    &= [\prod_{w} P(E(D_{cor,w})|Z_w, H_w, \theta,c,p_c) \prod_{w} P(Z_w|Z_{w-1}, \theta)] P(Z_0| \theta)
+\end{align}
+
+We calculate $E(D_{cor,w})$ as shown:
+
+\begin{align}
+    E[N_e] + E[N_c] = N\\
+    E[D_e] + E[D_c] = D
+\end{align}
+
+Probability of comparing endogenous reads at a site where we see a difference:
+$$P(\xi | N = 1,D = 1)=\frac{(1-c) p_e}{c p_c + (1-c) p_e} $$
+Probability of comparing endogenous reads at a site where we see no difference:
+$$P(\xi | N = 1,D = 0)=\frac{(1-c)(1-p_e)}{c (1-p_c) + (1-c) (1-p_e)} $$
+Expectation of number of endogenous comparisons in a window:
+\begin{align}
+    E[\xi | N, D] = D E[\xi | N=1, D=1] + (N-D) E[\xi | N=1, D=0]\\
+    = D P(\xi | N=1, D=1) + (N-D) P(\xi | N=1, D=0)
+\end{align}
+
+Total number of endogenous sites showing a difference in a window: 
+$$D_{cor,w} = D_w* \frac{p_e(1-c)}{p_e(1-c) + c p_c} $$
+Total number of endogenous sites showing 0 difference in a window: 
+$$S_{cor,w} = S_w* \frac{(1-p_e)(1-c)}{(1-p_e)(1-c) + c (1-p_c)} $$
+Total number of endogenous sites in a window:
+$$N_{cor,w} = D_{cor,w} + S_{cor,w}$$
+
+This model outputs the contamination-corrected number of differences and overlapping sites in each window.
+
 We added $D_w^{'}$ to $S_w^{'}$ to get the corrected number of overlapping sites $N_w^{'}$
@@
 
+We predict different cases of relatedness with KIN, and test it's performance on simulated pedigrees.
+We use these simulations to find the limits of KIN, and compare it to READ and lcmlkin CITE, two existing methods.
+Finally we apply KIN to published data and show that it can give reliable information about genetic kinship.
+
 \paragraph{Motivating Example}
-We predict different cases of relatedness with KIN, and test it's performance on simulated pedigrees.
-We use these simulations to find the limits of KIN as well as ROH HMM.
-We introduce cases of contamination, ascertainment, ROH, and low coverages to compare our method to READ.
-Finally we apply KIN to published data and show that it can give reliable information about genetic kinship, that may not be available from other available methods.
-Here, in Fig.\ref{fig1:ibd} we show the output of different Relatedness HMMs when they are applied to a pair of siblings.
+In Fig.\ref{fig1:ibd} we show the output of different relatedness cases when they are applied to data from a pair of siblings.
 
@@
 
-We see that in this case, the siblings model predictions match true IBD states the most, as reflected by the highest log-likelihood, and the close match between the estimated and true IBD states.
-
-\paragraph{IBD prediction}
-We compared the IBD assignment for simulated pairs of individuals with different relations.
+In this case, the siblings-model predictions match true IBD states the most, as reflected by the highest log-likelihood.
 
 \paragraph{Critical Values}
- To investigate the limits of our method, we plotted the true positive and false positive rates with a cutoff on the log likelihood ratio in Fig.\ref{fig3:cutoff}.
-Here, and in Fig.\ref{fig4:Comparison_READ_KIN}, we have two relatedness labels: 'Unrelated' and 'Unrelated w/o 3rd Degree'.
-Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
-This is useful for comparison to READ, since READ classifies relatedness up to Second Degree.
-'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.
-Fig.\ref{fig3:cutoff} shows that for all relatedness cases except for $3^{rd}$ Degree, the false positive rate is below $5\%$ even when no cut off is used.
-We observe that for $3^{rd}$ Degree relation, using a cutoff of 1 brings down the false positive rate close to $5\%$ for all coverages except 0.05x.
-In conclusion, it is recommended to use a cutoff of 1 for most cases.
+ To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff( Fig.\ref{fig3:cutoff},  Fig.\ref{fig4:Comparison_READ_KIN}).
+Fig.\ref{fig3:cutoff} shows that for all relatedness cases except for $3^{rd}$-degree, the false positive rate is below $5\%$ even when simply selecting the model with the highest likelihood.
+We observe that for $3^{rd}$-degree relatives, using a cutoff of 1.0 brings down the false positive rate close to $5\%$ for all coverages except 0.05x.
+Thus, we  recommended to use a cutoff of 1 for all cases where ROH or contamination are not a concern.
 
@@
     \centering
-    \caption{False positive and True Positive rates as a function of cutoff on log likelihood ratio. In top panel, dotted line represents $5\%$}.
+    \caption{False positive and True Positive rates as a function of cutoff on log likelihood ratio.
+In top panel, dotted line represents $5\%$.
+Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
+'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
     \label{fig3:cutoff}
@@
 
-In our assessment of critical value for log likelihood ratio, we used the control simulations.
-However, we test performance of KIN on different simulation cases including case R, which is also based on performance of ROH HMM.
-
 \paragraph{Evaluating the ROH detection}
- Fig.\ref{fig2:ROH} shows the performance of ROH-HMM under cases of control, R, A, C and simulations (described in section \ref{simulat}), and for different coverages.
+We next evaluate the performance of the ROH-HMM( Fig.\ref{fig2:ROH}).
 We observe that as coverage decreases, the raw heterozygosity estimates become more noisy.
-In the control case (Fig \ref{fig2:ROH}, we correctly infer that there is no ROH for most of the time, but occasionally we erroneously call some ROH as shown in the figure.
-We observe that the model predicts ROH probabilities with biases of upto -0.007, -0.081, -0.083, and -0.091 and variances upto 0.006, 0.036, 0.036, and 0.039 in control, R, A, and C cases respectively as shown in supplementary table \ref{tab:Table 2}.
-The negative bias in all cases shows that on an average, the model over-estimates probability of ROH, and this effect is seen slightly more often at higer coverages.
-We do not see any large adverse effects of ascertainment bias and contamination on the performance of ROH HMM.
+In the case without ROH (Fig \ref{fig2:ROH}A), we correctly infer that there is no ROH for all but two windows.
+We observe that the model predicts ROH probabilities well in all cases, even at a coverage of 0.1 \ref{tab:Table 2}.
 
@@
     \centering
-    \caption{Estimation of ROH probabilities along the genome in simulations with different cases of ROH, contamination, and ascertainment.
+    \caption{Estimation of ROH probabilities along the genome in simulations.
 Top panel shows proportion of differences in a simulated individual in windows along the genome, and the bottom panel shows probability of seeing no ROH.
@@
 \end{figure}
-We use a cutoff of 1 to compare our method performance to READ, while using a cutoff of 1 standard deviation for READ.
-Fig.\ref{fig4:Comparison_READ_KIN} shows this comparison using false positive and true positive rates for both the methods at different coverages under different simulations.
-We show that both methods have similar performance in the control case, and cases with ROH and ascertainment, except for $2^{nd}$ Degree relation, where KIN has higher power, especially at low coverages.
+
+\paragraph{Comparing KIN-HMM with READ}
+In Figure Fig.
+\ref{fig4:Comparison_READ_KIN}, we compare the classification accuracy of KIN (cutoff: 1 log-likelihood unit) to that of READ (cutoff: 1 standard deviation)  under four different simulation scenarios.
+We show that both methods have similar performance for low-coverage shotgun data (``control''-case),  except for $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages.
 We show that adding contamination adversely affects both the False Positve and True Positive rates corresponding to 1st and 2nd Degree relations for READ, while this has no effect on KIN performance.
+Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
+'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.
 
@@
     \centering
-    \caption{Comparison of KIN with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH}
+    \caption{Comparison of KIN with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH.
+Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
+'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
     \label{fig4:Comparison_READ_KIN}
@@
 
-To test KIN on real ancient dataset, we applied it to Neandertal specimens from Chagyrskaya cave in Siberia, Russia.
-This dataset comprised of 14 specimens that came from neighbouring layers, and hence could possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya \cite{kolobova_archaeological_2020-1}.
-These specimens showed low coverage of nuclear genome (ranging from 0.01x to 12.34x), with 8 samples out of 14 showing $<1x$ coverage.
-Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena.
-All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes and common variations in Africans (Laurits et al.,2022), and we further ascertained this data to variable sites in high coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}.
-This dataset had all the issues that KIN deals with, and hence we decided to apply it here to see if we could find family relationships that would not be possible with other available methods.
+\paragraph{Evaluating IBD prediction}
+KIN-HMM co-estimates IBD states and relatedness, and hence it has a higher power to classify $2^{nd}$ Degree relatives, and can distinguish between Siblings and Parent-Child.
+We investigate our limit to predict IBD states along the genome by calculating the number of genomic windows where we correctly  predict IBD state for a pair of individuals.
+We calculate mean and standard deviation for this accuracy over 60 simulations of a pedigree (see section \ref{simulat}).
+Fig.
+shows that in the control case we can predict IBD states very well at high coverages, and our accuracy decreases with downsampling.
+
+\begin{figure}[h!]
+    \includegraphics[width=16cm]{plots/plotimg/}
+    \centering
+    \caption{Comparison of KIN with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH.
+Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
+'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
+    \label{fig4:Comparison_READ_KIN}
+\end{figure}
+
+\subsubsection{Application to Chagyrskaya Neanderthals}
+To test KIN on real ancient dataset, we applied it to a Neandertal data set  from Chagyrskaya cave in Siberia \cite{kolobova_archaeological_2020-1}, Russia.
+This dataset contains genetic data from  14 fossils that might  possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya .
+The data has low-to intermediate coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage.
+Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena (again cite Laurits paper).
+All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes and common variations in Africans (Laurits et al.,2022), and we further restricted  this data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}.
+is dataset has all the issues that KIN deals with, and hence we decided to apply it here to see if we could find family relationships that would not be possible with other available methods.
 Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}.
-We found Identical individuals (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), $1^{st}$ Degree relative (Chagyrskaya07-Chagyrskaya17), and a $2^{nd}$ Degree relative (Chagyrskaya01-Chagyrskaya60).
-These results agree with READ (fig.
-S\ref{figS2:Chagyrskaya_READ}).
-Both the methods classified Chagyrskaya06-Chagyrskaya14 to be $1^{st}$ Degree relatives with low confidence (table S1), but have been classified as Identical pair in the original study, since Chagyrskaya06 is a tooth sample that fits in mandible which is Chagyrskaya14.
+We found three fossils from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair  (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60).
+We also find that Chagyrskaya 06 and Chagyrskaya 14 are classified as parent-child with low confidence, but from the morphology they likely stem from the same individual.
 We believe that the contradiction is due to non human contamination present in these libraries ($1\%$ and $2.9\%$ respectively).
-Interestingly, for the pair of Chagyrskaya07-Chagyrskaya17, READ shows that they are first degree relatives, and KIN further identifies them as parent-child.
-Our result matches with the conclusions of the authors based on additional information from mtDNA.
-Further, we find a possible $3^{rd}$ degree relative (Chagyrskaya07-Chagyrskaya60).
+Further, we identify   Chagyrskaya07 and Chagyrskaya60 as 3rd-degree relatives.
+
+Our estimates are consistent with those obtained using READ, except that READ is unable to detect 3rd-degree relationships, and cannot distinguish between siblings and parent-child.
+
 We compared our IBD estimates to lcMLKIN for all pairs for whom READ results (with s.d.
->1) and KIN results ($\Lambda$>1) match (fig.
+>1) and KIN results ($\Delta LL$>1) match (fig.
 S\ref{figS6:Chagyrskaya_ibd}).
-Chagyrskaya06-Chagyrskaya14 is removed from the analysis.
 We find that lcMLKIN does create four different clusters for different relatedness.
-However, the clusters are quite deviated from the expected values, which makes it difficult to identify the relatedness.
-Identical individuals, for example, are expected to be at $k_0 = 0$ and $r = 1$, and Parent-Child at $k_0 = 0$ and $r = 0.5$.
-These deviations can be due to ascertainment, contamination, ROH as well as low coverage of samples.
+However, the clusters are strongly deviated from the expected values, likely due to the ROH.
+Fossils from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $r \approx 0.6, k_0 \approx 0.33$.
 
@@
 We compared our relatedness estimates to those obtained from the authors using READ and lcMLKIN, and found that in most cases the three methods agree.
-Out of 6903 pairs, 346 pairs show different relatedness for KIN and READ (table S2).
-However, there are only 31 pairs where KIN has $\Lambda >1$.
-We looked in detail at these pairs, and we found that it is possible to deduce the correct relatedness for several of these pairs by plotting the proportion of differences along the genome.
-In all the pairs that we analyzed in this way, KIN seems to predict the true relatedness.
-We show few examples of such cases in fig.
-S\ref{figS8:eg1}.
-We found 20 pairs in this subset of 346 pairs where lcMLKIN has sufficient data (Overlapping SNPs $>$ 10000).
-Among these 20 pairs, we found that lcMLKIN matches KIN estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLKIN predicts 3rd-5th degree, KIN predicts second degree) and pair AITI72-AITI77A (READ and lcMLKIN predict unrelated, while KIN estimates 3rd degree with log likelihood=1.7).
-We found 43 pairs for which READ estimates first degree, and compared the estimates of lcMLKIN and KIN for these cases (suppl table3).
+Out of 6903 pairwise comparisons, 346 pairs show different relatedness for KIN and READ (table S2).
+However, most of these are at very low confidence, and there are only 31 pairs where KIN has $\Lambda >1$.
+When we investigated some of these comparisons in more detail we found that in many cases, the proportion of differences along the genome align with the prediction by KIN (fig.
+S\ref{figS8:eg1}).
+Only 20 pairs in this subset of 346 pairs had sufficient data for analysis using lcMLKIN (more than 10,000 overlapping SNPs).
+Among these 20 pairs, we found that lcMLKIN matches KIN's estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLKIN predicts 3rd-5th degree, KIN predicts second degree) and pair AITI72-AITI77A (READ and lcMLKIN predict unrelated, while KIN estimates 3rd degree with log likelihood=1.7).
+We further found 43 pairs for which READ estimates first degree, and compared the estimates of lcMLKIN and KIN for these cases (suppl table3).
 There were a total of 20 pairs where lcMLKIN had sufficient data, and 26 pairs where $loglikelihood ratio>2$ for KIN.
-Among the 19 overlapping pairs, we found four pairs where KIN and lcMLKIN differ, and a visual inspection shows that in these cases, KIN seems to have the correct estimate (fig.
+Among the 19 overlapping pairs, we found four pairs where KIN and lcMLKIN differ, and again the heterozygosities along the genome match the inference of KIN (fig.
 S\ref{figS9:eg2}).
@@
 
-Several methods are available to estimate relatedness from modern as well as ancient DNA data.
-We developed a new method to estimate genetic KINship from low coverage data.
-We show that it is possible to accurately infer relatedness, along with IBD tracts, from such data in presence of long ROH, contamination, as well as ascertainment bias.
-Our method utilizes Hidden Markov Models to estimate IBD tracts, conditional on a particular relatedness, and we use log likelihoods of these HMMs to find the best fit.
-This approach of using Hidden Markov Model makes use of the fact that DNA is inherited in blocks, and hence the IBD state in a given window is dependent on the IBD state in the previous window.
-We use this approach to identify the following pairwise relations: Identical individuals, Parent-Child, Siblings, Second degree, Third degree, Unrelated individuals.
-We have developed a similar Hidden Markov Model to infer long ROH, and we use the posterior ROH probabilities to make better relatedness inferences.
-We further use a simple model to correct the pairwise differences in genomic windows in the presence of contamination.
+Here, we present a new method called KIN to estimate genetic kinship and the location IBD tracts from low coverage data, taking into account long ROH, contamination, as well as ascertainment bias.
+Our method utilizes a set of HMMs to estimate IBD tracts, conditional on a particular relatedness, and we use the log likelihoods of these HMMs to find the best fit.
 
@@
 We demonstrate that at low coverage (0.1x) we have higher power to detect first degree relatives compared to READ with similar false positives, while a higher power to identify second degree relatives at the cost of slightly higher false positive rate.
+
 Counter-intuitively, we see that adding long ROH ($~17\%$) to simulations improves the performance of both KIN and READ.
-READ's performance improves because ROH does not cause any change in pairwise differences between unrelated individuals, while decreasing pairwise differences between related individuals.
+READ's performance improves because ROH does not cause any change in the average pairwise differences between unrelated individuals, while decreasing pairwise differences between related individuals.
 The higher the relatedness, the higher is the decrease in pairwise differences in presence of ROH.
 And so, ROH causes higher power for READ which otherwise is a conservative method.
-KIN performs well for the same reasons, but can additionally differentiate between different combinations of IBD and ROH in genomic windows.
+KIN performs well for the same reasons, but distinguishing IBD and ROH becomes important for distinguishing different relationship cases within a degree, i.e.
+when distinguishing between parent-child and siblings.
+
 Contamination, on the other hand, can affect the performance of READ adversely depending on the amount of exogenous DNA, and the number of individuals contaminated.
-We show that even at contamination levels  $<=3\%$, and contamination in only 8 out of 17 individuals, READ performance falls.
-Contamination in more individuals causes a shift in median difference in the population and may cause non contaminated individuals to look more related, thus increasing the false positive rates.
-On the other hand, an increase in the amount of contamination in a few individuals can make them look more unrelated.
-KIN, with the implementation of a model to correct for contamination, performs well in this case.
-Ascertainment causes a decrease in power for both the methods, since it reduces the amount of data.
-However, the decrease in power is more for READ compared to KIN.
+We show that even at contamination levels  $\leq 3\%$, and contamination in only 8 out of 17 individuals, READ performance falls.
+Contamination in more individuals causes a shift in median difference in the population and may cause non-contaminated individuals to look more related, thus increasing the false positive rates.
+On the other hand, an increase in the amount of contamination in a few individuals can make them look less related to uncontaminated individuals.
+Assuming the contamination rates are correctly inferred, the correction we implemented leads to improved performance compare to naive methods such as READ, although they too could be ameliorated in a similar way (cite Laurits).
+SNP ascertainment causes a decrease in power for both  methods, since it reduces the amount of data available.
+However, the decrease in power is lower for KIN compared to READ.
 
-DNA data from Bronze age, Lech valley has low contamination, and no ROH.
-We demonstrate that KIN matches the results from READ and lcMLKIN for pairs with good amount of overlaps, while adds more information for low coverage samples.
-As expected from the simulations, KIN is helpful in differentiating between parent-child from siblings, and identifying second and third degree individuals in cases where there is an overlap of only few thousand sites or less, and the other methods are not informative.
-We show that when applied to Neandertal specimens from Chagyrskaya cave, READ and KIN agree in relatedness estimates for all pairs where both methods have sufficient data.
-KIN additionally identifies a pair of first degree relatives as parent-child, which is in agreement with mtDNA analysis from the authors (Laurits et al., 2022 (in press)).
-In addition, KIN identifies a third degree relative.
+The Lech valley data  has low contamination, and no ROH.
+For pairwise comparisons with large numbers of overlapping sites (>XXXX), KIN, READ and lcMLKIN all mostly agree.
+However, KIN is able to differentiate between parent-child and siblings, and identify second and third degree relationships from just a few thousand sites overlapping between samples.
+We show that when applied to Neandertal specimens from Chagyrskaya cave, KIN identifies a pair of first-degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples (Laurits et al., 2022 (in press)).
+In addition, KIN identifies a pair of third degree relatives.
 We compare IBD estimates from KIN to lcMLKIN and show that KIN's IBD estimates are much closer to the expected estimates, given the relatedness between individuals that has been estimated with both READ and KIN.
+
 The output of KIN is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio.
@@
 One limitation of our approach is that it assumes a single population.
-In case of population structure, where the sub-populations have very different diversity, KIN may show inaccurate inferences.
-Our method makes another assumption, that the dataset comprises of at least some unrelated individuals.
-We estimate $p_0$ as the median of all pairwise differences, and if more than half of the pairs are related to each other, $p_0$ estimate would be biased, and may affect the relatedness inference.
-In this study, we identify six different cases of relatedness, but in theory our approach can be extended to identify many more such cases, such as double first cousins, using a corresponding IBD state transition matrix.
+In case of a highly structured population, KIN may show inaccurate inferences.
+Also, our method makes the assumptions, that the median pairwise genetic difference reflects the population diversity, which fails if almost all individuals in the data set are related.
+The current implementation of KIN is restricted to five relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.
 
@@
 \subsection{Simulations}\label{simulat}
-
-We use simulations both for estimating the transition matrices for all relatedness cases, and for testing and validating our algorithm.
+We use simulations both for estimating the transition matrices and for testing and validating our algorithm.
 All simulations are performed by simulating unrelated individuals using $\texttt{msprime}$ \cite{kelleher_efficient_2016}, followed by an additional step where we simulate related individuals using a predetermined pedigree (Figure S\ref{figS10:pedigree}).
@@
 
-For our simulations of background diversity, we form a population (Pop1) with constant effective size of 10,000 and sample eight diploid individuals (each made up of two haploid individuals) from 2500 generations ago (fig.
+For our simulations of background diversity, we form a population (Pop1) with constant effective size of 10,000 and sample eight diploid individuals (each made up of two haploid individuals) (fig.
 \ref{figS10:pedigree}).
@@
 KIN requires a transition matrix for each relatedness case.
-For the simplest cases like grandparent-grandchild and siblings, it is possible to calculate the transition matrix from first principles (as shown in the section \ref{Aver}), but for most cases this is difficult  because the relatedness cases include multiple cases, or they are non-Markovian, and so for these cases we obtain transition matrices from our simulations.
-
-For each relatedness case, we estimate the transition matrix  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of XX simulations.
+which we estimate  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of XX simulations.
 
 \paragraph{Verifying accuracy of transition matrices}\label{A_ver}
-We verify that are estimation of transition matrices for grandparent-grandchild ($A_g$) and siblings ($A_s$) are very close to the respective transition matrices obtained from theoretical calculations.
+For the simplest cases like grandparent-grandchild and siblings, it is possible to calculate the transition matrix from first principles, which allows validation of our simulations on the caes of grandparent-grandchild ($A_g$) and siblings ($A_s$) are very close to the respective transition matrices obtained from theoretical calculations.
 To calculate $A_g$, we count the number of IBD state change events that happen along chromosomes from grandparent-grandchild pair given a recombination event happens with rate $r$.
@@
 We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from Pop1.
-We teste out the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a Markov chain using transition matrix shown below:
+This scenario roughly models the ascertainment of the Chagyrskaya  data.
+We tested the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a Markov chain using transition matrix shown below:
 
@@
 From the steps described above, we got genotypes of 17 individuals in Pop1 in presence/absence of ROH and ascertainment.
-We further simulated 10 haploids (5 diploid individuals) from Pop4 with split time of 20,000 generations with pop1, sampled from the present time.
+We further simulated five diploid individuals from Pop4 with split time of 20,000 generations with pop1, sampled from the present time.
 We calculated the allele frequency for Pop4 at each site, and used it to introduce contamination (with fixed contamination percentage for each individual ($<5\%$) in samples from Pop1.
-We generated reads (derived/ancestral) for different genomic coverages ranging from 4x to 0.03x at each position for each individual.
-We specifically generated reads for four different cases: Ancestral-Endogenous, Ancestral-Contaminant, Derived-Endogenous, Derived-Contaminant.
-The number of reads for each case was sampled from a Poisson distribution with $\lambda$ calculated from the coverage $\zeta$, allele frequency or genotype (G $\in$ 0,0.5,1), and contamination estimate:
-
-\begin{align}
-    \lambda_{DE} = \zeta (1-c) G\nonumber\\
-    \lambda_{AE} = \zeta (1-c) (1-G)\nonumber\\
-    \lambda_{DC} = \zeta c \upsilon\nonumber\\
-    \lambda_{AC} = \zeta c (1-\upsilon)
-\end{align}
+We generated reads (derived/ancestral) for different genomic coverages ranging from 4x to 0.03x, assuming a Poisson distribution.
 
@@
     \centering
-    \caption{Overall schematic of the method. Blue boxes show the data files, while green boxes represent scripts.}
+    \caption{Overall schematic of the method showing the entire pipeline from processed bam files to final relatedness and IBD estimates .
+Blue boxes show the data files, while green boxes represent scripts.}
     \label{figS0:schematic}
@@
 Top panel shows pairwise proportion of differences is close to expectation, but dips in some windows due to presence of ROH tracts.
-Bottom two panels show $P(Data|Z=2)$ without constraints and with constraints respectively.}
+Bottom two panels show $P(Data|Z_w=2)$ without constraints and with constraints respectively.}
     \label{figS3:bnds}
@@
     \caption{Comparison of beta distributions estimated with the KIN-HMM (left) without and (right) with variance constrained optimization of $\delta$ parameters.The histogram in each plot shows pairwise differences in genomic windows.
-Colored lines shows the Beta probability distribution for each IBD state.}
+Colored lines shows the Beta probability distribution for each IBD state $Z_w$.}
     \label{figS4:bndsbeta}
@@
 The histogram in each plot shows pairwise differences in genomic windows.
-Colored lines shows the Beta probability distribution for each homozygosity state.}
+Colored lines shows the Beta probability distribution for each homozygosity state $Y_w$.}
     \label{figS5:ROHforced}
@@
     \caption{Comparison of IBD states estimated for Chagyrskaya specimens using (A) lcMLKIN and (B) KIN.
-    The relatedness shown with different colors is estimated with and matched with READ and KIN.}
+Each plot shows relatedness coefficient (calculated as $k_2+k_1/2$) on y-axis plotted against $k_0$ (proportion of genome with no chromosomes shared).
+    The relatedness shown with different colors is estimated with both READ and KIN.}
     \label{figS6:Chagyrskaya_ibd}
@@
     \includegraphics[width=18cm]{supplementary_info/plots/degree2_10Mwin.png}
-    \caption{Avuncular and Grandparent-grandchild form different clusters, when total number of IBD fragments are plotted against total length of IBD fragments.}
+    \caption{Ploting total number of IBD fragments with the number of IBD fragments can tease apart avuncular relationship from grandparent-grandchild, since they form different clusters.
+Half-siblings, however, overlap considerably with both these relations.}
     \label{figS7:second_degree}

### 2022-04-04 plots/Rscripts/comparisonPlot.R
fixed label in comparison plot

@@ -49,7 +49,7 @@ comparison_plot <- function(inf,outf){
   plot<- plot + facet_grid(T_F ~ Relatedness, labeller = labeller(T_F = T_F.labs, Relatedness = Relatedness.labs))
   #plot + geom_hline(yintercept=0.05, linetype="dashed")
   plot <- plot + geom_hline(data = data.frame(T_F='False_positive'), aes(yintercept = 0.05), linetype = "dotted")
-  plot <- plot+ labs(y="Classification Proportions", x = "Log likelihood Ratio Cutoff")
+  plot <- plot+ labs(y="Classification Proportions", x = "Coverage")
   plot + theme_bw() + theme(
     strip.background = element_rect(
       color="transparent", fill="transparent", size=1.5, linetype="solid"

### 2022-04-04 paper_overleaf/main.tex

@@
 We predict different cases of relatedness with KIN, and test it's performance on simulated pedigrees.
-We use these simulations to find the limits of KIN, and compare it to READ and lcmlkin CITE, two existing methods.
+We use these simulations to find the limits of KIN, and compare it to READ and lcMLkin \cite{lipatov_maximum_2015}, two existing methods.
 Finally we apply KIN to published data and show that it can give reliable information about genetic kinship.
@@
 \paragraph{Critical Values}
- To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff( Fig.\ref{fig3:cutoff},  Fig.\ref{fig4:Comparison_READ_KIN}).
-Fig.\ref{fig3:cutoff} shows that for all relatedness cases except for $3^{rd}$-degree, the false positive rate is below $5\%$ even when simply selecting the model with the highest likelihood.
+ To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff( Fig.\ref{figS10:cutoff},  Fig.\ref{fig4:Comparison_READ_KIN}).
+Fig.\ref{figS10:cutoff} shows that for all relatedness cases except for $3^{rd}$-degree, the false positive rate is below $5\%$ even when simply selecting the model with the highest likelihood.
 We observe that for $3^{rd}$-degree relatives, using a cutoff of 1.0 brings down the false positive rate close to $5\%$ for all coverages except 0.05x.
 Thus, we  recommended to use a cutoff of 1 for all cases where ROH or contamination are not a concern.
-
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/plotimg/contam0_inbred0_model_performance_allroc_asc0_plot.png}
-    \centering
-    \caption{False positive and True Positive rates as a function of cutoff on log likelihood ratio.
-In top panel, dotted line represents $5\%$.
-Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
-'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
-    \label{fig3:cutoff}
-\end{figure}
 
@@
 In the case without ROH (Fig \ref{fig2:ROH}A), we correctly infer that there is no ROH for all but two windows.
-We observe that the model predicts ROH probabilities well in all cases, even at a coverage of 0.1 \ref{tab:Table 2}.
+We observe that the model predicts ROH probabilities well in all cases, even at a coverage of 0.1x \ref{tab:Table 2}.
 
@@
 We show that both methods have similar performance for low-coverage shotgun data (``control''-case),  except for $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages.
-We show that adding contamination adversely affects both the False Positve and True Positive rates corresponding to 1st and 2nd Degree relations for READ, while this has no effect on KIN performance.
-Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
-'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.
+We show that adding contamination adversely affects both the False Positve and True Positive rates corresponding to 1st and 2nd Degree relations for READ, while this does not affect KIN performance too much.
 
@@
 Fig.
-shows that in the control case we can predict IBD states very well at high coverages, and our accuracy decreases with downsampling.
+shows that we can predict IBD states for Unrelated $3^{rd}$ Degree, $2^nd Degree$, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling.
+On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03 x.
 
 \begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/plotimg/}
+    \includegraphics[width=16cm]{plots/plotimg/plot_IBDaccuracy.png}
     \centering
-    \caption{Comparison of KIN with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH.
-Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
-'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
+    \caption{Evaluation of IBD estimation at different coverages.
+y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis.
+The error bars are drawn at 1 standard deviation from the mean.}
     \label{fig4:Comparison_READ_KIN}
@@
 
+\begin{figure}[h!]
+    \includegraphics[width=16cm]{plots/plotimg/contam0_inbred0_model_performance_allroc_asc0_plot.png}
+    \centering
+    \caption{False positive and True Positive rates as a function of cutoff on log likelihood ratio.
+In top panel, dotted line represents $5\%$.
+Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
+'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
+    \label{figS10:cutoff}
+\end{figure}
+
 *Remember to use updated lcMLKIN in the plot.

### 2022-04-05 plots/Rscripts/IBD_accuracyPlot.R
changed size

@@ -55,7 +55,7 @@ IBD_accuracy_plot <- function(infile,outfile){
     theme(legend.position = "bottom")
 
   ggsave(outfile,
-         width = 10, height = 8, dpi = 150, units = "in", device='png')
+         width = 10, height = 4, dpi = 150, units = "in", device='png')
   }
 
 IBD_accuracy_plot(infile=snakemake@input[["inf"]], outfile=snakemake@output[["outf"]])

### 2022-04-05 paper_overleaf/contamination.tex
Updates from Overleaf

--- 2022-04-04 paper_overleaf/main.tex
@@
-\documentclass[12pt, letterpaper]{article}
-\usepackage[utf8]{inputenc}
-\usepackage[margin=1in]{geometry}
-\usepackage[super]{nth}
-\usepackage{hyperref}
-\usepackage{lineno}
-\usepackage[
-singlelinecheck=false
-]{caption}
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{bm}
-\usepackage{bbm}
-\usepackage{graphicx}
-\usepackage{csvsimple}
-\usepackage[section]{placeins}
-\usepackage{lineno}
-\linenumbers
-
-\title{KIN: Method to infer relatedness using low coverage ancient DNA}
-\author{Divya Ratan Popli, Stéphane Peyrégne, Benjamin M. Peter}
-\date{5 August 2021}
-\linenumbers
-
-\setlength{\parskip}{1em}
-\setlength{\parindent}{0em}
-
-\newcommand{\BZ}{\mathbf{Z}}
-\newcommand{\BD}{\mathbf{D}}
-\newcommand{\BN}{\mathbf{N}}
-\newcommand{\BH}{\mathbf{H}}
-\newcommand{\Btheta}{\pmb{\theta}}
-
-\begin{document}
-\nolinenumbers
-
-\maketitle
-
-\begin{abstract}
-
-\noindent Knowledge of familial relationships is an important part of several research fields.
-Knowing genetic kinship of ancient specimens can lead to interesting insights into the culture and social hierarchy of ancient populations, and is also relevant for downstream genetic analyses.
-However, estimating relatedness from ancient DNA can be difficult since DNA sequences extracted from ancient remains may have low coverage, ascertainment bias, and contamination from modern human DNA.
-In addition, the population of interest may have long runs of homozygosity (ROH) due to recent inbreeding or/and small population size.
-This can affect the relatedness estimates.
-Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness up to 3rd degree, while additionally differentiating between siblings and parent-child with as low as 0.05x coverage DNA data.
-The output can be further refined with another HMM that we developed to estimate ROH proportions.
-We focus on the application of KIN on ancient human DNA, but it is applicable to other diploid species as well.
-On a machine with 8 cores, the generation of input files from 17 bam files takes -- minutes, and running KIN software takes -- minutes.
-\end{abstract}
-
-\section{Introduction}
-
-\subsection{Why study relatedness?}
-
-Identifying related individuals is a common task in genetic studies.
-Relatedness is of direct interest in e.g.
-DNA forensics, where familial search can aid in solving criminal cases, and to identify unknown deceased persons \cite{murphy_law_2018,ram_genealogy_2018}.
-Genetic paternity tests have an important application in resolving family relation, e.g.
-in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}.
-It is also an essential  step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.
-
-In ancient DNA studies, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_KINship-based_2019,sikora_ancient_2017}.
-For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to avoid inbreeding and ensure diversity.
-~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}
-
-\subsection{Approaches to estimate relatedness from high-coverage data}
-
-Commonly, pairs of related individuals are identified by looking for parts of the genome that are identical by descent (IBD), ie.
-inherited from a recent common ancestor.
-Due to the laws of Mendelian segregation, each parent, for example, will share exactly one set of chromosomes IBD with their offspring.
-A grandparent will, on average share a quarter of their genome with a grand-child due to recombination.
-Along the genomes of a pair of diploid individuals, there are three IBD states possible at any given position: either no chromosomes are IBD, only one chromosome is IBD, or both chromosomes are IBD.
-The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals.
-For example, a pair of siblings are expected to have all three possible IBD states with proportions of 0.25,0.5,0.25, respectively (Fig \ref{fig0:schematic}).
-These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}.
-One can also use these probabilities to estimate the coefficient of relatedness $r$, which is defined as the proportion of the genome that is IBD.
-In the absence of inbreeding, this would be calculated as $r= k_1/2 + k_2$.
-
-However, it is not possible to directly observe IBD segments, and so a common approach is to first identify segments of the genome that are Identical by State (IBS) and using population allele frequencies obtained from a reference panel to calculate the probability of IBD given IBS \cite{vai_KINship_2020, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}.
-One exception is SNPduo, a software that uses IBS directly from a pair of individuals to visualize and analyze their relatedness \cite{roberson_visualization_2009}.
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib.png}
-    \centering
-    \caption{IBD sharing between siblings without and with ROH.
-In top row, recombined chromosomes inherited from parents by a pair of siblings are shown.
-On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes.
-The bottom row shows the expected proportion of differences (p) along the chromosome in both the cases.
-p can take values of $p_0$, $p_1$ or $p_2$ which are expected proportion of differences in IBD states 0,1 and 2 respectively.
-}
-    \label{fig0:schematic}
-\end{figure}
-
-These methods utilize population allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}.
-PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each Single Nucleotide Polymorphism (SNP) in the analyzed samples \cite{purcell_plink_2007}.
-KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies in the samples (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}.
-
-\begin{table}
-\caption{\label{tab:Table 1}IBD sharing probabilities for different relations in absence of inbreeding}
-\begin{tabular}{|c|c|c|c|}
-    \hline
-    Relatedness & $k_0$ & $k_1$ & $k_2$\\
-    \hline
-    Unrelated & 1 & 0 & 0\\
-    \hline
-    3rd Degree & 0.75 & 0.25 & 0\\
-    \hline
-    2nd Degree & 0.5 & 0.5 & 0\\
-    \hline
-    Siblings & 0.25 & 0.5 & 0.25\\
-    \hline
-    Parent-Child & 0 & 1 & 0\\
-    \hline
-    Identical/twins & 0 & 0 & 1\\
-    \hline
-\end{tabular}
-\label{table1}
-\end{table}
-
-\subsection{Approaches that address problems with ancient DNA data}
-One major issue with applying the above mentioned methods to ancient DNA data is that the sequence coverage is typically low, making it difficult to get genotype calls.
-Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}.
-In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods.
-However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be imprecise at lower coverages .
-Ancient DNA analyses often face additional challenges such as contamination from present-day DNA \cite{peyregne_authentict_2020}, absence of suitable reference panels, and ascertainment bias in the case of data generated from DNA capture.
-
-In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations.
-However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}.
-Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}.
-In this study, we show that contamination can cause some of the individuals in ancient populations to look more unrelated than they are.
-This happens because the DNA sequence of an ancient individual typically exhibits more differences to a contaminant sequence than that of another ancient individual from the same population.
-It is worth mentioning that different DNA libraries may have different levels of contamination, even when they have been sampled from the same specimen.
-In addition, some analyzed  genomes  may have long runs of homozygosity (ROH), for example due to a small population size, or due to recent inbreeding.
-Long ROH cause related individuals to seem genetically more similar to each other, but do not affect the genetic distance between unrelated individuals.
-
-Moreover, ancient DNA is commonly captured with a SNP array to enrich for sequences from the target species.
-Particularly methods based  on the fraction of sites in different IBS states are sensiteive to the ascertainment bias caused by this non-random selection of targted sites \cite{waples_allele_2019}.
-
-Several limitations with ancient DNA mentioned above, such as low coverage, the absence of population allele frequencies or a reference panel, and ascertainment bias are resolved by a software called READ (REF).
-This software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual.
-A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to calculate average pairwise genetic distances, which in turn are used to infer relatedness.
-However, READ only infers the degree of relatedness, and only up to second degree.
-
-\subsection{How our method works}
-Here, we present KIN (Kinship INference), a Hidden Markov Model (HMM)-based approach to estimate relatedness and IBD from low-coverage ancient DNA data.
-KIN can detect up to \nth{3}-degree relatives, and differentiates between siblings and parent-child.
-KIN is also able to take into account ROH, contamination and ascertainment.
-We validate the performance of KIN using simulations and show that we are able to infer relatedness in real data from two datasets: a group of Neandertals and a group of Bronze age individuals.
-
-\section{New Approaches}\label{new_approaches}
-
-To infer the case of relatedness, KIN fits one HMM for each pair of individual and for each relatedness case.
-The KIN-HMM infers IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts and contamination estimates in each individual into account.
-The best-fit model is then assigned as the inferred relatedness.
-If the locations of ROH tracts are unknown, we provide another HMM (ROH-HMM) to coarsely estimate the location of ROH for samples with sufficient coverage ($\geq 0.1x$).
-Our method is available on \url{https://github.com/DivyaratanPopli/KIN_snakemake} along with a \textt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.
-
-\subsection{Model description}\label{method_overview}
-
-The goal of KIN-HMM is to infer how two individuals are  relateded via the patterns of shared IBD-states along a pair of genomes.
-For this purpose,  we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb), and infer the pattern of IBD-sharing for each relatedness case we consider (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical).
-We then compare the likelihood between all models, and classify each pair to the model with the highest likelihood.
-We also return the most likely location of IBD tracts (using standard Viterbi algorithm), and the state posterior probablities $\BZ$.
-
-The details of the likelihood computation are given in section \ref{ll}.
-As any HMM, KIN-HMM  require a set of n emission (section \ref{B}) and transition probability matrices.
-We assume the transition matrix for each relatedness case is fixed.
-Our emissions include a vector of parameters ($\delta$) that describes the variance in the data for each IBD state that we infer from the data (section ADD REF).
-
-\subsection{Input of KIN-HMM}\label{hmm_input}
-The inputs of our algorithm are i) the number of overlapping sites for the $w$-th window $N_w$ for which both samples have at least one read available, ii) the expected number of pairwise differences $D_w$ at these sites, and iii) the probability of ROH in windows, by default obtained from ROH-HMM described in section \ref{roh}.
-
-$D_w$ is the number of sites that are different between a pair of individuals.
-For high-coverage data, this can be estimated accurately by comparing genotypes, but for low-coverage samples, $D_w$ needs to be estimated.
-The simplest approach is is to  randomly-sampled a read from each position ("pseudo-haploid" sequences) \cite{haak_massive_2015}.
-However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
-\begin{align}\label{eq:x}
-D_w &= \sum_{s=1}^{N_w} \nu_i(s) (1-\nu_j(s)) + (1-\nu_i(s)) \nu_j(s)
-\end{align}
-Here, $\nu_i(s)$ and $\nu_j(s)$ are the proportions of reads carrying the derived allele at SNP index $s$ for individuals i and j respectively.
-Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$.
-
-\subsection{Log likelihood of the KIN-HMM}\label{ll}
-The KIN-HMM uses $\BD$ and $\BN$ to classify each window into three hidden states $Z_w \in$ ($0$, $1$, $2$), reflecting zero, one or two shared chromosomes IBD, respectively.
-To take ROH into account, we define the variable $H_w \in$ ($0$, $1$, $2$) that designates that zero, one or both individuals are homozygous in window $w$.
-Since $H_w$ is unobserved, in practice we use the estimates from ROH-HMM $h_{wj} = P(H_w = j)$.
-
-There are three additional model parameters, $\pi, \mathbf{A}$ and $\delta$.
-The initial probabilities $\pi$ give the probabilities of being in each state $Z$ at the beginning of each chromosome, which we set to uniform for simplicity.
-The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case.
-The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).
-For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.
-
-Thus, the complete data likelihood for the HMM is
-
-\begin{align}\label{eq:1}
-\log P(\BD,\BZ|\BH, \theta, \delta) &= \log P(\BD|\BZ,\BH, \theta, \delta) + \log P(\BZ |\theta) \nonumber\\
-&= \sum_w \log P(D_w|Z_w,H_w, N_w, \delta) + \sum_w \log P(Z_w |Z_{w-1},A) + \log P(Z_0|\pi)\text{.}
-\end{align}
-
-\subsection{Emission probability}\label{B}
-
-Using this setup, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta, \delta)$ from equation \ref{eq:1} and optimize them for $\delta$.
-The simplest model is to assume that sites in each window are equally distributed and independent, in this case we could use the  binomial likelihood:
-$$P(D_w|Z_w, H_w, N_w) \sim \text{Binom}[D_w ; p(Z_w, H_w), N_w] \text{,}$$
-
-where, $p$ is the proportion of differences expected for a particular IBD and ROH state.
-If the two individuals are unrelated in a particular window (i.e.
-$Z_w = 0$), then the expected proportion of pairwise differences depends solely on the population history, and we denote this proportion with $p_0$.
-If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes \cite{kuhn_estimating_2018}.
-Thus, $p(Z_w=i, H_w=0) = p_i$.
-
-The proportion of differences between unrelated individuals $p_0$ is an important parameter.
-We follow READ \cite{kuhn_estimating_2018} and estimate $p_0$ as the median of differences for all possible pairs of individuals, which should work well if the majority of individuals in the samples are unrelated.
-We set $p_1$ to $(3/4)p_0$ and $p_2$ to $p_0/2$, because one, respectively two of the four compared chromosomes are identical in these cases.
-
-The presence of inbreeding adds an additional complication, as the number of shared chromosome may  change.
-For example, when considering two bones from the same individual, we would expect the entire genome to have a pairwise difference of $p_2$, because two out of the four compared chromosomes are between identical copies.
-However, in inbred regions all four chromosomes will be identical, and so the expected pairwise differences are zero.
-Note however, that $p_0$ does not depend on the presence of ROH, since all comparisons are between unrelated chromosomes even if both individuals are homozygous at a particular locus.
-
-Taken together, we can summarize $p$ in the following matrix, where rows give the state of $Z_w$, and columns of $H_w$:
-\begin{equation}\label{eq:2}
-    p(Z_w, H_w) = \left[\begin{array}
-{rrr}
-p_0 & p_0 & p_0 \\
-p_1 & p_2 & p_3 \\
-p_2 & p_3 & p_3
-\end{array}\right]
-\end{equation}
-
-As explained above, $p_3 = 0$.
-However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that we obtain better results by setting $p_3 = \frac{p_2}{2}$).
-
-The effect of these considerations is that even though we have nine possible combinations of $Z_w$ and $H_w$ for each window, there are actually only four unique $p$-parameters $p_i$ with $i$ $\in$ (0,1,2,3).
-
-\paragraph{Beta Binomial Model}
-The Binomial Model laid out above assumes that sites are independent.
-However, due to genetic linkage, nearby sites will be correlated.
-As a result, we empirically find that the data often has considerably higher variance than would be expected from a binomial model (fig.
-S\ref{figS1:binom}).
-We take this into account by adding an overdispersion parameter $\delta$.Just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ depends on the number of chromosomes compared and so each of the  four $p_i$ has a corresponding $\delta_i$ parameter
-
-Taken together, our emission probabilities are 
-\begin{align}\label{eq:3}
-P(D_{w}|Z_w,H_w,N_w, \delta) &\sim BB[D_w; p_i, \delta_i, N_w] \nonumber\\
-&= \binom{N_w}{D_w}\frac{B(D_w+p_i \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})},
-\end{align}
-where $i$ is fully determined by the combination of $Z_w$ and $H_w$ (see eq 3).
-
-This parameterization of the beta distribution in terms of expected value $p$ and overdispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parametrization in terms of $\alpha$ and $\beta$.
-We use this equation even if preprocessing steps result in non-integer $D_w$ and $N_w$, in which case we approximate the binomial coefficient using Gamma functions.
-
-\subsection{Estimation of $\delta$}\label{delta}
-We estimate the $\delta$-parameters using an Expectation-Maximization (EM) algorithm (REF). In an EM-algorithm, we iterate the Expecation and Maximiation Steps until the algorithm converges.
-
-\paragraph{Initialization}
-The value of $\delta_i$ is unknown to start with, and we set it to a random value between x and y.
-
-\paragraph{Expectation step}
-In the $t$-th iteration, we calculate the posterior probability of each IBD state in each window $\gamma^{(t)}_{wi} = P(Z_w=i | D_w, H_w, \theta_i, \delta_i^{(t)})$ using the forward-backward algorithm, where $\delta_i^{(t)}$ is the current guess for $\delta$ for a given IBD state.
-
-\paragraph{Maximization step}
-The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_i$.
-We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wi}$ and optionally the ROH state-probabilities $h_{w\omega}$ obtained from the ROH-HMM.
-
-\begin{align}\label{eq:4}
-\mathcal{C} &= \mathbb{E}[\log P(D_w|Z_w, H_w, \theta, \delta^{(t-1)}]\nonumber\\
-&= \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta_{i \omega}^{(t-1)}, p_{i \omega})h_{w\omega}\gamma_{wi}
-\end{align}
-
-Using equation \ref{eq:2}, we simplify this by grouping all the terms that would result in the same $p_i$ and $\delta_i$, i.e.
-
-\begin{align*}\label{eq:5}
-g_{w0} &= \gamma_{w0}\\
-g_{w1} &= \gamma_{w1} h_{w0}\\
-g_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\\
-g_{w3} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2}
-\end{align*}
-
-So we can rewrite equation \ref{eq:4} as:
-\begin{align}\label{eq:6}
-\mathcal{C} &= \sum_{i=0}^3 \sum_{w=1}^L \log P(D_{w}|Z_w, H_w, N_w, \delta_i, p_i)g_{wi}
-\end{align}
-
-The cost function \ref{eq:6} has one independent term for each $i$ and so we can separate them and estimate each $\delta_i$ independently using the XXX algorithm implemented in \texttt{scipy.optimize} (REF scipy).
-
-%\paragraph{Constraints on $\delta$ optimization}\label{delta_constraint}
-
-Unconstrained optimization of the $\delta_i$ could result in some confounding of cases, We realized that different cases of relatedness have different number of IBD states possible.
-For example, siblings may have all three IBD states present while a Parent-Child has only $Z_w=1$.
-However, the parent-child model could fit data generated under the sibling model by assigning it a very high $\delta_1$, which would reduce performance  \ref{figS4:bndsbeta}.
-We avoid this problem by constraining the $\delta$ such that the beta distributions for the different cases overlap by at most one standard deviation.
-
-\subsection{Model comparison}\label{model_comp}
-To infer the most probable relatedness case, we run our model on all cases ( section \ref{method_overview}) and compare the resulting likelihoods.
-We output the relatedness corresponding to the maximum likelihood model, and the confidence as given by the log likelihood ratio between the two highest likelihood models.
-The log likelihood-ratio is commonly used in likelihood-ratio tests to compare nested models.
-However, in our cases, models are not nested, and so standard likelihood-ratio theory does not apply.
-For this reason, we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig \ref{fig3:cutoff}).
-
-\paragraph{Grouping of cases}
-Particularly for low-quality data, we may not be able to distinguish all cases.
-Thus, we group $4^{th}$ and $5^{th}$ degree relatives together with unrelated.
-Similarly, we group  half-siblings, avuncular and grandparent-grandchild to  $2^{nd}$-degree relatives in the final results.
-We report the final pairwise classification in the following categories: Unrelated, Third Degree, Second Degree, Parent-Child, Siblings, Identical individuals.
-
-\subsection{ROH estimation model}\label{roh}
-Our HMM to detect ROH tracts works similar to the KIN-HMM described above, but we only consider one individual at a time, and only positions covered by at least two reads.
-For each site, we calculate the proportion of reads that carry different alleles, and sum them up in windows along the genome.
-We call the vector with the number of differences $\mathbf{\Delta}$, and the vector with the number of sites with at least two reads $\mathbf{N}$.
-Our model has two possible hidden states: homozygous state ($Y_w=3$), and non homozygous state ($Y-w=2$).
-As above, we collect the hidden states in a vector $\mathbf{Y} = (Y_1, \dots Y_w, \dots, Y_L)$.
-The complete data likelihood for the model in this case is then:
-
-\begin{align}\label{eq:10}
-    log P(\mathbf{\Delta},\mathbf{Y}|\Theta) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta) + log P(\mathbf{Y}|\Theta)\nonumber\\
- &= [\sum_{w} log P(\Delta_w|Y_w, \Theta) + \sum_{w} log P(Y_w|Y_w-1, \Theta)] + P(Y_0| \Theta)
-\end{align}
-
-Since the source of ROH may not be known,  we estimate both transitions and emissions.
-We calculate the emissions using a beta-binomial likelihood, and fix the mean of the distributions at expected proportion of differences in a homozygous tract ($p_3$) and expected proportion of differences in a non-homozygous tract ($p_2$).
-The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w=3$ or the state $Y_w=2$ in each window.
-The M-step for emissions in analogous to that in the KIN model (eq.
-\ref{eq:4}), and the optimization step here is done with the following cost function:
-
-\begin{align}\label{eq:11}
-\mathcal{C} = \sum_{w=1}^L \sum_{i=2}^3 P(\Delta_w|\delta_{i},p_{i}) \Gamma_{wi} ,
-\end{align}
-where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and overdispersion parameter $\eta$ similar to eq. \ref{eq:3}.
-
-To estimate transitions, we initialize the transition matrix with 0.2 for the off-diagonal entries, and update it using the standard Baum-Welch update step \cite{baum_maximization_1970}.
-
-Similar to the KIN-HMM, we avoid fitting issues by forcing all windows whose proportion of differences is larger than $p_2$ to be in the non-homozygous state (fig. S\ref{figS5:ROHforced}).
-
-\subsection{Contamination correction}\label{contam}
-Contamination by the DNA from present-day people is a common feature in human ancient DNA \cite{peyregne_present-day_2020}.
-Even levels of contamination may make samples look less similar, and thus distort relatedness analyses.
-This issue is particularly pronounced  when analyzing populations with low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations .
-To address this issue, we develop a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.
-
-We conceptualize contamination in the following way: If data from individual $i$ has a known contamination rate of $C_i$, this means that any particular read  will actually come from the desired individual with probability $1-C_i$.
-With probability $C_i$ the read comes from a contaminant.
-Thus, when comparing two reads, one drawn from individual $i$ and the other drawn from individual $j$, there are three possible cases: First, with probability $(1-C_i)(1-C_j)$, both reads are endogenous and we can use them to infer relatedness.
-With probability $C_i(1-C_j) + C_j(1-C_i)$, one read is endogenous and the other is contaminant.
-Finally, with probability $C_iC_j$, both reads are contaminant.
-If we make the simplifying assumption that both $C_i$ and $C_j$ are small, the $C_iC_j$ terms  can be ignored.
-Then, we compare contaminant read to endogenous read with the probability $C_{ij} = C_i + C_j$.
-
-We assume that the level of contamination is known (cite Stephane and Kay's review).
-We  estimate the expected number of differences from comparison of endogenous reads $\mathbb{E}[D_w^{'}]$, and the total number of sites with overlapping endogenous reads $\mathbb{E}[N_w^{'}]$.
 To estimate $\mathbb{E}[D_w^{'}]$, we first need to calculate $P(c=0|d=1)$.
@@
 We added $D_w^{'}$ to $S_w^{'}$ to get the corrected number of overlapping sites $N_w^{'}$
-
-\section{Results}\label{results}
-
-We predict different cases of relatedness with KIN, and test it's performance on simulated pedigrees.
-We use these simulations to find the limits of KIN, and compare it to READ and lcmlkin CITE, two existing methods.
-Finally we apply KIN to published data and show that it can give reliable information about genetic kinship.
-
-\paragraph{Motivating Example}
-In Fig.\ref{fig1:ibd} we show the output of different relatedness cases when they are applied to data from a pair of siblings.
-
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/plotimg/IBDplot.png}
-    \centering
-    \caption{\textbf{Comparison of pairwise difference data  and inferred IBD fragments}:  The top panel shows the proportion of differences in each window along the genome for a pair of simulated siblings.
-Dashed lines represent $p_0$, $p_1$ and $p_2$ estimates.
-The second panel shows the true IBD state for each window.
-The remaining panels show the IBD states predicted by particular relatedness models.
-The log likelihood value for each model is shown on upper left corner of the panel.
-Light and shaded background represent distinct chromosomes.}
-    \label{fig1:ibd}
-\end{figure}
-
-In this case, the siblings-model predictions match true IBD states the most, as reflected by the highest log-likelihood.
-
-\paragraph{Critical Values}
- To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff( Fig.\ref{fig3:cutoff},  Fig.\ref{fig4:Comparison_READ_KIN}).
-Fig.\ref{fig3:cutoff} shows that for all relatedness cases except for $3^{rd}$-degree, the false positive rate is below $5\%$ even when simply selecting the model with the highest likelihood.
-We observe that for $3^{rd}$-degree relatives, using a cutoff of 1.0 brings down the false positive rate close to $5\%$ for all coverages except 0.05x.
-Thus, we  recommended to use a cutoff of 1 for all cases where ROH or contamination are not a concern.
-
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/plotimg/contam0_inbred0_model_performance_allroc_asc0_plot.png}
-    \centering
-    \caption{False positive and True Positive rates as a function of cutoff on log likelihood ratio.
-In top panel, dotted line represents $5\%$.
-Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
-'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
-    \label{fig3:cutoff}
-\end{figure}
-
-\paragraph{Evaluating the ROH detection}
-We next evaluate the performance of the ROH-HMM( Fig.\ref{fig2:ROH}).
-We observe that as coverage decreases, the raw heterozygosity estimates become more noisy.
-In the case without ROH (Fig \ref{fig2:ROH}A), we correctly infer that there is no ROH for all but two windows.
-We observe that the model predicts ROH probabilities well in all cases, even at a coverage of 0.1 \ref{tab:Table 2}.
-
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot_final.png}
-    \centering
-    \caption{Estimation of ROH probabilities along the genome in simulations.
-Top panel shows proportion of differences in a simulated individual in windows along the genome, and the bottom panel shows probability of seeing no ROH.
-(A) Simulation with no ascertainment, contamination, or ROH.
-(B) Simulation with ROH.
-(C) Simulation with ROH and ascertainment.
-(D) Simulation with ROH and contamination.}
-    \label{fig2:ROH}
-\end{figure}
-
-\paragraph{Comparing KIN-HMM with READ}
-In Figure Fig.
-\ref{fig4:Comparison_READ_KIN}, we compare the classification accuracy of KIN (cutoff: 1 log-likelihood unit) to that of READ (cutoff: 1 standard deviation)  under four different simulation scenarios.
-We show that both methods have similar performance for low-coverage shotgun data (``control''-case),  except for $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages.
-We show that adding contamination adversely affects both the False Positve and True Positive rates corresponding to 1st and 2nd Degree relations for READ, while this has no effect on KIN performance.
-Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
-'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.
-
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/plotimg/comparison_plot.png}
-    \centering
-    \caption{Comparison of KIN with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH.
-Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
-'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
-    \label{fig4:Comparison_READ_KIN}
-\end{figure}
-
-\paragraph{Evaluating IBD prediction}
-KIN-HMM co-estimates IBD states and relatedness, and hence it has a higher power to classify $2^{nd}$ Degree relatives, and can distinguish between Siblings and Parent-Child.
-We investigate our limit to predict IBD states along the genome by calculating the number of genomic windows where we correctly  predict IBD state for a pair of individuals.
-We calculate mean and standard deviation for this accuracy over 60 simulations of a pedigree (see section \ref{simulat}).
-Fig.
-shows that in the control case we can predict IBD states very well at high coverages, and our accuracy decreases with downsampling.
-
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/plotimg/}
-    \centering
-    \caption{Comparison of KIN with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH.
-Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
-'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
-    \label{fig4:Comparison_READ_KIN}
-\end{figure}
-
-\subsubsection{Application to Chagyrskaya Neanderthals}
-To test KIN on real ancient dataset, we applied it to a Neandertal data set  from Chagyrskaya cave in Siberia \cite{kolobova_archaeological_2020-1}, Russia.
-This dataset contains genetic data from  14 fossils that might  possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya .
-The data has low-to intermediate coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage.
-Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena (again cite Laurits paper).
-All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes and common variations in Africans (Laurits et al.,2022), and we further restricted  this data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}.
-is dataset has all the issues that KIN deals with, and hence we decided to apply it here to see if we could find family relationships that would not be possible with other available methods.
-Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}.
-We found three fossils from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair  (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60).
-We also find that Chagyrskaya 06 and Chagyrskaya 14 are classified as parent-child with low confidence, but from the morphology they likely stem from the same individual.
-We believe that the contradiction is due to non human contamination present in these libraries ($1\%$ and $2.9\%$ respectively).
-Further, we identify   Chagyrskaya07 and Chagyrskaya60 as 3rd-degree relatives.
-
-Our estimates are consistent with those obtained using READ, except that READ is unable to detect 3rd-degree relationships, and cannot distinguish between siblings and parent-child.
-
-We compared our IBD estimates to lcMLKIN for all pairs for whom READ results (with s.d.
->1) and KIN results ($\Delta LL$>1) match (fig.
-S\ref{figS6:Chagyrskaya_ibd}).
-We find that lcMLKIN does create four different clusters for different relatedness.
-However, the clusters are strongly deviated from the expected values, likely due to the ROH.
-Fossils from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $r \approx 0.6, k_0 \approx 0.33$.
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/kinplot.png}
-    \centering
-    \caption{Application of KIN on Neandertal specimens from Chagyrskaya.
-The color of a square represents the relatedness, while the number within denotes log likelihood ratio between the two maximum likelihood models.}
-    \label{fig5:Chagyrskaya_KIN}
-\end{figure}
-
-We further applied KIN to a genome-wide dataset of 118 ancient individuals sequenced and analyzed by Mittnik et al.,2019.
-We compared our relatedness estimates to those obtained from the authors using READ and lcMLKIN, and found that in most cases the three methods agree.
-Out of 6903 pairwise comparisons, 346 pairs show different relatedness for KIN and READ (table S2).
-However, most of these are at very low confidence, and there are only 31 pairs where KIN has $\Lambda >1$.
-When we investigated some of these comparisons in more detail we found that in many cases, the proportion of differences along the genome align with the prediction by KIN (fig.
-S\ref{figS8:eg1}).
-Only 20 pairs in this subset of 346 pairs had sufficient data for analysis using lcMLKIN (more than 10,000 overlapping SNPs).
-Among these 20 pairs, we found that lcMLKIN matches KIN's estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLKIN predicts 3rd-5th degree, KIN predicts second degree) and pair AITI72-AITI77A (READ and lcMLKIN predict unrelated, while KIN estimates 3rd degree with log likelihood=1.7).
-We further found 43 pairs for which READ estimates first degree, and compared the estimates of lcMLKIN and KIN for these cases (suppl table3).
-There were a total of 20 pairs where lcMLKIN had sufficient data, and 26 pairs where $loglikelihood ratio>2$ for KIN.
-Among the 19 overlapping pairs, we found four pairs where KIN and lcMLKIN differ, and again the heterozygosities along the genome match the inference of KIN (fig.
-S\ref{figS9:eg2}).
-
-\section{Discussion}\label{discussion}
-
-Here, we present a new method called KIN to estimate genetic kinship and the location IBD tracts from low coverage data, taking into account long ROH, contamination, as well as ascertainment bias.
-Our method utilizes a set of HMMs to estimate IBD tracts, conditional on a particular relatedness, and we use the log likelihoods of these HMMs to find the best fit.
-
-We used simulated data to validate KIN's performance.
-We compared KIN's model performance to a widely used software READ to show that our approach gives similar predictions as READ on ideal data with 4x coverage, and without long ROH, ascertainment and contamination, while additionally distinguishing between parent-child and siblings, and identifying third degree relatives.
-We demonstrate that at low coverage (0.1x) we have higher power to detect first degree relatives compared to READ with similar false positives, while a higher power to identify second degree relatives at the cost of slightly higher false positive rate.
-
-Counter-intuitively, we see that adding long ROH ($~17\%$) to simulations improves the performance of both KIN and READ.
-READ's performance improves because ROH does not cause any change in the average pairwise differences between unrelated individuals, while decreasing pairwise differences between related individuals.
-The higher the relatedness, the higher is the decrease in pairwise differences in presence of ROH.
-And so, ROH causes higher power for READ which otherwise is a conservative method.
-KIN performs well for the same reasons, but distinguishing IBD and ROH becomes important for distinguishing different relationship cases within a degree, i.e.
-when distinguishing between parent-child and siblings.
-
-Contamination, on the other hand, can affect the performance of READ adversely depending on the amount of exogenous DNA, and the number of individuals contaminated.
-We show that even at contamination levels  $\leq 3\%$, and contamination in only 8 out of 17 individuals, READ performance falls.
-Contamination in more individuals causes a shift in median difference in the population and may cause non-contaminated individuals to look more related, thus increasing the false positive rates.
-On the other hand, an increase in the amount of contamination in a few individuals can make them look less related to uncontaminated individuals.
-Assuming the contamination rates are correctly inferred, the correction we implemented leads to improved performance compare to naive methods such as READ, although they too could be ameliorated in a similar way (cite Laurits).
-SNP ascertainment causes a decrease in power for both  methods, since it reduces the amount of data available.
-However, the decrease in power is lower for KIN compared to READ.
-
-The Lech valley data  has low contamination, and no ROH.
-For pairwise comparisons with large numbers of overlapping sites (>XXXX), KIN, READ and lcMLKIN all mostly agree.
-However, KIN is able to differentiate between parent-child and siblings, and identify second and third degree relationships from just a few thousand sites overlapping between samples.
-We show that when applied to Neandertal specimens from Chagyrskaya cave, KIN identifies a pair of first-degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples (Laurits et al., 2022 (in press)).
-In addition, KIN identifies a pair of third degree relatives.
-We compare IBD estimates from KIN to lcMLKIN and show that KIN's IBD estimates are much closer to the expected estimates, given the relatedness between individuals that has been estimated with both READ and KIN.
-
-The output of KIN is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio.
-Due to the tabular output, KIN is easy to automatise and apply to large datasets.
-One limitation of our approach is that it assumes a single population.
-In case of a highly structured population, KIN may show inaccurate inferences.
-Also, our method makes the assumptions, that the median pairwise genetic difference reflects the population diversity, which fails if almost all individuals in the data set are related.
-The current implementation of KIN is restricted to five relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.
-
-\section{Materials and methods}\label{method}
-
-\subsection{Simulations}\label{simulat}
-We use simulations both for estimating the transition matrices and for testing and validating our algorithm.
-All simulations are performed by simulating unrelated individuals using $\texttt{msprime}$ \cite{kelleher_efficient_2016}, followed by an additional step where we simulate related individuals using a predetermined pedigree (Figure S\ref{figS10:pedigree}).
-
-\paragraph{Simulating pedigrees}
-
-For our simulations of background diversity, we form a population (Pop1) with constant effective size of 10,000 and sample eight diploid individuals (each made up of two haploid individuals) (fig.
-\ref{figS10:pedigree}).
-For each individual, we simulate 22 chromosomes with length $L=100$ Mb and a recombination rate of $r=10^{-8}$ per base pair per generation.
-We introduce mutations using an infinite sites model with rate  $\mu= 10^{-8}$ per base pair per generation.
-
-For the pedigree simulations, we first simulate a recombined set of chromosomes for either parent, and combine them to create the progeny.
-We simulate recombination by first drawing the number of breakpoints as a  Poisson random variable with parameter $rL$, and use a uniform distribution on $[1, L]$ to sample the positions of recombination points.
-
-\paragraph{Estimating transition matrices}
-KIN requires a transition matrix for each relatedness case.
-which we estimate  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of XX simulations.
-
-\paragraph{Verifying accuracy of transition matrices}\label{A_ver}
-For the simplest cases like grandparent-grandchild and siblings, it is possible to calculate the transition matrix from first principles, which allows validation of our simulations on the caes of grandparent-grandchild ($A_g$) and siblings ($A_s$) are very close to the respective transition matrices obtained from theoretical calculations.
-To calculate $A_g$, we count the number of IBD state change events that happen along chromosomes from grandparent-grandchild pair given a recombination event happens with rate $r$.
-In this case, every recombination event in the recombined chromosome inherited from grandparent leads to a state change.
-So, we define the rate matrix ($Q$) for our continuous time markov chain as $$Q = \left[\begin{array}
-{rrr}
--r & r \\
-r & -r 
-\end{array}\right]$$
-
-Here the state space includes only $Z_w=0$ and $Z_w=1$. Similarly, for a pair of siblings, we calculate $$Q =  
-\left[\begin{array}
-{rrr}
--4r & 4r & 0\\
-2r & -4r & 2r\\
-0 & 4r & -4r
-\end{array}\right]$$
-
-The state space includes all IBD states in this case.
-Then, we get the transition matrix in each case as $e^{Qb}$, where b is the window size.
-We show a comparison between these theoretically obtained transition matrices with the simulated ones in supp table YY.
-
-\paragraph{Simulations for method evaluation}
-Apart from the related and unrelated individuals in Pop1, we simulated more haploids in three other populations to create scenarios with ascertainment bias and contamination (fig.
-\ref{figS10:pedigree}).
-We simulated two haploids to form an individual each from two other populations (Pop2, Pop3) with split time of 3500 and 4500 generations with Pop1, and sampling time of 4000 generations and 2000 generations ago respectively.
-We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from Pop1.
-This scenario roughly models the ascertainment of the Chagyrskaya  data.
-We tested the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a Markov chain using transition matrix shown below:
-
-$$\mathbf{A} = \left[\begin{array}
-{rr}
-1-10/L & 10/L \\
-2/L & 1-2/L  \\
-\end{array}\right]
-$$
-
-From the steps described above, we got genotypes of 17 individuals in Pop1 in presence/absence of ROH and ascertainment.
-We further simulated five diploid individuals from Pop4 with split time of 20,000 generations with pop1, sampled from the present time.
-We calculated the allele frequency for Pop4 at each site, and used it to introduce contamination (with fixed contamination percentage for each individual ($<5\%$) in samples from Pop1.
-We generated reads (derived/ancestral) for different genomic coverages ranging from 4x to 0.03x, assuming a Poisson distribution.
-
-We run simulations with different cases referred to as control (with no ROH, contamination or ascertainment bias), R (with ROH), A (with ascertainment bias), and C (with fixed levels of contamination for each individual).
-
-\section{Supplementary Information}
-
-\begin{table}
-\caption{\label{tab:Table 2}Model performance for ROH prediction}
-\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
-    \hline
-    Coverage & Control(bias) & Control(var) & R(bias) & R(var) & RA(bias) & RA(var) & RC(bias) & RC(var)\\
-    \hline
-    4x & -0.007 & 0.006 & -0.081 & 0.036 & -0.083 & 0.036 & -0.091 & 0.039\\
-    \hline
-    0.2x & -0.003 & 0.003 & -0.068 & 0.03 & -0.068 & 0.026 & -0.076 & 0.029\\
-    \hline
-    0.1x & -0.003 & 0.002 & -0.049 & 0.021 & -0.05 & 0.021 & -0.057 & 0.022\\
-    \hline
-\end{tabular}
-\label{table1}
-\end{table}
-
-\renewcommand{\figurename}{Fig. S}
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sup.png}
-    \centering
-    \caption{Overall schematic of the method showing the entire pipeline from processed bam files to final relatedness and IBD estimates .
-Blue boxes show the data files, while green boxes represent scripts.}
-    \label{figS0:schematic}
-\end{figure}
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/binom.png}
-    \centering
-    \caption{Comparison of fit with Beta-binomial and Binomial distributions.
-Data is represented by the histogram of proportion of differences from all windows of a (A) unrelated, (B) Parent-Child, (C) Identical pair of individuals.
-Binomial parameter p is calculated from the data, while Beta-binomial parameters are estimated with the relatedness HMM.}
-    \label{figS1:binom}
-
-\end{figure}
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/pedigree.png}
-    \centering
-    \caption{Overview of simulated dataset.
-We simulated Pop1, and artificially mated unrelated individuals to make a family with upto 5th degree relatives, along with other unrelated and identical individuals (not shown here).
-Pop2 and Pop3 are simulated so as to introduce ascertainment, while Pop4 is a present day population used to simulate contamination.
-Split times and sampling times of the populations are shown in generations.}
-    \label{figS10:pedigree}
-
-\end{figure}
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/fil0_read_plot.png}
-    \centering
-    \caption{Application of READ on Neandertal specimens from Chagyrskaya cave.
-Color of a square represents the relatedness, while the number denotes standard deviations away from the upper threshold (we show lower threshold for unrelated pairs since upper threshold is not available).}
-    \label{figS2:Chagyrskaya_READ}
-\end{figure}
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_emissions_bnds.png}
-    \centering
-    \caption{Application of identical relatedness HMM on a pair of identical individuals with low coverage (0.2x), and ROH tracts.
-Top panel shows pairwise proportion of differences is close to expectation, but dips in some windows due to presence of ROH tracts.
-Bottom two panels show $P(Data|Z_w=2)$ without constraints and with constraints respectively.}
-    \label{figS3:bnds}
-\end{figure}
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_betaplot.png}
-    \centering
-    \caption{Comparison of beta distributions estimated with the KIN-HMM (left) without and (right) with variance constrained optimization of $\delta$ parameters.The histogram in each plot shows pairwise differences in genomic windows.
-Colored lines shows the Beta probability distribution for each IBD state $Z_w$.}
-    \label{figS4:bndsbeta}
-\end{figure}
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_ind0_forced_roh.png}
-     \centering
-    \caption{Comparison of Beta distributions estimated with ROH-HMM (left) without and (right) with constrained emissions.
-The histogram in each plot shows pairwise differences in genomic windows.
-Colored lines shows the Beta probability distribution for each homozygosity state $Y_w$.}
-    \label{figS5:ROHforced}
-\end{figure}
-
-*Remember to use updated lcMLKIN in the plot.
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/lcPlot.png}
-    \caption{Comparison of IBD states estimated for Chagyrskaya specimens using (A) lcMLKIN and (B) KIN.
-Each plot shows relatedness coefficient (calculated as $k_2+k_1/2$) on y-axis plotted against $k_0$ (proportion of genome with no chromosomes shared).
-    The relatedness shown with different colors is estimated with both READ and KIN.}
-    \label{figS6:Chagyrskaya_ibd}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/degree2_10Mwin.png}
-    \caption{Ploting total number of IBD fragments with the number of IBD fragments can tease apart avuncular relationship from grandparent-grandchild, since they form different clusters.
-Half-siblings, however, overlap considerably with both these relations.}
-    \label{figS7:second_degree}
-\end{figure}
-
-\begin{figure}[!ht]
-    \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/egplot1.png}
-    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIN differs from READ.
-(A) AITI62B-OTTM156 (READ estimates Identical individuals, lcMLKIN does not have sufficient data, and KIN predicts parent-child.
-(B) UNTA5867-UNTA5868Sk1 (READ estimates Second Degree, lcMLKIN does not have sufficient data, and KIN predicts parent-child).
-(C) AITI40-AITI72 (READ estimates Unrelated individuals, lcMLKIN shows 3rd-5th degree, and KIN predicts 3rd degree.)
-    (D) AITI2-AITI55 (READ estimates Unrelated individuals, lcMLKIN and KIN predicts predict second degree.)}
-    \label{figS8:eg1}
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/egplot2.png}
-    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIN differs from lcMLKIN.
-(A) AITI43-AITI55 (READ estimates First degree, lcMLKIN predicts siblings, while KIN predicts parent-child.
-(B) AITI70-AITI72 (READ estimates First degree, lcMLKIN predicts parent-child, while KIN predicts siblings.}
-    \label{figS9:eg2}
-\end{figure}
-
-\bibliographystyle{plain}
-\bibliography{references.bib}
-
-\end{document}

### 2022-04-05 paper_overleaf/main.tex

@@
 It is also an essential  step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.
+For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to avoid inbreeding and ensure diversity ~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}.
 
 In ancient DNA studies, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_KINship-based_2019,sikora_ancient_2017}.
-For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to avoid inbreeding and ensure diversity.
-~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}
 
@@
 However, it is not possible to directly observe IBD segments, and so a common approach is to first identify segments of the genome that are Identical by State (IBS) and using population allele frequencies obtained from a reference panel to calculate the probability of IBD given IBS \cite{vai_KINship_2020, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}.
-One exception is SNPduo, a software that uses IBS directly from a pair of individuals to visualize and analyze their relatedness \cite{roberson_visualization_2009}.
-
 \begin{figure}[h!]
@@
 
-In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations.
-However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}.
 Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}.
@@
 Contamination by the DNA from present-day people is a common feature in human ancient DNA \cite{peyregne_present-day_2020}.
-Even levels of contamination may make samples look less similar, and thus distort relatedness analyses.
-This issue is particularly pronounced  when analyzing populations with low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations .
 To address this issue, we develop a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.
 
-We conceptualize contamination in the following way: If data from individual $i$ has a known contamination rate of $C_i$, this means that any particular read  will actually come from the desired individual with probability $1-C_i$.
-With probability $C_i$ the read comes from a contaminant.
-Thus, when comparing two reads, one drawn from individual $i$ and the other drawn from individual $j$, there are three possible cases: First, with probability $(1-C_i)(1-C_j)$, both reads are endogenous and we can use them to infer relatedness.
-With probability $C_i(1-C_j) + C_j(1-C_i)$, one read is endogenous and the other is contaminant.
-Finally, with probability $C_iC_j$, both reads are contaminant.
-If we make the simplifying assumption that both $C_i$ and $C_j$ are small, the $C_iC_j$ terms  can be ignored.
-Then, we compare contaminant read to endogenous read with the probability $C_{ij} = C_i + C_j$.
+We assume that contamination rates in both individuals are known and small, and set $C_{ij} \approx C_i + C_j$, where $C_i, C_j$ are the contamination estimates from the two individuals.
+We also assume the divergence $\phi$ between our target population and the putative contaminant is known.
+With probability $C_{ij}$, a comparison between two random reads from either individual will contain a contaminant read, and thus contain a difference with probability $\phi$, and with probability $1 - C_{ij}$ it will be between endogenous ones.
+The comparisons between two contaminant reads are ignored, which is why we assume $C_{ij}$ to be small.
 
-We assume that the level of contamination is known (cite Stephane and Kay's review).
 We  estimate the expected number of differences from comparison of endogenous reads $\mathbb{E}[D_w^{'}]$, and the total number of sites with overlapping endogenous reads $\mathbb{E}[N_w^{'}]$.
-To estimate $\mathbb{E}[D_w^{'}]$, we first need to calculate $P(c=0|d=1)$.
-Here, $c$ is a binary variable with value $0$ or $1$ denoting that the comparison at a site is between endogenous reads, or between endogenous and contaminant reads respectively.
-Similarly $d$ is a binary variable taking values 0 and 1 for seeing no difference, or a difference, when comparing the reads at a given site.
-We can further expand this probability:
 
-\begin{align}\label{eq:12a}
-    P(c=0|d=1) &= \frac{P(d=1|c=0)P(c=0)}{P(d=1)} \nonumber\\
+For any particular comparison showing a difference $D$, we calculate the probability of the event $E$ that it is between endogenous reads as 
+\begin{equation}
+    P(E|D) = \frac{P(D|E)P(E)}{P(D|E)P(E) + P(D| \neg E)P( \neg E)}.
+\end{equation}
+
+Then, by linearity of expectation, we obtain our estimators for the expected number of endogenous comparisons with a difference as 
+\begin{align}
+    \mathbb{E}[D_e] &= D_{obs} P(E|D) = D_{obs} \times \frac{P(D|E)P(E)}{P(D|E)P(E) + P(D| \neg E)P( \neg E)}
+\end{align}.
+
+Of these terms, $P(E)= 1 -C_{ij} = 1 - P(\neg E)$, and $P(D| \neg E) = \phi$.
+
+For $P(D|E)$, we use an estimator based on the genome-wide average:
+
+\begin{align}
+    P(D|E) = p_e = \frac{\frac{\sum_w D_w}{\sum_w N_w} - C_{ij} \phi}{1 - C_{ij}}.
 \end{align}
 
-Here, $P(d=1|c=0)$, referred later to as $\rho_{ij}$ for simplicity, is equivalent to the proportion of differences when we compare endogenous reads from individuals $i$ and $j$.
-This can be estimated by considering that the proportion of pairwise differences $P(d=1) =\frac{\sum_w D_w}{\sum_w N_w}$, referred to as $o_{ij}$ later, can be expressed as a linear combination of the proportion of differences between endogenous reads and that between endogenous and contaminant reads \cite{peyregne_present-day_2020}.
-\begin{align}\label{eq:13}
-    o_{i,j} = (1-C_{ij}) \rho_{ij} + C_{ij} \phi \nonumber\\
-    or, \rho_{ij} = \frac{o_{ij} - C_{ij} \phi} {1 - C_{ij}}
-\end{align}
-
-Here, $\phi$ is the probability of seeing a difference when an endogenous read is compared to a contaminant read, and is equivalent to the average sequence divergence between individuals of the target and contaminating populations.
-Given this information, we can rewrite eq. \ref{eq:12a}:
-\begin{align}\label{eq:12}
-    P(c=0|d=1) &= \frac{(1 - C_{ij}) \rho_{ij}}{(1 - C_{ij}) \rho_{ij} + C_{ij} \phi} \nonumber\\
-\end{align}
-We can further write $\mathbb{E}[D_w^{'}]$ as $D_w P(c=0|d=1)$:
-\begin{align}
-    \mathbb{E}[D_w^{'}] &= D_w\frac{(1-C_{ij})\rho_{ij}}{(1-C_{ij}) \rho_{ij} + C_{ij} \phi}\\
-\end{align}
-
-Similarly, we derive $\mathbb{E}[S_w^{'}]$ as the estimate for the number of sites that do not show pairwise differences.
-\begin{align}
-    \mathbb{E}[S_w^{'}] &= (N_w-D_w) P(e=1|d=0) = (N_w - D_w)\frac{(1-\rho_{ij})(1-C_{ij})}{(1-\rho_{ij})(1-C_{ij}) + C_{ij}(1-\phi)}\\
-\end{align}
-
-%adding back the previous versions:
-If we look at one site in the window, we could find either no difference in reads from individuals $i$ and $j$ ($d_s$=0), or a non-zero difference ($0<d_s<=1$).
-Here $d_s$ is the probability of seeing a difference at site $s$, and $\sum_{s \in w} d_s =D_w$.
-We ask, what is the
-
-If we make the simplifying assumption that both $C_i$ and $C_j$ are small, terms of the order of $C^2$ can be ignored and there are just two cases: with probability $C_{i,j} =C_i + C_j$, one of the reads stems from a contaminant, and we effectively compare a contaminant fragment with an endogenous one.
-The probability that they differ is the average divergence between target and contaminating population, which we denote with $\phi$.
-The other case (w.
-p.
-$(1-C_{i,j}$) is where we compare two endogenous reads, which will differ with probability
+Taken together,
 
 \begin{equation}
-    p_{e} = \frac{p_{obs}-\phi C_{i,j}}{1-C_{i,j}}
+ \mathbb{E}[D_e]= D_{obs}\frac{p_e (1-C_{ij})}{p_e(1-C_{ij}) + C_{ij}\phi}
  \end{equation}
 
-Here, $$p_{obs} = \frac{\sum_w D_w}{\sum_w N_w}$$
-
-We estimate $\hat{D_c} = C_{i,j} N_{tot} \phi$ and $\hat{D_e} = (1-C_{i,j}) N_{tot} p_{obs}\phi$.
-If there are $N_{obs}$ overlapping sites in a window, $N_{c} = C_{i,j}N_{obs}$ of them will include a contaminant allele, and $N_e = (1-C_{i,j})N_{obs}$ will be between endogenous alleles.
-
-For KIN, we can use $N_e$ as an input (for each window), but we also need the proportion of differences, $D_e$.
-For this, we use the assumption that the proportion of sites where the contaminant differs from our samples is
-
-$$\phi = \frac{D_c}{N_c} = \frac{D_c}{C_{i,j} N_{obs}}, $$
-and thus $D_c = C_{i,j} \phi N_{obs}$.
-
-We can think of $p_c$ and $\rho$ as empirical likelihoods that a particular site shows a difference between two reads, given that the comparisons contain only endogenous or one contaminant read, respectively.
-\begin{align}
-    P(E|D) &= \frac{P(D|E)P(E)}{P(D)} \nonumber\\
-    \mathbb{E}[D_e] &= D_{obs} P(E|D) = D_{obs}\frac{p_e (1-c)}{p_e(1-c) + c\rho}\\
-    \mathbb{E}[S_e] &= D_{obs} P(E|\text{not} D) = S_{obs}\frac{(1-p_e)(1-c)}{(1-p_e)(1-c) + c(1-\rho)}
-\end{align}
-
-And for calculating $N_{e}$, we added $D_{e}$ to $S_{e}$
-
-The goal is that given $C_{i,j}, p_e, \BN and \BD$ to calculate adjusted $\hat{\BN}$ and $\hat{\BD}$ that reflect the endogenous proportion only.
-
-, and the observed proportion of differences between individuals $i$ and $j$ as $p_{oij}$.
-We calculate average endogenous pairwise difference $p_{eij}$ for individuals i and j as follows \cite{noauthor_ancient_nodate}:
-\begin{align}
-    p_{oij} = C_{ij}  p_c + (1-C_{ij})  p_{eij}\\
-    or, p_{eij} = \frac{p_{oij} - C_{ij}  p_c}{1-C_{ij}} 
-\end{align}
-
-Given $p_{eij}$, we would ideally like to calculate the complete data log likelihood for the relatedness model while summing over all possible differences we could have in a window weighted by the probability of seeing that difference.
-
-\paragraph{Short version}
-With contamination, we do not observe the number of differences $D_{true}$ and total number of comparisons $N_{true}$ directly, but instead need to approximate them.
-The window-subscript is omitted in the following for clarity:
-\begin{equation}
-    P(D_{\text{obs}} | N_\text{obs}, Z, \theta) \approx 
-    P( \hat{D} | \hat{N}, Z, \theta)
-\end{equation}
-We do that by calculating the probability that a single read is endogenous given it has a difference.
-\begin{align}
-    \hat{D} = \mathbb{E}[D | D_{\text{obs}}, c] &= 
-    D_{\text{obs}} P(E_i | D_i=1) \\
-    &= D_\text{obs} \frac{P(D_i=1 | E_i)P(E_i)}{P(D_i=1)}\\
-    &= D_\text{obs} \frac{(1-c)p_e}{(1-c)p_e + c p_c}
-\end{align}
-
-and likewise for reads that do not have a difference
-\begin{align}
-    \hat{S} = \mathbb{E}[S | S_{\text{obs}}, c] &= 
-     S_\text{obs} \frac{(1-c)(1-p_e)}{(1-c)(1-p_e) + c (1-p_c)}
-\end{align}
-
-and $\hat{N} = \hat{S} + \hat{D}$
+ Analogous considerations lead to the expected number of endogenous comparisons that yield no difference:
 
  \begin{equation}
-    P(D_{\text{obs}} | Z, N_\text{obs}, \theta) = \sum_n\sum_d P(D_{\text{obs}}, N_\text{obs} | D_w=d, N_w, C_i, C_j)\underbrace{P(D_w=d | N_w, Z, \theta)}_\text{old likelihood}
+\mathbb{E}[S_e] = S_{obs} P(E| \neg D) = S_{obs} \frac{(1-p_e)(1-C_{ij})}{(1-p_e)(1-C_{ij}) + C_{ij}(1-\phi)}.
  \end{equation}
 
-i.e. we sum over all possible $D_w$, $N_w$. Instead, we use
-
-\begin{align}
-    P(D_{cor},Z|D,H,\theta,c,p_c) &= P(D_{cor}|Z,D,H,\theta,c,p_c) P(Z|D,H,\theta,c,p_c)\nonumber\\
-    &= [\prod_{w} [\sum_\kappa P(D_{cor,w}|Z_w, H_w, \theta,c,p_c) P(D_{cor,w}=\kappa)] \prod_{w} P(Z_w|Z_{w-1}, \theta)] P(Z_0| \theta)
-\end{align}
-
-We realize that this calculation would be time consuming, and would further complicate the HMM.
-To avoid this, instead of using weighted sum of all possible $\kappa$, we use $E(D_{cor,w}|Z_w, H_w, \theta,c,p_c)$.
-This simplifies the equation 21 to :
-
-\begin{align}
-    P(D_{cor},Z|D,H,\theta,c,p_c) &= P(D_{cor}|Z,D,H,\theta,c,p_c) P(Z|D,H,\theta,c,p_c)\nonumber\\
-    &= [\prod_{w} P(E(D_{cor,w})|Z_w, H_w, \theta,c,p_c) \prod_{w} P(Z_w|Z_{w-1}, \theta)] P(Z_0| \theta)
-\end{align}
-
-We calculate $E(D_{cor,w})$ as shown:
-
-\begin{align}
-    E[N_e] + E[N_c] = N\\
-    E[D_e] + E[D_c] = D
-\end{align}
-
-Probability of comparing endogenous reads at a site where we see a difference:
-$$P(\xi | N = 1,D = 1)=\frac{(1-c) p_e}{c p_c + (1-c) p_e} $$
-Probability of comparing endogenous reads at a site where we see no difference:
-$$P(\xi | N = 1,D = 0)=\frac{(1-c)(1-p_e)}{c (1-p_c) + (1-c) (1-p_e)} $$
-Expectation of number of endogenous comparisons in a window:
-\begin{align}
-    E[\xi | N, D] = D E[\xi | N=1, D=1] + (N-D) E[\xi | N=1, D=0]\\
-    = D P(\xi | N=1, D=1) + (N-D) P(\xi | N=1, D=0)
-\end{align}
-
-Total number of endogenous sites showing a difference in a window: 
-$$D_{cor,w} = D_w* \frac{p_e(1-c)}{p_e(1-c) + c p_c} $$
-Total number of endogenous sites showing 0 difference in a window: 
-$$S_{cor,w} = S_w* \frac{(1-p_e)(1-c)}{(1-p_e)(1-c) + c (1-p_c)} $$
-Total number of endogenous sites in a window:
-$$N_{cor,w} = D_{cor,w} + S_{cor,w}$$
-
-This model outputs the contamination-corrected number of differences and overlapping sites in each window.
-
-We added $D_w^{'}$ to $S_w^{'}$ to get the corrected number of overlapping sites $N_w^{'}$
+Finally, we set $\mathbb{E}[N_e] = \mathbb{E}[S_e] + \mathbb{E}[D_e]$
 
@@
 We show that both methods have similar performance for low-coverage shotgun data (``control''-case),  except for $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages.
-We show that adding contamination adversely affects both the False Positve and True Positive rates corresponding to 1st and 2nd Degree relations for READ, while this does not affect KIN performance too much.
+We show that adding contamination adversely affects the True Positive rate for $2^{nd}$ Degree, and both the False Positve and True Positive rates corresponding to 2nd Degree relations for READ, while this does not affect KIN performance too much.
 
@@
 'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
-    \label{fig4:Comparison_READ_KIN}
+    \label{fig3:Comparison_READ_KIN}
 \end{figure}
@@
 Fig.
-shows that we can predict IBD states for Unrelated $3^{rd}$ Degree, $2^nd Degree$, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling.
+\ref{fig4:IBDstate_accu} shows that we can predict IBD states for Unrelated $3^{rd}$ Degree, $2^nd Degree$, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling.
 On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03 x.
@@
 The error bars are drawn at 1 standard deviation from the mean.}
-    \label{fig4:Comparison_READ_KIN}
+    \label{fig4:IBDstate_accuracy}
 \end{figure}
@@
 To test KIN on real ancient dataset, we applied it to a Neandertal data set  from Chagyrskaya cave in Siberia \cite{kolobova_archaeological_2020-1}, Russia.
-This dataset contains genetic data from  14 fossils that might  possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya .
+This dataset contains genetic data from  14 fossils that might  possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya (Laurits et al., 2022, in press).
 The data has low-to intermediate coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage.
-Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena (again cite Laurits paper).
+Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena (Laurits et al., 2022, in press).
 All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes and common variations in Africans (Laurits et al.,2022), and we further restricted  this data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}.
-is dataset has all the issues that KIN deals with, and hence we decided to apply it here to see if we could find family relationships that would not be possible with other available methods.
 Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}.
@@
 We believe that the contradiction is due to non human contamination present in these libraries ($1\%$ and $2.9\%$ respectively).
-Further, we identify   Chagyrskaya07 and Chagyrskaya60 as 3rd-degree relatives.
+Further, we identify   Chagyrskaya 07 and Chagyrskaya 60 as 3rd-degree relatives.
 
@@
 We compared our IBD estimates to lcMLKIN for all pairs for whom READ results (with s.d.
->1) and KIN results ($\Delta LL$>1) match (fig.
+$>$1) and KIN results ($\Delta LL>$1) match (fig.
 S\ref{figS6:Chagyrskaya_ibd}).
 We find that lcMLKIN does create four different clusters for different relatedness.
-However, the clusters are strongly deviated from the expected values, likely due to the ROH.
+However, the clusters are strongly deviated from the expected values, likely due to the low coverage and ROH.
 Fossils from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $r \approx 0.6, k_0 \approx 0.33$.
@@
 
-We further applied KIN to a genome-wide dataset of 118 ancient individuals sequenced and analyzed by Mittnik et al.,2019.
+We further applied KIN to a genome-wide dataset of 118 ancient individuals sequenced and analyzed by Mittnik et al.,2019 \cite{mittnik_kinship-based_2019}.
 We compared our relatedness estimates to those obtained from the authors using READ and lcMLKIN, and found that in most cases the three methods agree.
@@
 We further found 43 pairs for which READ estimates first degree, and compared the estimates of lcMLKIN and KIN for these cases (suppl table3).
-There were a total of 20 pairs where lcMLKIN had sufficient data, and 26 pairs where $loglikelihood ratio>2$ for KIN.
+There were a total of 20 pairs where lcMLKIN had sufficient data, and 26 pairs where $\Delta LL>1$ for KIN.
 Among the 19 overlapping pairs, we found four pairs where KIN and lcMLKIN differ, and again the heterozygosities along the genome match the inference of KIN (fig.
@@
 \section{Discussion}\label{discussion}
+Here, we present a new method called KIN to estimate genetic kinship and the location of IBD tracts from low- coverage data, taking into account long ROH, contamination, as well as ascertainment bias.
+Our method utilizes a set of HMMs to estimate IBD tracts and uses them to classify each pair of individuals into a possible relatedness case, along with a measure of classification confidence.
+We evaluated the method performance of KIN, and compared it to that of READ using simulated pedigrees.
+Finally, we show applications of KIN on two ancient datasets.
 
-Here, we present a new method called KIN to estimate genetic kinship and the location IBD tracts from low coverage data, taking into account long ROH, contamination, as well as ascertainment bias.
-Our method utilizes a set of HMMs to estimate IBD tracts, conditional on a particular relatedness, and we use the log likelihoods of these HMMs to find the best fit.
+In our simulations, we find that adding long ROH ($~17\%$) to simulations improves the performance of  both KIN and READ, particularly at lower coverages.
+This is likely because ROH actually reduces the variance in differences, and so it becomes easier to estimate the location of IBD tracts.
+For READ, the presence of ROH causes a bias towards inferring closer relatedness classes, but the model we use in KIN reduces this bias  (see Fig.
+\ref{fig3:Comparison_READ_KIN}).
 
-We used simulated data to validate KIN's performance.
-We compared KIN's model performance to a widely used software READ to show that our approach gives similar predictions as READ on ideal data with 4x coverage, and without long ROH, ascertainment and contamination, while additionally distinguishing between parent-child and siblings, and identifying third degree relatives.
-We demonstrate that at low coverage (0.1x) we have higher power to detect first degree relatives compared to READ with similar false positives, while a higher power to identify second degree relatives at the cost of slightly higher false positive rate.
-
-Counter-intuitively, we see that adding long ROH ($~17\%$) to simulations improves the performance of both KIN and READ.
-READ's performance improves because ROH does not cause any change in the average pairwise differences between unrelated individuals, while decreasing pairwise differences between related individuals.
-The higher the relatedness, the higher is the decrease in pairwise differences in presence of ROH.
-And so, ROH causes higher power for READ which otherwise is a conservative method.
-KIN performs well for the same reasons, but distinguishing IBD and ROH becomes important for distinguishing different relationship cases within a degree, i.e.
-when distinguishing between parent-child and siblings.
-
-Contamination, on the other hand, can affect the performance of READ adversely depending on the amount of exogenous DNA, and the number of individuals contaminated.
-We show that even at contamination levels  $\leq 3\%$, and contamination in only 8 out of 17 individuals, READ performance falls.
-Contamination in more individuals causes a shift in median difference in the population and may cause non-contaminated individuals to look more related, thus increasing the false positive rates.
-On the other hand, an increase in the amount of contamination in a few individuals can make them look less related to uncontaminated individuals.
-Assuming the contamination rates are correctly inferred, the correction we implemented leads to improved performance compare to naive methods such as READ, although they too could be ameliorated in a similar way (cite Laurits).
-SNP ascertainment causes a decrease in power for both  methods, since it reduces the amount of data available.
+Contamination can affect the accuracy of relatedness inference for two reasons.
+For one, if a substantial fraction of samples is contaminated, then the estimation of $p_0$ becomes inaccurate, because the majority of pairwise differences will include at least one contaminated sample.
+The second issue is that contaminated samples will look less similar to other individuals, and thus cause a bias towards inferring them to be less related to other individuals.
+The amount of contamination does not need to be large for this to be important, in our simulations we find that even at contamination levels $\leq 3\%$, the performance of READ is substantially reduced.
+Assuming the contamination rates are correctly inferred, the correction we implemented leads to improved performance compared to naive methods such as READ, although they too could be ameliorated in a similar way (Laurits et al., 2022 (in press)).
+SNP ascertainment can cause a slight decrease in power for both  methods, since it reduces the amount of data available.
 However, the decrease in power is lower for KIN compared to READ.
@@
 The Lech valley data  has low contamination, and no ROH.
-For pairwise comparisons with large numbers of overlapping sites (>XXXX), KIN, READ and lcMLKIN all mostly agree.
+For pairwise comparisons with large numbers of overlapping sites (>10000), KIN, READ and lcMLKIN all mostly agree.
 However, KIN is able to differentiate between parent-child and siblings, and identify second and third degree relationships from just a few thousand sites overlapping between samples.
@@
 In addition, KIN identifies a pair of third degree relatives.
-We compare IBD estimates from KIN to lcMLKIN and show that KIN's IBD estimates are much closer to the expected estimates, given the relatedness between individuals that has been estimated with both READ and KIN.
+In this case of a population with large amounts of ROH, we find that the inference by lcMLKIN are heavily biased, but KIN's model takes ROH into account and both the coefficient of relatedness and $k_0$ are very close to what would be expected from the inference by both READ and KIN.
 
-The output of KIN is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio.
-Due to the tabular output, KIN is easy to automatise and apply to large datasets.
 One limitation of our approach is that it assumes a single population.
@@
 Also, our method makes the assumptions, that the median pairwise genetic difference reflects the population diversity, which fails if almost all individuals in the data set are related.
-The current implementation of KIN is restricted to five relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.
+The current implementation of KIN is restricted to six relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.
+
+While we have focused on the application of KIN on ancient human samples, but the model is not tied to this system.
+In addition, the output of KIN is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio.
+Thus, assuming we know the recombination rate, and hence can estimate the transition matrix \ref{method}, KIN can be widely applied to any diploid species  and is easy to automatise and apply to large datasets.
 
@@
 We use simulations both for estimating the transition matrices and for testing and validating our algorithm.
-All simulations are performed by simulating unrelated individuals using $\texttt{msprime}$ \cite{kelleher_efficient_2016}, followed by an additional step where we simulate related individuals using a predetermined pedigree (Figure S\ref{figS10:pedigree}).
+All simulations are performed in a scenario mimicking the analysis of a Neandertal population contaminated by modern humans.
+We simulate unrelated individuals using $\texttt{msprime}$ \cite{kelleher_efficient_2016}, followed by an additional step where we simulate related individuals using a predetermined pedigree (Figure S\ref{figS10:pedigree}).
 
@@
 
-For our simulations of background diversity, we form a population (Pop1) with constant effective size of 10,000 and sample eight diploid individuals (each made up of two haploid individuals) (fig.
+For our simulations of background diversity, we form a population (Pop1) with constant effective size of 10,000 and sample eight diploid individuals (each made up of two haploid individuals) from 2500 generations ago (fig.
 \ref{figS10:pedigree}).
@@
 
-\paragraph{Estimating transition matrices}
+\paragraph{Transition matrices}
 KIN requires a transition matrix for each relatedness case.
-which we estimate  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of XX simulations.
+which we estimate  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of 1000 simulations from our pedigree (Figure SXXXX).
+For two cases, siblings and grandparent-grandchild, it is possible to write down the theoretical expectation of the transition matrix:
 
-\paragraph{Verifying accuracy of transition matrices}\label{A_ver}
-For the simplest cases like grandparent-grandchild and siblings, it is possible to calculate the transition matrix from first principles, which allows validation of our simulations on the caes of grandparent-grandchild ($A_g$) and siblings ($A_s$) are very close to the respective transition matrices obtained from theoretical calculations.
-To calculate $A_g$, we count the number of IBD state change events that happen along chromosomes from grandparent-grandchild pair given a recombination event happens with rate $r$.
-In this case, every recombination event in the recombined chromosome inherited from grandparent leads to a state change.
-So, we define the rate matrix ($Q$) for our continuous time markov chain as $$Q = \left[\begin{array}
+For grandparent-grandchild, 
+ $$Q = \left[\begin{array}
 {rrr}
@@
 
-Here the state space includes only $Z_w=0$ and $Z_w=1$. Similarly, for a pair of siblings, we calculate $$Q =  
+Here the state space includes only $Z_w=0$ and $Z_w=1$. 
+Similarly, for a pair of siblings, we calculate $$Q =  
 \left[\begin{array}
@@
 
-The state space includes all IBD states in this case.
-Then, we get the transition matrix in each case as $e^{Qb}$, where b is the window size.
-We show a comparison between these theoretically obtained transition matrices with the simulated ones in supp table YY.
+The state space includes all IBD states in this case. For these two cases, we get the transition matrix in each case as $e^{Qb}$, where b is the window size.
 
@@
 
-From the steps described above, we got genotypes of 17 individuals in Pop1 in presence/absence of ROH and ascertainment.
-We further simulated five diploid individuals from Pop4 with split time of 20,000 generations with pop1, sampled from the present time.
+From the steps described above, we got genotypes of individuals in Pop1 in presence/absence of ROH and ascertainment.
+We further simulated five diploid individuals from Pop4 with split time of 20,000 generations with Pop1, sampled from the present time.
 We calculated the allele frequency for Pop4 at each site, and used it to introduce contamination (with fixed contamination percentage for each individual ($<5\%$) in samples from Pop1.

### 2022-04-05 plots/Rscripts/IBD_accuracyPlot.R
changed label size

@@ -52,7 +52,7 @@ IBD_accuracy_plot <- function(infile,outfile){
     geom_point(size = 1.5)
   f <- f + labs(y="Accuracy", x="Coverage") +
     scale_color_manual(values = colors1, name="Relatedness") + theme_bw() +geom_path(aes(x=cov, y=accuracy, group=rel)) +
-    theme(legend.position = "bottom")
+    theme(legend.position = "bottom", text = element_text(size=20))
 
   ggsave(outfile,
          width = 10, height = 4, dpi = 150, units = "in", device='png')

### 2022-04-05 plots/Rscripts/comparisonPlot.R
changed ylim, dotted line alpha

@@ -46,9 +46,9 @@ comparison_plot <- function(inf,outf){
 
   plot<-ggplot(data=data, aes(x=Coverage, y=classification_proportion, group=interaction(Simulation,Method), color=Simulation, lty=Method))
   plot<-plot+geom_line()
-  plot<- plot + facet_grid(T_F ~ Relatedness, labeller = labeller(T_F = T_F.labs, Relatedness = Relatedness.labs))
+  plot<- plot + facet_grid(T_F ~ Relatedness, labeller = labeller(T_F = T_F.labs, Relatedness = Relatedness.labs), scales="free")
   #plot + geom_hline(yintercept=0.05, linetype="dashed")
-  plot <- plot + geom_hline(data = data.frame(T_F='False_positive'), aes(yintercept = 0.05), linetype = "dotted")
+  plot <- plot + geom_hline(data = data.frame(T_F='False_positive'), aes(yintercept = 0.05), linetype = "dotted", alpha=0.5)
   plot <- plot+ labs(y="Classification Proportions", x = "Coverage")
   plot + theme_bw() + theme(
     strip.background = element_rect(

### 2022-04-05 paper_overleaf/main.tex
Updates from Overleaf

@@
 y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis.
-The error bars are drawn at 1 standard deviation from the mean.}
+The error bars are drawn at 1 standard deviation from the mean.
+Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
     \label{fig4:IBDstate_accuracy}
@@
 To test KIN on real ancient dataset, we applied it to a Neandertal data set  from Chagyrskaya cave in Siberia \cite{kolobova_archaeological_2020-1}, Russia.
-This dataset contains genetic data from  14 fossils that might  possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya (Laurits et al., 2022, in press).
+This dataset contains genetic data from  14 fossils that might  possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya \cite{laurits_skov_genetic_nodate}.
 The data has low-to intermediate coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage.
-Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena (Laurits et al., 2022, in press).
-All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes and common variations in Africans (Laurits et al.,2022), and we further restricted  this data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}.
+Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena \cite{laurits_skov_genetic_nodate}.
+All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes and common variations in Africans \cite{laurits_skov_genetic_nodate}, and we further restricted  this data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}.
 Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}.
@@
 The amount of contamination does not need to be large for this to be important, in our simulations we find that even at contamination levels $\leq 3\%$, the performance of READ is substantially reduced.
-Assuming the contamination rates are correctly inferred, the correction we implemented leads to improved performance compared to naive methods such as READ, although they too could be ameliorated in a similar way (Laurits et al., 2022 (in press)).
+Assuming the contamination rates are correctly inferred, the correction we implemented leads to improved performance compared to naive methods such as READ, although they too could be ameliorated in a similar way \cite{laurits_skov_genetic_nodate}.
 SNP ascertainment can cause a slight decrease in power for both  methods, since it reduces the amount of data available.
@@
 However, KIN is able to differentiate between parent-child and siblings, and identify second and third degree relationships from just a few thousand sites overlapping between samples.
-We show that when applied to Neandertal specimens from Chagyrskaya cave, KIN identifies a pair of first-degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples (Laurits et al., 2022 (in press)).
+We show that when applied to Neandertal specimens from Chagyrskaya cave, KIN identifies a pair of first-degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples \cite{laurits_skov_genetic_nodate}.
 In addition, KIN identifies a pair of third degree relatives.

### 2022-04-07 paper_overleaf/main.tex

@@
 
-\title{KIN: Method to infer relatedness using low coverage ancient DNA}
+\title{KIN: A method to infer relatedness from low-coverage ancient DNA}
 \author{Divya Ratan Popli, Stéphane Peyrégne, Benjamin M. Peter}
-\date{5 August 2021}
+%\date{5 August 2021}
 \linenumbers
@@
 
-\noindent Knowledge of familial relationships is an important part of several research fields.
-Knowing genetic kinship of ancient specimens can lead to interesting insights into the culture and social hierarchy of ancient populations, and is also relevant for downstream genetic analyses.
-However, estimating relatedness from ancient DNA can be difficult since DNA sequences extracted from ancient remains may have low coverage, ascertainment bias, and contamination from modern human DNA.
-In addition, the population of interest may have long runs of homozygosity (ROH) due to recent inbreeding or/and small population size.
-This can affect the relatedness estimates.
-Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness up to 3rd degree, while additionally differentiating between siblings and parent-child with as low as 0.05x coverage DNA data.
-The output can be further refined with another HMM that we developed to estimate ROH proportions.
-We focus on the application of KIN on ancient human DNA, but it is applicable to other diploid species as well.
-On a machine with 8 cores, the generation of input files from 17 bam files takes -- minutes, and running KIN software takes -- minutes.
+\noindent Inferring familial relationships is an important question in population genetics.
+Knowing genetic kinship of ancient individuals leads to insights into the culture and social hierarchy of ancient populations, and is  relevant for downstream genetic analyses.
+However, estimating relatedness from ancient DNA is difficult since DNA sequences extracted from ancient remains may have low coverage, ascertainment bias, or contamination from various sources.
+
+Here, we present KIN, a Hidden-Markov-Model-based approach to estimate identity-by descent-fragments and degree of relatedness up to 3rd-degree, from as low as 0.05x coverage ancient DNA data, and additionally differentiates between siblings and parent-child relationships.
+KIN also incorporates an adjustment for contamination, and an additional model to estimate the location of long runs of homozygosity, and uses it to improve classification accuracy.
+
+We show in simulations that KIN performs at near perfect classification accuracy at inferring first-degree relationships, and at 90\% accuracy for second degree even at 0.05x coverage.
+
+We illustrate the utility of KIN on two applications, one on a Neandertal population from Siberia, Russia, and one in a Bronze-Age population from the Lech Valley, Germany. 
+We find that in both cases, KIN is competitive with existing approaches and provides new insights.
+
+Thus, KIN is a useful tool for analyzing relatedness for large ancient-DNA datasets with low-coverage sequencing data that is applicable to any diploid species, without the requirement for a reference panel.
 \end{abstract}
@@
 
-However, it is not possible to directly observe IBD segments, and so a common approach is to first identify segments of the genome that are Identical by State (IBS) and using population allele frequencies obtained from a reference panel to calculate the probability of IBD given IBS \cite{vai_KINship_2020, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}.
+However, it is not possible to directly observe IBD segments, and so a common approach is to first identify segments of the genome that are Identical by State (IBS) and using population allele frequencies obtained from a reference panel to calculate the probability of IBD given IBS.
+There are several methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007,manichaikul_robust_2010}.
+
 \begin{figure}[h!]
@@
 \end{figure}
-
-These methods utilize population allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}.
-PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each Single Nucleotide Polymorphism (SNP) in the analyzed samples \cite{purcell_plink_2007}.
-KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies in the samples (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}.
 
@@
 In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods.
-However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be imprecise at lower coverages .
-Ancient DNA analyses often face additional challenges such as contamination from present-day DNA \cite{peyregne_authentict_2020}, absence of suitable reference panels, and ascertainment bias in the case of data generated from DNA capture.
+However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be imprecise at lower coverages \cite{korneliussen_angsd_2014} .
+Ancient DNA analyses often face additional challenges such as contamination from present-day DNA \cite{peyregne_present-day_2020-1}, absence of suitable reference panels, and ascertainment bias in the case of data generated from DNA capture \cite{prufer_computational_2010, vai_kinship_2020}.
 
 Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}.
-In this study, we show that contamination can cause some of the individuals in ancient populations to look more unrelated than they are.
-This happens because the DNA sequence of an ancient individual typically exhibits more differences to a contaminant sequence than that of another ancient individual from the same population.
-It is worth mentioning that different DNA libraries may have different levels of contamination, even when they have been sampled from the same specimen.
+A second issue is contamination.
+If the contamination stems from another population, contaminated data will look more dissimilar to other individuals from the analyzed population, and hence relatedness inference will be biased towards towards less relatedness.
 In addition, some analyzed  genomes  may have long runs of homozygosity (ROH), for example due to a small population size, or due to recent inbreeding.
@@
 Moreover, ancient DNA is commonly captured with a SNP array to enrich for sequences from the target species.
-Particularly methods based  on the fraction of sites in different IBS states are sensiteive to the ascertainment bias caused by this non-random selection of targted sites \cite{waples_allele_2019}.
+Particularly methods based  on the fraction of sites in different IBS states are sensitive to the ascertainment bias caused by this non-random selection of targted sites \cite{waples_allele_2019}.
 
-Several limitations with ancient DNA mentioned above, such as low coverage, the absence of population allele frequencies or a reference panel, and ascertainment bias are resolved by a software called READ (REF).
+Several limitations with ancient DNA mentioned above, such as low coverage, the absence of population allele frequencies or a reference panel, and ascertainment bias are resolved by a software called READ \cite{kuhn_estimating_2018}.
 This software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual.
@@
 
-The goal of KIN-HMM is to infer how two individuals are  relateded via the patterns of shared IBD-states along a pair of genomes.
+The goal of KIN-HMM is to infer how two individuals are related via the patterns of shared IBD-states along a pair of genomes.
 For this purpose,  we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb), and infer the pattern of IBD-sharing for each relatedness case we consider (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical).
@@
 The details of the likelihood computation are given in section \ref{ll}.
-As any HMM, KIN-HMM  require a set of n emission (section \ref{B}) and transition probability matrices.
+As any HMM, KIN-HMM requires a set of emission (section \ref{B}) and transition probability matrices.
 We assume the transition matrix for each relatedness case is fixed.
-Our emissions include a vector of parameters ($\delta$) that describes the variance in the data for each IBD state that we infer from the data (section ADD REF).
+Our emissions include a vector of parameters ($\delta$) that describe the variance in the data for each IBD state that we infer from the data (section \ref{B}).
 
@@
 For high-coverage data, this can be estimated accurately by comparing genotypes, but for low-coverage samples, $D_w$ needs to be estimated.
-The simplest approach is is to  randomly-sampled a read from each position ("pseudo-haploid" sequences) \cite{haak_massive_2015}.
+The simplest approach is is to  randomly-sample a read from each position to make "pseudo-haploid" sequences \cite{haak_massive_2015}.
 However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
@@
 To take ROH into account, we define the variable $H_w \in$ ($0$, $1$, $2$) that designates that zero, one or both individuals are homozygous in window $w$.
-Since $H_w$ is unobserved, in practice we use the estimates from ROH-HMM $h_{wj} = P(H_w = j)$.
+Since $H_w$ is unobserved, in practice we use the estimates from ROH-HMM: $h_{wj} = P(H_w = j)$.
 
 There are three additional model parameters, $\pi, \mathbf{A}$ and $\delta$.
-The initial probabilities $\pi$ give the probabilities of being in each state $Z$ at the beginning of each chromosome, which we set to uniform for simplicity.
+The initial probabilities $\pi$ give the probabilities of being in each state $Z_0$ (at the beginning of each chromosome), which we set to uniform for simplicity.
 The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case.
@@
 We follow READ \cite{kuhn_estimating_2018} and estimate $p_0$ as the median of differences for all possible pairs of individuals, which should work well if the majority of individuals in the samples are unrelated.
-We set $p_1$ to $(3/4)p_0$ and $p_2$ to $p_0/2$, because one, respectively two of the four compared chromosomes are identical in these cases.
 
@@
 \paragraph{Initialization}
-The value of $\delta_i$ is unknown to start with, and we set it to a random value between x and y.
+The value of $\delta_i$ is unknown to start with, and we set it to a random value between 0 and 1000.
 
@@
 The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_i$.
-We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wi}$ and optionally the ROH state-probabilities $h_{w\omega}$ obtained from the ROH-HMM.
+We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wj}$ and optionally the ROH state-probabilities $h_{w\omega}$ obtained from the ROH-HMM.
 
@@
 \mathcal{C} &= \mathbb{E}[\log P(D_w|Z_w, H_w, \theta, \delta^{(t-1)}]\nonumber\\
-&= \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta_{i \omega}^{(t-1)}, p_{i \omega})h_{w\omega}\gamma_{wi}
+&= \sum_{w=1}^L \sum_{j=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta_{j \omega}^{(t-1)}, p_{j \omega})h_{w\omega}\gamma_{wj}
 \end{align}
@@
 
-The cost function \ref{eq:6} has one independent term for each $i$ and so we can separate them and estimate each $\delta_i$ independently using the XXX algorithm implemented in \texttt{scipy.optimize} (REF scipy).
+The cost function \ref{eq:6} has one independent term for each $i$ and so we can separate them and estimate each $\delta_i$ independently using the minimize\_scalar algorithm implemented in \texttt{scipy.optimize} \cite{virtanen_scipy_2020}.
 
@@
 For example, siblings may have all three IBD states present while a Parent-Child has only $Z_w=1$.
-However, the parent-child model could fit data generated under the sibling model by assigning it a very high $\delta_1$, which would reduce performance  \ref{figS4:bndsbeta}.
+However, the parent-child model could fit data generated under the sibling model by assigning it a very high $\delta_1$, which would reduce performance  (for example, see fig.
+S \ref{figS4:bndsbeta}).
 We avoid this problem by constraining the $\delta$ such that the beta distributions for the different cases overlap by at most one standard deviation.
@@
 \subsection{Model comparison}\label{model_comp}
-To infer the most probable relatedness case, we run our model on all cases ( section \ref{method_overview}) and compare the resulting likelihoods.
+To infer the most probable relatedness case, we run our model on all cases (section \ref{method_overview}) and compare the resulting likelihoods.
 We output the relatedness corresponding to the maximum likelihood model, and the confidence as given by the log likelihood ratio between the two highest likelihood models.
 The log likelihood-ratio is commonly used in likelihood-ratio tests to compare nested models.
-However, in our cases, models are not nested, and so standard likelihood-ratio theory does not apply.
-For this reason, we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig \ref{fig3:cutoff}).
+However, in our case, models are not nested, and so standard likelihood-ratio theory does not apply.
+For this reason, we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig S \ref{figS10:cutoff}).
 
@@
 We call the vector with the number of differences $\mathbf{\Delta}$, and the vector with the number of sites with at least two reads $\mathbf{N}$.
-Our model has two possible hidden states: homozygous state ($Y_w=3$), and non homozygous state ($Y-w=2$).
+Our model has two possible hidden states: homozygous state ($Y_w=3$), and non homozygous state ($Y_w=2$).
 As above, we collect the hidden states in a vector $\mathbf{Y} = (Y_1, \dots Y_w, \dots, Y_L)$.
@@
 \begin{align}\label{eq:10}
-    log P(\mathbf{\Delta},\mathbf{Y}|\Theta) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta) + log P(\mathbf{Y}|\Theta)\nonumber\\
- &= [\sum_{w} log P(\Delta_w|Y_w, \Theta) + \sum_{w} log P(Y_w|Y_w-1, \Theta)] + P(Y_0| \Theta)
+    log P(\mathbf{\Delta},\mathbf{Y}|\Theta,\delta) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta, \delta) + log P(\mathbf{Y}|\Theta)\nonumber\\
+ &= [\sum_{w} log P(\Delta_w|Y_w, \Theta, \delta) + \sum_{w} log P(Y_w|Y_{w-1}, \Theta)] + P(Y_0| \Theta)
 \end{align}
@@
 \end{align}
-where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and overdispersion parameter $\eta$ similar to eq. \ref{eq:3}.
+where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and overdispersion parameter $\delta$ similar to eq. \ref{eq:3}.
 
-To estimate transitions, we initialize the transition matrix with 0.2 for the off-diagonal entries, and update it using the standard Baum-Welch update step \cite{baum_maximization_1970}.
+To estimate transitions, we initialize the transition matrix with the value 0.2 for the off-diagonal entries, and update it using the standard Baum-Welch update step \cite{baum_maximization_1970}.
 
@@
 Contamination by the DNA from present-day people is a common feature in human ancient DNA \cite{peyregne_present-day_2020}.
-To address this issue, we develop a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.
+To address this issue, we developed a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.
 
@@
 
-Then, by linearity of expectation, we obtain our estimators for the expected number of endogenous comparisons with a difference as 
+Then, by linearity of expectation, we obtain our estimator for the expected number of endogenous comparisons with a difference as 
 \begin{align}
-    \mathbb{E}[D_e] &= D_{obs} P(E|D) = D_{obs} \times \frac{P(D|E)P(E)}{P(D|E)P(E) + P(D| \neg E)P( \neg E)}
+    \mathbb{E}[D_w^{'}] &= D_{w} P(E|D) = D_{w} \times \frac{P(D|E)P(E)}{P(D|E)P(E) + P(D| \neg E)P( \neg E)}
 \end{align}.
@@
 \begin{align}
-    P(D|E) = p_e = \frac{\frac{\sum_w D_w}{\sum_w N_w} - C_{ij} \phi}{1 - C_{ij}}.
+    P(D|E) = \rho = \frac{\frac{\sum_w D_w}{\sum_w N_w} - C_{ij} \phi}{1 - C_{ij}}.
 \end{align}
@@
 \begin{equation}
- \mathbb{E}[D_e]= D_{obs}\frac{p_e (1-C_{ij})}{p_e(1-C_{ij}) + C_{ij}\phi}
+ \mathbb{E}[D_w^{'}]= D_{w}\frac{\rho (1-C_{ij})}{\rho(1-C_{ij}) + C_{ij}\phi}
  \end{equation}
@@
  \begin{equation}
-\mathbb{E}[S_e] = S_{obs} P(E| \neg D) = S_{obs} \frac{(1-p_e)(1-C_{ij})}{(1-p_e)(1-C_{ij}) + C_{ij}(1-\phi)}.
+\mathbb{E}[S_w^{'}] = S_{w} P(E| \neg D) = S_{w} \frac{(1-\rho)(1-C_{ij})}{(1-\rho)(1-C_{ij}) + C_{ij}(1-\phi)}.
  \end{equation}
 
-Finally, we set $\mathbb{E}[N_e] = \mathbb{E}[S_e] + \mathbb{E}[D_e]$
+Here, $S_w = N_w - D_w$. Hence, we set $\mathbb{E}[N_w^{'}] = \mathbb{E}[S_w^{'}] + \mathbb{E}[D_w^{'}]$. We do a similar contamination correction for the input of ROH-HMM.
 
@@
 \paragraph{Critical Values}
- To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff( Fig.\ref{figS10:cutoff},  Fig.\ref{fig4:Comparison_READ_KIN}).
+ To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff( Fig.\ref{figS10:cutoff},  Fig.\ref{fig3:Comparison_READ_KIN}).
 Fig.\ref{figS10:cutoff} shows that for all relatedness cases except for $3^{rd}$-degree, the false positive rate is below $5\%$ even when simply selecting the model with the highest likelihood.
@@
 In the case without ROH (Fig \ref{fig2:ROH}A), we correctly infer that there is no ROH for all but two windows.
-We observe that the model predicts ROH probabilities well in all cases, even at a coverage of 0.1x \ref{tab:Table 2}.
+We observe that the model predicts ROH probabilities well in all cases, even at a coverage of 0.1x (Table \ref{tab:Table 2}).
 
@@
 In Figure Fig.
-\ref{fig4:Comparison_READ_KIN}, we compare the classification accuracy of KIN (cutoff: 1 log-likelihood unit) to that of READ (cutoff: 1 standard deviation)  under four different simulation scenarios.
-We show that both methods have similar performance for low-coverage shotgun data (``control''-case),  except for $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages.
-We show that adding contamination adversely affects the True Positive rate for $2^{nd}$ Degree, and both the False Positve and True Positive rates corresponding to 2nd Degree relations for READ, while this does not affect KIN performance too much.
+\ref{fig3:Comparison_READ_KIN}, we compare the classification accuracy of KIN (cutoff: 1 log-likelihood unit) to that of READ (cutoff: 1 standard deviation)  under four different simulation scenarios.
+We show that both methods have similar performance for low-coverage shotgun data (``control''-case),  and for ascertained data.
+One exception are  $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages.
+We also find that READ is strongly impacted by contamination, but that the correction implemented in KIN is sufficient to remove this bias..
+For simulations with ROH, we find the overall classification accuracy is only slightly reduced.
+Since ROH makes individuals more similar, we see a bias towards more false positives, for first and second-degree relationships, which is slightly higher for READ than for KIN.
 
@@
 Fig.
-\ref{fig4:IBDstate_accu} shows that we can predict IBD states for Unrelated $3^{rd}$ Degree, $2^nd Degree$, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling.
+\ref{fig4:IBDstate_accuracy} shows that we can predict IBD states for Unrelated, $3^{rd}$ Degree, $2^{nd} Degree$, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling.
 On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03 x.
@@
 \subsubsection{Application to Chagyrskaya Neanderthals}
-To test KIN on real ancient dataset, we applied it to a Neandertal data set  from Chagyrskaya cave in Siberia \cite{kolobova_archaeological_2020-1}, Russia.
+To test KIN on real ancient dataset, we applied it to a Neandertal data set  from Chagyrskaya cave in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020}.
 This dataset contains genetic data from  14 fossils that might  possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya \cite{laurits_skov_genetic_nodate}.
-The data has low-to intermediate coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage.
+The data has low-to-intermediate coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage.
 Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena \cite{laurits_skov_genetic_nodate}.
@@
 We found three fossils from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair  (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60).
-We also find that Chagyrskaya 06 and Chagyrskaya 14 are classified as parent-child with low confidence, but from the morphology they likely stem from the same individual.
-We believe that the contradiction is due to non human contamination present in these libraries ($1\%$ and $2.9\%$ respectively).
 Further, we identify   Chagyrskaya 07 and Chagyrskaya 60 as 3rd-degree relatives.
@@
 Our estimates are consistent with those obtained using READ, except that READ is unable to detect 3rd-degree relationships, and cannot distinguish between siblings and parent-child.
+We also find that both  KIN and READ classify Chagyrskaya 06 and Chagyrskaya 14 as parent-child with low confidence, but from the morphology they likely stem from the same individual.
+We believe that this due to non-human contamination present in these libraries ($1\%$ and $2.9\%$ respectively).
 
-We compared our IBD estimates to lcMLKIN for all pairs for whom READ results (with s.d.
+We compared our IBD estimates to lcMLkin for all pairs for whom READ results (with s.d.
 $>$1) and KIN results ($\Delta LL>$1) match (fig.
 S\ref{figS6:Chagyrskaya_ibd}).
-We find that lcMLKIN does create four different clusters for different relatedness.
-However, the clusters are strongly deviated from the expected values, likely due to the low coverage and ROH.
+We find that lcMLkin does create four different clusters for different relatedness cases.
+However, the location of these clusters strongly deviate from the expected values, likely due to the low coverage and ROH.
 Fossils from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $r \approx 0.6, k_0 \approx 0.33$.
@@
     \caption{Application of KIN on Neandertal specimens from Chagyrskaya.
-The color of a square represents the relatedness, while the number within denotes log likelihood ratio between the two maximum likelihood models.}
+The color of a square represents the relatedness, while the number within denotes log likelihood ratio ($\Delta LL$) between the two maximum likelihood models.}
     \label{fig5:Chagyrskaya_KIN}
@@
 
-We further applied KIN to a genome-wide dataset of 118 ancient individuals sequenced and analyzed by Mittnik et al.,2019 \cite{mittnik_kinship-based_2019}.
-We compared our relatedness estimates to those obtained from the authors using READ and lcMLKIN, and found that in most cases the three methods agree.
+We further applied KIN to a genome-wide dataset of 118 ancient individuals from the Lech valley \cite{mittnik_kinship-based_2019}.
+We compared our relatedness estimates to those obtained from the authors using READ and lcMLkin, and found that in most cases the three methods agree.
 Out of 6903 pairwise comparisons, 346 pairs show different relatedness for KIN and READ (table S2).
-However, most of these are at very low confidence, and there are only 31 pairs where KIN has $\Lambda >1$.
+However, most of these are at very low confidence, and there are only 31 pairs where KIN has $\Delta LL>1$.
 When we investigated some of these comparisons in more detail we found that in many cases, the proportion of differences along the genome align with the prediction by KIN (fig.
 S\ref{figS8:eg1}).
-Only 20 pairs in this subset of 346 pairs had sufficient data for analysis using lcMLKIN (more than 10,000 overlapping SNPs).
-Among these 20 pairs, we found that lcMLKIN matches KIN's estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLKIN predicts 3rd-5th degree, KIN predicts second degree) and pair AITI72-AITI77A (READ and lcMLKIN predict unrelated, while KIN estimates 3rd degree with log likelihood=1.7).
-We further found 43 pairs for which READ estimates first degree, and compared the estimates of lcMLKIN and KIN for these cases (suppl table3).
-There were a total of 20 pairs where lcMLKIN had sufficient data, and 26 pairs where $\Delta LL>1$ for KIN.
-Among the 19 overlapping pairs, we found four pairs where KIN and lcMLKIN differ, and again the heterozygosities along the genome match the inference of KIN (fig.
+Only 20 pairs in this subset of 346 pairs had sufficient data for analysis using lcMLkin (more than 10,000 overlapping SNPs).
+Among these 20 pairs, we found that lcMLkin matches KIN's estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLkin predicts 3rd-5th degree, KIN predicts second degree) and pair AITI72-AITI77A (READ and lcMLkin predict unrelated, while KIN estimates 3rd degree with LL=1.7).
+We further found 43 pairs for which READ estimates First-Degree, and compared the estimates of lcMLkin and KIN for these cases (Table S3).
+There were a total of 20 pairs where lcMLkin had sufficient data, and 26 pairs where $\Delta LL>1$ for KIN.
+Among the 19 overlapping pairs, we found four pairs where KIN and lcMLkin differ, and again the heterozygosities along the genome match the inference of KIN (fig.
 S\ref{figS9:eg2}).
@@
 \section{Discussion}\label{discussion}
-Here, we present a new method called KIN to estimate genetic kinship and the location of IBD tracts from low- coverage data, taking into account long ROH, contamination, as well as ascertainment bias.
+Here, we present a new method called KIN to estimate genetic kinship and the location of IBD tracts from low-coverage data, taking into account long ROH, contamination, as well as ascertainment bias.
 Our method utilizes a set of HMMs to estimate IBD tracts and uses them to classify each pair of individuals into a possible relatedness case, along with a measure of classification confidence.
@@
 This is likely because ROH actually reduces the variance in differences, and so it becomes easier to estimate the location of IBD tracts.
-For READ, the presence of ROH causes a bias towards inferring closer relatedness classes, but the model we use in KIN reduces this bias  (see Fig.
+For READ, the presence of ROH causes a bias towards inferring closer relatedness cases, but the model we use in KIN reduces this bias  (see Fig.
 \ref{fig3:Comparison_READ_KIN}).
@@
 The Lech valley data  has low contamination, and no ROH.
-For pairwise comparisons with large numbers of overlapping sites (>10000), KIN, READ and lcMLKIN all mostly agree.
+For pairwise comparisons with large numbers of overlapping sites ($>10000$), KIN, READ and lcMLkin all mostly agree.
 However, KIN is able to differentiate between parent-child and siblings, and identify second and third degree relationships from just a few thousand sites overlapping between samples.
-We show that when applied to Neandertal specimens from Chagyrskaya cave, KIN identifies a pair of first-degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples \cite{laurits_skov_genetic_nodate}.
-In addition, KIN identifies a pair of third degree relatives.
-In this case of a population with large amounts of ROH, we find that the inference by lcMLKIN are heavily biased, but KIN's model takes ROH into account and both the coefficient of relatedness and $k_0$ are very close to what would be expected from the inference by both READ and KIN.
+We show that when applied to Neandertal specimens from Chagyrskaya cave, KIN identifies a pair of $1^{st}$-Degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples \cite{laurits_skov_genetic_nodate}.
+In addition, KIN identifies a pair of $3^{rd}$ degree relatives.
+In this case of a population with large amounts of ROH, we find that the inference by lcMLkin are heavily biased, but KIN's model takes ROH into account and both the coefficient of relatedness and $k_0$ are very close to what would be expected from the inference by both READ and KIN.
 
 One limitation of our approach is that it assumes a single population.
-In case of a highly structured population, KIN may show inaccurate inferences.
+In case of a highly structured population, KIN may show inaccurate inferences, similar to READ and lcMLkin.
 Also, our method makes the assumptions, that the median pairwise genetic difference reflects the population diversity, which fails if almost all individuals in the data set are related.
@@
 For our simulations of background diversity, we form a population (Pop1) with constant effective size of 10,000 and sample eight diploid individuals (each made up of two haploid individuals) from 2500 generations ago (fig.
-\ref{figS10:pedigree}).
+S \ref{figS10:pedigree}).
 For each individual, we simulate 22 chromosomes with length $L=100$ Mb and a recombination rate of $r=10^{-8}$ per base pair per generation.
@@
 We simulate recombination by first drawing the number of breakpoints as a  Poisson random variable with parameter $rL$, and use a uniform distribution on $[1, L]$ to sample the positions of recombination points.
+The pedigree simulations result in nine addition individuals, resulting in a total sample of 17 individuals.
 
@@
 KIN requires a transition matrix for each relatedness case.
-which we estimate  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of 1000 simulations from our pedigree (Figure SXXXX).
+which we estimate  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of 1000 simulations from our pedigree (Figure S \ref{figS10:pedigree}).
 For two cases, siblings and grandparent-grandchild, it is possible to write down the theoretical expectation of the transition matrix:
@@
 \paragraph{Simulations for method evaluation}
-Apart from the related and unrelated individuals in Pop1, we simulated more haploids in three other populations to create scenarios with ascertainment bias and contamination (fig.
+Apart from the related and unrelated individuals in Pop1, we simulated more haploid individuals in three other populations to create scenarios with ascertainment bias and contamination (fig.
 \ref{figS10:pedigree}).
-We simulated two haploids to form an individual each from two other populations (Pop2, Pop3) with split time of 3500 and 4500 generations with Pop1, and sampling time of 4000 generations and 2000 generations ago respectively.
+We simulated two individuals to form an individual each from two other populations (Pop2, Pop3) with split time of 3500 and 4500 generations with Pop1, and sampling time of 4000 generations and 2000 generations ago respectively.
 We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from Pop1.
 This scenario roughly models the ascertainment of the Chagyrskaya data.
-We tested the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a Markov chain using transition matrix shown below:
+We tested the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a Markov chain using the transition matrix
 
@@
 From the steps described above, we got genotypes of individuals in Pop1 in presence/absence of ROH and ascertainment.
-We further simulated five diploid individuals from Pop4 with split time of 20,000 generations with Pop1, sampled from the present time.
-We calculated the allele frequency for Pop4 at each site, and used it to introduce contamination (with fixed contamination percentage for each individual ($<5\%$) in samples from Pop1.
+We further simulated five diploid individuals from Pop4 with split time of 20,000 generations with Pop1, sampled from the present time, as a source of contamination.
+For cases including contamination, we contaminated eight individuals with varying amounts between 0.5\% to 3\% contamination, with the remaining  nine individuals did not have any contamination.
 We generated reads (derived/ancestral) for different genomic coverages ranging from 4x to 0.03x, assuming a Poisson distribution.
 
-We run simulations with different cases referred to as control (with no ROH, contamination or ascertainment bias), R (with ROH), A (with ascertainment bias), and C (with fixed levels of contamination for each individual).
+For testing, we replicate each simulation 60 times, and create data from the same base simulations at varying levels of coverages and different scenarios: We have a control scenario (without ascertainment bias, ROH or contamination), and individual scenarios where we add ROH, SNP ascertainment and contamination, respectively.
+For the evaluation of the ROH-HMM, we also combine ROH with ascertainment or contamination.
 
@@
 \begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/pedigree.png}
+    \includegraphics[width=18cm]{plots/inkscape_finalImg/pedigree_both.png}
     \centering
     \caption{Overview of simulated dataset.
-We simulated Pop1, and artificially mated unrelated individuals to make a family with upto 5th degree relatives, along with other unrelated and identical individuals (not shown here).
-Pop2 and Pop3 are simulated so as to introduce ascertainment, while Pop4 is a present day population used to simulate contamination.
-Split times and sampling times of the populations are shown in generations.}
+(A) We simulated four different populations with split times and sampling times (in generations) mimicking the populations of Chagyrskaya neandertals, Vindija neandertals, Altai neandertals and present day Africans \cite{mafessoni_high-coverage_2020}.
+We artificially mated individuals in Pop1 to create related individuals.
+We used Pop2 and Pop 3 to introduced ascertainment bias in the data, while we simulated contamination from modern humans using Pop4.
+(B) In Pop1, we sampled 8 diploid unrelated individuals, and artificially mated some of them to make a pedigree with upto $5^th$ Degree relatives.
+Circles represent females, and squares represent males.
+Dotted lines connect identical pair of individuals.}
     \label{figS10:pedigree}
@@
 
-*Remember to use updated lcMLKIN in the plot.
+*Remember to use updated lcMLkin in the plot.
 \begin{figure}[h!]
@@
     \includegraphics[width=18cm]{supplementary_info/plots/lcPlot.png}
-    \caption{Comparison of IBD states estimated for Chagyrskaya specimens using (A) lcMLKIN and (B) KIN.
+    \caption{Comparison of IBD states estimated for Chagyrskaya specimens using (A) lcMLkin and (B) KIN.
 Each plot shows relatedness coefficient (calculated as $k_2+k_1/2$) on y-axis plotted against $k_0$ (proportion of genome with no chromosomes shared).
@@
     \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIN differs from READ.
-(A) AITI62B-OTTM156 (READ estimates Identical individuals, lcMLKIN does not have sufficient data, and KIN predicts parent-child.
-(B) UNTA5867-UNTA5868Sk1 (READ estimates Second Degree, lcMLKIN does not have sufficient data, and KIN predicts parent-child).
-(C) AITI40-AITI72 (READ estimates Unrelated individuals, lcMLKIN shows 3rd-5th degree, and KIN predicts 3rd degree.)
-    (D) AITI2-AITI55 (READ estimates Unrelated individuals, lcMLKIN and KIN predicts predict second degree.)}
+(A) AITI62B-OTTM156 (READ estimates Identical individuals, lcMLkin does not have sufficient data, and KIN predicts parent-child.
+(B) UNTA5867-UNTA5868Sk1 (READ estimates Second Degree, lcMLkin does not have sufficient data, and KIN predicts parent-child).
+(C) AITI40-AITI72 (READ estimates Unrelated individuals, lcMLkin shows 3rd-5th degree, and KIN predicts 3rd degree.)
+    (D) AITI2-AITI55 (READ estimates Unrelated individuals, lcMLkin and KIN predicts predict second degree.)}
     \label{figS8:eg1}
@@
     \includegraphics[width=18cm]{supplementary_info/plots/egplot2.png}
-    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIN differs from lcMLKIN.
-(A) AITI43-AITI55 (READ estimates First degree, lcMLKIN predicts siblings, while KIN predicts parent-child.
-(B) AITI70-AITI72 (READ estimates First degree, lcMLKIN predicts parent-child, while KIN predicts siblings.}
+    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIN differs from lcMLkin.
+(A) AITI43-AITI55 (READ estimates First degree, lcMLkin predicts siblings, while KIN predicts parent-child.
+(B) AITI70-AITI72 (READ estimates First degree, lcMLkin predicts parent-child, while KIN predicts siblings.}
     \label{figS9:eg2}

### 2022-04-07 paper_overleaf/main.tex

@@
 \paragraph{Critical Values}
- To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff( Fig.\ref{figS10:cutoff},  Fig.\ref{fig3:Comparison_READ_KIN}).
-Fig.\ref{figS10:cutoff} shows that for all relatedness cases except for $3^{rd}$-degree, the false positive rate is below $5\%$ even when simply selecting the model with the highest likelihood.
+ To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff( fig.
+S \ref{figS10:cutoff}).
+The figure shows that for all relatedness cases except for $3^{rd}$-degree, the false positive rate is below $5\%$ even when simply selecting the model with the highest likelihood.
 We observe that for $3^{rd}$-degree relatives, using a cutoff of 1.0 brings down the false positive rate close to $5\%$ for all coverages except 0.05x.

### 2022-04-07 paper_overleaf/main.tex
Updates from Overleaf

@@
 We show that both methods have similar performance for low-coverage shotgun data (``control''-case),  and for ascertained data.
-One exception are  $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages.
-We also find that READ is strongly impacted by contamination, but that the correction implemented in KIN is sufficient to remove this bias..
+One exception is  $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages.
+We also find that READ is strongly impacted by contamination, but that the correction implemented in KIN is sufficient to remove this bias.
 For simulations with ROH, we find the overall classification accuracy is only slightly reduced.
@@
 Fig.
-\ref{fig4:IBDstate_accuracy} shows that we can predict IBD states for Unrelated, $3^{rd}$ Degree, $2^{nd} Degree$, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling.
+\ref{fig4:IBDstate_accuracy} shows that we can predict IBD states for Unrelated, $3^{rd}$ Degree, $2^{nd}$ Degree, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling.
 On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03 x.
@@
 Our estimates are consistent with those obtained using READ, except that READ is unable to detect 3rd-degree relationships, and cannot distinguish between siblings and parent-child.
-We also find that both  KIN and READ classify Chagyrskaya 06 and Chagyrskaya 14 as parent-child with low confidence, but from the morphology they likely stem from the same individual.
+We also find that both  KIN and READ classify Chagyrskaya06 and Chagyrskaya14 as parent-child with low confidence, but from the morphology they likely stem from the same individual.
 We believe that this due to non-human contamination present in these libraries ($1\%$ and $2.9\%$ respectively).
@@
 However, the location of these clusters strongly deviate from the expected values, likely due to the low coverage and ROH.
-Fossils from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $r \approx 0.6, k_0 \approx 0.33$.
+Fossils from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at k_0 \approx 0.33$, $r \approx 0.6.
 
@@
 However, most of these are at very low confidence, and there are only 31 pairs where KIN has $\Delta LL>1$.
-When we investigated some of these comparisons in more detail we found that in many cases, the proportion of differences along the genome align with the prediction by KIN (fig.
+When we investigated these comparisons in more detail we found that in many cases, the proportion of differences along the genome align with the prediction by KIN (fig.
 S\ref{figS8:eg1}).
@@
 
-In our simulations, we find that adding long ROH ($~17\%$) to simulations improves the performance of  both KIN and READ, particularly at lower coverages.
+In our simulations, we find that adding long ROH ($~17\%$) to simulations slightly improves the performance of  both KIN and READ, particularly at lower coverages.
 This is likely because ROH actually reduces the variance in differences, and so it becomes easier to estimate the location of IBD tracts.
@@
 In addition, the output of KIN is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio.
-Thus, assuming we know the recombination rate, and hence can estimate the transition matrix \ref{method}, KIN can be widely applied to any diploid species and is easy to automatise and apply to large datasets.
+Thus, assuming we know the recombination rate, and hence can estimate the transition matrix (see section \ref{method}), KIN can be widely applied to any diploid species and is easy to automatise and apply to large datasets.
 
@@
 KIN requires a transition matrix for each relatedness case.
-which we estimate  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of 1000 simulations from our pedigree (Figure S \ref{figS10:pedigree}).
+which we estimate  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of 1000 simulations from our pedigree (Fig.
+S \ref{figS10:pedigree}).
 For two cases, siblings and grandparent-grandchild, it is possible to write down the theoretical expectation of the transition matrix:
 
-For grandparent-grandchild, 
+For grandparent-grandchild, rate matrix is
  $$Q = \left[\begin{array}
@@
 
-*Remember to use updated lcMLkin in the plot.
 \begin{figure}[h!]

### 2022-04-07 supplementary_info/Rscripts/lcPlot.R
scale

@@ -40,7 +40,7 @@ lcplot <- function(lcres,likf,kinf,outplot){
 
   panel1<-ggplot(data=lc[which(lc$loglik_ratio >=1),], mapping=aes(x=k0_hat, y=r_hat, color=Relatedness)) +
   geom_point() + theme_bw() + labs(x=expression(k["0"]), y= "Coefficient of Relatedness") +
-  scale_color_manual(values = colors1) + ylim(0, 1) +
+  scale_color_manual(values = colors1) + ylim(0, 1) + xlim(0, 1) +
   theme(legend.position = "none")
 
 
@@ -71,12 +71,12 @@ lcplot <- function(lcres,likf,kinf,outplot){
   #alphas <-c("0"=0.01,"1"=1)
   panel2<-ggplot(data=newdf[which(newdf$loglik_ratio >=1),], mapping=aes(x=k0_prop, y=r_prop, color=Relatedness)) +
     geom_point() + theme_bw() + labs(x=expression(k["0"]), y= "Coefficient of Relatedness") +
-    scale_color_manual(values = colors) +  ylim(0, 1) +
+    scale_color_manual(values = colors) +  ylim(0, 1) + xlim(0, 1) +
     #scale_alpha_manual(values = alphas) +
     theme(axis.title.y = element_blank())
 
   plot_grid(panel1,panel2, labels= c("A","B"),
-            ncol=2, rel_widths = c(1,1.25), align = "hv", axis="bt")
+            ncol=2, rel_widths = c(1,1.3), align = "hv", axis="bt")
 
   ggsave(outplot,
          width = 8, height = 5, dpi = 150, units = "in", device='png')

### 2022-04-11 paper_overleaf/main.tex
Updates from Overleaf

@@
 \usepackage{lineno}
-\linenumbers
+\usepackage{natbib}
 
@@
 \begin{document}
-\nolinenumbers
 
@@
 
-Here, we present KIN, a Hidden-Markov-Model-based approach to estimate identity-by descent-fragments and degree of relatedness up to 3rd-degree, from as low as 0.05x coverage ancient DNA data, and additionally differentiates between siblings and parent-child relationships.
+Here, we present KIN, a Hidden-Markov-Model-based approach to identify identity-by-descent fragments and to estimate degree of relatedness.
+KIN can accurately determine up to 3rd-degree relatives and differentiate between sibling and parent-child relationships using ancient DNA sequence data with as little as 0.05x coverage.
 KIN also incorporates an adjustment for contamination, and an additional model to estimate the location of long runs of homozygosity, and uses it to improve classification accuracy.
@@
 
-We illustrate the utility of KIN on two applications, one on a Neandertal population from Siberia, Russia, and one in a Bronze-Age population from the Lech Valley, Germany. 
-We find that in both cases, KIN is competitive with existing approaches and provides new insights.
+We illustrate the utility of KIN on two applications, one on a Neandertal group from Siberia, Russia, and one in a Bronze-Age modern human population from the Lech Valley, Germany. 
+We find that in both cases, KIN matches the findings of existing approaches and provides new insights.
 
@@
 It is also an essential step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.
-For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to avoid inbreeding and ensure diversity ~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}.
+For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to avoid inbreeding and ensure diversity \cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}.
 
-In ancient DNA studies, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_KINship-based_2019,sikora_ancient_2017}.
+In ancient DNA studies, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_kinship-based_2019,sikora_ancient_2017}.
 
@@
 
-Commonly, pairs of related individuals are identified by looking for parts of the genome that are identical by descent (IBD), ie.
+Commonly, pairs of related individuals are identified by looking for parts of the genome that are identical by descent (IBD), i.e.
 inherited from a recent common ancestor.
-Due to the laws of Mendelian segregation, each parent, for example, will share exactly one set of chromosomes IBD with their offspring.
-A grandparent will, on average share a quarter of their genome with a grand-child due to recombination.
+Due to the laws of Mendelian segregation, each parent will share exactly one set of chromosomes IBD with their offspring, while subsequent recombination means that a grandparent will, on average share a quarter of their genome with a grand-child.
 Along the genomes of a pair of diploid individuals, there are three IBD states possible at any given position: either no chromosomes are IBD, only one chromosome is IBD, or both chromosomes are IBD.
@@
 
-However, it is not possible to directly observe IBD segments, and so a common approach is to first identify segments of the genome that are Identical by State (IBS) and using population allele frequencies obtained from a reference panel to calculate the probability of IBD given IBS.
+However, since it is not possible to directly observe IBD segments, a common approach is to first identify segments of the genome that are Identical by State (IBS) and to use population allele frequencies obtained from a reference panel to calculate the probability of IBD given IBS.
 There are several methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007,manichaikul_robust_2010}.
@@
 On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes.
-The bottom row shows the expected proportion of differences (p) along the chromosome in both the cases.
-p can take values of $p_0$, $p_1$ or $p_2$ which are expected proportion of differences in IBD states 0,1 and 2 respectively.
+The bottom row shows the expected proportion of differences (p) along the chromosome in both the cases p can take values of $p_0$, $p_1$ or $p_2$ which are the expected proportion of differences in IBD states 0,1 and 2 respectively.
 }
@@
 
-\begin{table}
+\begin{table}[h!]
 \caption{\label{tab:Table 1}IBD sharing probabilities for different relations in absence of inbreeding}
@@
 \subsection{Approaches that address problems with ancient DNA data}
-One major issue with applying the above mentioned methods to ancient DNA data is that the sequence coverage is typically low, making it difficult to get genotype calls.
+One major issue with applying the above-mentioned methods to ancient DNA data is that the sequence coverage is typically low, making it difficult to generate accurate genotype calls.
 Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}.
@@
 However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be imprecise at lower coverages \cite{korneliussen_angsd_2014} .
-Ancient DNA analyses often face additional challenges such as contamination from present-day DNA \cite{peyregne_present-day_2020-1}, absence of suitable reference panels, and ascertainment bias in the case of data generated from DNA capture \cite{prufer_computational_2010, vai_kinship_2020}.
+Ancient DNA analyses often face additional challenges such as contamination with present-day DNA \cite{peyregne_present-day_2020-1}, absence of suitable reference panels, and ascertainment bias in the case of data generated from DNA capture \cite{prufer_computational_2010, vai_kinship_2020}.
 
-Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}.
+Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of allele frequencies in the population from which the sampled individuals originate \cite{theunert_joint_2017}.
 A second issue is contamination.
-If the contamination stems from another population, contaminated data will look more dissimilar to other individuals from the analyzed population, and hence relatedness inference will be biased towards towards less relatedness.
-In addition, some analyzed  genomes  may have long runs of homozygosity (ROH), for example due to a small population size, or due to recent inbreeding.
+If the contamination stems from another population, contaminated data will look more dissimilar to other individuals from the analyzed population, and hence relatedness inference will be biased towards  less relatedness.
+In addition, some analyzed  genomes  may have long runs of homozygosity (ROH), for example due to a small population size, or recent inbreeding.
 Long ROH cause related individuals to seem genetically more similar to each other, but do not affect the genetic distance between unrelated individuals.
 
-Moreover, ancient DNA is commonly captured with a SNP array to enrich for sequences from the target species.
-Particularly methods based  on the fraction of sites in different IBS states are sensitive to the ascertainment bias caused by this non-random selection of targted sites \cite{waples_allele_2019}.
+Moreover, ancient DNA is commonly captured with a SNP array that enrich for informative variants.
+Particularly methods based on the fraction of sites in different IBS states are sensitive to the ascertainment bias caused by this non-random selection of targetted sites \cite{waples_allele_2019}.
 
-Several limitations with ancient DNA mentioned above, such as low coverage, the absence of population allele frequencies or a reference panel, and ascertainment bias are resolved by a software called READ \cite{kuhn_estimating_2018}.
-This software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual.
+READ \cite{kuhn_estimating_2018} addresses several of the issues encountered in the analysis of ancient DNA.
+In particular, the lack of  genotype calls is dealt with by randomly sampling alleles from each individual.
 A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to calculate average pairwise genetic distances, which in turn are used to infer relatedness.
@@
 Here, we present KIN (Kinship INference), a Hidden Markov Model (HMM)-based approach to estimate relatedness and IBD from low-coverage ancient DNA data.
-KIN can detect up to \nth{3}-degree relatives, and differentiates between siblings and parent-child.
+KIN can detect up to \nth{3}-degree relatives, and differentiates between siblings and parent-child relationships.
 KIN is also able to take into account ROH, contamination and ascertainment.
@@
 
-To infer the case of relatedness, KIN fits one HMM for each pair of individual and for each relatedness case.
+To infer the case of relatedness, KIN fits one HMM for each pair of individuals and for each possible relatedness case.
 The KIN-HMM infers IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts and contamination estimates in each individual into account.
@@
 If the locations of ROH tracts are unknown, we provide another HMM (ROH-HMM) to coarsely estimate the location of ROH for samples with sufficient coverage ($\geq 0.1x$).
-Our method is available on \url{https://github.com/DivyaratanPopli/KIN_snakemake} along with a \textt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.
+Our method is available on \url{https://github.com/DivyaratanPopli/KIN_snakemake} along with a \texttt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.
 
@@
 We then compare the likelihood between all models, and classify each pair to the model with the highest likelihood.
-We also return the most likely location of IBD tracts (using standard Viterbi algorithm), and the state posterior probablities $\BZ$.
+We also return the most likely locations of IBD tracts (using standard Viterbi algorithm), and the state posterior probabilities $\BZ$.
 
 The details of the likelihood computation are given in section \ref{ll}.
-As any HMM, KIN-HMM requires a set of emission (section \ref{B}) and transition probability matrices.
+As is the case for any HMM, KIN-HMM requires a set of emission (section \ref{B}) and transition probability matrices.
 We assume the transition matrix for each relatedness case is fixed.
@@
 For high-coverage data, this can be estimated accurately by comparing genotypes, but for low-coverage samples, $D_w$ needs to be estimated.
-The simplest approach is is to  randomly-sample a read from each position to make "pseudo-haploid" sequences \cite{haak_massive_2015}.
+The simplest approach is to  randomly-sample a read from each position to make "pseudo-haploid" sequences \cite{haak_massive_2015}.
 However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
@@
 
+Here, $\BZ$ is not dependent on $\BH$, $\theta$ and $\delta$.
+
 \subsection{Emission probability}\label{B}
@@
 For example, when considering two bones from the same individual, we would expect the entire genome to have a pairwise difference of $p_2$, because two out of the four compared chromosomes are between identical copies.
-However, in inbred regions all four chromosomes will be identical, and so the expected pairwise differences are zero.
+However, in inbred regions all four chromosomes will be identical, and so the expected pairwise differences are zero ($p_3$ in equation \ref{eq:2}).
 Note however, that $p_0$ does not depend on the presence of ROH, since all comparisons are between unrelated chromosomes even if both individuals are homozygous at a particular locus.
@@
 As explained above, $p_3 = 0$.
-However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that we obtain better results by setting $p_3 = \frac{p_2}{2}$).
+However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that we obtain better results by setting $p_3 = \frac{p_2}{2}$.
 
@@
 S\ref{figS1:binom}).
-We take this into account by adding an overdispersion parameter $\delta$.Just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ depends on the number of chromosomes compared and so each of the  four $p_i$ has a corresponding $\delta_i$ parameter
+We take this into account by adding an overdispersion parameter $\delta$.
+Just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ depends on the number of chromosomes compared, and so each of the  four $p_i$ has a corresponding $\delta_i$ parameter.
 
@@
 \end{align}
-where $i$ is fully determined by the combination of $Z_w$ and $H_w$ (see eq 3).
+where $i$ is fully determined by the combination of $Z_w$ and $H_w$ (see equation \ref{eq:2}).
 
@@
 \subsection{Estimation of $\delta$}\label{delta}
-We estimate the $\delta$-parameters using an Expectation-Maximization (EM) algorithm (REF). In an EM-algorithm, we iterate the Expecation and Maximiation Steps until the algorithm converges.
+We estimate the $\delta$-parameters using an Expectation-Maximization (EM) algorithm (REF). In an EM-algorithm, we iterate the Expectation and Maximization steps until the algorithm converges.
 
@@
 \mathcal{C} &= \mathbb{E}[\log P(D_w|Z_w, H_w, \theta, \delta^{(t-1)}]\nonumber\\
-&= \sum_{w=1}^L \sum_{j=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta_{j \omega}^{(t-1)}, p_{j \omega})h_{w\omega}\gamma_{wj}
+&= \sum_{w=1}^L \sum_{j=0}^2\sum^2_{\omega=0} \log P(D_{w}|N_w, \delta_{j \omega}^{(t-1)}, p_{j \omega})h_{w\omega}\gamma_{wj}
 \end{align}
@@
 \begin{align}\label{eq:6}
-\mathcal{C} &= \sum_{i=0}^3 \sum_{w=1}^L \log P(D_{w}|Z_w, H_w, N_w, \delta_i, p_i)g_{wi}
+\mathcal{C} &= \sum_{i=0}^3 \sum_{w=1}^L \log P(D_{w}|N_w, \delta_i, p_i)g_{wi}
 \end{align}
@@
 \subsection{Model comparison}\label{model_comp}
-To infer the most probable relatedness case, we run our model on all cases (section \ref{method_overview}) and compare the resulting likelihoods.
+To infer the most probable relatedness case, we run our model on all relatedness cases (section \ref{method_overview}) and compare the resulting likelihoods.
 We output the relatedness corresponding to the maximum likelihood model, and the confidence as given by the log likelihood ratio between the two highest likelihood models.
@@
 
-We assume that contamination rates in both individuals are known and small, and set $C_{ij} \approx C_i + C_j$, where $C_i, C_j$ are the contamination estimates from the two individuals.
+We assume that contamination rates in both individuals are known and small ($<5\%$), and set $C_{ij} \approx C_i + C_j$, where $C_i, C_j$ are the contamination estimates from the two individuals.
 We also assume the divergence $\phi$ between our target population and the putative contaminant is known.
@@
 \paragraph{Motivating Example}
-In Fig.\ref{fig1:ibd} we show the output of different relatedness cases when they are applied to data from a pair of siblings.
+In Fig.\ref{fig1:ibd} we show the inferred IBD fragments from different KIN-HMMs when they are applied to simulated data from a pair of siblings.
+In this case, the siblings-model predictions match true IBD states the most, as reflected by the highest log-likelihood.
 
@@
     \centering
-    \caption{\textbf{Comparison of pairwise difference data  and inferred IBD fragments}:  The top panel shows the proportion of differences in each window along the genome for a pair of simulated siblings.
+    \caption{Comparison of pairwise difference data and inferred IBD fragments.
+The top panel shows the proportion of differences in each window along the genome for a pair of simulated siblings.
 Dashed lines represent $p_0$, $p_1$ and $p_2$ estimates.
@@
 
-In this case, the siblings-model predictions match true IBD states the most, as reflected by the highest log-likelihood.
+\paragraph{Overview of simulations} 
+We simulated an ancient population (sampled from 2500 generations ago) with msprime \cite{kelleher_efficient_2016}, and artificially mated some of the individuals to form a pedigree (figure \ref{figS10:pedigree}).
+We downsampled this data to low coverages (4$x$ - 0.03$x$), and applied KIN to find out a suitable cut-off for a reliable log likelihood ratio.
+We also introduced ascertainment bias and contamination from modern humans (see section \ref{method}) to test the performance of KIN, and compare it to that of READ.
+\paragraph{Critical Values}
 
-\paragraph{Critical Values}
  To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff( fig.
@@
 One exception is  $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages.
-We also find that READ is strongly impacted by contamination, but that the correction implemented in KIN is sufficient to remove this bias.
+We also find that READ is strongly impacted by contamination ($<3\%$), but that the correction implemented in KIN is sufficient to remove this bias.
 For simulations with ROH, we find the overall classification accuracy is only slightly reduced.
@@
 \paragraph{Evaluating IBD prediction}
-KIN-HMM co-estimates IBD states and relatedness, and hence it has a higher power to classify $2^{nd}$ Degree relatives, and can distinguish between Siblings and Parent-Child.
+KIN co-estimates IBD states and relatedness, and hence it has a higher power to classify $2^{nd}$ degree relatives, and can distinguish between Siblings and Parent-Child.
 We investigate our limit to predict IBD states along the genome by calculating the number of genomic windows where we correctly  predict IBD state for a pair of individuals.
@@
 \ref{fig4:IBDstate_accuracy} shows that we can predict IBD states for Unrelated, $3^{rd}$ Degree, $2^{nd}$ Degree, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling.
-On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03 x.
+On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03x coverage depth.
 
@@
 \subsubsection{Application to Chagyrskaya Neanderthals}
-To test KIN on real ancient dataset, we applied it to a Neandertal data set  from Chagyrskaya cave in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020}.
-This dataset contains genetic data from  14 fossils that might  possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya \cite{laurits_skov_genetic_nodate}.
-The data has low-to-intermediate coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage.
-Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena \cite{laurits_skov_genetic_nodate}.
-All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes and common variations in Africans \cite{laurits_skov_genetic_nodate}, and we further restricted  this data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}.
+To test KIN on real ancient data, we applied it to a Neandertal data set  from Chagyrskaya Cave in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020}.
+This dataset contains genetic data from  14 skeletal remains that might  possibly belong to contemporary Neandertals who occupied the cave from 63 ± 4 to 48 ± 3 kya \cite{laurits_skov_genetic_nodate}.
+The data has low-to-intermediate depth of coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage.
+Some of these specimens showed signs of long ROH, and contamination from modern humans as well as hyenas \cite{laurits_skov_genetic_nodate}.
+All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes, and common variations in Africans \cite{laurits_skov_genetic_nodate}.
+We further restricted this data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017} to make the data for our relatedness analysis same as that used by the authors.
 Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}.
@@
 We also find that both  KIN and READ classify Chagyrskaya06 and Chagyrskaya14 as parent-child with low confidence, but from the morphology they likely stem from the same individual.
-We believe that this due to non-human contamination present in these libraries ($1\%$ and $2.9\%$ respectively).
+We believe that the low confidence miss-classification arises  due to non-human contamination present in these libraries ($1\%$ and $2.9\%$ respectively \cite{laurits_skov_genetic_nodate}), biasing the estimated differences between the individuals to higher values.
 
@@
 However, the location of these clusters strongly deviate from the expected values, likely due to the low coverage and ROH.
-Fossils from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at k_0 \approx 0.33$, $r \approx 0.6.
+Remains from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $k_0 \approx 0.33$, $r \approx 0.6$.
 
@@
 
-We further applied KIN to a genome-wide dataset of 118 ancient individuals from the Lech valley \cite{mittnik_kinship-based_2019}.
-We compared our relatedness estimates to those obtained from the authors using READ and lcMLkin, and found that in most cases the three methods agree.
+We further applied KIN to a genome-wide dataset of 118 ancient individuals from the Lech Valley \cite{mittnik_kinship-based_2019}.
+We compared our relatedness estimates to those obtained by the authors using READ and lcMLkin, and found that in most cases the three methods agree.
 Out of 6903 pairwise comparisons, 346 pairs show different relatedness for KIN and READ (table S2).
@@
 \section{Discussion}\label{discussion}
-Here, we present a new method called KIN to estimate genetic kinship and the location of IBD tracts from low-coverage data, taking into account long ROH, contamination, as well as ascertainment bias.
+Here, we present a new method called KIN to estimate genetic kinship and the location of IBD tracts from low-coverage data, in presence of long ROH, contamination, as well as ascertainment bias.
 Our method utilizes a set of HMMs to estimate IBD tracts and uses them to classify each pair of individuals into a possible relatedness case, along with a measure of classification confidence.
@@
 
-The Lech valley data  has low contamination, and no ROH.
+The Lech Valley data  has low contamination, and no ROH.
 For pairwise comparisons with large numbers of overlapping sites ($>10000$), KIN, READ and lcMLkin all mostly agree.
 However, KIN is able to differentiate between parent-child and siblings, and identify second and third degree relationships from just a few thousand sites overlapping between samples.
-We show that when applied to Neandertal specimens from Chagyrskaya cave, KIN identifies a pair of $1^{st}$-Degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples \cite{laurits_skov_genetic_nodate}.
+We show that when applied to Neandertal specimens from Chagyrskaya Cave, KIN identifies a pair of $1^{st}$-Degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples \cite{laurits_skov_genetic_nodate}.
 In addition, KIN identifies a pair of $3^{rd}$ degree relatives.
@@
 
-While we have focused on the application of KIN on ancient human samples, but the model is not tied to this system.
+While we have focused on the application of KIN on ancient human samples, the model is not tied to this system.
+Assuming we know the recombination rate, and hence can estimate the transition matrix (see section \ref{method}), KIN can be widely applied to any diploid species.
 In addition, the output of KIN is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio.
-Thus, assuming we know the recombination rate, and hence can estimate the transition matrix (see section \ref{method}), KIN can be widely applied to any diploid species and is easy to automatise and apply to large datasets.
+This makes KIN easy to automatize for large datasets.
+To make application of KIN user-friendly, we provide a python package to create input files for KIN from processed bamfiles, while optionally estimating ROH, and correcting for contamination estimates.
 
@@
 
+\renewcommand\thefigure{\arabic{figure}}    
 \section{Supplementary Information}
+\setcounter{figure}{0}
 
 \begin{table}
-\caption{\label{tab:Table 2}Model performance for ROH prediction}
+\caption{\label{tab:Table 2}Model performance for ROH prediction.
+Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination, R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
 \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
@@
     \centering
-    \caption{Overall schematic of the method showing the entire pipeline from processed bam files to final relatedness and IBD estimates .
+    \caption{Overall schematic of the method showing the entire pipeline from processed bam files to final relatedness and IBD estimates.
 Blue boxes show the data files, while green boxes represent scripts.}
@@
     \caption{Overview of simulated dataset.
-(A) We simulated four different populations with split times and sampling times (in generations) mimicking the populations of Chagyrskaya neandertals, Vindija neandertals, Altai neandertals and present day Africans \cite{mafessoni_high-coverage_2020}.
+(A) We simulated four different populations with split times and sampling times (in generations) mimicking the populations of Chagyrskaya Neandertals, Vindija Neandertals, Altai Neandertals and present day Africans \cite{mafessoni_high-coverage_2020}.
 We artificially mated individuals in Pop1 to create related individuals.
-We used Pop2 and Pop 3 to introduced ascertainment bias in the data, while we simulated contamination from modern humans using Pop4.
+We used Pop2 and Pop3 to introduced ascertainment bias in the data, while we simulated contamination from modern humans using Pop4.
 (B) In Pop1, we sampled 8 diploid unrelated individuals, and artificially mated some of them to make a pedigree with upto $5^th$ Degree relatives.
@@
     \centering
-    \caption{Application of READ on Neandertal specimens from Chagyrskaya cave.
+    \caption{Application of READ on Neandertal specimens from Chagyrskaya Cave.
 Color of a square represents the relatedness, while the number denotes standard deviations away from the upper threshold (we show lower threshold for unrelated pairs since upper threshold is not available).}
@@
     \centering
-    \caption{False positive and True Positive rates as a function of cutoff on log likelihood ratio.
+    \caption{False positive and true positive rates as a function of cutoff on log likelihood ratio.
 In top panel, dotted line represents $5\%$.
@@
     \includegraphics[width=18cm]{supplementary_info/plots/degree2_10Mwin.png}
-    \caption{Ploting total number of IBD fragments with the number of IBD fragments can tease apart avuncular relationship from grandparent-grandchild, since they form different clusters.
+    \caption{Plotting total number of IBD fragments with the number of IBD fragments can tease apart avuncular relationship from grandparent-grandchild, since they form different clusters.
 Half-siblings, however, overlap considerably with both these relations.}

### 2022-04-25 paper_overleaf/main.tex
Updates from Overleaf

@@
 \title{KIN: A method to infer relatedness from low-coverage ancient DNA}
-\author{Divya Ratan Popli, Stéphane Peyrégne, Benjamin M. Peter}
+\author{Divyaratan Popli, Stéphane Peyrégne, Benjamin M. Peter}
 %\date{5 August 2021}
@@
 
-\noindent Inferring familial relationships is an important question in population genetics.
+\noindent Inferring familial relationships is important in population genetics.
 Knowing genetic kinship of ancient individuals leads to insights into the culture and social hierarchy of ancient populations, and is relevant for downstream genetic analyses.
@@
 KIN can accurately determine up to 3rd-degree relatives and differentiate between sibling and parent-child relationships using ancient DNA sequence data with as little as 0.05x coverage.
-KIN also incorporates an adjustment for contamination, and an additional model to estimate the location of long runs of homozygosity, and uses it to improve classification accuracy.
+KIN also incorporates an adjustment for contamination, and an additional model to estimate the location of long runs of homozygosity, and uses them to improve classification accuracy.
 
@@
 
-We illustrate the utility of KIN on two applications, one on a Neandertal group from Siberia, Russia, and one in a Bronze-Age modern human population from the Lech Valley, Germany. 
-We find that in both cases, KIN matches the findings of existing approaches and provides new insights.
+We illustrate the utility of KIN on two applications, one on a Neandertal group from Siberia, Russia, and one in a group of Bronze-Age people from the Lech Valley, Germany. 
+We find that in both cases, KIN matches the findings of existing approaches and detects several additional related individuals.
 
@@
 
-However, since it is not possible to directly observe IBD segments, a common approach is to first identify segments of the genome that are Identical by State (IBS) and to use population allele frequencies obtained from a reference panel to calculate the probability of IBD given IBS.
-There are several methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007,manichaikul_robust_2010}.
+However, since it is not possible to directly observe IBD segments, a common approach is to first identify segments of the genome that are Identical by State (IBS) and to use population allele frequencies obtained from an out-of-sample reference panel to calculate the probability of IBD given IBS.
+There are several methods that incproporate reference panel allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007,manichaikul_robust_2010,gusev_whole_2009,nyerki_optimized_2022}.
 
@@
     \centering
-    \caption{IBD sharing between siblings without and with ROH.
+    \caption{IBD sharing between siblings without and with runs of homozygosity (ROH).
 In top row, recombined chromosomes inherited from parents by a pair of siblings are shown.
 On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes.
-The bottom row shows the expected proportion of differences (p) along the chromosome in both the cases p can take values of $p_0$, $p_1$ or $p_2$ which are the expected proportion of differences in IBD states 0,1 and 2 respectively.
+The bottom row shows the expected proportion of differences ($p$) along the chromosome in both the cases $p$ can take values of $p_0$, $p_1$ or $p_2$ which are the expected proportion of differences in IBD states 0,1 and 2 respectively.
 }
@@
 \subsection{Approaches that address problems with ancient DNA data}
-One major issue with applying the above-mentioned methods to ancient DNA data is that the sequence coverage is typically low, making it difficult to generate accurate genotype calls.
+One major issue with applying the above-mentioned methods to ancient DNA data is that the sequence coverage is typically low, making it difficult to obtain accurate genotype calls.
 Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}.
-In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods.
+In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes, weighted by their genotype likelihoods.
 However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be imprecise at lower coverages \cite{korneliussen_angsd_2014} .
-Ancient DNA analyses often face additional challenges such as contamination with present-day DNA \cite{peyregne_present-day_2020-1}, absence of suitable reference panels, and ascertainment bias in the case of data generated from DNA capture \cite{prufer_computational_2010, vai_kinship_2020}.
+Ancient DNA analyses often face additional challenges such as contamination with present-day DNA \cite{peyregne_present-day_2020-1}, the unavailability of  reference panels to estimate population allele frequencies, and an ascertainment bias caused by  DNA capture approaches \cite{prufer_computational_2010, vai_kinship_2020}.
 
-Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of allele frequencies in the population from which the sampled individuals originate \cite{theunert_joint_2017}.
+Several methods have been proposed to estimate relatedness without a reference panel from ancient DNA, but they  require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of allele frequencies in the population from which the sampled individuals originate \cite{theunert_joint_2017}.
 A second issue is contamination.
-If the contamination stems from another population, contaminated data will look more dissimilar to other individuals from the analyzed population, and hence relatedness inference will be biased towards  less relatedness.
+If the contamination stems from another population, contaminated data will look more dissimilar to other individuals from the analyzed population, and hence relatedness will be underestimated.
 In addition, some analyzed  genomes may have long runs of homozygosity (ROH), for example due to a small population size, or recent inbreeding.
@@
 
-Moreover, ancient DNA is commonly captured with a SNP array that enrich for informative variants.
-Particularly methods based on the fraction of sites in different IBS states are sensitive to the ascertainment bias caused by this non-random selection of targetted sites \cite{waples_allele_2019}.
+Moreover, ancient DNA is commonly captured with a SNP array that enriches for informative variants.
+Particularly methods based on the fraction of sites in different IBS states are sensitive to the ascertainment bias caused by this non-random selection of targeted sites \cite{waples_allele_2019}.
 
@@
 In particular, the lack of  genotype calls is dealt with by randomly sampling alleles from each individual.
-A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to calculate average pairwise genetic distances, which in turn are used to infer relatedness.
+A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to estimate  average pairwise genetic distances, which in turn are used to infer relatedness.
 However, READ only infers the degree of relatedness, and only up to second degree.
@@
 \subsection{How our method works}
-Here, we present KIN (Kinship INference), a Hidden Markov Model (HMM)-based approach to estimate relatedness and IBD from low-coverage ancient DNA data.
+Here, we present KIN (Kinship INference), a Hidden Markov Model (HMM)-based approach to estimate genetic kinship and IBD from low-coverage ancient DNA data.
 KIN can detect up to \nth{3}-degree relatives, and differentiates between siblings and parent-child relationships.
@@
 \section{New Approaches}\label{new_approaches}
-
 To infer the case of relatedness, KIN fits one HMM for each pair of individuals and for each possible relatedness case.
@@
 \subsection{Model description}\label{method_overview} 
-
 The goal of KIN-HMM is to infer how two individuals are related via the patterns of shared IBD-states along a pair of genomes.
@@
 For high-coverage data, this can be estimated accurately by comparing genotypes, but for low-coverage samples, $D_w$ needs to be estimated.
-The simplest approach is to  randomly-sample a read from each position to make "pseudo-haploid" sequences \cite{haak_massive_2015}.
+The simplest approach is to  randomly-sample a read from each position \cite{haak_massive_2015,prufer_complete_2014}.
 However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
@@
 The proportion of differences between unrelated individuals $p_0$ is an important parameter.
-We follow READ \cite{kuhn_estimating_2018} and estimate $p_0$ as the median of differences for all possible pairs of individuals, which should work well if the majority of individuals in the samples are unrelated.
+We follow READ \cite{kuhn_estimating_2018} and estimate $p_0$ as the median of differences for all possible pairs of individuals, which works well if the majority of individuals in the sample are unrelated.
 
-The presence of inbreeding adds an additional complication, as the number of shared chromosome may change.
-For example, when considering two bones from the same individual, we would expect the entire genome to have a pairwise difference of $p_2$, because two out of the four compared chromosomes are between identical copies.
+The presence of long tracts of homozygosity resulting from recent inbreeding adds an additional complication, as the number of shared chromosomes may be overestimated \cite{weir_genetic_2006}.
+For example, when considering two bones from the same individual, we would expect the entire genome to have a pairwise difference of $p_2$, because two out of the four compared chromosomes are identical copies.
 However, in inbred regions all four chromosomes will be identical, and so the expected pairwise differences are zero ($p_3$ in equation \ref{eq:2}).
@@
 
-As explained above, $p_3 = 0$.
+As explained above, we would expect $p_3$ to be zero.
 However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that we obtain better results by setting $p_3 = \frac{p_2}{2}$.
@@
 S\ref{figS1:binom}).
-We take this into account by adding an overdispersion parameter $\delta$.
+We take this into account by adding an over-dispersion parameter $\delta$.
 Just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ depends on the number of chromosomes compared, and so each of the  four $p_i$ has a corresponding $\delta_i$ parameter.
@@
 
-This parameterization of the beta distribution in terms of expected value $p$ and overdispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parametrization in terms of $\alpha$ and $\beta$.
+This parameterization of the beta distribution in terms of expected value $p$ and over-dispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parameterization in terms of $\alpha$ and $\beta$.
 We use this equation even if preprocessing steps result in non-integer $D_w$ and $N_w$, in which case we approximate the binomial coefficient using Gamma functions.
@@
 \paragraph{Expectation step}
-In the $t$-th iteration, we calculate the posterior probability of each IBD state in each window $\gamma^{(t)}_{wi} = P(Z_w=i | D_w, H_w, \theta_i, \delta_i^{(t)})$ using the forward-backward algorithm, where $\delta_i^{(t)}$ is the current guess for $\delta$ for a given IBD state.
+In the $t$-th iteration, we calculate the posterior probability of each IBD state in each window $\gamma^{(t)}_{wi} = P(Z_w=i | D_w, H_w, \theta_i, \delta_i^{(t)})$ using the forward-backward algorithm, where $\delta_i^{(t)}$ is the current estimate for $\delta$ for a given IBD state.
 
 \paragraph{Maximization step}
-The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_i$.
+The only free parameters we estimate in the maximization step are the over-dispersion parameters $\delta_i$.
 We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wj}$ and optionally the ROH state-probabilities $h_{w\omega}$ obtained from the ROH-HMM.
@@
 
-Unconstrained optimization of the $\delta_i$ could result in some confounding of cases, We realized that different cases of relatedness have different number of IBD states possible.
+Unconstrained optimization of the $\delta_i$ could result in some confounding of cases, We know that different cases of relatedness have different numbers of IBD states possible.
 For example, siblings may have all three IBD states present while a Parent-Child has only $Z_w=1$.
@@
 S \ref{figS4:bndsbeta}).
-We avoid this problem by constraining the $\delta$ such that the beta distributions for the different cases overlap by at most one standard deviation.
+We avoid this problem by constraining the $\delta$ such that the beta distributions for the different relatedness cases overlap by at most one standard deviation.
 
 \subsection{Model comparison}\label{model_comp}
-To infer the most probable relatedness case, we run our model on all relatedness cases (section \ref{method_overview}) and compare the resulting likelihoods.
-We output the relatedness corresponding to the maximum likelihood model, and the confidence as given by the log likelihood ratio between the two highest likelihood models.
-The log likelihood-ratio is commonly used in likelihood-ratio tests to compare nested models.
-However, in our case, models are not nested, and so standard likelihood-ratio theory does not apply.
-For this reason, we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig S \ref{figS10:cutoff}).
+To infer the most probable relatedness case, we run our model on all relatedness cases mentioned in section \ref{method_overview}, and compare the resulting likelihoods.
+We output the relatedness corresponding to the maximum likelihood model.
+Because the models we test are not nested, standard likelihood-ratio theory cannot be used to generate a confidence value.
+Instead we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig S \ref{figS10:cutoff}).
 
-\paragraph{Grouping of cases}
+\paragraph{Grouping of cases:}
 Particularly for low-quality data, we may not be able to distinguish all cases.
@@
 
+\paragraph{Critical Values:}
+
+To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff (fig.
+S \ref{figS10:cutoff}).
+The figure shows that for all relatedness cases except for $3^{rd}$-degree, the false positive rate is below $5\%$ even when simply selecting the model with the highest likelihood.
+We observe that for $3^{rd}$-degree relatives, using a cutoff of 1.0 brings down the false positive rate close to $5\%$ for all coverages except 0.05x.
+Thus, we  recommended using a cutoff of 1 for all cases where ROH and contamination are not a concern.
+
+\paragraph{Example case of siblings:}
+In Fig.\ref{fig1:ibd} we show the inferred IBD fragments from different KIN-HMMs when they are applied to simulated data from a pair of siblings.
+In this case, the siblings-model predictions match true IBD states the most, as reflected by the highest log-likelihood.
+
+\begin{figure}[h!]
+    \includegraphics[width=16cm]{plots/plotimg/IBDplot.png}
+    \centering
+    \caption{Comparison of pairwise difference data and inferred IBD fragments.
+The top panel shows the proportion of differences in each window along the genome for a pair of simulated siblings.
+Dashed lines represent $p_0$, $p_1$ and $p_2$ estimates.
+The second panel shows the true IBD state for each window.
+The remaining panels show the IBD states predicted by particular relatedness models.
+The log likelihood value for each model is shown on upper left corner of the panel.
+Light and shaded background represent distinct chromosomes.}
+    \label{fig1:ibd}
+\end{figure}
+
 \subsection{ROH estimation model}\label{roh}
-Our HMM to detect ROH tracts works similar to the KIN-HMM described above, but we only consider one individual at a time, and only positions covered by at least two reads.
+Our HMM to detect ROH tracts works similarly to the KIN-HMM described above, but we only consider one individual at a time, and only consider positions covered by at least two reads.
 For each site, we calculate the proportion of reads that carry different alleles, and sum them up in windows along the genome.
 We call the vector with the number of differences $\mathbf{\Delta}$, and the vector with the number of sites with at least two reads $\mathbf{N}$.
-Our model has two possible hidden states: homozygous state ($Y_w=3$), and non homozygous state ($Y_w=2$).
+Our model has two possible hidden states: homozygous state ($Y_w=3$), and non-homozygous state ($Y_w=2$).
 As above, we collect the hidden states in a vector $\mathbf{Y} = (Y_1, \dots Y_w, \dots, Y_L)$.
@@
 The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w=3$ or the state $Y_w=2$ in each window.
-The M-step for emissions in analogous to that in the KIN model (eq.
+The maximization step for emissions is analogous to that in the KIN model (eq.
 \ref{eq:4}), and the optimization step here is done with the following cost function:
@@
 \end{align}
-where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and overdispersion parameter $\delta$ similar to eq. \ref{eq:3}.
+where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and over-dispersion parameter $\delta$ similar to eq. \ref{eq:3}.
 
@@
 \subsection{Contamination correction}\label{contam}
-Contamination by the DNA from present-day people is a common feature in human ancient DNA \cite{peyregne_present-day_2020}.
-To address this issue, we developed a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.
+Contamination by DNA from present-day people is a common feature of human ancient DNA datasets \cite{peyregne_present-day_2020}.
+To address this issue, we developed a heuristic that adjusts both $D_w$ and $N_w$ to minimize the influence of contamination on the relatedness inference.
 
 We assume that contamination rates in both individuals are known and small ($<5\%$), and set $C_{ij} \approx C_i + C_j$, where $C_i, C_j$ are the contamination estimates from the two individuals.
-We also assume the divergence $\phi$ between our target population and the putative contaminant is known.
-With probability $C_{ij}$, a comparison between two random reads from either individual will contain a contaminant read, and thus contain a difference with probability $\phi$, and with probability $1 - C_{ij}$ it will be between endogenous ones.
+We also assume the divergence $\phi$ between our
+target population and the putative contaminant popoulation is known.
+With probability $C_{ij}$, a comparison between two random reads from the pair of tested individual will contain a contaminant read, and thus contain a difference with probability $\phi$, and with probability $1 - C_{ij}$ it will be between endogenous ones.
 The comparisons between two contaminant reads are ignored, which is why we assume $C_{ij}$ to be small.
@@
 
-We predict different cases of relatedness with KIN, and test it's performance on simulated pedigrees.
-We use these simulations to find the limits of KIN, and compare it to READ and lcMLkin \cite{lipatov_maximum_2015}, two existing methods.
+To test the performance of KIN, and compare it to a commonly used method, READ \cite{kuhn_estimating_2018}, we simulated an ancient population (sampled from 2500 generations ago) with msprime \cite{kelleher_efficient_2016}, and artificially mated some of the individuals to form a pedigree (figure \ref{figS10:pedigree}).
+We downsampled this data to low coverages (4$x$ - 0.03$x$), and introduced ascertainment bias and contamination from modern humans (see section \ref{method}) to understand how these factors affect the performance of KIN and READ.
 Finally we apply KIN to published data and show that it can give reliable information about genetic kinship.
-
-\paragraph{Motivating Example}
-In Fig.\ref{fig1:ibd} we show the inferred IBD fragments from different KIN-HMMs when they are applied to simulated data from a pair of siblings.
-In this case, the siblings-model predictions match true IBD states the most, as reflected by the highest log-likelihood.
-
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/plotimg/IBDplot.png}
-    \centering
-    \caption{Comparison of pairwise difference data and inferred IBD fragments.
-The top panel shows the proportion of differences in each window along the genome for a pair of simulated siblings.
-Dashed lines represent $p_0$, $p_1$ and $p_2$ estimates.
-The second panel shows the true IBD state for each window.
-The remaining panels show the IBD states predicted by particular relatedness models.
-The log likelihood value for each model is shown on upper left corner of the panel.
-Light and shaded background represent distinct chromosomes.}
-    \label{fig1:ibd}
-\end{figure}
-
-\paragraph{Overview of simulations} 
-We simulated an ancient population (sampled from 2500 generations ago) with msprime \cite{kelleher_efficient_2016}, and artificially mated some of the individuals to form a pedigree (figure \ref{figS10:pedigree}).
-We downsampled this data to low coverages (4$x$ - 0.03$x$), and applied KIN to find out a suitable cut-off for a reliable log likelihood ratio.
-We also introduced ascertainment bias and contamination from modern humans (see section \ref{method}) to test the performance of KIN, and compare it to that of READ.
-\paragraph{Critical Values}
-
- To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff( fig.
-S \ref{figS10:cutoff}).
-The figure shows that for all relatedness cases except for $3^{rd}$-degree, the false positive rate is below $5\%$ even when simply selecting the model with the highest likelihood.
-We observe that for $3^{rd}$-degree relatives, using a cutoff of 1.0 brings down the false positive rate close to $5\%$ for all coverages except 0.05x.
-Thus, we  recommended to use a cutoff of 1 for all cases where ROH or contamination are not a concern.
 
@@
 We next evaluate the performance of the ROH-HMM( Fig.\ref{fig2:ROH}).
-We observe that as coverage decreases, the raw heterozygosity estimates become more noisy.
-In the case without ROH (Fig \ref{fig2:ROH}A), we correctly infer that there is no ROH for all but two windows.
+We observe that as coverage decreases, the raw heterozygosity estimates become noisier.
+In the case without ROH (Fig \ref{fig2:ROH}A), we correctly infer that there is no ROH for all  windows (except for the case of 0.1x coverage where we miss-classify 2 windows out of 220).
 We observe that the model predicts ROH probabilities well in all cases, even at a coverage of 0.1x (Table \ref{tab:Table 2}).
@@
 
-\paragraph{Comparing KIN-HMM with READ}
-In Figure Fig.
-\ref{fig3:Comparison_READ_KIN}, we compare the classification accuracy of KIN (cutoff: 1 log-likelihood unit) to that of READ (cutoff: 1 standard deviation)  under four different simulation scenarios.
+\paragraph{Evaluating IBD prediction}
+We investigate our limit to predict IBD states along the genome by calculating the number of genomic windows where we correctly  predict IBD state for a pair of individuals.
+We calculate mean and standard deviation for this accuracy over 60 simulations of a pedigree (see section \ref{simulat}).
+Fig.
+\ref{fig4:IBDstate_accuracy} shows that we can predict IBD states for Unrelated, $3^{rd}$ Degree, $2^{nd}$ Degree, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling.
+On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03x coverage depth.
+
+\begin{figure}[h!]
+    \includegraphics[width=16cm]{plots/plotimg/plot_IBDaccuracy.png}
+    \centering
+    \caption{Evaluation of IBD estimation at different coverages.
+y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-Here,Here,axis.
+The error bars are drawn at 1 standard deviation from the mean.
+Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
+    \label{fig4:IBDstate_accuracy}
+\end{figure}
+
+\paragraph{Evaluation of KIN on simulations and comparison to READ}
+We next compare the classification accuracy of KIN (cutoff: 1 log-likelihood unit) to that of READ (cutoff: 1 standard deviation) under four different simulation scenarios (Fig.
+\ref{fig3:Comparison_READ_KIN}).
 We show that both methods have similar performance for low-coverage shotgun data (``control''-case), and for ascertained data.
@@
 For simulations with ROH, we find the overall classification accuracy is only slightly reduced.
-Since ROH makes individuals more similar, we see a bias towards more false positives, for first and second-degree relationships, which is slightly higher for READ than for KIN.
+Since ROH makes individuals more similar, we see a bias towards more false positives, for first and second-degree relationships, which is slightly higher for READ than for KIN (for details, see table S4).
 
@@
 
-\paragraph{Evaluating IBD prediction}
-KIN co-estimates IBD states and relatedness, and hence it has a higher power to classify $2^{nd}$ degree relatives, and can distinguish between Siblings and Parent-Child.
-We investigate our limit to predict IBD states along the genome by calculating the number of genomic windows where we correctly  predict IBD state for a pair of individuals.
-We calculate mean and standard deviation for this accuracy over 60 simulations of a pedigree (see section \ref{simulat}).
-Fig.
-\ref{fig4:IBDstate_accuracy} shows that we can predict IBD states for Unrelated, $3^{rd}$ Degree, $2^{nd}$ Degree, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling.
-On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03x coverage depth.
-
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/plotimg/plot_IBDaccuracy.png}
-    \centering
-    \caption{Evaluation of IBD estimation at different coverages.
-y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis.
-The error bars are drawn at 1 standard deviation from the mean.
-Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
-    \label{fig4:IBDstate_accuracy}
-\end{figure}
-
 \subsubsection{Application to Chagyrskaya Neanderthals}
-To test KIN on real ancient data, we applied it to a Neandertal data set  from Chagyrskaya Cave in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020}.
-This dataset contains genetic data from  14 skeletal remains that might  possibly belong to contemporary Neandertals who occupied the cave from 63 ± 4 to 48 ± 3 kya \cite{laurits_skov_genetic_nodate}.
-The data has low-to-intermediate depth of coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage.
+To test KIN on real ancient data, we applied it to a Neandertal data set  from Chagyrskaya Cave in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020,laurits_skov_genetic_nodate}.
+This dataset contains genetic data from  14 skeletal remains that likely belong to contemporary Neandertals who occupied the cave between 63 ± 4 and 48 ± 3 kya \cite{laurits_skov_genetic_nodate}.
+DNA extracted from each of these remains were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes, and common variations in Africans \cite{laurits_skov_genetic_nodate}, and sequenced  low-to-intermediate depth of coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage.
 Some of these specimens showed signs of long ROH, and contamination from modern humans as well as hyenas \cite{laurits_skov_genetic_nodate}.
-All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes, and common variations in Africans \cite{laurits_skov_genetic_nodate}.
-We further restricted this data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017} to make the data for our relatedness analysis same as that used by the authors.
+We further restricted these data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017} to make the data for our relatedness analyses match that used by the authors \cite{laurits_skov_genetic_nodate}.
 Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}.
-We found three fossils from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair  (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60).
+We found three specimens from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair  (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60).
 Further, we identify Chagyrskaya 07 and Chagyrskaya 60 as 3rd-degree relatives.
 
-Our estimates are consistent with those obtained using READ, except that READ is unable to detect 3rd-degree relationships, and cannot distinguish between siblings and parent-child.
+Our estimates are consistent with those obtained using READ, except that READ is unable to detect 3rd-degree relationships, and does not distinguish between sibling and parent-child relationships.
 We also find that both  KIN and READ classify Chagyrskaya06 and Chagyrskaya14 as parent-child with low confidence, but from the morphology they likely stem from the same individual.
-We believe that the low confidence miss-classification arises  due to non-human contamination present in these libraries ($1\%$ and $2.9\%$ respectively \cite{laurits_skov_genetic_nodate}), biasing the estimated differences between the individuals to higher values.
+We believe that the low confidence mis-classification may be due to non-human contamination present in these libraries ($1\%$ and $2.9\%$ respectively \cite{laurits_skov_genetic_nodate}), biasing the estimated differences between the individuals to higher values.
 
-We compared our IBD estimates to lcMLkin for all pairs for whom READ results (with s.d.
+We compared the IBD estimates obtained using KIN to those from lcMLkin for all pairs for whom READ results (with s.d.
 $>$1) and KIN results ($\Delta LL>$1) match (fig.
 S\ref{figS6:Chagyrskaya_ibd}).
-We find that lcMLkin does create four different clusters for different relatedness cases.
+We find that lcMLkin creates four different clusters for different relatedness cases (identical, paret-child, $2^{nd}$ degree and unrelated).
 However, the location of these clusters strongly deviate from the expected values, likely due to the low coverage and ROH.
@@
     \centering
-    \caption{Application of KIN on Neandertal specimens from Chagyrskaya.
+    \caption{Application of KIN to Neandertal remains from Chagyrskaya.
 The color of a square represents the relatedness, while the number within denotes log likelihood ratio ($\Delta LL$) between the two maximum likelihood models.}
@@
 
+\subsubsection{Application to ancient modern humans}
+
 We further applied KIN to a genome-wide dataset of 118 ancient individuals from the Lech Valley \cite{mittnik_kinship-based_2019}.
-We compared our relatedness estimates to those obtained by the authors using READ and lcMLkin, and found that in most cases the three methods agree.
-Out of 6903 pairwise comparisons, 346 pairs show different relatedness for KIN and READ (table S2).
-However, most of these are at very low confidence, and there are only 31 pairs where KIN has $\Delta LL>1$.
-When we investigated these comparisons in more detail we found that in many cases, the proportion of differences along the genome align with the prediction by KIN (fig.
+We compared our relatedness estimates to those obtained by the authors using READ and lcMLkin.
+We found that in most cases the three methods agree, as only $\approx5\%$ of pairs (346 out of 6903) show a different relatedness for KIN and READ (table S2).
+However, most of these are at very low confidence for both the methods ($LL<1$ for KIN and $sd<1$ for READ).
+There are 31 pairs where KIN has $\Delta LL>1$.
+When we investigated these comparisons in more detail we found that in many cases, the prediction by KIN is in agreement with the observed proportion of differences along the genome (fig.
 S\ref{figS8:eg1}).
 Only 20 pairs in this subset of 346 pairs had sufficient data for analysis using lcMLkin (more than 10,000 overlapping SNPs).
-Among these 20 pairs, we found that lcMLkin matches KIN's estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLkin predicts 3rd-5th degree, KIN predicts second degree) and pair AITI72-AITI77A (READ and lcMLkin predict unrelated, while KIN estimates 3rd degree with LL=1.7).
+Among these 20 pairs, we found that lcMLkin matches KIN's estimate in all cases except for pair POST131-POST28 where READ predicts unrelated individuals, lcMLkin predicts 3rd-5th degree, KIN predicts second degree, and pair AITI72-AITI77A where READ and lcMLkin predict unrelated, while KIN estimates 3rd degree with LL=1.7).
 We further found 43 pairs for which READ estimates First-Degree, and compared the estimates of lcMLkin and KIN for these cases (Table S3).
@@
 Here, we present a new method called KIN to estimate genetic kinship and the location of IBD tracts from low-coverage data, in presence of long ROH, contamination, as well as ascertainment bias.
-Our method utilizes a set of HMMs to estimate IBD tracts and uses them to classify each pair of individuals into a possible relatedness case, along with a measure of classification confidence.
+Our method utilizes a set of HMMs to estimate IBD tracts and uses them to classify each pair of individuals into a possible relatedness case, along with a measure of classification confidence (fig.
+S \ref{fig0:schematic}).
 We evaluated the method performance of KIN, and compared it to that of READ using simulated pedigrees.
@@
 For pairwise comparisons with large numbers of overlapping sites ($>10000$), KIN, READ and lcMLkin all mostly agree.
-However, KIN is able to differentiate between parent-child and siblings, and identify second and third degree relationships from just a few thousand sites overlapping between samples.
+However, KIN is able to differentiate between parent-child and siblings, and identify second degree relationship from just a few thousand polymorphic sites ($\approx$ 4000) overlapping between samples.
+KIN can also infer third degree relation with $\approx$ 50,000 overlapping polymorphic sites.
 We show that when applied to Neandertal specimens from Chagyrskaya Cave, KIN identifies a pair of $1^{st}$-Degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples \cite{laurits_skov_genetic_nodate}.
@@
 One limitation of our approach is that it assumes a single population.
-In case of a highly structured population, KIN may show inaccurate inferences, similar to READ and lcMLkin.
-Also, our method makes the assumptions, that the median pairwise genetic difference reflects the population diversity, which fails if almost all individuals in the data set are related.
+In case of a highly structured population, KIN may show inaccurate inference of $p_0$ causing inaccurate relatedness inference.
+Also, our method makes the assumption, that the median pairwise genetic difference in the population reflects the population diversity $p_0$, which fails if almost all individuals in the dataset are related.
 The current implementation of KIN is restricted to six relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.
@@
 \begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sup.png}
+    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic1.png}
     \centering

### 2022-04-27 paper_overleaf/main.tex
Updates from Overleaf

@@
 
-In ancient DNA studies, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_kinship-based_2019,sikora_ancient_2017}.
+In ancient DNA studies, relatedness can be used to identify bones and teeth belonging to the same individual.
+Given adequate familiarity with the subject, relatedness can provide an understanding of an ancient society's social structures, mobility and inheritance rules ~\cite{baca_ancient_2012,mittnik_kinship-based_2019,sikora_ancient_2017}.
 
@@
 
-In our simulations, we find that adding long ROH ($~17\%$) to simulations slightly improves the performance of both KIN and READ, particularly at lower coverages.
+For detecting ROH, there is only one method available that works with low coverage ancient DNA \cite{ringbauer_parental_2021}.
+This software can infer ROH at coverages $>0.3x$, but requires a reference panel.
+In the absence of a method to detect ROH at low coverages without a reference panel to detect long ROH, and we tested it on simulations to show that it can be used reliably at coverage $>=0.1x$.
+In our simulations, we find that adding long ROH ($~17\%$) to simulations slightly improves the power of both KIN and READ, particularly at lower coverages.
 This is likely because ROH actually reduces the variance in differences, and so it becomes easier to estimate the location of IBD tracts.
@@
 Also, our method makes the assumption, that the median pairwise genetic difference in the population reflects the population diversity $p_0$, which fails if almost all individuals in the dataset are related.
-The current implementation of KIN is restricted to six relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.
+The current implementation of KIN is restricted to six relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.Even though we do not differentiate between avuncular, grandparent-grandchild, half-siblings, we show that it is possible to differentiate between avuncular and grandparent-grandchild given that a pair is not half-sibling (see figure \ref{figS7:second_degree}).
 
@@
 For the evaluation of the ROH-HMM, we also combine ROH with ascertainment or contamination.
+
+\section{Acknowledgements}
+We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Fabrizio Mafessoni, Laurits Skov, Leonardo Iasi, Alba Bossom Mesa, Arev Sümer and Luca Traverso for helpful comments and discussions.
+This work was supported by the Max Planck Society and the European Research Council (Grant No.: 694707) to Svante Pääbo.
 

### 2022-04-29 plots/Rscripts/comparisonPlot.R
small fitting issue fixed

@@ -19,7 +19,7 @@ comparison_plot <- function(inf,outf){
 
   colnames(data)[4] <- "Method"
   data$Method=as.character(data$Method)
-  data[data$Method=='allLikelihoods_inphapProbs','Method'] = 'KIn'
+  data[data$Method=='allLikelihoods_inphapProbs','Method'] = 'KIN'
   data[data$Method=='read_inppshap','Method'] = 'READ'
 
 
@@ -53,7 +53,8 @@ comparison_plot <- function(inf,outf){
   plot + theme_bw() + theme(
     strip.background = element_rect(
       color="transparent", fill="transparent", size=1.5, linetype="solid"
-    ), legend.position="bottom", text = element_text(size=12), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
+    ), legend.position="bottom", text = element_text(size=12), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.title = element_text(size=10),
+        legend.text = element_text(size=8)
   )
   ggsave(outf,
          width = 8, height = 5, dpi = 150, units = "in", device='png')

### 2022-04-29 plots/Rscripts/comparisonPlot.R
small fitting issue fixed1

@@ -54,7 +54,7 @@ comparison_plot <- function(inf,outf){
     strip.background = element_rect(
       color="transparent", fill="transparent", size=1.5, linetype="solid"
     ), legend.position="bottom", text = element_text(size=12), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.title = element_text(size=10),
-        legend.text = element_text(size=8)
+        legend.text = element_text(size=9)
   )
   ggsave(outf,
          width = 8, height = 5, dpi = 150, units = "in", device='png')

### 2022-04-29 paper_overleaf/main.tex

@@
 This software can infer ROH at coverages $>0.3x$, but requires a reference panel.
-In the absence of a method to detect ROH at low coverages without a reference panel to detect long ROH, and we tested it on simulations to show that it can be used reliably at coverage $>=0.1x$.
+In the absence of a method to detect ROH at low coverages without a reference panel, and we tested it on simulations to show that it can be used reliably at coverage $\geq 0.1x$ to detect ROH regions at lengths on the order of 10cM.
 In our simulations, we find that adding long ROH ($~17\%$) to simulations slightly improves the power of both KIN and READ, particularly at lower coverages.
@@
 \section{Acknowledgements}
-We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Fabrizio Mafessoni, Laurits Skov, Leonardo Iasi, Alba Bossom Mesa, Arev Sümer and Luca Traverso for helpful comments and discussions.
+We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Laurits Skov, Leonardo Iasi, Alba Bossom Mesa, Arev Sümer, Guido Alberto Gnecchi Ruscone, Ke Wang and Luca Traverso for helpful comments and discussions.
 This work was supported by the Max Planck Society and the European Research Council (Grant No.: 694707) to Svante Pääbo.

### 2022-05-10 paper_overleaf/main.tex

@@
 We compared our relatedness estimates to those obtained by the authors using READ and lcMLkin.
-We found that in most cases the three methods agree, as only $\approx5\%$ of pairs (346 out of 6903) show a different relatedness for KIN and READ (table S2).
+We found that in most cases the three methods agree, as only $<5\%$ of pairs (311 out of 6903) show a different relatedness for KIN and READ (table S2).
 However, most of these are at very low confidence for both the methods ($LL<1$ for KIN and $sd<1$ for READ).
-There are 31 pairs where KIN has $\Delta LL>1$.
+There are 28 pairs where KIN has $\Delta LL>1$.
 When we investigated these comparisons in more detail we found that in many cases, the prediction by KIN is in agreement with the observed proportion of differences along the genome (fig.
 S\ref{figS8:eg1}).
-Only 20 pairs in this subset of 346 pairs had sufficient data for analysis using lcMLkin (more than 10,000 overlapping SNPs).
-Among these 20 pairs, we found that lcMLkin matches KIN's estimate in all cases except for pair POST131-POST28 where READ predicts unrelated individuals, lcMLkin predicts 3rd-5th degree, KIN predicts second degree, and pair AITI72-AITI77A where READ and lcMLkin predict unrelated, while KIN estimates 3rd degree with LL=1.7).
+Only 27 pairs in this subset of 311 pairs had sufficient data for analysis using lcMLkin (more than 10,000 overlapping SNPs).
+Among these 27 pairs, we found that lcMLkin matches KIN's estimate in all cases except for pair POST131-POST28 where READ predicts unrelated individuals, lcMLkin predicts 3rd-5th degree, KIN predicts second degree, and pair AITI72-AITI77A where READ and lcMLkin predict unrelated, while KIN estimates 3rd degree with LL=1.7).
 We further found 43 pairs for which READ estimates First-Degree, and compared the estimates of lcMLkin and KIN for these cases (Table S3).
@@
 \section{Acknowledgements}
-We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Laurits Skov, Leonardo Iasi, Alba Bossom Mesa, Arev Sümer, Guido Alberto Gnecchi Ruscone, Ke Wang and Luca Traverso for helpful comments and discussions.
+We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Johann Visagie, Laurits Skov, Leonardo Iasi, Alba Bossom Mesa, Arev Sümer, Zuzana Hofmanová, Guido Alberto Gnecchi Ruscone, Ke Wang and Luca Traverso for helpful comments and discussions.
 This work was supported by the Max Planck Society and the European Research Council (Grant No.: 694707) to Svante Pääbo.

### 2022-05-13 paper_overleaf/main.tex

@@
 
-\noindent Inferring familial relationships is important in population genetics.
-Knowing genetic kinship of ancient individuals leads to insights into the culture and social hierarchy of ancient populations, and is relevant for downstream genetic analyses.
-However, estimating relatedness from ancient DNA is difficult since DNA sequences extracted from ancient remains may have low coverage, ascertainment bias, or contamination from various sources.
+\noindent Inferring genetic kinship of ancient individuals can provide insights into their culture and social hierarchy, and is relevant for downstream genetic analyses.
+However, estimating relatedness from ancient DNA is difficult due to low coverage, ascertainment bias, or contamination from various sources.
 
-Here, we present KIN, a Hidden-Markov-Model-based approach to identify identity-by-descent fragments and to estimate degree of relatedness.
-KIN can accurately determine up to 3rd-degree relatives and differentiate between sibling and parent-child relationships using ancient DNA sequence data with as little as 0.05x coverage.
+Here, we present KIN, a Hidden-Markov-Model-based approach to estimate the relatedness of a pair of individuals and to identify the identical-by-descent segments they share.
+KIN can accurately classify up to 3rd-degree relatives and differentiate between sibling and parent-child relationships using ancient DNA sequence data with as little as 0.05x coverage.
 KIN also incorporates an adjustment for contamination, and an additional model to estimate the location of long runs of homozygosity, and uses them to improve classification accuracy.
 
-We show in simulations that KIN performs at near perfect classification accuracy at inferring first-degree relationships, and at 90\% accuracy for second degree even at 0.05x coverage.
+We show with simulations that KIN infers first-degree relationships with nearly perfect accuracy, and second degree relationships with 90\% accuracy, even at 0.05x coverage.
 
-We illustrate the utility of KIN on two applications, one on a Neandertal group from Siberia, Russia, and one in a group of Bronze-Age people from the Lech Valley, Germany. 
+To illustrate its utility, we apply KIN to a Neandertal group from Siberia, Russia, and a group of Bronze-Age people from the Lech Valley, Germany.
 We find that in both cases, KIN matches the findings of existing approaches and detects several additional related individuals.
 
-Thus, KIN is a useful tool for analyzing relatedness for large ancient-DNA datasets with low-coverage sequencing data that is applicable to any diploid species, without the requirement for a reference panel.
+KIN is a useful tool for analyzing relatedness for ancient DNA datasets of any size, even with low coverage, and is applicable to any diploid species, without the requirement for a reference panel.
 \end{abstract}
@@
 Due to the laws of Mendelian segregation, each parent will share exactly one set of chromosomes IBD with their offspring, while subsequent recombination means that a grandparent will, on average share a quarter of their genome with a grand-child.
-Along the genomes of a pair of diploid individuals, there are three IBD states possible at any given position: either no chromosomes are IBD, only one chromosome is IBD, or both chromosomes are IBD.
+Along the genomes of a pair of diploid individuals, there are three IBD states possible at any given position: the individuals share zero, one or two chromosomes IBD.
 The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals.
@@
 On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes.
-The bottom row shows the expected proportion of differences ($p$) along the chromosome in both the cases $p$ can take values of $p_0$, $p_1$ or $p_2$ which are the expected proportion of differences in IBD states 0,1 and 2 respectively.
+The bottom row shows the expected proportion of differences ($p$) along the chromosome.
+In both cases, $p$ can take values of $p_0$, $p_1$ or $p_2$, which are the expected proportion of differences in IBD states 0,1 and 2, respectively.
 }
@@
 
-\section{New Approaches}\label{new_approaches}
-To infer the case of relatedness, KIN fits one HMM for each pair of individuals and for each possible relatedness case.
+\section{Algorithm}\label{new_approaches}
+To infer relatedness, KIN fits one HMM for each pair of individuals and for each possible relatedness.
 The KIN-HMM infers IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts and contamination estimates in each individual into account.
@@
 The goal of KIN-HMM is to infer how two individuals are related via the patterns of shared IBD-states along a pair of genomes.
-For this purpose,  we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb), and infer the pattern of IBD-sharing for each relatedness case we consider (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical).
+For this purpose,  we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb), and infer the pattern of IBD-sharing for each relatedness case we consider (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{rd}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical).
 We then compare the likelihood between all models, and classify each pair to the model with the highest likelihood.
@@
 
-To test the performance of KIN, and compare it to a commonly used method, READ \cite{kuhn_estimating_2018}, we simulated an ancient population (sampled from 2500 generations ago) with msprime \cite{kelleher_efficient_2016}, and artificially mated some of the individuals to form a pedigree (figure \ref{figS10:pedigree}).
-We downsampled this data to low coverages (4$x$ - 0.03$x$), and introduced ascertainment bias and contamination from modern humans (see section \ref{method}) to understand how these factors affect the performance of KIN and READ.
-Finally we apply KIN to published data and show that it can give reliable information about genetic kinship.
+\subsection{Evaluation with simulations}
+We first tested the performance of KIN with simulated pedigrees.
+We performed coalescent simulations to generate 7 unrelated diploid genomes and artificially mated them to form pedigrees of 17 individuals with relationships up to $5^{th}$ degree (figure \ref{figS10:pedigree}).
+To evaluate the effect of sequence coverage on the performance of KIN, we generated artificial pileups at each polymorphic site for each individual following a Poisson distribution with 6 different average depths varying between 4$x$ and 0.03$x$ (see \ref{method}).
+To mimic ROH, for some pedigrees, we picked a single allele at heterozygous sites in some regions as determined by a Markov chain so that about $17\%$ of the genome is ROH.
+We also created versions with contamination, by introducing alleles from distantly related individuals, and ascertainment bias, by selecting polymorphic sites identified in a subset of the individuals (see \ref{method}).
+We created 60 pedigrees for each combination of average coverage and scenarios of presence/absence of ROH, contamination, and ascertainment bias, totalling 2,880 pedigrees.
 
-\paragraph{Evaluating the ROH detection}
-We next evaluate the performance of the ROH-HMM( Fig.\ref{fig2:ROH}).
-We observe that as coverage decreases, the raw heterozygosity estimates become noisier.
-In the case without ROH (Fig \ref{fig2:ROH}A), we correctly infer that there is no ROH for all  windows (except for the case of 0.1x coverage where we miss-classify 2 windows out of 220).
-We observe that the model predicts ROH probabilities well in all cases, even at a coverage of 0.1x (Table \ref{tab:Table 2}).
+\paragraph{ROH detection}
+We first evaluated the performance of the ROH-HMM ( Fig.\ref{fig2:ROH}).
+We computed the proportion of windows that are in ROH and compared it to the posterior probability of the ROH state, by computing the bias (i.e., the difference between both) and its variance across windows.
+Without ROH, we correctly inferred the absence of ROH as the observed bias ranges from $0.3$ to $0.7\%$ (variance $\leq0.006$; Table \ref{tab:Table 2}).
+With ROH, we observed a slight bias of $5$ to $9\%$ more ROH than simulated (variance $\leq0.004$).
+The largest bias observed ($9\%$) was for the datasets that included contamination.
+Ascertainment bias had a minimal effect on the detection of ROH (Table \ref{tab:Table 2}).
+Surprisingly, the bias was lower for lower coverages, perhaps because there is only one possible ROH state per window to model a continuous variable, viz.
+proportion of window in ROH, and this state is inferred more confidently with more data.
+Overall, the model predicts ROH well in all cases, even at a coverage of 0.1x (Fig.\ref{fig2:ROH}).
 
@@
 
-\paragraph{Evaluating IBD prediction}
-We investigate our limit to predict IBD states along the genome by calculating the number of genomic windows where we correctly  predict IBD state for a pair of individuals.
-We calculate mean and standard deviation for this accuracy over 60 simulations of a pedigree (see section \ref{simulat}).
-Fig.
-\ref{fig4:IBDstate_accuracy} shows that we can predict IBD states for Unrelated, $3^{rd}$ Degree, $2^{nd}$ Degree, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling.
-On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03x coverage depth.
+\paragraph{IBD prediction}
+We investigated the accuracy of IBD state prediction along the genome by counting the number of genomic windows where we correctly predict IBD state for different relatdness and coverages (Fig.
+\ref{fig4:IBDstate_accuracy}).
+The accuracy for coverages of 0.1x or higher is consistent and varies between 1.0 and 0.836, depending on relatedness.
+However, the accuracy decreases at lower coverages for most relatedness and ranges from 1.0 to 0.563 at 0.03x.
+On the other hand, the accuracy for identical individuals and parent-child pairs is nearly perfect for all investigated coverages, even at 0.03x coverage.
+[add some comments about contamination and ascertainment bias, with reference to supplementary figures]
 
@@
 
-\paragraph{Evaluation of KIN on simulations and comparison to READ}
-We next compare the classification accuracy of KIN (cutoff: 1 log-likelihood unit) to that of READ (cutoff: 1 standard deviation) under four different simulation scenarios (Fig.
+\paragraph{Relatedness classification}
+We finally evaluated the classification accuracy of KIN (cutoff: 1 log-likelihood unit) and compared it to that of READ (cutoff: 1 standard deviation)) (Fig.
 \ref{fig3:Comparison_READ_KIN}).
-We show that both methods have similar performance for low-coverage shotgun data (``control''-case), and for ascertained data.
-One exception is  $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages.
-We also find that READ is strongly impacted by contamination ($<3\%$), but that the correction implemented in KIN is sufficient to remove this bias.
-For simulations with ROH, we find the overall classification accuracy is only slightly reduced.
-Since ROH makes individuals more similar, we see a bias towards more false positives, for first and second-degree relationships, which is slightly higher for READ than for KIN (for details, see table S4).
+We first describe the results for relatedness cases detectable by READ, viz.
+identical, $1_{st} degree$, $2_{nd} degree$ and $unrelated$.
+In this case we show that both methods have similar performance for low-coverage shotgun data (``control''-case), and for ascertained data.
+The true positive rate is above 0.97 for both KIN and READ, while the false negative rate is below 0.02.
+One exception is  $2^{nd}$-degree relatives, where KIN has higher power than READ, and as coverage decreases from 4x to 0.05x, KIN's true positive rate decreases from 0.94 to 0.86, compared to range of 0.91 to 0.74 for READ).
+The false positive rate in this case remains below 0.02 for both methods.
+We also find that READ is strongly impacted by contamination ($<3\%$) as the true positive rate is in the range 0.89 to 0.78 and 0.53 to 0.46 for 1st and 2nd degree relatives, respectively, and the false positive rate reaches up to 0.04 and 0.17 for unrelated individuals and 2nd degree relatives, respectively.
+By contrast, the correction implemented in KIN is sufficient to remove this bias, and the true and false positive rates remain as in the control.
+For simulations with ROH, we find that KIN also outperforms READ for 1st and 2nd degree relatives, although both the true and false positive rates increased for both methods compared to the control for 2nd degree relatives.
+The increases in false positives is likely due to ROH making individuals more similar (for details, see table S4).
+Finally, we also found that KIN has good power to detect relatedness cases that are not detectable by READ, i.e.
+parent-child, siblings and, to a lower level, 3rd degree relatives (Fig.
+\ref{fig3:Comparison_READ_KIN})
 
@@
 
-\subsubsection{Application to Chagyrskaya Neanderthals}
-To test KIN on real ancient data, we applied it to a Neandertal data set  from Chagyrskaya Cave in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020,laurits_skov_genetic_nodate}.
+\subsubsection{Application to real data}
+
+\paragraph{Chagyrskaya Neandertals}
+To test KIN on real ancient data, we applied it to a Neandertal dataset from Chagyrskaya Cave in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020,laurits_skov_genetic_nodate}.
 This dataset contains genetic data from  14 skeletal remains that likely belong to contemporary Neandertals who occupied the cave between 63 ± 4 and 48 ± 3 kya \cite{laurits_skov_genetic_nodate}.
-DNA extracted from each of these remains were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes, and common variations in Africans \cite{laurits_skov_genetic_nodate}, and sequenced  low-to-intermediate depth of coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage.
-Some of these specimens showed signs of long ROH, and contamination from modern humans as well as hyenas \cite{laurits_skov_genetic_nodate}.
-We further restricted these data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017} to make the data for our relatedness analyses match that used by the authors \cite{laurits_skov_genetic_nodate}.
+DNA extracted from each of these remains were captured with an array targeting variable sites identified in high coverage Neandertal and Denisovan genomes, and common variations in Africans \cite{laurits_skov_genetic_nodate}.
+This genetic data has  low-to-intermediate depth of coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage.
+Some of these specimens showed signs of long ROH, and DNA contamination from modern humans as well as hyenas \cite{laurits_skov_genetic_nodate}.
+We focused our analysis to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017} as done by the authors for the relatedness analysis \cite{laurits_skov_genetic_nodate}.
 Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}.
 We found three specimens from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60).
-Further, we identify Chagyrskaya 07 and Chagyrskaya 60 as 3rd-degree relatives.
+Further, we identify Chagyrskaya17 and Chagyrskaya60 as 3rd-degree relatives.
 
@@
 
-\subsubsection{Application to ancient modern humans}
+\subsubsection{Ancient modern humans}
 
@@
 We compared our relatedness estimates to those obtained by the authors using READ and lcMLkin.
-We found that in most cases the three methods agree, as only $<5\%$ of pairs (311 out of 6903) show a different relatedness for KIN and READ (table S2).
-However, most of these are at very low confidence for both the methods ($LL<1$ for KIN and $sd<1$ for READ).
-There are 28 pairs where KIN has $\Delta LL>1$.
-When we investigated these comparisons in more detail we found that in many cases, the prediction by KIN is in agreement with the observed proportion of differences along the genome (fig.
+We found that in most cases the three methods agree.
+Only $<5\%$ of pairs (315 out of 6903) show a different relatedness for KIN and READ (table S2).
+However, most of these are at very low confidence for both the methods ($\Delta LL>1$ for KIN and $sd<1$ for READ).
+There are 28 pairs where KIN shows confident results, and when we plot the raw heterozygosity estimates along the genome, in many cases the prediction by KIN is in agreement with the observed proportion of differences along the genome (fig.
 S\ref{figS8:eg1}).
-Only 27 pairs in this subset of 311 pairs had sufficient data for analysis using lcMLkin (more than 10,000 overlapping SNPs).
-Among these 27 pairs, we found that lcMLkin matches KIN's estimate in all cases except for pair POST131-POST28 where READ predicts unrelated individuals, lcMLkin predicts 3rd-5th degree, KIN predicts second degree, and pair AITI72-AITI77A where READ and lcMLkin predict unrelated, while KIN estimates 3rd degree with LL=1.7).
-We further found 43 pairs for which READ estimates First-Degree, and compared the estimates of lcMLkin and KIN for these cases (Table S3).
-There were a total of 20 pairs where lcMLkin had sufficient data, and 26 pairs where $\Delta LL>1$ for KIN.
-Among the 19 overlapping pairs, we found four pairs where KIN and lcMLkin differ, and again the heterozygosities along the genome match the inference of KIN (fig.
+Comparisons with lcMLkin are spare since lcMLkin shows reliable results for $\appr 53\%$ pairs, and we find 16 pairs for which lcMLkin gives a reliable result (\cite{mittnik_kinship-based_2019}), and KIN shows $\Delta LL>$1.
+We again investigated these comparisons in more detail, and found that at least in cases where there is good amount of overlapping sites ($\geq20,000$), we can clearly see that proportion of differences along the genome aligns with prediction of KIN (fig.
 S\ref{figS9:eg2}).

### 2022-05-13 plots/Rscripts/supp_IBD.R
added ibd supplementary plot

@@ -1,0 +1,77 @@
+library(ggplot2)
+
+#infile1='/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/plots/plotdata/plot_IBDaccuracy_cont.csv'
+#infile2='/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/plots/plotdata/plot_IBDaccuracy_asc.csv'
+#infile3='/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/plots/plotdata/plot_IBDaccuracy_roh.csv'
+
+IBD_accuracy_plot <- function(infile1,infile2,infile3,outfile){
+  bigd <-data.frame()
+  for(j in c(1,2,3)) {
+  if (j==1){
+    infile=infile1
+    }else if (j==2){
+      infile=infile2
+    }else if (j==3){
+      infile=infile3
+    }
+  data=read.csv(file = infile, header = TRUE)
+  
+  
+  data[data$rel=='deg4','rel']='Unrelated'
+  data[data$rel=='deg5','rel']='Unrelated'
+  data[data$rel=='un','rel']='Unrelated'
+  
+  data[data$rel=='avu','rel']='2nd Degree'
+  data[data$rel=='gr','rel']='2nd Degree'
+  data[data$rel=='hsib','rel']='2nd Degree'
+  
+  data[data$rel=='id','rel']='Identical'
+  data[data$rel=='pc','rel']='Parent-Child'
+  data[data$rel=='sib','rel']='Siblings'
+  
+  data[data$rel=='deg3','rel']='3rd Degree'
+  
+  allrel=c("Unrelated", "3rd Degree", "2nd Degree", "Siblings", "Parent-Child", "Identical")
+  allcov=c(4,0.5,0.2,0.1,0.05,0.03)
+  
+  accuracy=c()
+  stdev=c()
+  rel=c()
+  cov=c()
+  for(cv in allcov) {
+    
+    datacv=data[data$cov==cv,]
+    for(i in allrel) {
+      temp=datacv[datacv$rel==i,]
+      accuracy=c(accuracy,mean(temp$accuracy))
+      stdev=c(stdev,sd(temp$accuracy))
+      rel=c(rel,i)
+      cov=c(cov,cv)
+    }
+  }
+  df=data.frame(accuracy,stdev,rel,cov)
+  
+  df$cov=as.factor(df$cov)
+  df$Scenario=j
[…]

### 2022-05-13 plots/plot_script.py

@@ -500,7 +500,7 @@ def comparison_plotf(hmmfold,outfold, outs):
                     supf.to_csv(outs, sep=',',index=False)
 
 
-def IBDstates(fold, outf, list_inds, runlist):
+def IBDstates(fold, outf, list_inds, runlist, var):
 
     accuracy=[]
     cov_all=[]
@@ -529,7 +529,15 @@ def IBDstates(fold, outf, list_inds, runlist):
                 tr=np.loadtxt(trf, dtype='float', delimiter = ",")
 
             for cov in [4,0.5,0.2,0.1,0.05,0.03]:
-                vname=fold+"hmm_numba_viterbi_all_win/contam0/inbred0/run%s/coverage%s/filtered0/asc0/relation_%s_file/res_inphapProbs/pw_%s.csv.gz" %(run,cov,rel,pair)
+                if var=='control':
+                    vname=fold+"hmm_numba_viterbi_all_win/contam0/inbred0/run%s/coverage%s/filtered0/asc0/relation_%s_file/res_inphapProbs/pw_%s.csv.gz" %(run,cov,rel,pair)
+                elif var=='c':
+                    vname=fold+"hmm_numba_viterbi_all_win/contam1/inbred0/run%s/coverage%s/filtered0/asc0/relation_%s_file/res_inphapProbs/pw_%s.csv.gz" %(run,cov,rel,pair)
+                elif var=='a':
+                    vname=fold+"hmm_numba_viterbi_all_win/contam0/inbred0/run%s/coverage%s/filtered0/asc2/relation_%s_file/res_inphapProbs/pw_%s.csv.gz" %(run,cov,rel,pair)
+                elif var=='r':
+                    vname=fold+"hmm_numba_viterbi_all_win/contam0/inbred1/run%s/coverage%s/filtered0/asc0/relation_%s_file/res_inphapProbs/pw_%s.csv.gz" %(run,cov,rel,pair)
+
                 viterbi=np.loadtxt(vname, dtype='float', delimiter = ",")
 
                 ac=sum(viterbi == tr)/len(viterbi)

### 2022-05-13 paper_overleaf/main.tex
Updates from Overleaf

@@
 
+\begin{table}[h!]{ l p{5cm}}
+\caption{\label{tab:Table 2}Model performance for ROH prediction.
+Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination, R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
+    \hline
+    Coverage & Control(bias) & Control(var) & R(bias) & R(var) & RA(bias) & RA(var) & RC(bias) & RC(var)\\
+    \hline
+    4x & -0.007 & 0.006 & -0.081 & 0.036 & -0.083 & 0.036 & -0.091 & 0.039\\
+    \hline
+    0.2x & -0.003 & 0.003 & -0.068 & 0.03 & -0.068 & 0.026 & -0.076 & 0.029\\
+    \hline
+    0.1x & -0.003 & 0.002 & -0.049 & 0.021 & -0.05 & 0.021 & -0.057 & 0.022\\
+    \hline
+\end{tabular}
+\label{table1}
+\end{table}
+
 \paragraph{IBD prediction}
@@
 On the other hand, the accuracy for identical individuals and parent-child pairs is nearly perfect for all investigated coverages, even at 0.03x coverage.
-[add some comments about contamination and ascertainment bias, with reference to supplementary figures]
+We find that adding contamination, ascertainment bias and ROH in our simulations does not affect IBD prediction for identical and parent-child pairs, but reduces accuracy for unrelated individuals by xx, yy and zz respectively, for second degree by xx,yy,zz and for siblings by xx,yy,zz respectively.
 
@@
     \caption{Evaluation of IBD estimation at different coverages.
-y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-Here,Here,axis.
+y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis.
 The error bars are drawn at 1 standard deviation from the mean.
@@
 
-\begin{table}
-\caption{\label{tab:Table 2}Model performance for ROH prediction.
-Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination, R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
-\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
-    \hline
-    Coverage & Control(bias) & Control(var) & R(bias) & R(var) & RA(bias) & RA(var) & RC(bias) & RC(var)\\
-    \hline
-    4x & -0.007 & 0.006 & -0.081 & 0.036 & -0.083 & 0.036 & -0.091 & 0.039\\
-    \hline
-    0.2x & -0.003 & 0.003 & -0.068 & 0.03 & -0.068 & 0.026 & -0.076 & 0.029\\
-    \hline
-    0.1x & -0.003 & 0.002 & -0.049 & 0.021 & -0.05 & 0.021 & -0.057 & 0.022\\
-    \hline
-\end{tabular}
-\label{table1}
-\end{table}
-
 \renewcommand{\figurename}{Fig. S}
@@
     \centering
+    \includegraphics[width=18cm]{plots/plotimg/plot_IBDaccuracy_asc.png}
+    \caption{Evaluation of IBD estimation in presence of ascertainment at different coverages.
+y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis.
+The error bars are drawn at 1 standard deviation from the mean.
+Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
+    \label{figS11:ibd_asc}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=18cm]{plots/plotimg/plot_IBDaccuracy_roh.png}
+    \caption{Evaluation of IBD estimation in presence of ROH at different coverages.
+y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis.
+The error bars are drawn at 1 standard deviation from the mean.
+Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
+    \label{figS11:ibd_asc}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=18cm]{plots/plotimg/plot_IBDaccuracy_cont.png}
+    \caption{Evaluation of IBD estimation in presence of contamination at different coverages.
+y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis.
+The error bars are drawn at 1 standard deviation from the mean.
+Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
+    \label{figS11:ibd_asc}
+\end{figure}
+\begin{figure}[h!]
+    \centering
     \includegraphics[width=18cm]{supplementary_info/plots/lcPlot.png}

### 2022-05-14 plots/Rscripts/calculating_IBD_supplementary_stats_for_text.R
added control

@@ -1,5 +1,8 @@
 library(ggplot2)
-
+#infile1='/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/plots/plotdata/plot_IBDaccuracy.csv'
+#infile2='/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/plots/plotdata/plot_IBDaccuracy_cont.csv'
+#infile3='/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/plots/plotdata/plot_IBDaccuracy_asc.csv'
+#infile4='/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/plots/plotdata/plot_IBDaccuracy_roh.csv'
 
 IBD_accuracy_plot <- function(infile1,infile2,infile3,infile4,outfile){
   bigd <-data.frame()
@@ -53,13 +56,14 @@ IBD_accuracy_plot <- function(infile1,infile2,infile3,infile4,outfile){
     df$cov=as.factor(df$cov)
     df$Scenario=j
     bigd=rbind(bigd,df)
-  
-  df=data.frame(accuracy,stdev,rel,cov)
-  
-  df$cov=as.factor(df$cov)
-  df$Scenario=j
-  bigd=rbind(bigd,df)
   }
+  
+  bigd_un=bigd[bigd$rel=='Siblings',]
+  bigd_un_control=bigd_un[bigd_un$Scenario==1,]
+  bigd_un_c=bigd_un[bigd_un$Scenario==2,]
+  bigd_un_a=bigd_un[bigd_un$Scenario==3,]
+  bigd_un_r=bigd_un[bigd_un$Scenario==4,]
+  
   bigd$Scenario=as.factor(bigd$Scenario)
   colors1 <- c("Unrelated" = "grey30", "3rd Degree" = "blue", "2nd Degree" = "darkorchid1", "Siblings"= "Green", "Parent-Child" = "orangered3", "Identical" = "darkgoldenrod1")
   line1 <- c("1" = "solid", "2" = "dashed", "3" = "dotted","4" = "12345678")
@@ -74,9 +78,9 @@ IBD_accuracy_plot <- function(infile1,infile2,infile3,infile4,outfile){
     scale_color_manual(values = colors1, name="Relatedness") + theme_bw() +geom_path(aes(x=cov, y=accuracy, group=interaction(rel,Scenario),colour=rel, linetype = Scenario)) +
[…]

### 2022-05-14 paper_overleaf/main.tex
Updates from Overleaf

--- 2022-05-13 paper_overleaf/main.tex
@@
 The accuracy for coverages of 0.1x or higher is consistent and varies between 1.0 and 0.836, depending on relatedness.
-However, the accuracy decreases at lower coverages for most relatedness and ranges from 1.0 to 0.563 at 0.03x.
-On the other hand, the accuracy for identical individuals and parent-child pairs is nearly perfect for all investigated coverages, even at 0.03x coverage.
-[add some comments about contamination and ascertainment bias, with reference to supplementary figures]
+However, the accuracy decreases at lower coverages for most relatedness cases and ranges from 1.0 to 0.563 at 0.03x.
+The exceptions are for identical individuals and parent-child pairs, where the accuracy is nearly perfect for all investigated coverages, even at 0.03x coverage.
+We find that adding contamination, ascertainment bias in our simulations does not affect IBD prediction.
+Adding ROH to the simulations does not affect identical, parent-child, unrelated and $3^{rd} degree$  pairs, but reduces the IBD prediction accuracy for second degree by $1.5$ to $5\%$ and for siblings by $5$ to $17\%$ (fig\ref{figS11:ibd_others}.
+We see this adverse effect of ROH in case of siblings, perhaps because only in case of siblings we do not introduce long ROH directly, but long and short ROH may be present through parents (see section \ref{method}).
+Unidentified short ROH may cause a difficulty in differentiating between different combinations of IBD ($Z_w$) and ROH states ($H_w$).
 
@@
 \paragraph{Relatedness classification}
-We finally evaluated the classification accuracy of KIN (cutoff: 1 log-likelihood unit) and compared it to that of READ (cutoff: 1 standard deviation)) (Fig.
+We evaluated the classification accuracy of KIN (cutoff: 1 log-likelihood unit) and compared it to that of READ (cutoff: 1 standard deviation)) (Fig.
 \ref{fig3:Comparison_READ_KIN}).
@@
 The true positive rate is above 0.97 for both KIN and READ, while the false negative rate is below 0.02.
-One exception is  $2^{nd}$-degree relatives, where KIN has higher power than READ, and as coverage decreases from 4x to 0.05x, KIN's true positive rate decreases from 0.94 to 0.86, compared to range of 0.91 to 0.74 for READ).
+One exception is  $2^{nd}$-degree relatives, where KIN has higher power than READ, and as coverage decreases from 4x to 0.05x, KIN's true positive rate decreases from 0.94 to 0.86, compared to range of 0.91 to 0.74 for READ.
 The false positive rate in this case remains below 0.02 for both methods.
@@
 Some of these specimens showed signs of long ROH, and DNA contamination from modern humans as well as hyenas \cite{laurits_skov_genetic_nodate}.
-We focused our analysis to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017} as done by the authors for the relatedness analysis \cite{laurits_skov_genetic_nodate}.
+We focused our analysis on the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017} as done by the authors for the relatedness analysis \cite{laurits_skov_genetic_nodate}.
 Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}.
@@
 Comparisons with lcMLkin are spare since lcMLkin shows reliable results for $\appr 53\%$ pairs, and we find 16 pairs for which lcMLkin gives a reliable result (\cite{mittnik_kinship-based_2019}), and KIN shows $\Delta LL>$1.
-We again investigated these comparisons in more detail, and found that at least in cases where there is good amount of overlapping sites ($\geq20,000$), we can clearly see that proportion of differences along the genome aligns with prediction of KIN (fig.
+We again investigated these comparisons in more detail, and found that at least in cases where there is good amount of overlapping sites ($\geq20,000$), we can clearly see that the proportion of differences along the genome aligns with the prediction of KIN (fig.
 S\ref{figS9:eg2}).
@@
 \left[\begin{array}
-{rrr}
+{rrr}    bigd=rbind(bigd,df)
+  }
 -4r & 4r & 0\\
@@
 \setcounter{figure}{0}
-
-\begin{table}
-\caption{\label{tab:Table 2}Model performance for ROH prediction.
-Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination, R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
-\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
-    \hline
-    Coverage & Control(bias) & Control(var) & R(bias) & R(var) & RA(bias) & RA(var) & RC(bias) & RC(var)\\
-    \hline
-    4x & -0.007 & 0.006 & -0.081 & 0.036 & -0.083 & 0.036 & -0.091 & 0.039\\
-    \hline
-    0.2x & -0.003 & 0.003 & -0.068 & 0.03 & -0.068 & 0.026 & -0.076 & 0.029\\
-    \hline
-    0.1x & -0.003 & 0.002 & -0.049 & 0.021 & -0.05 & 0.021 & -0.057 & 0.022\\
-    \hline
-\end{tabular}
-\label{table1}
-\end{table}
 
@@
     \centering
+    \includegraphics[width=18cm]{plots/plotimg/plot_IBDaccuracy_others_supp.png}
+    \caption{Evaluation of IBD estimation in presence of contamination, ascertainment or ROH (denoted in figure legend by C, A and R, respectively)at different coverages.
+y-axis shows accuracy calculated over 60 simulations for each relatedness case, for each scenario of C, A or R, and six different coverages shown on x-axis.
+The error bars are drawn at 1 standard deviation from the mean.
+Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
+    \label{figS11:ibd_others}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
     \includegraphics[width=18cm]{supplementary_info/plots/lcPlot.png}

### 2022-05-14 paper_overleaf/main.tex

@@
 S\ref{figS8:eg1}).
-Comparisons with lcMLkin are spare since lcMLkin shows reliable results for $\appr 53\%$ pairs, and we find 16 pairs for which lcMLkin gives a reliable result (\cite{mittnik_kinship-based_2019}), and KIN shows $\Delta LL>$1.
+Comparisons with lcMLkin are sparse since lcMLkin shows reliable results for $\appr 53\%$ pairs, and we find 16 pairs for which lcMLkin gives a reliable result (\cite{mittnik_kinship-based_2019}), and KIN shows $\Delta LL>$1.
 We again investigated these comparisons in more detail, and found that at least in cases where there is good amount of overlapping sites ($\geq20,000$), we can clearly see that the proportion of differences along the genome aligns with the prediction of KIN (fig.

### 2022-05-15 paper_overleaf/KIN_a_method_to_estimate_relatedness_from_low_coverage_data_supplements.tex
Updates from Overleaf

--- 2022-05-13 paper_overleaf/main.tex
@@
 \maketitle
-
-\begin{abstract}
-
-\noindent Inferring genetic kinship of ancient individuals can provide insights into their culture and social hierarchy, and is relevant for downstream genetic analyses.
-However, estimating relatedness from ancient DNA is difficult due to low coverage, ascertainment bias, or contamination from various sources.
-
-Here, we present KIN, a Hidden-Markov-Model-based approach to estimate the relatedness of a pair of individuals and to identify the identical-by-descent segments they share.
-KIN can accurately classify up to 3rd-degree relatives and differentiate between sibling and parent-child relationships using ancient DNA sequence data with as little as 0.05x coverage.
-KIN also incorporates an adjustment for contamination, and an additional model to estimate the location of long runs of homozygosity, and uses them to improve classification accuracy.
-
-We show with simulations that KIN infers first-degree relationships with nearly perfect accuracy, and second degree relationships with 90\% accuracy, even at 0.05x coverage.
-
-To illustrate its utility, we apply KIN to a Neandertal group from Siberia, Russia, and a group of Bronze-Age people from the Lech Valley, Germany.
-We find that in both cases, KIN matches the findings of existing approaches and detects several additional related individuals.
-
-KIN is a useful tool for analyzing relatedness for ancient DNA datasets of any size, even with low coverage, and is applicable to any diploid species, without the requirement for a reference panel.
-\end{abstract}
-
-\section{Introduction}
-
-\subsection{Why study relatedness?}
-
-Identifying related individuals is a common task in genetic studies.
-Relatedness is of direct interest in e.g.
-DNA forensics, where familial search can aid in solving criminal cases, and to identify unknown deceased persons \cite{murphy_law_2018,ram_genealogy_2018}.
-Genetic paternity tests have an important application in resolving family relation, e.g.
-in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}.
-It is also an essential step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.
-For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to avoid inbreeding and ensure diversity \cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}.
-
-In ancient DNA studies, relatedness can be used to identify bones and teeth belonging to the same individual.
-Given adequate familiarity with the subject, relatedness can provide an understanding of an ancient society's social structures, mobility and inheritance rules ~\cite{baca_ancient_2012,mittnik_kinship-based_2019,sikora_ancient_2017}.
-
-\subsection{Approaches to estimate relatedness from high-coverage data}
-
-Commonly, pairs of related individuals are identified by looking for parts of the genome that are identical by descent (IBD), i.e.
-inherited from a recent common ancestor.
-Due to the laws of Mendelian segregation, each parent will share exactly one set of chromosomes IBD with their offspring, while subsequent recombination means that a grandparent will, on average share a quarter of their genome with a grand-child.
-Along the genomes of a pair of diploid individuals, there are three IBD states possible at any given position: the individuals share zero, one or two chromosomes IBD.
-The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals.
-For example, a pair of siblings are expected to have all three possible IBD states with proportions of 0.25,0.5,0.25, respectively (Fig \ref{fig0:schematic}).
-These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}.
-One can also use these probabilities to estimate the coefficient of relatedness $r$, which is defined as the proportion of the genome that is IBD.
-In the absence of inbreeding, this would be calculated as $r= k_1/2 + k_2$.
-
-However, since it is not possible to directly observe IBD segments, a common approach is to first identify segments of the genome that are Identical by State (IBS) and to use population allele frequencies obtained from an out-of-sample reference panel to calculate the probability of IBD given IBS.
-There are several methods that incproporate reference panel allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007,manichaikul_robust_2010,gusev_whole_2009,nyerki_optimized_2022}.
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib.png}
-    \centering
-    \caption{IBD sharing between siblings without and with runs of homozygosity (ROH).
-In top row, recombined chromosomes inherited from parents by a pair of siblings are shown.
-On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes.
-The bottom row shows the expected proportion of differences ($p$) along the chromosome.
-In both cases, $p$ can take values of $p_0$, $p_1$ or $p_2$, which are the expected proportion of differences in IBD states 0,1 and 2, respectively.
-}
-    \label{fig0:schematic}
-\end{figure}
-
-\begin{table}[h!]
-\caption{\label{tab:Table 1}IBD sharing probabilities for different relations in absence of inbreeding}
-\begin{tabular}{|c|c|c|c|}
-    \hline
-    Relatedness & $k_0$ & $k_1$ & $k_2$\\
-    \hline
-    Unrelated & 1 & 0 & 0\\
-    \hline
-    3rd Degree & 0.75 & 0.25 & 0\\
-    \hline
-    2nd Degree & 0.5 & 0.5 & 0\\
-    \hline
-    Siblings & 0.25 & 0.5 & 0.25\\
-    \hline
-    Parent-Child & 0 & 1 & 0\\
-    \hline
-    Identical/twins & 0 & 0 & 1\\
-    \hline
-\end{tabular}
-\label{table1}
-\end{table}
-
-\subsection{Approaches that address problems with ancient DNA data}
-One major issue with applying the above-mentioned methods to ancient DNA data is that the sequence coverage is typically low, making it difficult to obtain accurate genotype calls.
-Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}.
-In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes, weighted by their genotype likelihoods.
-However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be imprecise at lower coverages \cite{korneliussen_angsd_2014} .
-Ancient DNA analyses often face additional challenges such as contamination with present-day DNA \cite{peyregne_present-day_2020-1}, the unavailability of  reference panels to estimate population allele frequencies, and an ascertainment bias caused by  DNA capture approaches \cite{prufer_computational_2010, vai_kinship_2020}.
-
-Several methods have been proposed to estimate relatedness without a reference panel from ancient DNA, but they  require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of allele frequencies in the population from which the sampled individuals originate \cite{theunert_joint_2017}.
-A second issue is contamination.
-If the contamination stems from another population, contaminated data will look more dissimilar to other individuals from the analyzed population, and hence relatedness will be underestimated.
-In addition, some analyzed  genomes may have long runs of homozygosity (ROH), for example due to a small population size, or recent inbreeding.
-Long ROH cause related individuals to seem genetically more similar to each other, but do not affect the genetic distance between unrelated individuals.
-
-Moreover, ancient DNA is commonly captured with a SNP array that enriches for informative variants.
-Particularly methods based on the fraction of sites in different IBS states are sensitive to the ascertainment bias caused by this non-random selection of targeted sites \cite{waples_allele_2019}.
-
-READ \cite{kuhn_estimating_2018} addresses several of the issues encountered in the analysis of ancient DNA.
-In particular, the lack of  genotype calls is dealt with by randomly sampling alleles from each individual.
-A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to estimate  average pairwise genetic distances, which in turn are used to infer relatedness.
-However, READ only infers the degree of relatedness, and only up to second degree.
-
-\subsection{How our method works}
-Here, we present KIN (Kinship INference), a Hidden Markov Model (HMM)-based approach to estimate genetic kinship and IBD from low-coverage ancient DNA data.
-KIN can detect up to \nth{3}-degree relatives, and differentiates between siblings and parent-child relationships.
-KIN is also able to take into account ROH, contamination and ascertainment.
-We validate the performance of KIN using simulations and show that we are able to infer relatedness in real data from two datasets: a group of Neandertals and a group of Bronze age individuals.
-
-\section{Algorithm}\label{new_approaches}
-To infer relatedness, KIN fits one HMM for each pair of individuals and for each possible relatedness.
-The KIN-HMM infers IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts and contamination estimates in each individual into account.
-The best-fit model is then assigned as the inferred relatedness.
-If the locations of ROH tracts are unknown, we provide another HMM (ROH-HMM) to coarsely estimate the location of ROH for samples with sufficient coverage ($\geq 0.1x$).
-Our method is available on \url{https://github.com/DivyaratanPopli/KIN_snakemake} along with a \texttt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.
-
-\subsection{Model description}\label{method_overview} 
-The goal of KIN-HMM is to infer how two individuals are related via the patterns of shared IBD-states along a pair of genomes.
-For this purpose,  we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb), and infer the pattern of IBD-sharing for each relatedness case we consider (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{rd}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical).
-We then compare the likelihood between all models, and classify each pair to the model with the highest likelihood.
-We also return the most likely locations of IBD tracts (using standard Viterbi algorithm), and the state posterior probabilities $\BZ$.
-
-The details of the likelihood computation are given in section \ref{ll}.
-As is the case for any HMM, KIN-HMM requires a set of emission (section \ref{B}) and transition probability matrices.
-We assume the transition matrix for each relatedness case is fixed.
-Our emissions include a vector of parameters ($\delta$) that describe the variance in the data for each IBD state that we infer from the data (section \ref{B}).
-
-\subsection{Input of KIN-HMM}\label{hmm_input}
-The inputs of our algorithm are i) the number of overlapping sites for the $w$-th window $N_w$ for which both samples have at least one read available, ii) the expected number of pairwise differences $D_w$ at these sites, and iii) the probability of ROH in windows, by default obtained from ROH-HMM described in section \ref{roh}.
-
-$D_w$ is the number of sites that are different between a pair of individuals.
-For high-coverage data, this can be estimated accurately by comparing genotypes, but for low-coverage samples, $D_w$ needs to be estimated.
-The simplest approach is to  randomly-sample a read from each position \cite{haak_massive_2015,prufer_complete_2014}.
-However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
-\begin{align}\label{eq:x}
-D_w &= \sum_{s=1}^{N_w} \nu_i(s) (1-\nu_j(s)) + (1-\nu_i(s)) \nu_j(s)
-\end{align}
-Here, $\nu_i(s)$ and $\nu_j(s)$ are the proportions of reads carrying the derived allele at SNP index $s$ for individuals i and j respectively.
-Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$.
-
-\subsection{Log likelihood of the KIN-HMM}\label{ll}
-The KIN-HMM uses $\BD$ and $\BN$ to classify each window into three hidden states $Z_w \in$ ($0$, $1$, $2$), reflecting zero, one or two shared chromosomes IBD, respectively.
-To take ROH into account, we define the variable $H_w \in$ ($0$, $1$, $2$) that designates that zero, one or both individuals are homozygous in window $w$.
-Since $H_w$ is unobserved, in practice we use the estimates from ROH-HMM: $h_{wj} = P(H_w = j)$.
-
-There are three additional model parameters, $\pi, \mathbf{A}$ and $\delta$.
-The initial probabilities $\pi$ give the probabilities of being in each state $Z_0$ (at the beginning of each chromosome), which we set to uniform for simplicity.
-The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case.
-The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).
-For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.
-
-Thus, the complete data likelihood for the HMM is
-
-\begin{align}\label{eq:1}
-\log P(\BD,\BZ|\BH, \theta, \delta) &= \log P(\BD|\BZ,\BH, \theta, \delta) + \log P(\BZ |\theta) \nonumber\\
-&= \sum_w \log P(D_w|Z_w,H_w, N_w, \delta) + \sum_w \log P(Z_w |Z_{w-1},A) + \log P(Z_0|\pi)\text{.}
-\end{align}
-
-Here, $\BZ$ is not dependent on $\BH$, $\theta$ and $\delta$.
-
-\subsection{Emission probability}\label{B}
-
-Using this setup, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta, \delta)$ from equation \ref{eq:1} and optimize them for $\delta$.
-The simplest model is to assume that sites in each window are equally distributed and independent, in this case we could use the  binomial likelihood:
-$$P(D_w|Z_w, H_w, N_w) \sim \text{Binom}[D_w ; p(Z_w, H_w), N_w] \text{,}$$
-
-where, $p$ is the proportion of differences expected for a particular IBD and ROH state.
-If the two individuals are unrelated in a particular window (i.e.
-$Z_w = 0$), then the expected proportion of pairwise differences depends solely on the population history, and we denote this proportion with $p_0$.
-If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes \cite{kuhn_estimating_2018}.
-Thus, $p(Z_w=i, H_w=0) = p_i$.
-
-The proportion of differences between unrelated individuals $p_0$ is an important parameter.
-We follow READ \cite{kuhn_estimating_2018} and estimate $p_0$ as the median of differences for all possible pairs of individuals, which works well if the majority of individuals in the sample are unrelated.
-
-The presence of long tracts of homozygosity resulting from recent inbreeding adds an additional complication, as the number of shared chromosomes may be overestimated \cite{weir_genetic_2006}.
-For example, when considering two bones from the same individual, we would expect the entire genome to have a pairwise difference of $p_2$, because two out of the four compared chromosomes are identical copies.
-However, in inbred regions all four chromosomes will be identical, and so the expected pairwise differences are zero ($p_3$ in equation \ref{eq:2}).
-Note however, that $p_0$ does not depend on the presence of ROH, since all comparisons are between unrelated chromosomes even if both individuals are homozygous at a particular locus.
-
-Taken together, we can summarize $p$ in the following matrix, where rows give the state of $Z_w$, and columns of $H_w$:
-\begin{equation}\label{eq:2}
-    p(Z_w, H_w) = \left[\begin{array}
-{rrr}
-p_0 & p_0 & p_0 \\
-p_1 & p_2 & p_3 \\
-p_2 & p_3 & p_3
-\end{array}\right]
-\end{equation}
-
-As explained above, we would expect $p_3$ to be zero.
-However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that we obtain better results by setting $p_3 = \frac{p_2}{2}$.
-
-The effect of these considerations is that even though we have nine possible combinations of $Z_w$ and $H_w$ for each window, there are actually only four unique $p$-parameters $p_i$ with $i$ $\in$ (0,1,2,3).
-
-\paragraph{Beta Binomial Model}
-The Binomial Model laid out above assumes that sites are independent.
-However, due to genetic linkage, nearby sites will be correlated.
-As a result, we empirically find that the data often has considerably higher variance than would be expected from a binomial model (fig.
-S\ref{figS1:binom}).
-We take this into account by adding an over-dispersion parameter $\delta$.
-Just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ depends on the number of chromosomes compared, and so each of the  four $p_i$ has a corresponding $\delta_i$ parameter.
-
-Taken together, our emission probabilities are 
-\begin{align}\label{eq:3}
-P(D_{w}|Z_w,H_w,N_w, \delta) &\sim BB[D_w; p_i, \delta_i, N_w] \nonumber\\
-&= \binom{N_w}{D_w}\frac{B(D_w+p_i \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})},
-\end{align}
-where $i$ is fully determined by the combination of $Z_w$ and $H_w$ (see equation \ref{eq:2}).
-
-This parameterization of the beta distribution in terms of expected value $p$ and over-dispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parameterization in terms of $\alpha$ and $\beta$.
-We use this equation even if preprocessing steps result in non-integer $D_w$ and $N_w$, in which case we approximate the binomial coefficient using Gamma functions.
-
-\subsection{Estimation of $\delta$}\label{delta}
-We estimate the $\delta$-parameters using an Expectation-Maximization (EM) algorithm (REF). In an EM-algorithm, we iterate the Expectation and Maximization steps until the algorithm converges.
-
-\paragraph{Initialization}
-The value of $\delta_i$ is unknown to start with, and we set it to a random value between 0 and 1000.
-
-\paragraph{Expectation step}
-In the $t$-th iteration, we calculate the posterior probability of each IBD state in each window $\gamma^{(t)}_{wi} = P(Z_w=i | D_w, H_w, \theta_i, \delta_i^{(t)})$ using the forward-backward algorithm, where $\delta_i^{(t)}$ is the current estimate for $\delta$ for a given IBD state.
-
-\paragraph{Maximization step}
-The only free parameters we estimate in the maximization step are the over-dispersion parameters $\delta_i$.
-We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wj}$ and optionally the ROH state-probabilities $h_{w\omega}$ obtained from the ROH-HMM.
-
-\begin{align}\label{eq:4}
-\mathcal{C} &= \mathbb{E}[\log P(D_w|Z_w, H_w, \theta, \delta^{(t-1)}]\nonumber\\
-&= \sum_{w=1}^L \sum_{j=0}^2\sum^2_{\omega=0} \log P(D_{w}|N_w, \delta_{j \omega}^{(t-1)}, p_{j \omega})h_{w\omega}\gamma_{wj}
-\end{align}
-
-Using equation \ref{eq:2}, we simplify this by grouping all the terms that would result in the same $p_i$ and $\delta_i$, i.e.
-
-\begin{align*}\label{eq:5}
-g_{w0} &= \gamma_{w0}\\
-g_{w1} &= \gamma_{w1} h_{w0}\\
-g_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\\
-g_{w3} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2}
-\end{align*}
-
-So we can rewrite equation \ref{eq:4} as:
-\begin{align}\label{eq:6}
-\mathcal{C} &= \sum_{i=0}^3 \sum_{w=1}^L \log P(D_{w}|N_w, \delta_i, p_i)g_{wi}
-\end{align}
-
-The cost function \ref{eq:6} has one independent term for each $i$ and so we can separate them and estimate each $\delta_i$ independently using the minimize\_scalar algorithm implemented in \texttt{scipy.optimize} \cite{virtanen_scipy_2020}.
-
-%\paragraph{Constraints on $\delta$ optimization}\label{delta_constraint}
-
-Unconstrained optimization of the $\delta_i$ could result in some confounding of cases, We know that different cases of relatedness have different numbers of IBD states possible.
-For example, siblings may have all three IBD states present while a Parent-Child has only $Z_w=1$.
-However, the parent-child model could fit data generated under the sibling model by assigning it a very high $\delta_1$, which would reduce performance  (for example, see fig.
-S \ref{figS4:bndsbeta}).
-We avoid this problem by constraining the $\delta$ such that the beta distributions for the different relatedness cases overlap by at most one standard deviation.
-
-\subsection{Model comparison}\label{model_comp}
-To infer the most probable relatedness case, we run our model on all relatedness cases mentioned in section \ref{method_overview}, and compare the resulting likelihoods.
-We output the relatedness corresponding to the maximum likelihood model.
-Because the models we test are not nested, standard likelihood-ratio theory cannot be used to generate a confidence value.
-Instead we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig S \ref{figS10:cutoff}).
-
-\paragraph{Grouping of cases:}
-Particularly for low-quality data, we may not be able to distinguish all cases.
-Thus, we group $4^{th}$ and $5^{th}$ degree relatives together with unrelated.
-Similarly, we group  half-siblings, avuncular and grandparent-grandchild to  $2^{nd}$-degree relatives in the final results.
-We report the final pairwise classification in the following categories: Unrelated, Third Degree, Second Degree, Parent-Child, Siblings, Identical individuals.
-
-\paragraph{Critical Values:}
-
-To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff (fig.
-S \ref{figS10:cutoff}).
-The figure shows that for all relatedness cases except for $3^{rd}$-degree, the false positive rate is below $5\%$ even when simply selecting the model with the highest likelihood.
-We observe that for $3^{rd}$-degree relatives, using a cutoff of 1.0 brings down the false positive rate close to $5\%$ for all coverages except 0.05x.
-Thus, we  recommended using a cutoff of 1 for all cases where ROH and contamination are not a concern.
-
-\paragraph{Example case of siblings:}
-In Fig.\ref{fig1:ibd} we show the inferred IBD fragments from different KIN-HMMs when they are applied to simulated data from a pair of siblings.
-In this case, the siblings-model predictions match true IBD states the most, as reflected by the highest log-likelihood.
-
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/plotimg/IBDplot.png}
-    \centering
-    \caption{Comparison of pairwise difference data and inferred IBD fragments.
-The top panel shows the proportion of differences in each window along the genome for a pair of simulated siblings.
-Dashed lines represent $p_0$, $p_1$ and $p_2$ estimates.
-The second panel shows the true IBD state for each window.
-The remaining panels show the IBD states predicted by particular relatedness models.
-The log likelihood value for each model is shown on upper left corner of the panel.
-Light and shaded background represent distinct chromosomes.}
-    \label{fig1:ibd}
-\end{figure}
-
-\subsection{ROH estimation model}\label{roh}
-Our HMM to detect ROH tracts works similarly to the KIN-HMM described above, but we only consider one individual at a time, and only consider positions covered by at least two reads.
-For each site, we calculate the proportion of reads that carry different alleles, and sum them up in windows along the genome.
-We call the vector with the number of differences $\mathbf{\Delta}$, and the vector with the number of sites with at least two reads $\mathbf{N}$.
-Our model has two possible hidden states: homozygous state ($Y_w=3$), and non-homozygous state ($Y_w=2$).
-As above, we collect the hidden states in a vector $\mathbf{Y} = (Y_1, \dots Y_w, \dots, Y_L)$.
-The complete data likelihood for the model in this case is then:
-
-\begin{align}\label{eq:10}
-    log P(\mathbf{\Delta},\mathbf{Y}|\Theta,\delta) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta, \delta) + log P(\mathbf{Y}|\Theta)\nonumber\\
- &= [\sum_{w} log P(\Delta_w|Y_w, \Theta, \delta) + \sum_{w} log P(Y_w|Y_{w-1}, \Theta)] + P(Y_0| \Theta)
-\end{align}
-
-Since the source of ROH may not be known, we estimate both transitions and emissions.
-We calculate the emissions using a beta-binomial likelihood, and fix the mean of the distributions at expected proportion of differences in a homozygous tract ($p_3$) and expected proportion of differences in a non-homozygous tract ($p_2$).
-The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w=3$ or the state $Y_w=2$ in each window.
-The maximization step for emissions is analogous to that in the KIN model (eq.
-\ref{eq:4}), and the optimization step here is done with the following cost function:
-
-\begin{align}\label{eq:11}
-\mathcal{C} = \sum_{w=1}^L \sum_{i=2}^3 P(\Delta_w|\delta_{i},p_{i}) \Gamma_{wi} ,
-\end{align}
-where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and over-dispersion parameter $\delta$ similar to eq. \ref{eq:3}.
-
-To estimate transitions, we initialize the transition matrix with the value 0.2 for the off-diagonal entries, and update it using the standard Baum-Welch update step \cite{baum_maximization_1970}.
-
-Similar to the KIN-HMM, we avoid fitting issues by forcing all windows whose proportion of differences is larger than $p_2$ to be in the non-homozygous state (fig. S\ref{figS5:ROHforced}).
-
-\subsection{Contamination correction}\label{contam}
-Contamination by DNA from present-day people is a common feature of human ancient DNA datasets \cite{peyregne_present-day_2020}.
-To address this issue, we developed a heuristic that adjusts both $D_w$ and $N_w$ to minimize the influence of contamination on the relatedness inference.
-
-We assume that contamination rates in both individuals are known and small ($<5\%$), and set $C_{ij} \approx C_i + C_j$, where $C_i, C_j$ are the contamination estimates from the two individuals.
-We also assume the divergence $\phi$ between our
-target population and the putative contaminant popoulation is known.
-With probability $C_{ij}$, a comparison between two random reads from the pair of tested individual will contain a contaminant read, and thus contain a difference with probability $\phi$, and with probability $1 - C_{ij}$ it will be between endogenous ones.
-The comparisons between two contaminant reads are ignored, which is why we assume $C_{ij}$ to be small.
-
-We  estimate the expected number of differences from comparison of endogenous reads $\mathbb{E}[D_w^{'}]$, and the total number of sites with overlapping endogenous reads $\mathbb{E}[N_w^{'}]$.
-
-For any particular comparison showing a difference $D$, we calculate the probability of the event $E$ that it is between endogenous reads as 
-\begin{equation}
-    P(E|D) = \frac{P(D|E)P(E)}{P(D|E)P(E) + P(D| \neg E)P( \neg E)}.
-\end{equation}
-
-Then, by linearity of expectation, we obtain our estimator for the expected number of endogenous comparisons with a difference as 
-\begin{align}
-    \mathbb{E}[D_w^{'}] &= D_{w} P(E|D) = D_{w} \times \frac{P(D|E)P(E)}{P(D|E)P(E) + P(D| \neg E)P( \neg E)}
-\end{align}.
-
-Of these terms, $P(E)= 1 -C_{ij} = 1 - P(\neg E)$, and $P(D| \neg E) = \phi$.
-
-For $P(D|E)$, we use an estimator based on the genome-wide average:
-
-\begin{align}
-    P(D|E) = \rho = \frac{\frac{\sum_w D_w}{\sum_w N_w} - C_{ij} \phi}{1 - C_{ij}}.
-\end{align}
-
-Taken together,
-
-\begin{equation}
- \mathbb{E}[D_w^{'}]= D_{w}\frac{\rho (1-C_{ij})}{\rho(1-C_{ij}) + C_{ij}\phi}
- \end{equation}
-
- Analogous considerations lead to the expected number of endogenous comparisons that yield no difference:
-
- \begin{equation}
-\mathbb{E}[S_w^{'}] = S_{w} P(E| \neg D) = S_{w} \frac{(1-\rho)(1-C_{ij})}{(1-\rho)(1-C_{ij}) + C_{ij}(1-\phi)}.
- \end{equation}
-
-Here, $S_w = N_w - D_w$. Hence, we set $\mathbb{E}[N_w^{'}] = \mathbb{E}[S_w^{'}] + \mathbb{E}[D_w^{'}]$. We do a similar contamination correction for the input of ROH-HMM.
-
-\section{Results}\label{results}
-
-\subsection{Evaluation with simulations}
-We first tested the performance of KIN with simulated pedigrees.
-We performed coalescent simulations to generate 7 unrelated diploid genomes and artificially mated them to form pedigrees of 17 individuals with relationships up to $5^{th}$ degree (figure \ref{figS10:pedigree}).
-To evaluate the effect of sequence coverage on the performance of KIN, we generated artificial pileups at each polymorphic site for each individual following a Poisson distribution with 6 different average depths varying between 4$x$ and 0.03$x$ (see \ref{method}).
-To mimic ROH, for some pedigrees, we picked a single allele at heterozygous sites in some regions as determined by a Markov chain so that about $17\%$ of the genome is ROH.
-We also created versions with contamination, by introducing alleles from distantly related individuals, and ascertainment bias, by selecting polymorphic sites identified in a subset of the individuals (see \ref{method}).
-We created 60 pedigrees for each combination of average coverage and scenarios of presence/absence of ROH, contamination, and ascertainment bias, totalling 2,880 pedigrees.
-
-\paragraph{ROH detection}
-We first evaluated the performance of the ROH-HMM ( Fig.\ref{fig2:ROH}).
-We computed the proportion of windows that are in ROH and compared it to the posterior probability of the ROH state, by computing the bias (i.e., the difference between both) and its variance across windows.
-Without ROH, we correctly inferred the absence of ROH as the observed bias ranges from $0.3$ to $0.7\%$ (variance $\leq0.006$; Table \ref{tab:Table 2}).
-With ROH, we observed a slight bias of $5$ to $9\%$ more ROH than simulated (variance $\leq0.004$).
-The largest bias observed ($9\%$) was for the datasets that included contamination.
-Ascertainment bias had a minimal effect on the detection of ROH (Table \ref{tab:Table 2}).
-Surprisingly, the bias was lower for lower coverages, perhaps because there is only one possible ROH state per window to model a continuous variable, viz.
-proportion of window in ROH, and this state is inferred more confidently with more data.
-Overall, the model predicts ROH well in all cases, even at a coverage of 0.1x (Fig.\ref{fig2:ROH}).
-
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot_final.png}
-    \centering
-    \caption{Estimation of ROH probabilities along the genome in simulations.
-Top panel shows proportion of differences in a simulated individual in windows along the genome, and the bottom panel shows probability of seeing no ROH.
-(A) Simulation with no ascertainment, contamination, or ROH.
-(B) Simulation with ROH.
-(C) Simulation with ROH and ascertainment.
-(D) Simulation with ROH and contamination.}
-    \label{fig2:ROH}
-\end{figure}
-
-\begin{table}[h!]{ l p{5cm}}
-\caption{\label{tab:Table 2}Model performance for ROH prediction.
-Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination, R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
-\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
-    \hline
-    Coverage & Control(bias) & Control(var) & R(bias) & R(var) & RA(bias) & RA(var) & RC(bias) & RC(var)\\
-    \hline
-    4x & -0.007 & 0.006 & -0.081 & 0.036 & -0.083 & 0.036 & -0.091 & 0.039\\
-    \hline
-    0.2x & -0.003 & 0.003 & -0.068 & 0.03 & -0.068 & 0.026 & -0.076 & 0.029\\
-    \hline
-    0.1x & -0.003 & 0.002 & -0.049 & 0.021 & -0.05 & 0.021 & -0.057 & 0.022\\
-    \hline
-\end{tabular}
-\label{table1}
-\end{table}
-
-\paragraph{IBD prediction}
-We investigated the accuracy of IBD state prediction along the genome by counting the number of genomic windows where we correctly predict IBD state for different relatdness and coverages (Fig.
-\ref{fig4:IBDstate_accuracy}).
-The accuracy for coverages of 0.1x or higher is consistent and varies between 1.0 and 0.836, depending on relatedness.
-However, the accuracy decreases at lower coverages for most relatedness and ranges from 1.0 to 0.563 at 0.03x.
-On the other hand, the accuracy for identical individuals and parent-child pairs is nearly perfect for all investigated coverages, even at 0.03x coverage.
-We find that adding contamination, ascertainment bias and ROH in our simulations does not affect IBD prediction for identical and parent-child pairs, but reduces accuracy for unrelated individuals by xx, yy and zz respectively, for second degree by xx,yy,zz and for siblings by xx,yy,zz respectively.
-
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/plotimg/plot_IBDaccuracy.png}
-    \centering
-    \caption{Evaluation of IBD estimation at different coverages.
-y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis.
-The error bars are drawn at 1 standard deviation from the mean.
-Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
-    \label{fig4:IBDstate_accuracy}
-\end{figure}
-
-\paragraph{Relatedness classification}
-We finally evaluated the classification accuracy of KIN (cutoff: 1 log-likelihood unit) and compared it to that of READ (cutoff: 1 standard deviation)) (Fig.
-\ref{fig3:Comparison_READ_KIN}).
-We first describe the results for relatedness cases detectable by READ, viz.
-identical, $1_{st} degree$, $2_{nd} degree$ and $unrelated$.
-In this case we show that both methods have similar performance for low-coverage shotgun data (``control''-case), and for ascertained data.
-The true positive rate is above 0.97 for both KIN and READ, while the false negative rate is below 0.02.
-One exception is  $2^{nd}$-degree relatives, where KIN has higher power than READ, and as coverage decreases from 4x to 0.05x, KIN's true positive rate decreases from 0.94 to 0.86, compared to range of 0.91 to 0.74 for READ).
-The false positive rate in this case remains below 0.02 for both methods.
-We also find that READ is strongly impacted by contamination ($<3\%$) as the true positive rate is in the range 0.89 to 0.78 and 0.53 to 0.46 for 1st and 2nd degree relatives, respectively, and the false positive rate reaches up to 0.04 and 0.17 for unrelated individuals and 2nd degree relatives, respectively.
-By contrast, the correction implemented in KIN is sufficient to remove this bias, and the true and false positive rates remain as in the control.
-For simulations with ROH, we find that KIN also outperforms READ for 1st and 2nd degree relatives, although both the true and false positive rates increased for both methods compared to the control for 2nd degree relatives.
-The increases in false positives is likely due to ROH making individuals more similar (for details, see table S4).
-Finally, we also found that KIN has good power to detect relatedness cases that are not detectable by READ, i.e.
-parent-child, siblings and, to a lower level, 3rd degree relatives (Fig.
-\ref{fig3:Comparison_READ_KIN})
-
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/plotimg/comparison_plot.png}
-    \centering
-    \caption{Comparison of KIN with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH.
-Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
-'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
-    \label{fig3:Comparison_READ_KIN}
-\end{figure}
-
-\subsubsection{Application to real data}
-
-\paragraph{Chagyrskaya Neandertals}
-To test KIN on real ancient data, we applied it to a Neandertal dataset from Chagyrskaya Cave in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020,laurits_skov_genetic_nodate}.
-This dataset contains genetic data from  14 skeletal remains that likely belong to contemporary Neandertals who occupied the cave between 63 ± 4 and 48 ± 3 kya \cite{laurits_skov_genetic_nodate}.
-DNA extracted from each of these remains were captured with an array targeting variable sites identified in high coverage Neandertal and Denisovan genomes, and common variations in Africans \cite{laurits_skov_genetic_nodate}.
-This genetic data has  low-to-intermediate depth of coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage.
-Some of these specimens showed signs of long ROH, and DNA contamination from modern humans as well as hyenas \cite{laurits_skov_genetic_nodate}.
-We focused our analysis to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017} as done by the authors for the relatedness analysis \cite{laurits_skov_genetic_nodate}.
-Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}.
-We found three specimens from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60).
-Further, we identify Chagyrskaya17 and Chagyrskaya60 as 3rd-degree relatives.
-
-Our estimates are consistent with those obtained using READ, except that READ is unable to detect 3rd-degree relationships, and does not distinguish between sibling and parent-child relationships.
-We also find that both KIN and READ classify Chagyrskaya06 and Chagyrskaya14 as parent-child with low confidence, but from the morphology they likely stem from the same individual.
-We believe that the low confidence mis-classification may be due to non-human contamination present in these libraries ($1\%$ and $2.9\%$ respectively \cite{laurits_skov_genetic_nodate}), biasing the estimated differences between the individuals to higher values.
-
-We compared the IBD estimates obtained using KIN to those from lcMLkin for all pairs for whom READ results (with s.d.
-$>$1) and KIN results ($\Delta LL>$1) match (fig.
-S\ref{figS6:Chagyrskaya_ibd}).
-We find that lcMLkin creates four different clusters for different relatedness cases (identical, paret-child, $2^{nd}$ degree and unrelated).
-However, the location of these clusters strongly deviate from the expected values, likely due to the low coverage and ROH.
-Remains from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $k_0 \approx 0.33$, $r \approx 0.6$.
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/kinplot.png}
-    \centering
-    \caption{Application of KIN to Neandertal remains from Chagyrskaya.
-The color of a square represents the relatedness, while the number within denotes log likelihood ratio ($\Delta LL$) between the two maximum likelihood models.}
-    \label{fig5:Chagyrskaya_KIN}
-\end{figure}
-
-\subsubsection{Ancient modern humans}
-
-We further applied KIN to a genome-wide dataset of 118 ancient individuals from the Lech Valley \cite{mittnik_kinship-based_2019}.
-We compared our relatedness estimates to those obtained by the authors using READ and lcMLkin.
-We found that in most cases the three methods agree.
-Only $<5\%$ of pairs (315 out of 6903) show a different relatedness for KIN and READ (table S2).
-However, most of these are at very low confidence for both the methods ($\Delta LL>1$ for KIN and $sd<1$ for READ).
-There are 28 pairs where KIN shows confident results, and when we plot the raw heterozygosity estimates along the genome, in many cases the prediction by KIN is in agreement with the observed proportion of differences along the genome (fig.
-S\ref{figS8:eg1}).
-Comparisons with lcMLkin are spare since lcMLkin shows reliable results for $\appr 53\%$ pairs, and we find 16 pairs for which lcMLkin gives a reliable result (\cite{mittnik_kinship-based_2019}), and KIN shows $\Delta LL>$1.
-We again investigated these comparisons in more detail, and found that at least in cases where there is good amount of overlapping sites ($\geq20,000$), we can clearly see that proportion of differences along the genome aligns with prediction of KIN (fig.
-S\ref{figS9:eg2}).
-
-\section{Discussion}\label{discussion}
-Here, we present a new method called KIN to estimate genetic kinship and the location of IBD tracts from low-coverage data, in presence of long ROH, contamination, as well as ascertainment bias.
-Our method utilizes a set of HMMs to estimate IBD tracts and uses them to classify each pair of individuals into a possible relatedness case, along with a measure of classification confidence (fig.
-S \ref{fig0:schematic}).
-We evaluated the method performance of KIN, and compared it to that of READ using simulated pedigrees.
-Finally, we show applications of KIN on two ancient datasets.
-
-For detecting ROH, there is only one method available that works with low coverage ancient DNA \cite{ringbauer_parental_2021}.
-This software can infer ROH at coverages $>0.3x$, but requires a reference panel.
-In the absence of a method to detect ROH at low coverages without a reference panel, and we tested it on simulations to show that it can be used reliably at coverage $\geq 0.1x$ to detect ROH regions at lengths on the order of 10cM.
-In our simulations, we find that adding long ROH ($~17\%$) to simulations slightly improves the power of both KIN and READ, particularly at lower coverages.
-This is likely because ROH actually reduces the variance in differences, and so it becomes easier to estimate the location of IBD tracts.
-For READ, the presence of ROH causes a bias towards inferring closer relatedness cases, but the model we use in KIN reduces this bias (see Fig.
-\ref{fig3:Comparison_READ_KIN}).
-
-Contamination can affect the accuracy of relatedness inference for two reasons.
-For one, if a substantial fraction of samples is contaminated, then the estimation of $p_0$ becomes inaccurate, because the majority of pairwise differences will include at least one contaminated sample.
-The second issue is that contaminated samples will look less similar to other individuals, and thus cause a bias towards inferring them to be less related to other individuals.
-The amount of contamination does not need to be large for this to be important, in our simulations we find that even at contamination levels $\leq 3\%$, the performance of READ is substantially reduced.
-Assuming the contamination rates are correctly inferred, the correction we implemented leads to improved performance compared to naive methods such as READ, although they too could be ameliorated in a similar way \cite{laurits_skov_genetic_nodate}.
-SNP ascertainment can cause a slight decrease in power for both  methods, since it reduces the amount of data available.
-However, the decrease in power is lower for KIN compared to READ.
-
-The Lech Valley data  has low contamination, and no ROH.
-For pairwise comparisons with large numbers of overlapping sites ($>10000$), KIN, READ and lcMLkin all mostly agree.
-However, KIN is able to differentiate between parent-child and siblings, and identify second degree relationship from just a few thousand polymorphic sites ($\approx$ 4000) overlapping between samples.
-KIN can also infer third degree relation with $\approx$ 50,000 overlapping polymorphic sites.
-We show that when applied to Neandertal specimens from Chagyrskaya Cave, KIN identifies a pair of $1^{st}$-Degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples \cite{laurits_skov_genetic_nodate}.
-In addition, KIN identifies a pair of $3^{rd}$ degree relatives.
-In this case of a population with large amounts of ROH, we find that the inference by lcMLkin are heavily biased, but KIN's model takes ROH into account and both the coefficient of relatedness and $k_0$ are very close to what would be expected from the inference by both READ and KIN.
-
-One limitation of our approach is that it assumes a single population.
-In case of a highly structured population, KIN may show inaccurate inference of $p_0$ causing inaccurate relatedness inference.
-Also, our method makes the assumption, that the median pairwise genetic difference in the population reflects the population diversity $p_0$, which fails if almost all individuals in the dataset are related.
-The current implementation of KIN is restricted to six relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.Even though we do not differentiate between avuncular, grandparent-grandchild, half-siblings, we show that it is possible to differentiate between avuncular and grandparent-grandchild given that a pair is not half-sibling (see figure \ref{figS7:second_degree}).
-
-While we have focused on the application of KIN on ancient human samples, the model is not tied to this system.
-Assuming we know the recombination rate, and hence can estimate the transition matrix (see section \ref{method}), KIN can be widely applied to any diploid species.
-In addition, the output of KIN is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio.
-This makes KIN easy to automatize for large datasets.
-To make application of KIN user-friendly, we provide a python package to create input files for KIN from processed bamfiles, while optionally estimating ROH, and correcting for contamination estimates.
-
-\section{Materials and methods}\label{method}
-
-\subsection{Simulations}\label{simulat}
-We use simulations both for estimating the transition matrices and for testing and validating our algorithm.
-All simulations are performed in a scenario mimicking the analysis of a Neandertal population contaminated by modern humans.
-We simulate unrelated individuals using $\texttt{msprime}$ \cite{kelleher_efficient_2016}, followed by an additional step where we simulate related individuals using a predetermined pedigree (Figure S\ref{figS10:pedigree}).
-
-\paragraph{Simulating pedigrees}
-
-For our simulations of background diversity, we form a population (Pop1) with constant effective size of 10,000 and sample eight diploid individuals (each made up of two haploid individuals) from 2500 generations ago (fig.
-S \ref{figS10:pedigree}).
-For each individual, we simulate 22 chromosomes with length $L=100$ Mb and a recombination rate of $r=10^{-8}$ per base pair per generation.
-We introduce mutations using an infinite sites model with rate  $\mu= 10^{-8}$ per base pair per generation.
-
-For the pedigree simulations, we first simulate a recombined set of chromosomes for either parent, and combine them to create the progeny.
-We simulate recombination by first drawing the number of breakpoints as a  Poisson random variable with parameter $rL$, and use a uniform distribution on $[1, L]$ to sample the positions of recombination points.
-The pedigree simulations result in nine addition individuals, resulting in a total sample of 17 individuals.
-
-\paragraph{Transition matrices}
-KIN requires a transition matrix for each relatedness case.
-which we estimate  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of 1000 simulations from our pedigree (Fig.
-S \ref{figS10:pedigree}).
-For two cases, siblings and grandparent-grandchild, it is possible to write down the theoretical expectation of the transition matrix:
-
-For grandparent-grandchild, rate matrix is
- $$Q = \left[\begin{array}
-{rrr}
--r & r \\
-r & -r 
-\end{array}\right]$$
-
-Here the state space includes only $Z_w=0$ and $Z_w=1$. 
-Similarly, for a pair of siblings, we calculate $$Q =  
-\left[\begin{array}
-{rrr}
--4r & 4r & 0\\
-2r & -4r & 2r\\
-0 & 4r & -4r
-\end{array}\right]$$
-
-The state space includes all IBD states in this case. For these two cases, we get the transition matrix in each case as $e^{Qb}$, where b is the window size.
-
-\paragraph{Simulations for method evaluation}
-Apart from the related and unrelated individuals in Pop1, we simulated more haploid individuals in three other populations to create scenarios with ascertainment bias and contamination (fig.
-\ref{figS10:pedigree}).
-We simulated two individuals to form an individual each from two other populations (Pop2, Pop3) with split time of 3500 and 4500 generations with Pop1, and sampling time of 4000 generations and 2000 generations ago respectively.
-We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from Pop1.
-This scenario roughly models the ascertainment of the Chagyrskaya data.
-We tested the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a Markov chain using the transition matrix
-
-$$\mathbf{A} = \left[\begin{array}
-{rr}
-1-10/L & 10/L \\
-2/L & 1-2/L  \\
-\end{array}\right]
-$$
-
-From the steps described above, we got genotypes of individuals in Pop1 in presence/absence of ROH and ascertainment.
-We further simulated five diploid individuals from Pop4 with split time of 20,000 generations with Pop1, sampled from the present time, as a source of contamination.
-For cases including contamination, we contaminated eight individuals with varying amounts between 0.5\% to 3\% contamination, with the remaining  nine individuals did not have any contamination.
-We generated reads (derived/ancestral) for different genomic coverages ranging from 4x to 0.03x, assuming a Poisson distribution.
-
-For testing, we replicate each simulation 60 times, and create data from the same base simulations at varying levels of coverages and different scenarios: We have a control scenario (without ascertainment bias, ROH or contamination), and individual scenarios where we add ROH, SNP ascertainment and contamination, respectively.
-For the evaluation of the ROH-HMM, we also combine ROH with ascertainment or contamination.
-
-\section{Acknowledgements}
-We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Johann Visagie, Laurits Skov, Leonardo Iasi, Alba Bossom Mesa, Arev Sümer, Zuzana Hofmanová, Guido Alberto Gnecchi Ruscone, Ke Wang and Luca Traverso for helpful comments and discussions.
-This work was supported by the Max Planck Society and the European Research Council (Grant No.: 694707) to Svante Pääbo.
-
-\renewcommand\thefigure{\arabic{figure}}    
-\section{Supplementary Information}
-\setcounter{figure}{0}
+%\renewcommand\thefigure{\arabic{figure}}    
+\section*{Supplementary Figures}
+%\setcounter{figure}{0}
 
@@
     \centering
-    \caption{Application of READ on Neandertal specimens from Chagyrskaya Cave.
+    \caption{Application of READ on Neandertal specimens from Chagyrskaya and Okladnikov Caves.
 Color of a square represents the relatedness, while the number denotes standard deviations away from the upper threshold (we show lower threshold for unrelated pairs since upper threshold is not available).}
@@
     \centering
-    \caption{Application of identical relatedness HMM on a pair of identical individuals with low coverage (0.2x), and ROH tracts.
+    \caption{Application of Identical KIN-HMM on a pair of identical individuals with low coverage (0.2x), and ROH tracts.
 Top panel shows pairwise proportion of differences is close to expectation, but dips in some windows due to presence of ROH tracts.
@@
     \centering
-    \caption{Comparison of beta distributions estimated with the KIN-HMM (left) without and (right) with variance constrained optimization of $\delta$ parameters.The histogram in each plot shows pairwise differences in genomic windows.
+    \caption{Comparison of beta distributions estimated with the Identical KIN-HMM (left) without and (right) with variance constrained optimization of $\delta$ parameters.The histogram in each plot shows pairwise differences in genomic windows for identical individuals.
 Colored lines shows the Beta probability distribution for each IBD state $Z_w$.}
@@
     \caption{Comparison of Beta distributions estimated with ROH-HMM (left) without and (right) with constrained emissions.
-The histogram in each plot shows pairwise differences in genomic windows.
+The histogram in each plot shows pairwise differences in genomic windows for identical individuals.
 Colored lines shows the Beta probability distribution for each homozygosity state $Y_w$.}
@@
     \centering
-    \includegraphics[width=18cm]{plots/plotimg/plot_IBDaccuracy_asc.png}
-    \caption{Evaluation of IBD estimation in presence of ascertainment at different coverages.
-y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis.
+    \includegraphics[width=18cm]{plots/plotimg/plot_IBDaccuracy_others_supp.png}
+    \caption{Comparison of IBD estimation in control simulations and in presence of contamination, ascertainment or ROH (denoted in figure legend by Control, C, A and R, respectively) at different coverages.
+y-axis shows accuracy calculated over 60 simulations for each relatedness case, for each scenario (control, C, A or R), and six different coverages shown on x-axis.
 The error bars are drawn at 1 standard deviation from the mean.
 Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
-    \label{figS11:ibd_asc}
+    \label{figS11:ibd_others}
 \end{figure}
 
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=18cm]{plots/plotimg/plot_IBDaccuracy_roh.png}
-    \caption{Evaluation of IBD estimation in presence of ROH at different coverages.
-y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis.
-The error bars are drawn at 1 standard deviation from the mean.
-Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
-    \label{figS11:ibd_asc}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=18cm]{plots/plotimg/plot_IBDaccuracy_cont.png}
-    \caption{Evaluation of IBD estimation in presence of contamination at different coverages.
-y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis.
-The error bars are drawn at 1 standard deviation from the mean.
-Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
-    \label{figS11:ibd_asc}
-\end{figure}
 \begin{figure}[h!]
@@
     \label{figS6:Chagyrskaya_ibd}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/degree2_10Mwin.png}
-    \caption{Plotting total number of IBD fragments with the number of IBD fragments can tease apart avuncular relationship from grandparent-grandchild, since they form different clusters.
-Half-siblings, however, overlap considerably with both these relations.}
-    \label{figS7:second_degree}
 \end{figure}
@@
 
-\bibliographystyle{plain}
-\bibliography{references.bib}
-
 \end{document}

### 2022-05-15 paper_overleaf/main.tex
Updates from Overleaf

--- 2022-05-14 paper_overleaf/main.tex
@@
 
-We show with simulations that KIN infers first-degree relationships with nearly perfect accuracy, and second degree relationships with 90\% accuracy, even at 0.05x coverage.
+We show with simulations that KIN infers first-degree relationships with nearly perfect accuracy, and second degree relationships with over $90\%$ accuracy, even at 0.05x coverage.
 
@@
 The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals.
-For example, a pair of siblings are expected to have all three possible IBD states with proportions of 0.25,0.5,0.25, respectively (Fig \ref{fig0:schematic}).
+For example, a pair of siblings are expected to have all three possible IBD states with proportions of 0.25,0.5,0.25, respectively (Fig.
+\ref{fig0:schematic}).
 These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}.
@@
 However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be imprecise at lower coverages \cite{korneliussen_angsd_2014} .
-Ancient DNA analyses often face additional challenges such as contamination with present-day DNA \cite{peyregne_present-day_2020-1}, the unavailability of  reference panels to estimate population allele frequencies, and an ascertainment bias caused by  DNA capture approaches \cite{prufer_computational_2010, vai_kinship_2020}.
+Ancient DNA analyses often face additional challenges such as the unavailability of  reference panels to estimate population allele frequencies, contamination with present-day DNA \cite{peyregne_present-day_2020-1}, and an ascertainment bias caused by  DNA capture approaches \cite{prufer_computational_2010, vai_kinship_2020}.
 
@@
 KIN can detect up to \nth{3}-degree relatives, and differentiates between siblings and parent-child relationships.
-KIN is also able to take into account ROH, contamination and ascertainment.
+KIN is also able to take into account ROH and contamination, and is not sensitive to SNP ascertainment.
 We validate the performance of KIN using simulations and show that we are able to infer relatedness in real data from two datasets: a group of Neandertals and a group of Bronze age individuals.
@@
 If the locations of ROH tracts are unknown, we provide another HMM (ROH-HMM) to coarsely estimate the location of ROH for samples with sufficient coverage ($\geq 0.1x$).
-Our method is available on \url{https://github.com/DivyaratanPopli/KIN_snakemake} along with a \texttt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.
+Our method is available on \url{https://github.com/DivyaratanPopli/Kinship_Inference} along with a python package (KINgaroo) to generate the input files for the models directly from bam files.
 
 \subsection{Model description}\label{method_overview} 
-The goal of KIN-HMM is to infer how two individuals are related via the patterns of shared IBD-states along a pair of genomes.
+The goal of KIN-HMM is to infer how two individuals are related via the patterns of shared IBD states along a pair of genomes.
 For this purpose,  we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb), and infer the pattern of IBD-sharing for each relatedness case we consider (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{rd}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical).
@@
 For high-coverage data, this can be estimated accurately by comparing genotypes, but for low-coverage samples, $D_w$ needs to be estimated.
-The simplest approach is to  randomly-sample a read from each position \cite{haak_massive_2015,prufer_complete_2014}.
+The simplest approach is to randomly-sample a read from each position \cite{haak_massive_2015,prufer_complete_2014,green_draft_2010}.
 However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
@@
 The initial probabilities $\pi$ give the probabilities of being in each state $Z_0$ (at the beginning of each chromosome), which we set to uniform for simplicity.
-The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case.
+The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case (section \ref{simulat}).
 The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).
@@
 
-Here, $\BZ$ is not dependent on $\BH$, $\theta$ and $\delta$.
+Here, $\BZ$ is not dependent on $\BH$ and $\delta$.
 
@@
 
-\paragraph{Beta Binomial Model}
+\paragraph{Beta Binomial Model:}
 The Binomial Model laid out above assumes that sites are independent.
 However, due to genetic linkage, nearby sites will be correlated.
-As a result, we empirically find that the data often has considerably higher variance than would be expected from a binomial model (fig.
-S\ref{figS1:binom}).
+As a result, we empirically find that the data often has considerably higher variance than would be expected from a binomial model (Fig.
+S2).
 We take this into account by adding an over-dispersion parameter $\delta$.
@@
 \subsection{Estimation of $\delta$}\label{delta}
-We estimate the $\delta$-parameters using an Expectation-Maximization (EM) algorithm (REF). In an EM-algorithm, we iterate the Expectation and Maximization steps until the algorithm converges.
+We estimate the $\delta$-parameters using an Expectation-Maximization (EM) algorithm \cite{dempster_maximum_1977}.
 
-\paragraph{Initialization}
+\paragraph{Initialization:}
 The value of $\delta_i$ is unknown to start with, and we set it to a random value between 0 and 1000.
 
-\paragraph{Expectation step}
+\paragraph{Expectation step:}
 In the $t$-th iteration, we calculate the posterior probability of each IBD state in each window $\gamma^{(t)}_{wi} = P(Z_w=i | D_w, H_w, \theta_i, \delta_i^{(t)})$ using the forward-backward algorithm, where $\delta_i^{(t)}$ is the current estimate for $\delta$ for a given IBD state.
 
-\paragraph{Maximization step}
+\paragraph{Maximization step:}
 The only free parameters we estimate in the maximization step are the over-dispersion parameters $\delta_i$.
@@
 \mathcal{C} &= \mathbb{E}[\log P(D_w|Z_w, H_w, \theta, \delta^{(t-1)}]\nonumber\\
-&= \sum_{w=1}^L \sum_{j=0}^2\sum^2_{\omega=0} \log P(D_{w}|N_w, \delta_{j \omega}^{(t-1)}, p_{j \omega})h_{w\omega}\gamma_{wj}
+&= \sum_{w=1}^L \sum_{j=0}^2\sum^2_{\omega=0} \log P(D_{w}|N_w, \delta(j ,\omega)^{(t-1)}, p(j,\omega))h_{w\omega}\gamma_{wj}
 \end{align}
@@
 
-Unconstrained optimization of the $\delta_i$ could result in some confounding of cases, We know that different cases of relatedness have different numbers of IBD states possible.
+We constrain the optimization space of the $\delta_i$, as unconstrained optimization could result in some confounding of cases.
+We know that different cases of relatedness have different numbers of IBD states possible.
 For example, siblings may have all three IBD states present while a Parent-Child has only $Z_w=1$.
-However, the parent-child model could fit data generated under the sibling model by assigning it a very high $\delta_1$, which would reduce performance  (for example, see fig.
-S \ref{figS4:bndsbeta}).
+However, the parent-child model could fit data generated under the sibling model by assigning it a very high $\delta_1$, which would reduce performance (for example, see Fig.
+S5, S6).
 We avoid this problem by constraining the $\delta$ such that the beta distributions for the different relatedness cases overlap by at most one standard deviation.
@@
 \subsection{Model comparison}\label{model_comp}
-To infer the most probable relatedness case, we run our model on all relatedness cases mentioned in section \ref{method_overview}, and compare the resulting likelihoods.
+To infer the most likely relatedness case, we run our model on all relatedness cases mentioned in section \ref{method_overview}, and compare the resulting likelihoods.
 We output the relatedness corresponding to the maximum likelihood model.
 Because the models we test are not nested, standard likelihood-ratio theory cannot be used to generate a confidence value.
-Instead we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig S \ref{figS10:cutoff}).
+Instead we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig.
+S8).
 
@@
 
-To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff (fig.
-S \ref{figS10:cutoff}).
+To investigate the limits of our method, we plotted the true-positive and false-positive rates for classification of different relatedness in control simulations (without contamination, ROH and ascertainment bias, see section \ref{simulat}) when we use a particular difference in log-likelihood as a cutoff (fig.
+S8).
 The figure shows that for all relatedness cases except for $3^{rd}$-degree, the false positive rate is below $5\%$ even when simply selecting the model with the highest likelihood.
@@
 \subsection{ROH estimation model}\label{roh}
-Our HMM to detect ROH tracts works similarly to the KIN-HMM described above, but we only consider one individual at a time, and only consider positions covered by at least two reads.
+Our HMM to detect ROH tracts works similarly to the KIN-HMM described above, but in this case we only consider one individual at a time, and only consider positions covered by at least two reads.
 For each site, we calculate the proportion of reads that carry different alleles, and sum them up in windows along the genome.
-We call the vector with the number of differences $\mathbf{\Delta}$, and the vector with the number of sites with at least two reads $\mathbf{N}$.
+We call the vector with the number of differences $\mathbf{\Delta}$, and the vector with the number of sites with at least two reads $\mathbf{M}$.
 Our model has two possible hidden states: homozygous state ($Y_w=3$), and non-homozygous state ($Y_w=2$).
@@
     log P(\mathbf{\Delta},\mathbf{Y}|\Theta,\delta) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta, \delta) + log P(\mathbf{Y}|\Theta)\nonumber\\
- &= [\sum_{w} log P(\Delta_w|Y_w, \Theta, \delta) + \sum_{w} log P(Y_w|Y_{w-1}, \Theta)] + P(Y_0| \Theta)
+ &= [\sum_{w} log P(\Delta_w|Y_w, M, \delta) + \sum_{w} log P(Y_w|Y_{w-1}, A)] + P(Y_0| \pi),
 \end{align}
 
+where $\Theta$ is a vector of initial probabilities ($\pi$), transition matrix ($A$) and $M$.
+
 Since the source of ROH may not be known, we estimate both transitions and emissions.
-We calculate the emissions using a beta-binomial likelihood, and fix the mean of the distributions at expected proportion of differences in a homozygous tract ($p_3$) and expected proportion of differences in a non-homozygous tract ($p_2$).
+We calculate the emissions using a beta-binomial likelihood, and fix the mean of the distributions corresponding to $Y_w$ = 3 and $Y_w$ = 2 at expected proportion of differences in a homozygous tract ($p_3$) and expected proportion of differences in a non-homozygous tract ($p_2$) respectively.
 The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w=3$ or the state $Y_w=2$ in each window.
@@
 \end{align}
-where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and over-dispersion parameter $\delta$ similar to eq. \ref{eq:3}.
+where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and over-dispersion parameter $\delta$ similar to eq.
+\ref{eq:3}, and $\Gamma_{wi}$ is the posterior probability of homozygosity state $i$ at window $w$.
 
@@
 
-Similar to the KIN-HMM, we avoid fitting issues by forcing all windows whose proportion of differences is larger than $p_2$ to be in the non-homozygous state (fig. S\ref{figS5:ROHforced}).
+Similar to the KIN-HMM, we avoid fitting issues by forcing all windows whose proportion of differences is larger than $p_2$ to be in the non-homozygous state (Fig. S7).
 
@@
 With probability $C_{ij}$, a comparison between two random reads from the pair of tested individual will contain a contaminant read, and thus contain a difference with probability $\phi$, and with probability $1 - C_{ij}$ it will be between endogenous ones.
-The comparisons between two contaminant reads are ignored, which is why we assume $C_{ij}$ to be small.
+The comparisons between two contaminant reads are ignored, since we assume $C_{ij}$ to be small.
 
@@
 We first tested the performance of KIN with simulated pedigrees.
-We performed coalescent simulations to generate 7 unrelated diploid genomes and artificially mated them to form pedigrees of 17 individuals with relationships up to $5^{th}$ degree (figure \ref{figS10:pedigree}).
-To evaluate the effect of sequence coverage on the performance of KIN, we generated artificial pileups at each polymorphic site for each individual following a Poisson distribution with 6 different average depths varying between 4$x$ and 0.03$x$ (see \ref{method}).
+We performed coalescent simulations to generate 8 unrelated diploid genomes and artificially mated them to form pedigrees of 17 individuals with relationships up to $5^{th}$ degree (Fig.
+S3).
+To evaluate the effect of sequence coverage on the performance of KIN, we generated artificial pileups at each polymorphic site for each individual following a Poisson distribution with 6 different average depths varying between 4$x$ and 0.03$x$ (see section \ref{method}).
 To mimic ROH, for some pedigrees, we picked a single allele at heterozygous sites in some regions as determined by a Markov chain so that about $17\%$ of the genome is ROH.
-We also created versions with contamination, by introducing alleles from distantly related individuals, and ascertainment bias, by selecting polymorphic sites identified in a subset of the individuals (see \ref{method}).
+We also created versions with contamination, by introducing alleles from distantly related individuals, and ascertainment bias, by selecting polymorphic sites identified in a subset of the individuals (see section \ref{method}).
 We created 60 pedigrees for each combination of average coverage and scenarios of presence/absence of ROH, contamination, and ascertainment bias, totalling 2,880 pedigrees.
 
-\paragraph{ROH detection}
+\paragraph{ROH detection:}
 We first evaluated the performance of the ROH-HMM ( Fig.\ref{fig2:ROH}).
-We computed the proportion of windows that are in ROH and compared it to the posterior probability of the ROH state, by computing the bias (i.e., the difference between both) and its variance across windows.
+We computed the proportion of ROH in all windows and compared it to the posterior probability of the ROH state, by computing the bias (i.e., the difference between both) and its variance across windows.
 Without ROH, we correctly inferred the absence of ROH as the observed bias ranges from $0.3$ to $0.7\%$ (variance $\leq0.006$; Table \ref{tab:Table 2}).
@@
 
-\begin{table}[h!]{ l p{5cm}}
+\begin{table}[h!]
 \caption{\label{tab:Table 2}Model performance for ROH prediction.
@@
 
-\paragraph{IBD prediction}
+\paragraph{IBD prediction:}
 We investigated the accuracy of IBD state prediction along the genome by counting the number of genomic windows where we correctly predict IBD state for different relatdness and coverages (Fig.
@@
 We find that adding contamination, ascertainment bias in our simulations does not affect IBD prediction.
-Adding ROH to the simulations does not affect identical, parent-child, unrelated and $3^{rd} degree$  pairs, but reduces the IBD prediction accuracy for second degree by $1.5$ to $5\%$ and for siblings by $5$ to $17\%$ (fig\ref{figS11:ibd_others}.
+Adding ROH to the simulations does not affect identical, parent-child, unrelated and $3^{rd} degree$  pairs, but reduces the IBD prediction accuracy for second degree by $1.5$ to $5\%$ and for siblings by $5$ to $17\%$ (Fig.
+S9).
 We see this adverse effect of ROH in case of siblings, perhaps because only in case of siblings we do not introduce long ROH directly, but long and short ROH may be present through parents (see section \ref{method}).
 Unidentified short ROH may cause a difficulty in differentiating between different combinations of IBD ($Z_w$) and ROH states ($H_w$).
+However, this does not affect the power to identify siblings in presence of ROH even at $0.05x$ (see figure \ref{fig3:Comparison_READ_KIN})
 
@@
 
-\paragraph{Relatedness classification}
+\paragraph{Relatedness classification:}
 We evaluated the classification accuracy of KIN (cutoff: 1 log-likelihood unit) and compared it to that of READ (cutoff: 1 standard deviation)) (Fig.
@@
 We first describe the results for relatedness cases detectable by READ, viz.
-identical, $1_{st} degree$, $2_{nd} degree$ and $unrelated$.
+identical, $1^{st}$ degree, $2^{nd}$ degree and unrelated.
 In this case we show that both methods have similar performance for low-coverage shotgun data (``control''-case), and for ascertained data.
@@
 For simulations with ROH, we find that KIN also outperforms READ for 1st and 2nd degree relatives, although both the true and false positive rates increased for both methods compared to the control for 2nd degree relatives.
-The increases in false positives is likely due to ROH making individuals more similar (for details, see table S4).
+The increases in false positives is likely due to ROH making related individuals more similar.
 Finally, we also found that KIN has good power to detect relatedness cases that are not detectable by READ, i.e.
 parent-child, siblings and, to a lower level, 3rd degree relatives (Fig.
-\ref{fig3:Comparison_READ_KIN})
+\ref{fig3:Comparison_READ_KIN}, for details see table S4).
 
@@
     \caption{Comparison of KIN with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH.
-Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
+Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated (for fair comparison with READ).
 'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
@@
 
-\subsubsection{Application to real data}
+\subsection{Application to real data}
 
-\paragraph{Chagyrskaya Neandertals}
-To test KIN on real ancient data, we applied it to a Neandertal dataset from Chagyrskaya Cave in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020,laurits_skov_genetic_nodate}.
-This dataset contains genetic data from  14 skeletal remains that likely belong to contemporary Neandertals who occupied the cave between 63 ± 4 and 48 ± 3 kya \cite{laurits_skov_genetic_nodate}.
+\paragraph{Chagyrskaya Neandertals:}
+To test KIN on real ancient data, we applied it to a Neandertal dataset from Chagyrskaya and Okladnikov Caves in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020,laurits_skov_genetic_nodate}.
+This dataset contains genetic data from a total of 16 skeletal remains that likely belong to contemporary Neandertals who occupied the Chagyrskaya and Okladnikov caves between 59 and 51 kya, and atleast 44 kya respectively.
+\cite{laurits_skov_genetic_nodate}.
 DNA extracted from each of these remains were captured with an array targeting variable sites identified in high coverage Neandertal and Denisovan genomes, and common variations in Africans \cite{laurits_skov_genetic_nodate}.
@@
 We focused our analysis on the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017} as done by the authors for the relatedness analysis \cite{laurits_skov_genetic_nodate}.
-Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}.
+Our results for pairwise relatedness for these individuals are shown in Fig.\ref{fig5:Chagyrskaya_KIN}.
 We found three specimens from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60).
-Further, we identify Chagyrskaya17 and Chagyrskaya60 as 3rd-degree relatives.
+Further, we identify Chagyrskaya17 and Chagyrskaya60 as 3rd-degree relatives (table S1).
 
@@
 We also find that both KIN and READ classify Chagyrskaya06 and Chagyrskaya14 as parent-child with low confidence, but from the morphology they likely stem from the same individual.
-We believe that the low confidence mis-classification may be due to non-human contamination present in these libraries ($1\%$ and $2.9\%$ respectively \cite{laurits_skov_genetic_nodate}), biasing the estimated differences between the individuals to higher values.
+We believe that the low confidence mis-classification may be due to uncorrected non-human contamination present in these libraries ($1\%$ and $2.9\%$ respectively \cite{laurits_skov_genetic_nodate}), biasing the estimated differences between the individuals to higher values.
 
 We compared the IBD estimates obtained using KIN to those from lcMLkin for all pairs for whom READ results (with s.d.
-$>$1) and KIN results ($\Delta LL>$1) match (fig.
-S\ref{figS6:Chagyrskaya_ibd}).
-We find that lcMLkin creates four different clusters for different relatedness cases (identical, paret-child, $2^{nd}$ degree and unrelated).
+$>$1) and KIN results ($\Delta LL>$1) match (Fig.
+S10).
+We find that lcMLkin creates four different clusters for different relatedness cases (identical, parent-child, $2^{nd}$ degree and unrelated).
 However, the location of these clusters strongly deviate from the expected values, likely due to the low coverage and ROH.
-Remains from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $k_0 \approx 0.33$, $r \approx 0.6$.
+Remains from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $k_0 \approx 0.3$, $r \approx 0.6$.
 
@@
     \centering
-    \caption{Application of KIN to Neandertal remains from Chagyrskaya.
+    \caption{Application of KIN to Neandertal remains from Chagyrskaya and Okladnikov Caves.
 The color of a square represents the relatedness, while the number within denotes log likelihood ratio ($\Delta LL$) between the two maximum likelihood models.}
@@
 We found that in most cases the three methods agree.
-Only $<5\%$ of pairs (315 out of 6903) show a different relatedness for KIN and READ (table S2).
-However, most of these are at very low confidence for both the methods ($\Delta LL>1$ for KIN and $sd<1$ for READ).
-There are 28 pairs where KIN shows confident results, and when we plot the raw heterozygosity estimates along the genome, in many cases the prediction by KIN is in agreement with the observed proportion of differences along the genome (fig.
-S\ref{figS8:eg1}).
-Comparisons with lcMLkin are spare since lcMLkin shows reliable results for $\appr 53\%$ pairs, and we find 16 pairs for which lcMLkin gives a reliable result (\cite{mittnik_kinship-based_2019}), and KIN shows $\Delta LL>$1.
-We again investigated these comparisons in more detail, and found that at least in cases where there is good amount of overlapping sites ($\geq20,000$), we can clearly see that the proportion of differences along the genome aligns with the prediction of KIN (fig.
-S\ref{figS9:eg2}).
+Only $<5\%$ of pairs (315 out of 6903) show a different relatedness for KIN and READ.
+However, most of these are at very low confidence for both the methods ($\Delta LL<1$ for KIN and $sd<1$ for READ).
+There are 28 pairs where KIN shows confident results (table S2), and when we plot the raw heterozygosity estimates along the genome, in many cases it is possible to see that the prediction by KIN is in agreement with the observed proportion of differences along the genome (Fig.
+S11).
+Comparisons with lcMLkin are sparse since lcMLkin shows reliable results for $\appr 53\%$ pairs (\cite{mittnik_kinship-based_2019}), and we found 16 pairs for which lcMLkin gives a reliable result, and KIN shows $\Delta LL>$1 (table S3).
+We again investigated these comparisons in more detail, and found that at least in cases where there is good amount of overlapping sites ($\geq35,000$), we can clearly see that the proportion of differences along the genome aligns with the prediction of KIN (Fig.
+S12).
 
@@
 Here, we present a new method called KIN to estimate genetic kinship and the location of IBD tracts from low-coverage data, in presence of long ROH, contamination, as well as ascertainment bias.
-Our method utilizes a set of HMMs to estimate IBD tracts and uses them to classify each pair of individuals into a possible relatedness case, along with a measure of classification confidence (fig.
-S \ref{fig0:schematic}).
+Our method utilizes a set of HMMs to estimate IBD tracts and uses them to classify each pair of individuals into a possible relatedness case, along with a measure of classification confidence (Fig.
+S1).
 We evaluated the method performance of KIN, and compared it to that of READ using simulated pedigrees.
@@
 This software can infer ROH at coverages $>0.3x$, but requires a reference panel.
-In the absence of a method to detect ROH at low coverages without a reference panel, and we tested it on simulations to show that it can be used reliably at coverage $\geq 0.1x$ to detect ROH regions at lengths on the order of 10cM.
+In the absence of a method to detect long ROH at low coverages without a reference panel, we developed a HMM based method, and tested it on simulations to show that it can be used reliably at coverage $\geq 0.1x$ to detect ROH regions with lengths on the order of 10cM.
 In our simulations, we find that adding long ROH ($~17\%$) to simulations slightly improves the power of both KIN and READ, particularly at lower coverages.
-This is likely because ROH actually reduces the variance in differences, and so it becomes easier to estimate the location of IBD tracts.
+This is likely because ROH actually reduces the variance in differences depending on the relatedness, and makes it easier to correctly classify relatedness.
 For READ, the presence of ROH causes a bias towards inferring closer relatedness cases, but the model we use in KIN reduces this bias (see Fig.
+\ref{fig3:Comparison_READ_KIN}).
+
+We show that KIN reliably detects IBD tracts for upto 0.1x coverage even in presence of contamination and ascertainment bias, and the accuracy for IBD detection reduces with lower coverages.
+Adding ROH to the simulations adversely affects IBD prediction in siblings (Fig.
+S9), but this does not affect the power to correctly classify siblings (fig.
 \ref{fig3:Comparison_READ_KIN}).
@@
 Assuming the contamination rates are correctly inferred, the correction we implemented leads to improved performance compared to naive methods such as READ, although they too could be ameliorated in a similar way \cite{laurits_skov_genetic_nodate}.
-SNP ascertainment can cause a slight decrease in power for both  methods, since it reduces the amount of data available.
-However, the decrease in power is lower for KIN compared to READ.
 
@@
 However, KIN is able to differentiate between parent-child and siblings, and identify second degree relationship from just a few thousand polymorphic sites ($\approx$ 4000) overlapping between samples.
-KIN can also infer third degree relation with $\approx$ 50,000 overlapping polymorphic sites.
-We show that when applied to Neandertal specimens from Chagyrskaya Cave, KIN identifies a pair of $1^{st}$-Degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples \cite{laurits_skov_genetic_nodate}.
+KIN can also infer third degree relation with $\approx$ 30,000 overlapping polymorphic sites.
+We show that when applied to Neandertal specimens from Chagyrskaya and Okladnikov Caves, KIN identifies a pair of $1^{st}$-Degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples \cite{laurits_skov_genetic_nodate}.
 In addition, KIN identifies a pair of $3^{rd}$ degree relatives.
@@
 Also, our method makes the assumption, that the median pairwise genetic difference in the population reflects the population diversity $p_0$, which fails if almost all individuals in the dataset are related.
-The current implementation of KIN is restricted to six relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.Even though we do not differentiate between avuncular, grandparent-grandchild, half-siblings, we show that it is possible to differentiate between avuncular and grandparent-grandchild given that a pair is not half-sibling (see figure \ref{figS7:second_degree}).
+The current implementation of KIN is restricted to six relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.
 
@@
 This makes KIN easy to automatize for large datasets.
-To make application of KIN user-friendly, we provide a python package to create input files for KIN from processed bamfiles, while optionally estimating ROH, and correcting for contamination estimates.
+To make application of KIN user-friendly, we provide a python package (KINgaroo) to create input files for KIN from processed bamfiles, while optionally estimating ROH, and correcting for contamination estimates.
 
@@
 All simulations are performed in a scenario mimicking the analysis of a Neandertal population contaminated by modern humans.
-We simulate unrelated individuals using $\texttt{msprime}$ \cite{kelleher_efficient_2016}, followed by an additional step where we simulate related individuals using a predetermined pedigree (Figure S\ref{figS10:pedigree}).
+We simulate unrelated individuals using $\texttt{msprime}$ \cite{kelleher_efficient_2016}, followed by an additional step where we simulate related individuals using a predetermined pedigree (Fig.
+S1).
 
-\paragraph{Simulating pedigrees}
+\paragraph{Simulating pedigrees:}
 
 For our simulations of background diversity, we form a population (Pop1) with constant effective size of 10,000 and sample eight diploid individuals (each made up of two haploid individuals) from 2500 generations ago (fig.
-S \ref{figS10:pedigree}).
+S1).
 For each individual, we simulate 22 chromosomes with length $L=100$ Mb and a recombination rate of $r=10^{-8}$ per base pair per generation.
@@
 
-\paragraph{Transition matrices}
-KIN requires a transition matrix for each relatedness case.
-which we estimate  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of 1000 simulations from our pedigree (Fig.
-S \ref{figS10:pedigree}).
+\paragraph{Transition matrices:}
+KIN requires a transition matrix for each relatedness case, which we estimate by counting the transition between IBD states for all pairs of individuals with that relatedness in a training set of 1000 simulations from our pedigree (Fig.
+S1).
 For two cases, siblings and grandparent-grandchild, it is possible to write down the theoretical expectation of the transition matrix:
@@
 
-Here the state space includes only $Z_w=0$ and $Z_w=1$. 
 Similarly, for a pair of siblings, we calculate $$Q =  
 \left[\begin{array}
-{rrr}    bigd=rbind(bigd,df)
-  }
+{rrr}
 -4r & 4r & 0\\
@@
 
-\paragraph{Simulations for method evaluation}
-Apart from the related and unrelated individuals in Pop1, we simulated more haploid individuals in three other populations to create scenarios with ascertainment bias and contamination (fig.
-\ref{figS10:pedigree}).
+\paragraph{Simulations for method evaluation:}
+Apart from the related and unrelated individuals in Pop1, we simulated more haploid individuals in three other populations to create scenarios with ascertainment bias and contamination (Fig.
+S1).
 We simulated two individuals to form an individual each from two other populations (Pop2, Pop3) with split time of 3500 and 4500 generations with Pop1, and sampling time of 4000 generations and 2000 generations ago respectively.
 We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from Pop1.
-This scenario roughly models the ascertainment of the Chagyrskaya data.
-We tested the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a Markov chain using the transition matrix
+This scenario roughly models the ascertainment of the Chagyrskaya and Okladnikov Caves data.
+We tested the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity in unrelated individuals with a Markov chain using the transition matrix:
 
@@
 
+It is worth noting that we introduce long ROH in unrelated individuals, before artificially mating them to form pedigrees, which means that we do not directly introduce ROH in progeny, but it still affects relatedness inference among progeny as shown in Fig.
+\ref{fig0:schematic}
 From the steps described above, we got genotypes of individuals in Pop1 in presence/absence of ROH and ascertainment.
 We further simulated five diploid individuals from Pop4 with split time of 20,000 generations with Pop1, sampled from the present time, as a source of contamination.
-For cases including contamination, we contaminated eight individuals with varying amounts between 0.5\% to 3\% contamination, with the remaining  nine individuals did not have any contamination.
+For cases including contamination, we contaminated eight individuals with varying amounts between 0.5\% to 3\% contamination, while the remaining  nine individuals did not have any contamination.
 We generated reads (derived/ancestral) for different genomic coverages ranging from 4x to 0.03x, assuming a Poisson distribution.
 
-For testing, we replicate each simulation 60 times, and create data from the same base simulations at varying levels of coverages and different scenarios: We have a control scenario (without ascertainment bias, ROH or contamination), and individual scenarios where we add ROH, SNP ascertainment and contamination, respectively.
+For testing, we replicate our simulation 60 times, and create data from the same base simulations at varying levels of coverages and different scenarios: We have a control scenario (without ascertainment bias, ROH or contamination), and individual scenarios where we add ROH, SNP ascertainment and contamination.
 For the evaluation of the ROH-HMM, we also combine ROH with ascertainment or contamination.
@@
 
-\renewcommand\thefigure{\arabic{figure}}    
-\section{Supplementary Information}
-\setcounter{figure}{0}
+\section{Contributions}
+Conceptualization (Design of study): B.M.P.; Software: D.P.; Methodology—lead: B.M.P.; Methodology—support: D.P.; Formal Analysis: D.P.; Visualization-lead: D.P.; Visualization-support: S.P.; Data Curation: D.P.; Writing (original draft preparation)—lead: D.P.; Writing (original draft preparation)—support: B.M.P.; Writing (review and editing): input from all authors; Supervision: B.M.P.
 
-\renewcommand{\figurename}{Fig. S}
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic1.png}
-    \centering
-    \caption{Overall schematic of the method showing the entire pipeline from processed bam files to final relatedness and IBD estimates.
-Blue boxes show the data files, while green boxes represent scripts.}
-    \label{figS0:schematic}
-\end{figure}
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/binom.png}
-    \centering
-    \caption{Comparison of fit with Beta-binomial and Binomial distributions.
-Data is represented by the histogram of proportion of differences from all windows of a (A) unrelated, (B) Parent-Child, (C) Identical pair of individuals.
-Binomial parameter p is calculated from the data, while Beta-binomial parameters are estimated with the relatedness HMM.}
-    \label{figS1:binom}
-
-\end{figure}
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/pedigree_both.png}
-    \centering
-    \caption{Overview of simulated dataset.
-(A) We simulated four different populations with split times and sampling times (in generations) mimicking the populations of Chagyrskaya Neandertals, Vindija Neandertals, Altai Neandertals and present day Africans \cite{mafessoni_high-coverage_2020}.
-We artificially mated individuals in Pop1 to create related individuals.
-We used Pop2 and Pop3 to introduced ascertainment bias in the data, while we simulated contamination from modern humans using Pop4.
-(B) In Pop1, we sampled 8 diploid unrelated individuals, and artificially mated some of them to make a pedigree with upto $5^th$ Degree relatives.
-Circles represent females, and squares represent males.
-Dotted lines connect identical pair of individuals.}
-    \label{figS10:pedigree}
-
-\end{figure}
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/fil0_read_plot.png}
-    \centering
-    \caption{Application of READ on Neandertal specimens from Chagyrskaya Cave.
-Color of a square represents the relatedness, while the number denotes standard deviations away from the upper threshold (we show lower threshold for unrelated pairs since upper threshold is not available).}
-    \label{figS2:Chagyrskaya_READ}
-\end{figure}
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_emissions_bnds.png}
-    \centering
-    \caption{Application of identical relatedness HMM on a pair of identical individuals with low coverage (0.2x), and ROH tracts.
-Top panel shows pairwise proportion of differences is close to expectation, but dips in some windows due to presence of ROH tracts.
-Bottom two panels show $P(Data|Z_w=2)$ without constraints and with constraints respectively.}
-    \label{figS3:bnds}
-\end{figure}
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_betaplot.png}
-    \centering
-    \caption{Comparison of beta distributions estimated with the KIN-HMM (left) without and (right) with variance constrained optimization of $\delta$ parameters.The histogram in each plot shows pairwise differences in genomic windows.
-Colored lines shows the Beta probability distribution for each IBD state $Z_w$.}
-    \label{figS4:bndsbeta}
-\end{figure}
-
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_ind0_forced_roh.png}
-     \centering
-    \caption{Comparison of Beta distributions estimated with ROH-HMM (left) without and (right) with constrained emissions.
-The histogram in each plot shows pairwise differences in genomic windows.
-Colored lines shows the Beta probability distribution for each homozygosity state $Y_w$.}
-    \label{figS5:ROHforced}
-\end{figure}
-
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/plotimg/contam0_inbred0_model_performance_allroc_asc0_plot.png}
-    \centering
-    \caption{False positive and true positive rates as a function of cutoff on log likelihood ratio.
-In top panel, dotted line represents $5\%$.
-Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated.
-'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
-    \label{figS10:cutoff}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=18cm]{plots/plotimg/plot_IBDaccuracy_others_supp.png}
-    \caption{Evaluation of IBD estimation in presence of contamination, ascertainment or ROH (denoted in figure legend by C, A and R, respectively)at different coverages.
-y-axis shows accuracy calculated over 60 simulations for each relatedness case, for each scenario of C, A or R, and six different coverages shown on x-axis.
-The error bars are drawn at 1 standard deviation from the mean.
-Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
-    \label{figS11:ibd_others}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/lcPlot.png}
-    \caption{Comparison of IBD states estimated for Chagyrskaya specimens using (A) lcMLkin and (B) KIN.
-Each plot shows relatedness coefficient (calculated as $k_2+k_1/2$) on y-axis plotted against $k_0$ (proportion of genome with no chromosomes shared).
-    The relatedness shown with different colors is estimated with both READ and KIN.}
-    \label{figS6:Chagyrskaya_ibd}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/degree2_10Mwin.png}
-    \caption{Plotting total number of IBD fragments with the number of IBD fragments can tease apart avuncular relationship from grandparent-grandchild, since they form different clusters.
-Half-siblings, however, overlap considerably with both these relations.}
-    \label{figS7:second_degree}
-\end{figure}
-
-\begin{figure}[!ht]
-    \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/egplot1.png}
-    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIN differs from READ.
-(A) AITI62B-OTTM156 (READ estimates Identical individuals, lcMLkin does not have sufficient data, and KIN predicts parent-child.
-(B) UNTA5867-UNTA5868Sk1 (READ estimates Second Degree, lcMLkin does not have sufficient data, and KIN predicts parent-child).
-(C) AITI40-AITI72 (READ estimates Unrelated individuals, lcMLkin shows 3rd-5th degree, and KIN predicts 3rd degree.)
-    (D) AITI2-AITI55 (READ estimates Unrelated individuals, lcMLkin and KIN predicts predict second degree.)}
-    \label{figS8:eg1}
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/egplot2.png}
-    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIN differs from lcMLkin.
-(A) AITI43-AITI55 (READ estimates First degree, lcMLkin predicts siblings, while KIN predicts parent-child.
-(B) AITI70-AITI72 (READ estimates First degree, lcMLkin predicts parent-child, while KIN predicts siblings.}
-    \label{figS9:eg2}
-\end{figure}
+\section{Competing interests}
+The authors declare that they have no competing interests.
 

### 2022-05-16 supplementary_info/Rscripts/egplot2.R
added another pc-sib

@@ -10,6 +10,7 @@ egplot2 <- function(p1f, datadf, datatf, chrmf, outimg){
 
   ind1=1128+1 #AITI43_AITI55
   ind2=2247+1 #AITI70_AITI72
+  ind3=4801+1 #OBKR76_POST99
 
   p1 <- read_file(p1f)
   p1 <- as.double(gsub("[\r\n]", "", p1))
@@ -22,13 +23,14 @@ egplot2 <- function(p1f, datadf, datatf, chrmf, outimg){
 
   d1=datad[,ind1]/datat[,ind1]
   d2=datad[,ind2]/datat[,ind2]
-
+  d3=datad[,ind3]/datat[,ind3]
 
   wins=seq(1,300,1)
 
   chrm1=chrm%%2
   df1 <- data.frame(d1,wins,chrm1)
   df2 <- data.frame(d2,wins,chrm1)
+  df3 <- data.frame(d3,wins,chrm1)
 
 
   plot1 <- ggplot(data=df1, aes(x=wins,y=d1)) +
@@ -74,8 +76,8 @@ egplot2 <- function(p1f, datadf, datatf, chrmf, outimg){
     geom_line() +
     geom_segment(aes(x=1,xend=300,y=p1,yend=p1),linetype='dotted') + geom_segment(aes(x=1,xend=300,y=p12,yend=p12),linetype='dotted') +
     geom_segment(aes(x=1,xend=300,y=p34,yend=p34),linetype='dotted') +theme_bw() +
-    theme(legend.position = "none", axis.ticks.x=element_blank(), strip.text = element_text(size = 8))+
-    labs(y=expression(D[w]/N[w]), x = "Windows along the genome") +
+    theme(legend.position = "none", axis.ticks.x=element_blank(), , axis.title.x = element_blank(), strip.text = element_text(size = 8))+
+    labs(y=expression(D[w]/N[w])) +
     theme(panel.grid.major = element_blank(),
           panel.grid.minor = element_blank(),
           strip.background = element_blank(),
@@ -89,7 +91,37 @@ egplot2 <- function(p1f, datadf, datatf, chrmf, outimg){
              color="black", size=3)
 
 
-  panel<-plot_grid(plot1,plot2, labels= c("A","B"), ncol=1, rel_widths = c(1,1),rel_heights = c(1,1.1), align="v")
+ plot3 <- ggplot(data=df3, aes(x=wins,y=d3)) +
[…]

### 2022-05-16 paper_overleaf/KIN_a_method_to_estimate_relatedness_from_low_coverage_data_supplements.tex

@@
     \caption{Comparison of beta distributions estimated with the Identical KIN-HMM (left) without and (right) with variance constrained optimization of $\delta$ parameters.The histogram in each plot shows pairwise differences in genomic windows for identical individuals.
-Colored lines shows the Beta probability distribution for each IBD state $Z_w$.}
+Colored lines shows the Beta probability distribution for each $i$, where $i$ is the index used for combinations of $Z_W$ and $H_w$.}
     \label{figS4:bndsbeta}
@@
     \caption{Comparison of IBD estimation in control simulations and in presence of contamination, ascertainment or ROH (denoted in figure legend by Control, C, A and R, respectively) at different coverages.
-y-axis shows accuracy calculated over 60 simulations for each relatedness case, for each scenario (control, C, A or R), and six different coverages shown on x-axis.
-The error bars are drawn at 1 standard deviation from the mean.
+The y-axis shows accuracy calculated over 60 simulations for each relatedness case, for each scenario (control, C, A or R), and six different coverages shown on the x-axis.
+The error bars display one standard deviation from the mean.
 Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}

### 2022-05-16 paper_overleaf/main.tex

@@
 
-\paragraph{ROH detection:}
+\paragraph{ROH detection}
 We first evaluated the performance of the ROH-HMM ( Fig.\ref{fig2:ROH}).
@@
 The exceptions are for identical individuals and parent-child pairs, where the accuracy is nearly perfect for all investigated coverages, even at 0.03x coverage.
-We find that adding contamination, ascertainment bias in our simulations does not affect IBD prediction.
-Adding ROH to the simulations does not affect identical, parent-child, unrelated and $3^{rd} degree$  pairs, but reduces the IBD prediction accuracy for second degree by $1.5$ to $5\%$ and for siblings by $5$ to $17\%$ (Fig.
+We find that adding contamination, ascertainment bias in our simulations has little effect on IBD prediction.
+Adding ROH to the simulations reduces the IBD prediction in two cases: the accuracy for second degree is decreased by $1.5$ to $5\%$ and for siblings by $5$ to $17\%$ (Fig.
 S9).
@@
 We compared our relatedness estimates to those obtained by the authors using READ and lcMLkin.
-We found that in most cases the three methods agree.
-Only $<5\%$ of pairs (315 out of 6903) show a different relatedness for KIN and READ.
-However, most of these are at very low confidence for both the methods ($\Delta LL<1$ for KIN and $sd<1$ for READ).
-There are 28 pairs where KIN shows confident results (table S2), and when we plot the raw heterozygosity estimates along the genome, in many cases it is possible to see that the prediction by KIN is in agreement with the observed proportion of differences along the genome (Fig.
-S11).
-Comparisons with lcMLkin are sparse since lcMLkin shows reliable results for $\appr 53\%$ pairs (\cite{mittnik_kinship-based_2019}), and we found 16 pairs for which lcMLkin gives a reliable result, and KIN shows $\Delta LL>$1 (table S3).
-We again investigated these comparisons in more detail, and found that at least in cases where there is good amount of overlapping sites ($\geq35,000$), we can clearly see that the proportion of differences along the genome aligns with the prediction of KIN (Fig.
-S12).
+We found that KIN was able to confidently classify $85\%$ of the 6903 possible comparisons ($\Delta LL<1$), READ $94\%$ ($sd<1$) and lcMLkin 53\% of pairs (\cite{mittnik_kinship-based_2019}).
+
+Only for 28 pairs, KIN has confident classifications that differ from those obtained using READ (table S2, figure S11).
+For 20 of these comparisons, KIN predicts a third-degree relationship, which READ classifies as unrelated.
+
+lcMLkin predicts $3^{rd}$ - $5^th$ in 14 of these cases, unrelated in 1 case, second degree in 1 case, and does not have enough data for classification in 4 cases.
+Since READ does not infer third degree, and lcMLkin infers a group of relatedness together as $3^{rd}$ - $5^th$, it is likely that KIN is able to correctly resolve these cases.
+For 10 additional pairs, lcMLkin predicts $3^{rd}$ - $5^th$ degree, but KIN infers them to be unrelated.
+Thus, these are likely examples of $4^{th}$ or $5^{th}$ degree relations.
+
+For less than third-degree relations, only eight cases are classified differently between KIN and READ.
+lcMLkin matches KIN's prediction in two cases, and does not have enough data in 5 cases.
+In the last case, all three methods differ (table S3).
+Finally, there are three disagreements between KIN and lcMLkin in classification of parent-child versus siblings (READ predicts first degree for all three pairs).
+We plotted the pairwise differences for these three pairs in Fig.
+S12, and found that the proportion of differences along the genome aligns with the prediction of KIN.
 
@@
 S1).
-We simulated two individuals to form an individual each from two other populations (Pop2, Pop3) with split time of 3500 and 4500 generations with Pop1, and sampling time of 4000 generations and 2000 generations ago respectively.
+We simulated two individuals to form an individual each from two other populations (Pop2, Pop3) with split time of 3500 and 4500 generations with Pop1, and sampling time of 2000 generations and 4000 generations ago respectively.
 We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from Pop1.
@@
 \section{Acknowledgements}
-We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Johann Visagie, Laurits Skov, Leonardo Iasi, Alba Bossom Mesa, Arev Sümer, Zuzana Hofmanová, Guido Alberto Gnecchi Ruscone, Ke Wang and Luca Traverso for helpful comments and discussions.
+We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Johann Visagie, Laurits Skov, Leonardo N.
+M.
+Iasi, Alba Bossoms Mesa, Arev P.
+Sümer, Zuzana Hofmanová, Guido Alberto Gnecchi Ruscone, Ke Wang and Luca Traverso for helpful comments and discussions.
 This work was supported by the Max Planck Society and the European Research Council (Grant No.: 694707) to Svante Pääbo.
@@
 \section{Contributions}
-Conceptualization (Design of study): B.M.P.; Software: D.P.; Methodology—lead: B.M.P.; Methodology—support: D.P.; Formal Analysis: D.P.; Visualization-lead: D.P.; Visualization-support: S.P.; Data Curation: D.P.; Writing (original draft preparation)—lead: D.P.; Writing (original draft preparation)—support: B.M.P.; Writing (review and editing): input from all authors; Supervision: B.M.P.
+Conceptualization (Design of study): B.M.P.; Software: D.P.; Methodology—lead: D.P.; Methodology—support: B.M.P., S.P.; Formal Analysis: D.P.; Visualization-lead: D.P.; Visualization-support: S.P.; Data Curation: D.P.; Writing (original draft preparation)—lead: D.P.; Writing (original draft preparation)—support: B.M.P.; Writing (review and editing):  all authors; Supervision: B.M.P.
 

### 2022-05-16 supplementary_info/Rscripts/egplot1.R
changed pairs

@@ -8,10 +8,12 @@ library(readr)
 #chrmf <- "/mnt/diversity/divyaratan_popli/100arc/inbreeding/bronze_age_newSim/mergedwin_fil0/merged_chrm.csv"
 egplot1 <- function(p1f, datadf, datatf, chrmf, outimg){
 
-  ind1=1907+1 #AITI62B_OTTM156
-  ind2=6783+1 #UNTA5867_UNTA5868Sk1
-  ind3=1029+1 #AITI40_AITI72
-  ind4=244+1 #AITI2_AITI55
+  ind1=2987+1 #AITI95_AITI98
+  ind2=2343+1 #AITI72_AITI77A
+  ind3=6311+1 #POST131_POST28
+  ind4=3333+1 #ALT3_ALT4
+
+
   p1 <- read_file(p1f)
   p1 <- as.double(gsub("[\r\n]", "", p1))
   p12=p1/2
@@ -150,6 +152,8 @@ egplot1 <- function(p1f, datadf, datatf, chrmf, outimg){
 
 
   panel<-plot_grid(plot1,plot2,plot3,plot4, labels= c("A","B","C","D"), ncol=1, rel_widths = c(1,1,1,1),rel_heights = c(1,1,1,1.3), align="v")
+  #panel<-plot_grid(plot1,plot2,plot4, labels= c("A","B","C"), ncol=1, rel_widths = c(1,1,1,1),rel_heights = c(1,1,1,1.1), align="v")
+
   ggsave(outimg,
          width = 8, height = 5, dpi = 150, units = "in", device='png')
 }

### 2022-05-16 paper_overleaf/main.tex
Updates from Overleaf

@@
 
-Only for 28 pairs, KIN has confident classifications that differ from those obtained using READ (table S2, figure S11).
+Only for 28 pairs, KIN has confident classifications that differ from those obtained using READ (table S2, Fig.
+S11).
 For 20 of these comparisons, KIN predicts a third-degree relationship, which READ classifies as unrelated.
 
-lcMLkin predicts $3^{rd}$ - $5^th$ in 14 of these cases, unrelated in 1 case, second degree in 1 case, and does not have enough data for classification in 4 cases.
+lcMLkin predicts $3^{rd}$ - $5^th$ in 14 of these cases, unrelated in 1 case (see Fig.
+S11), second degree in 1 case (see Fig.
+S11), and does not have enough data for classification in 4 cases.
 Since READ does not infer third degree, and lcMLkin infers a group of relatedness together as $3^{rd}$ - $5^th$, it is likely that KIN is able to correctly resolve these cases.
@@
 lcMLkin matches KIN's prediction in two cases, and does not have enough data in 5 cases.
-In the last case, all three methods differ (table S3).
+In the last case, all three methods differ (table S3, Fig.
+S11).
 Finally, there are three disagreements between KIN and lcMLkin in classification of parent-child versus siblings (READ predicts first degree for all three pairs).

### 2022-05-16 paper_overleaf/KIN_a_method_to_estimate_relatedness_from_low_coverage_data_supplements.tex

@@
     \includegraphics[width=18cm]{supplementary_info/plots/egplot2.png}
-    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIN differs from lcMLkin.
-(A) AITI43-AITI55 (READ estimates First degree, lcMLkin predicts siblings, while KIN predicts parent-child.
-(B) AITI70-AITI72 (READ estimates First degree, lcMLkin predicts parent-child, while KIN predicts siblings.}
+    \caption{Plots showing proportion of differences in windows along the genome for the first degree relatives for which KIN differs from lcMLkin.
+We expect to see proportion of difference to be close to $p_0$ in $25\%$ windows, $p_1$ in $50\%$ windows, and $p_2$ in $25\%$ windows.
+For parent-child, we expect the proportion of differences to stay close to $p_1$ with some noise.
+(A) AITI43-AITI55 (READ estimates first degree, lcMLkin predicts siblings, while KIN predicts parent-child.
+(B) AITI70-AITI72 (READ estimates first degree, lcMLkin predicts parent-child, while KIN predicts siblings.
+(C) OBKR76-POST99 (READ estimates first degree, lcMLkin predicts siblings, while KIN predicts parent-child.}
     \label{figS9:eg2}

### 2022-05-16 plots/Rscripts/ROHlegend.R
made it high resolution

@@ -15,7 +15,7 @@ make_panel1 <- function(data,acrval,leg){
     )) +
     scale_fill_manual(values = c("grey90", "white")) +
     theme_void()
-  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.4, size=0.8)
+  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.7, size=0.8)
   plot1<-plot1 + theme_bw()+ theme(legend.position = "none", axis.title.x = element_blank())+labs(y= "Heterozygosity", x = "Windows along genome") +
     scale_color_manual(values=c("0.1"="Green", "0.2"="Blue","4"="Red")) +
     theme(panel.grid.major = element_blank(),
@@ -39,7 +39,7 @@ make_panel2 <- function(data,acrval,leg){
     theme_void()
   plot2<-plot2+geom_line(aes(x=Wins, y=True_ROH, linetype="Truth"))
   #plot2
-  plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.4, size=0.8)
+  plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.7, size=0.8)
 
   plot2<-plot2+scale_linetype_manual(values=c("Truth"="dashed", "Inferred"="solid"))
   plot2 <- plot2 + scale_color_manual(values=c("0.1"="Green", "0.2"="Blue","4"="Red"))

### 2022-05-16 plots/Rscripts/ROHplot.R

@@ -15,7 +15,7 @@ make_panel1 <- function(data,acrval,leg){
     )) +
     scale_fill_manual(values = c("grey90", "white")) +
     theme_void()
-  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.3, size=0.8)
+  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.7, size=0.8)
   plot1<-plot1 + theme_bw()+ theme(legend.position = "none", axis.title.x = element_blank())+labs(y= "Het.") +
     scale_color_manual(values=c("0.1"="Green", "0.2"="Blue","4"="Red")) +
     theme(panel.grid.major = element_blank(),
@@ -39,8 +39,8 @@ make_panel2 <- function(data,acrval,leg){
     theme_void()
   plot2<-plot2+geom_line(aes(x=Wins, y=True_ROH, linetype="Truth"))
   #plot2
-  plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.3, size=0.8)
-  
+  plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.7, size=0.8)
+
   plot2<-plot2+scale_linetype_manual(values=c("Truth"="dashed", "Inferred"="solid"))
   plot2 <- plot2 + scale_color_manual(values=c("0.1"="Green", "0.2"="Blue","4"="Red"))
   if (leg==1){
@@ -57,16 +57,16 @@ make_panel2 <- function(data,acrval,leg){
   #panel<-plot_grid(plot1,plot2, labels= c("A","B"), ncol=1, rel_widths = c(1,1))
   #panel
   return(plot2)
-  
+
 }
 
 
 ROH_plot <- function(infile,outfile){
-  
-  
+
+
   data=read.csv(file = infile, header = TRUE)
   data$Coverage <- as.factor(data$Coverage)
-  
+
   chrmlist=list()
   for (val in 1:264)
   {
@@ -76,27 +76,27 @@ ROH_plot <- function(infile,outfile){
     else if (val%%2==0){
       lchr=rep(1,10)
     }
-    
+
     chrmlist <- append(chrmlist, lchr)
   }
-  
-  
+
+
   data$chrm =as.integer(chrmlist)
-  
+
[…]

### 2022-05-16 paper_overleaf/KIN_a_method_to_estimate_relatedness_from_low_coverage_data_supplements.tex

@@
     \includegraphics[width=18cm]{supplementary_info/plots/egplot1.png}
-    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIN differs from READ.
-(A) AITI62B-OTTM156 (READ estimates Identical individuals, lcMLkin does not have sufficient data, and KIN predicts parent-child.
-(B) UNTA5867-UNTA5868Sk1 (READ estimates Second Degree, lcMLkin does not have sufficient data, and KIN predicts parent-child).
-(C) AITI40-AITI72 (READ estimates Unrelated individuals, lcMLkin shows 3rd-5th degree, and KIN predicts 3rd degree.)
-    (D) AITI2-AITI55 (READ estimates Unrelated individuals, lcMLkin and KIN predicts predict second degree.)}
+    \caption{Plots showing proportion of differences in windows along the genome for some pairs of relatives for which there is contradiction among KIN, READ and lcMLkin.
+(A) shows a case for which KIN and lcMLkin together disagree with READ, and (B) shows a case where KIN and lcMLkin differ, while READ does not have enough power.
+(C) and (D) show the two pairs for which all three methods disagree.
+We expect to see proportion of differences for second degree to be close to $p_0$ in $50\%$ windows, and $p_1$ in $50\%$ windows.
+For third degree we expect $75\%$ windows to be close to $p_0$, and $25\%$ windows to be close to $p_1$.
+For unrelated, we expect the proportion of differences to stay close to $p_0$ with some noise.
+(A) AITI95-AITI98 (READ estimates unrelated, lcMLkin and KIN predict second degree.
+(B) AITI72-AITI77A (READ estimates unrelated (READ does not distinguish between unrelated and third degree), lcMLkin shows unrelated, and KIN predicts third degree).
+(C) POST131-POST28 (READ estimates Unrelated, lcMLkin shows 3rd-5th degree, and KIN predicts second degree.) (D) ALT3-ALT4 (READ estimates unrelated, lcMLkin shows second degree and KIN predicts third degree.)}
     \label{figS8:eg1}
@@
     \caption{Plots showing proportion of differences in windows along the genome for the first degree relatives for which KIN differs from lcMLkin.
-We expect to see proportion of difference to be close to $p_0$ in $25\%$ windows, $p_1$ in $50\%$ windows, and $p_2$ in $25\%$ windows.
+We expect to see proportion of differences for siblings to be close to $p_0$ in $25\%$ windows, $p_1$ in $50\%$ windows, and $p_2$ in $25\%$ windows.
 For parent-child, we expect the proportion of differences to stay close to $p_1$ with some noise.

### 2022-05-17 plots/Rscripts/ROHlegend.R
Changed alpha

@@ -15,7 +15,7 @@ make_panel1 <- function(data,acrval,leg){
     )) +
     scale_fill_manual(values = c("grey90", "white")) +
     theme_void()
-  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.7, size=0.8)
+  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.5, size=0.8)
   plot1<-plot1 + theme_bw()+ theme(legend.position = "none", axis.title.x = element_blank())+labs(y= "Heterozygosity", x = "Windows along genome") +
     scale_color_manual(values=c("0.1"="Green", "0.2"="Blue","4"="Red")) +
     theme(panel.grid.major = element_blank(),
@@ -39,7 +39,7 @@ make_panel2 <- function(data,acrval,leg){
     theme_void()
   plot2<-plot2+geom_line(aes(x=Wins, y=True_ROH, linetype="Truth"))
   #plot2
-  plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.7, size=0.8)
+  plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.5, size=0.8)
 
   plot2<-plot2+scale_linetype_manual(values=c("Truth"="dashed", "Inferred"="solid"))
   plot2 <- plot2 + scale_color_manual(values=c("0.1"="Green", "0.2"="Blue","4"="Red"))

### 2022-05-17 plots/Rscripts/ROHplot.R

@@ -15,7 +15,7 @@ make_panel1 <- function(data,acrval,leg){
     )) +
     scale_fill_manual(values = c("grey90", "white")) +
     theme_void()
-  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.7, size=0.8)
+  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.5, size=0.8)
   plot1<-plot1 + theme_bw()+ theme(legend.position = "none", axis.title.x = element_blank())+labs(y= "Het.") +
     scale_color_manual(values=c("0.1"="Green", "0.2"="Blue","4"="Red")) +
     theme(panel.grid.major = element_blank(),
@@ -39,7 +39,7 @@ make_panel2 <- function(data,acrval,leg){
     theme_void()
   plot2<-plot2+geom_line(aes(x=Wins, y=True_ROH, linetype="Truth"))
   #plot2
-  plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.7, size=0.8)
+  plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.5, size=0.8)
 
   plot2<-plot2+scale_linetype_manual(values=c("Truth"="dashed", "Inferred"="solid"))
   plot2 <- plot2 + scale_color_manual(values=c("0.1"="Green", "0.2"="Blue","4"="Red"))

### 2022-05-17 paper_overleaf/KIN_a_method_to_estimate_relatedness_from_low_coverage_data_supplements.tex

@@
 Data is represented by the histogram of proportion of differences from all windows of a (A) unrelated, (B) Parent-Child, (C) Identical pair of individuals.
-Binomial parameter p is calculated from the data, while Beta-binomial parameters are estimated with the relatedness HMM.}
+Binomial parameter p is calculated from the data, while Beta-binomial parameters are estimated with corresponding KIN-HMM.}
     \label{figS1:binom}
@@
     \caption{Overview of simulated dataset.
-(A) We simulated four different populations with split times and sampling times (in generations) mimicking the populations of Chagyrskaya Neandertals, Vindija Neandertals, Altai Neandertals and present day Africans \cite{mafessoni_high-coverage_2020}.
+(A) We simulated four different populations with split times and sampling times (in generations) mimicking the populations of Chagyrskaya Neandertals, Vindija Neandertals, Altai Neandertals and present day Africans.
 We artificially mated individuals in Pop1 to create related individuals.
 We used Pop2 and Pop3 to introduced ascertainment bias in the data, while we simulated contamination from modern humans using Pop4.
-(B) In Pop1, we sampled 8 diploid unrelated individuals, and artificially mated some of them to make a pedigree with upto $5^th$ Degree relatives.
+(B) In Pop1, we sampled 8 diploid unrelated individuals, and artificially mated some of them to make a pedigree with upto $5^{th}$ Degree relatives.
 Circles represent females, and squares represent males.
@@
     \caption{Comparison of beta distributions estimated with the Identical KIN-HMM (left) without and (right) with variance constrained optimization of $\delta$ parameters.The histogram in each plot shows pairwise differences in genomic windows for identical individuals.
-Colored lines shows the Beta probability distribution for each $i$, where $i$ is the index used for combinations of $Z_W$ and $H_w$.}
+Colored lines shows the Beta probability distribution for each $i$, where $i$ is the index used for combinations of $Z_w$ and $H_w$.}
     \label{figS4:bndsbeta}
@@
     \caption{Plots showing proportion of differences in windows along the genome for some pairs of relatives for which there is contradiction among KIN, READ and lcMLkin.
-(A) shows a case for which KIN and lcMLkin together disagree with READ, and (B) shows a case where KIN and lcMLkin differ, while READ does not have enough power.
-(C) and (D) show the two pairs for which all three methods disagree.
+(A) shows a case for which KIN and lcMLkin together disagree with READ, and (B) shows a case where KIN and lcMLkin differ, while READ does not classify between contradicting relatedness categories.
+(C) and (D) show the two unresolved pairs for which all three methods disagree.
 We expect to see proportion of differences for second degree to be close to $p_0$ in $50\%$ windows, and $p_1$ in $50\%$ windows.
@@
 (A) AITI95-AITI98 (READ estimates unrelated, lcMLkin and KIN predict second degree.
-(B) AITI72-AITI77A (READ estimates unrelated (READ does not distinguish between unrelated and third degree), lcMLkin shows unrelated, and KIN predicts third degree).
+(B) AITI72-AITI77A (READ estimates unrelated, lcMLkin shows unrelated, and KIN predicts third degree).
 (C) POST131-POST28 (READ estimates Unrelated, lcMLkin shows 3rd-5th degree, and KIN predicts second degree.) (D) ALT3-ALT4 (READ estimates unrelated, lcMLkin shows second degree and KIN predicts third degree.)}

### 2022-05-18 pypackage/kin/KIN/__main__.py
changed contamfile head, added transitions,interval,p

@@ -16,10 +16,10 @@ from .hmm_scripts import helpers
 def cli():
     parser = argparse.ArgumentParser(description="Relatedness and IBD estimates")
 
-    parser.add_argument('-i', '--input_location',
+    parser.add_argument('-I', '--input_location',
                         type=str, metavar='',  required=True,
                         help='input files location')
-    parser.add_argument('-o', '--output_location',
+    parser.add_argument('-O', '--output_location',
                         type=str, metavar='', required=True,
                         help='Output files location')
     parser.add_argument('-T', '--target_location',
@@ -36,7 +36,10 @@ def cli():
                         help='Minimum number of sites in a window for ROH implementation')
     parser.add_argument('-p', '--diversity_parameter_p_0',
                         type=float, metavar='',
-                        help='Input p_0 parameter, if you do not want to calculate it from given samples')
+                        help='Input p_0 parameter, if you do not want to calculate it from given samples (Keep it same as that for KINgaroo)')
+    parser.add_argument('-i', '--interval',
+                        type=int, metavar='',
+                        help='Window size (Please choose the same size as for KINgaroo). By default:10000000')
 
     return parser.parse_args()
 
@@ -62,19 +65,25 @@ def main():
         p_0 = -9
     else:
         p_0 = args.diversity_parameter_p_0
+    if args.interval is None:
+        A_interval=C.A_ALL10
+    elif args.interval==10000000:
+        A_interval=C.A_ALL10
+    elif args.interval==1000000:
+        A_interval=C.A_ALL1
 
     helpers.hmm_all(
         targetfile=args.target_location,
         outfolder=args.output_location,
         datafolder=args.input_location,
-        paramfolder=C.TRANSITION,
         allrel=C.RELS,
         hbdfolder=hbdfolder,
         thresh=thresh,
         cores=cores,
         instates=C.STATES,
         totalch=C.CHRM,
-        p_0=p_0
+        p_0=p_0,
+        Afiles=A_interval
     )
 
 

### 2022-05-18 pypackage/kin/KIN/hmm_scripts/constants.py

@@ -4,7 +4,95 @@ Created on Fri Mar  4 10:44:57 2022
 
 @author: divyaratan_popli
 """
-TRANSITION="/mnt/diversity/divyaratan_popli/100arc/inbreeding/sim_relatedA/transitions_combined/"
+import numpy as np
+
 STATES=[0,1,2]
 CHRM=list(range(1,23))
 RELS=['un','deg5','deg4','deg3','gr','hsib','avu','sib','pc','id']
+
+#transitions for 10M windows
+A_un=np.array([[1.00000e-06, 9.99998e-01, 1.00000e-06],
+       [1.00000e-06, 9.99998e-01, 1.00000e-06],
+       [1.00000e-06, 9.99998e-01, 1.00000e-06]])
+
+A_deg5=np.array([[6.25593827e-01, 3.74405173e-01, 1.00000000e-06],
+       [2.51892991e-02, 9.74809701e-01, 1.00000000e-06],
+       [6.66666167e-01, 3.33332833e-01, 1.00000000e-06]])
+
+A_deg4=np.array([[7.53090860e-01, 2.46908140e-01, 1.00000000e-06],
+       [3.58180318e-02, 9.64180968e-01, 1.00000000e-06],
+       [6.66666167e-01, 3.33332833e-01, 1.00000000e-06]])
+
+A_deg3=np.array([[8.27500554e-01, 1.72498446e-01, 1.00000000e-06],
+       [5.86467863e-02, 9.41352214e-01, 1.00000000e-06],
+       [6.66666167e-01, 3.33332833e-01, 1.00000000e-06]])
+
+A_gr=np.array([[9.09364377e-01, 9.06346235e-02, 1.00000000e-06],
+       [9.06346235e-02, 9.09364377e-01, 1.00000000e-06],
+       [6.66665667e-01, 3.33333333e-01, 1.00000000e-06]])
+
+A_hsib=np.array([[8.35731279e-01, 1.64267721e-01, 1.00000000e-06],
+       [1.64387619e-01, 8.35611381e-01, 1.00000000e-06],
+       [6.66666167e-01, 3.33332833e-01, 1.00000000e-06]])
+
+A_avu=np.array([[8.05569361e-01, 1.94429639e-01, 1.00000000e-06],
+       [1.94943269e-01, 8.05055731e-01, 1.00000000e-06],
[…]

### 2022-05-18 pypackage/kin/KIN/hmm_scripts/helpers.py

@@ -12,20 +12,20 @@ import multiprocessing as mp
 from .hmm_functions import *
 
 
-def hmm_prep(targetfile, outfolder, datafolder, paramfolder, allrel, p_0):
-    
+def hmm_prep(targetfile, outfolder, datafolder, allrel, p_0):
+
     with open(targetfile) as f:
         libraries = [line.strip() for line in f]
-    
+
     listf=[]
     for i, l1 in enumerate(libraries):
         for l2 in libraries[(i+1):]:
             s = '%s_._%s' %(str(l1).strip("['']"), str(l2).strip("['']"))
             listf.append(s)
 
-    
+
     #make dirs
-    
+
     if not(os.path.isdir(outfolder)):
         os.mkdir(outfolder)
     if not(os.path.isdir(outfolder+"likfiles/")):
@@ -35,24 +35,23 @@ def hmm_prep(targetfile, outfolder, datafolder, paramfolder, allrel, p_0):
     if not(os.path.isdir(outfolder+"gammafiles/")):
         os.mkdir(outfolder+"gammafiles/")
 
-    
+
     if p_0==-9:
         pfile=datafolder + "hmm_parameters/p_0.txt"
         with open(pfile,"r") as f:
             p_0=float(f.read())
-    
+
     dfile=datafolder + "input_diffs_hmm.csv"
     tfile=datafolder + "input_total_hmm.csv"
-    
-    Afiles=[paramfolder + "transition_matrix_{}.csv".format(n) for n in allrel]
+
     p1thresh=datafolder+"input_hbd_hmm_total.csv"
-    
-    return libraries, listf, allrel, p_0, dfile, tfile, Afiles, p1thresh
+
+    return libraries, listf, allrel, p_0, dfile, tfile, p1thresh
 
 
 
 def hmm_run(hbdfolder, libraries, listf, instates, totalch, allrel, pfile, dfile, tfile, Afiles, p1thresh, outfolder, thresh, cores):
-    
+
     pool = mp.Pool(cores)
     [pool.apply_async(hmm, args=(listind, hbdfolder, dfile, tfile,
                             np.array(listf), np.array(libraries), pfile,
@@ -73,8 +72,8 @@ def hmm_results(outfolder, listf, allrel):
 
 
 
-def hmm_all(targetfile, outfolder, datafolder, paramfolder, allrel, hbdfolder, thresh, cores, instates, totalch, p_0):
[…]

### 2022-05-18 pypackage/kin/KIN/hmm_scripts/hmm_functions.py

@@ -359,9 +359,9 @@ def hmm(listind, hbdfolder, difffile, totalfile, listf, targets, pfile, Afiles, 
         chrmlist=chrmlist[gudwin]
 
         win=np.sum(gudwin)
-        
+
         p_1=float(pfile)
-    
+
         p_12=p_1/2
         inerror=p_12/2
         initial_p=np.array([[(p_1+p_12)/2,p_1,p_12],
@@ -379,8 +379,8 @@ def hmm(listind, hbdfolder, difffile, totalfile, listf, targets, pfile, Afiles, 
         likall=[]
         for rel_cnt in range(len(rels)):
 
-            Afile=Afiles[rel_cnt]
-            A=np.array(pd.read_csv(Afile,sep=',',header=None,index_col=False))
+            A=Afiles[rel_cnt]
+
             B = np.zeros((inbr_states,np.shape(A)[0],win)) #Emission probability
 
             [b0, b1, b2, b3]=[1000,1000,1000,1000]
@@ -391,7 +391,7 @@ def hmm(listind, hbdfolder, difffile, totalfile, listf, targets, pfile, Afiles, 
             B=makeB(data=data,xin=xin0, M=A.shape[0], T=len(data), inbr=inbr_states)
 
             np.argwhere(np.isnan(B))
-            
+
             gamma,A,B,up_p,lik, pi= baum_welch(data=data, hbd=hbd, A=A, B=B, pos=pos, p1avg=initial_p, inbr=inbr_states, x0=x0)
             res=viterbi(data, A, B, pi)
             likall.append(lik)

### 2022-05-18 pypackage/kingaroo/build/lib/KINgaroo/KINgaroo_scripts/helpers.py

@@ -180,7 +180,7 @@ def contamFile(infile, outfile, targets, idfile):
    for ch1 in range(len(cnf)-1):
        for ch2 in range(ch1+1,len(cnf)):
            names.append(cnf.loc[ch1, 'name'] + '_._' + cnf.loc[ch2, 'name'])
-           ctotal=cnf.loc[ch1, 'contamination'] + cnf.loc[ch2, 'contamination']
+           ctotal=cnf.loc[ch1, 'contamination (%)'] + cnf.loc[ch2, 'contamination (%)']
            contam.append(ctotal)
 
    df = pd.DataFrame(
@@ -190,7 +190,7 @@ def contamFile(infile, outfile, targets, idfile):
 
    df_id = pd.DataFrame(
        {'name': cnf['name'],
-        'contam': 2*cnf['contamination']
+        'contam': 2*cnf['contamination (%)']
        })
 
    with pd.option_context('display.max_rows', len(df.index), 'display.max_columns', len(df.columns)):

### 2022-05-18 pypackage/kingaroo/KINgaroo/KINgaroo_scripts/helpers.py
uncommented stuff

@@ -290,13 +290,13 @@ def pipeline1(targetsfile, bedfile, cores, rawbams, interval, splitbams, bedfile
 
     libraries, listf = prep_function(targetsfile, splitbams, bedfiles, hapProbs, hmm_param, hbdf, likf)
 
-    #split_bed(bedfile=bedfile, totalch=chrmf, bedfiles=bedfiles)
+    split_bed(bedfile=bedfile, totalch=chrmf, bedfiles=bedfiles)
 
-    #lib_chrm_all=create_lib_chrm(libraries=libraries, totalch=chrmf)
+    lib_chrm_all=create_lib_chrm(libraries=libraries, totalch=chrmf)
 
-    #parallel_indexes(rawbams=rawbams, libraries=libraries, cores=cores)
+    parallel_indexes(rawbams=rawbams, libraries=libraries, cores=cores)
 
-    #parallel_bamfilter(rawbams=rawbams, splitbams=splitbams, bedfiles=bedfiles, hapProbs=hapProbs, lib_chrm_all=lib_chrm_all, cores=cores)
+    parallel_bamfilter(rawbams=rawbams, splitbams=splitbams, bedfiles=bedfiles, hapProbs=hapProbs, lib_chrm_all=lib_chrm_all, cores=cores)
 
     dwins,twins,id_dwins,id_twins, chrmlist = parallel_mergedchrm(libraries=libraries, totalch=chrmf, interval=interval, cores=cores)
 

### 2022-05-18 paper_overleaf/main.tex

@@
 
-We show with simulations that KIN infers first-degree relationships with nearly perfect accuracy, and second degree relationships with over $90\%$ accuracy, even at 0.05x coverage.
+We show with simulations that KIN infers first-degree relationships with nearly perfect accuracy, and second degree relationships with about $90\%$ accuracy, even at 0.05x coverage.
 
 To illustrate its utility, we apply KIN to a Neandertal group from Siberia, Russia, and a group of Bronze-Age people from the Lech Valley, Germany.
-We find that in both cases, KIN matches the findings of existing approaches and detects several additional related individuals.
+We find that in both cases, KIN matches the findings of existing approaches and detects several additional relationships.
 
@@
 However, since it is not possible to directly observe IBD segments, a common approach is to first identify segments of the genome that are Identical by State (IBS) and to use population allele frequencies obtained from an out-of-sample reference panel to calculate the probability of IBD given IBS.
-There are several methods that incproporate reference panel allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007,manichaikul_robust_2010,gusev_whole_2009,nyerki_optimized_2022}.
+There are several methods that incproporate reference panel allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007,manichaikul_robust_2010,gusev_whole_2009,nyerki_optimized_2022}.
 
@@
 Moreover, ancient DNA is commonly captured with a SNP array that enriches for informative variants.
-Particularly methods based on the fraction of sites in different IBS states are sensitive to the ascertainment bias caused by this non-random selection of targeted sites \cite{waples_allele_2019}.
+Particularly, methods based on the fraction of sites in different IBS states are sensitive to the ascertainment bias caused by this non-random selection of targeted sites \cite{waples_allele_2019}.
 
@@
 We then compare the likelihood between all models, and classify each pair to the model with the highest likelihood.
-We also return the most likely locations of IBD tracts (using standard Viterbi algorithm), and the state posterior probabilities $\BZ$.
+We also return the most likely locations of IBD tracts (using standard Viterbi algorithm), and the IBD state posterior probabilities.
 
@@
 \begin{align}\label{eq:4}
-\mathcal{C} &= \mathbb{E}[\log P(D_w|Z_w, H_w, \theta, \delta^{(t-1)}]\nonumber\\
+\mathcal{C} &= \mathbb{E}[\log P(D_w|Z_w, H_w, \theta, \delta^{(t-1)})]\nonumber\\
 &= \sum_{w=1}^L \sum_{j=0}^2\sum^2_{\omega=0} \log P(D_{w}|N_w, \delta(j ,\omega)^{(t-1)}, p(j,\omega))h_{w\omega}\gamma_{wj}
@@
 
-The cost function \ref{eq:6} has one independent term for each $i$ and so we can separate them and estimate each $\delta_i$ independently using the minimize\_scalar algorithm implemented in \texttt{scipy.optimize} \cite{virtanen_scipy_2020}.
+The cost function (equation \ref{eq:6}) has one independent term for each $i$ and so we can separate them and estimate each $\delta_i$ independently using the minimize\_scalar algorithm implemented in \texttt{scipy.optimize} \cite{virtanen_scipy_2020}.
 
@@
 
-To investigate the limits of our method, we plotted the true-positive and false-positive rates for classification of different relatedness in control simulations (without contamination, ROH and ascertainment bias, see section \ref{simulat}) when we use a particular difference in log-likelihood as a cutoff (fig.
+To investigate the limits of our method, we plotted the true-positive and false-positive rates for classification of different relatedness in control simulations (without contamination, ROH and ascertainment bias, see section \ref{simulat}) when we use a particular difference in log-likelihood as a cutoff (Fig.
 S8).
@@
 \end{align}
-where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and over-dispersion parameter $\delta$ similar to eq.
-\ref{eq:3}, and $\Gamma_{wi}$ is the posterior probability of homozygosity state $i$ at window $w$.
+where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and over-dispersion parameter $\delta$ similar to eq. \ref{eq:3}.
 
@@
 
-\paragraph{ROH detection}
+\paragraph{ROH detection:}
 We first evaluated the performance of the ROH-HMM ( Fig.\ref{fig2:ROH}).
@@
 \begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot_final.png}
+    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot.png}
     \centering
@@
     \hline
-    4x & -0.007 & 0.006 & -0.081 & 0.036 & -0.083 & 0.036 & -0.091 & 0.039\\
+    4x & 0.007 & 0.006 & 0.081 & 0.036 & 0.083 & 0.036 & 0.091 & 0.039\\
     \hline
-    0.2x & -0.003 & 0.003 & -0.068 & 0.03 & -0.068 & 0.026 & -0.076 & 0.029\\
+    0.2x & 0.003 & 0.003 & 0.068 & 0.03 & 0.068 & 0.026 & 0.076 & 0.029\\
     \hline
-    0.1x & -0.003 & 0.002 & -0.049 & 0.021 & -0.05 & 0.021 & -0.057 & 0.022\\
+    0.1x & 0.003 & 0.002 & 0.049 & 0.021 & 0.05 & 0.021 & 0.057 & 0.022\\
     \hline
@@
 Unidentified short ROH may cause a difficulty in differentiating between different combinations of IBD ($Z_w$) and ROH states ($H_w$).
-However, this does not affect the power to identify siblings in presence of ROH even at $0.05x$ (see figure \ref{fig3:Comparison_READ_KIN})
+However, this does not affect the power to identify siblings in presence of ROH even at $0.05x$ (see Fig.
+\ref{fig3:Comparison_READ_KIN})
 
@@
 
-\paragraph{Chagyrskaya Neandertals:}
+\paragraph{Chagyrskaya and Okladnikov Neandertals:}
 To test KIN on real ancient data, we applied it to a Neandertal dataset from Chagyrskaya and Okladnikov Caves in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020,laurits_skov_genetic_nodate}.
-This dataset contains genetic data from a total of 16 skeletal remains that likely belong to contemporary Neandertals who occupied the Chagyrskaya and Okladnikov caves between 59 and 51 kya, and atleast 44 kya respectively.
-\cite{laurits_skov_genetic_nodate}.
+This dataset contains genetic data from a total of 16 skeletal remains that likely belong to contemporary Neandertals who occupied the Chagyrskaya and Okladnikov caves between 59 and 51 kya, and atleast 44 kya respectively \cite{laurits_skov_genetic_nodate}.
 DNA extracted from each of these remains were captured with an array targeting variable sites identified in high coverage Neandertal and Denisovan genomes, and common variations in Africans \cite{laurits_skov_genetic_nodate}.
@@
 
-lcMLkin predicts $3^{rd}$ - $5^th$ in 14 of these cases, unrelated in 1 case (see Fig.
+lcMLkin predicts $3^{rd}$ - $5^{th}$ in 14 of these cases, unrelated in 1 case (see Fig.
 S11), second degree in 1 case (see Fig.
 S11), and does not have enough data for classification in 4 cases.
-Since READ does not infer third degree, and lcMLkin infers a group of relatedness together as $3^{rd}$ - $5^th$, it is likely that KIN is able to correctly resolve these cases.
-For 10 additional pairs, lcMLkin predicts $3^{rd}$ - $5^th$ degree, but KIN infers them to be unrelated.
+Since READ does not infer third degree, and lcMLkin infers a group of relatedness together as $3^{rd}$ - $5^{th}$, it is likely that KIN is able to correctly resolve these cases.
+For 10 additional pairs, lcMLkin predicts $3^{rd}$ - $5^{th}$ degree, but KIN infers them to be unrelated.
 Thus, these are likely examples of $4^{th}$ or $5^{th}$ degree relations.
@@
 For detecting ROH, there is only one method available that works with low coverage ancient DNA \cite{ringbauer_parental_2021}.
-This software can infer ROH at coverages $>0.3x$, but requires a reference panel.
+This software can infer ROH at coverages $\geq0.3x$, but requires a reference panel.
 In the absence of a method to detect long ROH at low coverages without a reference panel, we developed a HMM based method, and tested it on simulations to show that it can be used reliably at coverage $\geq 0.1x$ to detect ROH regions with lengths on the order of 10cM.
@@
 Adding ROH to the simulations adversely affects IBD prediction in siblings (Fig.
-S9), but this does not affect the power to correctly classify siblings (fig.
+S9), but this does not affect the power to correctly classify siblings (Fig.
 \ref{fig3:Comparison_READ_KIN}).
@@
 We use simulations both for estimating the transition matrices and for testing and validating our algorithm.
-All simulations are performed in a scenario mimicking the analysis of a Neandertal population contaminated by modern humans.
+All simulations are performed in a scenario mimicking the analysis of a Neandertal population contaminated by modern humans \cite{mafessoni_high-coverage_2020}.
 We simulate unrelated individuals using $\texttt{msprime}$ \cite{kelleher_efficient_2016}, followed by an additional step where we simulate related individuals using a predetermined pedigree (Fig.
@@
 It is worth noting that we introduce long ROH in unrelated individuals, before artificially mating them to form pedigrees, which means that we do not directly introduce ROH in progeny, but it still affects relatedness inference among progeny as shown in Fig.
-\ref{fig0:schematic}
+\ref{fig0:schematic}.
 From the steps described above, we got genotypes of individuals in Pop1 in presence/absence of ROH and ascertainment.
@@
 \section{Contributions}
-Conceptualization (Design of study): B.M.P.; Software: D.P.; Methodology—lead: D.P.; Methodology—support: B.M.P., S.P.; Formal Analysis: D.P.; Visualization-lead: D.P.; Visualization-support: S.P.; Data Curation: D.P.; Writing (original draft preparation)—lead: D.P.; Writing (original draft preparation)—support: B.M.P.; Writing (review and editing):  all authors; Supervision: B.M.P.
+Conceptualization (Design of study): B.M.P.; Software: D.P.; Methodology—lead: D.P.; Methodology—support: B.M.P., S.P.; Formal Analysis: D.P.; Visualization-lead: D.P.; Visualization-support: S.P.; Data Curation: D.P.; Writing—lead: D.P.; Writing—support: B.M.P., S.P.; Supervision: B.M.P.
 

### 2022-05-18 Simulations/KIN_hmm/hmm_inbred_constVar_testing.py
added data generation files with example

@@ -1,0 +1,579 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Jan 21 21:36:58 2021
+
+@author: divyaratan_popli
+"""
+
+
+import scipy
+import numpy as np
+import random
+import pandas as pd
+from scipy.stats import binom
+from math import isclose
+from scipy.special import gammaln
+from scipy import optimize
+import math
+import matplotlib.pyplot as plt
+from scipy.special import loggamma
+from scipy.special import beta as Betaf
+from scipy.optimize import minimize_scalar
+import pylab
+from numpy.linalg import matrix_power
+from scipy.special import logsumexp
+import numba
+
+#print(scipy.__version__)
+
+
+
+"""
+#to stop the program if there is a warning
+import warnings
+
+with warnings.catch_warnings():
+    warnings.simplefilter('error')
+    for chrm in range(chrm_no):
+
+            for t in range(pos[chrm],pos[chrm+1]-1):
+
+
+                for i in range(M):
+                    print.info(chrm,t,i)
+                    numerator_l = np.log(alpha[t, i]) + np.log(A[i, :]) + np.log(B[:, t + 1].T) + np.log(beta[t + 1, :].T) - np.log(scale[t+1])  #it is the probability that hmm path has particular arc at time t
+                    xi_l[i, :, t] = numerator_l
+
+
+"""
+@numba.vectorize
+def Betal(x, y):
+    # numba doesn't understand scipy.special.gammaln, but it understands math.lgamma,
+    # which is the same, except that it doesn't understand arrays.
+    return math.lgamma(x) + math.lgamma(y) - math.lgamma(x+y)
+
+@numba.vectorize
+def fact(x):
+    return math.lgamma(x + 1)
+
+@numba.njit
+def forward(A,B,pi,data):
+    N = A.shape[0]
+    T = data.shape[0]
+    alpha = np.zeros((T,N))
+    alpha_p = np.zeros((T,N))
[…]

### 2022-05-18 Simulations/KIN_hmm/roc/anotherloop.sh

@@ -1,0 +1,15 @@
+for cv in 4 0.5 0.2 0.1 0.05 0.03
+do
+	for a in 0 2
+	do
+		for i in 0 1
+		do
+			for cn in 0 1
+			do
+
+				#snakemake allt --cores 140 --config inbs=$i inb=$i cov=$cv asc=$a cnt=$cn folder=$fold runs=$r
+				snakemake all --cores 140 --config inbs=$i inb=$i cov=$cv asc=$a cnt=$cn folder=$fold runs=$r
+			done
+		done
+	done
+done

### 2022-05-18 Simulations/KIN_hmm/rocNo3deg/loopnew.sh

@@ -1,0 +1,18 @@
+for cv in 4 0.5 0.2 0.1 0.05 0.03
+do
+	for i in 0 1
+	do
+		for a in 0 2
+		do
+			for cnt in 0 1
+			do
+				for fold in allLikelihoods_inphapProbs
+				do
+					snakemake all --cores 55 --config inbs=$i cov=$cv asc=$a cnt=$cnt folder=$fold runs=$r --unlock
+					snakemake all --cores 55 --config inbs=$i cov=$cv asc=$a cnt=$cnt folder=$fold runs=$r
+					
+				done
+			done
+		done
+	done
+done

### 2022-05-18 Simulations/hbd_hmm/hmm_func_hbd.py

@@ -1,0 +1,542 @@
+
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+"""
+Created on Thu May  6 16:33:15 2021
+
+@author: divyaratan_popli
+"""
+
+
+
+import scipy
+import numpy as np
+import random
+import pandas as pd
+from scipy.stats import binom
+from math import isclose
+from scipy.special import gammaln
+from scipy import optimize
+import math
+import matplotlib.pyplot as plt
+from scipy.special import loggamma
+from scipy.special import beta as Betaf
+from scipy.optimize import minimize_scalar
+import pylab
+from numpy.linalg import matrix_power
+
+
+
+
+def Betal(x,y):
+    return gammaln(x)+gammaln(y) - gammaln(x+y)
+
+def fact(x):
+    return(gammaln(x+1))
+
+
+def scaleB(B):
+    scale=B.max(axis=0)
+    Bscaled=B/scale
+    return Bscaled, np.sum(np.log(scale))
+    #return B, np.zeros(np.shape(B)[1])
+
+def forward(A,B,pi,data):
+    N= np.shape(A)[0]
+    T= np.shape(data)[0]
+    alpha=np.zeros((T,N))
+    alpha_p=np.zeros((T,N))
+    scale=np.zeros(T)
+
+    alpha[0,:]= (pi) * B[:,0]
+    alpha_p[0,:]= alpha[0,:]/sum(alpha[0,:])
+    scale[0]=sum(alpha[0,:])
+
+    for t in range(1,T):
+        alpha[t,:]= (alpha_p[t-1,:] @ A) * B[:,t]
+        alpha_p[t,:]= alpha[t,:]/sum(alpha[t,:])
+        scale[t]=sum(alpha[t,:])
+
+    return alpha_p, scale
+
+
+
+def backward(A, B, data, scale):
+
+    N= np.shape(A)[0]
+    T= np.shape(data)[0]
+    beta=np.zeros((T,N))
+    beta_p=np.zeros((T,N))
+
+    beta[-1,:]=np.ones(N)
+    beta_p[-1,:]=np.ones(N)
+
+    for t in reversed(range(T-1)):
+        for n in range(N):
+            beta[t,n]= sum( A[n,:] * beta_p[t+1,:] * B[:,t+1])
+            beta_p[t,n]= beta[t,n]/scale[t+1]
+
+    return beta_p
+
+
+
[…]

### 2022-05-18 Simulations/hbd_hmm/loopd.sh

@@ -1,0 +1,14 @@
+for cnt in 0 1
+do
+	for i in 0 1
+	do
+		for a in 0 2 3
+		do
+			for inp in hapProbs pshap
+			do
+				echo "$i"
+				snakemake alltables --config i=$i a=$a inp=$inp cnt=$cnt --cores 120
+			done
+		done
+	done
+done

### 2022-05-18 Simulations/input_preparation_func.py

@@ -1,0 +1,615 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+"""
+Created on Mon Oct 14 13:22:51 2019
+
+@author: divyaratan_popli
+"""
+
+
+import os
+import msprime as msp
+import numpy as np
+import random
+import pandas as pd
+import matplotlib.pyplot as plt
+from scipy import stats
+from matplotlib import figure
+from scipy.stats import bernoulli
+import math
+#from astropy.stats import jackknife_resampling
+#from astropy.stats import jackknife_stats
+#from scipy.signal import savgol_filter
+from ast import literal_eval
+import pysam
+import pybedtools
+##########################################################start with simulation for relatedness distn
+
+def pileup2alleles(bam, snp):
+    """returns the number of ref and alt alleles at each position"""
+    ref, alt = snp.name, snp.score
+    ref_cnt, alt_cnt = 0, 0
+    for pileupcolumn in bam.pileup(snp.chrom, snp.start, snp.end):
+        if pileupcolumn.reference_pos == snp.start:     #find out the position in pilup that matches position in bed
+            for base in pileupcolumn.get_query_sequences():     #get all bases present at that position
+                if base.upper() == ref.upper(): ref_cnt +=1
+                if base.upper() == alt.upper(): alt_cnt +=1
+            return ref_cnt, alt_cnt
+    return 0, 0
+
+
+def makePseudohap(inbam,inbai,inbed,outps):
+    """takes bam, index, bed files to do random sampling and create pseudohaps"""
+    big_list=[]
+    chrm=[]
+    pos=[]
+    var=9
+    bed = pybedtools.BedTool(inbed)
+    with pysam.AlignmentFile(inbam) as bam:
+        for i, snp in enumerate(bed):
+
+            freq = pileup2alleles(bam, snp)
+            if freq[0]+freq[1]==0:
+                big_list.append(9)
+                var=9
+            else:
+                ref_alt=[snp.name, snp.score]
+                p=freq[1]/(freq[0]+freq[1])
[…]

### 2022-05-18 Simulations/sanity_check/sanitycheck_func.py

@@ -1,0 +1,580 @@
+import numpy as np
+import pandas as pd
+import matplotlib.pyplot as plt
+from matplotlib import figure
+import os
+import math
+
+#preparing the recombination points file
+def prep_rec(rdir):
+
+    files=np.sort(os.listdir(rdir))
+    l=len(files)
+    chrm=np.zeros(l)
+    x=np.zeros(l)
+    par=[]
+    rec=[]
+    for f in range(l):
+        rname=rdir+str(files[f])
+        rfile=np.loadtxt(rname,dtype='float', delimiter = ",")
+        chrm[f]=rname.split('par')[0].split('recomb/')[1]
+
+        x[f]=rname.split('.csv')[0].split('_')[3]
+        par.append(tuple(rname.split('par')[1].split('.csv')[0].split('_')[0].split(',')))
+        rec.append(rfile)
+    chrm=chrm.astype(int)
+    x=x.astype(int)
+    recomb_mat=pd.DataFrame(
+                {'chromosome': chrm,
+                 'parents': par,
+                 'recomb_points': rec,
+                 'random_chrm': x
+                 })
+    return recomb_mat
+
+
+#################################################### Finding recombination points
+
+################################ grandparent-grandchild
+
+def plot_gp(length, interval,filenamed,filenamet,rdir,g1,g2,p1,p2,outimg,outstate,rel23, ninds):
+
+    g1=str(g1)
+    g2=str(g2)
+    p1=str(p1)
+    p2=str(p2)
+
+    win=math.ceil(length/interval)
+
+    rec=prep_rec(rdir)
+
+    j=4
+    i=0
+    state_32=np.zeros(int(rec.loc[j,"recomb_points"][-1]))
+
+    while i<np.shape(rec.loc[j,"recomb_points"])[0]-1:
+
+        if i%2==0:
+            state_32[int(rec.loc[j,"recomb_points"][i]):int(rec.loc[j,"recomb_points"][i+1])]=rec.loc[j,"parents"][rec['random_chrm'][j]]
+        elif i%2==1:
+            state_32[int(rec.loc[j,"recomb_points"][i]):int(rec.loc[j,"recomb_points"][i+1])]=rec.loc[j,"parents"][abs(rec['random_chrm'][j]-1)]
+        i=i+1
+
+
+    #plt.plot(state_32)
+
+    j=5
[…]

### 2022-05-18 Simulations/simulation_hapcallingall_function.py

@@ -1,0 +1,645 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+"""
+Created on Mon Oct 14 13:22:51 2019
+
+@author: divyaratan_popli
+"""
+
+
+import os
+import msprime as msp
+import numpy as np
+import random
+import pandas as pd
+import matplotlib.pyplot as plt
+from scipy import stats
+
+from matplotlib import figure
+from scipy.stats import bernoulli
+
+import copy
+from ast import literal_eval
+import shutil
+import scipy.special as scp
+##########################################################start with simulation for relatedness distn
+
+
+def get_recomb(recombination_rate_child,length):
+
+        lam1=recombination_rate_child*length
+        recomb_no1=np.random.poisson(lam1)
+        recomb_pos1=np.zeros(recomb_no1)
+        recomb_pos1=np.random.uniform(0,length,recomb_no1)
+        recomb_pos1=np.sort(recomb_pos1)
+
+        recomb_interval1=np.zeros(recomb_no1+2)
+        recomb_interval1[0]=0
+        recomb_interval1[-1]=length
+        recomb_interval1[1:recomb_no1+1]=recomb_pos1
+
+        lam2=recombination_rate_child*length
+        recomb_no2=np.random.poisson(lam2)
+        recomb_pos2=np.zeros(recomb_no2)
+        recomb_pos2=np.random.uniform(0,length,recomb_no2)
+        recomb_pos2=np.sort(recomb_pos2)
+
+
+        recomb_interval2=np.zeros(recomb_no2+2)
+        recomb_interval2[0]=0
+        recomb_interval2[-1]=length
+        recomb_interval2[1:recomb_no2+1]=recomb_pos2
+
+        return recomb_no1, recomb_interval1, recomb_no2, recomb_interval2
+
+
+
+
+def make_baby(gen, pos, parent1, parent2, recomb_no1, recomb_interval1, recomb_no2, recomb_interval2, x1, x2):
+    #recombination_rate_child=1e-8     #increasing recombinations for artificial kids 2e-6, 4e-7 for len=1e7
+
+
+    c1=np.zeros(np.shape(gen)[0])
+    c1alt=np.zeros(np.shape(gen)[0])
+    for i in range(recomb_no1+1):
[…]

### 2022-05-18 KIN_snakemake_version/cutoffPlot.R
cleaned up the repository

@@ -1,0 +1,45 @@
+#inf='/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz/hmm_numba_incorrect_roh_i1_i1_insteadOf_i1_i2/roc/contam0_inbred0_model_performance_allroc_asc0.csv.gz'
+
+cutoff_plot <- function(infile,outfile){
+  
+
+  compinp=read.csv(file = inf, header = TRUE)
+  compinp$Relatedness <- factor(compinp$Relatedness, levels = c("id", "fir", "sec", 'un_all','pc','sib','deg3','un_all_withdeg3'))
+  
+  data <- compinp %>%
+    pivot_longer(c(True_positive,False_positive), names_to = "T_F", values_to = "classification_proportion")#Plot
+  data$T_F <- factor(data$T_F, levels= c('True_positive','False_positive'))
+  
+  data<-data[!(data$method=="read_inppshap" | data$method=="allLikelihoods_inppshap"),]
+  #data = subset(data, select = -c(method) )
+  
+  data$Coverage <- as.factor(data$Coverage)
+  
+  #changing some column names:
+  
+  Relatedness.labs <- c("Unrelated", "Unrelated w/o \n 3rd Degree", "3rd Degree", "2nd Degree", "1st Degree", "Siblings", "Parent-Child", "Identical")
+  names(Relatedness.labs) <- c("un_all", "un_all_withdeg3", "deg3", "sec", "fir", "sib", "pc", "id")
+  
+  T_F.labs <- c("False Positive", "True Positive")
+  names(T_F.labs) <- c("False_positive", "True_positive")
+  
+  plot<-ggplot(data=data, aes(x=cutoff, y=classification_proportion, color=Coverage))
+  plot<-plot+geom_line()
+  plot<- plot + facet_grid(T_F ~ Relatedness, labeller = labeller(T_F = T_F.labs, Relatedness = Relatedness.labs))
+  #plot + geom_hline(yintercept=0.05, linetype="dashed")
+  plot <- plot + geom_hline(data = data.frame(T_F='False_positive'), aes(yintercept = 0.05), linetype = "dotted")
+  plot <- plot+ labs(y="Classification Proportions", x = "Log likelihood Ratio Cutoff")
+  plot + theme_bw() + theme(
+    strip.background = element_rect(
+      color="transparent", fill="transparent", size=1.5, linetype="solid"
[…]

### 2022-05-18 hmm_numba/run_snakemake_eva.sh

@@ -1,4 +1,0 @@
-snakemake -w 60 --jobname "{rule}_{jobid}" -j800 \
-    --cluster-config cluster.yaml \
-    --cluster "qsub -q all.q -l h_vmem={cluster.mem} -o autosnake/ -e autosnake/ -cwd -V -S /bin/bash" $*  
-

### 2022-05-18 plots/IBDplot.R

@@ -1,59 +1,0 @@
-library(cowplot)
-library(ggplot2)
-library(dplyr)
-library(tidyr)
-library(readr)
-#fname='/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz/plot_ibd.csv'
-#p1f='/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz/contam0/inbred0/run1/coverage4/asc0/inputMode_hapProbs_fil0/hmm_parameters/p1_file.gz'
-
-ibd_plot <- function(inf,p1f,outf){
-
-  p1 <- read_file(p1f)
-  p1 <- as.double(gsub("[\r\n]", "", p1))
-
-  p12=p1/2
-  p34=p1*3/4
-
-  data=read.csv(file = inf, header = TRUE)
-  colnames(data) <- c("X", "Wins","Data","True","Unrelated","Degree3","Degree2","Parent-Child","Siblings","Identical")
-
-  require(tidyverse)
-
-  d2 = rbind(data, c(0, NA, NA, rep(0.5,7)), c(0,NA, NA, rep(1,7))) %>% as_tibble
-
-  d3 <-d2 %>% pivot_longer(Data:Identical)
-  d3$name_f = factor(d3$name, levels=c('Data','True','Identical','Parent-Child','Siblings','Degree2','Degree3','Unrelated'))
-
-  d_ibd<-subset(d3, name!="Data")
-  d_prop<-subset(d3, name=="Data")
-
-  dat_text <- data.frame(
-    label = as.factor(c("", "LL=-1478.24", "LL=-1259.44", "LL=-1213.61", "LL=-1236.09", "LL=-1261.19", "LL=-1281.77")),
-    name_f = as.factor(c("True", "Identical", "Parent-Child", "Siblings", "Degree2", "Degree3", "Unrelated")),
-    x     = c(5,5,5,5,5,5,5),
-    y     = c(0.85,0.85,0.85,0.85,0.85,0.85,0.85)
-  )
-
-
-  plot1i<-ggplot(data=d_ibd,aes(x=Wins, y=value)) + geom_line() + facet_grid(name_f~., scales='free') +
-  scale_y_continuous(breaks=c(0.5,0.75,1.0), labels = c("2","1","0")) +theme_bw() + labs(y="IBD state", x = "Windows along the genome") +
-  theme(strip.text = element_text(size = 6))
-
-  plot1 <- plot1i + geom_text(
-    data    = dat_text,
[…]

### 2022-05-18 plots/ROHplot_old.R

@@ -1,108 +1,0 @@
-library(ggplot2)
-library(cowplot)
-
-
-
-make_panel1 <- function(data,acrval,leg){
-  plot1<-ggplot(data=data[data$acr==acrval,], aes(x=Wins, y=Prop)) +
-    geom_rect(aes(
-      xmin = Wins-1,
-      xmax = Wins,
-      ymin = -Inf,
-      ymax = +Inf,
-      fill = factor(chrm),
-      #alpha=0.8
-    )) +
-    scale_fill_manual(values = c("grey90", "white")) +
-    theme_void()
-  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.3, size=0.8)
-  plot1<-plot1 + theme_bw()+ theme(legend.position = "none", axis.title.x = element_blank())+labs(y= "Heterozygosity") +
-    scale_color_manual(values=c("0.1"="Green", "0.2"="Blue","4"="Red")) +
-    theme(panel.grid.major = element_blank(),
-          panel.grid.minor = element_blank(),
-          strip.background = element_blank(),
-          panel.border = element_rect(colour = "black", fill = NA))
-  return(plot1)
-}
-
-make_panel2 <- function(data,acrval,leg){
-  plot2<-ggplot(data=data[data$acr==acrval,]) +
-    geom_rect(aes(
-      xmin = Wins-1,
-      xmax = Wins,
-      ymin = -Inf,
-      ymax = +Inf,
-      fill = factor(chrm),
-      #alpha=0.8
-    )) +
-    scale_fill_manual(values = c("grey90", "white")) +
-    theme_void()
-  plot2<-plot2+geom_line(aes(x=Wins, y=True_ROH, linetype="Truth"))
-  #plot2
-  plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.3, size=0.8)
-  
-  plot2<-plot2+scale_linetype_manual(values=c("Truth"="dashed", "Inferred"="solid"))
-  plot2 <- plot2 + scale_color_manual(values=c("0.1"="Green", "0.2"="Blue","4"="Red"))
-  if (leg==1){
-    plot2<-plot2 + theme_bw() + theme(legend.position = "bottom", axis.title.x = element_blank())+labs(y= "1 - P(ROH)")
[…]

### 2022-05-18 plots/Rscripts/supp_IBD.R

@@ -1,5 +1,8 @@
 library(ggplot2)
-
+#infile1='/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/plots/plotdata/plot_IBDaccuracy.csv'
+#infile2='/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/plots/plotdata/plot_IBDaccuracy_cont.csv'
+#infile3='/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/plots/plotdata/plot_IBDaccuracy_asc.csv'
+#infile4='/mnt/diversity/divyaratan_popli/100arc/inbreeding/model_git/plots/plotdata/plot_IBDaccuracy_roh.csv'
 
 IBD_accuracy_plot <- function(infile1,infile2,infile3,infile4,outfile){
   bigd <-data.frame()
@@ -53,13 +56,14 @@ IBD_accuracy_plot <- function(infile1,infile2,infile3,infile4,outfile){
     df$cov=as.factor(df$cov)
     df$Scenario=j
     bigd=rbind(bigd,df)
-  
-  df=data.frame(accuracy,stdev,rel,cov)
-  
-  df$cov=as.factor(df$cov)
-  df$Scenario=j
-  bigd=rbind(bigd,df)
   }
+  
+  #bigd_un=bigd[bigd$rel=='Siblings',]
+  #bigd_un_control=bigd_un[bigd_un$Scenario==1,]
+  #bigd_un_c=bigd_un[bigd_un$Scenario==2,]
+  #bigd_un_a=bigd_un[bigd_un$Scenario==3,]
+  #bigd_un_r=bigd_un[bigd_un$Scenario==4,]
+  
   bigd$Scenario=as.factor(bigd$Scenario)
   colors1 <- c("Unrelated" = "grey30", "3rd Degree" = "blue", "2nd Degree" = "darkorchid1", "Siblings"= "Green", "Parent-Child" = "orangered3", "Identical" = "darkgoldenrod1")
   line1 <- c("1" = "solid", "2" = "dashed", "3" = "dotted","4" = "12345678")
@@ -74,9 +78,9 @@ IBD_accuracy_plot <- function(infile1,infile2,infile3,infile4,outfile){
     scale_color_manual(values = colors1, name="Relatedness") + theme_bw() +geom_path(aes(x=cov, y=accuracy, group=interaction(rel,Scenario),colour=rel, linetype = Scenario)) +
[…]

### 2022-05-18 plots/adding_chrm_colors.R

@@ -1,120 +1,0 @@
-library(cowplot)
-library(ggplot2)
-library(dplyr)
-library(tidyr)
-library(readr)
-#fname='/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz/plot_ibd.csv'
-#p1f='/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz/contam0/inbred0/run1/coverage4/asc0/inputMode_hapProbs_fil0/hmm_parameters/p1_file.gz'
-
-ibd_plot <- function(inf,p1f,outf){
-  
-  p1 <- read_file(p1f)
-  p1 <- as.double(gsub("[\r\n]", "", p1))
-  
-  p12=p1/2
-  p34=p1*3/4
-  
-  data=read.csv(file = inf, header = TRUE)
-  colnames(data) <- c("X", "Wins","Data","True","Unrelated","Degree3","Degree2","Parent-Child","Siblings","Identical")
-  
-  require(tidyverse)
-  
-  d2 = rbind(data, c(0, NA, NA, rep(0.5,7)), c(0,NA, NA, rep(1,7))) %>% as_tibble
-  
-  d3 <-d2 %>% pivot_longer(Data:Identical)
-  d3$name_f = factor(d3$name, levels=c('Data','True','Identical','Parent-Child','Siblings','Degree2','Degree3','Unrelated'))
-  
-  chrmlist=list()
-  for (val in 1:22)
-  {
-    if (val%%2==1){
-      lchr=rep(0,10*8)
-    }
-    else if (val%%2==0){
-      lchr=rep(1,10*8)
-    }
-    chrmlist <- append(chrmlist, lchr)
-    
-  }
-  nalist=rep(NA,16)
-  chrmlist <-append(chrmlist,nalist)
-  
-  d3$chrm =as.integer(chrmlist)
-  
-  
-  d_ibd<-subset(d3, name!="Data")
-  d_prop<-subset(d3, name=="Data")
-  
-  dat_text <- data.frame(
-    label = as.factor(c("", "LL=-1478.24", "LL=-1259.44", "LL=-1213.61", "LL=-1236.09", "LL=-1261.19", "LL=-1281.77")),
-    name_f = as.factor(c("True", "Identical", "Parent-Child", "Siblings", "Degree2", "Degree3", "Unrelated")),
-    x     = c(5,5,5,5,5,5,5),
[…]

### 2022-05-18 plots/comparisonPlot.R

@@ -1,64 +1,0 @@
-library(ggplot2)
-#install.packages("magrittr")
-#install.packages("dplyr")
-library(magrittr)
-library(dplyr)
-library(tidyr)
-
-#inf='/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz/hmm_numba_incorrect_roh_i1_i1_insteadOf_i1_i2/roc/comparison_plot_data.csv.gz'
-
-comparison_plot <- function(inf,outf){
-
-
-  compinp=read.csv(file = inf, header = TRUE)
-  compinp$Relatedness <- factor(compinp$Relatedness, levels = c("id", "fir", "sec", 'un_all','pc','sib','deg3','un_all_withdeg3'))
-
-  data <- compinp %>%
-    pivot_longer(c(True_positive,False_positive), names_to = "T_F", values_to = "classification_proportion")#Plot
-  data$T_F <- factor(data$T_F, levels= c('True_positive','False_positive'))
-
-  colnames(data)[4] <- "Method"
-  data$Method=as.character(data$Method)
-  data[data$Method=='allLikelihoods_inphapProbs','Method'] = 'KIn'
-  data[data$Method=='read_inppshap','Method'] = 'READ'
-
-
-
-
-  colnames(data)[9] <- "Simulation"
-  data$Simulation=as.character(data$Simulation)
-  data[data$Simulation=='0_0_0','Simulation'] = 'Control'
-  data[data$Simulation=='0_0_1','Simulation'] = 'With ROH'
-  data[data$Simulation=='0_2_0','Simulation'] = 'With Ascertainment'
-  data[data$Simulation=='1_0_0','Simulation'] = 'With Contamination'
-
-  data$Coverage <- as.factor(data$Coverage)
-  data$Simulation <-as.factor(data$Simulation)
-
-  #changing some column names:
-
-  Relatedness.labs <- c("Unrelated", "Unrelated w/o \n 3rd Degree", "3rd Degree", "2nd Degree", "1st Degree", "Siblings", "Parent-Child", "Identical")
-  names(Relatedness.labs) <- c("un_all", "un_all_withdeg3", "deg3", "sec", "fir", "sib", "pc", "id")
-
-  T_F.labs <- c("False Positive", "True Positive")
-  names(T_F.labs) <- c("False_positive", "True_positive")
-
-
[…]

### 2022-05-18 plots/cutoffPlot.R

@@ -1,51 +1,0 @@
-
-library(ggplot2)
-#install.packages("magrittr")
-#install.packages("dplyr")
-library(magrittr)
-library(dplyr)
-library(tidyr)
-
-#inf='/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz/hmm_numba_incorrect_roh_i1_i1_insteadOf_i1_i2/roc/contam0_inbred0_model_performance_allroc_asc0.csv.gz'
-
-cutoff_plot <- function(infile,outfile){
-
-
-  compinp=read.csv(file = infile, header = TRUE)
-  compinp$Relatedness <- factor(compinp$Relatedness, levels = c("id", "fir", "sec", 'un_all','pc','sib','deg3','un_all_withdeg3'))
-
-  data <- compinp %>%
-    pivot_longer(c(True_positive,False_positive), names_to = "T_F", values_to = "classification_proportion")#Plot
-  data$T_F <- factor(data$T_F, levels= c('True_positive','False_positive'))
-
-  data<-data[!(data$method=="read_inppshap" | data$method=="allLikelihoods_inppshap"),]
-  #data = subset(data, select = -c(method) )
-
-  data$Coverage <- as.factor(data$Coverage)
-
-  #changing some column names:
-
-  Relatedness.labs <- c("Unrelated", "Unrelated w/o \n 3rd Degree", "3rd Degree", "2nd Degree", "1st Degree", "Siblings", "Parent-Child", "Identical")
-  names(Relatedness.labs) <- c("un_all", "un_all_withdeg3", "deg3", "sec", "fir", "sib", "pc", "id")
-
-  T_F.labs <- c("False Positive", "True Positive")
-  names(T_F.labs) <- c("False_positive", "True_positive")
-
-  plot<-ggplot(data=data, aes(x=cutoff, y=classification_proportion, color=Coverage))
-  plot<-plot+geom_line()
-  plot<- plot + facet_grid(T_F ~ Relatedness, labeller = labeller(T_F = T_F.labs, Relatedness = Relatedness.labs))
-  #plot + geom_hline(yintercept=0.05, linetype="dashed")
-  plot <- plot + geom_hline(data = data.frame(T_F='False_positive'), aes(yintercept = 0.05), linetype = "dotted")
-  plot <- plot+ labs(y="Classification Proportions", x = "Log likelihood Ratio Cutoff")
-  plot + theme_bw() + theme(
-    strip.background = element_rect(
[…]

### 2022-05-18 plots/plot_script.py

@@ -496,6 +496,8 @@ def comparison_plotf(hmmfold,outfold, outs):
     supf=supf.drop('cutoff', axis=1)
     supf.columns=['True_positive', 'False_positive', 'Relatedness', 'Coverage', 'Method','Ascertainment', 'Contamination', 'ROH']
 
+    supf['True_positive']=supf['True_positive'].astype(float).round(2)
+    supf['False_positive']=supf['False_positive'].astype(float).round(2)
     with pd.option_context('display.max_rows', len(supf.index), 'display.max_columns', len(supf.columns)):
                     supf.to_csv(outs, sep=',',index=False)
 

### 2022-05-18 prepare_4_kin.sh

@@ -1,48 +1,0 @@
-#!/bin/bash
-
-#launch as:
-#bash {this_script}.sh {path_to}/{LIST}
-#where LIST=a list containing the absoluthe path to all the bam files that are
-#to be used in the kinship analysis
-
-echo launching preparation to KIN_snakemake
-echo make sure to be in the dir where the snakemake pipeline will be launched
-echo make sure that the original file names are formatted as:
-echo {sample_name}{.yadayada}.bam
-echo and that it is unique
-
-x=0
-PATHLIST=$1
-
-mkdir -p bamfiles
-mkdir -p bamfiles/stephane
-rm targets.txt
-
-sed 's/.*\///g' $PATHLIST > filenames.tmp
-tot=$(cat filenames.tmp | wc -l)
-
-echo creating the symlinks
-for i in $(cat filenames.tmp | sed 's/\..*//g')
-do
-  export y=$(( ${x} + 1 ))
-  export x=$y
-  echo ${x}/${length} creating the symlink for ${i} 
-
-  path=$(grep $i $PATHLIST)
-  ln -s $path bamfiles/stephane/${i}.bam
-  echo ${i} >> targets.txt
-done
-
-echo creating the other lists
-cp targets.txt unrelated.txt
-cp targets.txt identical.txt
-
-cp /home/luca_traverso/scripts/allsites.bed .
-
-echo done!
-echo $(ls targets.txt) $(wc -l targets.txt)
-echo $(ls unrelated.txt) $(wc -l unrelated.txt)
-echo $(ls identical.txt) $(wc -l identical.txt)
-
-ls -altrh bamfiles/stephane/*.bam > samples_analyzed.txt
-echo a list of all the samples of this analysis is in samples_analyzed.txt

### 2022-05-18 pypackage/kin/build/lib/KIN/__main__.py

@@ -1,0 +1,91 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+"""
+Created on Tue Feb 22 17:52:45 2022
+
+@author: divyaratan_popli
+"""
+
+import argparse
+import multiprocessing as mp
+
+from .hmm_scripts import constants as C
+from .hmm_scripts import helpers
+
+
+def cli():
+    parser = argparse.ArgumentParser(description="Relatedness and IBD estimates")
+
+    parser.add_argument('-I', '--input_location',
+                        type=str, metavar='',  required=True,
+                        help='input files location')
+    parser.add_argument('-O', '--output_location',
+                        type=str, metavar='', required=True,
+                        help='Output files location')
+    parser.add_argument('-T', '--target_location',
+                        type=str, metavar='', required=True,
+                        help='file with input bamfile names')
+    parser.add_argument('-r', '--ROH_file_location',
+                        type=str, metavar='',
+                        help='ROH files location')
+    parser.add_argument('-c', '--cores',
+                        type=int, metavar='',
+                        help='Number of cores available')
+    parser.add_argument('-t', '--threshold',
+                        type=int, metavar='',
+                        help='Minimum number of sites in a window for ROH implementation')
+    parser.add_argument('-p', '--diversity_parameter_p_0',
+                        type=float, metavar='',
+                        help='Input p_0 parameter, if you do not want to calculate it from given samples (Keep it same as that for KINgaroo)')
+    parser.add_argument('-i', '--interval',
+                        type=int, metavar='',
+                        help='Window size (Please choose the same size as for KINgaroo). By default:10000000')
+
+    return parser.parse_args()
+
+
+def main():
+    args = cli()
+
+    if args.ROH_file_location is None:
+        hbdfolder = args.input_location
+    else:
+        hbdfolder = args.ROH_file_location
+
+    if args.threshold is None:
+        thresh = 10
+    else:
+        thresh = args.threshold
+
+    if args.cores is None:
+        cores = mp.cpu_count()
+    else:
+        cores = args.cores
+    if args.diversity_parameter_p_0 is None:
+        p_0 = -9
+    else:
+        p_0 = args.diversity_parameter_p_0
+    if args.interval is None:
+        A_interval=C.A_ALL10
+    elif args.interval==10000000:
+        A_interval=C.A_ALL10
[…]

### 2022-05-18 pypackage/kin/build/lib/KIN/hmm_scripts/constants.py

@@ -1,0 +1,98 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Fri Mar  4 10:44:57 2022
+
+@author: divyaratan_popli
+"""
+import numpy as np
+
+STATES=[0,1,2]
+CHRM=list(range(1,23))
+RELS=['un','deg5','deg4','deg3','gr','hsib','avu','sib','pc','id']
+
+#transitions for 10M windows
+A_un=np.array([[1.00000e-06, 9.99998e-01, 1.00000e-06],
+       [1.00000e-06, 9.99998e-01, 1.00000e-06],
+       [1.00000e-06, 9.99998e-01, 1.00000e-06]])
+
+A_deg5=np.array([[6.25593827e-01, 3.74405173e-01, 1.00000000e-06],
+       [2.51892991e-02, 9.74809701e-01, 1.00000000e-06],
+       [6.66666167e-01, 3.33332833e-01, 1.00000000e-06]])
+
+A_deg4=np.array([[7.53090860e-01, 2.46908140e-01, 1.00000000e-06],
+       [3.58180318e-02, 9.64180968e-01, 1.00000000e-06],
+       [6.66666167e-01, 3.33332833e-01, 1.00000000e-06]])
+
+A_deg3=np.array([[8.27500554e-01, 1.72498446e-01, 1.00000000e-06],
+       [5.86467863e-02, 9.41352214e-01, 1.00000000e-06],
+       [6.66666167e-01, 3.33332833e-01, 1.00000000e-06]])
+
+A_gr=np.array([[9.09364377e-01, 9.06346235e-02, 1.00000000e-06],
+       [9.06346235e-02, 9.09364377e-01, 1.00000000e-06],
+       [6.66665667e-01, 3.33333333e-01, 1.00000000e-06]])
+
+A_hsib=np.array([[8.35731279e-01, 1.64267721e-01, 1.00000000e-06],
+       [1.64387619e-01, 8.35611381e-01, 1.00000000e-06],
+       [6.66666167e-01, 3.33332833e-01, 1.00000000e-06]])
+
+A_avu=np.array([[8.05569361e-01, 1.94429639e-01, 1.00000000e-06],
+       [1.94943269e-01, 8.05055731e-01, 1.00000000e-06],
+       [6.66666167e-01, 3.33332833e-01, 1.00000000e-06]])
+
[…]

### 2022-05-18 pypackage/kin/build/lib/KIN/hmm_scripts/helpers.py

@@ -1,0 +1,79 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Fri Mar  4 11:33:59 2022
+
+@author: divyaratan_popli
+"""
+
+import os
+import numpy as np
+import multiprocessing as mp
+
+from .hmm_functions import *
+
+
+def hmm_prep(targetfile, outfolder, datafolder, allrel, p_0):
+
+    with open(targetfile) as f:
+        libraries = [line.strip() for line in f]
+
+    listf=[]
+    for i, l1 in enumerate(libraries):
+        for l2 in libraries[(i+1):]:
+            s = '%s_._%s' %(str(l1).strip("['']"), str(l2).strip("['']"))
+            listf.append(s)
+
+
+    #make dirs
+
+    if not(os.path.isdir(outfolder)):
+        os.mkdir(outfolder)
+    if not(os.path.isdir(outfolder+"likfiles/")):
+        os.mkdir(outfolder+"likfiles/")
+    if not(os.path.isdir(outfolder+"resfiles/")):
+        os.mkdir(outfolder+"resfiles/")
+    if not(os.path.isdir(outfolder+"gammafiles/")):
+        os.mkdir(outfolder+"gammafiles/")
+
+
+    if p_0==-9:
+        pfile=datafolder + "hmm_parameters/p_0.txt"
+        with open(pfile,"r") as f:
+            p_0=float(f.read())
+
+    dfile=datafolder + "input_diffs_hmm.csv"
+    tfile=datafolder + "input_total_hmm.csv"
+
+    p1thresh=datafolder+"input_hbd_hmm_total.csv"
+
+    return libraries, listf, allrel, p_0, dfile, tfile, p1thresh
+
+
+
+def hmm_run(hbdfolder, libraries, listf, instates, totalch, allrel, pfile, dfile, tfile, Afiles, p1thresh, outfolder, thresh, cores):
+
+    pool = mp.Pool(cores)
+    [pool.apply_async(hmm, args=(listind, hbdfolder, dfile, tfile,
+                            np.array(listf), np.array(libraries), pfile,
+                            Afiles, outfolder,
+                            p1thresh, thresh, instates, allrel)) for listind in listf]
+    pool.close()
+    pool.join()
+
+def hmm_results(outfolder, listf, allrel):
+
+    likfile=[outfolder + "likfiles/{}.csv".format(n) for n in listf]
[…]

### 2022-05-18 pypackage/kin/build/lib/KIN/hmm_scripts/hmm_functions.py

@@ -1,0 +1,554 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Tue Feb 22 15:56:06 2022
+
+@author: divyaratan_popli
+"""
+
+import os
+import numpy as np
+import pandas as pd
+from math import isclose
+import math
+from scipy.optimize import minimize_scalar
+from numpy.linalg import matrix_power
+import numba
+
+
+@numba.vectorize
+def Betal(x, y):
+    # numba doesn't understand scipy.special.gammaln, but it understands math.lgamma,
+    # which is the same, except that it doesn't understand arrays.
+    return math.lgamma(x) + math.lgamma(y) - math.lgamma(x+y)
+
+@numba.vectorize
+def fact(x):
+    return math.lgamma(x + 1)
+
+@numba.njit
+def forward(A,B,pi,data):
+    N = A.shape[0]
+    T = data.shape[0]
+    alpha = np.zeros((T,N))
+    alpha_p = np.zeros((T,N))
+    scale = np.zeros(T)
+
+    alpha[0,:] = pi * B[:,0]
+    alpha_p[0,:] = alpha[0,:] / alpha[0,:].sum()
+    scale[0] = alpha[0,:].sum()
+
+    for t in range(1,T):
+        alpha[t,:]= (alpha_p[t-1,:] @ A) * B[:,t]
+        alpha_p[t,:]= alpha[t,:] / alpha[t,:].sum()
+        scale[t] = alpha[t,:].sum()
+
+    return alpha_p, scale
+
+#@numba.njit
+def backward(A, B, data, scale):
+
+    N= np.shape(A)[0]
+    T= np.shape(data)[0]
+    beta=np.zeros((T,N))
+    beta_p=np.zeros((T,N))
+
+    beta[-1,:]=np.ones(N)
+    beta_p[-1,:]=np.ones(N)
+
+    for t in reversed(range(T-1)):
+        for n in range(N):
+            beta[t,n]= sum( A[n,:] * beta_p[t+1,:] * B[:,t+1])
+            beta_p[t,n]= beta[t,n]/scale[t+1]
+
+    return beta_p
+
+
+
+@numba.njit
+def objective(a, meanp, k, data):
+    cost=0
+
+    b=a * (1-meanp)/meanp
+
+    for w in range(len(data)):
+
[…]

### 2022-05-18 run_snakemake_eva.sh

@@ -1,11 +1,0 @@
-#!/bin/bash
-
-#run with is_contam=0
-
-snakemake all \
--w 60 \
---jobname "{rule}_{jobid}" \
--j 20 \
---cluster-config cluster.yaml \
---cluster "qsub -q all.q -l h_vmem={cluster.mem} -o autosnake/smk.out -e autosnake/smk.err -cwd -V -S /bin/bash" $* \
---config is_contam=0

### 2022-05-18 supplementary_info/beta_bnds.py

@@ -639,22 +639,22 @@ def plot_betas(dfile, tfile, pair, thresh, rel, libraries, listf, Afile, pfile, 
 
         a=betas[2]
         b=betas[3]
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'r-', lw=3, alpha=0.6, label="$Z_w$=0")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'r-', lw=3, alpha=0.6, label="i=0")
         #plt.hist(diff/total,alpha=0.3)
 
         a=betas[0]
         b=betas[1]
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'b-', lw=3, alpha=0.6, label="$Z_w$=1")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'b-', lw=3, alpha=0.6, label="i=1")
 
         a=betas[4]
         b=betas[5]
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'g-', lw=3, alpha=0.6, label="$Z_w$=2")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'g-', lw=3, alpha=0.6, label="i=2")
         #plt.hist(diff/total,alpha=0.3)
 
         a=betas[6]
         b=betas[7]
 
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'y-', lw=3, alpha=0.6, label="$Z_w$=4")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'y-', lw=3, alpha=0.6, label="i=3")
         ax[i].hist(diff/total,alpha=0.3,weights=np.ones(len(total)) / len(total))
         ax[i].set_ylim(0,0.4)
 

### 2022-05-18 supplementary_info/supplementary_scripts.py

@@ -34,29 +34,66 @@ def suppTable2(resultsf,kinf,outf2,outf3):
     kin.loc[kin['relatedness']=='un','relatedness_l'] ='Unrelated'
     kin.loc[kin['relatedness']=='deg3','relatedness_l'] ='Third Degree'
 
+    a=list(set(kin['pair']) - set(res['pair']))
+    for i in a:
+        kin.loc[kin['pair']==i,'pair']=i.split('_')[1] + '_' + i.split('_')[0]
 
     ares=pd.merge(res, kin, how='inner', on=['pair'])
+
     ares.loc[ares['Relationship (r/k0)* ']=='unrelated', 'Relationship (r/k0)* ']='Unrelated'
     ares.loc[ares['Relationship (r/k0)* ']=='2nd degree', 'Relationship (r/k0)* ']='Second Degree'
-
+    ares.loc[ares['Relationship (r/k0)* ']=='3rd-5th degree', 'Relationship (r/k0)* ']='Third Degree'
+    #rounding off
+    ares['loglik_ratio']=round(ares['loglik_ratio'],2)
+    ares['Z_lower']=ares['Z_lower'].astype(float).round(2)
+    ares['Z_upper']=ares['Z_upper'].astype(float).round(2)
     #where do read and kin differ?
 
     diff=ares.loc[ares['relatedness_r']!=ares['Relationship (READ)'],['relatedness_r','relatedness_l','loglik_ratio','Relationship (READ)','Z_upper','Z_lower','Relationship (r/k0)* ','pair','nSNPs']]
     diff.reset_index(drop=True, inplace=True)
 
+    diff1=diff.loc[diff['loglik_ratio']>1,:]
+    diff1.reset_index(drop=True, inplace=True)
 
-    dtable=diff.copy()
-    dtable.columns=['relatedness_r', 'Relatedness KIn', 'Log liklihood ratio (KIn)', 'Relatedness READ','Z_upper (READ)', 'Z_lower (READ)', 'Relatedness lcMLkin', 'pair', 'nSNPs']
+    diff2=diff1.loc[~((abs(diff1['Z_lower'])<1) | (abs(diff1['Z_upper'])<1)),:]
+    diff2=diff2.loc[diff2['Relationship (r/k0)* ']!='too low coverage',:]
+
+    #where do lcmlkin and kin differ?
[…]

### 2022-05-18 paper_overleaf/main.tex
Updates from Overleaf

@@
 Genetic paternity tests have an important application in resolving family relation, e.g.
-in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}.
+in establishing relationship between an individual applying for immigration and the claimed relatives \cite{egeland_beyond_2000}.
 It is also an essential step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.
@@
 However, since it is not possible to directly observe IBD segments, a common approach is to first identify segments of the genome that are Identical by State (IBS) and to use population allele frequencies obtained from an out-of-sample reference panel to calculate the probability of IBD given IBS.
-There are several methods that incproporate reference panel allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007,manichaikul_robust_2010,gusev_whole_2009,nyerki_optimized_2022}.
+There are several methods that incproporate reference panel allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, purcell_plink_2007,manichaikul_robust_2010,gusev_whole_2009,nyerki_optimized_2022,browning_fast_2011,li_accurate_2014}.
 
@@
 Also, our method makes the assumption, that the median pairwise genetic difference in the population reflects the population diversity $p_0$, which fails if almost all individuals in the dataset are related.
+We may get around this problem by using an estimate of $p_0$, calculated from known a pair of unrelated individuals from same population, or another population with similar diversity.
 The current implementation of KIN is restricted to six relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.
@@
 
-\section{Competing interests}
+\section{Data and material availability}
+An open-source implementation of KIN and KINgaroo in python along with a toy example dataset, and the scripts to generate our test simulations are available on GitHub \url{https://github.com/DivyaratanPopli/Kinship_Inference}.
+We have deposited the version of software used in the manuscript on Zenodo ().
+The analysed datasets from Bronze Age Lech Valley were generated in a previous study \cite{mittnik_kinship-based_2019}.
+Chagyrskaya and Okladnikov dataset were generated from a study currently in press \cite{laurits_skov_genetic_nodate}, and the data will be uploaded to European Nucleotide Archive upon publication.
+
+\section{Ethics declarations}
+\subsection{Ethics approval and consent to participate}
+Not applicable.
+
+\subsection{Consent for publication}
+Not applicable.
+
+\subsection{Competing interests}
 The authors declare that they have no competing interests.

### 2022-08-18 paper_overleaf/main.tex
Updates from Overleaf

@@
 However, since it is not possible to directly observe IBD segments, a common approach is to first identify segments of the genome that are Identical by State (IBS) and to use population allele frequencies obtained from an out-of-sample reference panel to calculate the probability of IBD given IBS.
-There are several methods that incproporate reference panel allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, purcell_plink_2007,manichaikul_robust_2010,gusev_whole_2009,nyerki_optimized_2022,browning_fast_2011,li_accurate_2014}.
+There are several methods that incorporate reference panel allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, purcell_plink_2007,manichaikul_robust_2010,gusev_whole_2009,nyerki_optimized_2022,browning_fast_2011,li_accurate_2014}.
 
 \begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib.png}
+    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib_rev.png}
     \centering
@@
 \subsection{How our method works}
-Here, we present KIN (Kinship INference), a Hidden Markov Model (HMM)-based approach to estimate genetic kinship and IBD from low-coverage ancient DNA data.
+Here, we present KIN (Kinship Inference), a Hidden Markov Model (HMM)-based approach to estimate genetic kinship and IBD from low-coverage ancient DNA data.
 KIN can detect up to \nth{3}-degree relatives, and differentiates between siblings and parent-child relationships.
@@
 \paragraph{Beta Binomial Model:}
-The Binomial Model laid out above assumes that sites are independent.
-However, due to genetic linkage, nearby sites will be correlated.
-As a result, we empirically find that the data often has considerably higher variance than would be expected from a binomial model (Fig.
+We empirically find that the data often has considerably higher variance than would be expected from a binomial model (Fig.
 S2).
@@
 This parameterization of the beta distribution in terms of expected value $p$ and over-dispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parameterization in terms of $\alpha$ and $\beta$.
-We use this equation even if preprocessing steps result in non-integer $D_w$ and $N_w$, in which case we approximate the binomial coefficient using Gamma functions.
+We use this equation even if preprocessing steps (see equation \ref{eq:x} and section \ref{contam}) result in non-integer $D_w$ and $N_w$, in which case we approximate the binomial coefficient using Gamma functions.
 
@@
     log P(\mathbf{\Delta},\mathbf{Y}|\Theta,\delta) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta, \delta) + log P(\mathbf{Y}|\Theta)\nonumber\\
- &= [\sum_{w} log P(\Delta_w|Y_w, M, \delta) + \sum_{w} log P(Y_w|Y_{w-1}, A)] + P(Y_0| \pi),
+ &= [\sum_{w} log P(\Delta_w|Y_w, M_w, \delta) + \sum_{w} log P(Y_w|Y_{w-1}, A)] + P(Y_0| \pi),
 \end{align}
@@
 
-We assume that contamination rates in both individuals are known and small ($<5\%$), and set $C_{ij} \approx C_i + C_j$, where $C_i, C_j$ are the contamination estimates from the two individuals.
+We assume that contamination rates in both individuals are known and small ($<5\%$), and set $C_{ij} = C_i + C_j$, where $C_i, C_j$ are the contamination estimates from the two individuals.
 We also assume the divergence $\phi$ between our
@@
 
-For any particular comparison showing a difference $D$, we calculate the probability of the event $E$ that it is between endogenous reads as 
+For any particular comparison showing a difference $D$ (not to be confused with number of differences $D_w$), we calculate the probability of the event $E$ that it is between endogenous reads as 
 \begin{equation}
@@
 \caption{\label{tab:Table 2}Model performance for ROH prediction.
-Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination, R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
+Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination), R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
 \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
@@
 \paragraph{IBD prediction:}
-We investigated the accuracy of IBD state prediction along the genome by counting the number of genomic windows where we correctly predict IBD state for different relatdness and coverages (Fig.
+We investigated the accuracy of IBD state prediction along the genome by counting the number of genomic windows where we correctly predict IBD state for different relatedness and coverages (Fig.
 \ref{fig4:IBDstate_accuracy}).
@@
 The error bars are drawn at 1 standard deviation from the mean.
-Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
+Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.
+Accuracy is defined as the sum of true positives and true negatives divided by the total number of windows.
+We consider a prediction is true if it matches the IBD state at the central position of the window.}
     \label{fig4:IBDstate_accuracy}
@@
 The false positive rate in this case remains below 0.02 for both methods.
-We also find that READ is strongly impacted by contamination ($<3\%$) as the true positive rate is in the range 0.89 to 0.78 and 0.53 to 0.46 for 1st and 2nd degree relatives, respectively, and the false positive rate reaches up to 0.04 and 0.17 for unrelated individuals and 2nd degree relatives, respectively.
+We also find that READ is strongly impacted by contamination (proportion of contamination$<3\%$) as the true positive rate is in the range 0.89 to 0.78 and 0.53 to 0.46 for 1st and 2nd degree relatives, respectively, and the false positive rate reaches up to 0.04 and 0.17 for unrelated individuals and 2nd degree relatives, respectively.
 By contrast, the correction implemented in KIN is sufficient to remove this bias, and the true and false positive rates remain as in the control.
@@
 To test KIN on real ancient data, we applied it to a Neandertal dataset from Chagyrskaya and Okladnikov Caves in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020,laurits_skov_genetic_nodate}.
-This dataset contains genetic data from a total of 16 skeletal remains that likely belong to contemporary Neandertals who occupied the Chagyrskaya and Okladnikov caves between 59 and 51 kya, and atleast 44 kya respectively \cite{laurits_skov_genetic_nodate}.
+This dataset contains genetic data from a total of 16 skeletal remains that likely belong to contemporary Neandertals who occupied the Chagyrskaya and Okladnikov caves between 59 and 51 kya, and at least 44 kya respectively \cite{laurits_skov_genetic_nodate}.
 DNA extracted from each of these remains were captured with an array targeting variable sites identified in high coverage Neandertal and Denisovan genomes, and common variations in Africans \cite{laurits_skov_genetic_nodate}.
@@
 We compared our relatedness estimates to those obtained by the authors using READ and lcMLkin.
-We found that KIN was able to confidently classify $85\%$ of the 6903 possible comparisons ($\Delta LL<1$), READ $94\%$ ($sd<1$) and lcMLkin 53\% of pairs (\cite{mittnik_kinship-based_2019}).
+We found that KIN was able to confidently classify $85\%$ of the 6903 possible comparisons ($\Delta LL>1$), READ $94\%$ ($s.d.>1$) and lcMLkin 53\% of pairs (\cite{mittnik_kinship-based_2019}).
 
@@
 In the last case, all three methods differ (table S3, Fig.
-S11).
+S11), and the true relatedness is unresolved in this case.
 Finally, there are three disagreements between KIN and lcMLkin in classification of parent-child versus siblings (READ predicts first degree for all three pairs).
@@
 
-We show that KIN reliably detects IBD tracts for upto 0.1x coverage even in presence of contamination and ascertainment bias, and the accuracy for IBD detection reduces with lower coverages.
+We show that KIN reliably detects IBD tracts for up to 0.1x coverage even in presence of contamination and ascertainment bias, and the accuracy for IBD detection reduces with lower coverages.
 Adding ROH to the simulations adversely affects IBD prediction in siblings (Fig.
@@
 We may get around this problem by using an estimate of $p_0$, calculated from known a pair of unrelated individuals from same population, or another population with similar diversity.
+We provide the user with an option to provide an estimate of $p_0$.
 The current implementation of KIN is restricted to six relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.
@@
 This makes KIN easy to automatize for large datasets.
-To make application of KIN user-friendly, we provide a python package (KINgaroo) to create input files for KIN from processed bamfiles, while optionally estimating ROH, and correcting for contamination estimates.
+To make application of KIN user-friendly, we provide a python package (KINgaroo) to create input files for KIN from processed bam files, while optionally estimating ROH, and correcting for contamination estimates.
 
@@
 \section{Acknowledgements}
-We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Johann Visagie, Laurits Skov, Leonardo N.
+We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Johann Visagie, Zbigniew Jędrzejewski-Szmek, Laurits Skov, Leonardo N.
 M.
@@
 \section{Data and material availability}
-An open-source implementation of KIN and KINgaroo in python along with a toy example dataset, and the scripts to generate our test simulations are available on GitHub \url{https://github.com/DivyaratanPopli/Kinship_Inference}.
-We have deposited the version of software used in the manuscript on Zenodo ().
-The analysed datasets from Bronze Age Lech Valley were generated in a previous study \cite{mittnik_kinship-based_2019}.
+An open-source implementation of KIN and KINgaroo in python along with a toy example dataset, and the scripts to generate our test simulations are available on GitHub \cite{divyaratan_popli_httpsgithubcomdivyaratanpoplikinship_inferencereleasestagv312_2022}.
+We have deposited the version of the software used in the manuscript, along with the above mentioned files on Zenodo \cite{divyaratan_popli_httpsdoiorg105281zenodo6561015_nodate}.
+The analysed datasets from Bronze Age Lech Valley were generated in a previous study (ENA, accession no.
+PRJEB34400).
 Chagyrskaya and Okladnikov dataset were generated from a study currently in press \cite{laurits_skov_genetic_nodate}, and the data will be uploaded to European Nucleotide Archive upon publication.

### 2022-08-18 supplementary_info/beta_bnds.py

@@ -639,22 +639,22 @@ def plot_betas(dfile, tfile, pair, thresh, rel, libraries, listf, Afile, pfile, 
 
         a=betas[2]
         b=betas[3]
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'r-', lw=3, alpha=0.6, label="i=0")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'r-', lw=3, alpha=0.6, label="$Z_w$=0")
         #plt.hist(diff/total,alpha=0.3)
 
         a=betas[0]
         b=betas[1]
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'b-', lw=3, alpha=0.6, label="i=1")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'b-', lw=3, alpha=0.6, label="$Z_w$=1")
 
         a=betas[4]
         b=betas[5]
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'g-', lw=3, alpha=0.6, label="i=2")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'g-', lw=3, alpha=0.6, label="$Z_w$=2")
         #plt.hist(diff/total,alpha=0.3)
 
         a=betas[6]
         b=betas[7]
 
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'y-', lw=3, alpha=0.6, label="i=3")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'y-', lw=3, alpha=0.6, label="$Z_w$=4")
         ax[i].hist(diff/total,alpha=0.3,weights=np.ones(len(total)) / len(total))
         ax[i].set_ylim(0,0.4)
 

### 2022-08-18 paper_overleaf/main.tex
some naming issue

@@
     \caption{IBD sharing between siblings without and with runs of homozygosity (ROH).
-In top row, recombined chromosomes inherited from parents by a pair of siblings are shown.
-On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes.
+(A) Schematic of chromosomes for parent 1 (Par1)and parent 2 (Par2).
+On the left, the parents have unrelated chromosomes depicted with different colors, and on the right parent 1 has a homozygous pair of chromosomes.
+(B) In top row, recombined chromosomes inherited from parents by a pair of siblings are shown.
 The bottom row shows the expected proportion of differences ($p$) along the chromosome.

### 2022-09-06 supplementary_info/supplementary_scripts.py
added plot for contamination missspecification

@@ -60,3 +60,114 @@ def suppTable2(resultsf,kinf,outf2,outf3):
 
     with pd.option_context('display.max_rows', len(rtable.index), 'display.max_columns', len(rtable.columns)):
                 rtable.to_csv(outf3, sep=',',index=False)
+
+
+
+
+
+
+
+def cnt_manipulation3(t1):
+
+    t=t1.copy()
+    t['un_all']=t['un']+t['deg5']+t['deg4']
+    t=t.drop(['un','deg5','deg4'],axis=1)
+
+    t['fir']=t['sib']+t['pc']
+    t=t.drop(['pc','sib'],axis=1)
+
+    t['sec']=t['gr']+t['avu']+t['hsib']
+    t=t.drop(['gr','avu','hsib'],axis=1)
+
+
+    t = t.reindex(index=['un_all', 'deg3', 'sec', 'fir', 'id'], columns=['un_all', 'deg3', 'sec', 'fir', 'id', 'NF'])
+
+    t.loc['un_all',:]=t.loc['un_all',:] * 8
+    t.loc['deg3',:]=t.loc['deg3',:] * 1
+    t.loc['sec',:]=t.loc['sec',:] * 2
+    t.loc['fir',:]=t.loc['fir',:] * 4
+    t.loc['id',:]=t.loc['id',:] * 0
+
+    return t
+
+def cntall_manipulation(t1):
+    t=t1.copy()
+    t.loc['un_all',:]=t.loc['un',:]
+    t=t.drop(['un','deg4','deg5'],axis=0)
+    t.loc['sec',:]=(t.loc['gr',:] + t.loc['hsib',:]) / 2
+    t=t.drop(['gr','hsib','avu'],axis=0)
+    t.loc['fir',:]=(t.loc['pc',:]* 3 + t.loc['sib',:]* 1)/4
+    t=t.drop(['pc','sib'],axis=0)
+
+    return t
+
+
+def misscontam_plotf(outf,covs,ctlist):
+
+
+
+    Tr=[]
+    Fa=[]
+    rl=[]
+    cv=[]
+    met=[]
+    cnt=[]
+
+    for cov in covs:
+            for ct in ctlist:
+
+                table="/mnt/diversity/divyaratan_popli/review_sim/hmm_numba/miss_contam/output/missC%s_10/contam1/inbred0/model_performance_allLikelihoods_inphapProbs/coverage%s/asc0/filtered0_cut1.0.csv.gz" %(ct,cov)
+
[…]

### 2022-09-06 paper_overleaf/main.tex
Updates from Overleaf

@@
 \noindent Inferring genetic kinship of ancient individuals can provide insights into their culture and social hierarchy, and is relevant for downstream genetic analyses.
-However, estimating relatedness from ancient DNA is difficult due to low coverage, ascertainment bias, or contamination from various sources.
-
-Here, we present KIN, a Hidden-Markov-Model-based approach to estimate the relatedness of a pair of individuals and to identify the identical-by-descent segments they share.
-KIN can accurately classify up to 3rd-degree relatives and differentiate between sibling and parent-child relationships using ancient DNA sequence data with as little as 0.05x coverage.
-KIN also incorporates an adjustment for contamination, and an additional model to estimate the location of long runs of homozygosity, and uses them to improve classification accuracy.
-
-We show with simulations that KIN infers first-degree relationships with nearly perfect accuracy, and second degree relationships with about $90\%$ accuracy, even at 0.05x coverage.
-
-To illustrate its utility, we apply KIN to a Neandertal group from Siberia, Russia, and a group of Bronze-Age people from the Lech Valley, Germany.
-We find that in both cases, KIN matches the findings of existing approaches and detects several additional relationships.
-
-KIN is a useful tool for analyzing relatedness for ancient DNA datasets of any size, even with low coverage, and is applicable to any diploid species, without the requirement for a reference panel.
+However, estimating relatedness from ancient DNA is difficult due to low-coverage, ascertainment bias, or contamination from various sources.
+Here, we present KIN, a method to estimate the relatedness of a pair of individuals from the identical-by-descent segments they share.
+KIN accurately classifies up to 3rd-degree relatives using >0.05x sequence coverage and differentiates siblings from parent-child.
+It incorporates additional models to adjust for contamination and detect inbreeding, which improves classification accuracy, and can be applied to any diploid species.
 \end{abstract}
@@
 \begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/plotimg/IBDplot.png}
+    \includegraphics[width=16cm]{plots/inkscape_finalImg/IBDplot_title.png}
     \centering
@@
 
-For any particular comparison showing a difference $D$ (not to be confused with number of differences $D_w$), we calculate the probability of the event $E$ that it is between endogenous reads as 
+For any particular comparison showing a difference $D$ (not to be confused with the number of differences $D_w$), we calculate the probability of the event $E$ that it is between endogenous reads as 
 \begin{equation}
@@
 \begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot.png}
+    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot_final1.png}
     \centering
@@
 S11), and does not have enough data for classification in 4 cases.
-Since READ does not infer third degree, and lcMLkin infers a group of relatedness together as $3^{rd}$ - $5^{th}$, it is likely that KIN is able to correctly resolve these cases.
 For 10 additional pairs, lcMLkin predicts $3^{rd}$ - $5^{th}$ degree, but KIN infers them to be unrelated.
-Thus, these are likely examples of $4^{th}$ or $5^{th}$ degree relations.
 

### 2022-09-07 paper_overleaf/KIN_a_method_to_estimate_relatedness_from_low_coverage_data_supplements.tex
Updates from Overleaf

@@
 
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=18cm]{supplementary_info/plots/contam_missspecification.png}
+    \caption{False positive and true positive rates as a function of incorrect contamination estimates provided to KINgaroo.
+Dashed vertical line represents true contamination (2.5\%) in simulations, and dotted horizontal line in top panel represents $5\%$ false positive rate.}
+    \label{figS10:cnt}
+\end{figure}
+
 \end{document}

### 2022-09-07 paper_overleaf/main.tex

@@
 The amount of contamination does not need to be large for this to be important, in our simulations we find that even at contamination levels $\leq 3\%$, the performance of READ is substantially reduced.
-Assuming the contamination rates are correctly inferred, the correction we implemented leads to improved performance compared to naive methods such as READ, although they too could be ameliorated in a similar way \cite{laurits_skov_genetic_nodate}.
+When contamination rates are correctly inferred, the correction we implemented leads to improved performance compared to naive methods such as READ, although they too could be ameliorated in a similar way \cite{laurits_skov_genetic_nodate}.
+However, in many cases contamination estimates may be uncertain or inaccurate, and we show in fig.
+S13 that KIN's performance is robust to small deviations (small compared to average pairwise heterozygosity) in contamination estimates.
 

### 2022-09-07 paper_overleaf/main.tex
Updates from Overleaf

@@
 Without ROH, we correctly inferred the absence of ROH as the observed bias ranges from $0.3$ to $0.7\%$ (variance $\leq0.006$; Table \ref{tab:Table 2}).
-With ROH, we observed a slight bias of $5$ to $9\%$ more ROH than simulated (variance $\leq0.004$).
-The largest bias observed ($9\%$) was for the datasets that included contamination.
+With ROH, we observed a slight bias of $5$ to $9\%$ more ROH than simulated (variance $\leq0.04$).
+The largest bias observed ($10.1\%$) was for the datasets that included contamination.
 Ascertainment bias had a minimal effect on the detection of ROH (Table \ref{tab:Table 2}).
@@
 \begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot_final1.png}
+    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot_review.png}
     \centering
@@
     \hline
-    4x & 0.007 & 0.006 & 0.081 & 0.036 & 0.083 & 0.036 & 0.091 & 0.039\\
+    4x & 0.007 & 0.006 & 0.087 & 0.038 & 0.089 & 0.039 & 0.101 & 0.042\\
     \hline
-    0.2x & 0.003 & 0.003 & 0.068 & 0.03 & 0.068 & 0.026 & 0.076 & 0.029\\
+    0.5x & 0.006 & 0.005 & 0.085 & 0.036 & 0.087 & 0.036 & 0.098 & 0.039\\
     \hline
-    0.1x & 0.003 & 0.002 & 0.049 & 0.021 & 0.05 & 0.021 & 0.057 & 0.022\\
+    0.2x & 0.004 & 0.003 & 0.073 & 0.03 & 0.073 & 0.028 & 0.084 & 0.031\\
+    \hline
+    0.1x & 0.003 & 0.002 & 0.054 & 0.023 & 0.054 & 0.023 & 0.065 & 0.025\\
     \hline

### 2022-09-07 paper_overleaf/main.tex
Updates from Overleaf

@@
 The true positive rate is above 0.97 for both KIN and READ, while the false negative rate is below 0.02.
-One exception is  $2^{nd}$-degree relatives, where KIN has higher power than READ, and as coverage decreases from 4x to 0.05x, KIN's true positive rate decreases from 0.94 to 0.86, compared to range of 0.91 to 0.74 for READ.
-The false positive rate in this case remains below 0.02 for both methods.
-We also find that READ is strongly impacted by contamination (proportion of contamination$<3\%$) as the true positive rate is in the range 0.89 to 0.78 and 0.53 to 0.46 for 1st and 2nd degree relatives, respectively, and the false positive rate reaches up to 0.04 and 0.17 for unrelated individuals and 2nd degree relatives, respectively.
+One exception is  $2^{nd}$-degree relatives, where KIN has higher power than READ, and as coverage decreases from 4x to 0.05x, KIN's true positive rate decreases from 0.95 to 0.89, compared to range of 0.91 to 0.75 for READ.
+The false positive rate in this case remains below 0.03 for both methods.
+We also find that READ is strongly impacted by contamination (proportion of contamination$<3\%$) as the true positive rate is in the range 0.89 to 0.78 and 0.54 to 0.46 for 1st and 2nd degree relatives, respectively, and the false positive rate reaches up to 0.03 and 0.16 for unrelated individuals and 2nd degree relatives, respectively.
 By contrast, the correction implemented in KIN is sufficient to remove this bias, and the true and false positive rates remain as in the control.
 For simulations with ROH, we find that KIN also outperforms READ for 1st and 2nd degree relatives, although both the true and false positive rates increased for both methods compared to the control for 2nd degree relatives.
-The increases in false positives is likely due to ROH making related individuals more similar.
+The increase in false positives is likely due to ROH making related individuals more similar.
 Finally, we also found that KIN has good power to detect relatedness cases that are not detectable by READ, i.e.

### 2022-09-08 pypackage/kin/KIN/hmm_scripts/hmm_functions.py
changed pi to equilibrium value for transitions

@@ -374,13 +374,13 @@ def hmm(listind, hbdfolder, difffile, totalfile, listf, targets, pfile, Afiles, 
         pos=np.insert(pos, 0, 0, axis=0)
 
 
-        pi= np.array([1/3,1/3,1/3])
+        #pi= np.array([1/3,1/3,1/3])
 
         likall=[]
         for rel_cnt in range(len(rels)):
 
             A=Afiles[rel_cnt]
-
+            pi = matrix_power(A,10000)[0]
             B = np.zeros((inbr_states,np.shape(A)[0],win)) #Emission probability
 
             [b0, b1, b2, b3]=[1000,1000,1000,1000]

### 2022-09-08 pypackage/kingaroo/KINgaroo/KINgaroo_scripts/helpers.py

@@ -135,7 +135,7 @@ def get_merged_chrm(libraries, chrm, interval):
 
     id_diffs =[("hapProbs/hapProbs_{}_chrm%s_diffs.csv" %(chrm)).format(n) for n in libraries]
     id_diffs_list= hapProbsAll(haplist=id_diffs, hap='id')
-    print("id_diffs_list...............................................................", id_diffs_list)
+    #print("id_diffs_list...............................................................", id_diffs_list)
     diffs_list = findDiff(inds=probs_list, posall=pos_list)
     id_diffs_list[id_diffs_list==-9]=np.nan
 

### 2022-09-08 KIN_snakemake_version/hmm_inbred_constVar_testing.py
changed pi

@@ -416,7 +416,7 @@ def hmm(difffile, targets, totalfile, listind, listf, pfile, Afiles, resfiles, g
 
 
 
-        pi= np.array([1/3,1/3,1/3])
+        #pi= np.array([1/3,1/3,1/3])
 
         #A=A1
         likall=[]
@@ -424,7 +424,7 @@ def hmm(difffile, targets, totalfile, listind, listf, pfile, Afiles, resfiles, g
 
             Afile=Afiles[rel_cnt]
             A=np.array(pd.read_csv(Afile,sep=',',header=None,index_col=False))
-
+            pi=matrix_power(A,10000)[0]
             Bu = np.zeros((inbr_states,np.shape(A)[0],win)) #Emission probability
             #for st in instates:
              #   for re in range(np.shape(A)[0]):

### 2022-09-08 Simulations/KIN_hmm/hmm_inbred_constVar_testing.py

@@ -417,7 +417,7 @@ def hmm(difffile, targets, totalfile, listind, listf, pfile, Afiles, resfiles, l
 
 
 
-    pi= np.array([1/3,1/3,1/3])
+    #pi= np.array([1/3,1/3,1/3])
 
     #A=A1
     likall=[]
@@ -426,7 +426,7 @@ def hmm(difffile, targets, totalfile, listind, listf, pfile, Afiles, resfiles, l
         Afile=Afiles[rel_cnt]
 
         A=np.array(pd.read_csv(Afile,sep=',',header=None,index_col=False))
-
+        pi=matrix_power(A,10000)[0]
         Bu = np.zeros((inbr_states,np.shape(A)[0],win)) #Emission probability
         #for st in instates:
          #   for re in range(np.shape(A)[0]):

### 2022-09-08 paper_overleaf/main.tex

@@
     \caption{IBD sharing between siblings without and with runs of homozygosity (ROH).
-(A) Schematic of chromosomes for parent 1 (Par1)and parent 2 (Par2).
-On the left, the parents have unrelated chromosomes depicted with different colors, and on the right parent 1 has a homozygous pair of chromosomes.
-(B) In top row, recombined chromosomes inherited from parents by a pair of siblings are shown.
-The bottom row shows the expected proportion of differences ($p$) along the chromosome.
+(A) Schematic of chromosomes for two parents (Par1, Par2).
+(B) Schematics of recombinant chromosomes of two children (Sib1, Sib2).
+(C)  Expected differences between Sib1 and Sib2 ($p$) along the chromosome.
 In both cases, $p$ can take values of $p_0$, $p_1$ or $p_2$, which are the expected proportion of differences in IBD states 0,1 and 2, respectively.
+(D-F): Same as A-C, except parent 1 is assumed to be homozygous.
 }
@@
 \subsection{Input of KIN-HMM}\label{hmm_input}
-The inputs of our algorithm are i) the number of overlapping sites for the $w$-th window $N_w$ for which both samples have at least one read available, ii) the expected number of pairwise differences $D_w$ at these sites, and iii) the probability of ROH in windows, by default obtained from ROH-HMM described in section \ref{roh}.
+The inputs of our algorithm are i) the number of overlapping sites for the $w$-th window $N_w$ for which both samples have at least one read available, ii) the number of pairwise differences $D_w$ at these sites, and iii) the probability of ROH in windows, by default obtained from ROH-HMM described in section \ref{roh}.
 
-$D_w$ is the number of sites that are different between a pair of individuals.
-For high-coverage data, this can be estimated accurately by comparing genotypes, but for low-coverage samples, $D_w$ needs to be estimated.
+For high-coverage data, $D_w$ can be directly obtained by comparing genotypes, but for low-coverage samples, $D_w$ needs to be estimated from the sequencing data.
 The simplest approach is to randomly-sample a read from each position \cite{haak_massive_2015,prufer_complete_2014,green_draft_2010}.
-However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
+However, such an approach may result in loss of data, and hence we estimate $D_w$ by implicitly summing over all possible samplings:
 \begin{align}\label{eq:x}
@@
 There are three additional model parameters, $\pi, \mathbf{A}$ and $\delta$.
-The initial probabilities $\pi$ give the probabilities of being in each state $Z_0$ (at the beginning of each chromosome), which we set to uniform for simplicity.
 The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case (section \ref{simulat}).
+The initial probabilities $\pi$ give the probabilities of being in each state $Z_0$ (at the beginning of each chromosome), which we set to the stationary distribution of the transition matrix for simplicity.
 The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).
@@
 For example, when considering two bones from the same individual, we would expect the entire genome to have a pairwise difference of $p_2$, because two out of the four compared chromosomes are identical copies.
-However, in inbred regions all four chromosomes will be identical, and so the expected pairwise differences are zero ($p_3$ in equation \ref{eq:2}).
+However, in inbred regions all four chromosomes will be identical, and so the expected pairwise differences are zero ($p_4$ in equation \ref{eq:2}).
 Note however, that $p_0$ does not depend on the presence of ROH, since all comparisons are between unrelated chromosomes even if both individuals are homozygous at a particular locus.
@@
 p_0 & p_0 & p_0 \\
-p_1 & p_2 & p_3 \\
-p_2 & p_3 & p_3
+p_1 & p_2 & p_4 \\
+p_2 & p_4 & p_4
 \end{array}\right]
@@
 
-As explained above, we would expect $p_3$ to be zero.
-However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that we obtain better results by setting $p_3 = \frac{p_2}{2}$.
+As explained above, we would expect $p_4$ to be zero.
+However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that we obtain better results by setting $p_4 = \frac{p_2}{2}$, to take into account that many windows will only partially have four comparisons between identical chromosomes.
 
@@
 
-Using equation \ref{eq:2}, we simplify this by grouping all the terms that would result in the same $p_i$ and $\delta_i$, i.e.
+Using equation \ref{eq:2}, we simplify this by grouping all the terms that have the same number of pairwise comparisons between identical chromosomes, which would result in the same $p_i$ and $\delta_i$, i.e.
 
@@
 g_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\\
-g_{w3} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2}
+g_{w3} &= 0 \\
+g_{w4} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2},
 \end{align*}
+
+where $g_{w3}$ is always 0 because there is no case that leads to only three comparisons between identical chromosomes.
 
@@
 \begin{align}\label{eq:6}
-\mathcal{C} &= \sum_{i=0}^3 \sum_{w=1}^L \log P(D_{w}|N_w, \delta_i, p_i)g_{wi}
+\mathcal{C} &= \sum_{i=0}^4 \sum_{w=1}^L \log P(D_{w}|N_w, \delta_i, p_i)g_{wi}
 \end{align}
@@
 We output the relatedness corresponding to the maximum likelihood model.
-Because the models we test are not nested, standard likelihood-ratio theory cannot be used to generate a confidence value.
+Since we compare models where the  parameters are not subset of each other, standard likelihood-ratio theory for nested models cannot be used to obtain confidence intervals.
 Instead we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig.
@@
 In Fig.\ref{fig1:ibd} we show the inferred IBD fragments from different KIN-HMMs when they are applied to simulated data from a pair of siblings.
-In this case, the siblings-model predictions match true IBD states the most, as reflected by the highest log-likelihood.
+The models for identical, parent-child or unrelated relationships allow for just one IBD state, resulting in a flat line and low likelihood for this data.
+The other three cases all allow for different IBD-states, but the siblings-model predictions match true IBD states the most, as reflected by the highest log-likelihood and the close correspondence of the inferred and true IBD states.
 
@@
 The remaining panels show the IBD states predicted by particular relatedness models.
-The log likelihood value for each model is shown on upper left corner of the panel.
-Light and shaded background represent distinct chromosomes.}
+The log-likelihood value for each model is shown on upper left corner of the panel.
+Light and shaded backgrounds represent distinct chromosomes.}
     \label{fig1:ibd}
@@
 Since the source of ROH may not be known, we estimate both transitions and emissions.
-We calculate the emissions using a beta-binomial likelihood, and fix the mean of the distributions corresponding to $Y_w$ = 3 and $Y_w$ = 2 at expected proportion of differences in a homozygous tract ($p_3$) and expected proportion of differences in a non-homozygous tract ($p_2$) respectively.
+We calculate the emissions using a beta-binomial likelihood, and fix the mean of the distributions corresponding to $Y_w$ = 3 and $Y_w$ = 2 at expected proportion of differences in a homozygous tract ($p_4$) and expected proportion of differences in a non-homozygous tract ($p_2$) respectively.
 The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w=3$ or the state $Y_w=2$ in each window.
@@
 
+\begin{table}[h!]
+\caption{\label{tab:Table 3}Model performance for ROH prediction.
+Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination), R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
+    \hline
+    Coverage & R(Sensitivity) & R(Specificity) & RA(Sensitivity) & RA(Specificity) & RC(Sensitivity) & RC(Specificity)\\
+    \hline
+    4x & 0.007 & 0.006 & 0.087 & 0.038 & 0.089 & 0.039 & 0.101 & 0.042\\
+    \hline
+    0.5x & 0.006 & 0.005 & 0.085 & 0.036 & 0.087 & 0.036 & 0.098 & 0.039\\
+    \hline
+    0.2x & 0.004 & 0.003 & 0.073 & 0.03 & 0.073 & 0.028 & 0.084 & 0.031\\
+    \hline
+    0.1x & 0.003 & 0.002 & 0.054 & 0.023 & 0.054 & 0.023 & 0.065 & 0.025\\
+    \hline
+\end{tabular}
+\label{table2}
+\end{table}
+
 \paragraph{IBD prediction:}
@@
 \ref{fig4:IBDstate_accuracy}).
-The accuracy for coverages of 0.1x or higher is consistent and varies between 1.0 and 0.836, depending on relatedness.
-However, the accuracy decreases at lower coverages for most relatedness cases and ranges from 1.0 to 0.563 at 0.03x.
+The accuracy for coverages of 0.1x or higher is consistent and varies between 1.0 and 0.78, depending on relatedness.
+However, the accuracy decreases at lower coverages for most relatedness cases and ranges from 1.0 to 0.48 at 0.03x.
 The exceptions are for identical individuals and parent-child pairs, where the accuracy is nearly perfect for all investigated coverages, even at 0.03x coverage.
@@
 Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.
-Accuracy is defined as the sum of true positives and true negatives divided by the total number of windows.
-We consider a prediction is true if it matches the IBD state at the central position of the window.}
+Accuracy is defined as the proportion of correctly predicted IBD states, when compared to the central position of the window.}
     \label{fig4:IBDstate_accuracy}
@@
 S10).
-We find that lcMLkin creates four different clusters for different relatedness cases (identical, parent-child, $2^{nd}$ degree and unrelated).
+We find that lcMLkin creates four different clusters based on the coefficient of relatedness ($r$) and the proportion of the genome in the unrelated state ($k_0)$ for different relatedness cases (identical, parent-child, $2^{nd}$ degree and unrelated).
 However, the location of these clusters strongly deviate from the expected values, likely due to the low coverage and ROH.
-Remains from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $k_0 \approx 0.3$, $r \approx 0.6$.
+Remains from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $k_0 \approx 0.3$, $r \approx 0.6$ for lcMLkin (Figure S10).
 
@@
 S11).
-For 20 of these comparisons, KIN predicts a third-degree relationship, which READ classifies as unrelated.
-
+Twenty of these comparisons, KIN predicts to be third-degree relationships, and  READ concordantly classifies them as unrelated (as it only infers first and second degree relationships).
 lcMLkin predicts $3^{rd}$ - $5^{th}$ in 14 of these cases, unrelated in 1 case (see Fig.
@@
 For 10 additional pairs, lcMLkin predicts $3^{rd}$ - $5^{th}$ degree, but KIN infers them to be unrelated.
+There are a total of 83 pairs where READ obtains a confident call ($s.d.>1$), but KIN does not ($\Delta LL<1$).
+In 80 of these cases, READ classifies them as unrelated, KIN classifies $3^{rd}$ degree, while lcMLkin predicts $3^{rd}$ - $5^{th}$ degree.
+In rest 3 cases READ classifies unrelated, while KIN and lcMLkin agree on $2^{nd}$ degree.
 
@@
 For detecting ROH, there is only one method available that works with low coverage ancient DNA \cite{ringbauer_parental_2021}.
-This software can infer ROH at coverages $\geq0.3x$, but requires a reference panel.
-In the absence of a method to detect long ROH at low coverages without a reference panel, we developed a HMM based method, and tested it on simulations to show that it can be used reliably at coverage $\geq 0.1x$ to detect ROH regions with lengths on the order of 10cM.
+This software can infer ROH at coverages $\geq0.3x$, but requires a large reference panel.
+Instead, our method detects long ROH based on just the expected heterozygosity $p_0$, which can be estimated from a small number of unrelated individuals.
+On simulations, we find that our method reliably inferrs ROH regions  with lengths on the order of 10cM from samples sequenced to coverage $\geq 0.1x$ .
 In our simulations, we find that adding long ROH ($~17\%$) to simulations slightly improves the power of both KIN and READ, particularly at lower coverages.

### 2022-09-11 supplementary_info/beta_bnds.py
replaced i=3 by i=4

@@ -639,22 +639,22 @@ def plot_betas(dfile, tfile, pair, thresh, rel, libraries, listf, Afile, pfile, 
 
         a=betas[2]
         b=betas[3]
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'r-', lw=3, alpha=0.6, label="$Z_w$=0")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'r-', lw=3, alpha=0.6, label="i=0")
         #plt.hist(diff/total,alpha=0.3)
 
         a=betas[0]
         b=betas[1]
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'b-', lw=3, alpha=0.6, label="$Z_w$=1")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'b-', lw=3, alpha=0.6, label="i=1")
 
         a=betas[4]
         b=betas[5]
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'g-', lw=3, alpha=0.6, label="$Z_w$=2")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'g-', lw=3, alpha=0.6, label="i=2")
         #plt.hist(diff/total,alpha=0.3)
 
         a=betas[6]
         b=betas[7]
 
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'y-', lw=3, alpha=0.6, label="$Z_w$=4")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'y-', lw=3, alpha=0.6, label="i=4")
         ax[i].hist(diff/total,alpha=0.3,weights=np.ones(len(total)) / len(total))
         ax[i].set_ylim(0,0.4)
 

### 2022-09-11 paper_overleaf/main.tex

@@
 
-\noindent Inferring genetic kinship of ancient individuals can provide insights into their culture and social hierarchy, and is relevant for downstream genetic analyses.
+\noindent Genetic kinship of ancient individuals can provide insights into their culture and social hierarchy, and is relevant for downstream genetic analyses.
 However, estimating relatedness from ancient DNA is difficult due to low-coverage, ascertainment bias, or contamination from various sources.
 Here, we present KIN, a method to estimate the relatedness of a pair of individuals from the identical-by-descent segments they share.
-KIN accurately classifies up to 3rd-degree relatives using >0.05x sequence coverage and differentiates siblings from parent-child.
-It incorporates additional models to adjust for contamination and detect inbreeding, which improves classification accuracy, and can be applied to any diploid species.
+KIN accurately classifies up to 3rd-degree relatives using $>0.05x$ sequence coverage and differentiates siblings from parent-child.
+It incorporates additional models to adjust for contamination and detect inbreeding, which improves classification accuracy.
 \end{abstract}
@@
 \ref{fig0:schematic}).
-These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}.
+These IBD probabilities can directly be used to categorize their relatedness as shown in Table \ref{tab:Table 1}.
 One can also use these probabilities to estimate the coefficient of relatedness $r$, which is defined as the proportion of the genome that is IBD.
@@
 
-There are three additional model parameters, $\pi, \mathbf{A}$ and $\delta$.
+There are three additional model parameters, $\bm{\pi}, \mathbf{A}$ and $\bm{\delta}$.
 The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case (section \ref{simulat}).
-The initial probabilities $\pi$ give the probabilities of being in each state $Z_0$ (at the beginning of each chromosome), which we set to the stationary distribution of the transition matrix for simplicity.
-The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).
-For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.
+The initial probabilities $\bm{\pi}$ give the probabilities of being in each state $Z_0$ (at the beginning of each chromosome), which we set to the stationary distribution of the transition matrix for simplicity.
+The overdispersion-parameter $\bm{\delta}$ takes into account that SNPs in each window vary in their allele frequency (see next setion).
+For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \bm{\pi})$.
 
@@
 \begin{align}\label{eq:1}
-\log P(\BD,\BZ|\BH, \theta, \delta) &= \log P(\BD|\BZ,\BH, \theta, \delta) + \log P(\BZ |\theta) \nonumber\\
-&= \sum_w \log P(D_w|Z_w,H_w, N_w, \delta) + \sum_w \log P(Z_w |Z_{w-1},A) + \log P(Z_0|\pi)\text{.}
+\log P(\BD,\BZ|\BH, \theta, \bm{\delta}) &= \log P(\BD|\BZ,\BH, \theta, \bm{\delta}) + \log P(\BZ |\theta) \nonumber\\
+&= \sum_w \log P(D_w|Z_w,H_w, N_w, \bm{\delta}) + \sum_w \log P(Z_w |Z_{w-1},\bm{A}) + \log P(Z_0|\bm{\pi})\text{.}
 \end{align}
 
-Here, $\BZ$ is not dependent on $\BH$ and $\delta$.
+Here, $\BZ$ is not dependent on $\BH$ and $\bm{\delta}$.
 
@@
 
-Using this setup, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta, \delta)$ from equation \ref{eq:1} and optimize them for $\delta$.
+Using this setup, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta, \bm{\delta})$ from equation \ref{eq:1} and optimize them for $\bm{\delta}$.
 The simplest model is to assume that sites in each window are equally distributed and independent, in this case we could use the  binomial likelihood:
@@
 
-Taken together, we can summarize $p$ in the following matrix, where rows give the state of $Z_w$, and columns of $H_w$:
+Taken together, we can summarize $\bm{p}$ in the following matrix, where rows give the state of $Z_w$, and columns of $H_w$:
 \begin{equation}\label{eq:2}
@@
 S2).
-We take this into account by adding an over-dispersion parameter $\delta$.
+We take this into account by adding an over-dispersion parameter $\bm{\delta}$.
 Just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ depends on the number of chromosomes compared, and so each of the  four $p_i$ has a corresponding $\delta_i$ parameter.
@@
 \begin{align}\label{eq:3}
-P(D_{w}|Z_w,H_w,N_w, \delta) &\sim BB[D_w; p_i, \delta_i, N_w] \nonumber\\
+P(D_{w}|Z_w,H_w,N_w, \bm{\delta}) &\sim BB[D_w; p_i, \delta_i, N_w] \nonumber\\
 &= \binom{N_w}{D_w}\frac{B(D_w+p_i \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})},
@@
 S5, S6).
-We avoid this problem by constraining the $\delta$ such that the beta distributions for the different relatedness cases overlap by at most one standard deviation.
+We avoid this problem by constraining the $\bm{\delta}$ such that the beta distributions for the different relatedness cases overlap by at most one standard deviation.
 
@@
 We call the vector with the number of differences $\mathbf{\Delta}$, and the vector with the number of sites with at least two reads $\mathbf{M}$.
-Our model has two possible hidden states: homozygous state ($Y_w=3$), and non-homozygous state ($Y_w=2$).
+Our model has two possible hidden states: homozygous state ($Y_w=4$), and non-homozygous state ($Y_w=2$).
 As above, we collect the hidden states in a vector $\mathbf{Y} = (Y_1, \dots Y_w, \dots, Y_L)$.
@@
 \begin{align}\label{eq:10}
-    log P(\mathbf{\Delta},\mathbf{Y}|\Theta,\delta) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta, \delta) + log P(\mathbf{Y}|\Theta)\nonumber\\
- &= [\sum_{w} log P(\Delta_w|Y_w, M_w, \delta) + \sum_{w} log P(Y_w|Y_{w-1}, A)] + P(Y_0| \pi),
+    log P(\mathbf{\Delta},\mathbf{Y}|\Theta,\bm{\delta}) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta, \bm{\delta}) + log P(\mathbf{Y}|\Theta)\nonumber\\
+ &= [\sum_{w} log P(\Delta_w|Y_w, M_w, \bm{\delta}) + \sum_{w} log P(Y_w|Y_{w-1}, \bm{A})] + P(Y_0| \bm{\pi}),
 \end{align}
 
-where $\Theta$ is a vector of initial probabilities ($\pi$), transition matrix ($A$) and $M$.
+where $\Theta$ is a vector of initial probabilities ($\bm{\pi}$), transition matrix ($\bm{A}$) and $\bm{M}$.
 
 Since the source of ROH may not be known, we estimate both transitions and emissions.
-We calculate the emissions using a beta-binomial likelihood, and fix the mean of the distributions corresponding to $Y_w$ = 3 and $Y_w$ = 2 at expected proportion of differences in a homozygous tract ($p_4$) and expected proportion of differences in a non-homozygous tract ($p_2$) respectively.
-The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w=3$ or the state $Y_w=2$ in each window.
+We calculate the emissions using a beta-binomial likelihood, and fix the mean of the distributions corresponding to $Y_w$ = 4 and $Y_w$ = 2 at expected proportion of differences in a homozygous tract ($p_4$) and expected proportion of differences in a non-homozygous tract ($p_2$) respectively.
+The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w=4$ or the state $Y_w=2$ in each window.
 The maximization step for emissions is analogous to that in the KIN model (eq.
@@
 \begin{align}\label{eq:11}
-\mathcal{C} = \sum_{w=1}^L \sum_{i=2}^3 P(\Delta_w|\delta_{i},p_{i}) \Gamma_{wi} ,
+\mathcal{C} = \sum_{w=1}^L \sum_{i} P(\Delta_w|\delta_{i},p_{i}) \Gamma_{wi} ,
 \end{align}
-where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and over-dispersion parameter $\delta$ similar to eq. \ref{eq:3}.
+where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and over-dispersion parameter $\delta$ similar to eq.
+\ref{eq:3}, and $i$ can take values 2 and 4 corresponding to the hidden states $Y_w$.
 
@@
 To evaluate the effect of sequence coverage on the performance of KIN, we generated artificial pileups at each polymorphic site for each individual following a Poisson distribution with 6 different average depths varying between 4$x$ and 0.03$x$ (see section \ref{method}).
-To mimic ROH, for some pedigrees, we picked a single allele at heterozygous sites in some regions as determined by a Markov chain so that about $17\%$ of the genome is ROH.
+To mimic ROH, for some pedigrees, we picked a single allele at heterozygous sites in some regions as determined by a Markov chain so that on an average about $17\%$ of the genome is ROH.
 We also created versions with contamination, by introducing alleles from distantly related individuals, and ascertainment bias, by selecting polymorphic sites identified in a subset of the individuals (see section \ref{method}).
@@
 \paragraph{ROH detection:}
-We first evaluated the performance of the ROH-HMM ( Fig.\ref{fig2:ROH}).
-We computed the proportion of ROH in all windows and compared it to the posterior probability of the ROH state, by computing the bias (i.e., the difference between both) and its variance across windows.
-Without ROH, we correctly inferred the absence of ROH as the observed bias ranges from $0.3$ to $0.7\%$ (variance $\leq0.006$; Table \ref{tab:Table 2}).
-With ROH, we observed a slight bias of $5$ to $9\%$ more ROH than simulated (variance $\leq0.04$).
-The largest bias observed ($10.1\%$) was for the datasets that included contamination.
-Ascertainment bias had a minimal effect on the detection of ROH (Table \ref{tab:Table 2}).
-Surprisingly, the bias was lower for lower coverages, perhaps because there is only one possible ROH state per window to model a continuous variable, viz.
-proportion of window in ROH, and this state is inferred more confidently with more data.
-Overall, the model predicts ROH well in all cases, even at a coverage of 0.1x (Fig.\ref{fig2:ROH}).
+In Fig.
+\ref{fig2:ROH}, we present an example of data and inference of the ROH-HMM for simulations with and without ROH, potentially with ascertainment bias and contamination.
+In all cases, we find that the inferred ROH closely matches the simulations, but the confidence in the classification tends to increase with the simulated coverage.
+
+A systematic evaluation of the performance of the ROH-HMM is given in Table \ref{tab:Table 2}: For the purpose of this analysis, we classified all windows that were at least 20\% homozygous as ROH, and the remainder as non-ROH.
+Likewise, we classified all windows with a posterior probability of ROH of at least 20\% as ROH.
+We used these cases to compute sensitivity (Se) and specificity (Sp).
+In the control case (simulations with no ascertainment, contamination, or ROH) we see that the specificity remains $\geq0.99$ for all coverages.
+In the cases where we simulate ROH, we find that sensitivity decreases from 97\% at 4x coverage to less than 65\% at 0.05x coverage, but specificity remains high at above 0.96 for all cases, suggesting that the number of erroneously called ROH segments is low.
 
@@
     \caption{Estimation of ROH probabilities along the genome in simulations.
-Top panel shows proportion of differences in a simulated individual in windows along the genome, and the bottom panel shows probability of seeing no ROH.
+The top row in each panel shows the proportion of differences in a simulated individual along the genome.
+In the bottom row, the dotted line shows the proportion of each window in ROH, and the solid lines show the estimated probability of not being in the ROH state.
 (A) Simulation with no ascertainment, contamination, or ROH.
@@
 Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination), R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
-\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|c|c|}
     \hline
-    Coverage & Control(bias) & Control(var) & R(bias) & R(var) & RA(bias) & RA(var) & RC(bias) & RC(var)\\
+    Coverage & Control (Sp) & R (Se) & R (Sp) & RA (Se) & RA (Sp) & RC (Se) & RC (Sp)\\
     \hline
-    4x & 0.007 & 0.006 & 0.087 & 0.038 & 0.089 & 0.039 & 0.101 & 0.042\\
+    4x & $>$0.99 & 0.97 & 0.98 & 0.97 & 0.98 & 0.97 & 0.97\\
     \hline
-    0.5x & 0.006 & 0.005 & 0.085 & 0.036 & 0.087 & 0.036 & 0.098 & 0.039\\
+    0.5x & $>$0.99 & 0.96 & 0.98 & 0.96 & 0.98 & 0.96 & 0.97\\
     \hline
-    0.2x & 0.004 & 0.003 & 0.073 & 0.03 & 0.073 & 0.028 & 0.084 & 0.031\\
+    0.2x & $>$0.99 & 0.92 & 0.98 & 0.91 & 0.98 & 0.93 & 0.97\\
     \hline
-    0.1x & 0.003 & 0.002 & 0.054 & 0.023 & 0.054 & 0.023 & 0.065 & 0.025\\
+    0.1x & $>$0.99 & 0.85 & 0.98 & 0.83 & 0.97 & 0.82 & 0.97\\
     \hline
-\end{tabular}
-\label{table1}
-\end{table}
-
-\begin{table}[h!]
-\caption{\label{tab:Table 3}Model performance for ROH prediction.
-Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination), R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
-\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
-    \hline
-    Coverage & R(Sensitivity) & R(Specificity) & RA(Sensitivity) & RA(Specificity) & RC(Sensitivity) & RC(Specificity)\\
-    \hline
-    4x & 0.007 & 0.006 & 0.087 & 0.038 & 0.089 & 0.039 & 0.101 & 0.042\\
-    \hline
-    0.5x & 0.006 & 0.005 & 0.085 & 0.036 & 0.087 & 0.036 & 0.098 & 0.039\\
-    \hline
-    0.2x & 0.004 & 0.003 & 0.073 & 0.03 & 0.073 & 0.028 & 0.084 & 0.031\\
-    \hline
-    0.1x & 0.003 & 0.002 & 0.054 & 0.023 & 0.054 & 0.023 & 0.065 & 0.025\\
+    0.05x & 0.99 & 0.65 & 0.96 & 0.55 & 0.96 & 0.57 & 0.96\\
     \hline
@@
 We find that adding contamination, ascertainment bias in our simulations has little effect on IBD prediction.
-Adding ROH to the simulations reduces the IBD prediction in two cases: the accuracy for second degree is decreased by $1.5$ to $5\%$ and for siblings by $5$ to $17\%$ (Fig.
+Adding ROH to the simulations reduces the IBD prediction in two cases: the average accuracy for second degree decreases from 0.89 to 0.85 and for siblings from 0.81 to 0.70 (Fig.
 S9).
@@
 The false positive rate in this case remains below 0.03 for both methods.
-We also find that READ is strongly impacted by contamination (proportion of contamination$<3\%$) as the true positive rate is in the range 0.89 to 0.78 and 0.54 to 0.46 for 1st and 2nd degree relatives, respectively, and the false positive rate reaches up to 0.03 and 0.16 for unrelated individuals and 2nd degree relatives, respectively.
-By contrast, the correction implemented in KIN is sufficient to remove this bias, and the true and false positive rates remain as in the control.
+
+To investigate the impact of contamination, we performed simulations where we added up to 3\% contamination to some individuals.
+We also find that READ is strongly impacted by contamination  as the true positive rate is in the range 0.89 to 0.78 and 0.54 to 0.46 for 1st and 2nd degree relatives, respectively, and the false positive rate reaches up to 0.04 and 0.17 for unrelated individuals and 2nd degree relatives, respectively.
+In comparison, we also ran KIN, giving it the simulated contamination amounts for each individual.
+We found that the correction implemented in KIN is sufficient to remove this bias, and the true and false positive rates remain as in the control (Fig.
+\ref{fig3:Comparison_READ_KIN}).
+We also perform an analysis where we misspecify the contamination, and find that the effect of modest misspecification is small (Fig.
+S13).
+
 For simulations with ROH, we find that KIN also outperforms READ for 1st and 2nd degree relatives, although both the true and false positive rates increased for both methods compared to the control for 2nd degree relatives.
@@
 parent-child, siblings and, to a lower level, 3rd degree relatives (Fig.
-\ref{fig3:Comparison_READ_KIN}, for details see table S4).
+\ref{fig3:Comparison_READ_KIN}).
 
@@
 In 80 of these cases, READ classifies them as unrelated, KIN classifies $3^{rd}$ degree, while lcMLkin predicts $3^{rd}$ - $5^{th}$ degree.
-In rest 3 cases READ classifies unrelated, while KIN and lcMLkin agree on $2^{nd}$ degree.
+The remaining cases READ classifies unrelated, while KIN and lcMLkin call a $2^{nd}$-degree relationship.
+Thus, the vast majority of differences can be explained by READ not considering 3rd-degree relationships.
 
@@
 On simulations, we find that our method reliably inferrs ROH regions  with lengths on the order of 10cM from samples sequenced to coverage $\geq 0.1x$ .
-In our simulations, we find that adding long ROH ($~17\%$) to simulations slightly improves the power of both KIN and READ, particularly at lower coverages.
+In our simulations, we find that adding long ROH ($\approx17\%$) to simulations slightly improves the power of both KIN and READ, particularly at lower coverages.
 This is likely because ROH actually reduces the variance in differences depending on the relatedness, and makes it easier to correctly classify relatedness.
@@
 
-We show that KIN reliably detects IBD tracts for up to 0.1x coverage even in presence of contamination and ascertainment bias, and the accuracy for IBD detection reduces with lower coverages.
+We show that KIN reliably detects IBD tracts for $\geq0.1x$ coverage even in the presence of contamination and ascertainment bias, and the accuracy for IBD detection reduces with lower coverages.
 Adding ROH to the simulations adversely affects IBD prediction in siblings (Fig.
@@
 KIN can also infer third degree relation with $\approx$ 30,000 overlapping polymorphic sites.
-We show that when applied to Neandertal specimens from Chagyrskaya and Okladnikov Caves, KIN identifies a pair of $1^{st}$-Degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples \cite{laurits_skov_genetic_nodate}.
+We show that when applied to Neandertal specimens from Chagyrskaya and Okladnikov Caves, KIN identifies a pair of $1^{st}$-Degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples (one sample is male and the other female) \cite{laurits_skov_genetic_nodate}.
 In addition, KIN identifies a pair of $3^{rd}$ degree relatives.
@@
 We may get around this problem by using an estimate of $p_0$, calculated from known a pair of unrelated individuals from same population, or another population with similar diversity.
-We provide the user with an option to provide an estimate of $p_0$.
+We provide the user with an option to give an estimate of $p_0$.
 The current implementation of KIN is restricted to six relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.
@@
 S1).
-For each individual, we simulate 22 chromosomes with length $L=100$ Mb and a recombination rate of $r=10^{-8}$ per base pair per generation.
+For each individual, we simulate 22 chromosomes with length $L\approx96$ Mb (same as chromosome 13) and a recombination rate of $r=10^{-8}$ per base pair per generation.
 We introduce mutations using an infinite sites model with rate  $\mu= 10^{-8}$ per base pair per generation.
@@
 For the pedigree simulations, we first simulate a recombined set of chromosomes for either parent, and combine them to create the progeny.
-We simulate recombination by first drawing the number of breakpoints as a  Poisson random variable with parameter $rL$, and use a uniform distribution on $[1, L]$ to sample the positions of recombination points.
+There are two different ways in which we generate recombination points.
+For the estimation of transition probabilities, we simulate recombination by first drawing the number of breakpoints as a Poisson random variable with parameter $rL$, and use a uniform distribution on $[1, L]$ to sample the positions of recombination points.
+For the testing of our method, we use Ped-sim \cite{caballero_crossover_2019} to simulate recombination points.
+This allows us to take into account sex-specific recombination rates and crossover interference, and thus is expected to give a more realistic recombination landscape.
 The pedigree simulations result in nine addition individuals, resulting in a total sample of 17 individuals.
@@
 KIN requires a transition matrix for each relatedness case, which we estimate by counting the transition between IBD states for all pairs of individuals with that relatedness in a training set of 1000 simulations from our pedigree (Fig.
-S1).
+S3).
 For two cases, siblings and grandparent-grandchild, it is possible to write down the theoretical expectation of the transition matrix:
@@
 Apart from the related and unrelated individuals in Pop1, we simulated more haploid individuals in three other populations to create scenarios with ascertainment bias and contamination (Fig.
-S1).
+S3).
 We simulated two individuals to form an individual each from two other populations (Pop2, Pop3) with split time of 3500 and 4500 generations with Pop1, and sampling time of 2000 generations and 4000 generations ago respectively.
@@
 \section{Data and material availability}
-An open-source implementation of KIN and KINgaroo in python along with a toy example dataset, and the scripts to generate our test simulations are available on GitHub \cite{divyaratan_popli_httpsgithubcomdivyaratanpoplikinship_inferencereleasestagv312_2022}.
-We have deposited the version of the software used in the manuscript, along with the above mentioned files on Zenodo \cite{divyaratan_popli_httpsdoiorg105281zenodo6561015_nodate}.
+An open-source implementation of KIN and KINgaroo in python along with a toy example dataset, and the scripts to generate our simulations are available on GitHub \cite{divyaratan_popli_httpsgithubcomdivyaratanpoplikinship_inferencereleasestagv313_2022}.
+We have deposited the version of the software used in the manuscript, along with the above mentioned files on Zenodo \cite{divyaratan_popli_httpsdoiorg105281zenodo7067142_nodate}.
 The analysed datasets from Bronze Age Lech Valley were generated in a previous study (ENA, accession no.

### 2022-09-11 supplementary_info/Rscripts/misscontam_plot.R
Updates from Overleaf

@@ -1,53 +1,0 @@
-
-library(ggplot2)
-#install.packages("magrittr")
-#install.packages("dplyr")
-library(magrittr)
-library(dplyr)
-library(tidyr)
-
-#inf='/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz/hmm_numba_incorrect_roh_i1_i1_insteadOf_i1_i2/roc/contam0_inbred0_model_performance_allroc_asc0.csv.gz'
-#infile='/mnt/diversity/divyaratan_popli/review_sim/hmm_numba/miss_contam/output/missC10/contam1_inbred0_model_performance_allLikelihoods_inphapProbs_asc0.csv'
-
-contamination_plot <- function(infile,outfile){
-  
-  
-  compinp=read.csv(file = infile, header = TRUE)
-  compinp$Relatedness <- factor(compinp$Relatedness, levels = c("fir", "sec", 'un_all','pc','sib','deg3'))
-  
-  data <- compinp %>%
-    pivot_longer(c(True_positive,False_positive), names_to = "T_F", values_to = "classification_proportion")#Plot
-  data$T_F <- factor(data$T_F, levels= c('True_positive','False_positive'))
-  
-  #data<-data[!(data$method=="read_inppshap" | data$method=="allLikelihoods_inppshap"),]
-  #data = subset(data, select = -c(method) )
-  
-  data$Coverage <- as.factor(data$Coverage)
-  
-  #changing some column names:
-  
-  Relatedness.labs <- c("Unrelated", "3rd Degree", "2nd Degree", "1st Degree", "Siblings", "Parent-Child")
-  names(Relatedness.labs) <- c("un_all", "deg3", "sec", "fir", "sib", "pc")
-  
-  T_F.labs <- c("False Positive", "True Positive")
-  names(T_F.labs) <- c("False_positive", "True_positive")
-  
-  plot<-ggplot(data=data, aes(x=contamination, y=classification_proportion, color=Coverage))
-  plot<-plot+geom_line()
-  plot<- plot + facet_grid(T_F ~ Relatedness, labeller = labeller(T_F = T_F.labs, Relatedness = Relatedness.labs), scales="free")
-  #plot + geom_hline(yintercept=0.05, linetype="dashed")
-  plot <- plot + geom_hline(data = data.frame(T_F='False_positive'), aes(yintercept = 0.05), linetype = "dotted")
-  plot <- plot + geom_vline(data = data, aes(xintercept = 2.5), linetype = "dashed")
-  
[…]

### 2022-09-11 supplementary_info/forcedROH.py
changed Y_w label

@@ -464,7 +464,7 @@ def plotbetas(dfile, tfile, ind, chrmf, p1file, upA, targets, outplot):
 
         a=betas[0]
         b=betas[1]
-        ax[i].plot(x, beta.pdf(x, a, b)/500,'b-', lw=3, alpha=0.6, label="$Y_w$=3")
+        ax[i].plot(x, beta.pdf(x, a, b)/500,'b-', lw=3, alpha=0.6, label="$Y_w$=4")
 
         a=betas[2]
         b=betas[3]

### 2022-09-11 paper_overleaf/KIN_a_method_to_estimate_relatedness_from_low_coverage_data_supplements.tex
Updates from Overleaf

@@
 \begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_betaplot.png}
+    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_betaplot1.png}
     \centering
@@
     \includegraphics[width=18cm]{supplementary_info/plots/contam_missspecification.png}
-    \caption{False positive and true positive rates as a function of incorrect contamination estimates provided to KINgaroo.
+    \caption{False positive and true positive rates as a function of incorrect contamination estimates provided to the ROH-HMM.
 Dashed vertical line represents true contamination (2.5\%) in simulations, and dotted horizontal line in top panel represents $5\%$ false positive rate.}

### 2022-09-11 paper_overleaf/KIN_a_method_to_estimate_relatedness_from_low_coverage_data_supplements.tex
Updates from Overleaf

@@
     \caption{False positive and true positive rates as a function of incorrect contamination estimates provided to the ROH-HMM.
-Dashed vertical line represents true contamination (2.5\%) in simulations, and dotted horizontal line in top panel represents $5\%$ false positive rate.}
+Dashed vertical line represents true contamination (2.5\%) in simulations, and dotted horizontal line in top panel represents $5\%$ false positive rate.
+For this analysis we varied the contamination level in one individual from 0.5\% to 5.5\% while keeping contamination estimates for everyone else constant.
+We then analyzed the relatedness classification for this individual with all other individuals for power analysis.}
     \label{figS10:cnt}

### 2022-11-17 plots/Rscripts/ROHlegend.R
changed color to colorblind friendly

@@ -15,9 +15,9 @@ make_panel1 <- function(data,acrval,leg){
     )) +
     scale_fill_manual(values = c("grey90", "white")) +
     theme_void()
-  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.5, size=0.8)
+  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.7, size=0.8)
   plot1<-plot1 + theme_bw()+ theme(legend.position = "none", axis.title.x = element_blank())+labs(y= "Heterozygosity", x = "Windows along genome") +
-    scale_color_manual(values=c("0.1"="Green", "0.2"="Blue","4"="Red")) +
+    scale_color_manual(values=c("0.1"="#999933", "0.2"="#117733","4"="#661100")) +
     theme(panel.grid.major = element_blank(),
           panel.grid.minor = element_blank(),
           strip.background = element_blank(),
@@ -39,10 +39,10 @@ make_panel2 <- function(data,acrval,leg){
     theme_void()
   plot2<-plot2+geom_line(aes(x=Wins, y=True_ROH, linetype="Truth"))
   #plot2
-  plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.5, size=0.8)
+  plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.7, size=0.8)
 
   plot2<-plot2+scale_linetype_manual(values=c("Truth"="dashed", "Inferred"="solid"))
-  plot2 <- plot2 + scale_color_manual(values=c("0.1"="Green", "0.2"="Blue","4"="Red"))
+  plot2 <- plot2 + scale_color_manual(values=c("0.1"="#999933", "0.2"="#117733","4"="#661100"))
   if (leg==1){
     plot2<-plot2 + theme_bw() + theme(legend.position = "bottom")+labs(y= "1 - P(ROH)", x = "Windows along genome")
     plot2<- plot2+ guides(linetype = guide_legend(nrow = 1, byrow = TRUE))+ guides(colour = guide_legend(nrow = 1, byrow = TRUE))

### 2022-11-17 plots/Rscripts/ROHplot.R

@@ -15,9 +15,9 @@ make_panel1 <- function(data,acrval,leg){
     )) +
     scale_fill_manual(values = c("grey90", "white")) +
     theme_void()
-  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.5, size=0.8)
+  plot1<-plot1+geom_line(aes(colour=Coverage),alpha=0.7, size=0.8)
   plot1<-plot1 + theme_bw()+ theme(legend.position = "none", axis.title.x = element_blank())+labs(y= "Het.") +
-    scale_color_manual(values=c("0.1"="Green", "0.2"="Blue","4"="Red")) +
+    scale_color_manual(values=c("0.1"="#999933", "0.2"="#117733","4"="#661100")) +
     theme(panel.grid.major = element_blank(),
           panel.grid.minor = element_blank(),
           strip.background = element_blank(),
@@ -39,10 +39,10 @@ make_panel2 <- function(data,acrval,leg){
     theme_void()
   plot2<-plot2+geom_line(aes(x=Wins, y=True_ROH, linetype="Truth"))
   #plot2
-  plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.5, size=0.8)
+  plot2<-plot2+geom_line(aes(x=Wins, y=Model_pred, linetype="Inferred", colour=Coverage), alpha=0.7, size=0.8)
 
   plot2<-plot2+scale_linetype_manual(values=c("Truth"="dashed", "Inferred"="solid"))
-  plot2 <- plot2 + scale_color_manual(values=c("0.1"="Green", "0.2"="Blue","4"="Red"))
+  plot2 <- plot2 + scale_color_manual(values=c("0.1"="#999933", "0.2"="#117733","4"="#661100"))
   if (leg==1){
     plot2<-plot2 + theme_bw() + theme(legend.position = "bottom", axis.title.x = element_blank())+labs(y= "1 - P(ROH)")
     plot2<- plot2+ guides(linetype = guide_legend(nrow = 1, byrow = TRUE))+ guides(colour = guide_legend(nrow = 1, byrow = TRUE))

### 2022-11-17 paper_overleaf/main.tex
updated from overleaf

@@
 \usepackage[super]{nth}
-\usepackage{hyperref}
+
 \usepackage{lineno}
@@
 \usepackage{natbib}
+\usepackage[T1]{fontenc}
+\usepackage[breaklinks]{hyperref}
+\usepackage{microtype}
 
@@
 
+\hbadness=99999
 \begin{document}
@@
 
-\noindent Genetic kinship of ancient individuals can provide insights into their culture and social hierarchy, and is relevant for downstream genetic analyses.
-However, estimating relatedness from ancient DNA is difficult due to low-coverage, ascertainment bias, or contamination from various sources.
-Here, we present KIN, a method to estimate the relatedness of a pair of individuals from the identical-by-descent segments they share.
-KIN accurately classifies up to 3rd-degree relatives using $>0.05x$ sequence coverage and differentiates siblings from parent-child.
-It incorporates additional models to adjust for contamination and detect inbreeding, which improves classification accuracy.
+\noindent Genetic kinship of ancient individuals can provide insights into their culture and social
+hierarchy, and is relevant for downstream genetic analyses. However, estimating
+relatedness from ancient DNA is difficult due to low-coverage, ascertainment bias, or
+contamination from various sources. Here, we present KIN, a method to estimate the
+relatedness of a pair of individuals from the identical-by-descent segments they share.
+KIN accurately classifies up to 3rd-degree relatives using at least 0.05x sequence coverage
+and differentiates siblings from parent-child. It incorporates additional models to adjust
+for contamination and detect inbreeding, which improves classification accuracy.
 \end{abstract}
@@
 
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib_rev.png}
+\begin{figure}[ht!]
+    \includegraphics[width=16.5cm]{plots/inkscape_finalImg/schematic_sib_rev.png}
     \centering
@@
 In both cases, $p$ can take values of $p_0$, $p_1$ or $p_2$, which are the expected proportion of differences in IBD states 0,1 and 2, respectively.
-(D-F): Same as A-C, except parent 1 is assumed to be homozygous.
-}
+(D-F): Same as A-C, except parent 1 is assumed to be homozygous.}
     \label{fig0:schematic}
@@
 
-\begin{table}[h!]
+\begin{table}[ht!]
 \caption{\label{tab:Table 1}IBD sharing probabilities for different relations in absence of inbreeding}
@@
 
-The effect of these considerations is that even though we have nine possible combinations of $Z_w$ and $H_w$ for each window, there are actually only four unique $p$-parameters $p_i$ with $i$ $\in$ (0,1,2,3).
+The effect of these considerations is that even though we have nine possible combinations of $Z_w$ and $H_w$ for each window, there are actually only four unique $p$-parameters $p_i$ with $i$ $\in$ (0,1,2,4).
 
@@
 
-\subsection{Estimation of $\delta$}\label{delta}
+\subsection{Estimation of \texorpdfstring{$\delta$}{}}\label{delta}
 We estimate the $\delta$-parameters using an Expectation-Maximization (EM) algorithm \cite{dempster_maximum_1977}.
@@
 
-\begin{align*}\label{eq:5}
-g_{w0} &= \gamma_{w0}\\
-g_{w1} &= \gamma_{w1} h_{w0}\\
-g_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\\
-g_{w3} &= 0 \\
+\begin{align}\label{eq:5}
+g_{w0} &= \gamma_{w0}\nonumber\\
+g_{w1} &= \gamma_{w1} h_{w0}\nonumber\\
+g_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\nonumber\\
+g_{w3} &= 0 \nonumber\\
 g_{w4} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2},
-\end{align*}
+\end{align}
 
@@
 
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/inkscape_finalImg/IBDplot_title.png}
+\begin{figure}[ht!]
+    \includegraphics[width=16.5cm]{plots/inkscape_finalImg/IBDplot_title.png}
     \centering
@@
 We calculate the emissions using a beta-binomial likelihood, and fix the mean of the distributions corresponding to $Y_w$ = 4 and $Y_w$ = 2 at expected proportion of differences in a homozygous tract ($p_4$) and expected proportion of differences in a non-homozygous tract ($p_2$) respectively.
-The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w=4$ or the state $Y_w=2$ in each window.
+The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w = 4$ or the state $Y_w = 2$ in each window.
 The maximization step for emissions is analogous to that in the KIN model (eq.
@@
 To estimate transitions, we initialize the transition matrix with the value 0.2 for the off-diagonal entries, and update it using the standard Baum-Welch update step \cite{baum_maximization_1970}.
-
-Similar to the KIN-HMM, we avoid fitting issues by forcing all windows whose proportion of differences is larger than $p_2$ to be in the non-homozygous state (Fig. S7).
+Similar to the KIN-HMM, we avoid fitting issues by forcing all windows whose proportion of differences is larger than $p_2$ to be in the non-homozygous state (Fig.
+S7).
 
@@
 
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot_review.png}
+\begin{figure}[ht!]
+    \includegraphics[width=16.5cm]{pl}
     \centering
@@
 
-\begin{table}[h!]
+\begin{table}[ht!]
 \caption{\label{tab:Table 2}Model performance for ROH prediction.
@@
 The exceptions are for identical individuals and parent-child pairs, where the accuracy is nearly perfect for all investigated coverages, even at 0.03x coverage.
+We note that the accuracy of IBD prediction is lowest for siblings, followed by 2nd degree and 3rd degree relatives.
+This is because IBD in siblings is more variable, and therefore harder to predict.
+A pair of 3rd-degree relatives, for example, are expected to only share $12.5\%$ of their genome IBD, with the rest being unrelated.
+Therefore, even a naive classifier that classified everything as unrelated would have an accuracy of $87.5\%$.
+In contrast, siblings are expected to share one chromosome IBD for half their genomes, and both chromosomes for another $25\%$.
+This higher variability makes predictions harder.
+
 We find that adding contamination, ascertainment bias in our simulations has little effect on IBD prediction.
@@
 
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/plotimg/plot_IBDaccuracy.png}
+\begin{figure}[ht!]
+    \includegraphics[width=16.5cm]{plot_IBDaccuracy.png}
     \centering
@@
 
-\begin{figure}[h!]
-    \includegraphics[width=16cm]{plots/plotimg/comparison_plot.png}
+\begin{figure}[ht!]
+    \includegraphics[width=16.5cm]{comparison_plot.png}
     \centering
@@
 \paragraph{Chagyrskaya and Okladnikov Neandertals:}
-To test KIN on real ancient data, we applied it to a Neandertal dataset from Chagyrskaya and Okladnikov Caves in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020,laurits_skov_genetic_nodate}.
-This dataset contains genetic data from a total of 16 skeletal remains that likely belong to contemporary Neandertals who occupied the Chagyrskaya and Okladnikov caves between 59 and 51 kya, and at least 44 kya respectively \cite{laurits_skov_genetic_nodate}.
-DNA extracted from each of these remains were captured with an array targeting variable sites identified in high coverage Neandertal and Denisovan genomes, and common variations in Africans \cite{laurits_skov_genetic_nodate}.
+To test KIN on real ancient data, we applied it to a Neandertal dataset from Chagyrskaya and Okladnikov Caves in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020,skov_genetic_2022}.
+This dataset contains genetic data from a total of 16 skeletal remains that likely belong to contemporary Neandertals who occupied the Chagyrskaya and Okladnikov caves between 59 and 51 kya, and at least 44 kya respectively \cite{skov_genetic_2022}.
+DNA extracted from each of these remains were captured with an array targeting variable sites identified in high coverage Neandertal and Denisovan genomes, and common variations in Africans \cite{skov_genetic_2022}.
 This genetic data has  low-to-intermediate depth of coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage.
-Some of these specimens showed signs of long ROH, and DNA contamination from modern humans as well as hyenas \cite{laurits_skov_genetic_nodate}.
-We focused our analysis on the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017} as done by the authors for the relatedness analysis \cite{laurits_skov_genetic_nodate}.
+Some of these specimens showed signs of long ROH, and DNA contamination from modern humans as well as hyenas \cite{skov_genetic_2022}.
+We focused our analysis on the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017} as done by the authors for the relatedness analysis \cite{skov_genetic_2022}.
 Our results for pairwise relatedness for these individuals are shown in Fig.\ref{fig5:Chagyrskaya_KIN}.
@@
 We also find that both KIN and READ classify Chagyrskaya06 and Chagyrskaya14 as parent-child with low confidence, but from the morphology they likely stem from the same individual.
-We believe that the low confidence mis-classification may be due to uncorrected non-human contamination present in these libraries ($1\%$ and $2.9\%$ respectively \cite{laurits_skov_genetic_nodate}), biasing the estimated differences between the individuals to higher values.
+We believe that the low confidence mis-classification may be due to uncorrected non-human contamination present in these libraries ($1\%$ and $2.9\%$ respectively \cite{skov_genetic_2022}), biasing the estimated differences between the individuals to higher values.
 
@@
 However, the location of these clusters strongly deviate from the expected values, likely due to the low coverage and ROH.
-Remains from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $k_0 \approx 0.3$, $r \approx 0.6$ for lcMLkin (Figure S10).
+Remains from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $k_0 \approx 0.3$, $r \approx 0.6$ for lcMLkin (Fig.
+S10).
 
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/kinplot.png}
+\begin{figure}[ht!]
+    \includegraphics[width=16.5cm]{kinplot.png}
     \centering
@@
 The amount of contamination does not need to be large for this to be important, in our simulations we find that even at contamination levels $\leq 3\%$, the performance of READ is substantially reduced.
-When contamination rates are correctly inferred, the correction we implemented leads to improved performance compared to naive methods such as READ, although they too could be ameliorated in a similar way \cite{laurits_skov_genetic_nodate}.
+When contamination rates are correctly inferred, the correction we implemented leads to improved performance compared to naive methods such as READ, although they too could be ameliorated in a similar way \cite{skov_genetic_2022}.
 However, in many cases contamination estimates may be uncertain or inaccurate, and we show in fig.
@@
 KIN can also infer third degree relation with $\approx$ 30,000 overlapping polymorphic sites.
-We show that when applied to Neandertal specimens from Chagyrskaya and Okladnikov Caves, KIN identifies a pair of $1^{st}$-Degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples (one sample is male and the other female) \cite{laurits_skov_genetic_nodate}.
+We show that when applied to Neandertal specimens from Chagyrskaya and Okladnikov Caves, KIN identifies a pair of $1^{st}$-Degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples (one sample is male and the other female) \cite{skov_genetic_2022}.
 In addition, KIN identifies a pair of $3^{rd}$ degree relatives.
@@
 We simulate unrelated individuals using $\texttt{msprime}$ \cite{kelleher_efficient_2016}, followed by an additional step where we simulate related individuals using a predetermined pedigree (Fig.
-S1).
+S3).
 
@@
 
-For our simulations of background diversity, we form a population (Pop1) with constant effective size of 10,000 and sample eight diploid individuals (each made up of two haploid individuals) from 2500 generations ago (fig.
-S1).
+For our simulations of background diversity, we form a population (Pop1) with constant effective size of 10,000 and sample eight diploid individuals (each made up of two haploid individuals) from 2500 generations ago (Fig.
+S3).
 For each individual, we simulate 22 chromosomes with length $L\approx96$ Mb (same as chromosome 13) and a recombination rate of $r=10^{-8}$ per base pair per generation.
@@
 \section{Acknowledgements}
-We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Johann Visagie, Zbigniew Jędrzejewski-Szmek, Laurits Skov, Leonardo N.
+We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Johann Visagie, Zbigniew \linebreak Jędrzejewski-Szmek, Laurits Skov, Leonardo N.
 M.
@@
 We have deposited the version of the software used in the manuscript, along with the above mentioned files on Zenodo \cite{divyaratan_popli_httpsdoiorg105281zenodo7067142_nodate}.
-The analysed datasets from Bronze Age Lech Valley were generated in a previous study (ENA, accession no.
-PRJEB34400).
-Chagyrskaya and Okladnikov dataset were generated from a study currently in press \cite{laurits_skov_genetic_nodate}, and the data will be uploaded to European Nucleotide Archive upon publication.
+The analysed datasets were generated in previous studies, and are available in European Nucleotide Archive \cite{mittnik_kinship-based_nodate,skov_genetic_nodate}.
 

### 2022-11-17 plots/Rscripts/IBD_accuracyPlot.R
changed colors

@@ -42,7 +42,7 @@ IBD_accuracy_plot <- function(infile,outfile){
 
   df$cov=as.factor(df$cov)
 
-  colors1 <- c("Unrelated" = "grey30", "3rd Degree" = "blue", "2nd Degree" = "darkorchid1", "Siblings"= "Green", "Parent-Child" = "orangered3", "Identical" = "darkgoldenrod1")
+  colors1 <- c("Unrelated" = "#999999", "3rd Degree" = "#E69F00", "2nd Degree" = "#009E73", "Siblings"= "#0072B2", "Parent-Child" = "#D55E00", "Identical" = "#CC79A7")
 
   f <- ggplot(
     df,

### 2022-11-17 plots/Rscripts/comparisonPlot.R
colors

@@ -43,11 +43,14 @@ comparison_plot <- function(inf,outf){
   T_F.labs <- c("False Positive", "True Positive")
   names(T_F.labs) <- c("False_positive", "True_positive")
 
+  colors1 <- c("Control" = "#999999", "With ROH" = "#E69F00", "With Ascertainment" = "#D55E00", "With Contamination"= "#0072B2")
+
 
   plot<-ggplot(data=data, aes(x=Coverage, y=classification_proportion, group=interaction(Simulation,Method), color=Simulation, lty=Method))
   plot<-plot+geom_line()
-  plot<- plot + facet_grid(T_F ~ Relatedness, labeller = labeller(T_F = T_F.labs, Relatedness = Relatedness.labs), scales="free")
+  plot<- plot + facet_grid(T_F ~ Relatedness, labeller = labeller(T_F = T_F.labs, Relatedness = Relatedness.labs), scales="free") +
   #plot + geom_hline(yintercept=0.05, linetype="dashed")
+  scale_color_manual(values = colors1)
   plot <- plot + geom_hline(data = data.frame(T_F='False_positive'), aes(yintercept = 0.05), linetype = "dotted", alpha=0.5)
   plot <- plot+ labs(y="Classification Proportions", x = "Coverage")
   plot + theme_bw() + theme(

### 2022-11-17 plots/Rscripts/kinPlot.R
colors

@@ -38,7 +38,7 @@ bubbleplot_kin <- function(overlap_table, relatable, out_rel, out_over) {
   matr1$lib2 <- factor(matr1$lib2, levels=c("Chagyrskaya01","Chagyrskaya06","Chagyrskaya14","Chagyrskaya12","Chagyrskaya41","Chagyrskaya20","Chagyrskaya1141","Chagyrskaya13","Chagyrskaya19","Chagyrskaya07","Chagyrskaya17","Chagyrskaya18","Chagyrskaya02","Chagyrskaya60","Okladnikov11","Okladnikov15"))#data$SP_V1 <- ifelse(data$V1 == “A29253”, “red”, “black”)
   matr1$lib1 <- factor(matr1$lib1, levels=c("Chagyrskaya01","Chagyrskaya06","Chagyrskaya14","Chagyrskaya12","Chagyrskaya41","Chagyrskaya20","Chagyrskaya1141","Chagyrskaya13","Chagyrskaya19","Chagyrskaya07","Chagyrskaya17","Chagyrskaya18","Chagyrskaya02","Chagyrskaya60","Okladnikov11","Okladnikov15"))#data$SP_V1 <- ifelse(data$V1 == “A29253”, “red”, “black”)
 
-  colors <- c("seagreen4","steelblue2","mediumblue","steelblue3","lightpink3","thistle3","gray45","lightsalmon3","darkred","firebrick","tan","black","gray70","lightcyan3","darkorchid","pink3","pink4")#data[, c(1, 2)] <- as.data.frame(t(apply(data[, c(1, 2)], 1, sort)))#Plot all the infered relatedness
+  colors <- c("black","black","black","black","black","black","black","black","black","black","black","black","black","black","black","black","black")#data[, c(1, 2)] <- as.data.frame(t(apply(data[, c(1, 2)], 1, sort)))#Plot all the infered relatedness
 
   ##########trying heatmap.2
   #i1=unique(matr1$lib1)
@@ -57,7 +57,8 @@ bubbleplot_kin <- function(overlap_table, relatable, out_rel, out_over) {
 
   gfig_rel=ggplot(matr1, aes(lib1, lib2), col= "white") +
     geom_tile(aes(fill=relatedness), col="white")  +
-    scale_fill_manual(values = c("Identical"="khaki1","Parent-Child"="orangered3", "Siblings"="orange2","Unrelated"="gray80","Second Degree"="darkorchid1","Third Degree"="aquamarine2")) +
[…]

### 2022-11-17 supplementary_info/Rscripts/readPlot.R
color

@@ -46,7 +46,7 @@ bubbleplot_read <- function(overlap_table, relatable, out_rel, out_over) {
   matr1$lib2 <- factor(matr1$lib2, levels=c("Chagyrskaya01","Chagyrskaya06","Chagyrskaya14","Chagyrskaya12","Chagyrskaya41","Chagyrskaya20","Chagyrskaya1141","Chagyrskaya13","Chagyrskaya19","Chagyrskaya07","Chagyrskaya17","Chagyrskaya18","Chagyrskaya02","Chagyrskaya60","Okladnikov11","Okladnikov15"))#data$SP_V1 <- ifelse(data$V1 == “A29253”, “red”, “black”)
   matr1$lib1 <- factor(matr1$lib1, levels=c("Chagyrskaya01","Chagyrskaya06","Chagyrskaya14","Chagyrskaya12","Chagyrskaya41","Chagyrskaya20","Chagyrskaya1141","Chagyrskaya13","Chagyrskaya19","Chagyrskaya07","Chagyrskaya17","Chagyrskaya18","Chagyrskaya02","Chagyrskaya60","Okladnikov11","Okladnikov15"))#data$SP_V1 <- ifelse(data$V1 == “A29253”, “red”, “black”)
 
-  colors <- c("seagreen4","steelblue2","mediumblue","steelblue3","lightpink3","thistle3","gray45","lightsalmon3","darkred","firebrick","tan","black","gray70","lightcyan3","darkorchid","pink3","pink4")#data[, c(1, 2)] <- as.data.frame(t(apply(data[, c(1, 2)], 1, sort)))#Plot all the infered relatedness
+  colors <- c("black","black","black","black","black","black","black","black","black","black","black","black","black","black","black","black","black")#data[, c(1, 2)] <- as.data.frame(t(apply(data[, c(1, 2)], 1, sort)))#Plot all the infered relatedness
 
   ##########trying heatmap.2
   #i1=unique(matr1$lib1)
@@ -65,7 +65,7 @@ bubbleplot_read <- function(overlap_table, relatable, out_rel, out_over) {
 
   gfig_rel=ggplot(matr1, aes(lib1, lib2), col= "white") +
     geom_tile(aes(fill=relatedness), col="white")  +
-    scale_fill_manual(values = c("khaki1","orangered3","mediumorchid4","gray80")) +
+    scale_fill_manual(values = c("Unrelated" = "#999999", "Second Degree" = "#009E73", "First Degree" = "#D55E00", "Identical" = "#CC79A7")) +
[…]

### 2022-11-17 supplementary_info/beta_bnds.py
colors

@@ -639,22 +639,22 @@ def plot_betas(dfile, tfile, pair, thresh, rel, libraries, listf, Afile, pfile, 
 
         a=betas[2]
         b=betas[3]
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'r-', lw=3, alpha=0.6, label="$Z_w$=0")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'#377eb8', lw=3, alpha=0.6, label="$Z_w$=0")
         #plt.hist(diff/total,alpha=0.3)
 
         a=betas[0]
         b=betas[1]
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'b-', lw=3, alpha=0.6, label="$Z_w$=1")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'#ff7f00', lw=3, alpha=0.6, label="$Z_w$=1")
 
         a=betas[4]
         b=betas[5]
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'g-', lw=3, alpha=0.6, label="$Z_w$=2")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'#4daf4a', lw=3, alpha=0.6, label="$Z_w$=2")
         #plt.hist(diff/total,alpha=0.3)
 
         a=betas[6]
         b=betas[7]
 
-        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'y-', lw=3, alpha=0.6, label="$Z_w$=4")
+        ax[i].plot(x, beta.pdf(x, a, b)/533.33,'#f781bf', lw=3, alpha=0.6, label="$Z_w$=4")
         ax[i].hist(diff/total,alpha=0.3,weights=np.ones(len(total)) / len(total))
         ax[i].set_ylim(0,0.4)
 

### 2022-11-17 plots/Rscripts/cutoffPlot.R
color

@@ -31,9 +31,12 @@ cutoff_plot <- function(infile,outfile){
   T_F.labs <- c("False Positive", "True Positive")
   names(T_F.labs) <- c("False_positive", "True_positive")
 
+  colors1 <- c("0.1"="#999933", "0.2"="#117733","4"="#661100", "0.05"="#0072B2")
+
   plot<-ggplot(data=data, aes(x=cutoff, y=classification_proportion, color=Coverage))
   plot<-plot+geom_line()
-  plot<- plot + facet_grid(T_F ~ Relatedness, labeller = labeller(T_F = T_F.labs, Relatedness = Relatedness.labs), scales="free")
+  plot<- plot + facet_grid(T_F ~ Relatedness, labeller = labeller(T_F = T_F.labs, Relatedness = Relatedness.labs), scales="free") +
+  scale_color_manual(values = colors1)
   #plot + geom_hline(yintercept=0.05, linetype="dashed")
   plot <- plot + geom_hline(data = data.frame(T_F='False_positive'), aes(yintercept = 0.05), linetype = "dotted")
   plot <- plot+ labs(y="Classification Proportions", x = "Log likelihood Ratio Cutoff")

### 2022-11-18 plots/Rscripts/supp_IBD.R
color

@@ -17,31 +17,31 @@ IBD_accuracy_plot <- function(infile1,infile2,infile3,infile4,outfile){
       infile=infile4
     }
     data=read.csv(file = infile, header = TRUE)
-    
-    
+
+
     data[data$rel=='deg4','rel']='Unrelated'
     data[data$rel=='deg5','rel']='Unrelated'
     data[data$rel=='un','rel']='Unrelated'
-    
+
     data[data$rel=='avu','rel']='2nd Degree'
     data[data$rel=='gr','rel']='2nd Degree'
     data[data$rel=='hsib','rel']='2nd Degree'
-    
+
     data[data$rel=='id','rel']='Identical'
     data[data$rel=='pc','rel']='Parent-Child'
     data[data$rel=='sib','rel']='Siblings'
-    
+
     data[data$rel=='deg3','rel']='3rd Degree'
-    
+
     allrel=c("Unrelated", "3rd Degree", "2nd Degree", "Siblings", "Parent-Child", "Identical")
     allcov=c(4,0.5,0.2,0.1,0.05,0.03)
-    
+
     accuracy=c()
     stdev=c()
     rel=c()
     cov=c()
     for(cv in allcov) {
-      
+
       datacv=data[data$cov==cv,]
       for(i in allrel) {
         temp=datacv[datacv$rel==i,]
@@ -52,22 +52,22 @@ IBD_accuracy_plot <- function(infile1,infile2,infile3,infile4,outfile){
       }
     }
     df=data.frame(accuracy,stdev,rel,cov)
-    
+
     df$cov=as.factor(df$cov)
     df$Scenario=j
     bigd=rbind(bigd,df)
   }
-  
+
   #bigd_un=bigd[bigd$rel=='Siblings',]
   #bigd_un_control=bigd_un[bigd_un$Scenario==1,]
   #bigd_un_c=bigd_un[bigd_un$Scenario==2,]
   #bigd_un_a=bigd_un[bigd_un$Scenario==3,]
   #bigd_un_r=bigd_un[bigd_un$Scenario==4,]
-  
+
   bigd$Scenario=as.factor(bigd$Scenario)
-  colors1 <- c("Unrelated" = "grey30", "3rd Degree" = "blue", "2nd Degree" = "darkorchid1", "Siblings"= "Green", "Parent-Child" = "orangered3", "Identical" = "darkgoldenrod1")
[…]

### 2022-11-18 paper_overleaf/KIN_a_method_to_estimate_relatedness_from_low_coverage_data_supplements.tex
Updates from Overleaf

@@
 \renewcommand{\figurename}{Fig. S}
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic1.png}
+\begin{figure}[ht!]
+    \includegraphics[width=16cm]{plots/inkscape_finalImg/schematic1.png}
     \centering
@@
 
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/binom.png}
+\begin{figure}[ht!]
+    \includegraphics[width=16cm]{supplementary_info/plots/binom.png}
     \centering
@@
 
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{plots/inkscape_finalImg/pedigree_both.png}
+\begin{figure}[ht!]
+    \includegraphics[width=16cm]{plots/inkscape_finalImg/pedigree_both.png}
     \centering
@@
 
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/fil0_read_plot.png}
+\begin{figure}[ht!]
+    \includegraphics[width=16cm]{supplementary_info/plots/readplot_sub.png}
     \centering
@@
 
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_emissions_bnds.png}
+\begin{figure}[ht!]
+    \includegraphics[width=16cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_emissions_bnds.png}
     \centering
@@
 
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_betaplot1.png}
+\begin{figure}[ht!]
+    \includegraphics[width=16cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_betaplot1.png}
     \centering
@@
 
-\begin{figure}[h!]
-    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_ind0_forced_roh.png}
+\begin{figure}[ht!]
+    \includegraphics[width=16cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_ind0_forced_roh.png}
      \centering
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \includegraphics[width=16cm]{plots/plotimg/contam0_inbred0_model_performance_allroc_asc0_plot.png}
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \centering
-    \includegraphics[width=18cm]{plots/plotimg/plot_IBDaccuracy_others_supp.png}
+    \includegraphics[width=16cm]{plots/plotimg/plot_IBDaccuracy_others_supp.png}
     \caption{Comparison of IBD estimation in control simulations and in presence of contamination, ascertainment or ROH (denoted in figure legend by Control, C, A and R, respectively) at different coverages.
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/lcPlot.png}
+    \includegraphics[width=16cm]{supplementary_info/plots/lcPlot.png}
     \caption{Comparison of IBD states estimated for Chagyrskaya specimens using (A) lcMLkin and (B) KIN.
@@
     \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/egplot1.png}
+    \includegraphics[width=16cm]{supplementary_info/plots/egplot1.png}
     \caption{Plots showing proportion of differences in windows along the genome for some pairs of relatives for which there is contradiction among KIN, READ and lcMLkin.
@@
     \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/egplot2.png}
+    \includegraphics[width=16cm]{supplementary_info/plots/egplot2.png}
     \caption{Plots showing proportion of differences in windows along the genome for the first degree relatives for which KIN differs from lcMLkin.
@@
     \centering
-    \includegraphics[width=18cm]{supplementary_info/plots/contam_missspecification.png}
+    \includegraphics[width=16cm]{supplementary_info/plots/contam_missspecification.png}
     \caption{False positive and true positive rates as a function of incorrect contamination estimates provided to the ROH-HMM.

### 2022-11-18 supplementary_info/Rscripts/lcPlot.R
color change

@@ -36,7 +36,8 @@ lcplot <- function(lcres,likf,kinf,outplot){
 
   lc=lc[lc$pair!='Chagyrskaya06_Chagyrskaya14',]
 
-  colors1 <- c("Unrelated" = "grey80", "Degree2" = "darkorchid1", "Parent-Child" = "orangered3", "Identical" = "darkgoldenrod1")
+  #colors1 <- c("Unrelated" = "grey80", "Degree2" = "darkorchid1", "Parent-Child" = "orangered3", "Identical" = "darkgoldenrod1")
+  colors1 <- c("Unrelated" = "#999999", "Degree2" = "#009E73", "Parent-Child" = "#D55E00", "Identical" = "#CC79A7")
 
   panel1<-ggplot(data=lc[which(lc$loglik_ratio >=1),], mapping=aes(x=k0_hat, y=r_hat, color=Relatedness)) +
   geom_point() + theme_bw() + labs(x=expression(k["0"]), y= "Coefficient of Relatedness") +
@@ -67,7 +68,7 @@ lcplot <- function(lcres,likf,kinf,outplot){
 
   #newdf[newdf$loglik_ratio<1,]=NA
 
-  colors <- c("Unrelated" = "grey80", "Degree2" = "darkorchid1", "Parent-Child" = "orangered3", "Identical" = "darkgoldenrod1")
+  colors <- c("Unrelated" = "#999999", "Degree2" = "#009E73", "Parent-Child" = "#D55E00", "Identical" = "#CC79A7")
   #alphas <-c("0"=0.01,"1"=1)
   panel2<-ggplot(data=newdf[which(newdf$loglik_ratio >=1),], mapping=aes(x=k0_prop, y=r_prop, color=Relatedness)) +
     geom_point() + theme_bw() + labs(x=expression(k["0"]), y= "Coefficient of Relatedness") +

### 2022-11-18 supplementary_info/Rscripts/misscontam_plot.R
color change

@@ -1,0 +1,56 @@
+
+library(ggplot2)
+#install.packages("magrittr")
+#install.packages("dplyr")
+library(magrittr)
+library(dplyr)
+library(tidyr)
+
+#inf='/mnt/diversity/divyaratan_popli/100arc/inbreeding/fastsim_gz/hmm_numba_incorrect_roh_i1_i1_insteadOf_i1_i2/roc/contam0_inbred0_model_performance_allroc_asc0.csv.gz'
+#infile='/mnt/diversity/divyaratan_popli/review_sim/hmm_numba/miss_contam/output/missC10/contam1_inbred0_model_performance_allLikelihoods_inphapProbs_asc0.csv'
+
+contamination_plot <- function(infile,outfile){
+
+
+  compinp=read.csv(file = infile, header = TRUE)
+  compinp$Relatedness <- factor(compinp$Relatedness, levels = c("fir", "sec", 'un_all','pc','sib','deg3'))
+
+  data <- compinp %>%
+    pivot_longer(c(True_positive,False_positive), names_to = "T_F", values_to = "classification_proportion")#Plot
+  data$T_F <- factor(data$T_F, levels= c('True_positive','False_positive'))
+
+  #data<-data[!(data$method=="read_inppshap" | data$method=="allLikelihoods_inppshap"),]
+  #data = subset(data, select = -c(method) )
+
+  data$Coverage <- as.factor(data$Coverage)
+
+  #changing some column names:
+
+  Relatedness.labs <- c("Unrelated", "3rd Degree", "2nd Degree", "1st Degree", "Siblings", "Parent-Child")
+  names(Relatedness.labs) <- c("un_all", "deg3", "sec", "fir", "sib", "pc")
+
+  T_F.labs <- c("False Positive", "True Positive")
+  names(T_F.labs) <- c("False_positive", "True_positive")
+
+  colors1 <- c("0.1"="#999933", "0.2"="#117733","4"="#661100", "0.05"="#0072B2", "0.5"="#E69500")
+
+  plot<-ggplot(data=data, aes(x=contamination, y=classification_proportion, color=Coverage))
+  plot<-plot+geom_line()
+  plot<- plot + facet_grid(T_F ~ Relatedness, labeller = labeller(T_F = T_F.labs, Relatedness = Relatedness.labs), scales="free") +
+  scale_color_manual(values = colors1)
+  #plot + geom_hline(yintercept=0.05, linetype="dashed")
[…]
